### 2011-07-11 conditional/trunk/selectivity-hexastore.tex
Just a start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@2 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass[12pt]{article}
\usepackage{cite}
\title{Conditional selectivity for Hexastores}

\author{Kjetil Kjernsmo}

\begin{document}
\maketitle

\begin{abstract}
In this paper we see how the common assumption of independence of the
selectivity of RDF nodes can be abandoned. We shall also find it
instructive to explore how selectivity can be determined exactly by
using index size for a Hexastore triple store. Finally, we will
suggest further directions for selectivity estimation in the
conditional regime.

\end{abstract}

\section{Introduction}
Several authors have studied the use of selectivity estimation for
optimizing SPARQL query execution plans, both in the context of Basic
Graph Patterns \cite{Stocker:2008:SBG:1367497.1367578}.

\bibliography{selectivity}{}
\bibliographystyle{plain}

\end{document}

### 2011-07-11 conditional/trunk/selectivity-hexastore.tex
Conditional sel
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@3 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{cite}
+
+\newcommand{\rdfnode}[1]{\texttt{#1}}
+
 \title{Conditional selectivity for Hexastores}
@@
 selectivity of RDF nodes can be abandoned. We shall also find it
-instructive to explore how selectivity can be determined exactly by
-using index size for a Hexastore triple store. Finally, we will
-suggest further directions for selectivity estimation in the
-conditional regime.
+instructive to explore how selectivity can be determined exactly when
+a SPARQL query is executed on a Hexastore triple store
+\cite{Weiss:2008:HSI:1453856.1453965}. Finally, we will suggest
+further directions for selectivity estimation in the conditional
+regime.
 
@@
 \section{Introduction}
+
 Several authors have studied the use of selectivity estimation for
 optimizing SPARQL query execution plans, both in the context of Basic
-Graph Patterns \cite{Stocker:2008:SBG:1367497.1367578}.
+Graph Patterns \cite{Stocker:2008:SBG:1367497.1367578} and in more
+general cases. % TODO more references. OptARQ?
 
-\bibliography{selectivity}{}
+The above authors assume that the selectivity of RDF nodes are
+independent, admit that this assumption is easily violated and
+continue to use a conditional estimate for the object node.
+
+It is easy to point out cases where the independence assumption is
+unlikely to hold: If a predicate is \rdfnode{rdf:type}, then only
+classes are likely objects, and their frequency  would likely be
+different from e.g. literals.
+
+\section{Conditional selectivity}
+
+Previous authors have taken selectivity as analogous to the
+probability of matching a SPARQL node or triple with a node or triple
+in the data. For a triple, $<S, P, O>$ the selectivity is the
+intersection of the selectivity of each node.
+
+From elementary statistics, we know that the intersection probability
+can be written as
+\begin{equation}
+Pr(A_1 \cap A_2  \cap \ldots \cap A_n) = 
+  Pr(A_n | A_1 \cap A_2  \cap \ldots \cap A_{n-1}) \\
+  \cdot Pr(A_{n-1} | A_1 \cap A_2  \cap \ldots \cap A_{n-2}) 
+  \cdot \ldots \cdot Pr(A_2 | A_1) \cdot Pr(A_1) ,
+\end{equation}
+where the $|$ symbol is read as ``given'', e.g. ``given a predicate
+\rdfnode{rdf:type} what is the probability that the object is \rdfnode{foaf:Person}?''
+
+Analogously, the conditional selectivity for a triple $T$ can be written as 
+\begin{equation}
+sel(T) = sel(S \cap P \cap O) = sel(O | S \cap P) \cdot sel(P | S)
+\cdot sel(S) . 
+\end{equation}
+
 \bibliographystyle{plain}
+\bibliography{selectivity}
 

### 2011-07-11 conditional/trunk/selectivity-hexastore.tex
commutativity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@4 43ef1190-90fd-4dd7-a769-d0121f02b80e
unit for unbound
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@5 43ef1190-90fd-4dd7-a769-d0121f02b80e
Hexastore start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@6 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-\documentclass[12pt]{article}
+\documentclass[12pt, draft]{article}
 \usepackage{cite}
@@
 \end{equation}
+We note that intersection is commutative, thus similar
+relations can be shown for each permutation of nodes.
+
+Finally, note that for unbound nodes (variables), the selectivity will
+always be unity, whether conditional or independent, since it will
+match all possible RDF nodes.
+
+% TODO example?
+
+\section{Exact selectivity for Hexastore}
+
+The Hexastore was first described in
+\cite{Weiss:2008:HSI:1453856.1453965}, which sought to exploit a key
+strength of the RDF triple model by using six-way indices for
+efficient lookup.
+
+A Hexastore contains the six indices \textsf{spo}, \textsf{pso},
+\textsf{osp}, \textsf{sop}, \textsf{pos} and \textsf{ops}. Each index
+has the following structure, taking \textsf{spo} as an example: ``A
+subject key $s_i$ is associated to a sorted vector of $n_i$ property
+keys, $\{p_1^i , p_2^i , \ldots , p^i_{n_i} \}$. Each property key $p_j^i$
+is, in its turn, linked to an associated sorted list of $k_{i,j}$
+object keys.''\cite{Weiss:2008:HSI:1453856.1453965}, section~4.1.
+
+Now, given a query, we shall explain how the selectivity can be
+computed exactly rather than being estimated. First, we define a
+function $cm(x)$ which returns the number of matches of a given SPARQL
+node $x$. % TODO: needs to be a lot more precise
 

### 2011-07-11 conditional/trunk/selectivity-hexastore.tex
TODOs and status
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@7 43ef1190-90fd-4dd7-a769-d0121f02b80e
Add more todos and sel-equations
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@8 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\rdfnode}[1]{\texttt{#1}}
+\newcommand{\todo}[1]{\tiny{TODO}\normalsize\footnote{\textbf{TODO:}~#1}}
 
@@
 \maketitle
+
+\section*{Status of this document}
+
+This is currently at braindump status. Do not distribute without the
+permission of the author.
 
@@
 Graph Patterns \cite{Stocker:2008:SBG:1367497.1367578} and in more
-general cases. % TODO more references. OptARQ?
+general cases. \todo{more references. OptARQ?}
 
@@
 Previous authors have taken selectivity as analogous to the
-probability of matching a SPARQL node or triple with a node or triple
+probability of matching a SPARQL node\todo{terminology, (SPARQL nodes
+  vs. RDF nodes) could be better} or triple with a node or triple
 in the data. For a triple, $<S, P, O>$ the selectivity is the
@@
 
-Analogously, the conditional selectivity for a triple $T$ can be written as 
+Analogously\todo{anybody good at arguing this mathematically?}, the conditional selectivity for a triple $T$ can be written as 
 \begin{equation}
@@
 
-% TODO example?
+\todo{example?}
 
@@
 function $cm(x)$ which returns the number of matches of a given SPARQL
-node $x$. % TODO: needs to be a lot more precise
+node $x$. \todo{needs to be a lot more precise}
+
+Then, 
+\begin{equation}
+sel(S) = \frac{cm(s_i)}{i} ,
+\end{equation}
+\begin{equation}
+sel(P) = \frac{cm(p_{n_i}^i)}{n_i}
+\end{equation}
+and
+\begin{equation}
+sel(O) = \frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}}
+\end{equation}
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
correct left-hand
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@9 43ef1190-90fd-4dd7-a769-d0121f02b80e
Improve TODO command
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@10 43ef1190-90fd-4dd7-a769-d0121f02b80e
RDF node -> RDF term
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@13 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\rdfnode}[1]{\texttt{#1}}
-\newcommand{\todo}[1]{\tiny{TODO}\normalsize\footnote{\textbf{TODO:}~#1}}
+\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 In this paper we see how the common assumption of independence of the
-selectivity of RDF nodes can be abandoned. We shall also find it
+selectivity of RDF terms can be abandoned. We shall also find it
 instructive to explore how selectivity can be determined exactly when
@@
 
-The above authors assume that the selectivity of RDF nodes are
+The above authors assume that the selectivity of RDF terms are
 independent, admit that this assumption is easily violated and
@@
 probability of matching a SPARQL node\todo{terminology, (SPARQL nodes
-  vs. RDF nodes) could be better} or triple with a node or triple
+  vs. RDF terms) could be better} or triple with a node or triple
 in the data. For a triple, $<S, P, O>$ the selectivity is the
@@
 always be unity, whether conditional or independent, since it will
-match all possible RDF nodes.
+match all possible RDF terms.
 
@@
 \begin{equation}
-sel(P) = \frac{cm(p_{n_i}^i)}{n_i}
+sel(P|S) = \frac{cm(p_{n_i}^i)}{n_i}
 \end{equation}
@@
 \begin{equation}
-sel(O) = \frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}}
+sel(O|S \cap P) = \frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}}
 \end{equation}

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
single triple patterns
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@14 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
+This paper explores selectivity estimation for single triple patterns
+in SPARQL queries.
 In this paper we see how the common assumption of independence of the

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
reindent
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@15 43ef1190-90fd-4dd7-a769-d0121f02b80e
make sense of matching
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@16 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\newcommand{\rdfnode}[1]{\texttt{#1}}
+\newcommand{\rdfterm}[1]{\texttt{#1}}
 \newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
@@
 \begin{abstract}
-This paper explores selectivity estimation for single triple patterns
-in SPARQL queries.
-In this paper we see how the common assumption of independence of the
-selectivity of RDF terms can be abandoned. We shall also find it
-instructive to explore how selectivity can be determined exactly when
-a SPARQL query is executed on a Hexastore triple store
+  This paper explores selectivity estimation for single triple
+  patterns in SPARQL queries. In this paper we see how the common
+  assumption of independence of the selectivity of RDF terms can be
+  abandoned. We shall also find it instructive to explore how
+  selectivity can be determined exactly when a SPARQL query is
+  executed on a Hexastore triple store
   \cite{Weiss:2008:HSI:1453856.1453965}. Finally, we will suggest
@@
 It is easy to point out cases where the independence assumption is
-unlikely to hold: If a predicate is \rdfnode{rdf:type}, then only
+unlikely to hold: If a predicate is \rdfterm{rdf:type}, then only
 classes are likely objects, and their frequency  would likely be
@@
 Previous authors have taken selectivity as analogous to the
-probability of matching a SPARQL node\todo{terminology, (SPARQL nodes
-  vs. RDF terms) could be better} or triple with a node or triple
-in the data. For a triple, $<S, P, O>$ the selectivity is the
-intersection of the selectivity of each node.
+probability of matching a SPARQL term (RDF Term or variable) with an
+RDF term in the data or a SPARQL triple pattern with a triple in the
+data.  For a triple, $<S, P, O>$ the selectivity is the intersection
+of the selectivity of each node.
 
@@
 where the $|$ symbol is read as ``given'', e.g. ``given a predicate
-\rdfnode{rdf:type} what is the probability that the object is \rdfnode{foaf:Person}?''
+\rdfterm{rdf:type} what is the probability that the object is \rdfterm{foaf:Person}?''
 
@@
 function $cm(x)$ which returns the number of matches of a given SPARQL
-node $x$. \todo{needs to be a lot more precise}
+term $x$. \todo{needs to be a lot more precise}
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
eq formating
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@17 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 can be written as
-\begin{equation}
-Pr(A_1 \cap A_2  \cap \ldots \cap A_n) = 
-  Pr(A_n | A_1 \cap A_2  \cap \ldots \cap A_{n-1}) \\
-  \cdot Pr(A_{n-1} | A_1 \cap A_2  \cap \ldots \cap A_{n-2}) 
-  \cdot \ldots \cdot Pr(A_2 | A_1) \cdot Pr(A_1) ,
-\end{equation}
+\begin{eqnarray}
+Pr(A_1 \cap A_2  \cap \ldots \cap A_n) &=& 
+  Pr(A_n | A_1 \cap A_2  \cap \ldots \cap A_{n-1}) \nonumber\\
+  &&\cdot Pr(A_{n-1} | A_1 \cap A_2  \cap \ldots \cap A_{n-2}) \cdot \ldots \nonumber\\
+  && \cdot Pr(A_2 | A_1) \cdot Pr(A_1) ,
+\end{eqnarray}
 where the $|$ symbol is read as ``given'', e.g. ``given a predicate

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
almost eradicate "node"
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@18 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 data.  For a triple, $<S, P, O>$ the selectivity is the intersection
-of the selectivity of each node.
+of the selectivity of each term.
 
@@
 We note that intersection is commutative, thus similar
-relations can be shown for each permutation of nodes.
+relations can be shown for each permutation of terms.
 
-Finally, note that for unbound nodes (variables), the selectivity will
-always be unity, whether conditional or independent, since it will
-match all possible RDF terms.
+Finally, note that for variables, the selectivity will always be
+unity, whether conditional or independent, since it will match all
+possible RDF terms.
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
Add simple example
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@19 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Analogously\todo{anybody good at arguing this mathematically?}, the conditional selectivity for a triple $T$ can be written as 
-\begin{equation}
+Analogously\todo{anybody good at arguing this mathematically?}, the
+conditional selectivity for a triple pattern $T$ can be written as 
+\begin{equation}\ref{eq:sel}
 sel(T) = sel(S \cap P \cap O) = sel(O | S \cap P) \cdot sel(P | S)
@@
 
-\todo{example?}
+\subsection{Example}
+
+Consider the following RDF graph, expressed in Turtle with base URI
+and namespace declarations omitted for brevity:
+
+\begin{verbatim}
+</alice> a foaf:Person ;
+         foaf:name "Alice Example" ;
+         rel:spouseOf </bob> .
+</bob>   a foaf:Person ;
+         foaf:firstName "Bob" ;
+         foaf:based_near </place> .
+</place> rdfs:label "Our House" .
+\end{verbatim}
+
+A very simple example is to find the selectivity for the triple pattern
+\begin{verbatim}
+?person a foaf:Person .
+\end{verbatim}
+We can easily see that $sel(T) = 2/7$ since two out of the 7 triples
+match the pattern. To see how Eq.~\cite{eq:sel} is used, we note that
+$sel(S) = 1$ since it is a variable. $sel(P | S) = 2/7$ since the
+predicate matches the two occurences of \rdfterm{rdf:type} and the
+subject variable doesn't influence this. Finally, $sel(O | S \cap P) =
+1$ since the only two occurences of \rdfterm{foaf:Person} occurs in
+the same triples as \rdfterm{rdf:type}. Thus, we arrive at the
+expected result.
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
latex fix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@20 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 conditional selectivity for a triple pattern $T$ can be written as 
-\begin{equation}\ref{eq:sel}
+\begin{equation}\label{eq:sel}
 sel(T) = sel(S \cap P \cap O) = sel(O | S \cap P) \cdot sel(P | S)
@@
 We can easily see that $sel(T) = 2/7$ since two out of the 7 triples
-match the pattern. To see how Eq.~\cite{eq:sel} is used, we note that
+match the pattern. To see how Eq.~\ref{eq:sel} is used, we note that
 $sel(S) = 1$ since it is a variable. $sel(P | S) = 2/7$ since the

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
another example
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@21 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 </alice> a foaf:Person ;
-         foaf:name "Alice Example" ;
+         foaf:firstName "Alice" ;
          rel:spouseOf </bob> .
@@
 expected result.
+
+In a slightly harder example:
+\begin{verbatim}
+?person foaf:firstName "Bob" .
+\end{verbatim}
+We still have $sel(S) = 1$ as above, and there are still two
+predicates that match, so $sel(P | S) = 2/7$, but only half of those
+match the literal, so $sel(O | S \cap P) = 1/2$, giving us the
+expected $sel(T) = 1/7$.
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
Add variables to selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@22 43ef1190-90fd-4dd7-a769-d0121f02b80e
add todo section
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@23 43ef1190-90fd-4dd7-a769-d0121f02b80e
Add example to hexastore section
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@24 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsection{Example}
+\subsection{Example}\label{sec:example}
 
@@
 
+Let us return to the example in section~\ref{sec:example}: We have
+$s_1 = \texttt{key-for-alice}$ and $i=2$. Further, $p_1^1 =
+\texttt{key-for-foaf:firstName}$, $p_2^1 = \texttt{key-for-rdf:type}$ and $p_3^1 =
+\texttt{key-for-rel:spouseOf}$ and $n_1 = 3$. For each predicate, there is only
+one object, so $k_{1,1} = 1$ and $o_{1_{1,1_1}}^{1,1_1} = \texttt{key-for-Alice-literal}$.
+
 Now, given a query, we shall explain how the selectivity can be
 computed exactly rather than being estimated. First, we define a
-function $cm(x)$ which returns the number of matches of a given SPARQL
-term $x$. \todo{needs to be a lot more precise}
+function $cm(x)$ which returns the number of matches of a given key
+$x$\todo{argue the term-key-connection}. \todo{needs to be a lot more precise}
 
@@
 \begin{equation}
-sel(S) = \frac{cm(s_i)}{i} ,
+sel(S = s_i) = \frac{cm(s_i)}{i} ,
 \end{equation}
 \begin{equation}
-sel(P|S) = \frac{cm(p_{n_i}^i)}{n_i}
+sel(P = p_{n_i}^i|S = s_i) = \frac{cm(p_{n_i}^i)}{n_i}
 \end{equation}
@@
 \begin{equation}
-sel(O|S \cap P) = \frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}}
+sel(O = o_{k_{i,n_i}}^{i,n_i} |S = s_i \cap P = p_{n_i}^i) =
+\frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}} .
 \end{equation}
+
+\section{Further work on conditional selectivity}
+
+\todo{this section}
 

### 2011-07-20 conditional/trunk/selectivity-hexastore.tex
more explaining of exact selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@25 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Now, given a query, we shall explain how the selectivity can be
-computed exactly rather than being estimated. First, we define a
-function $cm(x)$ which returns the number of matches of a given key
-$x$\todo{argue the term-key-connection}. \todo{needs to be a lot more precise}
+The pattern that emerges is that since the index for the predicate is
+already dependent on the match for the subject, the conditional
+selectivity depends only on the number of matches of a term and the
+total number of keys in the index. Thus, the selectivity can be
+computed \emph{exactly} based on the data only. The selectivity of a
+query then depends only on what indices are used.
+
+Formalizing this notion, we first define a function $cm(x)$ which
+returns the number of matches of a given key $x$\todo{argue the
+  term-key-connection}.
 
@@
 \end{equation}
+\todo{surely needs more work}
+
+With this, it should be possible to store the selectivity along with
+the indices in a Hexastore.
 

### 2011-11-14 background/phd-description1.tex
Some stuff for a description
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@32 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass[a4paper,english,10pt]{article}
\usepackage{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}

\title{Problem statement and suggested project outline}
\author{Kjetil Kjernsmo}

\begin{document}

\maketitle

Query federation has been an active field for some time, but has until
the advent of the Semantic Web not been used in a highly diverse set
of endpoints, commonly they have been under the control of a single
institution. RDF has a triple-based data model for the Semantic Web,
and SPARQL is a standardized query language to query such data.

Query federation with SPARQL has attracted much attention
from industry and academia alike, and four implementations of basic
query federation were submitted to the SPARQL 1.1 Working Group as
input for the forthcoming work. The basic query federation feature was
supported by a large number of group members, and the latest working
draft of the proposed standard was published on June 1st 2010 and is
expected to enter a Last Call review period shortly.

There is a very strong industrial use case for SPARQL query
federation, as it is a straightforward way to combine data from the
large number of SPARQL Endpoints. With the emergence of the Linked
Data Web, URIs serves as ``global keys'', and makes data combination
across data sources trivial if a working SPARQL query federation
engine can be used.

While the basic feature set of the proposed standard can enable users
to create federated queries, it is not of great use as it requires
extensive prior knowledge of both the data to be queried and
performance characteristics of the involved query engines.

I intend to investigate possible remedies to this problem by using
statistical techniques.

An important technique execution optimization is known as
``selectivity estimation''. This technique seeks to optimize the order
of which parts of the query is executed. In the greater database
literature, this has been widely studied, but the SPARQL spesific
literature leaves many open questions.

An oft-cited paper is \cite{Stocker:2008:SBG:1367497.1367578}. In that
paper, the authors assume that the selectivity of RDF terms are
independent, admit that this assumption is easily violated and
continue to use a conditional estimate for the object node. This is
clearly flawed, and I have already started writing a first paper to
address this. Thus, it is intended that my first paper will address
open issues on selectivity estimation in the single-database
conditional case.  Several directions can be explored, possibly
involving Bayesian Networks, user configurable common SPARQL patterns,
estimation based on known ontologies, etc.

\end{document}

### 2011-11-14 background/phd-description1.tex
Section titles, empirical evaluation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@33 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 extensive prior knowledge of both the data to be queried and
-performance characteristics of the involved query engines.
+performance characteristics of the involved query engines. Without
+this knowledge, the overall performance is insufficient for any
+practical applications.
 
@@
 statistical techniques.
+
+\section*{Paper 1: Conditional selectivity estimation on a single database}
 
@@
 
+\section*{Paper 2: Empirical evaluation of implementations}
+
+Since the main objective of the proposed work is to create systems
+that have sufficient performance for practical applications, it is of
+paramount importance to have methodology that can falsify any
+conjecture about an implementation's performance.
+
+In a federated regime, any query is likely to execute queries on
+several different SPARQL implementations. Many of those are free
+software so the algorithms they use could in principle be analyzed,
+but in practice, this would be very time-consuming. Some may be
+undocumented or even trade secrets, thus precluding any such
+analysis. It is therefore my intention to treat the individual SPARQL
+implementations as ``black boxes'' and exclusively evaluate the
+performance characteristics by empirical means, aka ``benchmarking''.
+
 \end{document}

### 2011-11-15 background/phd-description1.tex
paper 2
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@34 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+I note that none of the established SPARQL benchmarks, or any other
+benchmark I've seen, regard performance (however it is measured) as a
+stochastic variable. It clearly is --- other things running on the
+benchmark system, randomness in algorithms, unknown data structure,
+etc. --- all contribute to random fluctuations in
+performance. Consequently, none of the benchmarks use well-established
+statistical methodology for hypothesis testing. They should, as
+randomness could influence the conclusions drawn. Rectifying this
+problem is a first step in the second paper I plan to write.
+
+However, I wish to perform a much deeper investigation in the same
+direction. In physical science and engineering, conventional wisdom
+has been that you should only vary one variable at a time to study the
+effects of that one variable. In medical science, this has been
+abandoned several decades ago, thanks to advances in statistics. In
+e.g. a case where the researcher administrates different treatments to
+terminally ill patients, some of which may be painful or shorten their
+lives, experimental economy is extremely important.
+
+In SPARQL benchmarking, the state of the art is that one compiles data
+which one hopes is representative for a wide selection of use cases,
+and a select few queries. The performance is then measured. Recently,
+\cite{Duan:2011:AOC:1989323.1989340} showed that the data used by all
+existing benchmarks is far from representative, leaving many open
+problems.
+
+With this in mind, I plan to use ideas from statistical experimental
+design to investigate whether it is possible to cover the entire
+(realistic) parameter space, thus resolving the problem of choosing
+hopefully representative data and queries. This would then be the main
+topic of paper 2.
+
 \end{document}

### 2011-11-15 background/phd-description1.tex
paper 3
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@36 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section*{Paper 3: Conditional selectivity in the federated case}
+
+In paper~1, I explored selectivity estimation as a technique for query
+optimization. In a federated regime, selectivity estimation is
+important because without it, large amounts of data may needlessly be
+passed back and forth between different endpoints. However, the
+federation engine has no way of estimating the selectivity of the
+entire query since it has no record of the structure of the data.
+
+The third paper will explore ways to remedy this situation, e.g. by
+compiling statistics that may be exposed in a service description. In
+particular, parameterized statistics may be useful.
+
+Finally, techniques developed in paper~2 will be used to evaluate the
+efficiency of various techniques.
+
 \end{document}

### 2011-11-28 background/phd-description1.tex
cite package, compiles ok
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@37 43ef1190-90fd-4dd7-a769-d0121f02b80e
litt korrektur
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@42 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{times}
+\usepackage{cite}
 
@@
 
-An important technique execution optimization is known as
+An important technique for execution optimization is known as
 ``selectivity estimation''. This technique seeks to optimize the order
@@
 that have sufficient performance for practical applications, it is of
-paramount importance to have methodology that can falsify any
+paramount importance to have methodology that can falsify a
 conjecture about an implementation's performance.
@@
 software so the algorithms they use could in principle be analyzed,
-but in practice, this would be very time-consuming. Some may be
-undocumented or even trade secrets, thus precluding any such
-analysis. It is therefore my intention to treat the individual SPARQL
-implementations as ``black boxes'' and exclusively evaluate the
-performance characteristics by empirical means, aka ``benchmarking''.
+but in practice, this would be very time-consuming. Some
+implementations may be undocumented or even trade secrets, thus
+precluding any such analysis. It is therefore my intention to treat
+the individual SPARQL implementations as ``black boxes'' and
+exclusively evaluate the performance characteristics by empirical
+means, aka ``benchmarking''.
 

### 2012-01-03 eswc-2012/phdsymp.tex
Stub LNCS LaTeX
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@46 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass{llncs}

\begin{document}
\title{}
\author{Kjetil Kjernsmo}
\institute{Department of Informatics,
Postboks 1080 Blindern,
0316 Oslo, Norway}

\email{kjekje@ifi.uio.no}

\maketitle

\begin{abstract}

### 2012-01-05 eswc-2012/phdsymp.tex
Disposisjon
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@47 43ef1190-90fd-4dd7-a769-d0121f02b80e
Paste description text
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@48 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{document}
-\title{}
+\title{Sharing statistics for SPARQL Federation optimization, with
+  emphasis on benchmark quality}
 \author{Kjetil Kjernsmo}
@@
 \begin{abstract}
+
+\end{abstract}
+
+\section{Introduction}
+
+Query federation has been an active field for some time, but has until
+the advent of the Semantic Web not been used in a highly diverse set
+of endpoints, commonly they have been under the control of a single
+institution. RDF has a triple-based data model for the Semantic Web,
+and SPARQL is a standardized query language to query such data.
+
+Query federation with SPARQL has attracted much attention
+from industry and academia alike, and four implementations of basic
+query federation were submitted to the SPARQL 1.1 Working Group as
+input for the forthcoming work. The basic query federation feature was
+supported by a large number of group members, and the latest working
+draft of the proposed standard was published on June 1st 2010 and is
+expected to enter a Last Call review period shortly. % TODO: Check
+                                % status
+
+While the basic feature set of the proposed standard can enable users
+to create federated queries, it is not of great use as it requires
+extensive prior knowledge of both the data to be queried and
+performance characteristics of the involved query engines. Without
+this knowledge, the overall performance is insufficient for any
+practical applications.
+
+I intend to investigate possible remedies to this problem by using
+statistical techniques.
+
+Since the main objective of the proposed work is to create systems
+that have sufficient performance for practical applications, it is of
+paramount importance to have methodology that can falsify a
+conjecture about an implementation's performance.
+
+In a federated regime, any query is likely to execute queries on
+several different SPARQL implementations. Many of those are free
+software so the algorithms they use could in principle be analyzed,
+but in practice, this would be very time-consuming. Some
+implementations may be undocumented or even trade secrets, thus
+precluding any such analysis. It is therefore my intention to treat
+the individual SPARQL implementations as ``black boxes'' and
+exclusively evaluate the performance characteristics by empirical
+means, aka ``benchmarking''.
+
+However, I have found the current state of the art in benchmarking
+lacking in its use of statistics, which motivates two distincts
+directions of work: \emph{Statistical experimental planning and execution in
+benchmarking} and \emph{statistical methods to optimize SPARQL queries
+in a federated regime}.
+
+\section{Problem Definition}
+
+\section{State of the Art}
+
+\section{Proposed Approach and Methodology}
+
+\section{Conclusion}
+
+\end{document}

### 2012-01-09 eswc-2012/phdsymp.tex
More disposition
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@49 43ef1190-90fd-4dd7-a769-d0121f02b80e
Elevator pitches
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@50 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\maketitle
+
 \email{kjekje@ifi.uio.no}
-
-\maketitle
 
@@
 query federation were submitted to the SPARQL 1.1 Working Group as
-input for the forthcoming work. The basic query federation feature was
-supported by a large number of group members, and the latest working
-draft of the proposed standard was published on June 1st 2010 and is
-expected to enter a Last Call review period shortly. % TODO: Check
-                                % status
+input for the forthcoming
+work\footnote{http://www.w3.org/2009/sparql/wiki/Feature:BasicFederatedQuery}. 
+The basic query federation feature was
+supported by a large number of group members, and the Last Call working
+draft of the proposed standard was published on 17 November 2011.
 
@@
 paramount importance to have methodology that can falsify a
-conjecture about an implementation's performance.
+hypothesis about an implementation's performance.
 
@@
 
+\subsection{In SPARQL Federation}
+
+This part of the thesis addresses problems of query optimization in a
+federated regime.
+
+\subsection{In Benchmarking}
+
+The contemporary problem in benchmarking is to distinguish
+implementations when the measured difference is little or susceptible
+to random noise.
+
 \section{State of the Art}
 
+\subsection{In SPARQL Federation}
+
+\subsubsection{Technology}
+
+I take the technology state of the art to be represented by the
+current basic SPARQL 1.1 Federated Query Working
+Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}. For
+an academic treatment of the current specification, see
+\cite{springerlink:10.1007/978-3-642-21064-8_1} and references therein.
+
+\subsubsection{Science}
+
+In \cite{springerlink:10.1007/978-3-642-21064-8_1}, the authors show
+an optimization strategy based on execution order of different
+algebraic elements. % TODO: wording here
+
+In \cite{5337556}, it is shown how histograms can be computed to
+assist the SemWIQ optimizer, written by the same authors.
+% RDFStats
+
+\subsection{In Benchmarking}
+
+Numerous benchmarks have been developed for SPARQL, but
+\cite{Duan:2011:AOC:1989323.1989340} showed that currently, most
+benchmarks poorly represent the typical data and queries that is used
+on the Semantic Web.
+
+Most recently, \cite{mxro:Morsey2011DBpedia} addressed some of these
+problems by using real data and real queries from DBpedia.
+
+The problems addressed by these studies are almost orthogonal to the
+problems considered by my proposed project.
+
 \section{Proposed Approach and Methodology}
+
+\subsection{In SPARQL Federation}
+
+There are many possible approaches for this part of the thesis. At
+this stage, it seems that use of selectivity estimation may be the
+most feasible direction.
+
+\subsection{In Benchmarking}
+
+This part of the thesis seeks to use advances from statistical
+experimental planning to improve accuracy and dependability of benchmarking.
 

### 2012-01-10 eswc-2012/phdsymp.tex
benchmarking sota
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@52 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 The problems addressed by these studies are almost orthogonal to the
-problems considered by my proposed project.
+problems considered by my proposed project. I have not to date seen
+any work towards using contemporary statistical methods to evaluate
+the performance of software, but I shall admit that my literature
+study have been limited to RDF/OWL databases, and practical benchmarks
+of databases, file systems, scientific software, etc. Thus, the
+existence of relevant references in the deeper computer science
+literature is a key issue I would appreciate discussing at the
+Symposium.
 

### 2012-01-10 eswc-2012/phdsymp.tex
stochastic variable
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@53 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-\documentclass{llncs}
+%\documentclass{llncs}
+\documentclass{article}
 
@@
 performance characteristics of the involved query engines. Without
-this knowledge, the overall performance is insufficient for any
+this knowledge, the overall performance is insufficient for many
 practical applications.
@@
 to random noise.
+
+Current common practice in benchmarking SPARQL-enabled systems is to
+use or synthesize a certain dataset, then formulate a number of
+queries seen as representative of SPARQL use in some way. These
+queries are then executed, and some characteristic of performance is
+measured, for example the time it takes for the engine to return the
+full result. Commonly, this process is repeated a number of times, and
+an average response time is found. The performance of different
+engines can then be compared based on these averages.
+
+In many cases, this is sufficient. Sometimes, one engine can execute a
+query an order of magnitude faster than another, in which case, there
+is hardly reasonable doubt which is faster.
+
+When the difference is little, we must consider the possibility that
+the measurements are impacted by random noise. Computer systems are
+generally very deterministic systems, but there are nevertheless many
+such scenarios: The underlying system may be doing other tasks. This
+is likely to be particularly important in enterprise environments
+where high-load applications are often running on virtual
+machines. The engine may use randomized algorithms, or the data may be
+of such a nature it triggers worst-case behaviour in situations the
+query writer cannot foresee. All this contributes to that we should
+consider the performance metric, whatever metric is used, as a
+\emph{stochastic variable}.
+
+Doing this opens new methodological possibilities, first and foremost
+using well-established statistical hypothesis testing rather than just
+comparing averages.
 

### 2012-01-10 eswc-2012/phdsymp.tex
experimental design
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@54 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+With the currently published benchmarks, it is up to the user to
+select an engine with the best characteristics relative to their own
+expected application. Another possibility that opens is to better
+answer the question ``what is the best general-purpose triple
+store?''. I.e. can we collapse all the metrics into a single
+\emph{test statistic} that can summarize the performance of the engine
+as a whole? Also, can we provide a system to developers that they can
+use to see if any changes they do to the engine has unexpected adverse
+effects on other parts of the system?
+
+The above relies on that the test data and queries are in fact
+representative of the system as a whole, but this is usually not proven.
+In physical science and engineering, conventional wisdom
+has been that you should only vary one variable at a time to study the
+effects of that one variable. In medical science, this has been
+abandoned several decades ago, thanks to advances in statistics. In
+e.g. a case where the researcher administrates different treatments to
+terminally ill patients, some of which may be painful or shorten their
+lives, experimental economy is extremely important.
+
+Using techniques from statistical experimental design, I conjecture
+that it is possible to design an experiment (i.e. a benchmark) so
+that it provably representative of all realistic situations. This
+would permit us to make rather categorical statements regarding
+overall query engine performance.
+
 \section{State of the Art}

### 2012-01-10 eswc-2012/phdsymp.tex
tilde
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@55 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 supported by a large number of group members, and the Last Call working
-draft of the proposed standard was published on 17 November 2011.
+draft of the proposed standard was published on 17~November~2011.
 

### 2012-01-11 eswc-2012/phdsymp.tex
problem def for SPARQL
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@56 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-%\documentclass{llncs}
-\documentclass{article}
+\documentclass{llncs}
+%\documentclass{article}
 
@@
 federated regime.
+
+An intrinsic problem of the current federated regime is that the
+system first executing the federated query has no information of the
+characteristics of neither the data nor the performance of the systems
+it further submits queries to. This limits the techniques it can use
+to optimize the query before submitting.
+
+One aspect of this problem can be illustrated by a self-experienced
+practical problem: In a federated conjuctive query, we created a query
+to ask for certain books written in a certain
+language. Some of the book metadata resided on one endpoint, the
+language metadata was on another endpoint. Unbeknowst to us, the
+distribution of languages were extremely skewed, i.e. the vast
+majority of books were written in one language. Thus, sending the
+language query first would in some cases be favourable, returning only
+a handful of results, in other cases, it would result in a very large
+number of results, which would then need to be bound in queries sent
+to other endpoints, causing weak performance.
+
+In the non-federated case, this can be mitigated by the well-known
+histogram technique for subsequent selectivity estimation at query
+execution time. In the federated case, this is not possible as nothing
+is known about the distribution of data on individual
+endpoints. \cite{5337556} addressed this problem by making it possible
+to exchange histograms, but in the general case this is untenable as
+the number of histograms to be computed and the number
+of bins in each histogram to represent highly heterogenous data would
+be very high. This would also cause the amount of statistical data to
+be exchanged between endpoints to be very high.
 

### 2012-01-11 eswc-2012/phdsymp.tex
long term goal
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@57 43ef1190-90fd-4dd7-a769-d0121f02b80e
uncommon hypothesis test
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@58 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 in a federated regime}.
+
+I am still in an early phase of my work, and I have not yet started to
+explore the scientific literature around SPARQL Federation to any
+depth, currently focusing my efforts on benchmarking. Therefore, this
+paper also focuses on benchmarking rather than the more long-term goal
+of SPARQL Federation.
 
@@
 The problems addressed by these studies are almost orthogonal to the
-problems considered by my proposed project. I have not to date seen
-any work towards using contemporary statistical methods to evaluate
-the performance of software, but I shall admit that my literature
-study have been limited to RDF/OWL databases, and practical benchmarks
-of databases, file systems, scientific software, etc. Thus, the
-existence of relevant references in the deeper computer science
-literature is a key issue I would appreciate discussing at the
-Symposium.
+problems considered by my proposed project. While I have heard of some
+cases where a hypothesis test is used to demonstrate that one
+implementation is better than another, it is seemingly not common
+practice. Furthermore, I have not to date seen any work towards using
+contemporary statistical methods to evaluate the performance of
+software, but I shall admit that my literature study have been limited
+to RDF/OWL databases, and practical benchmarks of databases, file
+systems, scientific software, etc. Thus, the existence of relevant
+references in the deeper computer science literature is a key issue I
+would appreciate discussing at the Symposium.
 

### 2012-01-11 eswc-2012/phdsymp.tex
gc may be random
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@59 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 where high-load applications are often running on virtual
-machines. The engine may use randomized algorithms, or the data may be
-of such a nature it triggers worst-case behaviour in situations the
-query writer cannot foresee. All this contributes to that we should
-consider the performance metric, whatever metric is used, as a
-\emph{stochastic variable}.
+machines. The garbage collection of some languages may trigger on not
+easily foreseeable points. The engine may use randomized algorithms,
+or the data may be of such a nature it triggers worst-case behaviour
+in situations the query writer cannot foresee. All this contributes to
+that we should consider the performance metric, whatever metric is
+used, as a \emph{stochastic variable}.
 

### 2012-01-11 eswc-2012/phdsymp.tex
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@61 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 The above relies on that the test data and queries are in fact
-representative of the system as a whole, but this is usually not proven.
-In physical science and engineering, conventional wisdom
-has been that you should only vary one variable at a time to study the
-effects of that one variable. In medical science, this has been
-abandoned several decades ago, thanks to advances in statistics. In
-e.g. a case where the researcher administrates different treatments to
-terminally ill patients, some of which may be painful or shorten their
-lives, experimental economy is extremely important.
+representative of the system as a whole. It is often asserted that it
+is not possible to create a ``one-size-fits-all'' benchmark (see
+e.g. \cite{Schmidt:2011:FBS:2063016.2063054}) to cover all aspects of
+an application. While that may be true in the general case, I want to
+challenge the assumption that the best we can do is to accommodate a
+multitude of dimensions and essential challenges in the case of
+SPARQL.
+
+In physical science and engineering, conventional wisdom has been that
+you should only vary one variable at a time to study the effects of
+that one variable. In medical science, this has been abandoned several
+decades ago, thanks to advances in statistics. In e.g. a case where
+the researcher administrates different treatments to terminally ill
+patients, some of which may be painful or shorten their lives,
+experimental economy is extremely important.
 
 Using techniques from statistical experimental design, I conjecture
-that it is possible to design an experiment (i.e. a benchmark) so
-that it provably representative of all realistic situations. This
-would permit us to make rather categorical statements regarding
-overall query engine performance.
+that it is possible to design an experiment (i.e. a benchmark) so that
+we can prove it covers all realistic cases and that we can justify
+why the remaining corner cases are unlikely to influence the result.
 
@@
 \subsection{In SPARQL Federation}
-
-\subsubsection{Technology}
 
@@
 \cite{springerlink:10.1007/978-3-642-21064-8_1} and references therein.
-
-\subsubsection{Science}
 

### 2012-01-11 eswc-2012/phdsymp.tex
fed sota
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@62 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 current basic SPARQL 1.1 Federated Query Working
-Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}. For
-an academic treatment of the current specification, see
-\cite{springerlink:10.1007/978-3-642-21064-8_1} and references therein.
+Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}.
 
-In \cite{springerlink:10.1007/978-3-642-21064-8_1}, the authors show
-an optimization strategy based on execution order of different
-algebraic elements. % TODO: wording here
+A recent scientific treatment of the current specification, is in
+\cite{springerlink:10.1007/978-3-642-21064-8_1}. In this paper, the
+authors also show an optimization strategy based on execution order of
+different algebraic elements. % TODO: wording here
 
-In \cite{5337556}, it is shown how histograms can be computed to
+In \cite{5337556}, it is shown how various statistics-gathering
+techniques, including histograms can be computed to
 assist the SemWIQ optimizer, written by the same authors.
-% RDFStats
+
+For benchmarking federated query processing strategies,
+\cite{Schmidt:2011:FBS:2063016.2063054} report the development of \emph{FedBench}.
 

### 2012-01-11 eswc-2012/phdsymp.tex
Remove stuff from fed problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@63 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-An intrinsic problem of the current federated regime is that the
-system first executing the federated query has no information of the
-characteristics of neither the data nor the performance of the systems
-it further submits queries to. This limits the techniques it can use
-to optimize the query before submitting.
+When using FedBench, (see \cite{Schmidt:2011:FBS:2063016.2063054}),
+the authors note that the absence of statistics has very adverse
+effects on performance as join order is impossible to optimize. It is
+also pointed out that it causes other problems as well, for example a
+high number of requests to federation members.
 
-One aspect of this problem can be illustrated by a self-experienced
-practical problem: In a federated conjuctive query, we created a query
-to ask for certain books written in a certain
-language. Some of the book metadata resided on one endpoint, the
-language metadata was on another endpoint. Unbeknowst to us, the
-distribution of languages were extremely skewed, i.e. the vast
-majority of books were written in one language. Thus, sending the
-language query first would in some cases be favourable, returning only
-a handful of results, in other cases, it would result in a very large
-number of results, which would then need to be bound in queries sent
-to other endpoints, causing weak performance.
-
-In the non-federated case, this can be mitigated by the well-known
-histogram technique for subsequent selectivity estimation at query
-execution time. In the federated case, this is not possible as nothing
-is known about the distribution of data on individual
-endpoints. \cite{5337556} addressed this problem by making it possible
-to exchange histograms, but in the general case this is untenable as
-the number of histograms to be computed and the number
-of bins in each histogram to represent highly heterogenous data would
-be very high. This would also cause the amount of statistical data to
-be exchanged between endpoints to be very high.
+This problem has already received some attention, \cite{5337556}
+addressed this problem by making it possible to exchange histograms,
+but I suspect this will prove to become untenable as the number of
+histograms to be computed and the number of bins in each histogram to
+represent highly heterogenous data would be very high. This would also
+cause the amount of statistical data to be exchanged between endpoints
+to be very high.
 

### 2012-01-11 eswc-2012/phdsymp.tex
use little space
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@65 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 This problem has already received some attention, \cite{5337556}
-addressed this problem by making it possible to exchange histograms,
-but I suspect this will prove to become untenable as the number of
-histograms to be computed and the number of bins in each histogram to
-represent highly heterogenous data would be very high. This would also
-cause the amount of statistical data to be exchanged between endpoints
-to be very high.
+addressed this problem by making it possible to exchange histograms
+and other data, and \cite{Harth:2010:DSO:1772690.1772733} introduced
+QTrees. I suspect these approaches will suffer from either involving
+too much statistical data or if constraining the data size, leaving
+too much out. The problem becomes how to digest the most useful
+statistics in as little as possible space.
 

### 2012-01-12 eswc-2012/phdsymp.tex
comment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@66 43ef1190-90fd-4dd7-a769-d0121f02b80e
histograms grow large
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@67 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-01-11 eswc-2012/phdsymp.tex
@@
 This problem has already received some attention, \cite{5337556}
-addressed this problem by making it possible to exchange histograms,
-but I suspect this will prove to become untenable as the number of
-histograms to be computed and the number of bins in each histogram to
-represent highly heterogenous data would be very high. This would also
-cause the amount of statistical data to be exchanged between endpoints
-to be very high.
+addressed this problem by making it possible to exchange histograms
+and other data. However, histogram approaches generally suffers from
+the problem that they grow too large or become a insufficiently
+accurate digest, especially in face of very heterogenous data.
 
@@
 
+\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees to alleviate
+the problem, but it is not clear to me if it will solve it.
+
 \subsection{In Benchmarking}

### 2012-01-12 eswc-2012/phdsymp.tex
move qtree to sota
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@68 43ef1190-90fd-4dd7-a769-d0121f02b80e
TeXing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@70 43ef1190-90fd-4dd7-a769-d0121f02b80e
fed approaches
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@71 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\documentclass{article}
+\usepackage{cite}
 
-\begin{document}
 \title{Sharing statistics for SPARQL Federation optimization, with
@@
 
+\begin{document}
+
 \maketitle
-
 \email{kjekje@ifi.uio.no}
@@
 A recent scientific treatment of the current specification, is in
-\cite{springerlink:10.1007/978-3-642-21064-8_1}. In this paper, the
+\cite{springerlink:10.1007/978-3-642-21064-8-1}. In this paper, the
 authors also show an optimization strategy based on execution order of
@@
 In \cite{5337556}, it is shown how various statistics-gathering
-techniques, including histograms can be computed to
-assist the SemWIQ optimizer, written by the same authors.
+techniques, including histograms can be computed to assist the SemWIQ
+optimizer, written by the same authors.
+\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees, which may
+alleviate the problem of histogram size, but it may not solve it.
 
@@
 \cite{Schmidt:2011:FBS:2063016.2063054} report the development of \emph{FedBench}.
-
-\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees to alleviate
-the problem, but it is not clear to me if it will solve it.
 
@@
 
-There are many possible approaches for this part of the thesis. At
-this stage, it seems that use of selectivity estimation may be the
-most feasible direction.
+There are many possible approaches for this part of the thesis. As I
+expect great advances to be made before I start tackling this problem,
+I have not chosen any methodology, but work to find more
+space-efficient ways to expose statistics in the service description
+and standardize those seems like an interesting direction.
+
+To this end, I briefly looked into two approaches:
+\cite{Getoor:2001:SEU:375663.375727} used Bayesian Networks and
+Probabilistic Relational Models to efficiently represent the joint
+distribution of database tables, a formalism that could be extended to
+RDF databases.
+
+Another approach that I have not seen used in the literature is to use
+parameterized statistics. This would amount to attempt fitting data to
+a known distribution function and expose which distribution and its
+parameters in the service description.
+
+Finally, I have seen little work on the problem of rapidly changing
+data, so the adaption of existing techniques to such situations may
+also be a interesting problem.
 
@@
 
+\bibliographystyle{plain}
+\bibliography{selectivity,federation,benchmarks}
+
 \end{document}

### 2012-01-12 eswc-2012/phdsymp.tex
Benchmark statistics design
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@72 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Using techniques from statistical experimental design, I conjecture
+Using techniques from statistical experimental design, I propose
 that it is possible to design an experiment (i.e. a benchmark) so that
@@
 This part of the thesis seeks to use advances from statistical
-experimental planning to improve accuracy and dependability of benchmarking.
+experimental planning to improve accuracy and dependability of
+benchmarking. Already in 1926, Ronald~Fischer noted that complex
+experiments can be much more efficient than simple
+ones~\cite{Fisher_1926}\footnote{Cited in
+  http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
+experimental design field. One of the simpler designs is ``fractional
+factorial design'', in which several ``factors'' are studied by using
+different ``treatments''. In terms of SPARQL execution, one factor can
+for example be the level of nested \texttt{OPTIONAL}s, number of
+triples in a basic graph pattern, etc, and the treatment is to vary
+these numbers. The key to understanding why this can be efficient is
+that these variations need not occur in the same experiment. Thus,
+many combinations of factors can be studied by carefully designing
+queries to cover different factors, and a formalism called
+``resolution'' has been developed to classify how well this has been
+achieved.
+
+Fractional factorial design is covered in elementary text-books in
+statistics, and I intend to go further into the statistical literature
+to see if there is methodology that is even better suited to the
+problem. Armed with this, I hope to use amongst other things, the
+complexity analysis of SPARQL (see
+e.g. \cite{Schmidt:2010:FSQ:1804669.1804675}) to find suitable factors
+to see if my admittedly bold proposition that it is possible to design
+a benchmark to cover all realistic cases.
 
@@
 \bibliographystyle{plain}
-\bibliography{selectivity,federation,benchmarks}
+\bibliography{selectivity,federation,benchmarks,statistics,optimization}
 

### 2012-01-12 eswc-2012/phdsymp.tex
texism
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@73 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   emphasis on benchmark quality}
-\author{Kjetil Kjernsmo}
+\titlerunning{Sharing statistics and benchmarking for SPARQL federation}
+\author{Kjetil Kjernsmo\inst{1}}
 \institute{Department of Informatics,
 Postboks 1080 Blindern,
-0316 Oslo, Norway}
+0316 Oslo, Norway
+\email{kjekje@ifi.uio.no}}
 
@@
 \maketitle
-\email{kjekje@ifi.uio.no}
 
@@
 experiments can be much more efficient than simple
-ones~\cite{Fisher_1926}\footnote{Cited in
+ones\footnote{Cited in
   http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
@@
 \bibliographystyle{plain}
-\bibliography{selectivity,federation,benchmarks,statistics,optimization}
+\bibliography{selectivity,federation,benchmarks,optimization}
 

### 2012-01-12 eswc-2012/phdsymp.tex
abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@74 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
+Federation of semantic data on SPARQL endpoints will allow data to
+remain distributed so that they can be controlled by local curators
+and swiftly updated. It creates considerable performance problems,
+however, that is intended in the suggested work to be addressed mainly
+by computation and exposure of statistical digests to assist
+selectivity estimation.
 
+Discontent with current state of the art in benchmarking prompts the
+author to devote much of the time to the evaluation problem and to
+challenge the notion that a ``one-size-fits-all'' benchmark is
+impossible.
 \end{abstract}
@@
 
-\section{Conclusion}
+%\section{Conclusion}
 

### 2012-01-12 eswc-2012/phdsymp.tex
speling
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@75 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 However, I have found the current state of the art in benchmarking
-lacking in its use of statistics, which motivates two distincts
+lacking in its use of statistics, which motivates two distinct
 directions of work: \emph{Statistical experimental planning and execution in
@@
 the problem that they grow too large or become a insufficiently
-accurate digest, especially in face of very heterogenous data.
+accurate digest, especially in face of very heterogeneous data.
 
@@
 Another approach that I have not seen used in the literature is to use
-parameterized statistics. This would amount to attempt fitting data to
+parametrized statistics. This would amount to attempt fitting data to
 a known distribution function and expose which distribution and its

### 2012-01-12 eswc-2012/phdsymp.tex
well-designed patterns
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@76 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 authors also show an optimization strategy based on execution order of
-different algebraic elements. % TODO: wording here
+so-called well-designed patterns.
 
@@
 benchmarking. Already in 1926, Ronald~Fischer noted that complex
-experiments can be much more efficient than simple
-ones\footnote{Cited in
-  http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
+experiments can be much more efficient than simple ones\footnote{Cited
+  in http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
 experimental design field. One of the simpler designs is ``fractional
 factorial design'', in which several ``factors'' are studied by using
-different ``treatments''. In terms of SPARQL execution, one factor can
-for example be the level of nested \texttt{OPTIONAL}s, number of
-triples in a basic graph pattern, etc, and the treatment is to vary
-these numbers. The key to understanding why this can be efficient is
-that these variations need not occur in the same experiment. Thus,
-many combinations of factors can be studied by carefully designing
-queries to cover different factors, and a formalism called
-``resolution'' has been developed to classify how well this has been
-achieved.
+different ``treatments''. In terms of SPARQL execution, the SPARQL
+engine is clearly a factor, but also for example the nestedness of
+\texttt{OPTIONAL}s can be a factor, number of triples in a basic graph
+pattern, etc, and the treatment is to vary these numbers to different
+``levels''. The key to understanding why this can be efficient is that
+these variations need not occur in the same experiment. Thus, many
+combinations of factors can be studied by carefully designing queries
+to cover different factors, and a formalism called ``resolution'' has
+been developed to classify how well this has been achieved.
 
 Fractional factorial design is covered in elementary text-books in
-statistics, and I intend to go further into the statistical literature
-to see if there is methodology that is even better suited to the
-problem. Armed with this, I hope to use amongst other things, the
-complexity analysis of SPARQL (see
-e.g. \cite{Schmidt:2010:FSQ:1804669.1804675}) to find suitable factors
-to see if my admittedly bold proposition that it is possible to design
-a benchmark to cover all realistic cases.
+statistics but is most likely inadequate for this purpose, so I
+intend to go further into the statistical literature to see if there
+is methodology that is even better suited to the problem. Armed with
+this, I hope to use amongst other things, the complexity analysis of
+SPARQL (see e.g. \cite{Schmidt:2010:FSQ:1804669.1804675}) to find
+suitable factors to see if my admittedly bold proposition that it is
+possible to design a benchmark to cover all realistic cases.
 

### 2012-01-12 eswc-2012/phdsymp.tex
simplify abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@77 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   remain distributed so that they can be controlled by local curators
-and swiftly updated. It creates considerable performance problems,
-however, that is intended in the suggested work to be addressed mainly
-by computation and exposure of statistical digests to assist
-selectivity estimation.
+  and swiftly updated. There are considerable performance problems,
+  intended in the suggested work to be addressed mainly by computation
+  and exposure of statistical digests to assist selectivity
+  estimation.
 

### 2012-01-13 eswc-2012/phdsymp.tex
Tilbakemelding fra Martin
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@78 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   and swiftly updated. There are considerable performance problems,
-  intended in the suggested work to be addressed mainly by computation
+  which the present work proposes to address mainly by computation
   and exposure of statistical digests to assist selectivity
@@
   Discontent with current state of the art in benchmarking prompts the
-  author to devote much of the time to the evaluation problem and to
-  challenge the notion that a ``one-size-fits-all'' benchmark is
-  impossible.
+  author to focus on the evaluation problem and to challenge the
+  notion that a ``one-size-fits-all'' benchmark is impossible.
 \end{abstract}
@@
 performance characteristics of the involved query engines. Without
-this knowledge, the overall performance is insufficient for many
+this knowledge, the overall performance is insufficient for most
 practical applications.
@@
 that have sufficient performance for practical applications, it is of
-paramount importance to have methodology that can falsify a
-hypothesis about an implementation's performance.
+paramount importance to have methodology that can rigorously rank
+implementations based on performance.
 
@@
 
-However, I have found the current state of the art in benchmarking
+However, I have found the current state of the art in SPARQL benchmarking
 lacking in its use of statistics, which motivates two distinct
@@
 
+Thus, the core problem is: How do we compute and expose a digest that
+is of optimal size for the query performance problem?
+
 \subsection{In Benchmarking}
@@
 The contemporary problem in benchmarking is to distinguish
-implementations when the measured difference is little or susceptible
+implementations when the measured difference is small or susceptible
 to random noise.
@@
 
-When the difference is little, we must consider the possibility that
+When the difference is small, we must consider the possibility that
 the measurements are impacted by random noise. Computer systems are
 generally very deterministic systems, but there are nevertheless many
-such scenarios: The underlying system may be doing other tasks. This
-is likely to be particularly important in enterprise environments
-where high-load applications are often running on virtual
-machines. The garbage collection of some languages may trigger on not
-easily foreseeable points. The engine may use randomized algorithms,
-or the data may be of such a nature it triggers worst-case behaviour
-in situations the query writer cannot foresee. All this contributes to
+such scenarios: The underlying system may be doing other tasks. The
+garbage collection of some languages may trigger at not easily
+foreseeable points. The engine may use randomized algorithms, or the
+data may be of such a nature that it triggers worst-case behaviour in
+situations the query writer cannot foresee. All this contributes to
 that we should consider the performance metric, whatever metric is
@@
 
-The above relies on that the test data and queries are in fact
+The above relies on data and queries that are in fact
 representative of the system as a whole. It is often asserted that it

### 2012-01-13 eswc-2012/phdsymp.tex
Mer svar til Martin om one-size-fits-all
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@79 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 select an engine with the best characteristics relative to their own
-expected application. Another possibility that opens is to better
-answer the question ``what is the best general-purpose triple
-store?''. I.e. can we collapse all the metrics into a single
-\emph{test statistic} that can summarize the performance of the engine
-as a whole? Also, can we provide a system to developers that they can
-use to see if any changes they do to the engine has unexpected adverse
-effects on other parts of the system?
+expected application. However, I am often asked the question: ``what
+is the best general-purpose triple store?'' and those who ask are not
+prepared to provide any further qualification. This prompts the
+question: Can we collapse all the metrics into a single \emph{test
+  statistic} that can summarize the performance of the engine as a
+whole? Also, can we provide a system to developers that they can use
+to see if any changes they do to the engine has unexpected adverse
+effects on other parts of the system? For those who are able to
+qualify their inquiry, weighing different parts of the system or
+metrics is more useful and will yield a more tenable answer, and
+motivates the question: Can we use the system to better find the
+distinguishing characteristics so that we can return a question that
+will actually help new users of SPARQL engines to chose?
 

### 2012-01-13 eswc-2012/phdsymp.tex
Tilbakemelding fra Audun
Biblografistyle

benchmarking for optimization


git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@80 43ef1190-90fd-4dd7-a769-d0121f02b80e
Tilbakemelding fra Audun
kommenter små forskjeller


git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@81 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+These small differences may become important if they are systematic,
+even if one engine is dramatically better than another in one case,
+small deficiencies may add up to make the other a better choice for
+most applications anyway.
+
 Doing this opens new methodological possibilities, first and foremost
@@
 we can prove it covers all realistic cases and that we can justify
-why the remaining corner cases are unlikely to influence the result.
+why the remaining corner cases are unlikely to influence the
+result. For further elaboration, see Section~\ref{sec:benchmethod}.
+
+Thus far, the benchmarking problem has been seen as a software testing
+problem, but we could also ask if the existence of benchmark data can
+be exposed to help federation query optimizers along with a
+statistical digest.
 
@@
 
-\subsection{In Benchmarking}
+\subsection{In Benchmarking}\label{sec:benchmethod}
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
+\bibliographystyle{plain-csmin}
+%\bibliographystyle{splncs03}
 \bibliography{selectivity,federation,benchmarks,optimization}

### 2012-01-13 eswc-2012/phdsymp.tex
reorg
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@82 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 query an order of magnitude faster than another, in which case, there
-is hardly reasonable doubt which is faster.
+is hardly reasonable doubt which is faster. Small differences may seem
+unimportant but may become important if they are systematic. Even if
+one engine is dramatically better than another in one case, small
+deficiencies may add up to make the other a better choice for most
+applications anyway.
 
@@
 used, as a \emph{stochastic variable}.
-
-These small differences may become important if they are systematic,
-even if one engine is dramatically better than another in one case,
-small deficiencies may add up to make the other a better choice for
-most applications anyway.
 
@@
 %\bibliographystyle{plain}
-\bibliographystyle{plain-csmin}
-%\bibliographystyle{splncs03}
+%\bibliographystyle{plain-csmin}
+\bibliographystyle{splncs03}
 \bibliography{selectivity,federation,benchmarks,optimization}

### 2012-01-13 eswc-2012/phdsymp.tex
snip
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@83 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Query federation has been an active field for some time, but has until
-the advent of the Semantic Web not been used in a highly diverse set
-of endpoints, commonly they have been under the control of a single
-institution. RDF has a triple-based data model for the Semantic Web,
-and SPARQL is a standardized query language to query such data.
-
 Query federation with SPARQL has attracted much attention
@@
 implementations based on performance.
-
-In a federated regime, any query is likely to execute queries on
-several different SPARQL implementations. Many of those are free
-software so the algorithms they use could in principle be analyzed,
-but in practice, this would be very time-consuming. Some
-implementations may be undocumented or even trade secrets, thus
-precluding any such analysis. It is therefore my intention to treat
-the individual SPARQL implementations as ``black boxes'' and
-exclusively evaluate the performance characteristics by empirical
-means, aka ``benchmarking''.
 
@@
 
-When the difference is small, we must consider the possibility that
-the measurements are impacted by random noise. Computer systems are
-generally very deterministic systems, but there are nevertheless many
-such scenarios: The underlying system may be doing other tasks. The
-garbage collection of some languages may trigger at not easily
-foreseeable points. The engine may use randomized algorithms, or the
-data may be of such a nature that it triggers worst-case behaviour in
-situations the query writer cannot foresee. All this contributes to
-that we should consider the performance metric, whatever metric is
-used, as a \emph{stochastic variable}.
+In this case, we must consider the possibility that the measurements
+are impacted by random noise. Computer systems are generally very
+deterministic systems, but there are nevertheless many such scenarios:
+The underlying system may be doing other tasks. The garbage collection
+of some languages may trigger at not easily foreseeable points. The
+engine may use randomized algorithms, or the data may be of such a
+nature that it triggers worst-case behaviour in situations the query
+writer cannot foresee. All this contributes to that we should consider
+the performance metric, whatever metric is used, as a \emph{stochastic
+  variable}.
 

### 2012-01-13 eswc-2012/phdsymp.tex
rewrite introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@84 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-I intend to investigate possible remedies to this problem by using
-statistical techniques.
+To help optimize, SPARQL endpoints should expose details about both
+data and performance characteristics of the engine itself. The
+proposed work has two focal points: \emph{Statistical digests of
+  data for optimizations} and \emph{benchmarking SPARQL engines}.
 
-Since the main objective of the proposed work is to create systems
-that have sufficient performance for practical applications, it is of
-paramount importance to have methodology that can rigorously rank
-implementations based on performance.
-
-However, I have found the current state of the art in SPARQL benchmarking
-lacking in its use of statistics, which motivates two distinct
-directions of work: \emph{Statistical experimental planning and execution in
-benchmarking} and \emph{statistical methods to optimize SPARQL queries
-in a federated regime}.
+The focus on SPARQL benchmarking is not just motivated from the
+optimization perspective, as I have found the current state of the art
+in SPARQL benchmarking lacking in its use of statistics. The emphasis
+in the present paper is on statistics in benchmarking with the purpose
+of putting it on a firmer foundation from a hypothesis testing
+perspective.
 
@@
 explore the scientific literature around SPARQL Federation to any
-depth, currently focusing my efforts on benchmarking. Therefore, this
-paper also focuses on benchmarking rather than the more long-term goal
-of SPARQL Federation.
+depth, currently focusing my efforts on benchmarking. The long-term
+goal of my work is SPARQL Federation, but it is a minor concern in
+this paper.
 

### 2012-01-13 eswc-2012/phdsymp.tex
snip
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@85 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In this case, we must consider the possibility that the measurements
-are impacted by random noise. Computer systems are generally very
-deterministic systems, but there are nevertheless many such scenarios:
-The underlying system may be doing other tasks. The garbage collection
-of some languages may trigger at not easily foreseeable points. The
-engine may use randomized algorithms, or the data may be of such a
-nature that it triggers worst-case behaviour in situations the query
-writer cannot foresee. All this contributes to that we should consider
-the performance metric, whatever metric is used, as a \emph{stochastic
+are impacted by random noise. For example: The underlying system may
+be doing other tasks. The garbage collection of some languages may
+trigger at not easily foreseeable points. The engine may use
+randomized algorithms, or the data may be of such a nature that it
+triggers worst-case behaviour in situations the query writer cannot
+foresee. All this contributes to that we should consider the
+performance metric, whatever metric is used, as a \emph{stochastic
   variable}.
@@
 Doing this opens new methodological possibilities, first and foremost
-using well-established statistical hypothesis testing rather than just
-comparing averages.
+using well-established statistical hypothesis testing or ranking
+rather than just comparing averages.
 
@@
 
-\subsection{In SPARQL Federation}
+\paragraph{In SPARQL Federation}
 
@@
 
-\subsection{In Benchmarking}
+\paragraph{In Benchmarking}
 
@@
 %\bibliographystyle{plain}
-%\bibliographystyle{plain-csmin}
-\bibliographystyle{splncs03}
+\bibliographystyle{plain-csmin}
+%\bibliographystyle{splncs03}
 \bibliography{selectivity,federation,benchmarks,optimization}

### 2012-01-13 eswc-2012/phdsymp.tex
snip
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@86 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-This part of the thesis addresses problems of query optimization in a
-federated regime.
-
 When using FedBench, (see \cite{Schmidt:2011:FBS:2063016.2063054}),
@@
 
-The above relies on data and queries that are in fact
-representative of the system as a whole. It is often asserted that it
-is not possible to create a ``one-size-fits-all'' benchmark (see
+The above relies on data and queries that are in fact representative
+of the system as a whole. It is often asserted that it is not possible
+to create a ``one-size-fits-all'' benchmark (see
 e.g. \cite{Schmidt:2011:FBS:2063016.2063054}) to cover all aspects of
 an application. While that may be true in the general case, I want to
-challenge the assumption that the best we can do is to accommodate a
-multitude of dimensions and essential challenges in the case of
-SPARQL.
+challenge the assumption that the best we can do is testing queries on
+a best effort basis in the case of SPARQL.
 

### 2012-01-13 eswc-2012/phdsymp.tex
snip reference
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@87 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Thus far, the benchmarking problem has been seen as a software testing
-problem, but we could also ask if the existence of benchmark data can
-be exposed to help federation query optimizers along with a
-statistical digest.
+problem, but as stated in the introduction this is not the only
+objective, we may now see if benchmark data can be exposed to help
+federation query optimizers along with a statistical digest.
 
@@
 Fractional factorial design is covered in elementary text-books in
-statistics but is most likely inadequate for this purpose, so I
-intend to go further into the statistical literature to see if there
-is methodology that is even better suited to the problem. Armed with
+statistics but is most likely inadequate for this purpose, so I intend
+to go further into the statistical literature to see if there is
+methodology that is even better suited to the problem. Armed with
 this, I hope to use amongst other things, the complexity analysis of
-SPARQL (see e.g. \cite{Schmidt:2010:FSQ:1804669.1804675}) to find
-suitable factors to see if my admittedly bold proposition that it is
-possible to design a benchmark to cover all realistic cases.
+SPARQL to find suitable factors to see if my admittedly bold
+proposition that it is possible to design a benchmark to cover all
+realistic cases.
 

### 2012-01-13 eswc-2012/phdsymp.tex
snip
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@88 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The contemporary problem in benchmarking is to distinguish
-implementations when the measured difference is small or susceptible
-to random noise.
-
 Current common practice in benchmarking SPARQL-enabled systems is to
@@
 In this case, we must consider the possibility that the measurements
-are impacted by random noise. For example: The underlying system may
-be doing other tasks. The garbage collection of some languages may
-trigger at not easily foreseeable points. The engine may use
-randomized algorithms, or the data may be of such a nature that it
-triggers worst-case behaviour in situations the query writer cannot
-foresee. All this contributes to that we should consider the
+are impacted by random noise. Sources of random noise may be the
+overall system load, garbage collection triggers, use of randomized
+algorithms, or the data of such a nature that it triggers worst-case
+behaviour, etc. All this contributes to that we should consider the
 performance metric, whatever metric is used, as a \emph{stochastic

### 2012-01-13 eswc-2012/phdsymp.tex
editorial
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@89 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 distinguishing characteristics so that we can return a question that
-will actually help new users of SPARQL engines to chose?
+will actually help new users of SPARQL engines to chose one that meets
+their need?
 

### 2012-01-13 eswc-2012/phdsymp.tex
trim factorial
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@92 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Fractional factorial design is covered in elementary text-books in
+Factorial design is covered in elementary text-books in
 statistics but is most likely inadequate for this purpose, so I intend

### 2012-01-13 eswc-2012/phdsymp.tex
refactor federation sota
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@93 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Introduction}
+\section{Motivation}
 
@@
 
-\section{Problem Definition}
+\section{State of the Art and Open Problems}
 
@@
 
-When using FedBench, (see \cite{Schmidt:2011:FBS:2063016.2063054}),
-the authors note that the absence of statistics has very adverse
-effects on performance as join order is impossible to optimize. It is
-also pointed out that it causes other problems as well, for example a
-high number of requests to federation members.
+I take the technology state of the art to be represented by the
+current basic SPARQL 1.1 Federated Query Working
+Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}.
 
-This problem has already received some attention, \cite{5337556}
-addressed this problem by making it possible to exchange histograms
-and other data. However, histogram approaches generally suffers from
-the problem that they grow too large or become a insufficiently
+A recent scientific treatment of the current specification is in
+\cite{springerlink:10.1007/978-3-642-21064-8-1}. In this paper, the
+authors also show an optimization strategy based on execution order of
+so-called well-designed patterns.
+
+For benchmarking federated query processing strategies,
+\cite{Schmidt:2011:FBS:2063016.2063054} report the development of
+\emph{FedBench}. In this paper, the authors note that the
+absence of statistics has very adverse effects on performance as join
+order is impossible to optimize. It is also pointed out that it causes
+other problems as well, for example a high number of requests to
+federation members. I take this as a key research problem.
+
+This problem has already received some attention, in \cite{5337556} it
+is shown how various statistics-gathering techniques, including
+histograms can be computed to assist the SemWIQ optimizer, written by
+the same authors.  However, histogram approaches generally suffers
+from the problem that they grow too large or become a insufficiently
 accurate digest, especially in face of very heterogeneous data.
+\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees, which may
+alleviate the problem of histogram size, but it may not solve it.
 
@@
 federation query optimizers along with a statistical digest.
-
-\section{State of the Art}
-
-\paragraph{In SPARQL Federation}
-
-I take the technology state of the art to be represented by the
-current basic SPARQL 1.1 Federated Query Working
-Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}.
-
-A recent scientific treatment of the current specification, is in
-\cite{springerlink:10.1007/978-3-642-21064-8-1}. In this paper, the
-authors also show an optimization strategy based on execution order of
-so-called well-designed patterns.
-
-In \cite{5337556}, it is shown how various statistics-gathering
-techniques, including histograms can be computed to assist the SemWIQ
-optimizer, written by the same authors.
-\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees, which may
-alleviate the problem of histogram size, but it may not solve it.
-
-For benchmarking federated query processing strategies,
-\cite{Schmidt:2011:FBS:2063016.2063054} report the development of \emph{FedBench}.
 

### 2012-01-13 eswc-2012/phdsymp.tex
reorg benchmark sota
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@94 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Query federation with SPARQL has attracted much attention
+Query federation with SPARQL, which is a standardized query language
+for the Semantic Web, has attracted much attention
 from industry and academia alike, and four implementations of basic
@@
 \subsection{In Benchmarking}
+
+Numerous benchmarks have been developed for SPARQL, but
+\cite{Duan:2011:AOC:1989323.1989340} showed that currently, most
+benchmarks poorly represent the typical data and queries that is used
+on the Semantic Web.
+
+Most recently, \cite{mxro:Morsey2011DBpedia} addressed some of these
+problems by using real data and real queries from DBpedia.
 
@@
 
-\paragraph{In Benchmarking}
-
-Numerous benchmarks have been developed for SPARQL, but
-\cite{Duan:2011:AOC:1989323.1989340} showed that currently, most
-benchmarks poorly represent the typical data and queries that is used
-on the Semantic Web.
-
-Most recently, \cite{mxro:Morsey2011DBpedia} addressed some of these
-problems by using real data and real queries from DBpedia.
-
-The problems addressed by these studies are almost orthogonal to the
-problems considered by my proposed project. While I have heard of some
-cases where a hypothesis test is used to demonstrate that one
-implementation is better than another, it is seemingly not common
-practice. Furthermore, I have not to date seen any work towards using
-contemporary statistical methods to evaluate the performance of
-software, but I shall admit that my literature study have been limited
-to RDF/OWL databases, and practical benchmarks of databases, file
-systems, scientific software, etc. Thus, the existence of relevant
-references in the deeper computer science literature is a key issue I
-would appreciate discussing at the Symposium.
+The problems addressed by existing benchmarks such as the one quoted
+above are almost orthogonal to the problems considered by my proposed
+project. While I have heard of some cases where a hypothesis test is
+used to demonstrate that one implementation is better than another, it
+is seemingly not common practice. Furthermore, I have not to date seen
+any work towards using contemporary statistical methods to evaluate
+the performance of software, but I shall admit that my literature
+study have been limited to RDF/OWL databases, and practical benchmarks
+of databases, file systems, scientific software, etc. Thus, the
+existence of relevant references in the deeper computer science
+literature is a key issue I would appreciate discussing at the
+Symposium.
 

### 2012-01-13 eswc-2012/phdsymp.tex
methods in fed eval
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@95 43ef1190-90fd-4dd7-a769-d0121f02b80e
skip treatment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@96 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 overall system load, garbage collection triggers, use of randomized
-algorithms, or the data of such a nature that it triggers worst-case
-behaviour, etc. All this contributes to that we should consider the
-performance metric, whatever metric is used, as a \emph{stochastic
-  variable}.
+algorithms, or that the data are of such a nature that it triggers
+worst-case behaviour, etc. All this contributes to that we should
+consider the performance metric, whatever metric is used, as a
+\emph{stochastic variable}.
 
@@
 
-The problems addressed by existing benchmarks such as the one quoted
+The problems addressed by existing benchmarks such as the one cited
 above are almost orthogonal to the problems considered by my proposed
@@
 
+The evaluation methodology for the SPARQL federation work of the
+thesis will largely be covered by running the elaborate benchmark
+designs of the thesis.
+
 \subsection{In Benchmarking}\label{sec:benchmethod}
@@
 experimental design field. One of the simpler designs is ``fractional
-factorial design'', in which several ``factors'' are studied by using
-different ``treatments''. In terms of SPARQL execution, the SPARQL
-engine is clearly a factor, but also for example the nestedness of
-\texttt{OPTIONAL}s can be a factor, number of triples in a basic graph
-pattern, etc, and the treatment is to vary these numbers to different
-``levels''. The key to understanding why this can be efficient is that
-these variations need not occur in the same experiment. Thus, many
-combinations of factors can be studied by carefully designing queries
-to cover different factors, and a formalism called ``resolution'' has
-been developed to classify how well this has been achieved.
+factorial design'', in which several ``factors'' are studied. In terms
+of SPARQL execution, the SPARQL engine is clearly a factor, but also
+for example the nestedness of \texttt{OPTIONAL}s can be a factor,
+number of triples in a basic graph pattern, etc, and these numbers are
+varied to different ``levels''. The key to understanding why this can
+be efficient is that these variations need not occur in the same
+experiment. Thus, many combinations of factors can be studied by
+carefully designing queries to cover different factors, and a
+formalism called ``resolution'' has been developed to classify how
+well this has been achieved, partly answering the question of
+evaluation methodology for this part of the thesis.
 

### 2012-01-13 eswc-2012/phdsymp.tex
Helens tilbakemeldinger
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@97 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   Federation of semantic data on SPARQL endpoints will allow data to
-  remain distributed so that they can be controlled by local curators
+  remain distributed so that it can be controlled by local curators
   and swiftly updated. There are considerable performance problems,
-  which the present work proposes to address mainly by computation
+  which the present work proposes to address, mainly by computation
   and exposure of statistical digests to assist selectivity
@@
 
-To help optimize, SPARQL endpoints should expose details about both
+To aid optimization, SPARQL endpoints should expose details about both
 data and performance characteristics of the engine itself. The
@@
 
-The focus on SPARQL benchmarking is not just motivated from the
-optimization perspective, as I have found the current state of the art
+The focus on SPARQL benchmarking is not only motivated from the
+perspective of optimization, as I have found the current state of the art
 in SPARQL benchmarking lacking in its use of statistics. The emphasis
@@
 
-I am still in an early phase of my work, and I have not yet started to
-explore the scientific literature around SPARQL Federation to any
-depth, currently focusing my efforts on benchmarking. The long-term
-goal of my work is SPARQL Federation, but it is a minor concern in
-this paper.
+I have not yet started to explore the scientific literature around
+SPARQL Federation in any depth as I am still in an early phase of my
+work. I am currently focusing my efforts on benchmarking. The
+long-term goal of my work is SPARQL Federation, but that is a minor
+concern in this paper.
 
@@
 \emph{FedBench}. In this paper, the authors note that the
-absence of statistics has very adverse effects on performance as join
+absence of statistics has very adverse effects on performance as the join
 order is impossible to optimize. It is also pointed out that it causes
@@
 is shown how various statistics-gathering techniques, including
-histograms can be computed to assist the SemWIQ optimizer, written by
-the same authors.  However, histogram approaches generally suffers
+histograms, can be computed to assist the SemWIQ optimizer, written by
+the same authors.  However, histogram approaches generally suffer
 from the problem that they grow too large or become a insufficiently
-accurate digest, especially in face of very heterogeneous data.
+accurate digest, especially in the face of very heterogeneous data.
 \cite{Harth:2010:DSO:1772690.1772733} introduced QTrees, which may
-alleviate the problem of histogram size, but it may not solve it.
+alleviate the problem of histogram size, but which may not solve it.
 
-Thus, the core problem is: How do we compute and expose a digest that
+Therefore, the core problem is: How do we compute and expose a digest that
 is of optimal size for the query performance problem?
@@
 Numerous benchmarks have been developed for SPARQL, but
-\cite{Duan:2011:AOC:1989323.1989340} showed that currently, most
-benchmarks poorly represent the typical data and queries that is used
+\cite{Duan:2011:AOC:1989323.1989340} showed that currently most
+benchmarks poorly represent the typical data and queries that are used
 on the Semantic Web.
@@
 measured, for example the time it takes for the engine to return the
-full result. Commonly, this process is repeated a number of times, and
+full result. Commonly, this process is repeated a number of times and
 an average response time is found. The performance of different
-engines can then be compared based on these averages.
+engines can then be compared, based on these averages.
 
 In many cases, this is sufficient. Sometimes, one engine can execute a
-query an order of magnitude faster than another, in which case, there
-is hardly reasonable doubt which is faster. Small differences may seem
+query in an order of magnitude faster than another, in which case, there
+is hardly reasonable doubt as to which is faster. Small differences may seem
 unimportant but may become important if they are systematic. Even if
@@
 overall system load, garbage collection triggers, use of randomized
-algorithms, or that the data are of such a nature that it triggers
+algorithms, or that the data is of such a nature that it triggers
 worst-case behaviour, etc. All this contributes to that we should
@@
 
-Doing this opens new methodological possibilities, first and foremost
+This opens new methodological possibilities, first and foremost
 using well-established statistical hypothesis testing or ranking
@@
 
-With the currently published benchmarks, it is up to the user to
-select an engine with the best characteristics relative to their own
+Within the currently published benchmarks, it is up to the user to
+select an engine with the best characteristics, relative to their own
 expected application. However, I am often asked the question: ``what
-is the best general-purpose triple store?'' and those who ask are not
-prepared to provide any further qualification. This prompts the
-question: Can we collapse all the metrics into a single \emph{test
+is the best general-purpose triple store?'' Those who ask are not
+prepared to expand on the question. Thus, we should ask ourselves: 
+Can we collapse all the metrics into a single \emph{test
   statistic} that can summarize the performance of the engine as a
-whole? Also, can we provide a system to developers that they can use
-to see if any changes they do to the engine has unexpected adverse
+whole? Also, can we provide a system for developers that they can use
+to see if any changes they do to the engine can have unexpected adverse
 effects on other parts of the system? For those who are able to
@@
 will actually help new users of SPARQL engines to chose one that meets
-their need?
+their needs?
 
-The above relies on data and queries that are in fact representative
+The above relies on data and queries that are representative
 of the system as a whole. It is often asserted that it is not possible
@@
 that one variable. In medical science, this has been abandoned several
-decades ago, thanks to advances in statistics. In e.g. a case where
+decades ago, thanks to advances in statistics. In for example a case where
 the researcher administrates different treatments to terminally ill
@@
 that it is possible to design an experiment (i.e. a benchmark) so that
-we can prove it covers all realistic cases and that we can justify
+we can prove it covers all realistic cases and with which we can justify
 why the remaining corner cases are unlikely to influence the
@@
 
-Thus far, the benchmarking problem has been seen as a software testing
+So far, the benchmarking problem has been seen as a software testing
 problem, but as stated in the introduction this is not the only
@@
 used to demonstrate that one implementation is better than another, it
-is seemingly not common practice. Furthermore, I have not to date seen
+seems not to be common practice. Furthermore, I have not to date seen
 any work towards using contemporary statistical methods to evaluate
 the performance of software, but I shall admit that my literature
-study have been limited to RDF/OWL databases, and practical benchmarks
-of databases, file systems, scientific software, etc. Thus, the
+study has been limited to RDF/OWL databases, and practical benchmarks
+of databases, file systems, scientific software, etc. Therefore, the
 existence of relevant references in the deeper computer science
@@
 There are many possible approaches for this part of the thesis. As I
-expect great advances to be made before I start tackling this problem,
-I have not chosen any methodology, but work to find more
-space-efficient ways to expose statistics in the service description
-and standardize those seems like an interesting direction.
+expect substantial advances to be made before I start tackling this
+problem, I have not chosen any methodology, but an interesting
+direction for work seem to be to find more space-efficient ways to
+expose statistics in the service description and standardize them.
 
-To this end, I briefly looked into two approaches:
+To this end, I have briefly looked into two approaches:
 \cite{Getoor:2001:SEU:375663.375727} used Bayesian Networks and
@@
 Another approach that I have not seen used in the literature is to use
-parametrized statistics. This would amount to attempt fitting data to
+parametrized statistics. This would amount to an attempt to fit data to
 a known distribution function and expose which distribution and its
@@
 data, so the adaption of existing techniques to such situations may
-also be a interesting problem.
+also be an interesting problem.
 
@@
 factorial design'', in which several ``factors'' are studied. In terms
-of SPARQL execution, the SPARQL engine is clearly a factor, but also
-for example the nestedness of \texttt{OPTIONAL}s can be a factor,
-number of triples in a basic graph pattern, etc, and these numbers are
+of SPARQL execution, the SPARQL engine is clearly a factor, but also,
+for example, the nestedness of \texttt{OPTIONAL}s can be a factor,
+or the number of triples in a basic graph pattern, etc. These numbers are
 varied to different ``levels''. The key to understanding why this can
 be efficient is that these variations need not occur in the same
-experiment. Thus, many combinations of factors can be studied by
+experiment. Therefore, many combinations of factors can be studied by
 carefully designing queries to cover different factors, and a
@@
 statistics but is most likely inadequate for this purpose, so I intend
-to go further into the statistical literature to see if there is
-methodology that is even better suited to the problem. Armed with
-this, I hope to use amongst other things, the complexity analysis of
+to go further into the statistical literature to see if there is a
+methodology that is even better suited to the problem. Using
+this, I hope to use, amongst other things, the complexity analysis of
 SPARQL to find suitable factors to see if my admittedly bold
 proposition that it is possible to design a benchmark to cover all
-realistic cases.
+realistic cases can be shown to hold.
 

### 2012-01-16 eswc-2012/phdsymp.tex
start on punchline
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@98 43ef1190-90fd-4dd7-a769-d0121f02b80e
coefficients and stuff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@99 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+With this analysis, I speculate based on superficial experience that
+certain estimated coefficients can be exposed in the service
+description to give federated query engines what they need to optimize
+for performance characteristics of certain SPARQL
+implementations. Adjusting how different features are weighed will
+help those who know details of their expected SPARQL use and
+normalization-techniques of the weights will answer the special case
+of ``the best general-purpose SPARQL engine''.
+
 %\section{Conclusion}

### 2012-01-16 eswc-2012/phdsymp.tex
compare existing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@100 43ef1190-90fd-4dd7-a769-d0121f02b80e
remove TeX comments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@101 43ef1190-90fd-4dd7-a769-d0121f02b80e
trimming
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@102 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 work\footnote{http://www.w3.org/2009/sparql/wiki/Feature:BasicFederatedQuery}. 
-The basic query federation feature was
-supported by a large number of group members, and the Last Call working
+This feature was
+supported by a many group members, and the Last Call working
 draft of the proposed standard was published on 17~November~2011.
@@
 Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}. 
-
 A recent scientific treatment of the current specification is in
@@
 on the Semantic Web.
-
 Most recently, \cite{mxro:Morsey2011DBpedia} addressed some of these
@@
 measured, for example the time it takes for the engine to return the
-full result. Commonly, this process is repeated a number of times and
-an average response time is found. The performance of different
-engines can then be compared, based on these averages.
+full result. Since there is a certain randomness in query times, this
+process is repeated a number of times and an average response time is
+found. The performance of different engines can then be compared,
+based on these averages.
 
@@
 
-In this case, we must consider the possibility that the measurements
-are impacted by random noise. Sources of random noise may be the
-overall system load, garbage collection triggers, use of randomized
-algorithms, or that the data is of such a nature that it triggers
-worst-case behaviour, etc. All this contributes to that we should
-consider the performance metric, whatever metric is used, as a
-\emph{stochastic variable}.
+In this case, we must consider the possibility that the random noise
+can influence the conclusions. Whatever metric is used, it should be
+treated as a \emph{stochastic variable}.
 
@@
 
-This part of the thesis seeks to use advances from statistical
-experimental planning to improve accuracy and dependability of
-benchmarking. Already in 1926, Ronald~Fischer noted that complex
-experiments can be much more efficient than simple ones\footnote{Cited
-  in http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
+Already in 1926, Ronald~Fischer noted that complex experiments can be
+much more efficient than simple ones\footnote{Cited in
+  http://en.wikipedia.org/wiki/Factorial\_experiment}, starting the
 experimental design field. One of the simpler designs is ``fractional
@@
 of SPARQL execution, the SPARQL engine is clearly a factor, but also,
-for example, the nestedness of \texttt{OPTIONAL}s can be a factor,
-or the number of triples in a basic graph pattern, etc. These numbers are
+for example, the nestedness of \texttt{OPTIONAL}s can be a factor, or
+the number of triples in a basic graph pattern, etc. These numbers are
 varied to different ``levels''. The key to understanding why this can
@@
 well this has been achieved, partly answering the question of
-evaluation methodology for this part of the thesis.
+evaluation methodology for this part of the thesis. We should also
+validate by comparing this benchmark with conclusions from existing
+benchmarks.
 
@@
 
-%\section{Conclusion}
-
-%\bibliographystyle{plain}
 \bibliographystyle{plain-csmin}

### 2012-01-17 eswc-2012/phdsymp.tex
siste Audun-rettinger
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@103 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 This feature was
-supported by a many group members, and the Last Call working
+supported by many group members, and the Last Call working
 draft of the proposed standard was published on 17~November~2011.
@@
 the same authors.  However, histogram approaches generally suffer
-from the problem that they grow too large or become a insufficiently
+from the problem that they grow too large or become an insufficiently
 accurate digest, especially in the face of very heterogeneous data.
@@
 
-With this analysis, I speculate based on superficial experience that
-certain estimated coefficients can be exposed in the service
-description to give federated query engines what they need to optimize
-for performance characteristics of certain SPARQL
-implementations. Adjusting how different features are weighed will
-help those who know details of their expected SPARQL use and
-normalization-techniques of the weights will answer the special case
-of ``the best general-purpose SPARQL engine''.
+With this analysis, I speculate based on superficial experience with
+factorial designs and analysis of variance that certain estimated
+coefficients can be exposed in the service description to give
+federated query engines assistance in optimizing for performance
+characteristics of certain SPARQL implementations. Adjusting how
+different features are weighed will help those who know details of
+their expected SPARQL use and normalization-techniques of the weights
+will answer the special case of ``the best general-purpose SPARQL
+engine''.
 

### 2012-01-17 eswc-2012/phdsymp.tex
siste småplukk
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@104 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 expected application. However, I am often asked the question: ``what
-is the best general-purpose triple store?'' Those who ask are not
+is the best general-purpose triple store?'' Those who ask are usually not
 prepared to expand on the question. Thus, we should ask ourselves: 
@@
 
-The above relies on data and queries that are representative
-of the system as a whole. It is often asserted that it is not possible
-to create a ``one-size-fits-all'' benchmark (see
+The above relies on data and queries that are representative of the
+system as a whole. It is often asserted that it is not possible to
+create a ``one-size-fits-all'' benchmark (see
 e.g. \cite{Schmidt:2011:FBS:2063016.2063054}) to cover all aspects of
 an application. While that may be true in the general case, I want to
-challenge the assumption that the best we can do is testing queries on
-a best effort basis in the case of SPARQL.
+challenge the assumption that a best effort approach is all we can do
+in the case of SPARQL.
 

### 2012-01-20 eswc-2012/phdsymp.tex
no entailment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@105 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 challenge the assumption that a best effort approach is all we can do
-in the case of SPARQL.
+in the case of SPARQL (excluding entailment).
 
@@
 \bibliographystyle{plain-csmin}
+%\bibliographystyle{jbact}
 %\bibliographystyle{splncs03}

### 2012-01-20 eswc-2012/phdsymp.tex
review paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@106 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 current basic SPARQL 1.1 Federated Query Working
-Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}. 
+Draft\footnote{http://www.w3.org/TR/2011/WD-sparql11-federated-query-20111117/}. In
+addition, many have implemented federation that doesn't require 
+explicit references to service endpoints, e.g.~\cite{springerlink:10.1007/978-3-642-25073-6-38}.
 A recent scientific treatment of the current specification is in
@@
 
-For benchmarking federated query processing strategies,
-\cite{Schmidt:2011:FBS:2063016.2063054} report the development of
-\emph{FedBench}. In this paper, the authors note that the
-absence of statistics has very adverse effects on performance as the join
-order is impossible to optimize. It is also pointed out that it causes
-other problems as well, for example a high number of requests to
-federation members. I take this as a key research problem.
+A recent review of the state of the art is in
+\cite{springerlink:10.1007/978-3-642-17551-0-5}. In addition,
+\cite{springerlink:10.1007/978-3-642-25073-6-38} proposes
+\emph{bound joins} and proves it can dramatically reduce the number of
+requests needed to federation members as well as the implementation of
+FedX.
 
-This problem has already received some attention, in \cite{5337556} it
-is shown how various statistics-gathering techniques, including
-histograms, can be computed to assist the SemWIQ optimizer, written by
-the same authors.  However, histogram approaches generally suffer
-from the problem that they grow too large or become an insufficiently
-accurate digest, especially in the face of very heterogeneous data.
-\cite{Harth:2010:DSO:1772690.1772733} introduced QTrees, which may
-alleviate the problem of histogram size, but which may not solve it.
+It has been my intention to focus on the two problems listed in
+section~3.3.1
+in~\cite{springerlink:10.1007/978-3-642-17551-0-5}. Notably, histogram
+approaches generally suffer from the problem that they grow too large
+or become an insufficiently accurate digest, especially in the face of
+very heterogeneous data.  \cite{Harth:2010:DSO:1772690.1772733}
+introduced QTrees, which may alleviate the problem of histogram size,
+but which may not solve it.
 
@@
 Most recently, \cite{mxro:Morsey2011DBpedia} addressed some of these
-problems by using real data and real queries from DBpedia.
+problems by using real data and real queries from
+DBpedia. \cite{Schmidt:2011:FBS:2063016.2063054} has developed a
+benchmark for the federated case.
 

### 2012-01-20 eswc-2012/phdsymp.tex
ease the hard formulations above benchmarks somewhat
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@108 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 system as a whole. It is often asserted that it is not possible to
-create a ``one-size-fits-all'' benchmark (see
-e.g. \cite{Schmidt:2011:FBS:2063016.2063054}) to cover all aspects of
-an application. While that may be true in the general case, I want to
-challenge the assumption that a best effort approach is all we can do
-in the case of SPARQL (excluding entailment).
+create a ``one-size-fits-all'' benchmark but answering the above
+question requires it.
 
@@
 Using techniques from statistical experimental design, I propose
-that it is possible to design an experiment (i.e. a benchmark) so that
-we can prove it covers all realistic cases and with which we can justify
+that it is possible to design an experiment (i.e. a benchmark) which
+makes it possible to cover all realistic cases and with which we can justify
 why the remaining corner cases are unlikely to influence the

### 2012-01-20 eswc-2012/phdsymp.tex
differences in query times
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@109 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In many cases, this is sufficient. Sometimes, one engine can execute a
-query in an order of magnitude faster than another, in which case, there
-is hardly reasonable doubt as to which is faster. Small differences may seem
-unimportant but may become important if they are systematic. Even if
-one engine is dramatically better than another in one case, small
-deficiencies may add up to make the other a better choice for most
-applications anyway.
+query in an order of magnitude faster than another. If this happens
+systematically for many different queries, there is hardly reasonable
+doubt as to which is faster. In most cases, the query response times
+differs little, however. Small differences may seem unimportant
+but may become important if they are systematic. Even if one engine is
+dramatically better than another in one case, small deficiencies may
+add up to make the other a better choice for most applications anyway.
 
@@
 create a ``one-size-fits-all'' benchmark but answering the above
-question requires it.
+question requires such a benchmark.
 

### 2012-01-20 eswc-2012/phdsymp.tex
plural
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@110 43ef1190-90fd-4dd7-a769-d0121f02b80e
minor cleanup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@111 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Within the currently published benchmarks, it is up to the user to
+With the currently published benchmarks, it is up to the user to
 select an engine with the best characteristics, relative to their own
@@
 
-The problems addressed by existing benchmarks such as the one cited
+The problems addressed by existing benchmarks such as the ones cited
 above are almost orthogonal to the problems considered by my proposed
@@
 
-Factorial design is covered in elementary text-books in
-statistics but is most likely inadequate for this purpose, so I intend
-to go further into the statistical literature to see if there is a
-methodology that is even better suited to the problem. Using
-this, I hope to use, amongst other things, the complexity analysis of
-SPARQL to find suitable factors to see if my admittedly bold
-proposition that it is possible to design a benchmark to cover all
-realistic cases can be shown to hold.
+Factorial design inspires the present work and is covered in
+elementary text-books in statistics but is inadequate for this
+purpose, so I intend to go further into the statistical literature to
+see if there is a methodology that is better suited to the
+problem. Using this, I hope to use, amongst other things, the
+complexity analysis of SPARQL to find suitable factors to see if my
+admittedly bold proposition that it is possible to design a benchmark
+to cover all realistic cases can be shown to hold.
 

### 2012-02-17 eswc-2012/workshop.tex
Stub workshop paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@113 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass{llncs}
%\documentclass{article}
\usepackage{cite}

\title{Practical implications of HATEOAS on a bridge between LOD and SPARQL}
\titlerunning{HATEOAS and a bridge between LOD and SPARQL}
\author{Kjetil Kjernsmo\inst{1}}
\institute{Department of Informatics,
Postboks 1080 Blindern,
0316 Oslo, Norway
\email{kjekje@ifi.uio.no}}

\begin{document}

\maketitle

\begin{abstract}
This paper argues two things: 1) SPARQL is a necessity for non-trivial
navigation in Linked Data, 2) The community needs to pay attention to
the HATEOAS constraint of REST and its practical implications for
making a bridge between LOD and SPARQL possible.

\end{abstract}

%\bibliographystyle{plain-csmin}
%\bibliographystyle{jbact}
%\bibliographystyle{splncs03}
%\bibliography{selectivity,federation,benchmarks,optimization}

\end{document}

### 2012-02-27 eswc-2012/workshop.tex
introduction to HATEOAS
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@115 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
-This paper argues two things: 1) SPARQL is a necessity for non-trivial
+This paper makes two points: 1) SPARQL is a necessity for non-trivial
 navigation in Linked Data, 2) The community needs to pay attention to
@@
 \end{abstract}
+
+\section{Introduction}
+
+Some bloggers in the developer community, see e.g.~\cite{sunsetonmvc},
+is currently predicting a paradigm shift in common practices, prompted
+by the trend towards richer clients. This has caused a departure from
+the time-tested Model-View-Controller pattern that has been the
+dominant paradigm in web development for more than a decade.
+
+Semantic Web services is likely to be an important source of data for
+future applications, but for foreseeable future just one of many. A
+key promise of Linked Open Data (LOD) in particular, and the Semantic
+Web in general is the ability to integrate many data sources
+easily. However, this integration requires links and it requires that
+the resulting graph can be traversed by reasonable means. While links
+are quite abundant across the LOD cloud, this paper will argue that
+key links are missing to make it possible to traverse the resulting
+graph. Moreover, we will argue that this shortcoming is due to that
+the community does not adequately take into account the constraint
+known as ``Hypermedia As The Engine Of Application State'' (commonly
+abbreviated HATEOAS) from the REST architectural style, see
+\cite{fielding}~Chapter~5.
+
+The HATEOAS constraint requires that the application can navigate from
+one resources to another using the hypermedia links in the present
+resource \emph{only}, it should not require any out-of-band
+information to do so. Thus, a RESTful API should not require prior
+knowledge of a URI structure, but it does not forbid it as long as the
+same information can be gleened from the resources. Nor does it
+require that all resources can be reached from any other resource. As
+such, the constraint is not very strict, one could argue that as long
+as there are some links, the API satisfies the HATEOAS constraint.cc
+% TODO: It is practical
 

### 2012-02-27 eswc-2012/workshop.tex
practical applications is important
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@116 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 such, the constraint is not very strict, one could argue that as long
-as there are some links, the API satisfies the HATEOAS constraint.cc
-% TODO: It is practical
+as there are some links, the API satisfies the HATEOAS constraint.
+
+Since RDF is built on URIs, many of which can be dereferenced to
+obtain further links, it lends itself well to RESTful APIs, but it is
+debateable whether an API can be said to be RESTful if the links
+doesn't enable any useful interactions. Therefore, whether the HATEOAS
+constraint is satisfied must be judged on the basis of the practical
+applications the API enables.
 

### 2012-02-27 eswc-2012/workshop.tex
lowercase
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@117 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the community does not adequately take into account the constraint
-known as ``Hypermedia As The Engine Of Application State'' (commonly
+known as ``Hypermedia as the Engine of Application State'' (commonly
 abbreviated HATEOAS) from the REST architectural style, see

### 2012-02-27 eswc-2012/workshop.tex
minor stuff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@118 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \end{abstract}
+
+\section*{BRAINDUMP STATUS}
 
@@
 information to do so. Thus, a RESTful API should not require prior
-knowledge of a URI structure, but it does not forbid it as long as the
-same information can be gleened from the resources. Nor does it
-require that all resources can be reached from any other resource. As
-such, the constraint is not very strict, one could argue that as long
-as there are some links, the API satisfies the HATEOAS constraint.
+knowledge of a URI structure, but one may define the API in terms of
+URI structure as long as the same information can be gleened from the
+resources. Nor does it require that all resources can be reached from
+any other resource. As such, the constraint is not very strict, one
+could argue that as long as there are some links, the API satisfies
+the HATEOAS constraint.
 

### 2012-02-27 eswc-2012/workshop.tex
one question; more links
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@119 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+The LOD Cloud consists of well-linked datasets and to find simple
+answers by simply dereferencing URIs, parse the returned content and
+address it by conventional methods provided by the programming
+language of choice, and this pattern has proved very useful. However,
+for slightly more complex applications, one has to traverse a
+relatively large graph to find the data one requires. In most cases,
+one would need to match the graph with a pattern, which is essentially
+what the SPARQL query language does. The main problem is that there
+may not be a link present that makes it possible for the application
+to find the query endpoint without requiring out-of-band information,
+which implies that non-REST interactions are required. To remedy this
+situation, we would need to add more links.
+
+\section{Questions for discussion}
+
+\begin{enumerate}
+\item TimBL suggests in the Linked Data Design
+  Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} the
+  use of ``Minimum Spanning Graph'', what other triples would humans
+  and machines find useful?
+
+\end{enumerate}
+
 %\bibliographystyle{plain-csmin}

### 2012-02-27 eswc-2012/workshop.tex
add timbl quote
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@120 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 one would need to match the graph with a pattern, which is essentially
-what the SPARQL query language does. The main problem is that there
-may not be a link present that makes it possible for the application
-to find the query endpoint without requiring out-of-band information,
-which implies that non-REST interactions are required. To remedy this
-situation, we would need to add more links.
+what the SPARQL query language does. In the Linked Data Design
+Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} Tim
+Berners-Lee notes ``To make the data be effectively linked, someone
+who only has the URI of something must be able to find their way the
+SPARQL endpoint.'', but this is generally not possible today without
+requiring out-of-band information, which implies that non-REST
+interactions are required. To remedy this situation, we would need to
+add more links.
 
@@
 \begin{enumerate}
-\item TimBL suggests in the Linked Data Design
+\item Tim Berners-Lee suggests in the Linked Data Design
   Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} the

### 2012-02-27 eswc-2012/workshop.tex
section heads
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@121 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
-This paper makes two points: 1) SPARQL is a necessity for non-trivial
-navigation in Linked Data, 2) The community needs to pay attention to
+This paper makes two points: 1) SPARQL is a necessity % TODO: relax 
+for non-trivial navigation in Linked Data, 
+2) The community needs to pay attention to
 the HATEOAS constraint of REST and its practical implications for
@@
 
+\section{Non-trivial graph traversal}
+
+\section{Finding the way with voiD}
+
 \section{Questions for discussion}

### 2012-02-27 eswc-2012/workshop.tex
Turtle that provides some links
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@122 43ef1190-90fd-4dd7-a769-d0121f02b80e
more concise in data set
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@123 43ef1190-90fd-4dd7-a769-d0121f02b80e
not pubby
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@124 43ef1190-90fd-4dd7-a769-d0121f02b80e
bashing graph store
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@126 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 abbreviated HATEOAS) from the REST architectural style, see
-\cite{fielding}~Chapter~5.
+\cite{Fielding_2000_Architectural-Styles}~Chapter~5.
 
@@
 
+Note that the bridge between LOD and SPARQL discussed in this paper is
+different from the bridge provided by e.g. Pubby or RDF::LinkedData
+and from the Linked Data API that is written to simplify use of linked
+data by web developers\footnote{http://code.google.com/p/linked-data-api/}.
+
 \section{Non-trivial graph traversal}
 
-\section{Finding the way with voiD}
+\section{Lacking links in standards}
+
+The SPARQL 1.1 Graph Store HTTP
+Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},
+currently with a Last Call status, claims to define a protocol in the
+REST architectural style. It is not clear how it satisfies the
+constraints of REST. On one hand, the strong ties to HTTP is overly
+restrictive from a REST point of view, on the other, the specification
+contains no mention today of how hypermedia (i.e. RDF in this case)
+will drive interaction, which is the critical part of a using the REST
+architectural style. Moreover, it defines out-of-band URI keywords
+(\texttt{graph}). As mentioned in the introduction, this is not
+necessarily a violation of architectural constraints, but the
+specification has no mention of how applications can interact to
+achieve the same tasks as it could with this out-of-band information.
+
+This represents a lost opportunity to consider the practical
+implications of HATEOAS to come up with a clearly RESTful specification.
+
+% TODO: more than one URI for graph traversal
+
+\section{Finding the way with VoID}
+
+<> void:inDataset [ void:sparqlEndpoint </sparql> . ] .
 
@@
 
-%\bibliographystyle{plain-csmin}
+\bibliographystyle{plain-csmin}
 %\bibliographystyle{jbact}
 %\bibliographystyle{splncs03}
-%\bibliography{selectivity,federation,benchmarks,optimization}
+\bibliography{federation,webarch,webframeworks}
 

### 2012-02-29 eswc-2012/workshop.tex
more headers
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@127 43ef1190-90fd-4dd7-a769-d0121f02b80e
radical RESTafarianism
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@128 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Finding the way with VoID}
+\section{Proposed solutions}
 
+Prefixes are in the following omitted for brevity.
+
+\subsection{Simple hypermedia vocabulary}
+
+To help drive interaction through hypermedia, the RDF information
+resources has to contain information about the URIs that can be used
+to perform operations other than pure reads. To this end, we suggest a
+simple vocabulary that can be used to supply some statements with all
+resources that the client may be allowed to manipulate.
+
+In keeping with the REST philosophy that the representation should be
+independent of the protocol that is used to transfer it, the
+vocabulary is independent of HTTP, but the explanatory comments uses
+HTTP examples, as this makes the discussion clearer.
+
+\begin{verbatim}
+hm:operations a rdf:Property ;
+              rdfs:range foaf:Document ;
+              rdfs:domain hm:Operations .
+\end{verbatim}
+\texttt{hm:operations} link the resource we want to manipulate,
+typically the current information resource to the operations we may be
+allowed to perform on it.
+
+In all the following examples, the object of the statement is the
+Request-URI that the server suggests is used to manipulate the resource.
+\begin{verbatim}
+hm:mergeInto a rdf:Property ;
+             rdfs:range hm:Operations ;
+             rdfs:domain foaf:Document .
+\end{verbatim}
+\texttt{hm:mergeInto} is used to indicate that a HTTP POST operation
+will do an RDF merge of the request body with the content at the
+Request-URI.
+
+\begin{verbatim}
+hm:serverCreate a rdf:Property ;
+                rdfs:range hm:Operations ;
+                rdfs:domain foaf:Document .
+\end{verbatim}
+\texttt{hm:serverCreate} is used to indicate that a HTTP POST
+operation to the Request-URI will create a new resource with a
+server-supplied URI.
+
+\begin{verbatim}
+hm:delete a rdf:Property ;
+          rdfs:range hm:Operations ;
+          rdfs:domain foaf:Document .
+\end{verbatim}
+\texttt{hm:delete} is used to indicate that the resource should be
+deleted when issuing a HTTP DELETE operation to the Request-URI.
+
+\begin{verbatim}
+hm:put a rdf:Property ;
+       rdfs:range hm:Operations ;
+       rdfs:domain foaf:Document .
+\end{verbatim}
+\texttt{hm:put} is used to indicate that the resource should be
+created using the request body or updated with it if it already exists
+when issuing a HTTP PUT operation to the Request-URI.
+
+\begin{verbatim}
+hm:Operations a rdfs:Class .
+\end{verbatim}
+The class of possible operations.
+
+To manipulate default graphs, properties \texttt{hm:mergeIntoDefault},
+\texttt{hm:deleteDefault} and \texttt{hm:putDefault} are defined
+similarly.
+
+In most cases, the current information resource will be the subject of
+the \texttt{hm:operations} \strong{but also} the object of the allowed
+operations. A typical information resource will supply the following
+triples (given a suitable base URI):
+
+\begin{verbatim}
+<> hm:operations
+   [ a hm:Operations ;
+     hm:mergeInto <> ;
+     hm:delete <> ;
+     hm:put <> ;
+     hm:serverCreate <../> .
+   ] .
+\end{verbatim}
+This represents the common case when you GET a resource, then update
+it with PUT or DELETE or append triples with POST.
+
+\subsection{Finding the way with VoID}
+
+\begin{verbatim}
 <> void:inDataset [ void:sparqlEndpoint </sparql> . ] .
+\end{verbatim}
 
@@
 
+\item What can we do to add more H Factors\footnote{see
+  \url{http://amundsen.com/hypermedia/hfactor/}} to RDF?
+
 \end{enumerate}

### 2012-02-29 eswc-2012/workshop2.tex
Fresh start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@129 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-02-27 eswc-2012/workshop.tex
@@
 
-\title{Practical implications of HATEOAS on a bridge between LOD and SPARQL}
-\titlerunning{HATEOAS and a bridge between LOD and SPARQL}
+\title{The necessity of hypermedia RDF and an approach to achieve it}
+\titlerunning{A hypermedia RDF approach}
 \author{Kjetil Kjernsmo\inst{1}}
@@
 \begin{abstract}
-This paper makes two points: 1) SPARQL is a necessity % TODO: relax 
-for non-trivial navigation in Linked Data, 
-2) The community needs to pay attention to
-the HATEOAS constraint of REST and its practical implications for
-making a bridge between LOD and SPARQL possible.
 
@@
 \section{Introduction}
-
-Some bloggers in the developer community, see e.g.~\cite{sunsetonmvc},
-is currently predicting a paradigm shift in common practices, prompted
-by the trend towards richer clients. This has caused a departure from
-the time-tested Model-View-Controller pattern that has been the
-dominant paradigm in web development for more than a decade.
-
-Semantic Web services is likely to be an important source of data for
-future applications, but for foreseeable future just one of many. A
-key promise of Linked Open Data (LOD) in particular, and the Semantic
-Web in general is the ability to integrate many data sources
-easily. However, this integration requires links and it requires that
-the resulting graph can be traversed by reasonable means. While links
-are quite abundant across the LOD cloud, this paper will argue that
-key links are missing to make it possible to traverse the resulting
-graph. Moreover, we will argue that this shortcoming is due to that
-the community does not adequately take into account the constraint
-known as ``Hypermedia as the Engine of Application State'' (commonly
-abbreviated HATEOAS) from the REST architectural style, see
-\cite{fielding}~Chapter~5.
-
-The HATEOAS constraint requires that the application can navigate from
-one resources to another using the hypermedia links in the present
-resource \emph{only}, it should not require any out-of-band
-information to do so. Thus, a RESTful API should not require prior
-knowledge of a URI structure, but one may define the API in terms of
-URI structure as long as the same information can be gleened from the
-resources. Nor does it require that all resources can be reached from
-any other resource. As such, the constraint is not very strict, one
-could argue that as long as there are some links, the API satisfies
-the HATEOAS constraint.
-
-Since RDF is built on URIs, many of which can be dereferenced to
-obtain further links, it lends itself well to RESTful APIs, but it is
-debateable whether an API can be said to be RESTful if the links
-doesn't enable any useful interactions. Therefore, whether the HATEOAS
-constraint is satisfied must be judged on the basis of the practical
-applications the API enables.
-
-The LOD Cloud consists of well-linked datasets and to find simple
-answers by simply dereferencing URIs, parse the returned content and
-address it by conventional methods provided by the programming
-language of choice, and this pattern has proved very useful. However,
-for slightly more complex applications, one has to traverse a
-relatively large graph to find the data one requires. In most cases,
-one would need to match the graph with a pattern, which is essentially
-what the SPARQL query language does. In the Linked Data Design
-Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} Tim
-Berners-Lee notes ``To make the data be effectively linked, someone
-who only has the URI of something must be able to find their way the
-SPARQL endpoint.'', but this is generally not possible today without
-requiring out-of-band information, which implies that non-REST
-interactions are required. To remedy this situation, we would need to
-add more links.
-
-\section{Non-trivial graph traversal}
-
-\section{Finding the way with voiD}
-
-\section{Questions for discussion}
-
-\begin{enumerate}
-\item Tim Berners-Lee suggests in the Linked Data Design
-  Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} the
-  use of ``Minimum Spanning Graph'', what other triples would humans
-  and machines find useful?
-
-\end{enumerate}
-
-%\bibliographystyle{plain-csmin}
-%\bibliographystyle{jbact}
-%\bibliographystyle{splncs03}
-%\bibliography{selectivity,federation,benchmarks,optimization}
-
-\end{document}

### 2012-02-29 eswc-2012/workshop.tex
strong->emph
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@130 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In most cases, the current information resource will be the subject of
-the \texttt{hm:operations} \strong{but also} the object of the allowed
+the \texttt{hm:operations} \emph{but also} the object of the allowed
 operations. A typical information resource will supply the following

### 2012-03-01 eswc-2012/workshop2.tex
comment vocab
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@131 43ef1190-90fd-4dd7-a769-d0121f02b80e
introduction to classifications
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@132 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
-
+This paper will give an overview of the practical implications of the
+HATEOAS constraint of the REST architectural style, and in that light argue why
+hypermedia RDF is a practical necessity. We will then sketch a
+vocabulary for hypermedia RDF using Mike Amundsen's H~Factor
+classification as motivator. Finally, we will briefly argue that
+SPARQL is important when making non-trivial traversals of Linked Data
+graphs, and see how to a bridge between Linked Data and SPARQL may be
+created with hypermedia RDF.
 \end{abstract}
@@
 \section{Introduction}
+
+%<> hm:can-be hm:put-to ;
+% hm:can-be hm:deleted .
+
+Mike Amundsen defines hypermedia types\footnote{\url{http://amundsen.com/hypermedia/}} as 
+\begin{quote}
+Hypermedia Types are MIME media types that contain native
+hyper-linking semantics that induce application flow. For example,
+HTML is a hypermedia type; XML is not.
+\end{quote}
+Furthermore, he defines a classification scheme called H~Factor as ``a
+measurement of the level of hypermedia support and sophistication of a
+media-type.'' The REST & WOA Wiki defines ``the Hypermedia
+Scale''\footnote{\url{http://restpatterns.org/Articles/The_Hypermedia_Scale}},
+where the categorization is based on capabilities for create, read,
+update and delete operations.  Standing on it's own, RDF is a
+hypermedia type, but only at the \textsf{LO} and \textsf{CL} levels
+on, and just an R~Type (read) on the Hypermedia Scale. We aim at improving
+this situation.

### 2012-03-01 eswc-2012/workshop2.tex
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@133 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 update and delete operations.  Standing on it's own, RDF is a
-hypermedia type, but only at the \textsf{LO} and \textsf{CL} levels
-on, and just an R~Type (read) on the Hypermedia Scale. We aim at improving
+hypermedia type, but only at the \textsf{LO} and \textsf{CL} levels,
+and just an R~Type (read) on the Hypermedia Scale. We aim at improving
 this situation.

### 2012-03-01 eswc-2012/workshop2.tex
rework some older text about requirements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@134 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 this situation.
+
+Amundsen argues in a blog
+post\footnote{\url{http://amundsen.com/blog/archives/1083}} that the
+Semantic Web community should not pursue an API for RDF, but rather
+make sure RDF is a more powerful hypermedia type, like HTML or
+Atom. He argues that a key factor in the success of the Web is that
+messages not only contain data but also application control
+information, and this is needed for the Web to scale.
+
+We also note that some bloggers in the developer community, see
+e.g.~\cite{sunsetonmvc}, is currently predicting a paradigm shift in
+common practices, prompted by the trend towards richer clients. This
+has caused a departure from the time-tested Model-View-Controller
+pattern that has been the dominant paradigm in web development for
+more than a decade.
+
+Semantic Web services is likely to be an important source of data for
+future applications, but for foreseeable future just one of many. A
+key promise of Linked Open Data (LOD) in particular, and the Semantic
+Web in general is the ability to integrate many data sources
+easily. There are two key issues, first read-write operations must be
+similar to how it is done with other hypermedia types to make it
+possible to use generic code to minimize the effort required to
+support RDF. Secondly, it requires relevant links and that the
+resulting graph can be traversed by reasonable means.
+
+While links are highly abundant across the LOD cloud, this paper will
+argue that key links are missing to make it possible to traverse the
+resulting graph and to enable read-write operations. Moreover, we will
+argue that this shortcoming is due to that the community does not
+adequately take into account the constraint known as ``Hypermedia as
+the Engine of Application State'' (commonly abbreviated HATEOAS) from
+the REST architectural style, see
+\cite{Fielding_2000_Architectural-Styles}~Chapter~5.

### 2012-03-01 eswc-2012/workshop2.tex
more updates from the old paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@135 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \cite{Fielding_2000_Architectural-Styles}~Chapter~5.
+
+In a read-only situation, the HATEOAS constraint requires that the
+application can navigate from one resources to another using the
+hypermedia links in the present resource \emph{only}, it should not
+require any out-of-band information to do so. Thus, a RESTful protocol
+should not require prior knowledge of a URI structure, but one may
+define the protocol in terms of URI structure as long as the same
+information can be gleened from the resources. Nor does it require
+that all resources can be reached from any other resource. As such,
+the constraint is not very strict, one could argue that as long as
+there are some links, the protocol satisfies the HATEOAS constraint.
+
+Since RDF is built on URIs, many of which can be dereferenced to
+obtain further links, it lends itself well to RESTful protocols, but
+it is debateable whether a protocol can be said to be RESTful if the
+links doesn't enable any useful interactions. Importantly, if there is
+to be such a thing as a RESTful read-write Semantic Web protocol,
+there must be something in the RDF itself that can be used by
+practical applications to write data. Therefore, whether the HATEOAS
+constraint is satisfied must be judged on the basis of the practical
+applications the protocol enables.

### 2012-03-01 eswc-2012/workshop2.tex
minor language
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@136 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Semantic Web community should not pursue an API for RDF, but rather
-make sure RDF is a more powerful hypermedia type, like HTML or
+work to make RDF a more powerful hypermedia type, like HTML or
 Atom. He argues that a key factor in the success of the Web is that

### 2012-03-01 eswc-2012/workshop2.tex
single endpoint?
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@137 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 applications the protocol enables.
+
+\section{Required links}
+
+As noted in \cite{sunsetonmvc}, AtomPub has a single service endpoint,
+and you can navigate to what you need from there. This motivates the
+first discussion topic of this paper:
+
+\begin{question}
+Is a single service endpoint enough for Linked Data?
+\end{question}
+
+We note that the distributed graph structure of the LOD Cloud makes
+this awkward: For every new data source encountered in a graph
+traversal, a new service endpoint must be queried. Moreover, it would
+be awkward if fine-grained access controls for writing is being used
+to record permissions for all resources in a single service
+description. It seems likely that a few triples attached to each
+information resource is better suited in most cases.

### 2012-03-01 eswc-2012/workshop2.tex
hypermedia RDF
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@138 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 measurement of the level of hypermedia support and sophistication of a
-media-type.'' The REST & WOA Wiki defines ``the Hypermedia
-Scale''\footnote{\url{http://restpatterns.org/Articles/The_Hypermedia_Scale}},
+media-type.'' The REST \& WOA Wiki defines ``the Hypermedia
+Scale''\footnote{\url{http://restpatterns.org/Articles/The\_Hypermedia\_Scale}},
 where the categorization is based on capabilities for create, read,
@@
 information resource is better suited in most cases.
+
+\section{Defining hypermedia RDF}
+
+We noted that for a read-write RESTful protocol, we need to say in the
+RDF message itself what can of operations can be made. We also note
+that only information resources can be manipulated, and we argued that
+it should be possible to add the required triples to every
+resource. We should be able to define hypermedia RDF in terms of a
+minimal vocabulary, but like the H~Factor web page, we shall find it
+instructive to use examples. In the following, we use Turtle syntax
+with prefixes omitted for brevity. Also, we note that in many cases,
+we are making statements about the current resource, which given a
+reasonable base URI can be written as \texttt{<>}. These are the H~Factors:
+
+\begin{description}
+\item[LE] Support for embedded links (HTTP GET)
+
+Not relevant for RDF.
+
+\item[LO] Support for out-bound navigational links (HTTP GET)
+
+Trivially supported by RDF.
+
+\item[LT] Support for templated queries (HTTP GET)
+
+Undefined by this paper. Possibly supported by RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?
+
+\item[LN] Support for non-idempotent updates (HTTP POST)
+
+\begin{verbatim}
+<> hm:canBe hm:mergedInto ;
+   hm:createSimilarAt <../> .
+\end{verbatim}
+
+The first triple says that the current resource can be merged with
+another resource. In a typical HTTP case, this would be achieved by
+POSTing to the current resource with an RDF payload, and the server
+would perform a RDF merge of the payload with the current resource.
+
+The second triple exists to make it possible (in the typical HTTP
+case) to POST an RDF payload to some URI and the server will itself
+assign a URI to the posted RDF payload.
+
+\item[LI] Support for idempotent updates (HTTP PUT, DELETE)
+
+\begin{verbatim}
+<> hm:canBe hm:replaced ,
+            hm:deleted .
+\end{verbatim}
+
+The first triple says that the current resource may be replaced by, in
+the HTTP case, PUTting an RDF payload to the resource URI. The second
+triple says that the resource may be deleted, typically by a HTTP
+DELETE on the resource URI.
+
+\item[CR] Support for modifying control data for read requests
+  (e.g. HTTP Accept-* headers).
+
+\begin{verbatim}
+<> hm:acceptsFormats <http://www.w3.org/ns/formats/Turtle> .
+\end{verbatim}
+
+\item[CU] Support for modifying control data for update requests
+  (e. g. Content-* headers).
+
+Undefined by this paper.
+
+\item[CM] Support for indiciating the interface method for requests
+  (e.g. HTTP GET, POST, PUT, DELETE methods).
+
+E.g.
+
+\begin{verbatim}
+hm:replaced        hm:httpMethod "PUT" .
+hm:deleted         hm:httpMethod "DELETE" .
+hm:mergedInto      hm:httpMethod "POST" .
+hm:createSimilarAt hm:httpMethod "POST" .
+\end{verbatim}
+
+\item[CL] Support for adding semantic meaning to link elements using
+  link relations (e.g. HTML rel attribute).
+
+Trivially supported by RDF.
+
+\end{description}
+
+We see that with a simple vocabulary, RDF can become a very powerful
+hypermedia type. We also note that this vocabulary does enable agents
+to do the same operations as the SPARQL  1.1 Graph Store HTTP
+Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},
+but being fully RESTful. The specification is currently not RESTful
+per the discussion above, even though it was a key design goal. It
+should also be possible to use this on the default (nameless) graph by
+assigning it a name in the service description rather than defining it
+in a out-of-band specification like it is currently being done.
+
+\bibliographystyle{plain-csmin}
+%\bibliographystyle{jbact}
+%\bibliographystyle{splncs03}
+\bibliography{federation,webarch,webframeworks}
+
+\end{document}

### 2012-03-01 eswc-2012/workshop2.tex
use footnotes for all URL refs
Another two questions


git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@139 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 We also note that some bloggers in the developer community, see
-e.g.~\cite{sunsetonmvc}, is currently predicting a paradigm shift in
-common practices, prompted by the trend towards richer clients. This
-has caused a departure from the time-tested Model-View-Controller
-pattern that has been the dominant paradigm in web development for
-more than a decade.
+e.g. a post by Gregg
+Caines\footnote{\url{http://caines.ca/blog/programming/the-sun-is-setting-on-rails-style-mvc-frameworks/}},
+%\cite{sunsetonmvc} 
+is currently predicting a paradigm shift in common practices, prompted
+by the trend towards richer clients. This has caused a departure from
+the time-tested Model-View-Controller pattern that has been the
+dominant paradigm in web development for more than a decade.
 
@@
 
-As noted in \cite{sunsetonmvc}, AtomPub has a single service endpoint,
+%As noted in \cite{sunsetonmvc} 
+AtomPub has a single service endpoint,
 and you can navigate to what you need from there. This motivates the
@@
 
+\begin{question}
+Should the SPARQL 1.1 Graph Store HTTP Protocol be replaced by a
+vocabulary like the above?
+\end{question}
+
+\begin{question}
+Can we create a better vocabulary for hypermedia RDF than the sketch above?
+\end{question}
+
 \bibliographystyle{plain-csmin}

### 2012-03-01 eswc-2012/workshop2.tex
lod sparql bridge
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@140 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Bridging LOD and SPARQL}
+
+In many cases, it is sufficient to get a small number of resources to
+collect the data needed for a given usage, but slightly more involved
+queries (e.g. ``what kind of connections exists between Kate Bush, Roy
+Harper and bands that have sold more than 200 million albums?'') would
+likely cause thousands of resources to be downloaded and examined. For
+this, more advanced graph pattern matching as well as more advanced
+mechanisms for source selection is
+required. \cite{springerlink:10.1007/978-3-642-25073-6-38} provides
+some techniques that should prove very valuable in this respect and so
+it becomes important that SPARQL can be used with Linked Data in a
+RESTful manner.
+
+In the Linked Data Design
+Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} Tim
+Berners-Lee notes ``To make the data be effectively linked, someone
+who only has the URI of something must be able to find their way the
+SPARQL endpoint.'', but this is generally not possible today without
+requiring out-of-band information, which implies that non-REST
+interactions are required.
+
+The triples needed to do this is already defined in
+VoID\footnote{\url{http://vocab.deri.ie/void}} and in use:
+\begin{verbatim}
+<> void:inDataset [ void:sparqlEndpoint </sparql> . ] .
+\end{verbatim}
+
 \bibliographystyle{plain-csmin}

### 2012-03-01 eswc-2012/workshop2.tex
serializations are media types
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@141 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section*{BRAINDUMP STATUS}
-
 \section{Introduction}
-
-%<> hm:can-be hm:put-to ;
-% hm:can-be hm:deleted .
 
@@
 where the categorization is based on capabilities for create, read,
-update and delete operations.  Standing on it's own, RDF is a
-hypermedia type, but only at the \textsf{LO} and \textsf{CL} levels,
-and just an R~Type (read) on the Hypermedia Scale. We aim at improving
-this situation.
+update and delete operations.  Standing on it's own, RDF
+serializations are hypermedia types, but only at the \textsf{LO} and
+\textsf{CL} levels, and just an R~Type (read) on the Hypermedia
+Scale. We aim at improving this situation.
 
@@
 Semantic Web community should not pursue an API for RDF, but rather
-work to make RDF a more powerful hypermedia type, like HTML or
+work to make RDF serializations more powerful hypermedia types, like HTML or
 Atom. He argues that a key factor in the success of the Web is that
@@
 
-We see that with a simple vocabulary, RDF can become a very powerful
-hypermedia type. We also note that this vocabulary does enable agents
-to do the same operations as the SPARQL  1.1 Graph Store HTTP
+We see that with a simple vocabulary, RDF serializations can become a
+very powerful hypermedia types. We also note that this vocabulary does
+enable agents to do the same operations as the SPARQL 1.1 Graph Store
+HTTP
 Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},

### 2012-03-01 eswc-2012/workshop2.tex
cleaning spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@142 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-While links are highly abundant across the LOD cloud, this paper will
+While links are abundant across the LOD cloud, this paper will
 argue that key links are missing to make it possible to traverse the
@@
 define the protocol in terms of URI structure as long as the same
-information can be gleened from the resources. Nor does it require
+information can be gleaned from the resources. Nor does it require
 that all resources can be reached from any other resource. As such,
@@
 obtain further links, it lends itself well to RESTful protocols, but
-it is debateable whether a protocol can be said to be RESTful if the
+it is debatable whether a protocol can be said to be RESTful if the
 links doesn't enable any useful interactions. Importantly, if there is
@@
 
-\item[CM] Support for indiciating the interface method for requests
+\item[CM] Support for indicating the interface method for requests
   (e.g. HTTP GET, POST, PUT, DELETE methods).

### 2012-03-01 eswc-2012/workshop2.tex
conclusion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@143 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Conclusion}
+
+We have shown how RDF serializations can become very powerful
+hypermedia types by using a simple vocabulary. We have argued briefly
+that some triples should be added to every information resource, but
+note that most triples are only relevant to write-operations and
+should only appear in that case. We also noted that SPARQL is
+important for non-trivial traversal of Linked Data. In summary, the
+addition of the following triples would make life easier for
+developers of applications that use Semantic Web data:
+
+\begin{verbatim}
+<> hm:canBe hm:mergedInto, hm:replaced , hm:deleted ;
+   hm:createSimilarAt <../> ;
+   hm:acceptsFormats <http://www.w3.org/ns/formats/Turtle> ;
+   void:inDataset [ void:sparqlEndpoint </sparql> . ] .
+\end{verbatim}
+
 \bibliographystyle{plain-csmin}

### 2012-03-01 eswc-2012/workshop2.tex
collapse hfactors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@144 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 we are making statements about the current resource, which given a
-reasonable base URI can be written as \texttt{<>}. These are the H~Factors:
+reasonable base URI can be written as \texttt{<>}. We have not found
+the following factors to be relevant to RDF: \textsf{LE}, \textsf{CU}
+and \textsf{LT} (though the latter may be supported by
+RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?)  Then
+\textsf{LO} and \textsf{CL} are trivially supported. The remaining are:
 
 \begin{description}
-\item[LE] Support for embedded links (HTTP GET)
 
-Not relevant for RDF.
-
-\item[LO] Support for out-bound navigational links (HTTP GET)
-
-Trivially supported by RDF.
-
-\item[LT] Support for templated queries (HTTP GET)
-
-Undefined by this paper. Possibly supported by RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?
-
-\item[LN] Support for non-idempotent updates (HTTP POST)
+\item[\textsf{LN}] Support for non-idempotent updates (HTTP POST)
 
@@
 
-\item[LI] Support for idempotent updates (HTTP PUT, DELETE)
+\item[\textsf{LI}] Support for idempotent updates (HTTP PUT, DELETE)
 
@@
 
-\item[CR] Support for modifying control data for read requests
+\item[\textsf{CR}] Support for modifying control data for read requests
   (e.g. HTTP Accept-* headers).
@@
 
-\item[CU] Support for modifying control data for update requests
-  (e. g. Content-* headers).
-
-Undefined by this paper.
-
-\item[CM] Support for indicating the interface method for requests
+\item[\textsf{CM}] Support for indicating the interface method for requests
   (e.g. HTTP GET, POST, PUT, DELETE methods).
@@
 \end{verbatim}
-
-\item[CL] Support for adding semantic meaning to link elements using
-  link relations (e.g. HTML rel attribute).
-
-Trivially supported by RDF.
 

### 2012-03-01 eswc-2012/workshop2.tex
grammar...
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@145 43ef1190-90fd-4dd7-a769-d0121f02b80e
gregg gone
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@146 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We also note that some bloggers in the developer community, see
-e.g. a post by Gregg
-Caines\footnote{\url{http://caines.ca/blog/programming/the-sun-is-setting-on-rails-style-mvc-frameworks/}},
-%\cite{sunsetonmvc} 
-is currently predicting a paradigm shift in common practices, prompted
-by the trend towards richer clients. This has caused a departure from
-the time-tested Model-View-Controller pattern that has been the
-dominant paradigm in web development for more than a decade.
-
-Semantic Web services is likely to be an important source of data for
+Semantic Web services are likely to be an important source of data for
 future applications, but for foreseeable future just one of many. A

### 2012-03-01 eswc-2012/workshop2.tex
trim fluff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@147 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Semantic Web community should not pursue an API for RDF, but rather
-work to make RDF serializations more powerful hypermedia types, like HTML or
-Atom. He argues that a key factor in the success of the Web is that
-messages not only contain data but also application control
-information, and this is needed for the Web to scale.
+make RDF serializations more powerful hypermedia types. He argues that
+a key factor in the success of the Web is that messages not only
+contain data but also application control information, and this is
+needed for the Web to scale.
 
@@
 future applications, but for foreseeable future just one of many. A
-key promise of Linked Open Data (LOD) in particular, and the Semantic
-Web in general is the ability to integrate many data sources
-easily. There are two key issues, first read-write operations must be
-similar to how it is done with other hypermedia types to make it
-possible to use generic code to minimize the effort required to
+key promise of the Semantic Web is the ability to integrate many data
+sources easily. There are two key issues, first read-write operations
+must be similar to how it is done with other hypermedia types to make
+it possible to use generic code to minimize the effort required to
 support RDF. Secondly, it requires relevant links and that the
@@
 
-While links are abundant across the LOD cloud, this paper will
+While links are abundant across the Linked Open Data (LOD) cloud, this paper will
 argue that key links are missing to make it possible to traverse the
@@
 SPARQL endpoint.'', but this is generally not possible today without
-requiring out-of-band information, which implies that non-REST
-interactions are required.
+requiring out-of-band information.
 

### 2012-03-01 eswc-2012/workshop2.tex
trim paragraph about URI structure trim triples
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@148 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 hypermedia links in the present resource \emph{only}, it should not
-require any out-of-band information to do so. Thus, a RESTful protocol
-should not require prior knowledge of a URI structure, but one may
-define the protocol in terms of URI structure as long as the same
-information can be gleaned from the resources. Nor does it require
-that all resources can be reached from any other resource. As such,
-the constraint is not very strict, one could argue that as long as
-there are some links, the protocol satisfies the HATEOAS constraint.
+require any out-of-band information.
 
@@
 \begin{verbatim}
-<> hm:canBe hm:replaced ,
-            hm:deleted .
+<> hm:canBe hm:replaced, hm:deleted .
 \end{verbatim}
@@
 
-E.g.
-
+For example (with similar triples for other HTTP methods):
 \begin{verbatim}
 hm:replaced hm:httpMethod "PUT" .
-hm:deleted         hm:httpMethod "DELETE" .
-hm:mergedInto      hm:httpMethod "POST" .
-hm:createSimilarAt hm:httpMethod "POST" .
 \end{verbatim}

### 2012-03-01 eswc-2012/workshop2.tex
punctuation fix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@149 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 and \textsf{LT} (though the latter may be supported by
-RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?)  Then
+RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?); and
 \textsf{LO} and \textsf{CL} are trivially supported. The remaining are:

### 2012-03-05 eswc-2012/workshop2.tex
clarifying which spec
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@150 43ef1190-90fd-4dd7-a769-d0121f02b80e
Helen's suggestions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@151 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 where the categorization is based on capabilities for create, read,
-update and delete operations.  Standing on it's own, RDF
+update and delete operations.  Considered on their own, RDF
 serializations are hypermedia types, but only at the \textsf{LO} and
@@
 a key factor in the success of the Web is that messages not only
-contain data but also application control information, and this is
+contain data but also application control information, and that this is
 needed for the Web to scale.
@@
 Semantic Web services are likely to be an important source of data for
-future applications, but for foreseeable future just one of many. A
+future applications, but for foreseeable future they are just one of many. A
 key promise of the Semantic Web is the ability to integrate many data
 sources easily. There are two key issues, first read-write operations
-must be similar to how it is done with other hypermedia types to make
+must be similar to how it is done with other hypermedia types, to make
 it possible to use generic code to minimize the effort required to
@@
 application can navigate from one resources to another using the
-hypermedia links in the present resource \emph{only}, it should not
+hypermedia links in the present resource \emph{only}, they should not
 require any out-of-band information.
@@
 it is debatable whether a protocol can be said to be RESTful if the
-links doesn't enable any useful interactions. Importantly, if there is
+links don't enable any useful interactions. Importantly, if there is
 to be such a thing as a RESTful read-write Semantic Web protocol,
@@
 traversal, a new service endpoint must be queried. Moreover, it would
-be awkward if fine-grained access controls for writing is being used
+be awkward if fine-grained access controls for writing are being used
 to record permissions for all resources in a single service
 description. It seems likely that a few triples attached to each
-information resource is better suited in most cases.
+information resource are better suited in most cases.
 
@@
 We noted that for a read-write RESTful protocol, we need to say in the
-RDF message itself what can of operations can be made. We also note
+RDF message itself what kind of operations can be made. We also note
 that only information resources can be manipulated, and we argued that
@@
 resource. We should be able to define hypermedia RDF in terms of a
-minimal vocabulary, but like the H~Factor web page, we shall find it
-instructive to use examples. In the following, we use Turtle syntax
+minimal vocabulary, but like the H~Factor web page, we will find it
+instructive to use examples. In the following, we use Turtle syntax,
 with prefixes omitted for brevity. Also, we note that in many cases,
@@
 We see that with a simple vocabulary, RDF serializations can become a
-very powerful hypermedia types. We also note that this vocabulary does
-enable agents to do the same operations as the SPARQL 1.1 Graph Store
+very powerful hypermedia types. We also note that this vocabulary
+enables agents to do the same operations as the SPARQL 1.1 Graph Store
 HTTP
 Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},
-but being fully RESTful. The specification is currently not RESTful
+while being fully RESTful. The Protocol specification is currently not RESTful
 per the discussion above, even though it was a key design goal. It
@@
 assigning it a name in the service description rather than defining it
-in a out-of-band specification like it is currently being done.
+in an out-of-band specification like is currently being done.
 
@@
 likely cause thousands of resources to be downloaded and examined. For
-this, more advanced graph pattern matching as well as more advanced
-mechanisms for source selection is
+this, more advanced graph pattern matching, as well as more advanced
+mechanisms for source selection, is
 required. \cite{springerlink:10.1007/978-3-642-25073-6-38} provides
@@
 
-The triples needed to do this is already defined in
-VoID\footnote{\url{http://vocab.deri.ie/void}} and in use:
+The triples needed to do this are already defined in
+VoID\footnote{\url{http://vocab.deri.ie/void}} and are in use:
 \begin{verbatim}
@@
 should only appear in that case. We also noted that SPARQL is
-important for non-trivial traversal of Linked Data. In summary, the
+important for non-trivial traversal of Linked Data. To sum up, the
 addition of the following triples would make life easier for

### 2012-03-05 eswc-2012/workshop2.tex
for doing CRUD ?
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@152 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Scale''\footnote{\url{http://restpatterns.org/Articles/The\_Hypermedia\_Scale}},
-where the categorization is based on capabilities for create, read,
+where the categorization is based on capabilities for doing create, read,
 update and delete operations.  Considered on their own, RDF

### 2012-03-05 eswc-2012/workshop2.tex
CRUD operations
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@153 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-03-05 eswc-2012/workshop2.tex
@@
 Scale''\footnote{\url{http://restpatterns.org/Articles/The\_Hypermedia\_Scale}},
-where the categorization is based on capabilities for create, read,
-update and delete operations.  Considered on their own, RDF
+where the categorization is based on capabilities to do CRUD (Create, Read,
+Update and Delete) operations.  Considered on their own, RDF
 serializations are hypermedia types, but only at the \textsf{LO} and

### 2012-03-06 eswc-2012/workshop2.tex
point to footnotes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@154 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-03-05 eswc-2012/workshop2.tex
@@
 Scale''\footnote{\url{http://restpatterns.org/Articles/The\_Hypermedia\_Scale}},
-where the categorization is based on capabilities for create, read,
-update and delete operations.  Considered on their own, RDF
+where the categorization is based on capabilities to do CRUD (Create,
+Read, Update, Delete) operations.  Considered on their own, RDF
 serializations are hypermedia types, but only at the \textsf{LO} and
-\textsf{CL} levels, and just an R~Type (read) on the Hypermedia
-Scale. We aim at improving this situation.
+\textsf{CL} levels (see footnotes for details), and just an R~Type
+(read) on the Hypermedia Scale. We aim at improving this situation.
 

### 2012-03-06 eswc-2012/phdsymp.tex
hypothesis testing vs evidence
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@155 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 in the present paper is on statistics in benchmarking with the purpose
-of putting it on a firmer foundation from a hypothesis testing
-perspective.
+of providing a firmer foundation on which assertions about engine
+performance can be backed with evidence.
 

### 2012-03-06 eswc-2012/phdsymp.tex
reformulate engine selection
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@156 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 select an engine with the best characteristics, relative to their own
-expected application. However, I am often asked the question: ``what
-is the best general-purpose triple store?'' Those who ask are usually not
-prepared to expand on the question. Thus, we should ask ourselves: 
+expected application. New users often are unable to formulate
+\emph{any} relevant expectations. For these users, the wide variety of
+benchmarks, for instance by the TPC, offers little help as a thorough
+understanding of the benchmarks and what they measure is required to
+select an engine. To help these users, we should ask ourselves:
 Can we collapse all the metrics into a single \emph{test
@@
 to see if any changes they do to the engine can have unexpected adverse
-effects on other parts of the system? For those who are able to
-qualify their inquiry, weighing different parts of the system or
+effects on other parts of the system? For those who can formulate
+relevant expectations, weighing different parts of the system or
 metrics is more useful and will yield a more tenable answer, and
@@
 create a ``one-size-fits-all'' benchmark but answering the above
-question requires such a benchmark.
+question requires such a benchmark. The relative simplicity of RDF and
+SPARQL may make this feasible.
 

### 2012-03-06 eswc-2012/phdsymp.tex
mention factorial designs early and restrict
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@157 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 seems not to be common practice. Furthermore, I have not to date seen
-any work towards using contemporary statistical methods to evaluate
-the performance of software, but I shall admit that my literature
-study has been limited to RDF/OWL databases, and practical benchmarks
-of databases, file systems, scientific software, etc. Therefore, the
-existence of relevant references in the deeper computer science
-literature is a key issue I would appreciate discussing at the
-Symposium.
+any work towards using methods like factorial designs to evaluate
+the performance of software, but there may be a limit in terms of
+complexity for where it is feasible, and I will restrict myself to
+SPARQL for this thesis.
 

### 2012-03-06 eswc-2012/phdsymp.tex
cite alioth shootout
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@158 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 above are almost orthogonal to the problems considered by my proposed
-project. While I have heard of some cases where a hypothesis test is
-used to demonstrate that one implementation is better than another, it
-seems not to be common practice. Furthermore, I have not to date seen
+project. While I have seen some cases that compares performance based
+on box plots\footnote{See \url{http://shootout.alioth.debian.org/} for example}
+but it seems not to be common practice. Furthermore, I have not to date seen
 any work towards using methods like factorial designs to evaluate

### 2012-03-06 eswc-2012/phdsymp.tex
advices in abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@159 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   Discontent with current state of the art in benchmarking prompts the
-  author to focus on the evaluation problem and to challenge the
-  notion that a ``one-size-fits-all'' benchmark is impossible.
+  author to focus on the evaluation problem and to investigate whether
+  we can provide better advices to new users selecting a SPARQL engine
+  to use.
 \end{abstract}

### 2012-03-06 eswc-2012/workshop2.tex
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@161 43ef1190-90fd-4dd7-a769-d0121f02b80e
remove space
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@162 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 serializations are hypermedia types, but only at the \textsf{LO} and
-\textsf{CL} levels (see footnotes for details), and just an R~Type
+\textsf{CL} levels (see footnotes for details on this notation), and just an R~Type
 (read) on the Hypermedia Scale. We aim at improving this situation.
@@
 \begin{verbatim}
-<> hm:canBe hm:mergedInto, hm:replaced , hm:deleted ;
+<> hm:canBe hm:mergedInto, hm:replaced, hm:deleted ;
    hm:createSimilarAt <../> ;

### 2012-03-06 eswc-2012/workshop2.tex
why hypermedia RDF
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@163 43ef1190-90fd-4dd7-a769-d0121f02b80e
contrast graphstore
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@164 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 it possible to use generic code to minimize the effort required to
-support RDF. Secondly, it requires relevant links and that the
+support RDF. Importantly, if interacting with Linked Data requires
+extensive out-of-band information, it will be harder to use than media
+types that does not. Secondly, it requires relevant links and that the
 resulting graph can be traversed by reasonable means.
@@
 while being fully RESTful. The Protocol specification is currently not RESTful
-per the discussion above, even though it was a key design goal. It
+per the discussion above as it requires extensive out-of-band information, even though it was a key design goal. It
 should also be possible to use this on the default (nameless) graph by

### 2012-03-07 eswc-2012/workshop2.tex
grammartypo trim some HTTP qualifications
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@166 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 extensive out-of-band information, it will be harder to use than media
-types that does not. Secondly, it requires relevant links and that the
+types that do not. Secondly, it requires relevant links and that the
 resulting graph can be traversed by reasonable means.
@@
 adequately take into account the constraint known as ``Hypermedia as
-the Engine of Application State'' (commonly abbreviated HATEOAS) from
+the Engine of Application State'' (abbreviated HATEOAS) from
 the REST architectural style, see
@@
 
-The second triple exists to make it possible (in the typical HTTP
-case) to POST an RDF payload to some URI and the server will itself
-assign a URI to the posted RDF payload.
+The second triple exists to make it possible to POST an RDF payload to
+some URI and the server will itself assign a URI to the posted RDF
+payload.
 
@@
 
-The first triple says that the current resource may be replaced by, in
-the HTTP case, PUTting an RDF payload to the resource URI. The second
+The first triple says that the current resource may be replaced by PUTting an RDF payload to the resource URI. The second
 triple says that the resource may be deleted, typically by a HTTP

### 2012-03-09 eswc-2012/phdsymp.tex
bullet list of contributions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@160 43ef1190-90fd-4dd7-a769-d0121f02b80e
Cleanup to shorten references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@168 43ef1190-90fd-4dd7-a769-d0121f02b80e
bullett point
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@169 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 long-term goal of my work is SPARQL Federation, but that is a minor
-concern in this paper.
+concern in this paper. Overall, the expected contributions are:
+
+\begin{itemize}
+\item Benchmarks that are able to cover all realistic
+  performance-influencing parameters for SPARQL engines and make it
+  possible to weigh different parameters, as well as quantify
+  unexplained differences.
+
+\item To assist optimization, enable endpoint service descriptions to
+  expose:
+\begin{itemize}
+\item performance characteristics,
+\item statistical digests of data that are optimized with respect to
+  size and query performance.
+\end{itemize}
+\end{itemize}
 
@@
 \bibliographystyle{plain-csmin}
+%\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}

### 2012-03-09 eswc-2012/phdsymp.tex
Martins enklere tilbakemelding
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@170 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-I take the technology state of the art to be represented by the
+I take the state of the art in technology to be represented by the
 current basic SPARQL 1.1 Federated Query Working
@@
 A recent scientific treatment of the current specification is in
-\cite{springerlink:10.1007/978-3-642-21064-8-1}. In this paper, the
+\cite{springerlink:10.1007/978-3-642-21064-8-1}. In that paper, the
 authors also show an optimization strategy based on execution order of
@@
 \cite{springerlink:10.1007/978-3-642-25073-6-38} proposes
-\emph{bound joins} and proves it can dramatically reduce the number of
-requests needed to federation members as well as the implementation of
+\emph{bound joins} and proves they can dramatically reduce the number of
+requests to federation members that are needed, as well as the implementation of
 FedX.
@@
 section~3.3.1
-in~\cite{springerlink:10.1007/978-3-642-17551-0-5}. Notably, histogram
+in~\cite{springerlink:10.1007/978-3-642-17551-0-5}, i.e. strike a
+balance between accuracy and index size, and updating statistics as
+data changes. Notably, histogram
 approaches generally suffer from the problem that they grow too large
@@
 above are almost orthogonal to the problems considered by my proposed
-project. While I have seen some cases that compares performance based
-on box plots\footnote{See \url{http://shootout.alioth.debian.org/} for example}
-but it seems not to be common practice. Furthermore, I have not to date seen
+project. While I have seen some cases that compare performance based
+on box plots\footnote{See \url{http://shootout.alioth.debian.org/} for
+  example}, it seems not to be common practice. Furthermore, I have not to date seen
 any work towards using methods like factorial designs to evaluate
@@
 problem, I have not chosen any methodology, but an interesting
-direction for work seem to be to find more space-efficient ways to
+direction for work seems to be to find more space-efficient ways to
 expose statistics in the service description and standardize them.
@@
 see if there is a methodology that is better suited to the
-problem. Using this, I hope to use, amongst other things, the
-complexity analysis of SPARQL to find suitable factors to see if my
+problem. I also intend to use existing results on complexity bounds
+for SPARQL query answering to find suitable factors to see if my
 admittedly bold proposition that it is possible to design a benchmark

### 2012-03-09 eswc-2012/phdsymp.tex
against SQL
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@171 43ef1190-90fd-4dd7-a769-d0121f02b80e
Comment out sections that should be changed
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@172 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-  Discontent with current state of the art in benchmarking prompts the
-  author to focus on the evaluation problem and to investigate whether
-  we can provide better advices to new users selecting a SPARQL engine
-  to use.
+%   Discontent with current state of the art in benchmarking prompts the
+%   author to focus on the evaluation problem and to investigate whether
+%   we can provide better advices to new users selecting a SPARQL engine
+%   to use.
 \end{abstract}
@@
 performance can be backed with evidence.
+
+The prior art in database theory is extensive, but I intend to focus
+on aspects that sets SPARQL apart from e.g. SQL, like the quad model,
+that it is a Web technology, or that data are commonly very
+heterogenous.
 
@@
 
-With the currently published benchmarks, it is up to the user to
-select an engine with the best characteristics, relative to their own
-expected application. New users often are unable to formulate
-\emph{any} relevant expectations. For these users, the wide variety of
-benchmarks, for instance by the TPC, offers little help as a thorough
-understanding of the benchmarks and what they measure is required to
-select an engine. To help these users, we should ask ourselves:
-Can we collapse all the metrics into a single \emph{test
-  statistic} that can summarize the performance of the engine as a
-whole? Also, can we provide a system for developers that they can use
-to see if any changes they do to the engine can have unexpected adverse
-effects on other parts of the system? For those who can formulate
-relevant expectations, weighing different parts of the system or
-metrics is more useful and will yield a more tenable answer, and
-motivates the question: Can we use the system to better find the
-distinguishing characteristics so that we can return a question that
-will actually help new users of SPARQL engines to chose one that meets
-their needs?
+% With the currently published benchmarks, it is up to the user to
+% select an engine with the best characteristics, relative to their own
+% expected application. New users often are unable to formulate
+% \emph{any} relevant expectations. For these users, the wide variety of
+% benchmarks, for instance by the TPC, offers little help as a thorough
+% understanding of the benchmarks and what they measure is required to
+% select an engine. To help these users, we should ask ourselves:
+% Can we collapse all the metrics into a single \emph{test
+%   statistic} that can summarize the performance of the engine as a
+% whole? Also, can we provide a system for developers that they can use
+% to see if any changes they do to the engine can have unexpected adverse
+% effects on other parts of the system? For those who can formulate
+% relevant expectations, weighing different parts of the system or
+% metrics is more useful and will yield a more tenable answer, and
+% motivates the question: Can we use the system to better find the
+% distinguishing characteristics so that we can return a question that
+% will actually help new users of SPARQL engines to chose one that meets
+% their needs?
 
-The above relies on data and queries that are representative of the
-system as a whole. It is often asserted that it is not possible to
-create a ``one-size-fits-all'' benchmark but answering the above
-question requires such a benchmark. The relative simplicity of RDF and
-SPARQL may make this feasible.
+% The above relies on data and queries that are representative of the
+% system as a whole. It is often asserted that it is not possible to
+% create a ``one-size-fits-all'' benchmark but answering the above
+% question requires such a benchmark. The relative simplicity of RDF and
+% SPARQL may make this feasible.
 
@@
 federated query engines assistance in optimizing for performance
-characteristics of certain SPARQL implementations. Adjusting how
-different features are weighed will help those who know details of
-their expected SPARQL use and normalization-techniques of the weights
-will answer the special case of ``the best general-purpose SPARQL
-engine''.
+characteristics of certain SPARQL implementations. 
+%Adjusting how
+%different features are weighed will help those who know details of
+%their expected SPARQL use and normalization-techniques of the weights
+%will answer the special case of ``the best general-purpose SPARQL
+%engine''.
 

### 2012-03-09 eswc-2012/phdsymp.tex
delete old text...
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@173 43ef1190-90fd-4dd7-a769-d0121f02b80e
merely anecdotal
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@174 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-%   Discontent with current state of the art in benchmarking prompts the
-%   author to focus on the evaluation problem and to investigate whether
-%   we can provide better advices to new users selecting a SPARQL engine
-%   to use.
+% TODO Here
 \end{abstract}
@@
 
-% With the currently published benchmarks, it is up to the user to
-% select an engine with the best characteristics, relative to their own
-% expected application. New users often are unable to formulate
-% \emph{any} relevant expectations. For these users, the wide variety of
-% benchmarks, for instance by the TPC, offers little help as a thorough
-% understanding of the benchmarks and what they measure is required to
-% select an engine. To help these users, we should ask ourselves:
-% Can we collapse all the metrics into a single \emph{test
-%   statistic} that can summarize the performance of the engine as a
-% whole? Also, can we provide a system for developers that they can use
-% to see if any changes they do to the engine can have unexpected adverse
-% effects on other parts of the system? For those who can formulate
-% relevant expectations, weighing different parts of the system or
-% metrics is more useful and will yield a more tenable answer, and
-% motivates the question: Can we use the system to better find the
-% distinguishing characteristics so that we can return a question that
-% will actually help new users of SPARQL engines to chose one that meets
-% their needs?
+Furthermore, the current approach presents merely anecdotal evidence
+that one engine is better than another with respect to
+performance. It may be that while the benchmarking queries do seem to
+favor one engine, other queries that have not been tried may reveal
+that there are strong adverse effects that may not have been
+anticipated. A more systematic approach is needed to provide a
+comprehensive and objective basis for comparing SPARQL engines.
 
-% The above relies on data and queries that are representative of the
-% system as a whole. It is often asserted that it is not possible to
-% create a ``one-size-fits-all'' benchmark but answering the above
-% question requires such a benchmark. The relative simplicity of RDF and
-% SPARQL may make this feasible.
+% TODO here
 
@@
 characteristics of certain SPARQL implementations.
-%Adjusting how
-%different features are weighed will help those who know details of
-%their expected SPARQL use and normalization-techniques of the weights
-%will answer the special case of ``the best general-purpose SPARQL
-%engine''.
+
+% TODO here
 

### 2012-03-12 eswc-2012/phdsymp.tex
value for new users and devs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@175 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Ultimately, this approach may be of great value to new users of SPARQL
+engines, as it can point out key differences between engines and
+therefore help users identify aspects that they should investigate
+with respect to their own expected uses. By normalizing weights
+between features, we may also attempt to advice those who do not know
+anything about their expected usage. Similarly, it may of great use to
+developers as it would quantify any adverse effects a modification may
+have.
+
 % TODO here

### 2012-03-12 eswc-2012/phdsymp.tex
systematic approach
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@176 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+A systematic approach based on statistical experimental design would
+imply that all possible axes of variation are identified, to identify
+possible contributions from different hardware platforms, operating
+system constraints, backend tuning, triple store features, to SPARQL
+language features and protocols. To make this benchmark feasible most
+variations of parameters must be done by a software suite designed for
+this purpose as the number of experiments that results will be very
+large.
+
 Factorial design inspires the present work and is covered in

### 2012-03-12 eswc-2012/phdsymp.tex
more contributions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@177 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \end{itemize}
+\item SPARQL engine developers can use the benchmark to reliably
+  quantify unexpected adverse effects from a given modification.
+\item New SPARQL engine users can identify key differences between
+  different engines, and therefore be able to more wisely choose
+  engine to use.
 \end{itemize}
@@
 
-Ultimately, this approach may be of great value to new users of SPARQL
-engines, as it can point out key differences between engines and
-therefore help users identify aspects that they should investigate
-with respect to their own expected uses. By normalizing weights
-between features, we may also attempt to advice those who do not know
-anything about their expected usage. Similarly, it may of great use to
-developers as it would quantify any adverse effects a modification may
-have.
-
 % TODO here

### 2012-03-12 eswc-2012/phdsymp.tex
just trimming
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@178 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-% TODO Here
+  For an objective evaluation as well as comparison of engines,
+  benchmarks that exhaustively covers the parameter space is
+  required. We propose an investigation into this problem using
+  statistical experimental planning.
 \end{abstract}
@@
 process is repeated a number of times and an average response time is
-found. The performance of different engines can then be compared,
-based on these averages.
+found. Different engines can be compared based on these averages.
 
@@
 treated as a \emph{stochastic variable}.
-
 This opens new methodological possibilities, first and foremost
@@
 comprehensive and objective basis for comparing SPARQL engines.
-
-% TODO here
 
@@
 data, so the adaption of existing techniques to such situations may
-also be an interesting problem.
+also be a interesting.
 
@@
 
-% TODO here
-
 \bibliographystyle{plain-csmin}

### 2012-03-12 eswc-2012/phdsymp.tex
typo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@179 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 data, so the adaption of existing techniques to such situations may
-also be a interesting.
+also be interesting.
 

### 2012-03-12 eswc-2012/phdsymp.tex
rewrite systematic paragraph
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@180 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 be efficient is that these variations need not occur in the same
-experiment. Therefore, many combinations of factors can be studied by
-carefully designing queries to cover different factors, and a
-formalism called ``resolution'' has been developed to classify how
-well this has been achieved, partly answering the question of
-evaluation methodology for this part of the thesis. We should also
-validate by comparing this benchmark with conclusions from existing
-benchmarks.
+experiment. Therefore, for the SPARQL language, many combinations of
+factors can be studied by carefully designing queries to cover
+different factors, and a formalism called ``resolution'' has been
+developed to classify how well this has been achieved, partly
+answering the question of evaluation methodology for this part of the
+thesis. We should also validate by comparing this benchmark with
+conclusions from existing benchmarks.
 
-A systematic approach based on statistical experimental design would
-imply that all possible axes of variation are identified, to identify
-possible contributions from different hardware platforms, operating
-system constraints, backend tuning, triple store features, to SPARQL
-language features and protocols. To make this benchmark feasible most
-variations of parameters must be done by a software suite designed for
-this purpose as the number of experiments that results will be very
-large.
+A systematic approach based on statistical experimental design means
+developing methods to add factors (not only language features will be
+factors), and then use this methodology to add and then vary all the
+factors in an optimal fashion. Developing a software suite to perform
+most of the experiments is a key requirement as the number of
+experiments that results will be very large.
 

### 2012-03-12 eswc-2012/phdsymp.tex
sort of a conclusion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@181 43ef1190-90fd-4dd7-a769-d0121f02b80e
titlerunning
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@182 43ef1190-90fd-4dd7-a769-d0121f02b80e
paste bibliography for submission
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@183 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   emphasis on benchmark quality}
-\titlerunning{Sharing statistics and benchmarking for SPARQL federation}
+\titlerunning{Sharing statistics and benchmarks for SPARQL federation}
 \author{Kjetil Kjernsmo\inst{1}}
@@
 
-\bibliographystyle{plain-csmin}
+This approach will yield the contributions listed above if completely
+successful, but will also advance the state of the art if only
+partially successful.
+
+%\bibliographystyle{plain-csmin}
 %\bibliographystyle{abbrv}
@@
 %\bibliographystyle{splncs03}
-\bibliography{selectivity,federation,benchmarks,optimization}
+%\bibliography{selectivity,federation,benchmarks,optimization}
+
+\begin{thebibliography}{1}
+
+\bibitem{springerlink:10.1007/978-3-642-21064-8-1}
+C. Buil-Aranda, M. Arenas, and O. Corcho.
+\newblock Semantics and optimization of the {SPARQL} 1.1 federation extension.
+\newblock In {\em The Semanic Web: Research and Applications}, LNCS\,6644:1-15,
+  2011.
+
+\bibitem{Duan:2011:AOC:1989323.1989340}
+S. Duan, A. Kementsietsidis, K. Srinivas, and O. Udrea.
+\newblock Apples and oranges: a comparison of {RDF} benchmarks and real {RDF}
+  datasets.
+\newblock In {\em Proc. of the 2011 Int. Conf. on Management of data}, SIGMOD
+  '11, pages 145--156, 2011. ACM.
+
+\bibitem{Getoor:2001:SEU:375663.375727}
+L. Getoor, B. Taskar, and D. Koller.
+\newblock Selectivity estimation using probabilistic models.
+\newblock In {\em Proc. of the 2001 ACM SIGMOD Int. Conf. on Management of
+  data}, SIGMOD '01, pages 461--472, 2001. ACM.
+
+\bibitem{springerlink:10.1007/978-3-642-17551-0-5}
+O. G\"orlitz and S. Staab.
+\newblock Federated data management and query optimization for linked open
+  data.
+\newblock In {\em New Directions in Web Data Management 1}, volume 331 of {\em
+  Studies in Computational Intelligence}, pages 109--137. Springer, 2011.
+
+\bibitem{Harth:2010:DSO:1772690.1772733}
+A. Harth, K. Hose, M. Karnstedt, A. Polleres, K.-U. Sattler, and J. Umbrich.
+\newblock Data summaries for on-demand queries over linked data.
+\newblock In {\em Proc. of the 19th Int. Conf. on World wide web}, WWW '10,
+  pages 411--420, 2010. ACM.
+
+\bibitem{mxro:Morsey2011DBpedia}
+M. Morsey, J. Lehmann, S. Auer, and A.-C.~N. Ngomo.
+\newblock {{DBpedia} {SPARQL} Benchmark - Performance Assessment with Real
+  Queries on Real Data}.
+\newblock In {\em 10th International Semantic Web Conference (ISWC2011)}, 2011.
+
+\bibitem{Schmidt:2011:FBS:2063016.2063054}
+M. Schmidt, O. G\"{o}rlitz, P. Haase, G. Ladwig, A. Schwarte, and T. Tran.
+\newblock {FedBench}: a benchmark suite for federated semantic data query
+  processing.
+\newblock In {\em Proc. of the 10th Int. Conf. on The semantic web - Volume
+  Part I}, LNCS\,7031:585--600, 2011.
+
+\bibitem{springerlink:10.1007/978-3-642-25073-6-38}
+A. Schwarte, P. Haase, K. Hose, R. Schenkel, and M. Schmidt.
+\newblock {FedX}: Optimization techniques for federated query processing on
+  linked data.
+\newblock In {\em The Semantic Web - ISWC 2011}, LNCS\,7031:601-616.
+
+\end{thebibliography}
 

### 2012-03-19 benchmark/dbpedia-in-R/curl-test.R
indentation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@189 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -2,19 +2,16 @@ library(RCurl)
 
 #h = getCurlHandle()
 
-timeQuery <- function(endpoint, query, ..., curlh = getCurlHandle())
-  {
-    getForm(c(endpoint), query=query, ..., curl = curlh)
-    info <- getCurlInfo(curlh)
-    list(
-         endpoint = info$total.time - info$namelookup.time,
-         connection = info$connect.time - info$namelookup.time,
-         firstbyte = info$starttransfer.time - info$connect.time,
-         alldata = info$total.time - info$starttransfer.time
-         )
-
-
-  }
+timeQuery <- function(endpoint, query, ..., curlh = getCurlHandle()) {
+  getForm(c(endpoint), query=query, ..., curl = curlh)
+  info <- getCurlInfo(curlh)
+  list(
+       endpoint = info$total.time - info$namelookup.time,
+       connection = info$connect.time - info$namelookup.time,
+       firstbyte = info$starttransfer.time - info$connect.time,
+       alldata = info$total.time - info$starttransfer.time
+       )
+}
 
 timeQuery("http://dbpedia.org/sparql", "SELECT * WHERE { ?concept a skos:Concept } LIMIT 100");
 

### 2012-03-19 benchmark/dbpedia-in-R/compileQueries.R
First stab at compiling queries
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@191 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,11 @@
+library(SPARQL)
+
+queries <- read.csv("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/benchmark.dbpedia.org/Queries.txt", header=FALSE, col.names=c('name', 'query', 'auxquery'), row.names=1, sep="\t")
+
+singleQuery <- function(endpoint, queries, runs = 5) {
+  auxdata <- SPARQL(url=endpoint, query=queries$auxquery)
+  using <- auxdata$results[sample((1:nrow(auxdata$results)), runs),]
+  using
+}
+
+singleQuery("http://dbpedia.org/sparql", queries[1,])

### 2012-03-22 eswc-2012/workshop2.tex
note read-write scenario in conclusion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@167 43ef1190-90fd-4dd7-a769-d0121f02b80e
Ooops
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@192 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 SPARQL is important when making non-trivial traversals of Linked Data
-graphs, and see how to a bridge between Linked Data and SPARQL may be
+graphs, and see how a bridge between Linked Data and SPARQL may be
 created with hypermedia RDF.
@@
 We have shown how RDF serializations can become very powerful
-hypermedia types by using a simple vocabulary. We have argued briefly
-that some triples should be added to every information resource, but
-note that most triples are only relevant to write-operations and
-should only appear in that case. We also noted that SPARQL is
-important for non-trivial traversal of Linked Data. To sum up, the
-addition of the following triples would make life easier for
-developers of applications that use Semantic Web data:
+hypermedia types by using a simple vocabulary. As a consequence,
+developers can use Linked Data in a read-write scenario without
+specialized knowledge about RDF APIs or other out-of-band
+information. We have argued briefly that some triples should be added
+to every information resource, but note that most triples are only
+relevant to write-operations and should only appear in that case. We
+also noted that SPARQL is important for non-trivial traversal of
+Linked Data. To sum up, the addition of the following triples would
+make life easier for developers of applications that use Semantic Web
+data:
 

### 2012-03-22 benchmark/dbpedia-in-R/compileQueries.R
running auxillary query
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@193 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,11 +1,13 @@
 library(SPARQL)
 
-queries <- read.csv("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/benchmark.dbpedia.org/Queries.txt", header=FALSE, col.names=c('name', 'query', 'auxquery'), row.names=1, sep="\t")
+queries <- read.delim("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/benchmark.dbpedia.org/Queries.txt", header=FALSE, col.names=c('name', 'query', 'auxquery'), row.names=1, colClasses = "character")
 
 singleQuery <- function(endpoint, queries, runs = 5) {
   auxdata <- SPARQL(url=endpoint, query=queries$auxquery)
-  using <- auxdata$results[sample((1:nrow(auxdata$results)), runs),]
+  using <- as.vector(auxdata$results[sample((1:nrow(auxdata$results)), runs),]) # TODO: Support multivars
   using
 }
+
+
 
 singleQuery("http://dbpedia.org/sparql", queries[1,])

### 2012-03-22 benchmark/dbpedia-in-R/compileQueries.R
replace and create queries to run
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@194 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -5,9 +5,11 @@ queries <- read.delim("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/bench
 singleQuery <- function(endpoint, queries, runs = 5) {
   auxdata <- SPARQL(url=endpoint, query=queries$auxquery)
   using <- as.vector(auxdata$results[sample((1:nrow(auxdata$results)), runs),]) # TODO: Support multivars
-  using
+  sapply(using, insertAuxillaries, queries$query)
 }
 
-
+insertAuxillaries <- function(var, query) {
+  sub("%%var%%", paste("<", var, ">", sep="", collapse=""), query, fixed=TRUE)
+}
 
 singleQuery("http://dbpedia.org/sparql", queries[1,])

### 2012-03-22 benchmark/dbpedia-in-R/compileQueries.R
time a single query
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@195 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -5,11 +5,13 @@ queries <- read.delim("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/bench
 singleQuery <- function(endpoint, queries, runs = 5) {
   auxdata <- SPARQL(url=endpoint, query=queries$auxquery)
   using <- as.vector(auxdata$results[sample((1:nrow(auxdata$results)), runs),]) # TODO: Support multivars
-  sapply(using, insertAuxillaries, queries$query)
+  runqueries <- sapply(using, runQueries, queries$query)
+  runqueries
 }
 
-insertAuxillaries <- function(var, query) {
-  sub("%%var%%", paste("<", var, ">", sep="", collapse=""), query, fixed=TRUE)
+runQueries <- function(var, queryWithVar) {
+  query <- sub("%%var%%", paste("<", var, ">", sep="", collapse=""), queryWithVar, fixed=TRUE)
+  timeQuery("http://dbpedia.org/sparql", query)
 }
 
 singleQuery("http://dbpedia.org/sparql", queries[1,])

### 2012-04-12 eswc-2012/workshop2.tex
move CR into a new subsection
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@209 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 reasonable base URI can be written as \texttt{<>}. We have not found
-the following factors to be relevant to RDF: \textsf{LE}, \textsf{CU}
-and \textsf{LT} (though the latter may be supported by
+the following factors to be relevant to RDF: \textsf{LE}, \textsf{CU},
+\textsf{CR} and \textsf{LT} (though the latter may be supported by
 RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?); and
-\textsf{LO} and \textsf{CL} are trivially supported. The remaining are:
+\textsf{LO} and \textsf{CL} are trivially supported. The remaining
+are:
 
@@
 
-\item[\textsf{CR}] Support for modifying control data for read requests
-  (e.g. HTTP Accept-* headers).
-
-\begin{verbatim}
-<> hm:acceptsFormats <http://www.w3.org/ns/formats/Turtle> .
-\end{verbatim}
-
 \item[\textsf{CM}] Support for indicating the interface method for requests
@@
 \end{question}
+
+\subsection{New H~Factors}
+
+\begin{itemize}
+\item Support for supplying control data relevant to subsequent requests.
+
+\begin{verbatim}
+<> hm:acceptsFormat <http://www.w3.org/ns/formats/Turtle> .
+\end{verbatim}
+
+\item Support for self-description.
+
+\end{itemize}
 

### 2012-04-12 eswc-2012/workshop2.tex
expand on level factors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@210 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Read, Update, Delete) operations.  Considered on their own, RDF
-serializations are hypermedia types, but only at the \textsf{LO} and
-\textsf{CL} levels (see footnotes for details on this notation), and just an R~Type
-(read) on the Hypermedia Scale. We aim at improving this situation.
+serializations are hypermedia types, but only at the \textsf{LO}
+(outbound links) and \textsf{CL} (control data for links) levels (see
+footnotes for details on this notation), and just an R~Type (read) on
+the Hypermedia Scale. We aim at improving this situation.
 
@@
 reasonable base URI can be written as \texttt{<>}. We have not found
-the following factors to be relevant to RDF: \textsf{LE}, \textsf{CU},
-\textsf{CR} and \textsf{LT} (though the latter may be supported by
-RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?); and
-\textsf{LO} and \textsf{CL} are trivially supported. The remaining
+the following factors to be relevant to RDF: \textsf{LE} (embedded
+links), \textsf{CR}, \textsf{CU} (control data for read and update
+requests respectively), and \textsf{LT} (templated queries, though the latter may be
+supported by RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?);
+and \textsf{LO} and \textsf{CL} are trivially supported. The remaining
 are:

### 2012-04-12 eswc-2012/workshop2.tex
note hm:mergedInto
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@211 43ef1190-90fd-4dd7-a769-d0121f02b80e
Make RaUL a LT remove RDForms
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@212 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the following factors to be relevant to RDF: \textsf{LE} (embedded
-links), \textsf{CR}, \textsf{CU} (control data for read and update
-requests respectively), and \textsf{LT} (templated queries, though the latter may be
-supported by RDForms\footnote{\url{http://vocab.deri.ie/rdforms}}?);
-and \textsf{LO} and \textsf{CL} are trivially supported. The remaining
-are:
+links), \textsf{CR} and \textsf{CU} (control data for read and update
+requests respectively); and \textsf{LO} and \textsf{CL} are trivially
+supported. The remaining are:
 
@@
 payload.
+
+\item[\textsf{LT}] Support for templated queries (HTTP GET)
+
+The RaUL\footnote{\url{http://vocab.deri.ie/raul}} goes beyond HTTP
+GET queries, but can be used to formulate templated queries similar to
+those in HTML, or indeed to generate HTML fragments for the same purpose.
 
@@
 in an out-of-band specification like is currently being done.
+
+Also note the \texttt{hm:mergedInto} term was motivated by that POST
+operations in this specification results in an RDF Merge of the
+payload into the named graph.
 

### 2012-04-12 eswc-2012/workshop2.tex
todo comments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@213 43ef1190-90fd-4dd7-a769-d0121f02b80e
reformat long lines
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@214 43ef1190-90fd-4dd7-a769-d0121f02b80e
move RaUL again, it is not really what we want
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@215 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{cite}
+
+% TODO:
+% Explain HATEOAS
+% More on new H-factors
+% References
+% POST in relation to graph store
+% Weakness of RDF
 
@@
 
-While links are abundant across the Linked Open Data (LOD) cloud, this paper will
-argue that key links are missing to make it possible to traverse the
-resulting graph and to enable read-write operations. Moreover, we will
-argue that this shortcoming is due to that the community does not
-adequately take into account the constraint known as ``Hypermedia as
-the Engine of Application State'' (abbreviated HATEOAS) from
-the REST architectural style, see
+While links are abundant across the Linked Open Data (LOD) cloud, this
+paper will argue that key links are missing to make it possible to
+traverse the resulting graph and to enable read-write
+operations. Moreover, we will argue that this shortcoming is due to
+that the community does not adequately take into account the
+constraint known as ``Hypermedia as the Engine of Application State''
+(abbreviated HATEOAS) from the REST architectural style, see
 \cite{Fielding_2000_Architectural-Styles}~Chapter~5.
@@
 
-%As noted in \cite{sunsetonmvc} 
-AtomPub has a single service endpoint,
-and you can navigate to what you need from there. This motivates the
-first discussion topic of this paper:
+% TODO: rough start
+AtomPub has a single service endpoint, and you can navigate to what
+you need from there. This motivates the first discussion topic of this
+paper:
 
@@
 the following factors to be relevant to RDF: \textsf{LE} (embedded
-links), \textsf{CR} and \textsf{CU} (control data for read and update
-requests respectively); and \textsf{LO} and \textsf{CL} are trivially
-supported. The remaining are:
+links), \textsf{CR}, \textsf{CU} (control data for read and update
+requests respectively) and \textsf{LT} (templated queries); and
+\textsf{LO} and \textsf{CL} are trivially supported. The remaining
+are:
 
@@
 
-\item[\textsf{LT}] Support for templated queries (HTTP GET)
-
-The RaUL\footnote{\url{http://vocab.deri.ie/raul}} goes beyond HTTP
-GET queries, but can be used to formulate templated queries similar to
-those in HTML, or indeed to generate HTML fragments for the same purpose.
-
 \item[\textsf{LI}] Support for idempotent updates (HTTP PUT, DELETE)
@@
 
-The first triple says that the current resource may be replaced by PUTting an RDF payload to the resource URI. The second
-triple says that the resource may be deleted, typically by a HTTP
-DELETE on the resource URI.
+The first triple says that the current resource may be replaced by
+PUTting an RDF payload to the resource URI. The second triple says
+that the resource may be deleted, typically by a HTTP DELETE on the
+resource URI.
 
@@
 Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},
-while being fully RESTful. The Protocol specification is currently not RESTful
-per the discussion above as it requires extensive out-of-band information, even though it was a key design goal. It
-should also be possible to use this on the default (nameless) graph by
-assigning it a name in the service description rather than defining it
-in an out-of-band specification like is currently being done.
+while being fully RESTful. The Protocol specification is currently not
+RESTful per the discussion above as it requires extensive out-of-band
+information, even though it was a key design goal. It should also be
+possible to use this on the default (nameless) graph by assigning it a
+name in the service description rather than defining it in an
+out-of-band specification like is currently being done.
 
@@
 \item Support for self-description.
+
+\item A factor for RaUL templates.
+
+The RaUL\footnote{\url{http://vocab.deri.ie/raul}} goes beyond HTTP
+GET queries, but can be used to formulate templated queries similar to
+those in HTML, or indeed to generate HTML fragments for the same purpose.
 
@@
    hm:createSimilarAt <../> ;
-   hm:acceptsFormats <http://www.w3.org/ns/formats/Turtle> ;
+   hm:acceptsFormat <http://www.w3.org/ns/formats/Turtle> ;
    void:inDataset [ void:sparqlEndpoint </sparql> . ] .

### 2012-04-12 eswc-2012/workshop2.tex
more on HATEOAS
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@216 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 % Weakness of RDF
+% Reformulate something on SPARQL about enabling
 
@@
 hypermedia links in the present resource \emph{only}, they should not
-require any out-of-band information.
+require any out-of-band information. It is very common that
+specifications that advertise themselves as RESTful has a read-write
+API that is specified in terms of URI structure as well as HTTP
+verbs. This is not a violation of the HATEOAS constraint \textit{per
+  se}, but it would usually be superfluous as the same information
+\emph{must} be available in a hypermedia message in order to satisfy
+the HATEOAS constraint. The constraint does not require this
+information to be available in \emph{all} messages, nor does it
+require that all resources can be reached from any other resource, and
+so, it is not a very strict constraint.
 

### 2012-04-12 eswc-2012/workshop2.tex
improve atompub start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@217 43ef1190-90fd-4dd7-a769-d0121f02b80e
new h factors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@218 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 % TODO:
-% Explain HATEOAS
-% More on new H-factors
 % References
@@
 % Reformulate something on SPARQL about enabling
+% More writing around factors?
 
@@
 
-% TODO: rough start
-AtomPub has a single service endpoint, and you can navigate to what
-you need from there. This motivates the first discussion topic of this
-paper:
+AtomPub %TODO: ref
+is a widely cited as a good example of a RESTful protocol and
+we note that it has a single service endpoint. From there, you can
+navigate to what you need. This motivates the first discussion topic
+of this paper:
 
@@
 
+Mike Amundsen solicits feedback on the completeness of his
+classification scheme. I suggest the following as discussion items:
+
 \begin{itemize}
+
+\item Support for self-description.
+
+The self-describing nature of RDF is an important characteristic of
+the model and arguably an important characteristic of RDF as
+hypermedia as agents may be able to infer possible interactions based
+on vocabulary definitions.
+
 \item Support for supplying control data relevant to subsequent requests.
+
+It may be useful for agents to know e.g. what formats they can send to
+a given resource before the request is made. This may require a new
+control data factor. An example of this use may be e.g.:
 
@@
 
-\item Support for self-description.
-
-\item A factor for RaUL templates.
+\item A factor encompassing RaUL templates.
 
   The RaUL\footnote{\url{http://vocab.deri.ie/raul}} goes beyond HTTP
-GET queries, but can be used to formulate templated queries similar to
-those in HTML, or indeed to generate HTML fragments for the same purpose.
+  GET queries, which is what the \textsf{LT} factor is about, but can
+  be used to formulate templated queries similar to those in HTML, or
+  indeed to generate HTML fragments for the same purpose. This is also
+  enables useful interactions, albeit quite different from those of
+  this paper.
 

### 2012-04-12 eswc-2012/workshop2.tex
Add references for everything
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@219 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Mike Amundsen defines hypermedia types\footnote{\url{http://amundsen.com/hypermedia/}} as 
+Mike Amundsen defines hypermedia types\cite{hypermediatypes} as 
 \begin{quote}
@@
 media-type.'' The REST \& WOA Wiki defines ``the Hypermedia
-Scale''\footnote{\url{http://restpatterns.org/Articles/The\_Hypermedia\_Scale}},
+Scale''\cite{hypermediascale},
 where the categorization is based on capabilities to do CRUD (Create,
@@
 (outbound links) and \textsf{CL} (control data for links) levels (see
-footnotes for details on this notation), and just an R~Type (read) on
+\cite{hypermediatypes} for details on this notation), and just an R~Type (read) on
 the Hypermedia Scale. We aim at improving this situation.
@@
 Amundsen argues in a blog
-post\footnote{\url{http://amundsen.com/blog/archives/1083}} that the
+post\cite{nordfapi} that the
 Semantic Web community should not pursue an API for RDF, but rather
@@
 enables agents to do the same operations as the SPARQL 1.1 Graph Store
-HTTP
-Protocol\footnote{\url{http://www.w3.org/TR/2011/WD-sparql11-http-rdf-update-20110512/}},
-while being fully RESTful. The Protocol specification is currently not
-RESTful per the discussion above as it requires extensive out-of-band
-information, even though it was a key design goal. It should also be
-possible to use this on the default (nameless) graph by assigning it a
-name in the service description rather than defining it in an
-out-of-band specification like is currently being done.
+HTTP Protocol\cite{graphstore}, while being fully RESTful. The
+Protocol specification is currently not RESTful per the discussion
+above as it requires extensive out-of-band information, even though it
+was a key design goal. It should also be possible to use this on the
+default (nameless) graph by assigning it a name in the service
+description rather than defining it in an out-of-band specification
+like is currently being done.
 
@@
 
-  The RaUL\footnote{\url{http://vocab.deri.ie/raul}} goes beyond HTTP
-  GET queries, which is what the \textsf{LT} factor is about, but can
-  be used to formulate templated queries similar to those in HTML, or
-  indeed to generate HTML fragments for the same purpose. This is also
-  enables useful interactions, albeit quite different from those of
-  this paper.
+  The RaUL\cite{raul} goes beyond HTTP GET queries, which is what the
+  \textsf{LT} factor is about, but can be used to formulate templated
+  queries similar to those in HTML, or indeed to generate HTML
+  fragments for the same purpose. This is also enables useful
+  interactions, albeit quite different from those of this paper.
 
@@
 In the Linked Data Design
-Issue\footnote{\url{http://www.w3.org/DesignIssues/LinkedData}} Tim
+Issue\cite{linkeddataissue} Tim
 Berners-Lee notes ``To make the data be effectively linked, someone
@@
 The triples needed to do this are already defined in
-VoID\footnote{\url{http://vocab.deri.ie/void}} and are in use:
+VoID (see e.g. \cite{Alexander09describinglinked}) and are in use:
 \begin{verbatim}
@@
 %\bibliographystyle{splncs03}
-\bibliography{federation,webarch,webframeworks}
+\bibliography{federation,webarch,vocabs,specs,hypermedia}
 

### 2012-04-12 eswc-2012/workshop2.tex
Add ref for atompub
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@220 43ef1190-90fd-4dd7-a769-d0121f02b80e
POST in relation to graph store
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@221 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 % TODO:
-% References
 % POST in relation to graph store
@@
 
-AtomPub %TODO: ref
-is a widely cited as a good example of a RESTful protocol and
-we note that it has a single service endpoint. From there, you can
-navigate to what you need. This motivates the first discussion topic
-of this paper:
+The Atom Publishing Protocol\cite{atompub} is a widely cited as a good
+example of a RESTful protocol and we note that it has a single service
+endpoint. From there, you can navigate to what you need. This
+motivates the first discussion topic of this paper:
 
@@
 
+This term is motivated by that POST operations in \cite{graphstore} is
+specified to results in an RDF Merge of the payload into the named
+graph. Amongst other possibilities is a union of graphs.
+
 The second triple exists to make it possible to POST an RDF payload to
@@
 like is currently being done.
-
-Also note the \texttt{hm:mergedInto} term was motivated by that POST
-operations in this specification results in an RDF Merge of the
-payload into the named graph.
 

### 2012-04-12 eswc-2012/workshop2.tex
Mention linked data basic profile
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@222 43ef1190-90fd-4dd7-a769-d0121f02b80e
trim a bit
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@223 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 % TODO:
-% POST in relation to graph store
-% Weakness of RDF
 % Reformulate something on SPARQL about enabling
@@
 
-The Atom Publishing Protocol\cite{atompub} is a widely cited as a good
-example of a RESTful protocol and we note that it has a single service
-endpoint. From there, you can navigate to what you need. This
+We note that the Atom Publishing Protocol\cite{atompub} has a single
+service endpoint. From there, you can navigate to what you need. This
 motivates the first discussion topic of this paper:
@@
 some URI and the server will itself assign a URI to the posted RDF
-payload.
+payload. This may for example be a named graph, as defined in
+\cite{graphstore} or a follow the protocol recently suggested in
+\cite{linkeddatabasicprofilesubmission}~Section~5.4.
 

### 2012-04-12 eswc-2012/workshop2.tex
trim more tomorrow :)
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@224 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{cite}
-
-% TODO:
-% Reformulate something on SPARQL about enabling
-% More writing around factors?
 
@@
 
-In the Linked Data Design
-Issue\cite{linkeddataissue} Tim
-Berners-Lee notes ``To make the data be effectively linked, someone
-who only has the URI of something must be able to find their way the
-SPARQL endpoint.'', but this is generally not possible today without
+In the \cite{linkeddataissue} Tim Berners-Lee
+notes ``To make the data be effectively linked, someone who only has
+the URI of something must be able to find their way the SPARQL
+endpoint.'', but this is generally not possible today without
 requiring out-of-band information.

### 2012-04-12 eswc-2012/workshop2.tex
-the
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@225 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In the \cite{linkeddataissue} Tim Berners-Lee
-notes ``To make the data be effectively linked, someone who only has
-the URI of something must be able to find their way the SPARQL
-endpoint.'', but this is generally not possible today without
-requiring out-of-band information.
+In \cite{linkeddataissue} Tim Berners-Lee notes ``To make the data be
+effectively linked, someone who only has the URI of something must be
+able to find their way the SPARQL endpoint.'', but this is generally
+not possible today without requiring out-of-band information.
 

### 2012-04-13 eswc-2012/workshop2.tex
grammar
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@226 43ef1190-90fd-4dd7-a769-d0121f02b80e
third person
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@227 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-This term is motivated by that POST operations in \cite{graphstore} is
-specified to results in an RDF Merge of the payload into the named
+This term is motivated by that POST operations in \cite{graphstore} are
+specified to result in an RDF Merge of the payload into the named
 graph. Amongst other possibilities is a union of graphs.
@@
 Mike Amundsen solicits feedback on the completeness of his
-classification scheme. I suggest the following as discussion items:
+classification scheme. The following are suggested as discussion
+items:
 

### 2012-04-13 eswc-2012/workshop2.tex
redo RaUL role again
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@228 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{itemize}
+\begin{enumerate}
 
@@
 
-\item A factor encompassing RaUL templates.
+\item A factor encompassing RaUL \cite{raul} templates.
 
-  The RaUL\cite{raul} goes beyond HTTP GET queries, which is what the
-  \textsf{LT} factor is about, but can be used to formulate templated
-  queries similar to those in HTML, or indeed to generate HTML
-  fragments for the same purpose. This is also enables useful
-  interactions, albeit quite different from those of this paper.
+  RaUL goes beyond HTTP GET queries, which is what the \textsf{LT}
+  factor is about. The interaction RaUL enables are quite different
+  from those discussed in this paper and are not captured by any of
+  the current factors.
 
-\end{itemize}
+\end{enumerate}
 

### 2012-04-13 eswc-2012/workshop2.tex
redo HATEOAS discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@229 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 hypermedia links in the present resource \emph{only}, they should not
-require any out-of-band information. It is very common that
-specifications that advertise themselves as RESTful has a read-write
+require any out-of-band information. Since RDF is built on URIs, many
+of which can be dereferenced to obtain further links, it will
+therefore almost always satisfy the HATEOAS constraint in the
+read-only case.
+
+Contrast the read-write situation: It is very common that
+specifications that advertise themselves as RESTful has an 
 API that is specified in terms of URI structure as well as HTTP
@@
 the HATEOAS constraint. The constraint does not require this
-information to be available in \emph{all} messages, nor does it
-require that all resources can be reached from any other resource, and
+information to be available in \emph{all} messages, and
 so, it is not a very strict constraint.
 
-Since RDF is built on URIs, many of which can be dereferenced to
-obtain further links, it lends itself well to RESTful protocols, but
-it is debatable whether a protocol can be said to be RESTful if the
-links don't enable any useful interactions. Importantly, if there is
+It is debatable whether a protocol can be said to be RESTful if the
+links don't enable any useful interactions. Even in the read-only
+case, we must carefully add links that enable the desired interactions
+to take place, e.g. linking a SPARQL endpoint. Importantly, if there is
 to be such a thing as a RESTful read-write Semantic Web protocol,

### 2012-04-17 benchmark/dbpedia-in-R/compileQueries.R
Simple R script for correlation between length of literal and frequency of presence
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@230 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -4,13 +4,14 @@ allqueries <- read.delim("/home/kjekje/DBPediaBenchmark/data/dbpedia.aksw.org/be
 
 singleQuery <- function(queries, endpoint, runs = 5) {
   auxdata <- SPARQL(url=endpoint, query=queries$auxquery)
+  if (length(auxdata$results) < 1) stop("Auxillary query had no results")
+#  browser()
   using <- as.matrix(auxdata$results[sample((1:nrow(auxdata$results)), runs),])
   runqueries <- t(apply(using, 1, runQueries, queries$query, endpoint))
   as.data.frame(runqueries)
 }
 
 runQueries <- function(var, queryWithVar, endpoint) {
-  if (length(var) < 1) stop("Auxillary query had no results")
   if (length(var) == 1) {
     substQuery <- sub("%%var%%", paste("<", var, ">", sep="", collapse=""), queryWithVar, fixed=TRUE)
   } else {
@@ -27,7 +28,7 @@ runQueries <- function(var, queryWithVar, endpoint) {
 allQueries <- function(allqueries, endpoint, runs = 5) {
   for(queryname in row.names(allqueries)) {
    singleq <- singleQuery(allqueries[queryname,], endpoint, runs)
-   browser()
+
    singleq
  }
 }

### 2012-04-17 misc/count-length-corr.R

@@ -1,0 +1,4 @@
+literals <- read.csv('/ifi/bifrost/a03/kjekje/Desktop/sparql')
+framelit <- as.data.frame(table(literals))
+framelit$length <- sapply(levels(framelit$literals), nchar)
+plot(framelit$Freq, framelit$length, log="xy")

### 2012-04-21 misc/extract-literals-btc.pl
Script to extract str literals from Nquads
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@234 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,14 @@
+#!/usr/bin/perl
+
+use strict;
+use warnings;
+
+use PerlIO::gzip;
+
+open (COMQ, "<:gzip", $ARGV[0]);
+while (<COMQ>) {
+	my ($lit) = m/\"(.+)\"/;
+	print "$lit\n" if ($lit);
+}
+
+binmode COMQ, ":gzip(none)";

### 2012-07-05 mnses/essay.tex
Start of essay
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@328 43ef1190-90fd-4dd7-a769-d0121f02b80e

I mars 2006 fikk jeg en mail der jeg ble bedt om å komme til London
for et jobbintervju påfølgende helg. I denne bransjen foregår selv
jobbintervjuer under taushetserklæring, men essensen var denne:
``Gründerne har nettopp solgt sitt forrige prosjekt for 2 milliarder
euro, og er klare for å revolusjonere TV-mediet, men skal gjøre det på
medieindustriens premisser, etter tidligere å ha vært i konflikt med
disse.'' Etter en prat ble jeg forespeilet at jeg kunne jobbe fra
hvor som helst, når som helst, og med den teknologien jeg selv syntes
mest interessant, ``bare si din pris''. Jeg avslo tilbudet.

### 2012-07-05 mnses/essay.tex
More intro
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@329 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 ``Gründerne har nettopp solgt sitt forrige prosjekt for 2 milliarder
-euro, og er klare for å revolusjonere TV-mediet, men skal gjøre det på
+euro, og er klare for å revolusjonere TV-mediet. De skal gjøre det på
 medieindustriens premisser, etter tidligere å ha vært i konflikt med
-disse.'' Etter en prat ble jeg forespeilet at jeg kunne jobbe fra
-hvor som helst, når som helst, og med den teknologien jeg selv syntes
-mest interessant, ``bare si din pris''. Jeg avslo tilbudet.
+disse.'' Etter en prat ble jeg forespeilet at jeg kunne jobbe fra hvor
+som helst, når som helst, og med den teknologien jeg selv syntes mest
+interessant (og som jeg nå skriver mitt doktorgradsarbeid om), ``bare
+si din pris''. Jeg avslo tilbudet.
+
+I dette essayet skal jeg forklare hvorfor de premissene
+medieindustrien legger er i vil ifrata befolkningen viktige friheter,
+stanse teknologisk utvikling, gi et inntrykk i befolkningen at
+vitenskapen er autoritær og ekskluderende, og dermed gjøre at all
+vitenskapelig virksomhet faller i vanry. Videre har dette vide
+implikasjoner for ytringsfrihet og demokrati. Men det verste er at
+politikerne har allerede tatt lange skritt i denne retning, og dette
+må reverseres.

### 2012-07-07 mnses/essay.tex
litt disp
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@330 43ef1190-90fd-4dd7-a769-d0121f02b80e
trimming
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@331 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 stanse teknologisk utvikling, gi et inntrykk i befolkningen at
-vitenskapen er autoritær og ekskluderende, og dermed gjøre at all
+vitenskapen er autoritær og ekskluderende, og dermed gjøre at
 vitenskapelig virksomhet faller i vanry. Videre har dette vide
-implikasjoner for ytringsfrihet og demokrati. Men det verste er at
-politikerne har allerede tatt lange skritt i denne retning, og dette
-må reverseres.
+implikasjoner for ytringsfrihet og demokrati.
+
+\paragraph{Relevante menneskerettigheter og min forståelse}
+
+\paragraph{Etikk}
+
+Min forståelse er at dette ikke er etiske spørsmål ettersom det ikke
+er en reell interessekonflikt mellom skapere og mottakere av
+verk. Disse må leve i et økosystem der de er gjensidig avhengige av
+hverandre.

### 2012-07-07 mnses/essay.tex
wikipedia og sånt
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@332 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Dog ligger det i sakens natur at jeg som teknolog og naturviter kun
+har overfladisk forståelse av de legale rammeverk som diskuteres, og
+derfor vil jeg også beskrive tvil og problemer med bildet jeg tegner.
+
 \paragraph{Relevante menneskerettigheter og min forståelse}
+
+Den universelle menneskerettighetskonvensjonen\cite{udhr} er et
+overordnet rammeverk for menneskerettigheter som er videre behandlet i
+en rekke internasjonale eller overnasjonale avtaler. Dette er en
+omfattende literatur jeg har lite grunnlag for å kjenne i detalj, og
+jeg tar derfor utgangspunkt i konvensjonen.
+
+I artikkel 29, 2. punkt kan man lese  
+\begin{quote}
+In the exercise of his rights and freedoms, everyone shall be subject
+only to such limitations as are determined by law solely for the
+purpose of securing due recognition and respect for the rights and
+freedoms of others and of meeting the just requirements of morality,
+public order and the general welfare in a democratic society.
+\end{quote}
+
+Det er min tolkning at begrensninger i individets frihet i form av lov
+kun kan gjøres gjeldende der det er et klart og godt begrunnet behov
+for dette. Videre må personene det gjelder faktisk ha mulighet til å
+vite hva loven sier og hvordan man skal oppføre seg for å ikke bryte
+den.
+
+Artikkel 27 har to deler:
+\begin{quote}
+\begin{enumerate}
+\item Everyone has the right freely to participate in the cultural
+  life of the community, to enjoy the arts and to share in scientific
+  advancement and its benefits.
+
+\item Everyone has the right to the protection of the moral and
+  material interests resulting from any scientific, literary or
+  artistic production of which he is the author.
+\end{enumerate}
+\end{quote}
+
+Opphavsrettslovgningen, i sine forskjellige former, skal gi den
+beskyttelsen del 2 krever, mens del 1 har såvidt jeg vet ingen
+eksplisitt beskyttelse. Dog begrunnes ofte sikring av forfatternes
+materielle rettigheter med at det er det som gjør at befolkningen har
+tilgang til verk.
+
+\paragraph{Bevisbasert avgrensning}
+
+Ettersom begrensninger i individets frihet må være klart begrunnet, og
+et forbud mot eksemplarframstilling er en slik begrensning, så må vi
+spørre om en del av utviklingen i senere år viser områder der slike
+begrensninger er unødvendig.
+
+Det er her naturlig å peke på Wikipedias suksess. Anekdotisk bevis
+viser at Wikipedia kan vurderes på linje med etablerte leksika, se
+omtale~i~\cite{naturewiki}. Hvis man godtar at Wikipedia er bra nok,
+så må man så spørre seg om dette kunne vært oppnådd uten de
+begrensninger åndsverksloven gir. Da må vi notere oss at Wikipedia
+også har lisensbetingelser, men disse regulerer kun forhold som er
+relatert til dagens opphavsrettsregime, de ville ikke være nødvendige
+i et regime der leksika ikke hadde vern, ei heller regulerer de
+eksemplarframstilling. Dermed bør Wikipedia tjene som et eksempel på
+at begrensninger i befolkningens frihet når det gjelder leksika
+\emph{ikke} er nødvendig.
+
+Vi ser det samme i mange andre felter, både innenfor informasjon, data
+og programvare. Man kan idag fint leve uten programvare som i et
+regime der programvare ikke var vernet likevel ville eksistert i sin
+nåværende form. Dette er programvare som gjerne benevnes som ``fri
+programvare'' . Jeg bruker selv nesten utelukkende fri programvare, og
+utgir selv det jeg skriver som fri programvare, og man kan finne alt
+fra meget enkle programmer til meget avanserte systemer for
+resonnering. Blant web-servere så er det to slike systemer som
+konkurrerer i popularitet med Microsofts begrensende
+tilbud\cite{netcraft}. Det er min oppfatning at dette betyr at
+programvare i hovedsak ikke bør kunne underlegges begrensninger.
+
+Dog ser jeg at det kan finnes unntak fra denne regelen. Jeg ser det
+som utvilsomt at det er enklere å tjene penger på programvare som er
+underlagt begrensninger basert på egen erfaring. Menneskerettighetene
+sier ikke at det skal være enkelt å tjene penger, men det kan tenkes
+at det blir så vanskelig å få til en investering i f.eks. veldig
+sektor-spesifik programvare at ingen vil ta seg råd til å utvikle
+den. Dette vil dermed gjøre at man ikke får til teknologisk utvikling.
+
+Vi må videre spørre, gitt suksessen til Wikipedia og til fri
+programvare, hvilke andre områder er det det er illegitimt å begrense
+befolkningens frihet? Hvilke andre områder vil man kunne ha samme eller
+forbedret kunstnerisk og vitenskapelig framgang uten slike begrensninger?
+
+Jeg foreslår ikke en radikal avskaffelse av opphavsretten,
+hovedpoenget er at enhver begrensning må begrunnes, og at man må være
+svært forsiktig med å legge gjennomgripende begrensninger. Ettersom
+enhver begrensning er en inngripen i befolkningens frihet, er det
+naturlig at det stilles beviskrav til den som foreslår en begrensning.
+
+\paragraph{Avgrensning av inngripende tekniske tiltak}
 

### 2012-07-07 mnses/essay.tex
anti-DRM
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@333 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\paragraph{Avgrensning av inngripende tekniske tiltak}
+\paragraph{Inngripende tekniske tiltak}
+
+Hittil har jeg diskutert kun klassisk opphavsrett, og medieindustriens
+rolle er her begrenset. For å forsto hvorfor det ble så viktig for meg
+å avslå et åpenbart godt tilbud, så må vi se til straffelovens
+§262. Denne paragrafen i seg selv er totalt uforståelig for meg sett i
+isolasjon, men forarbeidene til Åndsverksloven som kom noen år senere
+gir et bedre innblikk i hva jeg antar det er snakk om, se spesielt
+Ot.prp. nr. 46, seksjon 3.5.1 \footnote{se
+  \url{http://www.regjeringen.no/nb/dep/kud/dok/regpubl/otprp/20042005/otprp-nr-46-2004-2005-/3/5.html?id=396451}}. 
+Disse forarbeidene ser ut til å være noe mer nyansert enn
+Straffeloven, men man ender opp med å gi lovbeskyttelse til
+``effektive tekniske beskyttelsessystemer''. For å forstå hva dette er
+henvises det til et EU-direktiv fra 1997, men i store trekk er det
+systemer som kan brukes til å begrense tilgangen til et verk også
+etter at man har tilegnet seg det, lovlig eller ulovlig. Jeg har ikke
+lest EU-direktivet, fordi allerede i forarbeidene bærer det så galt av
+sted at det for meg er åpenbart at lovreguleringen er av en slik natur
+at det er umulig for befolkningen å vite om man bryter loven eller
+ikke, uansett om hvor godt man kjenner lovgivningen. Denne tvilen om
+hva som er lov og ikke er absolutt destruktiv.
+
+For å forstå mer av hva disse ``beskyttelsessystemene'' faktisk er, la
+oss se på hva medieindustrien har kommet med.  I juli 2009 fjernet
+Amazon e-bøker fra kundenes lesebrett selv om kundene hadde legitimt
+kjøpt disse bøkene, se f.eks.~\cite{nytamazon}. Ironien er at blant
+bøkene de fjernet var George Orwells ``Big Brother''. I denne boka har
+myndighetene lagd et ``memory hole'' som brukes til å skrive om
+historien. Essensen til tekniske beskyttelsessystemer er at man
+tillater programvare på sin egen datamaskin, lesebrett eller
+mobiltelefon, som man selv ikke har noen autoritet over. Denne
+autoriteten beholdes av noen andre, og man har selv ingen kontroll på
+hva de kan eller vil bruke denne autoriteten til. I tilfellet over er
+det brukt til nettopp å lage Orwells ``memory hole'', og per idag har
+altså lovgiver ment at slik bør det være. Vi har idag ingen sikkerhet
+mot misbruk overhodet.
+
+Det mest skremmende som har kommet er en serie med dokumenter skrevet
+av Motion Picture Association of America og adressert til det
+amerikanske Senatet, med tittel ``Content Protection Status
+Report''\footnote{Documentene var inntil nylig tilgjengelig fra
+  Senatets websider, men nå finner jeg kun en omtale på Wikipedia:
+  \url{http://en.wikipedia.org/wiki/Content_Protection_Status_Report}.
+I disse dokumentene ble det foreslått tiltak som at all elektronikk
+som ble lagd med analog-til-digitale konvertere måtte utstyres med noe
+som gjorde at de kunne slås av hvis det ble detektert at de ble brukt
+til å ta opp vernet verk. Bakgrunnen for forslaget er at man ønsker å
+gjøre det umulig for uautoriserte å lage en mikrofon som kan brukes
+til å gjøre digitale opptak. Grunnen til at dette er så skremmende er at
+det ville føre til at man lagde en infrastruktur der Hollywood ville
+fått absolutt autoritet over all produksjon. Dette er heldigvis ikke
+gjennomførbart, men det at det ble presentert og tatt seriøst viser
+hvor langt man er villig til å gå.
 

### 2012-07-07 mnses/essay.tex
NDS stuff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@334 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Jeg mener at befolkningen ikke skal trenge å godta slike inngripen i
+sin frihet. At man skal overgi autoritet over systemer som er så
+personlige som en mobiltelefon til en tredjepart, som i essens kan
+gjøre hva en vil med den, er for meg et sterkt inngripen i en persons
+integritet og frihet. I essens handler dette om myndighetene skal ha
+tillit til den enkelte borger. For å illustrere dette klarere, kan man
+ta ``TV-bokser'' som idag formidles av blant annet GET. De siste
+boksene fungerer både som modem og TV-dekoder. Dette vil føre til at
+boksen som helhet er beskyttet av Straffelovens §262, hvilket igjen
+betyr at man som borger ikke har noen praktisk mulighet til å
+undersøke hva boksen i sin egen stue faktisk gjør. Det betyr at
+produsenten av disse boksene har mulighet til å overvåke til minste
+detalj hva brukerne gjør, både på nett og hva de ser på på TV. Det vil
+kunne gjøres på en slik måte at det i praksis er umulig å oppdage uten
+å muligens bryte Straffelovens §262, og jeg vil anta at overvåkningen
+kan gjøres slik at veldig få vil trenge å vite om den, inkludert ingen
+i GET eller andre selgere. Hvis dette scenariet ikke er skremmende
+nok, så kan det nevnes at ifølge Wikipedia\cite{ndswiki}, så er firmaet som lager
+disse boksene et selskap i Rupert Murdochs imperium og ledet av en
+tidligere etterretningsoffiser.
+
+Det blir derfor legitimt å spørre hvor norske politikere skal ha
+større tillit til et medieimperium som beviselig har ulovlig overvåket
+flere titalls mennesker\cite{newscorpphone}. Det er per idag ingen
+indikasjoner på at dette faktisk skjer, men det viktige
+tillitsspørsmålet bør være tverrpolitisk interessant.
+
 \paragraph{Etikk}

### 2012-07-08 mnses/essay.tex
forbrukermakt
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@336 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Det er videre viktig å merke seg at i alle tilfeller der befolkningen
+som forbrukere har et reelt valg så velger de å ikke kjøpe produkter
+med slike begrensninger. Den siste CDen med ``kopisperre'' kom ut i
+2006\cite{aftkopi}, mens iTunes åpnet for å selge musikk uten sperrer året etter,
+mens alle sperrer ble fjernet i 2009\cite{ituneskopi}, da det ikke gikk lenger å
+konkurrere med slike sperrer. På DVD og TV har
+ikke befolkningen noen reell valgmulighetet, og de består dermed.
+
+Man må derfor stille spørsmålet: Når befolkningen som forbrukere med
+stort flertall har forkastet et tiltak, skal det da være mulig i et
+demokrati at politikerne opprettholder tiltaket når tiltaket begrenser
+befolkningens frihet? Etter mitt skjønn burde det være politisk
+umulig; fra et sosialdemokratisk perspektiv, der politikken anses å
+være en mulighet for å korrigere markedskreftenes negative innflytelse
+bør det være umulig å vedta og opprettholde lover som har større
+negativ innflytelse på befolkningen enn markedet, og fra et
+konservativt perspektiv, som vektlegger individets frihet, bør man
+stille seg negativt til begrensninger som har vist seg unødvendige.
+
+La oss så returnere til at jeg sa nei til jobbtilbudet: Den
+umiddelbare grunnen bør dermed være klar: Det var klart at uansett
+hvor revolusjonerende systemet var og hvor teknologisk interessant det
+var, så ville det aldri bli noe av, ettersom folk rett og slett ikke
+vil ha systemer som spiller etter underholdningsindustriens regler
+hvis de har reelle valg. Jeg var ikke interessert i å drive med ting
+som aldri vil bli noe av. Den ideelle grunnen bør også være klar, jeg
+ville ikke være med på å ta fra befolkningen rettigheter. Prosjektet
+kom imidlertid aldri så langt at det ble eksponert for store
+brukergrupper, det kollapset på at mediene uansett ikke ville være med
+på en revolusjon. Jeg har flere venner og bekjente som ble med, og de
+satt i månedsvis på høye lønninger uten å gjøre noe annet enn å vente
+på avklaringer.
+
+\paragraph{Forskningen når ikke fram}
+
 \paragraph{Etikk}

### 2012-07-08 mnses/essay.tex
stanser utvikling
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@337 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 gir et bedre innblikk i hva jeg antar det er snakk om, se spesielt
-Ot.prp. nr. 46, seksjon 3.5.1 \footnote{se
+Ot.prp. nr. 46 2004-2005, seksjon 3.5.1 \footnote{se
   \url{http://www.regjeringen.no/nb/dep/kud/dok/regpubl/otprp/20042005/otprp-nr-46-2004-2005-/3/5.html?id=396451}}. 
@@
 
+En del av folkene som hadde jobbet i dette prosjektet returnerte til
+akademia og vant et stort EU-prosjekt kalt
+NoTube\footnote{se \url{http://notube.tv/}}. Et av temaene for
+prosjektet var integrasjon av sosial web med TV-mediet. Det kan være
+så enkelt som at noen kommenterer på Twitter samtidig med at man ser
+på en sending, og dette er i dag vanlig. Men prosjektet gir også
+åpninger for langt kraftigere bruk: Gjennom åpne mediesenter-systemer
+kan man legge til systemer som er for kraftige for nettbrett eller
+liknende systemer man gjerne bruker til Twitter, f.eks. vil man kunne
+gjøre en semi-automatisk analyse av en politisk debatt på TV, og så
+kunne seerne samarbeide om å gjøre en faktasjekk i sanntid på det
+politikerne sier. Resultatet kunne så distribueres gjennom sosiale
+nettverk, og kunne også nå debattlederen, som kunne konfrontere
+politikerne med opplysningene. Teknisk er dette mulig, men er det
+mulig legalt?
+
+Det er to kritiske spørsmål i så måte: Det ene er om det er mulig
+innenfor plattformene som tilbys av f.eks. RiksTV, Get, osv. Det er
+kanskje mulig, men dette er per idag meget primitive systemer, og hva
+som skal inn der styres av kommersielle hensyn. I prinsippet er det
+mulig at disse selskapene kunne støttet slik teknologi, men i praksis
+ser det ikke ut til at dette skjer.
+
+NoTube-prosjektet har
+publisert\footnote{se \url{https://github.com/libbymiller/notube-jabber-mythtv}}
+kode som bruker MythTV, som er et av de mer avanserte
+mediesentersystemene på markedet. MythTV er fri programvare. For å
+bruke MythTV må man kjøre kode som dekrypterer TV-signalet. Dette er
+en nødvendighet for bruksscenariet over. Jeg har prøvd etter beste
+evne å forstå om dette er lov eller ikke, men konklusjonen er at det
+er det helt umulig å vite.
+
+Det er en lang rekke ting som gjør det umulig å vite. For det første
+er dette til privat bruk, og dermed er det ikke opphavsrettslig
+relevant. For det andre brukes kodekortet man har kjøpt og betalt for
+på måten det er ment å brukes, så man baner seg ikke vei til andres
+data, men på den annen side er det utvilsomt ment som et teknisk
+beskyttelsessystem.
+
+Hvis ikke det at det er ikke er opphavsrettslig relevant kan hjelpe,
+så blir spørsmålet om systemet er effektivt sentralt. Det er her
+forarbeidene i Ot. prp 46 2004-2005 går fullstendig feil. La oss
+behandle følgende sitat:
+
+\begin{quote}
+Departementet antar at det i dette ikke ligger veldig strenge krav til
+systemet, og at det f.eks. ikke kan kreves at det er så effektivt at
+det nærmest er umulig å omgå. Samtidig kan systemet ikke være altfor
+enkelt utformet, og dets formål må til en viss grad oppfylles. Det er
+departementets oppfatning at systemer som nøytraliseres ved et
+tusjstrøk på selve platen eller et tastetrykk ved innlesing i
+datamaskin ikke vil oppfylle forslagets krav for beskyttelse, og at
+systemer som er så enkle å omgå ikke vil være tekniske
+beskyttelsessystemer i lovens forstand.
+\end{quote}
+
+Det er min oppfattning at definisjonen av et effektivt teknisk
+beskyttelsessystem må være at det er umulig å omgå, og således bør det
+heller ikke trenge lovbeskyttelse\footnote{Det kan bemerkes at det
+virker som hele tanken med tekniske beskyttelsessystemer virker
+fullstendig feilslått, slik at det slikt system ikke kan eksistere, og
+at det er derfor man søker lovbeskyttelse.}. Etthvert system som skal
+være mulig for en sluttbruker å bruke må være utformet slik at
+brukeren oppfatter det som enkelt, dermed må det være en stor grad av
+automatisering. Etthvert system som kan knekkes, kan også knekkes med
+programvare som kan pakkes på en slik måte at en sluttbruker oppfatter
+det som enkelt (det er også tilfelle med MythTV). Dermed vil enhver
+teknisk beskyttelsesmekanisme være enten umulig eller enkel å
+omgå. Departementets antagelse blir dermed feil.
+
+Enkelte jurister jeg har snakket med har med største letthet foreslått
+at jeg bør ta saken til retten. Dette er en ekstremt navlebeskuende
+holdning, konsekvensen av å ikke få medhold er faktisk ett års fengsel
+og eller bøter i millionklassen, rett og slett personlig ruin. Det er
+ufattelig for meg at det finnes jurister som ikke innser at kostnaden
+er altfor høy, og at en minste tvil om legaliteten av et prosjekt vil
+føre til at man rett og slett ikke gjør det. Slik stanser loven
+teknologisk utvikling. Problemet er ikke nødvendigvis hva den sier,
+men at det er formulert på en slik måte at det er helt uforståelig for
+dem den retter seg mot\footnote{Forbrukermagasinet publiserte for noen
+år siden en oversikt der de hadde spurt en rekke organisasjoners
+jurister om hva de mente om hypotetiske problemstillinger liknende den
+over, og den viste at juristene heller ikke har særlig
+forståelse. Denne referansen er ikke lenger tilgjengelig på
+nettet.}. Såvidt jeg kan forstå bryter dette også mot artikkel 29 i
+menneskerettighetene, man må ha mulighet til å forstå en lov for å
+ikke bryte den, og lovgiver må ta byrden med å gjøre dette klart
+ettersom de ikke skal innføre begrensninger uten god grunn.
+
 \paragraph{Etikk}

### 2012-07-08 mnses/essay.tex
autoritær
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@338 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+For de fleste synes den teknologiske utviklingen å gå fort. Men for
+meg, som ser trender lenge før de kommer til folk flest, går den
+sakte. For meg var det forrige tiåret et tapt tiår. Alle de store
+tingene, som sosiale nettverk er bare en blek skygge av de ideer vi
+hadde før århundrets slutt, nettbrett og smarttelefoner er bare en
+konsekvens av miniatyriseringen, de fleste ideene var tenkt allerede
+på 60-tallet, osv. En av årsakene kan være at sinnene ikke er frie til
+å utvikle teknologien for vårt alles felles beste. Folk får ikke sin
+del av ``scientific advancement and its benefits''.
+
+\paragraph{Den autoritære vitenskapen}
+
+I sin forelesning viste Svein Sjøberg at ungdommer som hadde høy grad
+av eksponering og muligens også gode test-resultater for teknologi
+ofte hadde et antagonistisk eller likegyldig forhold til
+teknologien. Det er meget farlig om vitenskapen blir oppfattet som
+autoritær og ekskluderende, ikke bare for vitenskapen selv, men for
+oppfattningene av vitenskap i befolkningen.
+
+Om det blir slik at alle oppfatter at man ikke har kontrollen over sin
+egen datamaskin i sitt eget hjem eller sin egen mobiltelefon, som man
+alltid har med seg, som er en personlig ting, så vil det gi en sterk
+følelse i befolkningen av at teknologien er autoritær, man må gi seg
+hen til utenforstående autoriteter langt inne i sin private sfære, og
+uten noen praktisk klagemulighet. Gjennom gjeldende
+opphavsrettslovgivning har vi allerede gått langt i denne retning, og
+ingenting tyder på forbedring. Dette er etter min oppfatning meget
+alvorlig, og en situasjon alle som driver med vitenskap bør engasjere
+seg i.
+
+Uten at jeg skal kommentere på Feyerabends anarkistiske synspunkt, og
+heller ikke ta opp hvor jeg mener han tar feil om vitenskapens
+autoritære natur i dag i hans foredrag fra pensum\cite{feyer}, så må
+vi ta opp hans poeng om at det ikke finnes noe i vitenskapens natur som
+forhindrer den i å bli like autoritær som tidligere ideologier. Jeg
+har vist hvordan kommersielt og politisk press gjør at teknologien
+blir fientlig overfor befolkningen, og befolkningen vil gjøre helt
+rett hvis de avviser denne teknologien. Det er en fare at de også vil
+avvise vitenskapen bak teknologien som autoritær, selv om jeg mener
+det vil være en feilslutning. Jeg er ikke enig med Feyerabend at ``any
+falsehood that can aid us in the over throw of this tyrant is to be
+welcomed'', men jeg ser at vi er i ferd med å skape en slik tyrann, og
+jeg ser sannelig ikke hvordan vi skal argumentere for å slå ned denne
+tyrannen.
+
+\paragraph{Tvil og motargumenter}
+
+Bertrand Russell skal ha sagt: ``The fundamental cause of trouble in
+the world today is that the stupid are cocksure while the intelligent
+are full of doubt.'', så i et forsøk på å sette dette essayet på den
+intelligente siden skal jeg forsøke å fremme en rekke motargumenter:
+
 \paragraph{Etikk}
 
-Min forståelse er at dette ikke er etiske spørsmål ettersom det ikke
-er en reell interessekonflikt mellom skapere og mottakere av
-verk. Disse må leve i et økosystem der de er gjensidig avhengige av
-hverandre.
+Opphavsrettslovgivningen framstilles ofte som en avveining mellom
+skaperes rettigheter og den øvrige befolkningens rettigheter. Dermed
+kan man tenkes å behandle dette som et etisk spørsmål og det faller
+naturlig å behandle det innenfor utilitarianismen, ettersom det til
+syvende og sist handler om penger, om hvor mange som kan leve av å
+skape ikke-materielle goder.
+
+Problemet sies ofte også å være en avveining mellom ``piratkopiering''
+og teknikker for å begrense denne. Det er utenfor rekkevidden til
+dette essay å kommentere alle de forskjellige årsakene til at
+salg av åndsverk i mange tilfeller har blitt problematisk, men det
+kan listes mulige andre forklaringer som at produktene som har blitt
+tilbytt har vært dårlige, kvaliteten på åndsverk har gått ned,
+salgstall har vært kunstig høye på grunn av tidligere
+formattransisjoner, økt konkurranse med produkter brukerne allerede
+har på grunn av økt gjenfinnbarhet, økonomiske nedgangstider, flere
+konkurrerende nisjer, osv.
+
+Til syvende og sist er alle i samme båt: Alle ønsker kulturproduksjon,
+og alle har noe å hente fra den. Alle ønsker teknologisk utvikling, og
+mange er økonomisk avhengig av at den skjer. Alle ønsker vitenskap og
+forskning, og alle har noe å hente fra den. 
+Min oppfattning er derfor at dette ikke er snakk om en avveining
+mellom interessenter, og derfor må man finne en
+fellesskapsløsning. Det er ingen reell konflikt, og det er heller
+dermed ikke snakk om å gjøre avveininger for å maksimere felles
+lykke. Medieindustrien har ført en kampanje over flere
+tiår som har vært ekstremt polariserende, og denne må
+opphøre. Politikerne må igjen etablere et rammeverk der det blir mulig
+å etablere et økosystem der både skapende og andre lever i gjensidig
+avhengighet av hverandre. Den kan man ikke gjøre med uforståelige
+lover, fengselsstraff og bøter.

### 2012-07-09 mnses/essay.tex
tvil
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@339 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 are full of doubt.'', så i et forsøk på å sette dette essayet på den
-intelligente siden skal jeg forsøke å fremme en rekke motargumenter:
+intelligente siden skal jeg forsøke å gi en oversikt over hvor jeg kan
+ta feil:
+
+Essayets akilleshel er at jeg mangler relevante referanser til
+formuleringer av typen ``mange mener at'', som jeg så angriper. Når
+jeg påstår at skapernes rettigheter er begrunnet i befolkningens
+interesse, så finner man følgende i den amerikanske grunnloven: 
+\begin{quote}
+To promote the Progress of Science and useful Arts, by securing for
+limited Times to Authors and Inventors the exclusive Right to their
+respective Writings and Discoveries.
+\end{quote}
+Når jeg likevel ikke bruker denne som referanse, så er det fordi den
+ikke nødvendigvis er relevant for norske forhold (selv om det også kan
+argumenteres at dagens lovgivning er lobbiet igjennom av
+Hollywood). Den største risikoen man løper ved å mangle relevante
+referanser er at man kan argumentere mot en ``stråmann'', at ingen
+eller få faktisk holder den meningen man argumenterer mot.
+
+Selv om Hollywood kan ha lobbiet igjennom dagens lovgivning, så
+kanskje man kan stole på at tilsvarende motkrefter i samfunnet holder
+det langs en rimelig middelvei. I dag bruker f.eks. Google enorme
+summer på lobbyvirksomhet og argumenterer stort sett for en langt mer
+liberal opphavsrettslovgivning. Likevel vil jeg ønske meg en langt mer
+grasrotorientert politikk, et langt mer direkte demokrati der
+befolkningen er bedre orientert og mer involvert i politikk. Videre
+kan det synes som politikernes håndverk består i å ta litt her og gi
+litt der, slik at alle skal være litt misfornøyd, snarere enn å
+vurdere argumenter for hvorfor en part kan ta feil.
+
+Det er lett å peke på en manglende behandling av disse spørsmålene i
+media, og at dette er grunnen til at politikerne ikke ser
+motargumenter. Konspiratoriske stemmer kan påstå at dette er fordi
+massemediene også er med i medieindustrien. Blant oss som tviler kan
+dette heller forklares som en kognitiv effekt kjent som ``Hostile
+media effect''. Kanskje politikerne faktisk \emph{er} bedre orientert
+enn oss som er skeptiske til opphavsrettslovgivningen. Kanskje er
+miljøene enkelte teknologer som meg ferdes i blitt så konforme at vi
+er ute av stand til å se argumentene for dagens lovgivning, vi er
+drevet av en annen kognitiv effekt kjent som ``false consensus
+effect'', kanskje hele dette essayet kun er et resultat av
+forfatterens naive
+realisme\footnote{Se \url{http://en.wikipedia.org/wiki/Na\%C3\%AFve_realism_\%28psychology\%29}
+for en oversikt over dette begrepet og linker til referanser om
+effektene over.}.
+
+Jeg er også at det er fryktelig vanskelig å argumentere for at
+den teknologiske utviklingen nesten står stille, når de aller fleste i
+samfunnet føler at den går altfor fort. Den teknologiske utviklingen
+jeg ønsker meg er en annen retning, der befolkningens behov er i
+fokus. La meg derfor ta et eksempel på hvordan ``tekniske
+beskyttelsessystemer'' kunne være ønskelig: Si for eksempel at en ungdom
+flykter fra et voldelig hjem og får en ny identitet. Si så at begge
+identiteter finnes i sosiale nettverk, og at en venn uforvarende
+kommer til å si at disse to identitetene er samme person. Dette vil
+idag ikke være mulig å fjerne, like lite som et kompromitterende
+bilde, og en overgriper kunne da avsløre den nye identiteten, hvis da
+ikke myndighetene hadde en mulighet til å komme inn og fjerne denne
+opplysningen fra alles datamaskiner, akkurat som Amazon fjernet ``Big
+Brother'' fra kundenes Kindle. Jeg synes fortsatt at ``tekniske
+beskyttelsessystemer'' er noen god ide, fordi vi tenker oss at sosiale
+nettverk skulle kunne gi oss tilsvarende sikkerhet i en verden der den
+teknologiske utviklingen går mer i befolkningens favør.
+
+Kanskje er ikke teknologisk utvikling engang ønskelig. Jeg har
+allerede skrevet at det er lettere å tjene penger ved å begrense
+distribusjonen av programvare, og all denne innovasjonen vil skape
+enda mer disrupsjon. Folk må finne seg nye jobber, nye
+forretningsmodeller må finnes opp. Det vil bli perioder der man famler
+i mørket og etablert industri taper store penger. Kanskje er det
+legitimt for myndigheter å være autoritære for å sørge for
+stabilitet. Kanskje er menneskerettighetserklæringen skrevet i en tid
+der man hadde en naiv tro på framskritt, kanskje er det vitenskapens
+egen skyld at myndighetene må være så autoritære, folk flest kan ikke
+finne lykke i  så store omveltninger som vitenskapen krever av dem?
 

### 2012-07-09 mnses/essay.tex
texing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@340 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
+\documentclass[11pt,norsk,a4paper]{article}
+\usepackage{babel,cite,hyperref}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+
+\title{Menneskerettigheter og autoritær opphavsrettslovgivning}
+\author{Kjetil Kjernsmo}
+
+\begin{document}
+\maketitle
+
 I mars 2006 fikk jeg en mail der jeg ble bedt om å komme til London
@@
 amerikanske Senatet, med tittel ``Content Protection Status
-Report''\footnote{Documentene var inntil nylig tilgjengelig fra
+Report''\footnote{Dokumentene var inntil nylig tilgjengelig fra
   Senatets websider, men nå finner jeg kun en omtale på Wikipedia:
-  \url{http://en.wikipedia.org/wiki/Content_Protection_Status_Report}.
+  \url{http://en.wikipedia.org/wiki/Content_Protection_Status_Report}}.
 I disse dokumentene ble det foreslått tiltak som at all elektronikk
@@
 lover, fengselsstraff og bøter.
+
+\bibliography{essay}{}
+\bibliographystyle{plain}
+\end{document}

### 2012-07-09 mnses/essay.tex
Use URL field in article
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@341 43ef1190-90fd-4dd7-a769-d0121f02b80e
Angi dato
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@343 43ef1190-90fd-4dd7-a769-d0121f02b80e
første gjennomlesning
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@344 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Dog ser jeg at det kan finnes unntak fra denne regelen. Jeg ser det
-som utvilsomt at det er enklere å tjene penger på programvare som er
-underlagt begrensninger basert på egen erfaring. Menneskerettighetene
-sier ikke at det skal være enkelt å tjene penger, men det kan tenkes
-at det blir så vanskelig å få til en investering i f.eks. veldig
-sektor-spesifik programvare at ingen vil ta seg råd til å utvikle
-den. Dette vil dermed gjøre at man ikke får til teknologisk utvikling.
+som utvilsomt, basert på egen erfaring, at det er enklere å tjene
+penger på programvare som er underlagt
+begrensninger. Menneskerettighetene sier ikke at det skal være enkelt
+å tjene penger, men det kan tenkes at det blir så vanskelig å få til
+en investering i f.eks. veldig sektor-spesifik programvare at ingen
+vil ta seg råd til å utvikle den. Dette vil dermed gjøre at man ikke
+får til teknologisk utvikling.
 
 Vi må videre spørre, gitt suksessen til Wikipedia og til fri
-programvare, hvilke andre områder er det det er illegitimt å begrense
+programvare, \emph{hvilke andre områder er det det er illegitimt å begrense
 befolkningens frihet? Hvilke andre områder vil man kunne ha samme eller
-forbedret kunstnerisk og vitenskapelig framgang uten slike begrensninger?
+forbedret kunstnerisk og vitenskapelig framgang uten slike begrensninger?}
 
@@
 bøkene de fjernet var George Orwells ``Big Brother''. I denne boka har
-myndighetene lagd et ``memory hole'' som brukes til å skrive om
-historien. Essensen til tekniske beskyttelsessystemer er at man
-tillater programvare på sin egen datamaskin, lesebrett eller
-mobiltelefon, som man selv ikke har noen autoritet over. Denne
-autoriteten beholdes av noen andre, og man har selv ingen kontroll på
-hva de kan eller vil bruke denne autoriteten til. I tilfellet over er
-det brukt til nettopp å lage Orwells ``memory hole'', og per idag har
-altså lovgiver ment at slik bør det være. Vi har idag ingen sikkerhet
-mot misbruk overhodet.
+myndighetene lagd et ``memory hole'' som brukes til å fjerne
+ubehagelige fakta, slik at man skrive om historien. Essensen til
+tekniske beskyttelsessystemer er at man blir tvunget til å ha
+programvare på sin egen datamaskin, lesebrett eller mobiltelefon, som
+man selv ikke har noen autoritet over. Denne autoriteten beholdes av
+noen andre, og man har selv ingen kontroll på hva de kan eller vil
+bruke denne autoriteten til. I tilfellet over er det brukt til nettopp
+å lage Orwells ``memory hole'', og per idag har altså lovgiver ment at
+slik bør det være. Vi har idag ingen sikkerhet mot misbruk overhodet.
 
@@
 det ville føre til at man lagde en infrastruktur der Hollywood ville
-fått absolutt autoritet over all produksjon. Dette er heldigvis ikke
+fått absolutt autoritet over all produksjon, man ville i prinsippet
+kunne hindre alle ytringer. Dette er heldigvis ikke
 gjennomførbart, men det at det ble presentert og tatt seriøst viser
@@
 
+Jeg uttrykker frustrasjon over at man ikke får ta del i dagens
+forskning, men si at det bare tar ti år lenger før befolkningen får
+dem gjennom de vanlige kommersielle tilbyderne? Er det da bra nok til
+at menneskerettighetene ikke er brytt? Eller hva om tjenestene er der,
+de er bare langt mer primitive enn det de kunne ha vært om man fritt
+kunne skape tjenester? Spotify er et eksempel på en meget
+primitiv tjeneste som fyller mange behov, men som ikke tar inn over
+seg behovene til bredden av musikkartister.
+
 Kanskje er ikke teknologisk utvikling engang ønskelig. Jeg har
@@
 \bibliography{essay}{}
-\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 \end{document}

### 2012-07-09 mnses/essay.tex
emph spørsmål
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@345 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Det blir derfor legitimt å spørre hvor norske politikere skal ha
+\emph{Det blir derfor legitimt å spørre hvor norske politikere skal ha
 større tillit til et medieimperium som beviselig har ulovlig overvåket
-flere titalls mennesker\cite{newscorpphone}. Det er per idag ingen
+flere titalls mennesker\cite{newscorpphone}.} Det er per idag ingen
 indikasjoner på at dette faktisk skjer, men det viktige
@@
 
-Man må derfor stille spørsmålet: Når befolkningen som forbrukere med
+Man må derfor stille spørsmålet: \emph{Når befolkningen som forbrukere med
 stort flertall har forkastet et tiltak, skal det da være mulig i et
 demokrati at politikerne opprettholder tiltaket når tiltaket begrenser
-befolkningens frihet? Etter mitt skjønn burde det være politisk
+befolkningens frihet?} Etter mitt skjønn burde det være politisk
 umulig; fra et sosialdemokratisk perspektiv, der politikken anses å

### 2012-07-13 mnses/essay.tex
korrektur
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@346 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\emph{Det blir derfor legitimt å spørre hvor norske politikere skal ha
+\emph{Det blir derfor legitimt å spørre hvorfor norske politikere skal ha
 større tillit til et medieimperium som beviselig har ulovlig overvåket
-flere titalls mennesker\cite{newscorpphone}.} Det er per idag ingen
+flere titalls mennesker\cite{newscorpphone} enn til egne velgere.} Det er per idag ingen
 indikasjoner på at dette faktisk skjer, men det viktige
@@
 Brother'' fra kundenes Kindle. Jeg synes fortsatt at ``tekniske
-beskyttelsessystemer'' er noen god ide, fordi vi tenker oss at sosiale
+beskyttelsessystemer'' ikke er noen god ide, fordi vi tenker oss at sosiale
 nettverk skulle kunne gi oss tilsvarende sikkerhet i en verden der den
@@
 dem gjennom de vanlige kommersielle tilbyderne? Er det da bra nok til
-at menneskerettighetene ikke er brytt? Eller hva om tjenestene er der,
+at menneskerettighetene ikke er brutt? Eller hva om tjenestene er der,
 de er bare langt mer primitive enn det de kunne ha vært om man fritt
@@
 dette essay å kommentere alle de forskjellige årsakene til at
-salg av åndsverk i mange tilfeller har blitt problematisk, men det
+salg av åndsverk i mange tilfeller har blitt vanskelig men det
 kan listes mulige andre forklaringer som at produktene som har blitt
-tilbytt har vært dårlige, kvaliteten på åndsverk har gått ned,
+tilbudt har vært dårlige, kvaliteten på åndsverk har gått ned,
 salgstall har vært kunstig høye på grunn av tidligere

### 2012-07-13 mnses/essay.tex
trimme litt
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@347 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Vi ser det samme i mange andre felter, både innenfor informasjon, data
-og programvare. Man kan idag fint leve uten programvare som i et
-regime der programvare ikke var vernet likevel ville eksistert i sin
-nåværende form. Dette er programvare som gjerne benevnes som ``fri
-programvare'' . Jeg bruker selv nesten utelukkende fri programvare, og
-utgir selv det jeg skriver som fri programvare, og man kan finne alt
-fra meget enkle programmer til meget avanserte systemer for
-resonnering. Blant web-servere så er det to slike systemer som
-konkurrerer i popularitet med Microsofts begrensende
-tilbud\cite{netcraft}. Det er min oppfatning at dette betyr at
+og programvare. Det er min oppfatning at dette betyr at
 programvare i hovedsak ikke bør kunne underlegges begrensninger.
@@
 rolle er her begrenset. For å forsto hvorfor det ble så viktig for meg
-å avslå et åpenbart godt tilbud, så må vi se til straffelovens
+å avslå et åpenbart godt tilbud, så må vi se til Straffelovens
 §262. Denne paragrafen i seg selv er totalt uforståelig for meg sett i
@@
 kjøpt disse bøkene, se f.eks.~\cite{nytamazon}. Ironien er at blant
-bøkene de fjernet var George Orwells ``Big Brother''. I denne boka har
-myndighetene lagd et ``memory hole'' som brukes til å fjerne
-ubehagelige fakta, slik at man skrive om historien. Essensen til
-tekniske beskyttelsessystemer er at man blir tvunget til å ha
+bøkene de fjernet var George Orwells ``Big Brother''. I handlingen i
+denne boka har myndighetene lagd et ``memory hole'' som brukes til å
+fjerne ubehagelige fakta, slik at man skrive om historien. Essensen
+til tekniske beskyttelsessystemer er at man blir tvunget til å ha
 programvare på sin egen datamaskin, lesebrett eller mobiltelefon, som
@@
 det nærmest er umulig å omgå. Samtidig kan systemet ikke være altfor
-enkelt utformet, og dets formål må til en viss grad oppfylles. Det er
-departementets oppfatning at systemer som nøytraliseres ved et
-tusjstrøk på selve platen eller et tastetrykk ved innlesing i
-datamaskin ikke vil oppfylle forslagets krav for beskyttelse, og at
-systemer som er så enkle å omgå ikke vil være tekniske
-beskyttelsessystemer i lovens forstand.
+enkelt utformet, og dets formål må til en viss grad oppfylles. 
 \end{quote}

### 2012-07-13 mnses/essay.tex
trimming
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@348 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 at det er umulig for befolkningen å vite om man bryter loven eller
-ikke, uansett om hvor godt man kjenner lovgivningen. Denne tvilen om
-hva som er lov og ikke er absolutt destruktiv.
+ikke, uansett om hvor godt man kjenner lovgivningen.
 
@@
 slik bør det være. Vi har idag ingen sikkerhet mot misbruk overhodet.
-
-Det mest skremmende som har kommet er en serie med dokumenter skrevet
-av Motion Picture Association of America og adressert til det
-amerikanske Senatet, med tittel ``Content Protection Status
-Report''\footnote{Dokumentene var inntil nylig tilgjengelig fra
-  Senatets websider, men nå finner jeg kun en omtale på Wikipedia:
-  \url{http://en.wikipedia.org/wiki/Content_Protection_Status_Report}}.
-I disse dokumentene ble det foreslått tiltak som at all elektronikk
-som ble lagd med analog-til-digitale konvertere måtte utstyres med noe
-som gjorde at de kunne slås av hvis det ble detektert at de ble brukt
-til å ta opp vernet verk. Bakgrunnen for forslaget er at man ønsker å
-gjøre det umulig for uautoriserte å lage en mikrofon som kan brukes
-til å gjøre digitale opptak. Grunnen til at dette er så skremmende er at
-det ville føre til at man lagde en infrastruktur der Hollywood ville
-fått absolutt autoritet over all produksjon, man ville i prinsippet
-kunne hindre alle ytringer. Dette er heldigvis ikke
-gjennomførbart, men det at det ble presentert og tatt seriøst viser
-hvor langt man er villig til å gå.
 
@@
 større tillit til et medieimperium som beviselig har ulovlig overvåket
-flere titalls mennesker\cite{newscorpphone} enn til egne velgere.} Det er per idag ingen
-indikasjoner på at dette faktisk skjer, men det viktige
-tillitsspørsmålet bør være tverrpolitisk interessant.
+flere titalls mennesker\cite{newscorpphone} enn til egne velgere.}
 
@@
 ville ikke være med på å ta fra befolkningen rettigheter. Prosjektet
-kom imidlertid aldri så langt at det ble eksponert for store
-brukergrupper, det kollapset på at mediene uansett ikke ville være med
-på en revolusjon. Jeg har flere venner og bekjente som ble med, og de
-satt i månedsvis på høye lønninger uten å gjøre noe annet enn å vente
-på avklaringer.
+kollapset ble som forventet ingen suksess.
 
@@
 
-Uten at jeg skal kommentere på Feyerabends anarkistiske synspunkt, og
+Uten at jeg skal kommentere Feyerabends anarkistiske synspunkt, og
 heller ikke ta opp hvor jeg mener han tar feil om vitenskapens
@@
 jeg ønsker meg er en annen retning, der befolkningens behov er i
-fokus. La meg derfor ta et eksempel på hvordan ``tekniske
-beskyttelsessystemer'' kunne være ønskelig: Si for eksempel at en ungdom
-flykter fra et voldelig hjem og får en ny identitet. Si så at begge
-identiteter finnes i sosiale nettverk, og at en venn uforvarende
-kommer til å si at disse to identitetene er samme person. Dette vil
-idag ikke være mulig å fjerne, like lite som et kompromitterende
-bilde, og en overgriper kunne da avsløre den nye identiteten, hvis da
-ikke myndighetene hadde en mulighet til å komme inn og fjerne denne
-opplysningen fra alles datamaskiner, akkurat som Amazon fjernet ``Big
-Brother'' fra kundenes Kindle. Jeg synes fortsatt at ``tekniske
-beskyttelsessystemer'' ikke er noen god ide, fordi vi tenker oss at sosiale
-nettverk skulle kunne gi oss tilsvarende sikkerhet i en verden der den
-teknologiske utviklingen går mer i befolkningens favør.
+fokus. Det finnes  eksempler på hvordan ``tekniske
+beskyttelsessystemer'' kan være i befolkningens tjeneste, men de
+problematiske sidene ved dette gjør at jeg mener at det er vesentlig
+at det må innoveres uten å bruke slike virkemidler.
 

### 2012-09-06 conditional/trunk/selectivity-basic.tex
Rename paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@357 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2011-07-20 conditional/trunk/selectivity-hexastore.tex
@@
 Graph Patterns \cite{Stocker:2008:SBG:1367497.1367578} and in more
-general cases. \todo{more references. OptARQ?}
+general cases. \todo{more references. OptARQ? Some go into conditional
+regimes, actually.}
 
@@
 \end{verbatim}
-We can easily see that $sel(T) = 2/7$ since two out of the 7 triples
-match the pattern. To see how Eq.~\ref{eq:sel} is used, we note that
-$sel(S) = 1$ since it is a variable. $sel(P | S) = 2/7$ since the
-predicate matches the two occurences of \rdfterm{rdf:type} and the
-subject variable doesn't influence this. Finally, $sel(O | S \cap P) =
-1$ since the only two occurences of \rdfterm{foaf:Person} occurs in
-the same triples as \rdfterm{rdf:type}. Thus, we arrive at the
-expected result.
+We can easily see that the correct answer is $sel(T) = 2/7$ since two
+out of the 7 triples match the pattern. To see how Eq.~\ref{eq:sel} is
+used, we note that $sel(S) = 1$ since it is a variable. $sel(P | S) =
+2/7$ since the predicate matches the two occurences of
+\rdfterm{rdf:type} and the subject variable doesn't influence
+this. Finally, $sel(O | S \cap P) = 1$ since the only two occurences
+of \rdfterm{foaf:Person} occurs in the same triples as
+\rdfterm{rdf:type}. Thus, we arrive at the expected result.
 

### 2012-09-06 conditional/trunk/selectivity-basic.tex
related work stub
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@358 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\title{Conditional selectivity for Hexastores}
+\title{Conditional selectivity for triplestores}
 
@@
   assumption of independence of the selectivity of RDF terms can be
-  abandoned. We shall also find it instructive to explore how
-  selectivity can be determined exactly when a SPARQL query is
-  executed on a Hexastore triple store
-  \cite{Weiss:2008:HSI:1453856.1453965}. Finally, we will suggest
-  further directions for selectivity estimation in the conditional
-  regime.
+  abandoned. We proceed to extend SPLENDID\cite{splendid} using the
+  new formalism and demonstrate how the VoID data descriptions can be
+  further exploited in the conditional regime. Finally, we will
+  explore how selectivity can be determined exactly when a SPARQL
+  query is executed on a fully indexed triplestore, such as
+  Hexastore\cite{Weiss:2008:HSI:1453856.1453965}. Finally, we will
+  suggest further directions for selectivity estimation in the
+  conditional regime.
 
@@
 different from e.g. literals.
+
+\section{Related Work}
+
+\cite{splendid}, \cite{wodqa} \todo{write this}.
 

### 2012-09-06 conditional/trunk/selectivity-basic.tex
some reformulation about hexastore
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@359 43ef1190-90fd-4dd7-a769-d0121f02b80e
start void stuff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@362 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{cite}
+\usepackage[utf8]{inputenc}
 
@@
   abandoned. We proceed to extend SPLENDID\cite{splendid} using the
-  new formalism and demonstrate how the VoID data descriptions can be
+  new formalism and demonstrate how VoID data descriptions\cite{void} can be
   further exploited in the conditional regime. Finally, we will
@@
 
-\section{Conditional selectivity}
+\section{Conditional selectivity}\label{sec:cond}
 
@@
 
-\section{Exact selectivity for Hexastore}
+\section{Exploiting VoID descriptions}
 
-The Hexastore was first described in
+In SPLENDID\cite{splendid} the authors showed how statistics exposed
+using the VoID data description vocabulary\cite{void} could be
+exploited for query optimization. The authors made the assumption that
+subjects and objects were uniformly distributed and independent from
+the predicate. The present paper extends their work by relaxing these
+assumptions.
+
+In section~\ref{sec:cond}, we saw
+
+\section{Exact selectivity for fully indexed stores}
+
+Some triplestores fully index all terms. A prominent example is the
+Hexastore that was first described in
 \cite{Weiss:2008:HSI:1453856.1453965}, which sought to exploit a key
@@
 
+We also note that exposing a suitable VoID description should be cheap
+using the above formalism.
+
 \section{Further work on conditional selectivity}

### 2012-09-06 conditional/trunk/selectivity-basic.tex
conditional triples in a pattern
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@363 43ef1190-90fd-4dd7-a769-d0121f02b80e
first derived selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@365 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 classes are likely objects, and their frequency would likely be
-different from e.g. literals.
+different from e.g. literals. Moreover, it is easy to see when triples
+in a basic graph pattern are likely to be dependent: If one triple has
+\rdfterm{rdf:type} as predicate and \rdfterm{foaf:Person} as object,
+the next triple is much more likely to contain the very
+common predicate \rdfterm{foaf:name} than the esoteric
+\rdfterm{foaf:dnaChecksum}.
 
@@
 
-In section~\ref{sec:cond}, we saw
+In section~\ref{sec:cond}, we derived an equation for conditional
+selectivity, i.e. RDF terms are not independent of each
+other.
+
+In the follwing, we denote a bounded term with a lowercase letter, $s$
+for a subject, $p$ for a predicate and $o$ for object, and variables
+are denoted by a question mark, $?$. For example, $p =
+\rdfterm{rdf:type}$. We also adopt SPLENDIDs convention of denoting
+the number of all triples in certain data source $|d|$.
+
+We shall now systematically go through different triple patterns and
+see how the selectivity can be derived from VoID descriptions per the
+most recent vocabulary\cite{voidnote}:
+
+\paragraph{Subject unbound, object and predicate bound}
+
+In this case, we can use Eq.~\ref{eq:sel} directly. We again note that
+since the the unbound subject will always have selectivity of
+unity.
+Eq.~\ref{eq:sel} thus reduces to:
+
+\begin{equation}\label{eq:selsubjun}
+sel(T_{?~p~o}) = sel(O | P) \cdot sel(P) . 
+\end{equation}
+
+To relate this to VoID, we note that VoID includes the concept of
+\emph{property partitions} (see section~4.5 in \cite{voidnote}): ``A
+property-based partition contains only those triples of a dataset that
+use a particular predicate.'', and since a partition is itself a
+dataset, the same statistical properties can be applied, such as
+\rdfterm{void:triples}, \rdfterm{void:distinctSubjects} and
+\rdfterm{void:distinctObjects}.
+
+Since property partitions detail statistics given a certain predicate,
+they can be used in the conditional case. We can therefore write:
+
+\begin{equation}\label{eq:selsubjunp}
+sel(P = p) = \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|}
+\end{equation}
+and
+\begin{equation}\label{eq:selsubjunop}
+sel(O = o|P = p) =
+\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
+\end{equation}
+
+Thus,
+\begin{equation}\label{eq:selsubjunop}
+sel(T_{?~p~o}) =
+\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} .
+\end{equation}\todo{wtf}
 

### 2012-09-07 conditional/trunk/selectivity-basic.tex
fix label
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@366 43ef1190-90fd-4dd7-a769-d0121f02b80e
oooh, noooooes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@367 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{equation}\label{eq:selsubjun}
-sel(T_{?~p~o}) = sel(O | P) \cdot sel(P) . 
+sel(T_{\{?~p~o\}}) = sel(O | P) \cdot sel(P) . 
 \end{equation}
@@
 \end{equation}
+Thus,
+\begin{equation}\label{eq:selsubjunres}
+sel(T_{\{?~p~o\}}) =
+\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} ,
+\end{equation}\todo{WRONG!}
+where the notation is inspired by SPARQL 1.1 Property Paths,
+for example the relevant part of a VoID description for the second
+example in section~\ref{sec:example} would be:
+\begin{verbatim}
+</dataset> a void:Dataset ;
+           void:propertyPartition [
+                void:property foaf:firstName;
+                void:distinctObjects 2 .
+           ]
+\end{verbatim}
 
-Thus,
-\begin{equation}\label{eq:selsubjunop}
-sel(T_{?~p~o}) =
-\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} .
-\end{equation}\todo{wtf}
+there exists a \rdfterm{void:propertyPartition} predicate that
+has an object which is again linked to a dataset subset containing the
+property $p$ as predicate of \text
 

### 2012-09-17 conditional/trunk/selectivity-basic.tex
headline
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@368 43ef1190-90fd-4dd7-a769-d0121f02b80e
define \sel macro
Comment out blatantly flawed stuff


git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@369 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-09-06 conditional/trunk/selectivity-basic.tex
@@
 \newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
+\newcommand{\sel}[1]{\ensuremath{sel\left(#1\right)}}
 
@@
 \begin{equation}\label{eq:sel}
-sel(T) = sel(S \cap P \cap O) = sel(O | S \cap P) \cdot sel(P | S)
-\cdot sel(S) . 
+\sel{T} = \sel{S \cap P \cap O} = \sel{O | S \cap P} \cdot \sel{P | S}
+\cdot \sel{S} . 
 \end{equation}
@@
 \end{verbatim}
-We can easily see that the correct answer is $sel(T) = 2/7$ since two
+We can easily see that the correct answer is $\sel{T} = 2/7$ since two
 out of the 7 triples match the pattern. To see how Eq.~\ref{eq:sel} is
-used, we note that $sel(S) = 1$ since it is a variable. $sel(P | S) =
+used, we note that $\sel{S} = 1$ since it is a variable. $\sel{P | S} =
 2/7$ since the predicate matches the two occurences of
 \rdfterm{rdf:type} and the subject variable doesn't influence
-this. Finally, $sel(O | S \cap P) = 1$ since the only two occurences
+this. Finally, $\sel{O | S \cap P} = 1$ since the only two occurences
 of \rdfterm{foaf:Person} occurs in the same triples as
@@
 \end{verbatim}
-We still have $sel(S) = 1$ as above, and there are still two
-predicates that match, so $sel(P | S) = 2/7$, but only half of those
-match the literal, so $sel(O | S \cap P) = 1/2$, giving us the
-expected $sel(T) = 1/7$.
+We still have $\sel{S} = 1$ as above, and there are still two
+predicates that match, so $\sel{P | S} = 2/7$, but only half of those
+match the literal, so $\sel{O | S \cap P} = 1/2$, giving us the
+expected $\sel{T} = 1/7$.
 
@@
 
+\subsection{Terms that can be computed}
+
+\begin{equation}\label{eq:selpp}
+\sel{P = p} = 
+\end{equation}
+
+\subsection{Triple pattern cases}
+
 We shall now systematically go through different triple patterns and
@@
 
-\begin{equation}\label{eq:selsubjun}
-sel(T_{?~p~o}) = sel(O | P) \cdot sel(P) . 
-\end{equation}
+% \begin{equation}\label{eq:selsubjun}
+% \sel{T_{\{?~p~o\}}) = \sel{O | P) \cdot \sel{P) . 
+% \end{equation}
 
@@
 
-Since property partitions detail statistics given a certain predicate,
-they can be used in the conditional case. We can therefore write:
+% Since property partitions detail statistics given a certain predicate,
+% they can be used in the conditional case. We can therefore write:
 
-\begin{equation}\label{eq:selsubjunp}
-sel(P = p) = \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|}
-\end{equation}
-and
-\begin{equation}\label{eq:selsubjunop}
-sel(O = o|P = p) =
-\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
-\end{equation}
+% \begin{equation}\label{eq:selsubjunp}
+% \sel{P = p) = \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|}
+% \end{equation}
+% and
+% \begin{equation}\label{eq:selsubjunop}
+% \sel{O = o|P = p) =
+% \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
+% \end{equation}
+% Thus,
+% \begin{equation}\label{eq:selsubjunres}
+% \sel{T_{\{?~p~o\}}) =
+% \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} ,
+% \end{equation}\todo{WRONG!}
+% where the notation is inspired by SPARQL 1.1 Property Paths,
+% for example the relevant part of a VoID description for the second
+% example in section~\ref{sec:example} would be:
+% \begin{verbatim}
+% </dataset> a void:Dataset ;
+%            void:propertyPartition [
+%                 void:property foaf:firstName;
+%                 void:distinctObjects 2 .
+%            ]
+% \end{verbatim}
 
-Thus,
-\begin{equation}\label{eq:selsubjunop}
-sel(T_{?~p~o}) =
-\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} .
-\end{equation}\todo{wtf}
+% there exists a \rdfterm{void:propertyPartition} predicate that
+% has an object which is again linked to a dataset subset containing the
+% property $p$ as predicate of
 
@@
 \begin{equation}
-sel(S = s_i) = \frac{cm(s_i)}{i} ,
+\sel{S = s_i} = \frac{cm(s_i)}{i} ,
 \end{equation}
 \begin{equation}
-sel(P = p_{n_i}^i|S = s_i) = \frac{cm(p_{n_i}^i)}{n_i}
+\sel{P = p_{n_i}^i|S = s_i} = \frac{cm(p_{n_i}^i)}{n_i}
 \end{equation}
@@
 \begin{equation}
-sel(O = o_{k_{i,n_i}}^{i,n_i} |S = s_i \cap P = p_{n_i}^i) =
+\sel{O = o_{k_{i,n_i}}^{i,n_i} |S = s_i \cap P = p_{n_i}^i} =
 \frac{cm(o_{k_{i,n_i}}^{i,n_i})}{k_{i,n_i}} .

### 2012-09-17 conditional/trunk/selectivity-basic.tex
write down the equation parts that can be computed
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@370 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Finally, note that for variables, the selectivity will always be
-unity, whether conditional or independent, since it will match all
-possible RDF terms.
+unity, since it will match all possible RDF terms.
 
@@
 
-In the follwing, we denote a bounded term with a lowercase letter, $s$
+In the following, we shall have to introduce to more elaborate
+notation to make sure we maintain control of bounded and unbounded
+terms. We denote a bounded term with a lowercase letter, $s$
 for a subject, $p$ for a predicate and $o$ for object, and variables
@@
 
-\subsection{Terms that can be computed}
+\subsection{Parts that can be computed}
 
-\begin{equation}\label{eq:selpp}
+There will be a number of different equations that arises from using
+the commutative property of intersections, and some parts of these
+equations is possible to compute from VoID descriptions. We have
+already noted the most trivial case, for variables we have:
+
+\begin{equation}\label{eq:selvar}
+\sel{S = ?} = \sel{P = ?} = \sel{O = ?} = 1 .
+\end{equation}
+
+We note that VoID includes the concept of \emph{property partitions}
+(see section~4.5 in \cite{voidnote}): ``A property-based partition
+contains only those triples of a dataset that use a particular
+predicate.'', and since a partition is itself a dataset, the same
+statistical properties can be applied, such as \rdfterm{void:triples},
+\rdfterm{void:distinctSubjects} and \rdfterm{void:distinctObjects}.
+
+In the case where the predicate is known, we have
+\begin{equation}\label{eq:selPp}
  \sel{P = p} =
+   \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|} ,
 \end{equation}
+where the notation is inspired by SPARQL 1.1 Property Paths, for
+example the relevant part of a VoID description for the second example
+in section~\ref{sec:example} would be:
+\begin{verbatim}
+ </dataset> a void:Dataset ;
+            void:propertyPartition [
+                void:property foaf:firstName;
+                void:triples 2 .
+            ]
+\end{verbatim}
+where $p = \rdfterm{foaf:firstName}$.
+
+In the case where there is a given predicate but the subject is a
+variable, we can use the \rdfterm{void:distinctSubjects} property:
+\begin{equation}\label{eq:selSvPp}
+ \sel{S = ? | P = p} =
+   \frac{\rdfterm{void:propertyPartition/void:distinctSubjects}}{\rdfterm{void:propertyPartition/void:triples}} ,
+\end{equation}
+and similarly where the object is a variable:
+\begin{equation}\label{eq:selOvPp}
+ \sel{O = ? | P = p} =
+   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
+\end{equation}
+
+We also have in the special case of a given class, i.e. predicate
+\rdfterm{rdf:type} or $P = a$ and $O = c$, we can use
+\rdfterm{void:classPartition} (also defined in section~4.5 in
+\cite{voidnote}):
+\begin{equation}\label{eq:selSvPaOc}
+ \sel{S = ? | P = a \cap O = c } =
+    \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
+\end{equation}
+
+The above are the parts of the equations that can be computed directly
+from the VoID description, and in the next section we shall see how
+they can be exploited most efficiently.
 
@@
 % \end{equation}
-
-To relate this to VoID, we note that VoID includes the concept of
-\emph{property partitions} (see section~4.5 in \cite{voidnote}): ``A
-property-based partition contains only those triples of a dataset that
-use a particular predicate.'', and since a partition is itself a
-dataset, the same statistical properties can be applied, such as
-\rdfterm{void:triples}, \rdfterm{void:distinctSubjects} and
-\rdfterm{void:distinctObjects}.
 

### 2012-09-17 conditional/trunk/selectivity-basic.tex
use upright font for sel
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@371 43ef1190-90fd-4dd7-a769-d0121f02b80e
note |d| is void:triples
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@372 43ef1190-90fd-4dd7-a769-d0121f02b80e
Subject and object unbound, predicate bound
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@373 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
-\newcommand{\sel}[1]{\ensuremath{sel\left(#1\right)}}
+\newcommand{\sel}[1]{\ensuremath{\mathrm{sel}\left(#1\right)}}
 
@@
 \rdfterm{rdf:type}$. We also adopt SPLENDIDs convention of denoting
-the number of all triples in certain data source $|d|$.
+the number of all triples in certain data source $|d|$ (which can also
+be found as the \rdfterm{void:triples} of the dataset).
 
@@
 
+\paragraph{Subject and object unbound, predicate bound}
+
+By using Eq.~\ref{eq:sel} and swapping $S$ and $P$, we have 
+\begin{equation}\label{eq:selsuobunpb}
+\sel{T_{\{?~p~?\}}} = \sel{O = ? | P = p \cap S = ?} \cdot 
+\sel{S = ? | P = p} \cdot
+\sel{P = p} . 
+\end{equation}
+where we recognize Eqs.~\ref{eq:selSvPp}~and~\ref{eq:selPp} as the
+last two parts. The first part, $\sel{O = ? | P = p \cap S = ?}$ must
+be computed some other way, by employing heuristics or by using more
+advanced statistical methods.
+
 \paragraph{Subject unbound, object and predicate bound}
 
-In this case, we can use Eq.~\ref{eq:sel} directly. We again note that
+From Eq.~\ref{eq:sel} directly. We again note that
 since the the unbound subject will always have selectivity of

### 2012-09-17 conditional/trunk/selectivity-basic.tex
Subject unbound, object and predicate bound
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@374 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-From Eq.~\ref{eq:sel} directly. We again note that
-since the the unbound subject will always have selectivity of
-unity.
-Eq.~\ref{eq:sel} thus reduces to:
-
-% \begin{equation}\label{eq:selsubjun}
-% \sel{T_{\{?~p~o\}}) = \sel{O | P) \cdot \sel{P) . 
-% \end{equation}
-
-% Since property partitions detail statistics given a certain predicate,
-% they can be used in the conditional case. We can therefore write:
-
-% \begin{equation}\label{eq:selsubjunp}
-% \sel{P = p) = \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|}
-% \end{equation}
-% and
-% \begin{equation}\label{eq:selsubjunop}
-% \sel{O = o|P = p) =
-% \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
-% \end{equation}
-% Thus,
-% \begin{equation}\label{eq:selsubjunres}
-% \sel{T_{\{?~p~o\}}) =
-% \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} ,
-% \end{equation}\todo{WRONG!}
-% where the notation is inspired by SPARQL 1.1 Property Paths,
-% for example the relevant part of a VoID description for the second
-% example in section~\ref{sec:example} would be:
-% \begin{verbatim}
-% </dataset> a void:Dataset ;
-%            void:propertyPartition [
-%                 void:property foaf:firstName;
-%                 void:distinctObjects 2 .
-%            ]
-% \end{verbatim}
-
-% there exists a \rdfterm{void:propertyPartition} predicate that
-% has an object which is again linked to a dataset subset containing the
-% property $p$ as predicate of
+This case is similar to the above, and we can write 
+\begin{equation}\label{eq:selsuobunpb}
+\sel{T_{\{?~p~?\}}} = \sel{O = ? | P = p \cap S = s} \cdot 
+\sel{S = s | P = p} \cdot
+\sel{P = p} . 
+\end{equation}
+We note that we can't find the first two parts in the VoID
+description, but we also note with interest that the first part shows
+that only the marginal distribution of objects is needed to estimate
+this part.
 

### 2012-09-18 conditional/trunk/selectivity-basic.tex
cutnpasto
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@375 43ef1190-90fd-4dd7-a769-d0121f02b80e
cutnpasto
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@376 43ef1190-90fd-4dd7-a769-d0121f02b80e
void:triples -> void:entities
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@377 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 predicate.'', and since a partition is itself a dataset, the same
-statistical properties can be applied, such as \rdfterm{void:triples},
+statistical properties can be applied, such as \rdfterm{void:entities},
 \rdfterm{void:distinctSubjects} and \rdfterm{void:distinctObjects}.
@@
  \sel{P = p} =
-   \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|} ,
+   \frac{\rdfterm{void:propertyPartition/void:entities}}{|d|} ,
 \end{equation}
@@
                 void:property foaf:firstName;
-                void:triples 2 .
+                void:entities 2 .
             ]
@@
  \sel{S = ? | P = p} =
-   \frac{\rdfterm{void:propertyPartition/void:distinctSubjects}}{\rdfterm{void:propertyPartition/void:triples}} ,
+   \frac{\rdfterm{void:propertyPartition/void:distinctSubjects}}{\rdfterm{void:propertyPartition/void:entities}} ,
 \end{equation}
@@
  \sel{O = ? | P = p} =
-   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
+   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:entities}} .
 \end{equation}
@@
  \sel{S = ? | P = a \cap O = c } =
-    \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
+    \frac{\rdfterm{void:classPartition/void:entities}}{|d|} .
 \end{equation}
@@
 This case is similar to the above, and we can write 
-\begin{equation}\label{eq:selsuobunpb}
-\sel{T_{\{?~p~?\}}} = \sel{O = ? | P = p \cap S = s} \cdot 
+\begin{equation}\label{eq:selsuunobpb}
+\sel{T_{\{s~p~?\}}} = \sel{O = ? | P = p \cap S = s} \cdot 
 \sel{S = s | P = p} \cdot

### 2012-09-18 conditional/trunk/selectivity-basic.tex
revert. void:entities -> void:triples
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@378 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2012-09-17 conditional/trunk/selectivity-basic.tex
@@
 \end{equation}
-and similarly where the object is a variable:
+
+In the case where the object is a is a variable, there may be multiple
+objects for each predicate in the partition, so we can only say:
 \begin{equation}\label{eq:selOvPp}
- \sel{O = ? | P = p} =
+ \sel{O = ? | P = p} \ge
    \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
 \end{equation}
+We note that OWL cardinality constraints may constrain this problem
+and that \rdfterm{void:vocabulary} may be used to find the
+corresponding ontology.
 

### 2012-09-18 conditional/trunk/selectivity-basic.tex
add todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@379 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 from the VoID description, and in the next section we shall see how
-they can be exploited most efficiently.
+they can be exploited.
 
@@
 We note that we can't find the first two parts in the VoID
-description, but we also note with interest that the first part shows
+description\todo{2nd part always 1?}, but we also note with interest that the first part shows
 that only the marginal distribution of objects is needed to estimate

### 2012-09-18 conditional/trunk/selectivity-basic.tex
change todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@380 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \rdfterm{void:classPartition} (also defined in section~4.5 in
-\cite{voidnote}):
+\cite{voidnote}) to find the selectivity of the triple pattern:
 \begin{equation}\label{eq:selSvPaOc}
- \sel{S = ? | P = a \cap O = c } =
+ \sel{T_{\{?~a~c\}}} =
     \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
@@
 We note that we can't find the first two parts in the VoID
-description\todo{2nd part always 1?}, but we also note with interest that the first part shows
+description\todo{2nd part smaller than a fraction of distinctSubjects?}, but we also note with interest that the first part shows
 that only the marginal distribution of objects is needed to estimate

### 2012-09-18 conditional/trunk/selectivity-basic.tex
inequalities subsubsection
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@381 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   abandoned. We proceed to extend SPLENDID\cite{splendid} using the
-  new formalism and demonstrate how VoID data descriptions\cite{void} can be
-  further exploited in the conditional regime. Finally, we will
-  explore how selectivity can be determined exactly when a SPARQL
-  query is executed on a fully indexed triplestore, such as
+  new formalism and demonstrate how VoID data descriptions\cite{void}
+  can be further exploited in the conditional regime to optimize join
+  order. Finally, we will explore how selectivity can be determined
+  exactly when a SPARQL query is executed on a fully indexed
+  triplestore, such as
   Hexastore\cite{Weiss:2008:HSI:1453856.1453965}. Finally, we will
@@
 the commutative property of intersections, and some parts of these
-equations is possible to compute from VoID descriptions. We have
-already noted the most trivial case, for variables we have:
+equations is possible to compute from VoID descriptions.
+
+\subsubsection{Equalities}
+
+We have already noted the most trivial case, for variables we have:
 
@@
 
-In the case where the object is a is a variable, there may be multiple
-objects for each predicate in the partition, so we can only say:
-\begin{equation}\label{eq:selOvPp}
- \sel{O = ? | P = p} \ge
-   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
-\end{equation}
-We note that OWL cardinality constraints may constrain this problem
-and that \rdfterm{void:vocabulary} may be used to find the
-corresponding ontology.
-
 We also have in the special case of a given class, i.e. predicate
@@
 they can be exploited.
+
+\subsubsection{Inequalities}
+
+As noted in the abstract, the goal is to establish join order for
+optimization. For this to be successful, it is not critical to be able
+to estimate the selectivity in all cases. It may also suffice to
+establish that the selectivity of one triple pattern is greater or
+smaller than another, they can then be ordered.
+
+First, in the case where the object is a variable, there may be multiple
+objects for each predicate in the partition, so we can only say:
+\begin{equation}\label{eq:selOvPp}
+ \sel{O = ? | P = p} \ge
+   \frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
+\end{equation}
+We note that OWL cardinality constraints may constrain this problem
+and that \rdfterm{void:vocabulary} may be used to find the
+corresponding ontology.
 

### 2012-09-18 conditional/trunk/selectivity-basic.tex
section for class case
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@382 43ef1190-90fd-4dd7-a769-d0121f02b80e
mention bad spec on classpartitions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@383 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We also have in the special case of a given class, i.e. predicate
-\rdfterm{rdf:type} or $P = a$ and $O = c$, we can use
-\rdfterm{void:classPartition} (also defined in section~4.5 in
-\cite{voidnote}) to find the selectivity of the triple pattern:
-\begin{equation}\label{eq:selSvPaOc}
- \sel{T_{\{?~a~c\}}} =
-    \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
-\end{equation}
-
 The above are the parts of the equations that can be computed directly
@@
 this part.
+
+\paragraph{Subject unbound, class bound}
+
+The selectivity can be found directly in the case of a given class,
+i.e. predicate \rdfterm{rdf:type} or $P = a$ and $O = c$, we can use
+\rdfterm{void:classPartition} (also defined in section~4.5 in
+\cite{voidnote}) to find the selectivity of the triple
+pattern\footnote{The specification is unclear at this point, as it
+ mentions the concept of entities, in which case
+ \rdfterm{void:entities} might be used rather than
+ \rdfterm{void:triples}, but we believe that this is the correct usage}:
+\begin{equation}\label{eq:selSvPaOc}
+ \sel{T_{\{?~a~c\}}} =
+    \frac{\rdfterm{void:classPartition/void:triples}}{|d|} .
+\end{equation}
 

### 2012-09-18 conditional/trunk/selectivity-basic.tex
distribution void additions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@384 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Interesting distributions}
+
+Uniform distributions have no impact on join order\todo{argue}.
+\todo{Use parameterized stats for datatype properties?}
+
+\subsection{Suggested additions to the VoID vocabulary}
+
+Allthough a careful estimate of the selectivity in situations where we
+have e.g. $\sel{O = ? | P = p \cap S = ?}$ would require a two
+dimensional distribution function, the most significant impacts on
+join order is when the distribution is highly skewed\todo{argue}. A
+hint to the optimizer in such cases could be achieved by the addition
+of properties like \rdfterm{void:mostFrequentObjects},
+\rdfterm{void:mostFrequentSubjects},
+\rdfterm{void:leastFrequentObjects} and 
+\rdfterm{void:leastFrequentSubjects} to property
+partitions.\todo{mention the dc:language case in Sublima?}
+
 \section{Exact selectivity for fully indexed stores}

### 2012-09-19 conditional/trunk/selectivity-basic.tex
minor updates
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@385 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 where the $|$ symbol is read as ``given'', e.g. ``given a predicate
-\rdfterm{rdf:type} what is the probability that the object is \rdfterm{foaf:Person}?''
+\rdfterm{rdf:type} what is the probability that the object is
+\rdfterm{foaf:Person}?'' This is known as the ``chain rule''.
 
@@
 
-Uniform distributions have no impact on join order\todo{argue}.
+Uniform distributions have little interesting impact on join
+order\todo{argue}. \todo{fallback estimator: distinctSubjects/triples, etc?}
 \todo{Use parameterized stats for datatype properties?}
@@
 have e.g. $\sel{O = ? | P = p \cap S = ?}$ would require a two
-dimensional distribution function, the most significant impacts on
+dimensional joint distribution function, the most significant impacts on
 join order is when the distribution is highly skewed\todo{argue}. A

### 2012-09-19 conditional/trunk/selectivity-basic.tex
add BGP section
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@386 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Extension to Basic Graph Patterns}
+
+The statistics of Basic Graph Patterns, especially when involving
+filters, are even more complex, and is likely to involve the use of
+Bayesian Networks, which is beyond the scope of this paper. However,
+as noted by \cite{splendid}, the intuition is that for some conjuctive
+queries, the order of triples are not influenced by the conditionality
+of separate triples. We believe the intuition is correct, but that in
+detail, there are flaws in the argument of \cite{splendid}\todo{show
+  counterexample}\todo{improve our own argument}.
+
 \subsection{Suggested additions to the VoID vocabulary}

### 2012-10-19 dissertation/index.tex
Add stub on dissertation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@392 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass[a4paper, 12pt, draft]{scrartcl}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{times}
\usepackage{listings}
\usepackage{supertabular}
\usepackage{cite}
\include{notasjon}
\include{helpers}
%\usepackage[pdftex,linktocpage,ps2pdf,dvips]{hyperref}

\begin{document}
\title{Dissertation}
\author{Kjetil Kjernsmo}
%\institute{Department of Informatics,
%Postboks 1080 Blindern,
%0316 Oslo, Norway}

%\email{kjekje@ifi.uio.no}

\maketitle

\input{introsota}

\bibliography{selectivity}
\bibliographystyle{plainnat}

\end{document}

### 2012-10-19 dissertation/introsota.tex
Add stub on dissertation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@392 43ef1190-90fd-4dd7-a769-d0121f02b80e

\section{State of the art}

The state of the art.

### 2012-10-19 dissertation/introsota.tex
Fix citations; new label on PRM selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@393 43ef1190-90fd-4dd7-a769-d0121f02b80e
more on cond sel
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@394 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The state of the art.
+The following work serve as a starting point for the present work.
+
+\subsection{Conditional selectivity estimation in triple stores}
+
+Selectivity estimation has been a standard technique for optimization
+of query responses to databases for a long time. However, conditional
+statistics is rarely employed, as most relational systems assume both
+\emph{attribute value independence} and \emph{join uniformity}. For
+RDF databases the use of detailed statistics to relax these
+assumptions is a recent development.
+
+\cite{Lv:2009:SEC:1685170.1685590} first considered correlated
+properties for SPARQL query optimization. However, they only
+considered TBox statements, and so, no detailed statistics of ABox
+data was compiled.
+
+In SPLENDID\cite{splendid} the authors showed how statistics exposed
+using the VoID data description vocabulary\cite{void} could be
+exploited for query optimization. The authors exploited the fact that
+VoID descriptions may expose the number of distinct subjects or
+objects for a given predicate, and therefore partially used
+conditional statistics. However, they assumed that the distribution of
+subjects and objects was uniform.
+
+\subsection{Statistical Relational Learning}
+
+The idea to use Statistical Relational Learning techniques for
+selectivity estimation first originated in \cite{selectivityPRM}.

### 2012-10-23 dissertation/introproblem.tex
Add problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@395 43ef1190-90fd-4dd7-a769-d0121f02b80e

\section{Problem description}

The main topic of the present dissertation is SPARQL
Federation\todo{further discuss different interesting directions with
  SPARQL federation}.

### 2012-10-23 dissertation/introproblem.tex
sublima ref
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@396 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   SPARQL federation}.
+
+In general, heterogenous data often causes optimization
+problems. Since heterogenous data is important in Semantic Web
+applications, these problems become important to solve in a Semantic
+Web context.
+
+One problem can be motivated from previous implementation experience:
+In the SUBLIMA project, described in \ref{sublima}, entries were
+typically described has having a language, and in one deployment the
+vast majority were written in Norwegian, but a few in English, Danish
+or Swedish. With the SPARQL engines of the day, this heterogenity
+caused significant performance issues: If the language was joined
+first, and it was Norwegian, the query would be slow, whereas if it
+was joined last, it would be orders of magnitude faster. For English,
+Danish or Swedish, the situation was the opposite. The lesson was
+learned and fixed on an \textit{ad hoc} basis, the most restrictive
+term must be joined first. This issue will be referred to as the
+\langcase in this text as we shall return to it.
+
+In modern centralized SPARQL databases, this is less of a problem, but
+the problem rearises for distributed databases.

### 2012-10-24 dissertation/introproblem.tex
describe stats problem in fed cases
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@397 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In modern centralized SPARQL databases, this is less of a problem, but
-the problem rearises for distributed databases.
+the problem re-emerges for distributed databases. In the case where
+the \fedeng has no knowledge of the data on the \indendp, it cannot
+optimize the join order, as such details are unavailable. For example,
+in the \langcase, if the language annotations were on a different
+endpoint than other data, the effects would be disasterous, as if the
+Norwegian language resources was first retrieved from one endpoint,
+the query would retrieve a large result set, and then return the
+result set over the network to the \fedeng, with the high
+communication costs incurred, and finally the query passed on to the
+next federation member with the bound variables would be large, which
+would incur a large communication cost as well as the cost of parsing
+the query.
+
+Sparse statistics is not sufficient to solve the \langcase
+generally. We require knowledge not only of the statistics of the
+predicate, but also on the object for a given predicate.
+
+This is the main motivation of the present dissertation. To investigate
+possible solutions, we explore different directions in statistics,
+ranging from simple extensions to the VoID vocabulary\ref{void} to
+advanced \SRL.

### 2012-10-24 dissertation/introproblem.tex
fixups
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@400 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 One problem can be motivated from previous implementation experience:
-In the SUBLIMA project, described in \ref{sublima}, entries were
+In the SUBLIMA project, described in \cite{sublima}, entries were
 typically described has having a language, and in one deployment the
@@
 term must be joined first. This issue will be referred to as the
-\langcase in this text as we shall return to it.
+\langcase{} in this text as we shall return to it.
 
@@
 the problem re-emerges for distributed databases. In the case where
-the \fedeng has no knowledge of the data on the \indendp, it cannot
+the \fedeng{} has no knowledge of the data on the \indendp, it cannot
 optimize the join order, as such details are unavailable. For example,
@@
 
-Sparse statistics is not sufficient to solve the \langcase
+Sparse statistics is not sufficient to solve the \langcase{}
 generally. We require knowledge not only of the statistics of the
@@
 possible solutions, we explore different directions in statistics,
-ranging from simple extensions to the VoID vocabulary\ref{void} to
+ranging from simple extensions to the VoID vocabulary\cite{void} to
 advanced \SRL.

### 2012-10-24 dissertation/introsota.tex
Add problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@395 43ef1190-90fd-4dd7-a769-d0121f02b80e
more on SRL
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@401 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 statistics is rarely employed, as most relational systems assume both
-\emph{attribute value independence} and \emph{join uniformity}. For
-RDF databases the use of detailed statistics to relax these
-assumptions is a recent development.
+\emph{attribute value independence} and \emph{join
+  uniformity}\todo{explain what this means}. These are the assumptions
+that made the \langcase{} problem so significant in the SUBLIMA
+project, as both assumptions were violated by the mere fact that only
+languages can be values (i.e. be objects) of the language attribute
+(predicate) and the highly skewed value distribution for
+languages. Therefore, these assumptions must be relaxed. For RDF
+databases the use of detailed statistics to relax these assumptions is
+a recent development and the field is very open to further progress.
 
@@
 considered TBox statements, and so, no detailed statistics of ABox
-data was compiled.
+data was compiled. While this does prove valuable in some cases, it
+would do nothing to alleviate the \langcase{} problem.
 
@@
 The idea to use Statistical Relational Learning techniques for
-selectivity estimation first originated in \cite{selectivityPRM}.
+selectivity estimation was first proposed in \cite{selectivityPRM}.
+
+The procedure to do this for RDF databases is also quite clear: First,
+a network must be learned from the data, then exposed in a SPARQL
+endpoint service description so that it could be retrieved by the
+\fedeng{}, where the learned network would be looked up to the
+estimate the selectivity of a given graph pattern.
+
+Work has been published on different parts of this problem:
+\cite{Lin:2011:LRB:2063016.2063042} demonstrated learning relational
+bayesian classifiers from RDF data, while [?] \todo{lost reference}
+showed how a probabilistic  relational model could be expressed with
+OWL. Finally,  \cite{selectivityPRM} could be adapted to RDF according
+to the principles in these two papers.

### 2012-10-25 dissertation/introsota.tex
cyclicity trouble
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@404 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The idea to use Statistical Relational Learning techniques for
-selectivity estimation was first proposed in \cite{selectivityPRM}.
+\SRL, see \cite{srlbook} for a comprehensive overview, is a relatively
+new field, but encompasses well established techniques such as
+Bayesian Networks and recent developments, such as Relational Markov
+Networks.
 
-The procedure to do this for RDF databases is also quite clear: First,
+The idea to use \SRL-techniques for selectivity estimation was first
+proposed in \cite{selectivityPRM}.
+
+The procedure to use \SRL{} for RDF databases is also quite clear: First,
 a network must be learned from the data, then exposed in a SPARQL
@@
 showed how a probabilistic  relational model could be expressed with
-OWL. Finally,  \cite{selectivityPRM} could be adapted to RDF according
-to the principles in these two papers.
+OWL. Finally,  \cite{selectivityPRM} could be adapted to RDF following
+the principles in these two papers.
+
+There are, however, many problems: \SRL{} consists of many different
+but related methodologies, and so, proving compatibility between
+different methods and applicability to the problem at hand is
+important and not necessarily trivial.
+
+All methods used above is based on directed \emph{acyclic} graphs, but
+an RDF graph may be cyclic, which may present significant
+difficulties. \cite{Lin:2011:LRB:2063016.2063042} addresses this
+problem by constraining themselves to an ``Attribute Graph'', which
+they define to match a Basic Graph Pattern such as
+\begin{verbatim}
+x     p1 ?v1 .
+?v1   p2 ?v2 .
+...
+?vn-1 pn ?vn .
+\end{verbatim}
+(for the formal definition, see
+\cite{Lin:2011:LRB:2063016.2063042}). The resulting graph is a
+directed acyclic graph.
+
+\SRL{} methodologies can be classified in two groups based on whether
+they allow cyclicity or not. Bayesian Networks, Probabilistic
+Relational Models, etc. are acyclic, while e.g. Conditional Random
+Fields and Relational Markov Networks allow cyclic references. It is
+clear that to completely model statistical dependencies in an RDF
+graph, a cyclic model is required, but then a key question becomes
+whether the cyclic nature of RDF is necessary to take into account to
+satisfactorily address the optimization problem.

### 2012-10-25 dissertation/introsota.tex
SPLENDID and langcase
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@405 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 subjects and objects was uniform.
+
+We note that SPLENDID would not solve the \langcase, but since there
+were so few languages, there would be other objects that had a lot
+more distinct values. They would would be joined earlier in this
+specific case and we would be unlikely to see the worst-case
+behavior. However, this would be brittle and we seek a better
+solution.
 
@@
 (for the formal definition, see
-\cite{Lin:2011:LRB:2063016.2063042}). The resulting graph is a
-directed acyclic graph.
+\cite{Lin:2011:LRB:2063016.2063042}). This makes an Attribute Graph a
+Directed Acyclic Graph (DAG).
 

### 2012-10-25 dissertation/tmpsolutions.tex
Discussion of solutions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@407 43ef1190-90fd-4dd7-a769-d0121f02b80e

\section{Possible directions of work}

\emph{This section is a temporary discussion area}

We need to discuss the future strategy for this work. I think the main
discussion point is the tension between elaborately develop theory
that then can be applied to derive publishable results, vs. the
practical time constraints of Ph.D. work. Another issue is that I'd
like to be mindful of the realities of adoption in standards and
adoption of standards.

### 2012-10-25 dissertation/tmpsolutions.tex
further partitioning most frequent objects
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@408 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 adoption of standards.
+
+The SPLENDID paper provides a pretty clear path forward to really
+solve the \langcase: Certain statistics could be amended to a new
+release of VoID for cases where it very clearly would have impact. It
+would be up to the data management software to identify those case
+based on the work I publish.
+
+There are two basic classes of solutions to this problem: One is to
+allow further nesting of dataset partitions. VoID currently allow
+class and property partitions, where the class partitions are the most
+important in SPLENDID, property partitions is the most interesting for
+conditional statistics. Within property partitions, one could imagine
+a nested object partition, and the \langcase{} could be solved
+conclusively by specifying the number of triples containing each
+language. If we did the same thing with subject partitions, it would
+be a very good solution to this and similar problems. However, it
+would be a rather verbose solution.
+
+The second option is based on what the database literature knows as
+most-frequent-values histograms. This is a technique used by
+PostgreSQL. The (over-)simplified version of this is to say in a VoID
+description:
+
+\begin{verbatim}
+_:foo void:propertyPartition 
+  [
+    void:property dct:language ;
+    void:mostFrequentObjects lang:nb .
+].
+\end{verbatim}
+This would declare \rdfterm{lang:nb} as the most frequent object for
+\rdfterm{dct:language}, and this simple example would solve the
+\langcase.
+
+While rare objects might more seldomly occur in queries, they may
+contribute very significantly to the performance if they are joined
+first if they do. Thus, it might be worthwhile to have a property
+\rdfterm{void:leastFrequentOjects} too, as well as corresponding
+properties for subjects.

### 2012-10-25 dissertation/tmpsolutions.tex
headline
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@409 43ef1190-90fd-4dd7-a769-d0121f02b80e
more void solutions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@410 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 adoption of standards.
+
+\subsection{VoID extensions}
 
@@
 properties for subjects.
+
+However, this isn't a histogram, it merely mentioning the objects that
+would most likely be useful. To get a histogram we would need a
+similar \rdfterm{void:frequency} attribute and both this and
+\rdfterm{void:mostFrequentObjects} would have to accept ordered
+lists.
+
+Another option would be to do something like:
+\begin{verbatim}
+_:foo void:propertyPartition 
+  [
+    void:property dct:language ;
+    void:mostFrequentObjects 
+    [
+      void:object lang:nb ;
+      void:frequency "0.012" .
+    ]
+].
+\end{verbatim}
+but this solution would be less general but of similar complexity than
+the nested solution above.

### 2012-10-25 dissertation/tmpsolutions.tex
advocating void paper fully indexed stores
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@411 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the nested solution above.
+
+I find the prospect of writing a first paper to explore these different
+extensions to VoID very appealing. I would need to put down some
+substantial work on notation, but not intending an very elaborate
+theory, which is something that requires \SRL. The paper could discuss
+the different approaches above, their merits and drawbacks. After
+publishing, I could work with the W3C Semantic Web Interest Group to
+get one of the solutions into VoID.
+
+\subsection{Fully indexed stores}
+
+Some time ago, I wrote a partial paper about how fully indexed stores
+like Hexastore \cite{hexastore} could easily find the cardinality of
+any given triple pattern. It would be interesting to publish that soon
+in some form.

### 2012-10-25 dissertation/tmpsolutions.tex
srl discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@412 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{\SRL}
+
+Using \SRL{} in some form, whether it is in the familiar end of
+Bayesian Networks or in the relatively adventerous end of methods that
+support cycles, is much more elaborate than the above strategy. I'm
+quite convinced that I need to publish a paper on this topic as well,
+and arguably, if I should focus at the dissertation as a whole rather
+than the individual parts, it makes sense to develop this theory
+first, and then derive the VoID extensions from that.
+
+There's a whole zoo of methods in \cite{srlbook}, and at this point,
+just choosing a methodology is difficult. On page~191 in this book,
+there is a hypertext with typed links-example, which is very close to
+the RDF model, so it is clear that there are suitable examples to work
+from. Also, I have a feeling that a general methodology to learn a
+Conditional Random Field from a cyclic RDF graph would be a very
+significant addition to the literature, with broad applicability way
+beyond the selectivity optimization use case.
+
+However, there seems that nobody in Oslo know the CRF theory, it seems
+to have been an active field of study for only about 7~years, and we don't
+know anyone who works on this, so it seems to be very ambitious to
+embark on.
+
+In the acyclic case, as mentioned in section~\ref{sotasrl}, there
+seems to be a pretty clear avenue for progress, and substantial
+progress could be done in exploring acyclic graph patterns that are
+common and how to expose them compactly.
+
 \subsection{Fully indexed stores}

### 2012-10-25 dissertation/tmpsolutions.tex
conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@413 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 in some form.
+
+\subsection{Conclusion}
+
+These are my current opinions:
+
+At present, I feel the best way to progress is to publish a paper on
+VoID extensions as well as fully indexed stores in a conditional
+regime. I should do some work to get the notation and theory well
+founded, but defer \SRL{} for later.
+
+The 3.~semester evaluation committee also critisized the lack of a
+paper, and said that I should produce a paper, at least a draft before
+the end of the year. If I embark on a study of \SRL{}, this would not
+be possible.
+
+The study of \SRL{} is very interesting, so I should look into that
+once the first paper has been published, but possibly seek partners or
+an external resource.
+
+Amongst the below references, the most important are
+\cite{selectivityPRM}, \cite{Lin:2011:LRB:2063016.2063042} and
+\cite{splendid}.

### 2012-10-25 dissertation/tmpsolutions.tex
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@414 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{verbatim}
-_:foo void:propertyPartition 
+</foo> void:propertyPartition 
   [
@@
 Some time ago, I wrote a partial paper about how fully indexed stores
-like Hexastore \cite{hexastore} could easily find the cardinality of
+like Hexastore \cite{Weiss:2008:HSI:1453856.1453965} could easily find the cardinality of
 any given triple pattern. It would be interesting to publish that soon
@@
 Amongst the below references, the most important are
-\cite{selectivityPRM}, \cite{Lin:2011:LRB:2063016.2063042} and
-\cite{splendid}.
+\cite{selectivityPRM}, \cite{splendid} and \cite{Lin:2011:LRB:2063016.2063042}.

### 2012-10-25 dissertation/tmpsolutions.tex
just the object lang
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@416 43ef1190-90fd-4dd7-a769-d0121f02b80e
w3c
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@417 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 This would declare \rdfterm{lang:nb} as the most frequent object for
-\rdfterm{dct:language}, and this simple example would solve the
+\rdfterm{dct:language}, and so hint to the \fedeng that it should be
+treated as having low selectivity. This simple example would solve the
 \langcase.
@@
 publishing, I could work with the W3C Semantic Web Interest Group to
-get one of the solutions into VoID.
+get one of the solutions into VoID (when we join the W3C).
 

### 2013-04-24 exp1/run.sh
Add run shell script
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@455 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+killall 4s-backend 
+killall 4s-httpd
+cd 4store-unmodified/
+bin/4s-backend largeunmod
+bin/4s-httpd -s -1 -p 8012 largeunmod
+bin/4s-backend smallunmod
+bin/4s-httpd -s -1 -p 8011 smallunmod
+cd ../4store-sleep/
+bin/4s-backend largesleep
+bin/4s-backend smallsleep
+bin/4s-httpd -s -1 -p 8022 largesleep
+bin/4s-httpd -s -1 -p 8021 smallsleep
+ps ux | grep 4s
+exit 0

### 2013-04-25 exp1/generate-design.R
Add completely random design
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@458 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -11,3 +11,6 @@ sparqltest.sans.machine.design <- cross.design(
                                   fac.design(nlevels = 2, nfactors = 5,
                                              factor.names = c("BGPComp", "Lang", "Range", "Union", "Optional")),
                                   randomize = F)
+
+sparqltest.random.design <- fac.design(nlevels = 2, nfactors = 8,
+                                             factor.names = c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"))

### 2013-04-26 exp1/run.sh
Check if the port responds on startup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@459 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -13,4 +13,14 @@ bin/4s-backend smallsleep
 bin/4s-httpd -s -1 -p 8022 largesleep
 bin/4s-httpd -s -1 -p 8021 smallsleep
 ps ux | grep 4s
+
+echo 8012
+wget -O /dev/null http://localhost:8012/status/
+echo 8011
+wget -O /dev/null http://localhost:8011/status/
+echo 8022
+wget -O /dev/null http://localhost:8022/status/
+echo 8021
+wget -O /dev/null http://localhost:8021/status/
+
 exit 0

### 2013-04-26 exp1/iswc2013.tex
Start paper Add abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@460 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass{llncs}
%\documentclass{article}
\usepackage{cite}

\title{Introducing Statistical Design of Experiments to SPARQL
  Endpoint Evaluation}
\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
\author{Kjetil Kjernsmo\inst{1}}
\institute{Department of Informatics,
Postboks 1080 Blindern,
0316 Oslo, Norway
\email{kjekje@ifi.uio.no}}

\begin{document}

\maketitle

\begin{abstract}
This paper argues that the common practice of benchmarking is
inadequate as a scientific evaluation methodology. It further attempts
to introduce the empirical tradition of the physical sciences by using
techniques from Statistical Design of Experiments applied to the
example of SPARQL endpoint performance evaluation. It does so by
studying full as well as fractional factorial experiments designed to
evaluate an assertion that some change introduced in a system has
improved performance. This paper does not present a finished
experimental design, rather its main focus is didactical, to shift the
focus of the community away from benchmarking towards higher
scientific rigor.
\end{abstract}

\section{Introduction}

\section{Related work}

\section{Conclusion}

\section*{Acknowledgements}

\end{document}

### 2013-04-26 exp1/iswc2013.tex
Related work first stab
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@462 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+A literature survey as not revealed any direct prior art, however, the
+general approach has been well established, not only in statistics. A
+relatively well cited textbook is \cite{citeulike:5190414} but we have
+not found the parts discussed in the present paper to be widely
+adopted. Recently, an comprehensive text on experimental methods have
+been published in
+\cite{Springer-2010-Experimental-Methods-for-the-Analysis-of-Optimization-Algorithms}. 
+We have chosen to turn to statistical standard texts
+\cite{wu2009experiments} for this study.
+
+The problematic sides of the benchmarking have been noted in several
+papers, notably \cite{Duan:2011:AOC:1989323.1989340} and
+\cite{conf/semweb/MontoyaVCRA12}. We also acknowledge that great
+progress has been made to improve benchmarks, in particular, we are
+indebted to \cite{mxro:Morsey2011DBpedia}.
+
+We believe that automatic creation of benchmark queries, as
+pioneered by \cite{goerlitz2012splodge} is a critical ingredient for
+the application of DoE to be successful as a methodology.
+
+Finally, we acknowledge the efforts of the Linked Data Benchmark
+Council and the SEALS project. However, they appear to focus on
+industrial benchmarking rather than scientific evaluations.
+
 \section{Conclusion}
@@
 
+\bibliographystyle{plain}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{jbact}
+%\bibliographystyle{splncs03}
+\bibliography{selectivity,federation,benchmarks,optimization,experimentsperformance}
+
 \end{document}

### 2013-04-26 exp1/run.R
run a while loop around the host
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@467 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -45,7 +45,7 @@ experiment <- function(run, files) {
   querypattern <- apply(thisexperiment, 1, compose, type="query")
   ret <- apply(thisexperiment, 1, compose, type="endpointhost")
   host <- ret[!is.na(ret)]
-  if(length(host) == 0) {
+  while((length(host) == 0 || nchar(host) == 0)) {
     cat("No hostname found in experiment setup. ")
     host <- ask("Enter hostname: ")
   }

### 2013-04-27 exp1/run.R
just hardcode host for now
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@468 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -45,9 +45,8 @@ experiment <- function(run, files) {
   querypattern <- apply(thisexperiment, 1, compose, type="query")
   ret <- apply(thisexperiment, 1, compose, type="endpointhost")
   host <- ret[!is.na(ret)]
-  while((length(host) == 0 || nchar(host) == 0)) {
-    cat("No hostname found in experiment setup. ")
-    host <- ask("Enter hostname: ")
+  if(length(host) == 0) {
+    host <- "10.72.1.180"
   }
   if(length(host) > 1) {
     stop("We need exactly one host")

### 2013-04-27 exp1/run.R
print some status
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@469 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -60,6 +60,7 @@ experiment <- function(run, files) {
     cat("Previous Endpoint URL", endpointurl, "returned error\n")
     ask("Press <RETURN> to test again")
   }
+  cat(date(), " Sending ", run, " to ", endpointurl, "\n") 
   timeQuery(endpointurl, query)$endpoint
 }
 

### 2013-05-02 exp1/run.sh
switch sleep and unmod for more intuitive analysis
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@483 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -4,14 +4,14 @@ killall 4s-backend
 killall 4s-httpd
 cd 4store-unmodified/
 bin/4s-backend largeunmod
-bin/4s-httpd -s -1 -p 8012 largeunmod
+bin/4s-httpd -s -1 -p 8022 largeunmod
 bin/4s-backend smallunmod
-bin/4s-httpd -s -1 -p 8011 smallunmod
+bin/4s-httpd -s -1 -p 8021 smallunmod
 cd ../4store-sleep/
 bin/4s-backend largesleep
 bin/4s-backend smallsleep
-bin/4s-httpd -s -1 -p 8022 largesleep
-bin/4s-httpd -s -1 -p 8021 smallsleep
+bin/4s-httpd -s -1 -p 8012 largesleep
+bin/4s-httpd -s -1 -p 8011 smallsleep
 ps ux | grep 4s
 
 echo 8012

### 2013-05-02 exp1/run.sh
use curl for status check
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@484 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -14,13 +14,12 @@ bin/4s-httpd -s -1 -p 8012 largesleep
 bin/4s-httpd -s -1 -p 8011 smallsleep
 ps ux | grep 4s
 
-echo 8012
-wget -O /dev/null http://localhost:8012/status/
-echo 8011
-wget -O /dev/null http://localhost:8011/status/
-echo 8022
-wget -O /dev/null http://localhost:8022/status/
-echo 8021
-wget -O /dev/null http://localhost:8021/status/
+curl -s -o /dev/null -w "%{url_effective} %{http_code}\n" http://localhost:8011/status/
+
+curl -s -o /dev/null -w "%{url_effective} %{http_code}\n" http://localhost:8012/status/
+
+curl -s -o /dev/null -w "%{url_effective} %{http_code}\n" http://localhost:8021/status/
+
+curl -s -o /dev/null -w "%{url_effective} %{http_code}\n" http://localhost:8022/status/
 
 exit 0

### 2013-05-02 exp1/analysis.R
a start of some analysis code
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@485 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,7 @@
+coef.sorted <- function(model) {
+  sort(abs(coef(model)))
+}
+
+implementation.coef <- function(model) {
+  coef(model)[grep("Implement", names(coef(model)))]
+}

### 2013-05-02 exp1/analysis.R
Analysis with two-sample t-test
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@487 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -5,3 +5,18 @@ coef.sorted <- function(model) {
 implementation.coef <- function(model) {
   coef(model)[grep("Implement", names(coef(model)))]
 }
+
+fulllm <- function(results) {
+  lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
+}
+
+robust.single <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), ...) {
+  fm <- as.formula(paste("experiments ~", control, " * ", paste(noise, collapse=" * ")));
+  allmeans <- aggregate(fm, data=model, mean)
+  t.test(
+         allmeans[allmeans[[control]] == 2,"experiments"],
+         allmeans[allmeans[[control]] == 1,"experiments"],
+         alternative="less",
+         ...)
+}
+  

### 2013-05-02 exp1/analysis.R
refactor robust analysis
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@488 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,13 +10,13 @@ fulllm <- function(results) {
   lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
 }
 
-robust.single <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), ...) {
+robust <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), pairwise=TRUE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(noise, collapse=" * ")));
   allmeans <- aggregate(fm, data=model, mean)
-  t.test(
-         allmeans[allmeans[[control]] == 2,"experiments"],
-         allmeans[allmeans[[control]] == 1,"experiments"],
-         alternative="less",
-         ...)
+  oldimplement <- allmeans[allmeans[[control]] == 1,"experiments"]
+  newimplement <- allmeans[allmeans[[control]] == 2,"experiments"]
+  if(pairwise) {
+    t.test(newimplement, oldimplement, alternative="less", ...)
+  }# else {
 }
   

### 2013-05-02 exp1/analysis.R
trivial logic bug
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@489 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,13 +10,14 @@ fulllm <- function(results) {
   lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
 }
 
-robust <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), pairwise=TRUE, ...) {
+robust <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), pairwise=FALSE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(noise, collapse=" * ")));
   allmeans <- aggregate(fm, data=model, mean)
   oldimplement <- allmeans[allmeans[[control]] == 1,"experiments"]
   newimplement <- allmeans[allmeans[[control]] == 2,"experiments"]
   if(pairwise) {
+  } else {
     t.test(newimplement, oldimplement, alternative="less", ...)
-  }# else {
+  }
 }
   

### 2013-05-03 exp1/analysis.R
Check
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@490 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -16,6 +16,7 @@ robust <- function(model, control = "Implement", noise = c("TripleC", "Lang", "U
   oldimplement <- allmeans[allmeans[[control]] == 1,"experiments"]
   newimplement <- allmeans[allmeans[[control]] == 2,"experiments"]
   if(pairwise) {
+    browser()
   } else {
     t.test(newimplement, oldimplement, alternative="less", ...)
   }

### 2013-05-03 exp1/analysis.R
Put the inactive to iterate?
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@491 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,15 +10,10 @@ fulllm <- function(results) {
   lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
 }
 
-robust <- function(model, control = "Implement", noise = c("TripleC", "Lang", "Union", "Optional"), pairwise=FALSE, ...) {
-  fm <- as.formula(paste("experiments ~", control, " * ", paste(noise, collapse=" * ")));
+robust <- function(model, control = "Implement", inactive = c("Machine", "BGPComp", "Range"), pairwise=FALSE, ...) {
+  fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")));
   allmeans <- aggregate(fm, data=model, mean)
-  oldimplement <- allmeans[allmeans[[control]] == 1,"experiments"]
-  newimplement <- allmeans[allmeans[[control]] == 2,"experiments"]
-  if(pairwise) {
-    browser()
-  } else {
-    t.test(newimplement, oldimplement, alternative="less", ...)
-  }
+  allmeans
+
 }
   

### 2013-05-04 exp1/generate-design.R
Add the fractional 32 run design
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@492 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -14,3 +14,5 @@ sparqltest.sans.machine.design <- cross.design(
 
 sparqltest.random.design <- fac.design(nlevels = 2, nfactors = 8,
                                              factor.names = c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"))
+
+frac32.design <- FrF2(nruns=32, factor.names=c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"), estimable=formula("~Implement + TripleC+Machine+BGPComp+Lang+Range+Union+Optional + Implement:(TripleC+Machine+BGPComp+Lang+Range+Union+Optional)"), default.levels=c(1,2))

### 2013-05-04 exp1/run.sh
Support running huge too
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@494 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,16 +1,24 @@
 #!/bin/bash
 
+if [ $1 ] 
+  then
+	largename=$1
+  else
+	largename="large"
+fi
+
+
 killall 4s-backend 
 killall 4s-httpd
 cd 4store-unmodified/
-bin/4s-backend largeunmod
-bin/4s-httpd -s -1 -p 8022 largeunmod
+bin/4s-backend $largename"unmod"
+bin/4s-httpd -s -1 -p 8022 $largename"unmod"
 bin/4s-backend smallunmod
 bin/4s-httpd -s -1 -p 8021 smallunmod
 cd ../4store-sleep/
-bin/4s-backend largesleep
+bin/4s-backend $largename"sleep"
 bin/4s-backend smallsleep
-bin/4s-httpd -s -1 -p 8012 largesleep
+bin/4s-httpd -s -1 -p 8012 $largename"sleep"
 bin/4s-httpd -s -1 -p 8011 smallsleep
 ps ux | grep 4s
 

### 2013-05-04 exp1/iswc2013.tex
A bit of tex-reorg
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@464 43ef1190-90fd-4dd7-a769-d0121f02b80e
describe factors in paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@496 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-\documentclass{llncs}
-%\documentclass{article}
+%\documentclass{llncs}
+\documentclass{article}
 \usepackage{cite}
+\usepackage{verbatim}
 
@@
   Endpoint Evaluation}
-\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
 \author{Kjetil Kjernsmo\inst{1}}
-\institute{Department of Informatics,
-Postboks 1080 Blindern,
-0316 Oslo, Norway
-\email{kjekje@ifi.uio.no}}
+%\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
+%\institute{Department of Informatics,
+%Postboks 1080 Blindern,
+%0316 Oslo, Norway
+%\email{kjekje@ifi.uio.no}}
+
+%\maketitle
 
 \begin{document}
-
-\maketitle
 
@@
 papers, notably \cite{Duan:2011:AOC:1989323.1989340} and
-\cite{conf/semweb/MontoyaVCRA12}. We also acknowledge that great
+\cite{MontoyaVCRA12}. We also acknowledge that great
 progress has been made to improve benchmarks, in particular, we are
@@
 industrial benchmarking rather than scientific evaluations.
+
+\section{Experimental setup}
+
+As the focus of this paper is didactical, the actual measurements are
+much simpler than what has been used in benchmarking.
+
+To only deal with the relatively straightforward aspects of the DoE
+formalism, we have chosen to constrain the experiment strictly to 2
+levels. We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
+dataset, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
+where one machine runs GNU/Linux Debian Squeeze, has 16 GB RAM, an
+Intel Core2 Duo E8500 CPU and two Parallell-ATA disks in a RAID-1
+configuration. The other machine runs Debian Wheezy, has 8 GB RAM, an
+Intel Core i7-3520M CPU and a single SSD on SATA-III.
+
+Then, we test some language features: 
+``BGPComp'', which is a Basic Graph Pattern of varying complexity. The
+first level is:
+\verbatiminput{experiment/BGPComp-1}
+and the second level is 
+\verbatiminput{experiment/BGPComp-1}
+
+The following factors tests the absence (at level 1) or presence (at level 2) of the following clauses:
+The ``Lang'' factor is this FILTER:
+\verbatiminput{experiment/Lang-2}
+and similarly,  ``Range'' is this FILTER:
+\verbatiminput{experiment/Range-2}
+``Union'' tests the following:
+\verbatiminput{experiment/Union-2}
+and ``Optional''
+\verbatiminput{experiment/Optional-2}
+
+Finally, ``Implement'' is the implementation undergoing
+evaluation. Level 1 is running the new implementation, whereas Level 2
+is the old implementation. The null hypothesis $H_0$ is that the old
+implementation is as good as the new, and the alternative hypothesis
+$H_1$ is that the new implementation has overall improved the
+performance of the system.
+
+We have used the dataset of \cite{mxro:Morsey2011DBpedia}, where
 

### 2013-05-04 exp1/iswc2013.tex
more on experimental setup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@497 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{verbatim}
+\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 As the focus of this paper is didactical, the actual measurements are
-much simpler than what has been used in benchmarking.
+much simpler than what has been used in benchmarking. We have used the
+dataset of \cite{mxro:Morsey2011DBpedia}, but only consider the
+smallest of their datasets (that we found to me more than 15
+MTriples). Moreover, we have taken subsets of that dataset by using
+the last 1 or 2 MTriples their file.
 
@@
 
+These fragments has been carefully designed for illustrative utility,
+as well as suitable selectivity. When the experiment is run, they are
+combined to yield a query, which is then sent to a SPARQL
+endpoint. The SPARQL endpoint is itself set up using 4store\todo{ref}
+version 1.1.5. The ``Machine'' factor specifies a hostname, whereas
+``TripleC'' and ``Implement'' specifies a port number. On each host,
+four 4store instances runs independently on different
+ports. \todo{possible problem?}
+
 Finally, ``Implement'' is the implementation undergoing
@@
 
-We have used the dataset of \cite{mxro:Morsey2011DBpedia}, where
+However, as a real optimization is beyond the scope of this paper, we
+have instead created a simulation of an optimization. This has also
+the advantage that we can \textit{a priori} understand the effect of
+the changes. In a real-world case, there is often the case that an
+optimization has negative side-effects, and we would need to simulate
+both the optimization and the side-effects. To do that, we degraded
+the performance of a the 4store SPARQL implementation by inserting
+\texttt{sleep} statements. Specifically we inserted the C statement
+\texttt{sleep(2)} on line~920 in \texttt{src/frontend/query.c} on
+level~1, to simulate the optimization. This has the effect of delaying
+execution for 2~seconds for every block for all kinds of joins. On
+level 2, we inserted \texttt{usleep(2000)} on line 987 in
+\texttt{src/frontend/filter.c} to simulate the negative
+side-effect. This delays execution for 2 milliseconds every time the
+\texttt{langMatches} SPARQL function is called.
 

### 2013-05-04 exp1/iswc2013.tex
more on setup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@499 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Experimental setup}
+\section{Experimens}
+
+\subsection{Setup}
 
@@
 
+The experimentation is implemented in R\todo{ref}, which is a free
+software environment for statistical computing and graphics. Necessary
+tools for DoE has been implemented by the R community in packages
+called DoE.base, BsMD and FrF2. The experiments are run on a third
+computer in a 1 gigabit LAN with the two experiment hosts.  As
+response variable, we have chosen to use the time from the namespace
+server lookup to the data has been transferred. This choice is
+somewhat arbitrary, many other response variables could be chosen, and
+indeed, future work should explore multi-variable responses, but for
+simplicity, we think this reflects the total amount of work done by
+the SPARQL endpoint well. For the measurements, we have chosen to use
+the RCurl package, which is a thin wrapper around the curl
+library. Curl has  good facilities for timing requests, and so
+we rely on its measurements.
+
+\subsection{Full factorial experiment}
+
+\subsection{Fractional factorial experiment}
+
+\subsection{Fractional factorial experiment with more data}
+
 \section{Conclusion}

### 2013-05-05 exp1/generate-design.R
Add res 4 fractional
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@501 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -16,3 +16,5 @@ sparqltest.random.design <- fac.design(nlevels = 2, nfactors = 8,
                                              factor.names = c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"))
 
 frac32.design <- FrF2(nruns=32, factor.names=c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"), estimable=formula("~Implement + TripleC+Machine+BGPComp+Lang+Range+Union+Optional + Implement:(TripleC+Machine+BGPComp+Lang+Range+Union+Optional)"), default.levels=c(1,2))
+
+frac64.design <- FrF2(nruns=64, factor.names=c("Implement", "TripleC", "Machine", "BGPComp", "Lang", "Range", "Union", "Optional"), estimable=formula("~Implement + TripleC+Machine+BGPComp+Lang+Range+Union+Optional + Implement:(TripleC+Machine+BGPComp+Lang+Range+Union+Optional)"), default.levels=c(1,2))

### 2013-05-05 exp1/iswc2013.tex
stubs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@500 43ef1190-90fd-4dd7-a769-d0121f02b80e
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@502 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 much simpler than what has been used in benchmarking. We have used the
-dataset of \cite{mxro:Morsey2011DBpedia}, but only consider the
-smallest of their datasets (that we found to me more than 15
+dataset of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
+smallest of their datasets (that we found to be more than 15
 MTriples). Moreover, we have taken subsets of that dataset by using
-the last 1 or 2 MTriples their file.
+the last 1 or 2 MTriples from their file.
 
@@
 the RCurl package, which is a thin wrapper around the curl
-library. Curl has  good facilities for timing requests, and so
-we rely on its measurements.
+library. Curl has well developed facilities for timing requests and
+responses, and so we rely on its measurements.
 
@@
 
-\section{Conclusion}
+\section{Discussion}
+
+\paragraph{Summary and hypothesis test}
+
+\paragraph{Strategy for evaluating experiment}\todo{Say something
+  about examining variance} \todo{Say something about drilling down in
+  factors} \todo{Say something about spreading levels}
+
+\paragraph{Fold-over}
+
+\section{Future Work}
+
+\section{Conclusions}
 

### 2013-05-05 exp1/iswc2013.tex
Introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@504 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   Endpoint Evaluation}
-\author{Kjetil Kjernsmo\inst{1}}
+%\author{Kjetil Kjernsmo\inst{1}}
 %\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
@@
 
-%\maketitle
+\begin{document}
 
-\begin{document}
+\maketitle
 
@@
 \section{Introduction}
+
+The practice of benchmarking is widespread, both in industry to select
+the most appropriate implementation amongst several choices, and in
+academia as means to verify or refute an assertion on the performance
+of a certain practice or system. While we could contribute to the
+industrial cases, our focus is on the latter. A benchmark typically
+consists of several microbenchmarks, which are procedures for placing
+loads on the system to measure its performance as a function of the
+load. For SPARQL benchmarks, usually the performance is measured as a
+the number of queries the SPARQL implementation is able to answer in a
+certain time (the throughput) or the time required for it to respond
+to a query.
+
+However, there are many of problems with this approach when used to
+evaluate assertions about the overall performance of a system or
+several systems: 
+\begin{itemize}
+\item As the complexity increases, the number of parameters that
+one may wish to test increases dramatically.
+\item Nothing compels anyone to find microbenchmarks that
+  can refute an assertion, and there is no structured approach to
+  investigate whether a benchmark is flawed.
+\item There is no meaningful summary of the performance as a whole, so
+  that an assertion about performance cannot be refuted as an
+  optimization may have dentrimental side effects that cannot be
+  identified, or cannot be compared to possibly positive effects in a
+  different microbenchmark.
+\end{itemize}
+
+To address some of these problems, one may try to standardize
+benchmarks, to eliminate some parameters and identify others as
+believed to be important. Also, social pressure should then prompt
+developers to use a standard benchmark so that assertions may be
+refuted.
+
+While this is an improvement which is possibly sufficient for
+industrial uses, is it still flawed as empirical
+research, as it cannot capture the full diversity of the problem: For
+example, if a vendor claims that their implementation is better if
+used on a highly capable hardware platform, then the standardization
+of a hardware platform cannot test this assertion. Moreover, to test
+this assertion, one would need a summary technique that can identify
+the interaction between implementation and hardware platform and
+possibly other parameters.
+
+One important aspect of science is the structured approach to falsify
+hypotheses, and thus is important that assertions about performance is
+formulated as hypotheses that can be falsified.
+
+Thus, attacking the complexity problem by eliminating parameters we
+think influence the outcome for simplicity is indefensible as
+scientific practice. We must instead tackle this problem by enhancing
+our capability of testing complex systems and to identify situations
+where our assumptions are flawed. Such constraints can only be made
+when we have very good reason to suspect that a certain parameter does
+not have influence.
+
+In this paper, we employ techniques from the well-established field in
+statistics of Design of Experiments to point out a direction that
+promises to address these problems. First, we design a simple
+8-parameter experiment, and then we demonstrate how a full
+factorial experiment can meaningfully provide a summary of the
+statistics.
+
+Then, we demonstrate how a fractional factorial experiment can be used
+to maintain a structured approach to the problem while reducing the
+number of runs the experiment needs, and we discuss the tradeoffs
+involved. We then continue to discuss techniques that shows our
+simplistic experiment to be flawed, yet scientifically better than the
+practice of benchmarking. Finally, we outline a roadmap to establish
+the use of these techniques in SPARQL endpoint evaluations.
 

### 2013-05-05 exp1/iswc2013.tex
Small changes Note warm-up and caching
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@507 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 when we have very good reason to suspect that a certain parameter does
-not have influence.
+not have any influence.
 
@@
 practice of benchmarking. Finally, we outline a roadmap to establish
-the use of these techniques in SPARQL endpoint evaluations.
+the use of these techniques in SPARQL endpoint evaluations both as a
+practical tool and as a scientific more rigorous methodology.
 
@@
 
-To only deal with the relatively straightforward aspects of the DoE
-formalism, we have chosen to constrain the experiment strictly to 2
-levels. We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
+First, note that the concept of parameters as considered in the
+introduction is known as \emph{factors} in DoE. For each factor, we
+have chosen in this paper to only deal with the relatively
+straightforward aspects of the DoE formalism, and therefore 
+constrain the experiment strictly to 2 levels.
+
+We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
 dataset, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
@@
 These fragments has been carefully designed for illustrative utility,
-as well as suitable selectivity. When the experiment is run, they are
-combined to yield a query, which is then sent to a SPARQL
-endpoint. The SPARQL endpoint is itself set up using 4store\todo{ref}
-version 1.1.5. The ``Machine'' factor specifies a hostname, whereas
-``TripleC'' and ``Implement'' specifies a port number. On each host,
-four 4store instances runs independently on different
-ports. \todo{possible problem?}
+as well as suitable selectivity. They are not themselves important,
+they only serve to illustrate certain examples of possible
+factors. When the experiment is run, they are combined to yield a
+query, which is then sent to a SPARQL endpoint. The SPARQL endpoint is
+itself set up using 4store\todo{ref} version 1.1.5. The ``Machine''
+factor specifies a hostname, whereas ``TripleC'' and ``Implement''
+specifies a port number. On each host, four 4store instances runs
+independently on different ports. \todo{possible problem?}
 
@@
 
+Finally note that there are two common issues in benchmarking we do
+not consider: we do not allow a warm-up run, nor do we take into
+account that the server may cache all or parts of the result set. The
+reasons for this choice will be discussed later.\todo{lurking
+  variables}
+
 \subsection{Full factorial experiment}

### 2013-05-05 exp1/iswc2013.tex
minor corrections
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@508 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 called DoE.base, BsMD and FrF2. The experiments are run on a third
-computer in a 1 gigabit LAN with the two experiment hosts.  As
+computer in a 1 gigabit Ethernet LAN with the two experiment hosts.  As
 response variable, we have chosen to use the time from the namespace
-server lookup to the data has been transferred. This choice is
+server lookup finishes to the data has been transferred. This choice is
 somewhat arbitrary, many other response variables could be chosen, and

### 2013-05-06 exp1/analysis.R
Working simple t-test
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@511 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,10 +10,19 @@ fulllm <- function(results) {
   lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
 }
 
-robust <- function(model, control = "Implement", inactive = c("Machine", "BGPComp", "Range"), pairwise=FALSE, ...) {
-  fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")));
+robust <- function(model, control = "Implement",
+                   noise = c("TripleC", "Lang", "Union", "Optional"), 
+                   inactive = c("Machine", "BGPComp", "Range"),
+                   pairwise=FALSE, ...) {
+#  fm <- as.formula(paste("experiment~", paste(noise, collapse=" * ")))
+  fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
+#rowser()
+#  ddply(model, fm, function(subset) {
+#    browser()
+#  }
+#    )
   allmeans <- aggregate(fm, data=model, mean)
-  allmeans
+  t.test(allmeans[allmeans["Implement"] == 1,]$experiments, allmeans[allmeans["Implement"] == 2,]$experiments)
 
 }
   

### 2013-05-06 exp1/analysis.R
Pairwise t-test first example
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@512 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,7 +10,7 @@ fulllm <- function(results) {
   lm(formula = experiments ~ Implement * TripleC * BGPComp * Lang * Range * Union * Optional * Machine, data = results)
 }
 
-robust <- function(model, control = "Implement",
+robust.simple <- function(model, control = "Implement",
                    noise = c("TripleC", "Lang", "Union", "Optional"), 
                    inactive = c("Machine", "BGPComp", "Range"),
                    pairwise=FALSE, ...) {
@@ -25,4 +25,12 @@ robust <- function(model, control = "Implement",
   t.test(allmeans[allmeans["Implement"] == 1,]$experiments, allmeans[allmeans["Implement"] == 2,]$experiments)
 
 }
-  
+
+robust.pairwise <- function(model, control = "Implement", 
+                            noise = c("TripleC", "Lang", "Union", "Optional")) {
+  fm <- as.formula(paste("~", paste(noise, collapse=" * ")))
+  dlply(model, fm, function(tmp) {
+    t.test(tmp[tmp["Implement"] == 1,]$experiments, tmp[tmp["Implement"] == 2,]$experiments)
+  })
+}
+ 

### 2013-05-06 exp1/iswc2013.tex
explain implementation constraint
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@513 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Experimens}
+\section{Experiments}
+
+Common experiments include comparing several different
+implementations. However, we focus on a slightly different problem: We
+would like to compare the same implementation before and after some
+change has been made, typically with the intention to enhance the
+performance of the system. There are two reasons for this: One is that
+there are currently many SPARQL implementations available, also Free
+Software ones that can be modified by anyone, and the other is to
+constrain the scope of the study to comparing two different things.
 

### 2013-05-06 exp1/iswc2013.tex
start of full explanation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@514 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 and the second level is 
-\verbatiminput{experiment/BGPComp-1}
+\verbatiminput{experiment/BGPComp-2}
 
@@
 
+The code will be published\todo{github URL}.
+
 \subsection{Full factorial experiment}
+
+In the full factorial experiment, all combinations of the 8 factors
+are executed, in total 256 runs.\todo{concrete example?} This is
+called a $2^8$ factorial experiment, and we can see that the number of
+runs is grows exponentially with the number of factors, thus making
+large experiments prohibitively expensive. We shall nevertheless find
+the following small example instructive. Note that each combination is
+executed only once (i.e. it is \emph{unreplicated}, however, so even a
+full factorial experiment compares well to a typical benchmark in
+which each microbenchmark must be executed several times to obtain
+enough data to compute a mean with reasonable accuracy.
+
+The above experimental setup is executed across a design matrix
+generated with the DoE.base package, which returns a data frame
+consisting of the factors with a column containing the corresponding
+curl measurements. To analyze the experiment, we first generate a
+normal plot, see figure~\ref{fig:fullnormal}. In this plot, if the
+null hypothesis $H_0$ holds, all estimated effects will fall on a
+straight line, as their means would be zero. Any significant departure
+from a straight line can be interpreted as a significant effect (see
+Section~4.8 of \cite{wu2009experiments} for a detailed
+explanation). In our case, it is most important to note that any
+effect that is negative (i.e. has a negative slope) means the effect
+enhances the performance of the SPARQL endpoint, the runtime
+decreases. If ``Implement'' and the sum of its interactions are
+negative, then it supports the alternative hypothesis, i.e. we have
+successfully improved the performance.
+
+To proceed beyond a visual judgement of significance, we may use
+Lenth's method\todo{ref Lenth}, which is a simple method for
+estimating significance in an experiment with no replicates. We plot
+the results in figure~\ref{fig:fulllenth}. Visually, this is not as
+useful, since the large number\todo{really include?} We see that we
+have very significant effects. We also note that ``Implement'' itself
+is significantly negative. We also note that the ``Implement:Lang''
+interaction is significantly positive, as expected from our setup,
+where ``Lang'' was the detrimental side-effect we simulated.
+
+To investigate whether the detrimental side-effects cancels the
+positive effect of our simulated optimization, we have to formulate a
+hypothesis test. By inspecting the normal plot in figure~\ref{fig:fullnormal}
 

### 2013-05-06 exp1/iswc2013.tex
comment pending tripleC disaster
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@515 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 estimating significance in an experiment with no replicates. We plot
-the results in figure~\ref{fig:fulllenth}. Visually, this is not as
-useful, since the large number\todo{really include?} We see that we
-have very significant effects. We also note that ``Implement'' itself
-is significantly negative. We also note that the ``Implement:Lang''
-interaction is significantly positive, as expected from our setup,
-where ``Lang'' was the detrimental side-effect we simulated.
+the results in figure~\ref{fig:fulllenth}. In
+figure~\ref{fig:fullnormal}, Lenth's method is used to label the
+significant effects. Visually, this is not as useful, since the large
+number\todo{really include?} We see that we have very significant
+effects. We also note that ``Implement'' itself is significantly
+negative. We also note that the ``Implement:Lang'' interaction is
+significantly positive, as expected from our setup, where ``Lang'' was
+the detrimental side-effect we simulated.
 
@@
 positive effect of our simulated optimization, we have to formulate a
-hypothesis test. By inspecting the normal plot in figure~\ref{fig:fullnormal}
+hypothesis test. By inspecting the normal plot in
+figure~\ref{fig:fullnormal}, we see that only five factors are highly
+significant by either being a significant main effect or participate
+in a highly significant interaction. We say that the three factors
+``BGPComp'', ``Range'' and ``Machine'' are \emph{inactive} since they
+do not contribute significantly to the observed variation in
+performance. In practical situations, the four factors ``Lang'',
+``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
+i.e. we cannot change the size of the dataset without loss of
+information, or change the query, since it gives a different
+answer. In practice, we can only control ``Implement'', i.e. we can
+only change the implementation. In this context, we call ``Implement''
+a \emph{control} factor and the other four active factors
+\emph{environmental}.
+
+\todo{finish the t-test stuff once ready}
+
+The main objective of this paper is not to establish that the
+optimization is significant, it is merely a simulation. It is to
+establish a critical practice of evaluations. It is of critical
+importance that we understand why some effects are critical. That
+``Implement'' is significant is hardly a surprise, that is what we
+worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
+are strong is also due to that these are fairly demanding things to
+evaluate. Since the delay we inserted affects all kinds of joins, it
+is also intuitive that interactions between ``Implement'' and those
+that require joins are strong. The strong ``BGPComp:Lang'' and
+``Lang:Union'' interactions might be due to that the many more triples
+needs to be searched for a language tag in one case. However, the
+positive ``Implement:TripleC'' interaction evades such explanations;
+in fact, it hints that our optimization may not work as well for
+larger databases. This should cause some alarm. It is a fundamental
+property of two-level experiments that it can only model linear
+effects. If the effect is truly linear, and the variance is small, the
+two levels can be quite close together and still yield a reasonable
+estimate for the effect. However, if the effect is non-linear, the
+choice of levels becomes critical. In this case, the experimenter
+should be concerned that they may have a non-linear effect that is not
+well represented by the sizes of datasets. Such reasoning should be
+applied to all significant effects.
 

### 2013-05-06 exp1/iswc2013.tex
Describe fractional experiments describe 32frac
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@516 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 establish a critical practice of evaluations. It is of critical
-importance that we understand why some effects are critical. That
+importance that we understand why some effects are critical.\todo{add
+  table of effects} That
 ``Implement'' is significant is hardly a surprise, that is what we
@@
 \subsection{Fractional factorial experiment}
+
+As mentioned, the full factorial experiment goes as $2^n$ and becomes
+prohibitivly expensive when $n$ is a large number of
+factor. As the evaluation of SPARQL endpoints is an inherently complex
+problem, a large number of factors are needed, and so full experiments
+cannot scale. However, we may reduce the size of the experiment
+significantly by sacrificing some explanatory power, in the form of
+fractional factorial experiments.
+
+We loose explanatory power due to \emph{aliasing}, for example, the
+``TripleC:BGPComp'' interaction may be aliased with the
+``Machine:Range'' interaction. That is to say, a detected increase in
+run time from larger Basic Graph Patterns for large databases, can
+also be explained by a less powerful machine evaluation FILTER clauses
+with ranges. They are indistinguishable. This may or may not cause a
+problem. In many cases, we may not be interested in these effects, we
+may only be interested in ``Implement'' and its
+interactions. Moreover, it is possible (even easy using the FrF2
+package in R) to declare which effects must not be aliased, and
+determine the size of the experiment based on that. Such a main effect
+or two-factor interaction is called \emph{clear} if none of its
+aliases are other main effects or two-factor interactions. Another
+possibility to shrink the size of the experiment is if we \textit{a
+  priori} can say that some effects are negligible. We have not found
+such assumptions to be tenable in our case.
+
+We have made two fractional factorial designs: One with 32 runs and
+one with 64 runs. Again, the number of runs is in powers of 2, but
+much smaller. In both cases, we have specified that all the factors
+must be clear and also all two-factor interactions where ``Implement''
+is involved must be clear. \todo{declare generators?}
+
+The resulting normal plot is in \figure{fig:frac32normal}. We see that
+only two main effects are significant on a $\alpha = 0.05$ level
+according to the Lenth criterion, ``Implement'' is again deemed a
+significant improvement, and ``Union'' is deemed significantly the
+hardest operation for the query engine. We see that some other points
+also deviate from the straight line, but with as few runs as this, the
+total variance is great, so no other conclusions can be drawn. Also,
+the Student's t-test we used in the previous section cannot be
+employed, as we have no additional data to use for the test.
+
+For any other purpose than a rough idea of some key influencers on the
+overall performance of the endpoint, this little experiment is insufficient.
 

### 2013-05-06 exp1/iswc2013.tex
large dataset explanation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@517 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The resulting normal plot is in \figure{fig:frac32normal}. We see that
+The resulting normal plot is in figure~\ref{fig:frac32normal}. We see that
 only two main effects are significant on a $\alpha = 0.05$ level
@@
 For any other purpose than a rough idea of some key influencers on the
-overall performance of the endpoint, this little experiment is insufficient.
+overall performance of the endpoint for the given levels, this little
+experiment is insufficient.
+
+We turn to the 64-run experiment and its normal plot in
+figure~\ref{fig:frac64normalalpha005}. We see that the we get more significant
+effects, and that they correspond well to those we saw in the full
+factorial experiment. They also all have an intuitive
+explanation. However, we do not see the effects that we found
+worrisome in the full factorial design. The ``Implement:TripleC''
+interaction emerges only for a level $\alpha=0.15$, see
+figure~\ref{fig:frac64normalalpha015}. To ensure that we discover such
+cases will be a key challenge in further studies.
+
+\todo{Do t-tests}
 
 \subsection{Fractional factorial experiment with more data}
+
+The importance of finding flaws in the levels is highlighted by the
+normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
+experiment made it feasible to run with a larger dataset, this
+experiment uses the original 15~MTriples dataset of
+\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
+previous discussion, for larger ``TripleC'', our simulated performance
+is actually worse, as evident from the positive
+``Implement''. We also see that the interactions ``Implement:Lang''
+and ``Implement:TripleC'' is highly significant. It is clear what
+happened: For larger ``TripleC'', the delay we introduced in the
+langMatches function has a devastating effect, since it delays the
+execution for every returned triple, and thereby completely dominates
+the execution time.
+
+Moreover, we note that the ``Machine'' factor has become highly
+significant in the other end along with its ``Machine:TripleC''
+interaction. One should take careful notice that the ``Machine''
+factor encompasses many different things, that should be different
+factors; CPU, disks, available RAM, installed software, tuning
+parameters and so on. It is advantegous to have such broad factors in
+the design in early experiments, because they may point out unexpected
+effects that should be further scrutinized. The ``Machine'' factor is
+a good example of usage that also helps us manage
+complexity: If possible, one may group many factors into one if one is
+in doubt of its relevance and break it up in subsequent
+experiments. In this case, the cause of the performance problem was
+highly audible: It was the old P-ATA disks of the larger machine, but
+in general, such clues are not available to the experimenter.
+
+We are forced to conclude that the three preceeding experiments are
+severely flawed, as they failed to take into account the non-linear
+effects arising from larger data sizes on slow hardware.
+
+Note that we do not take advantage of the usual performance of 4store
+in this simulation, to the contrary, we chose 4store because of its
+relative simplicity to modify it to suite our needs for illustrative
+purposes, which consists of degrading it in many ways. The small
+datasets used here does not imply that the approach will not work for
+large datasets; properly configured, actual optimizations will give
+very different runtimes, and the small number of runs will make it
+possible to evaluate a very large number of factors.
 

### 2013-05-06 exp1/iswc2013.tex
small changes spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@518 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   that an assertion about performance cannot be refuted as an
-  optimization may have dentrimental side effects that cannot be
+  optimization may have detrimental side effects that cannot be
   identified, or cannot be compared to possibly positive effects in a
@@
 Thus, attacking the complexity problem by eliminating parameters we
-think influence the outcome for simplicity is indefensible as
+think influence the outcome for simplicity is untenable as
 scientific practice. We must instead tackle this problem by enhancing
@@
 In this paper, we employ techniques from the well-established field in
-statistics of Design of Experiments to point out a direction that
-promises to address these problems. First, we design a simple
-8-parameter experiment, and then we demonstrate how a full
+statistics known as Design of Experiments (DoE) to point out a
+direction that promises to address these problems. First, we design a
+simple 8-parameter experiment, and then we demonstrate how a full
 factorial experiment can meaningfully provide a summary of the
@@
 to maintain a structured approach to the problem while reducing the
-number of runs the experiment needs, and we discuss the tradeoffs
+number of runs the experiment needs, and we discuss the trade-offs
 involved. We then continue to discuss techniques that shows our
 simplistic experiment to be flawed, yet scientifically better than the
-practice of benchmarking. Finally, we outline a roadmap to establish
+practice of benchmarking. Finally, we outline a road-map to establish
 the use of these techniques in SPARQL endpoint evaluations both as a
-practical tool and as a scientific more rigorous methodology.
+practical tool and as a scientifically more rigorous methodology.
 
@@
 much simpler than what has been used in benchmarking. We have used the
-dataset of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
-smallest of their datasets (that we found to be more than 15
-MTriples). Moreover, we have taken subsets of that dataset by using
+data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
+smallest of their data-sets (that we found to be more than 15
+MTriples). Moreover, we have taken subsets of that data-set by using
 the last 1 or 2 MTriples from their file.
@@
 We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
-dataset, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
+data-set, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
 where one machine runs GNU/Linux Debian Squeeze, has 16 GB RAM, an
-Intel Core2 Duo E8500 CPU and two Parallell-ATA disks in a RAID-1
+Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
 configuration. The other machine runs Debian Wheezy, has 8 GB RAM, an
@@
 
-To proceed beyond a visual judgement of significance, we may use
+To proceed beyond a visual judgment of significance, we may use
 Lenth's method\todo{ref Lenth}, which is a simple method for
@@
 ``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
-i.e. we cannot change the size of the dataset without loss of
+i.e. we cannot change the size of the data-set without loss of
 information, or change the query, since it gives a different
@@
 should be concerned that they may have a non-linear effect that is not
-well represented by the sizes of datasets. Such reasoning should be
+well represented by the sizes of data-sets. Such reasoning should be
 applied to all significant effects.
@@
 As mentioned, the full factorial experiment goes as $2^n$ and becomes
-prohibitivly expensive when $n$ is a large number of
-factor. As the evaluation of SPARQL endpoints is an inherently complex
+prohibitively expensive when $n$ is a large number of
+factors. As the evaluation of SPARQL endpoints is an inherently complex
 problem, a large number of factors are needed, and so full experiments
@@
 
-For any other purpose than a rough idea of some key influencers on the
+For any other purpose than a rough idea of some key influences on the
 overall performance of the endpoint for the given levels, this little
@@
 factorial experiment. They also all have an intuitive
-explanation. However, we do not see the effects that we found
+explanation as above. However, we do not see the effects that we found
 worrisome in the full factorial design. The ``Implement:TripleC''
@@
 normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
-experiment made it feasible to run with a larger dataset, this
-experiment uses the original 15~MTriples dataset of
+experiment made it feasible to run with a larger data-set, this
+experiment uses the original 15~MTriples data-set of
 \cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
@@
 factors; CPU, disks, available RAM, installed software, tuning
-parameters and so on. It is advantegous to have such broad factors in
+parameters and so on. It is advantageous to have such broad factors in
 the design in early experiments, because they may point out unexpected
@@
 
-We are forced to conclude that the three preceeding experiments are
+We are forced to conclude that the three preceding experiments are
 severely flawed, as they failed to take into account the non-linear
@@
 purposes, which consists of degrading it in many ways. The small
-datasets used here does not imply that the approach will not work for
-large datasets; properly configured, actual optimizations will give
+data-sets used here does not imply that the approach will not work for
+large data-sets; properly configured, actual optimization will give
 very different runtimes, and the small number of runs will make it
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 

### 2013-05-07 exp1/analysis.R
remove outcommented code
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@519 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -14,16 +14,9 @@ robust.simple <- function(model, control = "Implement",
                    noise = c("TripleC", "Lang", "Union", "Optional"), 
                    inactive = c("Machine", "BGPComp", "Range"),
                    pairwise=FALSE, ...) {
-#  fm <- as.formula(paste("experiment~", paste(noise, collapse=" * ")))
   fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
-#rowser()
-#  ddply(model, fm, function(subset) {
-#    browser()
-#  }
-#    )
   allmeans <- aggregate(fm, data=model, mean)
   t.test(allmeans[allmeans["Implement"] == 1,]$experiments, allmeans[allmeans["Implement"] == 2,]$experiments)
-
 }
 
 robust.pairwise <- function(model, control = "Implement", 

### 2013-05-07 exp1/analysis.R
Refactor to generalize
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@520 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -11,19 +11,18 @@ fulllm <- function(results) {
 }
 
 robust.simple <- function(model, control = "Implement",
-                   noise = c("TripleC", "Lang", "Union", "Optional"), 
                    inactive = c("Machine", "BGPComp", "Range"),
                    pairwise=FALSE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
   allmeans <- aggregate(fm, data=model, mean)
-  t.test(allmeans[allmeans["Implement"] == 1,]$experiments, allmeans[allmeans["Implement"] == 2,]$experiments)
+  t.test(allmeans[allmeans[control] == 1,]$experiments, allmeans[allmeans[control] == 2,]$experiments)
 }
 
 robust.pairwise <- function(model, control = "Implement", 
                             noise = c("TripleC", "Lang", "Union", "Optional")) {
   fm <- as.formula(paste("~", paste(noise, collapse=" * ")))
-  dlply(model, fm, function(tmp) {
-    t.test(tmp[tmp["Implement"] == 1,]$experiments, tmp[tmp["Implement"] == 2,]$experiments)
+  dlply(model, fm, function(subset) {
+    t.test(subset[subset[control] == 1,]$experiments, subset[subset[control] == 2,]$experiments)
   })
 }
  

### 2013-05-07 exp1/analysis.R
do one-sided t tests
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@521 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -15,14 +15,18 @@ robust.simple <- function(model, control = "Implement",
                    pairwise=FALSE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
   allmeans <- aggregate(fm, data=model, mean)
-  t.test(allmeans[allmeans[control] == 1,]$experiments, allmeans[allmeans[control] == 2,]$experiments)
+  t.test(allmeans[allmeans[control] == 2,]$experiments,
+         allmeans[allmeans[control] == 1,]$experiments,
+         alternative="less")
 }
 
 robust.pairwise <- function(model, control = "Implement", 
                             noise = c("TripleC", "Lang", "Union", "Optional")) {
   fm <- as.formula(paste("~", paste(noise, collapse=" * ")))
   dlply(model, fm, function(subset) {
-    t.test(subset[subset[control] == 1,]$experiments, subset[subset[control] == 2,]$experiments)
+    t.test(subset[subset[control] == 2,]$experiments,
+           subset[subset[control] == 1,]$experiments,
+           alternative="less")
   })
 }
  

### 2013-05-07 exp1/iswc2013.tex
Small additions to rdata; todo on explaining non-linear
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@522 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 two levels can be quite close together and still yield a reasonable
-estimate for the effect. However, if the effect is non-linear, the
+estimate for the effect.\todo{ikke helt god diskusjon, øke avstand vil
+gi klarere effekter, i lineær sammenheng} However, if the effect is non-linear, the
 choice of levels becomes critical. In this case, the experimenter

### 2013-05-08 exp1/iswc2013.tex
note bonferroni problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@524 43ef1190-90fd-4dd7-a769-d0121f02b80e
overall hypothesis test
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@525 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\todo{finish the t-test stuff once ready}
+The presence of inactive factors makes it possible to use a standard
+t-test by averaging the inactive factors for each combination of
+control and environmental factors. This enables us to treat our
+experiment as a $2^5$ experiment replicated 8 times.
+
+We then extract two vectors, one for the each of the implementations,
+and then perform a one-sided hypothesis test where $H_0$ is that they
+are equal and $H_1$ is that the vector representing the new
+implementation is smaller. In this example, we find that the available
+data supports the assertion that the new implementation is better with a
+high probability, $p=2.9 \cdot 10^{-10}$.\todo{rationale could be stronger}
+
+For future larger experiments, one must investigate whether it is
+necessary to adjust the p-value due to the larger numbers of
+hypotheses tested.
 

### 2013-05-08 exp1/iswc2013.tex
Table with p-values
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@526 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 data supports the assertion that the new implementation is better with a
-high probability, $p=2.9 \cdot 10^{-10}$.\todo{rationale could be stronger}
+high probability, $p=2.9 \cdot 10^{-10}$.\todo{rationale could be
+  stronger}
+
+While it is interesting to test the overall hypothesis that the new
+implementation is an improvement, it is also interesting to know if
+there are cases that it fails. Again, we treat the experiment as a
+$2^5$ experiment replicated 8 times. Now, for each of the 16 resulting
+combinations of the environmental factors for each level of
+``Implement'', we extract 8 values. We may now run a t-test for each
+of the replications, as above. The results are tabulated in
+table~\ref{tab:pvaluesfull}. We see that there are significant
+improvements in most cases, one may want to further investigate the
+cases where they do not.
+
+\begin{table}[ht]
+\begin{center}
+\caption{p-values for different parts of the experiment}\label{tab:pvaluesfull}
+\begin{tabular}{ccccl}
+  \hline
+``TripleC'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
+  \hline
+  1 & 1 & 1 & 1 & 0.00032 \\ 
+  1 & 1 & 1 & 2 & $2 \cdot 10^{-14}$ \\ 
+  1 & 1 & 2 & 1 & $4.7 \cdot 10^{-14}$ \\ 
+  1 & 1 & 2 & 2 & $2 \cdot 10^{-18}$ \\ 
+  1 & 2 & 1 & 1 & $1.4 \cdot 10^{-06}$ \\ 
+  1 & 2 & 1 & 2 & $8.3 \cdot 10^{-11}$ \\ 
+  1 & 2 & 2 & 1 & 0.0017 \\ 
+  1 & 2 & 2 & 2 & $2.1 \cdot 10^{-06}$ \\ 
+  2 & 1 & 1 & 1 & 0.09 \\ 
+  2 & 1 & 1 & 2 & $1.6 \cdot 10^{-09}$ \\ 
+  2 & 1 & 2 & 1 & $8.4 \cdot 10^{-10}$ \\ 
+  2 & 1 & 2 & 2 & $2.5 \cdot 10^{-13}$ \\ 
+  2 & 2 & 1 & 1 & 0.29 \\ 
+  2 & 2 & 1 & 2 & $9.1 \cdot 10^{-07}$ \\ 
+  2 & 2 & 2 & 1 & 0.92 \\ 
+  2 & 2 & 2 & 2 & 0.071 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 

### 2013-05-08 exp1/iswc2013.tex
smaller changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@527 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 been published in
-\cite{Springer-2010-Experimental-Methods-for-the-Analysis-of-Optimization-Algorithms}. 
-We have chosen to turn to statistical standard texts
-\cite{wu2009experiments} for this study.
+\cite{Springer-2010-Experimental-Methods-for-the-Analysis-of-Optimization-Algorithms}
+demonstrating the broad applicability of the methodology we employ.
+We have chosen SPARQL Endpoint evaluation as an example in this study
+since SPARQL Federation is our main interest, and to turn to
+statistical standard texts \cite{wu2009experiments} for this study.
 
@@
 
-The code will be published\todo{github URL}.
+All the code to reproduce this study as well as detailed instructions
+is published on Github: \texttt{https://github.com/kjetilk/doe-sparql}.
 
@@
 which each microbenchmark must be executed several times to obtain
-enough data to compute a mean with reasonable accuracy.
+enough data to compute a mean with reasonable certainty.
 

### 2013-05-08 exp1/iswc2013.tex
htest for fractional
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@528 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 high probability, $p=2.9 \cdot 10^{-10}$.\todo{rationale could be
-  stronger}
+  stronger}\todo{figure?}
 
@@
 
-\todo{Do t-tests}
+Since the 64-run experiment also has 3 inactive factors, ``Machine'',
+``TripleC'' and ``Range'', we can again treat it as a $2^5$ experiment
+replicated twice. Using the same procedure as above, we find that
+again, the hypothesis test indicates that the new implementation is an
+improvement at $p = 5.9 \cdot 10^{-6}$. The comparatively higher p-value
+is due to the lower number of runs. It is also possible to perform the
+per-environmental test, but we refrain from that in the interest of
+brevity.\todo{good way to say it?}
 

### 2013-05-08 exp1/iswc2013.tex
small changes in the large fractional
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@529 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 previous discussion, for larger ``TripleC'', our simulated performance
-is actually worse, as evident from the positive
-``Implement''. We also see that the interactions ``Implement:Lang''
-and ``Implement:TripleC'' is highly significant. It is clear what
-happened: For larger ``TripleC'', the delay we introduced in the
-langMatches function has a devastating effect, since it delays the
-execution for every returned triple, and thereby completely dominates
-the execution time.
+is actually worse, as evident from the positive ``Implement''. We also
+see that the interactions ``Implement:Lang'' and ``Implement:TripleC''
+is highly significant. It is clear what happened: For larger
+``TripleC'', the delay we introduced in the langMatches function has a
+devastating effect, since it delays the execution for every triple
+that the query engine applies the filter to, and thereby completely
+dominates the execution time.
 
@@
 effects that should be further scrutinized. The ``Machine'' factor is
-a good example of usage that also helps us manage
-complexity: If possible, one may group many factors into one if one is
-in doubt of its relevance and break it up in subsequent
-experiments. In this case, the cause of the performance problem was
-highly audible: It was the old P-ATA disks of the larger machine, but
-in general, such clues are not available to the experimenter.
+a good example of usage that also helps us manage complexity: If
+possible, one may group many factors into one if one is in doubt of
+its relevance and break it up in subsequent experiments if deeper
+understanding is needed. In this case, the cause of the performance
+problem was highly audible: It was the old P-ATA disks of the larger
+machine, but in general, such clues are not available to the
+experimenter, therefore more factors are needed.
 
@@
 relative simplicity to modify it to suite our needs for illustrative
-purposes, which consists of degrading it in many ways. The small
+purposes, which consists of degrading it in several ways. The small
 data-sets used here does not imply that the approach will not work for

### 2013-05-08 exp1/generate-figures.R
generate some figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@530 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,27 @@
+library(FrF2)
+library(plyr)
+
+load("RelevantExperiments.RData")
+
+source("analysis.R")
+
+postscript(file="fullnormal.eps")
+DanielPlot(runbgplangf1.lm)
+dev.off()
+
+postscript(file="fulllenth.eps")
+LenthPlot(runbgplangf1.lm)
+dev.off()
+
+postscript(file="frac32normal.eps")
+DanielPlot(frac32r1.lm)
+dev.off()
+
+postscript(file="frac64normalalpha005.eps")
+DanielPlot(frac64r3.lm)
+dev.off()
+
+postscript(file="frac64normalalpha015l.eps")
+DanielPlot(frac64r3.lm, alpha=0.15)
+dev.off()
+

### 2013-05-08 exp1/generate-figures.R
all current figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@531 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -25,3 +25,6 @@ postscript(file="frac64normalalpha015l.eps")
 DanielPlot(frac64r3.lm, alpha=0.15)
 dev.off()
 
+postscript(file="frac64hugenormal.eps")
+DanielPlot(frac64huger1.lm)
+dev.off()

### 2013-05-08 exp1/iswc2013.tex
all figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@532 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{verbatim}
+\usepackage{graphicx}
 \newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
@@
 curl measurements. To analyze the experiment, we first generate a
-normal plot, see figure~\ref{fig:fullnormal}. In this plot, if the
+normal plot, see figure~\ref{fig:fullnormal}\todo{explain axes}. In this plot, if the
 null hypothesis $H_0$ holds, all estimated effects will fall on a
@@
 
+\begin{figure}[h]
+  \includegraphics[width=\textwidth]{fullnormal.pdf}
+  \caption{A normal plot of the Full Factorial Experiment. The
+    labelled points are considered significant by using the Lenth
+    criterion at a level $\alpha=0.05$.}\label{fig:fullnormal}
+\end{figure}
+
 To proceed beyond a visual judgment of significance, we may use
@@
 estimating significance in an experiment with no replicates. We plot
-the results in figure~\ref{fig:fulllenth}. In
+the results in figure~\ref{fig:fulllenth}\todo{relationship level-SME/ME}. In
 figure~\ref{fig:fullnormal}, Lenth's method is used to label the
@@
 the detrimental side-effect we simulated.
+
+\begin{figure}[h]
+  \includegraphics[width=\textwidth]{fulllenth.pdf}
+  \caption{A Lenth plot of the Full Factorial Experiment. The
+    conservative Lenth criterion is indicated by the horizontal lines
+    labelled SME, the liberal criterion is labelled ME. They are at
+    ±0.90 and ±0.46 respectively. }\label{fig:fulllenth}
+\end{figure}
 
@@
 
+\begin{figure}[hb]
+  \includegraphics[width=\textwidth]{frac32normal.pdf}
+  \caption{A normal plot of the Fractional Factorial Experiment with
+    32 runs. The labelled points are considered significant by using
+    the Lenth criterion at a level
+    $\alpha=0.05$.}\label{fig:frac32normal}
+\end{figure}
+
 The resulting normal plot is in figure~\ref{fig:frac32normal}. We see that
@@
 cases will be a key challenge in further studies.
+
+\begin{figure}[ht]
+  \includegraphics[width=\textwidth]{frac64normalalpha005.pdf}
+  \caption{A normal plot of the Fractional Factorial Experiment with
+    64 runs. The labelled points are considered significant by using
+    the Lenth criterion at a level
+    $\alpha=0.05$.}\label{fig:frac64normalalpha005}
+\end{figure}
+\begin{figure}[hb]
+  \includegraphics[width=\textwidth]{frac64normalalpha015.pdf}
+  \caption{A normal plot of the Fractional Factorial Experiment with
+    64 runs. The labelled points are considered significant by using
+    the Lenth criterion at a level
+    $\alpha=0.15$.}\label{fig:frac64normalalpha015}
+\end{figure}
 
@@
 dominates the execution time.
+
+\begin{figure}[h!]
+  \includegraphics[width=\textwidth]{frac64hugenormal.pdf}
+  \caption{A normal plot of the Fractional Factorial Experiment with
+    64 runs where the level 2 of ``TripleC'' is 15 MTriples. The
+    labelled points are considered significant by using the Lenth
+    criterion at a level $\alpha=0.05$.}\label{fig:frac64hugenormal}
+\end{figure}
 

### 2013-05-08 exp1/generate-figures.R
remove LenthPlot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@534 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -5,26 +5,22 @@ load("RelevantExperiments.RData")
 
 source("analysis.R")
 
-postscript(file="fullnormal.eps")
+pdf(file="fullnormal.pdf")
 DanielPlot(runbgplangf1.lm)
 dev.off()
 
-postscript(file="fulllenth.eps")
-LenthPlot(runbgplangf1.lm)
-dev.off()
-
-postscript(file="frac32normal.eps")
+pdf(file="frac32normal.pdf")
 DanielPlot(frac32r1.lm)
 dev.off()
 
-postscript(file="frac64normalalpha005.eps")
+pdf(file="frac64normalalpha005.pdf")
 DanielPlot(frac64r3.lm)
 dev.off()
 
-postscript(file="frac64normalalpha015l.eps")
+pdf(file="frac64normalalpha015.pdf")
 DanielPlot(frac64r3.lm, alpha=0.15)
 dev.off()
 
-postscript(file="frac64hugenormal.eps")
+pdf(file="frac64hugenormal.pdf")
 DanielPlot(frac64huger1.lm)
 dev.off()

### 2013-05-08 exp1/iswc2013.tex
remove LenthPlot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@533 43ef1190-90fd-4dd7-a769-d0121f02b80e
Fix all needed references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@535 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 query, which is then sent to a SPARQL endpoint. The SPARQL endpoint is
-itself set up using 4store\todo{ref} version 1.1.5. The ``Machine''
-factor specifies a hostname, whereas ``TripleC'' and ``Implement''
-specifies a port number. On each host, four 4store instances runs
-independently on different ports. \todo{possible problem?}
+itself set up using 4store (see \cite{harris20094store}) version
+1.1.5. The ``Machine'' factor specifies a hostname, whereas
+``TripleC'' and ``Implement'' specifies a port number. On each host,
+four 4store instances runs independently on different
+ports. \todo{possible problem?}
 
@@
 
-The experimentation is implemented in R\todo{ref}, which is a free
+The experimentation is implemented in R\cite{rmanual}, which is a free
 software environment for statistical computing and graphics. Necessary
 tools for DoE has been implemented by the R community in packages
-called DoE.base, BsMD and FrF2. The experiments are run on a third
+called DoE.base\cite{doebase} and FrF2\cite{frf2}. The experiments are run on a third
 computer in a 1 gigabit Ethernet LAN with the two experiment hosts.  As
@@
 the SPARQL endpoint well. For the measurements, we have chosen to use
-the RCurl package, which is a thin wrapper around the curl
+the RCurl package\cite{lang2007r}, which is a thin wrapper around the curl
 library. Curl has well developed facilities for timing requests and
@@
 To proceed beyond a visual judgment of significance, we may use
-Lenth's method\todo{ref Lenth}, which is a simple method for
-estimating significance in an experiment with no replicates. We plot
-the results in figure~\ref{fig:fulllenth}\todo{relationship level-SME/ME}. In
+Lenth's method\cite{lenthmethod}, which is a simple method for
+estimating significance in an experiment with no replicates. In
 figure~\ref{fig:fullnormal}, Lenth's method is used to label the
 significant effects. Visually, this is not as useful, since the large
-number\todo{really include?} We see that we have very significant
+number\todo{how to fix this...?} We see that we have very significant
 effects. We also note that ``Implement'' itself is significantly
@@
 the detrimental side-effect we simulated.
-
-\begin{figure}[h]
-  \includegraphics[width=\textwidth]{fulllenth.pdf}
-  \caption{A Lenth plot of the Full Factorial Experiment. The
-    conservative Lenth criterion is indicated by the horizontal lines
-    labelled SME, the liberal criterion is labelled ME. They are at
-    ±0.90 and ±0.46 respectively. }\label{fig:fulllenth}
-\end{figure}
 
@@
 %\bibliographystyle{splncs03}
-\bibliography{selectivity,federation,benchmarks,optimization,experimentsperformance}
+\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance}
 

### 2013-05-08 exp1/iswc2013.tex
Acknowledgements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@536 43ef1190-90fd-4dd7-a769-d0121f02b80e
future work start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@537 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsection{Fractional factorial experiment}
+\subsection{Fractional factorial experiment}\label{sec:frac}
 
@@
 
+The present experiment is very primitive in its choice of factors, and
+we believe that the strategy employed to construct suitable
+experiments will be the main determinant for the long-term feasibility
+of this direction. One strategy will be to start with broad factors
+such as ``Machine'' and refine them with experience. A more difficult
+issue is how to address the problem of testing the whole SPARQL
+language. We believe that the parameterization work inititated by
+\cite{goerlitz2012splodge} with their SPLODGE system is a key
+ingredient. They parameterized SPARQL queries in such a way they could
+be autogenerated. One suggestion is to parameterize queries based on
+the grammar, so that one factor becomes the number of e.g. UNIONs, and
+the levels are chosen. One problem one will soon run into is that
+parts of the language is recursive, e.g. a GroupGraphPattern can
+consist of several GroupGraphPatterns. However, we believe that
+pragmatic limits can be set, it is for example not of practical
+interest to nest OPTIONALs to any great depth even if it is allowed by
+the language.
+
+We saw in section~\ref{sec:frac} that with a smaller experiment, fewer
+interactions are significant, and so provides us with fewer clues to
+assess the soundness of the experiment. This should be a focus of
+further research.
+
+Finally, we have only employed the simplest parts of the DoE
+theory, what has been presented here is part of a much more general
+formalism known as Orthogonal Arrays. The use of orthogonal arrays
+allow for different numbers of levels, for much greater flexibility in
+total run size, and for irregular designs that can provide a good fit
+for the complex problem of SPARQL endpoint evaluations.
+
 \section{Conclusions}
 
+Feasible doing millions of runs.
+
 \section*{Acknowledgments}
+
+Kjetil Kjernsmo would like to thank his main supervisor Martin Giese
+for kind assistance in writing this paper. He would also like to
+thank Steve Harris for his generous support in degrading his excellent
+work on 4store.
 

### 2013-05-08 exp1/iswc2013.tex
sparql citation a bit on conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@539 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 issue is how to address the problem of testing the whole SPARQL
-language. We believe that the parameterization work inititated by
+language\cite{sparql11query}. We believe that the parameterization work inititated by
 \cite{goerlitz2012splodge} with their SPLODGE system is a key
@@
 
+We saw that while 2 level experiments can perform well in determining
+significant effects, they do not necessarily work well to find a
+model, as exemplified by failure to identify the probable non-linear
+effect of the P-ATA disks. This issue would be fixed by using an
+orthogonal array design.
+
 Feasible doing millions of runs.
@@
 %\bibliographystyle{splncs03}
-\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance}
+\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance,specs}
 

### 2013-05-09 exp1/iswc2013.tex
relax discussion of non-linearity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@545 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-%\documentclass{llncs}
-\documentclass{article}
+\documentclass{llncs}
+%\documentclass{article}
+\usepackage[utf8]{inputenc}
 \usepackage{cite}
@@
 All the code to reproduce this study as well as detailed instructions
-is published on Github: \texttt{https://github.com/kjetilk/doe-sparql}.
+is published on Github: \url{https://github.com/kjetilk/doe-sparql}.
 
@@
 explanation). In our case, it is most important to note that any
-effect that is negative (i.e. has a negative slope) means the effect
+effect that is negative means the effect
 enhances the performance of the SPARQL endpoint, the runtime
@@
 in fact, it hints that our optimization may not work as well for
-larger databases. This should cause some alarm. It is a fundamental
-property of two-level experiments that it can only model linear
-effects. If the effect is truly linear, and the variance is small, the
-two levels can be quite close together and still yield a reasonable
-estimate for the effect.\todo{ikke helt god diskusjon, øke avstand vil
-gi klarere effekter, i lineær sammenheng} However, if the effect is non-linear, the
-choice of levels becomes critical. In this case, the experimenter
-should be concerned that they may have a non-linear effect that is not
-well represented by the sizes of data-sets. Such reasoning should be
-applied to all significant effects.
+larger databases. This should cause some alarm. 
+It is a fundamental property of two-level experiments that it can only
+model linear effects. If the effect is truly linear, and the variance
+is small, the two levels can be quite close together and still yield a
+reasonable estimate for the effect, but the effect will be greater
+with greater separation. Additionally if the effect is non-linear, the
+choice of levels is critical. In this case, the experimenter should be
+concerned that the separation between levels for the sizes of the
+data-set is too small and should further investigate. Such reasoning
+should be applied to all significant effects.
 
@@
 the Student's t-test we used in the previous section cannot be
-employed, as we have no additional data to use for the test.
+employed, as we have no additional data to use for the
+test.\todo{actually, we can}
 
@@
 We are forced to conclude that the three preceding experiments are
-severely flawed, as they failed to take into account the non-linear
-effects arising from larger data sizes on slow hardware.
+severely flawed, as they failed to take into account the effects from
+larger data sizes on slow hardware.
 
@@
 of this direction. One strategy will be to start with broad factors
-such as ``Machine'' and refine them with experience. A more difficult
-issue is how to address the problem of testing the whole SPARQL
-language\cite{sparql11query}. We believe that the parameterization work inititated by
-\cite{goerlitz2012splodge} with their SPLODGE system is a key
-ingredient. They parameterized SPARQL queries in such a way they could
-be autogenerated. One suggestion is to parameterize queries based on
-the grammar, so that one factor becomes the number of e.g. UNIONs, and
-the levels are chosen. One problem one will soon run into is that
-parts of the language is recursive, e.g. a GroupGraphPattern can
-consist of several GroupGraphPatterns. However, we believe that
-pragmatic limits can be set, it is for example not of practical
-interest to nest OPTIONALs to any great depth even if it is allowed by
-the language.
+such as ``Machine'' and refine them with experience. Also, the data
+must be parameterized with many more factors than just the number of
+triples; data heterogenity \cite{Duan:2011:AOC:1989323.1989340} is one
+example, but also queries over highly skewed data is important. A more
+difficult issue is how to address the problem of testing the whole
+SPARQL language\cite{sparql11query}. We believe that the
+parameterization work inititated by \cite{goerlitz2012splodge} with
+their SPLODGE system is a key ingredient. They parameterized SPARQL
+queries in such a way they could be autogenerated. One suggestion is
+to parameterize queries based on the grammar, so that one factor
+becomes the number of e.g. UNIONs, and the levels are chosen. One
+problem one will soon run into is that parts of the language is
+recursive, e.g. a GroupGraphPattern can consist of several
+GroupGraphPatterns. However, we believe that pragmatic limits can be
+set, it is for example not of practical interest to nest OPTIONALs to
+any great depth even if it is allowed by the language.
 
@@
 significant effects, they do not necessarily work well to find a
-model, as exemplified by failure to identify the probable non-linear
+model, as exemplified by failure to identify the possibly non-linear
 effect of the P-ATA disks. This issue would be fixed by using an
 orthogonal array design.
+
+Also, experiments as small as the 32-run is not useful in estimating
+any detailed characteristics and is thus of little use for science,
+but could be useful in engineering: If the experiment has been
+validated by a larger experiment, it could serve well in a Continuous
+Integration system.
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 %\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}
-%\bibliographystyle{splncs03}
+\bibliographystyle{splncs03}
 \bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance,specs}

### 2013-05-09 exp1/iswc2013.tex
we can do t-test on 32
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@546 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The resulting normal plot is in figure~\ref{fig:frac32normal}. We see that
-only two main effects are significant on a $\alpha = 0.05$ level
+The resulting normal plot is in figure~\ref{fig:frac32normal}. We see
+that only two main effects are significant on a $\alpha = 0.05$ level
 according to the Lenth criterion, ``Implement'' is again deemed a
@@
 also deviate from the straight line, but with as few runs as this, the
-total variance is great, so no other conclusions can be drawn. Also,
-the Student's t-test we used in the previous section cannot be
-employed, as we have no additional data to use for the
-test.\todo{actually, we can}
+total variance is great, so no other conclusions can be drawn. We may
+use the t-test we used in the previous section, as we have only two
+active factors and so we may treat it as a $2^2$ experiment with 8
+repetitions, but we must expect a higher p-value due to the larger
+variance.
 

### 2013-05-09 exp1/iswc2013.tex
more on linearity problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@547 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsection{Full factorial experiment}
+\subsection{Full factorial experiment}\label{sec:full}
 
@@
 in fact, it hints that our optimization may not work as well for
-larger databases. This should cause some alarm. 
-It is a fundamental property of two-level experiments that it can only
-model linear effects. If the effect is truly linear, and the variance
-is small, the two levels can be quite close together and still yield a
-reasonable estimate for the effect, but the effect will be greater
-with greater separation. Additionally if the effect is non-linear, the
-choice of levels is critical. In this case, the experimenter should be
-concerned that the separation between levels for the sizes of the
-data-set is too small and should further investigate. Such reasoning
-should be applied to all significant effects.
+larger databases. This is a clear indication that the separation
+between levels for the sizes of the data-set is too small and should
+be investigated further. Such reasoning should be applied to all
+significant effects.
 
@@
 
+It is a fundamental property of two-level experiments that it can only
+model linear effects. Usually, it is quite sufficient to identify
+significant effects even though the underlying response may be
+non-linear. As we saw in section~\ref{sec:full}, the choice of levels
+may be critical, and if non-linearity is expected, more than two
+levels must be considered, or at the very least, attention must be
+paid to the choice of levels.\todo{this is a long discussion, how much
+  should be done?}
+
 \paragraph{Summary and hypothesis test}

### 2013-05-09 exp1/iswc2013.tex
lurking variables and randomization
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@548 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 account that the server may cache all or parts of the result set. The
-reasons for this choice will be discussed later.\todo{lurking
-  variables}
+reasons for this choice will be discussed later.
 
@@
 the following small example instructive. Note that each combination is
-executed only once (i.e. it is \emph{unreplicated}, however, so even a
+executed only once (i.e. it is \emph{unreplicated}), however, so even a
 full factorial experiment compares well to a typical benchmark in
@@
 
-\paragraph{Summary and hypothesis test}
+We have argued that fixing certain parameters is untenable, but we
+have allowed ourselves to disregard caching and warm-up runs. We are
+aware that caching may be important in 4store, as a preliminary
+experiment gave the results in \ref{tab:cacheexp}. We see that the
+response time was significantly lower when an OPTIONAL clause was
+added to the query. As this is not more restrictive query, the most
+likely explanation is that the result of the evaluation of the Basic
+Graph Pattern was cached, so only the OPTIONAL left join was necessary
+to add in the second query.
 
-\paragraph{Strategy for evaluating experiment}\todo{Say something
-  about examining variance} \todo{Say something about drilling down in
-  factors} \todo{Say something about spreading levels}
+\begin{table}[ht]
+\begin{center}
+\caption{Two experiments with and without OPTIONAL clause.}\label{tab:cacheexp}
+\begin{tabular}{ccccccccr}
+  \hline
+``Implement'' & ``TripleC'' & ``Machine'' & ``BGPComp'' & ``Lang'' & ``Range'' & ``Union'' & ``Optional'' & Time \\
+ \hline
+1   &    1  &     1  &     2 &   1 &    1   &  1  & 1 &  53.918049 \\
+1   &    1  &     1  &     2 &   1 &    1   &  1  & 2 &   1.252312 \\
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 
-\paragraph{Fold-over}
+It is \emph{randomization} that makes this permissible, i.e. the order
+of execution is random, so the benefit of caching is likely to apply
+randomly to different runs, for example, had the run order been
+reversed in the previous example, it could have been the query without
+OPTIONAL that would have benefitted. Unless the effect of caching is
+cumulative throughout the experiment (which is possible), the
+randomization will have the effect of reversing the runs randomly. The
+end result is that the effect of so-called ``lurking variables'' such
+as caching, warm-up effects, etc, contributes to total unexplained
+variance of the experiment.
+
+As the unexplained variance of the experiment may cloud important
+effects, notably effects that indicate flaws, it should be kept to a
+minimum. Thus, lurking variables should be turned into factors
+whenever possible, is they degrade the quality of the experiment, but
+in many cases, we can live with them if their contribution to the
+variance is small.
 

### 2013-05-09 exp1/iswc2013.tex
more on conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@549 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsection{Fractional factorial experiment with more data}
+\subsection{Fractional factorial experiment with more data}\label{sec:hugefrac}
 
@@
 SPARQL language\cite{sparql11query}. We believe that the
-parameterization work inititated by \cite{goerlitz2012splodge} with
+parameterization work initiated by \cite{goerlitz2012splodge} with
 their SPLODGE system is a key ingredient. They parameterized SPARQL
@@
 
+As evaluating SPARQL endpoints is inherently difficult a simplistic
+experiment such as the one we designed should not hold up to
+scrutiny. We set out to demonstrate how an experiment could be set up
+using DoE, and show how it can be analyzed. In
+sections~\ref{sec:full}~and~\ref{sec:frac}, we first saw how the
+analysis correctly pointed out the most important effects, under
+assumptions dictated by the factors and levels that were given. We saw
+how the formalism provided a comprehensive view of the
+experiment. Then, we saw that the formalism pointed out weaknesses
+that could invalidate the experiment, and in
+section~\ref{sec:hugefrac} we saw that the experiment did indeed not
+hold up to scrutiny.
+
 We saw that while 2 level experiments can perform well in determining
@@
 
-Feasible doing millions of runs.
+We have seen that we can perform a proper hypothesis test based on
+summary statistics, and by using our expertise, we showed how to show
+that the experiment was flawed, and we have cautioned that failure to
+control unexplained variance may compromize our ability to do so. We
+have also pointed out how this direction can provide more rigorous
+evaluation practices.
+
+Even if the complexity is great, with properly tuned endpoints, it is
+feasible to do millions of runs, and with orthogonoal arrays, this
+formalism can be extended to many different evaluation problems.
 

### 2013-05-09 exp1/iswc2013.tex
cosmetics
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@550 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 establish a critical practice of evaluations. It is of critical
-importance that we understand why some effects are critical.\todo{add
-  table of effects} That
-``Implement'' is significant is hardly a surprise, that is what we
-worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
-are strong is also due to that these are fairly demanding things to
-evaluate. Since the delay we inserted affects all kinds of joins, it
-is also intuitive that interactions between ``Implement'' and those
-that require joins are strong. The strong ``BGPComp:Lang'' and
-``Lang:Union'' interactions might be due to that the many more triples
-needs to be searched for a language tag in one case. However, the
-positive ``Implement:TripleC'' interaction evades such explanations;
-in fact, it hints that our optimization may not work as well for
-larger databases. This is a clear indication that the separation
-between levels for the sizes of the data-set is too small and should
-be investigated further. Such reasoning should be applied to all
+importance that we understand why some effects are
+significant.\todo{add table of effects} That ``Implement'' is
+significant is hardly a surprise, that is what we worked for. That the
+main effects ``Union'', ``Optional'' and ``Lang'' are strong is also
+due to that these are fairly demanding things to evaluate. Since the
+delay we inserted affects all kinds of joins, it is also intuitive
+that interactions between ``Implement'' and those that require joins
+are strong. The strong ``BGPComp:Lang'' and ``Lang:Union''
+interactions might be due to that the many more triples needs to be
+searched for a language tag in one case. However, the positive
+``Implement:TripleC'' interaction evades such explanations; in fact,
+it hints that our optimization may not work as well for larger
+databases. This is a clear indication that the separation between
+levels for the sizes of the data-set is too small and should be
+investigated further. Such reasoning should be applied to all
 significant effects.
@@
 Kjetil Kjernsmo would like to thank his main supervisor Martin Giese
-for kind assistance in writing this paper. He would also like to
+for kind assistance. He would also like to
 thank Steve Harris for his generous support in degrading his excellent

### 2013-05-09 exp1/generate-figures.R
Changes to the generated figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@551 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -5,22 +5,25 @@ load("RelevantExperiments.RData")
 
 source("analysis.R")
 
+
 pdf(file="fullnormal.pdf")
-DanielPlot(runbgplangf1.lm)
+par(mai=c(0.6,0.6,0.05,0.05))
+DanielPlot(runbgplangf1.lm, main="", cex.fac=0.8)
 dev.off()
 
-pdf(file="frac32normal.pdf")
-DanielPlot(frac32r1.lm)
+pdf(file="frac32normal.pdf", height=3)
+par(mai=c(0.6,0.6,0.05,0.05))
+DanielPlot(frac32r1.lm, main="", cex.fac=0.8)
 dev.off()
 
-pdf(file="frac64normalalpha005.pdf")
-DanielPlot(frac64r3.lm)
+pdf(file="frac64normal.pdf", height=5)
+par(mai=c(0.6,0.6,0.05,0.05))
+DanielPlot(frac64r3.lm, alpha=c(0.05, 0.15), main="", cex.fac=0.8)
 dev.off()
 
-pdf(file="frac64normalalpha015.pdf")
-DanielPlot(frac64r3.lm, alpha=0.15)
-dev.off()
 
 pdf(file="frac64hugenormal.pdf")
-DanielPlot(frac64huger1.lm)
+par(mai=c(0.6,0.6,0.05,0.05))
+DanielPlot(frac64huger1.lm, main="", cex.fac=0.8)
 dev.off()
+

### 2013-05-09 exp1/iswc2013.tex
Use normal plot with two levels
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@552 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 We turn to the 64-run experiment and its normal plot in
-figure~\ref{fig:frac64normalalpha005}. We see that the we get more significant
+figure~\ref{fig:frac64normal}. We see that the we get more significant
 effects, and that they correspond well to those we saw in the full
@@
 worrisome in the full factorial design. The ``Implement:TripleC''
-interaction emerges only for a level $\alpha=0.15$, see
-figure~\ref{fig:frac64normalalpha015}. To ensure that we discover such
+interaction emerges only for a level $\alpha=0.15$. To ensure that we discover such
 cases will be a key challenge in further studies.
@@
 \begin{figure}[ht]
-  \includegraphics[width=\textwidth]{frac64normalalpha005.pdf}
+  \includegraphics[width=\textwidth]{frac64normal.pdf}
   \caption{A normal plot of the Fractional Factorial Experiment with
     64 runs. The labelled points are considered significant by using
-    the Lenth criterion at a level
-    $\alpha=0.05$.}\label{fig:frac64normalalpha005}
-\end{figure}
-\begin{figure}[hb]
-  \includegraphics[width=\textwidth]{frac64normalalpha015.pdf}
-  \caption{A normal plot of the Fractional Factorial Experiment with
-    64 runs. The labelled points are considered significant by using
-    the Lenth criterion at a level
-    $\alpha=0.15$.}\label{fig:frac64normalalpha015}
+    the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
+    and crosses at $\alpha=0.05$.}\label{fig:frac64normal}
 \end{figure}

### 2013-05-10 exp1/iswc2013.tex
add authors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@553 43ef1190-90fd-4dd7-a769-d0121f02b80e
John's smaller changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@554 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   Endpoint Evaluation}
-%\author{Kjetil Kjernsmo\inst{1}}
-%\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
-%\institute{Department of Informatics,
-%Postboks 1080 Blindern,
-%0316 Oslo, Norway
-%\email{kjekje@ifi.uio.no}}
+\author{Kjetil Kjernsmo\inst{1} \and John S. Tyssedal\inst{2}}
+\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
+\institute{Department of Informatics,
+Postboks 1080 Blindern,
+N-0316 Oslo, Norway \email{kjekje@ifi.uio.no} \and Department of Mathematical Sciences,
+The Norwegian Institute of Technology,
+N-7491 Trondheim,
+Norway
+\email{john.tyssedal@math.ntnu.no}}
 
@@
 loads on the system to measure its performance as a function of the
-load. For SPARQL benchmarks, usually the performance is measured as a
+load. For SPARQL benchmarks, usually the performance is measured as 
 the number of queries the SPARQL implementation is able to answer in a
@@
 
-However, there are many of problems with this approach when used to
+However, there are many problems with this approach when used to
 evaluate assertions about the overall performance of a system or
@@
 While this is an improvement which is possibly sufficient for
-industrial uses, is it still flawed as empirical
+industrial uses, it is still flawed as empirical
 research, as it cannot capture the full diversity of the problem: For
@@
 not found the parts discussed in the present paper to be widely
-adopted. Recently, an comprehensive text on experimental methods have
+adopted. Recently, a comprehensive text on experimental methods have
 been published in
@@
 
-However, as a real optimization is beyond the scope of this paper, we
-have instead created a simulation of an optimization. This has also
-the advantage that we can \textit{a priori} understand the effect of
-the changes. In a real-world case, there is often the case that an
+A real optimization is beyond the scope of this paper. Instead we have
+performed a simulation that enable us to understand the effect of the
+changes.  
+In a real-world case, there is often the case that an
 optimization has negative side-effects, and we would need to simulate
@@
 called a $2^8$ factorial experiment, and we can see that the number of
-runs is grows exponentially with the number of factors, thus making
+runs grows exponentially with the number of factors, thus making
 large experiments prohibitively expensive. We shall nevertheless find
@@
 enhances the performance of the SPARQL endpoint, the runtime
-decreases. If ``Implement'' and the sum of its interactions are
-negative, then it supports the alternative hypothesis, i.e. we have
-successfully improved the performance.
+decreases. If ``Implement'' and its interactions are negative conditional on that
+the factors involved with its interactions are on their high level,
+then it supports the alternative hypothesis, i.e. we have successfully
+improved the performance.
 
@@
 significant effects. Visually, this is not as useful, since the large
-number\todo{how to fix this...?} We see that we have very significant
+number\todo{how to fix this...?} We see that we have many very significant
 effects. We also note that ``Implement'' itself is significantly
@@
 control and environmental factors. This enables us to treat our
-experiment as a $2^5$ experiment replicated 8 times.
+experiment as a $2^5$ experiment replicated 8 times.\todo{expand on this}
 
@@
 
-It is a fundamental property of two-level experiments that it can only
-model linear effects. Usually, it is quite sufficient to identify
-significant effects even though the underlying response may be
-non-linear. As we saw in section~\ref{sec:full}, the choice of levels
-may be critical, and if non-linearity is expected, more than two
-levels must be considered, or at the very least, attention must be
-paid to the choice of levels.\todo{this is a long discussion, how much
-  should be done?}
+It is a fundamental property of two-level experiments that they can
+only model linear effects, however, they are usually quite sufficient
+for identifying significant effects even though the underlying
+response may be non-linear.  As we saw in section~\ref{sec:full}, the
+choice of levels may be critical, and if non-linearity is expected,
+more than two levels must be considered, or at the very least,
+attention must be paid to the choice of levels.\todo{this is a long
+  discussion, how much should be done?}
 
@@
 allow for different numbers of levels, for much greater flexibility in
-total run size, and for irregular designs that can provide a good fit
+total run size, and for non-regular designs that can provide a good fit
 for the complex problem of SPARQL endpoint evaluations.

### 2013-05-10 exp1/FrF2-mod/DanielPlot.R
Check in parts of FrF2 to edit
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@555 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,0 +1,223 @@
+DanielPlot <- function(fit, ...){
+    UseMethod("DanielPlot")
+}
+DanielPlot.design <- function(fit, ..., response=NULL){
+    if (!"design" %in% class(fit)) 
+        stop("DanielPlot.design works for obj from class design only.")
+    di <- design.info(fit)
+    if (is.null(di$response.names)) 
+        stop("The design fit must have at least one response.")
+    if (!(is.null(response))) 
+      if (!response %in% di$response.names)
+        stop("Requested response is not a response variable in fit.")
+    if (!(length(grep("FrF2",di$type))>0 | 
+           length(grep("pb",di$type))>0)) { 
+           if (!(di$type=="full factorial" & all(di$nlevels==2)))
+        stop("The design obj must be of a type containing FrF2 or pb.")
+       }
+    grad <- 1
+    if (length(grep("pb",di$type)) > 0 & di$nfactors < di$nruns-1)
+          warning("Effects plots for Plackett-Burman designs must be done with nruns-1 effects! The error effects are missing!")
+    ## make sure there are as many effects as possible in the plots, redundant ones will not be shown
+    if (length(grep("FrF2",di$type)) > 0 | (di$type=="full factorial" & all(di$nlevels==2))){
+       grad <- 2
+       hilf <- lm(fit, degree=grad)
+       ncoef <- sum(!is.na(coef(hilf)))
+       nhilf <- di$nrun
+       if (!is.null(di$ncenter)) nhilf <- nhilf - di$ncenter
+       while (ncoef < nhilf){
+          grad <- grad+1
+          hilf <- lm(fit, degree=grad)
+          ncoef <- sum(!is.na(coef(hilf)))
+          }
+    }
+    subtext <- ""
+    if (length(grep("splitplot", di$type)) > 0){
+        ## splitplot situation
+        ## make sure to use different plot symbols for whole plot and split plot effects
+        ## important to distinguish wbreps and bbreps
+               mm <- model.matrix(hilf)
+               #coefs <- stats:::coef.default(fit)
+               coefs <- coef(hilf)[-1] ## only non-missing coefficients
+               nms <- names(coefs)
+               hilf <- apply(mm[,1+(1:di$nfac.WP),drop=FALSE],1,paste,collapse="")
+               pchs <- rep("*", length(coefs))
+               pchs[1:di$nfac.WP] <- "o"
+               
[…]

### 2013-05-10 exp1/FrF2-mod/DanielPlot.R
My change to DanielPlot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@556 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -83,11 +83,18 @@ function (fit, code = FALSE, autolab = TRUE, alpha=0.05,
     factor.effects <- factor.effects[!is.na(factor.effects)]
     plotmain <- paste("Normal Plot for", respnam)
     if (autolab) {
-        crit <- LenthPlot(factor.effects,alpha=alpha,plt=FALSE)["ME"]
+        crit <- sort(sapply(alpha, function(a) { LenthPlot(factor.effects,alpha=a,plt=FALSE)["ME"] }))
         if (!code)
-        faclab <- list(idx = which(crit<=abs(factor.effects)),
-               lab = names(factor.effects)[which(crit<=abs(factor.effects))])
-        plotmain <- paste(plotmain, ", ", "alpha=", alpha, sep="") 
+            idx <- which(min(crit)<=abs(factor.effects))
+            faclab <- list(idx = idx,
+                           lab = names(factor.effects)[idx])
+            if (length(alpha) > 1) {
+                pch <- rep(1, length(factor.effects))
+                for(i in 1:length(crit)) {
+                    pch[which(crit[i]<=abs(factor.effects))] <- i+1
+                }  
+            }
+            plotmain <- paste(plotmain, ", ", "alpha=", max(alpha), sep="") 
         }
     if (half) {
         tn <- list(x = qnorm(0.5 * ((rank(abs(factor.effects)) - 

### 2013-05-10 exp1/iswc2013.tex
invalidated due to absence of solution modifiers
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@557 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
-    and crosses at $\alpha=0.05$.}\label{fig:frac64normal}
+    and crosses at $\alpha=0.05$. The use of different symbols have
+    been added to FrF2 by the authors.}\label{fig:frac64normal}
 \end{figure}
@@
 
+Finally, we note that the most problematic cases are those that are
+covered neither by the broad factors, nor the lurking variables, or
+any of the specific factors. We cannot in the general case be sure
+that they are accounted for, so reviewers must remain vigilant that
+some factors are simply ignored. What finally invalidates the present
+experiment is the complete absence of certain obvious language
+features, such as solution modifiers.
+
 \section{Future Work}

### 2013-05-10 exp1/iswc2013.tex
add generators
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@559 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 must be clear and also all two-factor interactions where ``Implement''
-is involved must be clear. \todo{declare generators?}
+is involved must be clear. The resulting experiments is returned to us
+as design matrices, but they can also be described sufficiently for
+reproducability in terms of \emph{design generators}. These also
+declare aliasing relations for some of the main effects. The design
+generators are\\
+\begin{tabular}{ccl}
+``Range'' &=& ``TripleC'' ``Machine'' ``BGPComp'' \\  
+``Union'' &=& ``TripleC'' ``Machine'' ``Lang'' \\  
+``Optional'' &=& ``TripleC'' ``BGPComp'' ``Lang'' ``Implement'' 
+\end{tabular}
+\\and\\
+\begin{tabular}{ccl}
+``Union'' &=& ``Implement''  ``TripleC''  ``Machine'' ``BGPComp'' \\
+``Optional'' &=& ``Implement''  ``TripleC'' ``Lang''  ``Range'' 
+\end{tabular}
+\\for the 32 and 64 run experiments respectively.
 

### 2013-05-10 exp1/iswc2013.tex
Add table of effects in full run
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@560 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
   \includegraphics[width=\textwidth]{fullnormal.pdf}
@@
 estimating significance in an experiment with no replicates. In
+table~\ref{tab:effects}, the most important effects are listed.  In
 figure~\ref{fig:fullnormal}, Lenth's method is used to label the
-significant effects. Visually, this is not as useful, since the large
-number\todo{how to fix this...?} We see that we have many very significant
-effects. We also note that ``Implement'' itself is significantly
-negative. We also note that the ``Implement:Lang'' interaction is
-significantly positive, as expected from our setup, where ``Lang'' was
-the detrimental side-effect we simulated.
+significant effects. We see that we have many very significant effects
+(unfortunately so many some labels are illegible). We also note that
+``Implement'' itself is significantly negative. We also note that the
+``Implement:Lang'' interaction is significantly positive, as expected
+from our setup, where ``Lang'' was the detrimental side-effect we
+simulated.
+
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Fri May 10 16:31:49 2013
+% Modified by hand to remove 1's
+\begin{table}[hb]
+\begin{center}
+\caption{The magnitude of effects for some important main effects and interactions.}\label{tab:effects}
+\begin{tabular}{rr}
+  \hline
+Factors & Effect  \\ 
+  \hline
+Implement & -21.27 \\ 
+  Implement:Optional & -11.49 \\ 
+  Implement:Union & -6.21 \\ 
+  Implement:TripleC:Lang1 & 4.12 \\ 
+  TripleC:Lang & 4.20 \\ 
+  Implement:TripleC & 4.32 \\ 
+  Implement:BGPComp & 4.89 \\ 
+  Lang:Union & 5.16 \\ 
+  Implement:BGPComp:Lang & 5.16 \\ 
+  BGPComp:Lang & 5.25 \\ 
+  Implement:Lang:Union & 5.75 \\ 
+  TripleC & 7.06 \\ 
+  Implement:Lang & 7.73 \\ 
+  Lang & 8.62 \\ 
+  Optional & 11.30 \\ 
+  Union & 17.18 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 
-\begin{figure}[hb]
+\begin{figure}[h!]
   \includegraphics[width=\textwidth]{frac32normal.pdf}

### 2013-05-10 exp1/analysis.R
Let BGPComp be noise by default
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@561 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -11,7 +11,7 @@ fulllm <- function(results) {
 }
 
 robust.simple <- function(model, control = "Implement",
-                   inactive = c("Machine", "BGPComp", "Range"),
+                   inactive = c("Machine", "Range"),
                    pairwise=FALSE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
   allmeans <- aggregate(fm, data=model, mean)
@@ -21,7 +21,7 @@ robust.simple <- function(model, control = "Implement",
 }
 
 robust.pairwise <- function(model, control = "Implement", 
-                            noise = c("TripleC", "Lang", "Union", "Optional")) {
+                            noise = c("TripleC", "BGPComp", "Lang", "Union", "Optional")) {
   fm <- as.formula(paste("~", paste(noise, collapse=" * ")))
   dlply(model, fm, function(subset) {
     t.test(subset[subset[control] == 2,]$experiments,

### 2013-05-10 exp1/iswc2013.tex
BGComp now active in Full
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@563 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 hypothesis test. By inspecting the normal plot in
-figure~\ref{fig:fullnormal}, we see that only five factors are highly
-significant by either being a significant main effect or participate
-in a highly significant interaction. We say that the three factors
-``BGPComp'', ``Range'' and ``Machine'' are \emph{inactive} since they
-do not contribute significantly to the observed variation in
-performance. In practical situations, the four factors ``Lang'',
+figure~\ref{fig:fullnormal} and table~\ref{tab:effects}, we see that
+six factors are highly significant by either being a significant main
+effect or participate in a highly significant interaction. We say that
+the two factors ``Range'' and ``Machine'' are \emph{inactive} since
+they do not contribute significantly to the observed variation in
+performance. ``Machine'' is marginally significant, but the effect is
+so little, it will be of greater use for us in the hypothesis test. In
+practical situations, the five factors ``BGPComp'', ``Lang'',
 ``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
@@
 only change the implementation. In this context, we call ``Implement''
-a \emph{control} factor and the other four active factors
+a \emph{control} factor and the other five active factors
 \emph{environmental}.
 
-The presence of inactive factors makes it possible to use a standard
-t-test by averaging the inactive factors for each combination of
-control and environmental factors. This enables us to treat our
-experiment as a $2^5$ experiment replicated 8 times.\todo{expand on this}
+We would like an overall test to see whether the new implementation is
+better than the old, and the presence of inactive factors makes it
+possible to average the performance of the new and the old
+implementations into two distinct vectors. We do this by creating an
+average over all the 32 level-combinations of the five active
+environmental factors. This leaves us with 4 values for each of the
+two levels of ``Implement'' and a two-sample, t-test can be
+performed. Effectively, we treat our experiment as a $2^6$ experiment
+replicated 4 times.
 
-We then extract two vectors, one for the each of the implementations,
-and then perform a one-sided hypothesis test where $H_0$ is that they
-are equal and $H_1$ is that the vector representing the new
-implementation is smaller. In this example, we find that the available
-data supports the assertion that the new implementation is better with a
-high probability, $p=2.9 \cdot 10^{-10}$.\todo{rationale could be
-  stronger}\todo{figure?}
+We perform the test as a one-sided hypothesis test where $H_0$ is that
+the vectors are equal and $H_1$ is that the vectors representing the
+new implementation is smaller. In this example, we find that the
+available data supports the assertion that the new implementation is
+better with a high probability, $p=1.16 \cdot 10^{-07}$.\todo{figure?}
 
@@
 there are cases that it fails. Again, we treat the experiment as a
-$2^5$ experiment replicated 8 times. Now, for each of the 16 resulting
+$2^6$ experiment replicated 4 times. Now, for each of the 32 resulting
 combinations of the environmental factors for each level of
-``Implement'', we extract 8 values. We may now run a t-test for each
+``Implement'', we extract 4 values. We may now run a t-test for each
 of the replications, as above. The results are tabulated in
@@
 \caption{p-values for different parts of the experiment}\label{tab:pvaluesfull}
-\begin{tabular}{ccccl}
+\begin{tabular}{cccccl}
   \hline
-``TripleC'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
+``TripleC'' & ``BGPComp'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
   \hline
-  1 & 1 & 1 & 1 & 0.00032 \\ 
-  1 & 1 & 1 & 2 & $2 \cdot 10^{-14}$ \\ 
-  1 & 1 & 2 & 1 & $4.7 \cdot 10^{-14}$ \\ 
-  1 & 1 & 2 & 2 & $2 \cdot 10^{-18}$ \\ 
-  1 & 2 & 1 & 1 & $1.4 \cdot 10^{-06}$ \\ 
-  1 & 2 & 1 & 2 & $8.3 \cdot 10^{-11}$ \\ 
-  1 & 2 & 2 & 1 & 0.0017 \\ 
-  1 & 2 & 2 & 2 & $2.1 \cdot 10^{-06}$ \\ 
-  2 & 1 & 1 & 1 & 0.09 \\ 
-  2 & 1 & 1 & 2 & $1.6 \cdot 10^{-09}$ \\ 
-  2 & 1 & 2 & 1 & $8.4 \cdot 10^{-10}$ \\ 
-  2 & 1 & 2 & 2 & $2.5 \cdot 10^{-13}$ \\ 
-  2 & 2 & 1 & 1 & 0.29 \\ 
-  2 & 2 & 1 & 2 & $9.1 \cdot 10^{-07}$ \\ 
-  2 & 2 & 2 & 1 & 0.92 \\ 
-  2 & 2 & 2 & 2 & 0.071 \\ 
+  1 & 1 & 1 & 1 & 1 & 0.012 \\ 
+  1 & 1 & 1 & 1 & 2 & $1.4 \cdot 10^{-09}$ \\ 
+  1 & 1 & 1 & 2 & 1 & $3.1 \cdot 10^{-09}$ \\ 
+  1 & 1 & 1 & 2 & 2 & $6.1 \cdot 10^{-11}$ \\ 
+  1 & 1 & 2 & 1 & 1 & $3.3 \cdot 10^{-06}$ \\ 
+  1 & 1 & 2 & 1 & 2 & $2.7 \cdot 10^{-09}$ \\ 
+  1 & 1 & 2 & 2 & 1 & $2 \cdot 10^{-06}$ \\ 
+  1 & 1 & 2 & 2 & 2 & $3.6 \cdot 10^{-10}$ \\ 
+  1 & 2 & 1 & 1 & 1 & 0.014 \\ 
+  1 & 2 & 1 & 1 & 2 & $1.2 \cdot 10^{-10}$ \\ 
+  1 & 2 & 1 & 2 & 1 & $2.8 \cdot 10^{-14}$ \\ 
+  1 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
+  1 & 2 & 2 & 1 & 1 & $2.1 \cdot 10^{-05}$ \\ 
+  1 & 2 & 2 & 1 & 2 & $2.7 \cdot 10^{-07}$ \\ 
+  1 & 2 & 2 & 2 & 1 & 0.0072 \\ 
+  1 & 2 & 2 & 2 & 2 & $1.6 \cdot 10^{-05}$ \\ 
+  2 & 1 & 1 & 1 & 1 & 0.28 \\ 
+  2 & 1 & 1 & 1 & 2 & $3 \cdot 10^{-07}$ \\ 
+  2 & 1 & 1 & 2 & 1 & $3.3 \cdot 10^{-07}$ \\ 
+  2 & 1 & 1 & 2 & 2 & $1.7 \cdot 10^{-08}$ \\ 
+  2 & 1 & 2 & 1 & 1 & 0.0023 \\ 
+  2 & 1 & 2 & 1 & 2 & $6.5 \cdot 10^{-07}$ \\ 
+  2 & 1 & 2 & 2 & 1 & 0.00032 \\ 
+  2 & 1 & 2 & 2 & 2 & $1.3 \cdot 10^{-06}$ \\ 
+  2 & 2 & 1 & 1 & 1 & 0.013 \\ 
+  2 & 2 & 1 & 1 & 2 & $1 \cdot 10^{-11}$ \\ 
+  2 & 2 & 1 & 2 & 1 & $3.8 \cdot 10^{-11}$ \\ 
+  2 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
+  2 & 2 & 2 & 1 & 1 &   1 \\ 
+  2 & 2 & 2 & 1 & 2 & $2.3 \cdot 10^{-05}$ \\ 
+  2 & 2 & 2 & 2 & 1 &   1 \\ 
+  2 & 2 & 2 & 2 & 2 & 0.99 \\ 
    \hline

### 2013-05-10 exp1/iswc2013.tex
add some more p-values
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@564 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 active factors and so we may treat it as a $2^2$ experiment with 8
-repetitions, but we must expect a higher p-value due to the larger
-variance.
+replications, but the resulting $p = 0.00098$ must be expected with
+such a small number of runs and thus higher variance.
 
@@
 
-Since the 64-run experiment also has 3 inactive factors, ``Machine'',
+Since the 64-run experiment has 3 inactive factors, ``Machine'',
 ``TripleC'' and ``Range'', we can again treat it as a $2^5$ experiment
@@
 again, the hypothesis test indicates that the new implementation is an
-improvement at $p = 5.9 \cdot 10^{-6}$. The comparatively higher p-value
+improvement at $p = 5.9 \cdot 10^{-6}$. Again, the comparatively higher p-value
 is due to the lower number of runs. It is also possible to perform the
-per-environmental test, but we refrain from that in the interest of
-brevity.\todo{good way to say it?}
+per-environmental test.
 

### 2013-05-10 exp1/iswc2013.tex
bothersome questions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@565 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+To this end, the following questions must be asked:
+\begin{enumerate}
+\item Are there factors that cover all realistic features?
+\item If not, are they adequately covered by randomization?
+\item If so, would the variance resulting from randomization obscure
+  factors that could provide clues that the levels are wrongly set?
+\item By carefully examining interactions with ``Implement'', are
+  there any that are unaccounted for, and that could point out wrongly
+  set levels?
+\end{enumarate}
+
 Even if the complexity is great, with properly tuned endpoints, it is

### 2013-05-10 exp1/generate-figures.R
add some code for pairwise table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@567 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -3,8 +3,9 @@ library(plyr)
 
 load("RelevantExperiments.RData")
 
-source("analysis.R")
-
+source("FrF2-mod/remodel.R")
+source("FrF2-mod/check.R")
+source("FrF2-mod/DanielPlot.R")
 
 pdf(file="fullnormal.pdf")
 par(mai=c(0.6,0.6,0.05,0.05))
@@ -27,3 +28,10 @@ par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64huger1.lm, main="", cex.fac=0.8)
 dev.off()
 
+#sign <- LenthPlot(runbgplangf1.lm, alpha=0.05, plt=F)["ME"]
+sign <- 2 # We haven't got space for more
+coefs <- coef(runbgplangf1.lm)
+sigcoefs <- sort(coefs[abs(coefs) > sign])*2
+sigcoefs2 <- sigcoefs[names(sigcoefs) != "(Intercept)"]
+
+lapply(robust.pairwise(runbgplangf1), function(x) x$p.value) # Lots of manual work after this

### 2013-05-10 exp1/generate-figures.R
fixing the size of figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@568 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -23,7 +23,7 @@ DanielPlot(frac64r3.lm, alpha=c(0.05, 0.15), main="", cex.fac=0.8)
 dev.off()
 
 
-pdf(file="frac64hugenormal.pdf")
+pdf(file="frac64hugenormal.pdf", height=5.5)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64huger1.lm, main="", cex.fac=0.8)
 dev.off()

### 2013-05-10 exp1/iswc2013.tex
BGPComp is also broad
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@566 43ef1190-90fd-4dd7-a769-d0121f02b80e
use paragraph for acknowledgements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@569 43ef1190-90fd-4dd7-a769-d0121f02b80e
remove todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@571 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{graphicx}
-\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 four 4store instances runs independently on different
-ports. \todo{possible problem?}
+ports.
 
@@
 In the full factorial experiment, all combinations of the 8 factors
-are executed, in total 256 runs.\todo{concrete example?} This is
-called a $2^8$ factorial experiment, and we can see that the number of
-runs grows exponentially with the number of factors, thus making
-large experiments prohibitively expensive. We shall nevertheless find
-the following small example instructive. Note that each combination is
-executed only once (i.e. it is \emph{unreplicated}), however, so even a
-full factorial experiment compares well to a typical benchmark in
-which each microbenchmark must be executed several times to obtain
-enough data to compute a mean with reasonable certainty.
+are executed, in total 256 runs. This is called a $2^8$ factorial
+experiment, and we can see that the number of runs grows exponentially
+with the number of factors, thus making large experiments
+prohibitively expensive. We shall nevertheless find the following
+small example instructive. Note that each combination is executed only
+once (i.e. it is \emph{unreplicated}), however, so even a full
+factorial experiment compares well to a typical benchmark in which
+each microbenchmark must be executed several times to obtain enough
+data to compute a mean with reasonable certainty.
 
@@
 curl measurements. To analyze the experiment, we first generate a
-normal plot, see figure~\ref{fig:fullnormal}\todo{explain axes}. In this plot, if the
+normal plot, see figure~\ref{fig:fullnormal}. In this plot, if the
 null hypothesis $H_0$ holds, all estimated effects will fall on a
@@
 available data supports the assertion that the new implementation is
-better with a high probability, $p=1.16 \cdot 10^{-07}$.\todo{figure?}
+better with a high probability, $p=1.16 \cdot 10^{-07}$.
 
@@
 establish a critical practice of evaluations. It is of critical
-importance that we understand why some effects are
-significant.\todo{add table of effects} That ``Implement'' is
-significant is hardly a surprise, that is what we worked for. That the
-main effects ``Union'', ``Optional'' and ``Lang'' are strong is also
-due to that these are fairly demanding things to evaluate. Since the
-delay we inserted affects all kinds of joins, it is also intuitive
-that interactions between ``Implement'' and those that require joins
-are strong. The strong ``BGPComp:Lang'' and ``Lang:Union''
-interactions might be due to that the many more triples needs to be
-searched for a language tag in one case. However, the positive
-``Implement:TripleC'' interaction evades such explanations; in fact,
-it hints that our optimization may not work as well for larger
-databases. This is a clear indication that the separation between
-levels for the sizes of the data-set is too small and should be
-investigated further. Such reasoning should be applied to all
+importance that we understand why some effects are significant. That
+``Implement'' is significant is hardly a surprise, that is what we
+worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
+are strong is also due to that these are fairly demanding things to
+evaluate. Since the delay we inserted affects all kinds of joins, it
+is also intuitive that interactions between ``Implement'' and those
+that require joins are strong. The strong ``BGPComp:Lang'' and
+``Lang:Union'' interactions might be due to that the many more triples
+needs to be searched for a language tag in one case. However, the
+positive ``Implement:TripleC'' interaction evades such explanations;
+in fact, it hints that our optimization may not work as well for
+larger databases. This is a clear indication that the separation
+between levels for the sizes of the data-set is too small and should
+be investigated further. Such reasoning should be applied to all
 significant effects.
@@
 its relevance and break it up in subsequent experiments if deeper
-understanding is needed. In this case, the cause of the performance
+understanding is needed. To some extent, ``BGPComp'' is also a broad
+factor, since it encompasses many structures that is normally seen in
+a Basic Graph Pattern. In this case, the cause of the performance
 problem was highly audible: It was the old P-ATA disks of the larger
@@
 more than two levels must be considered, or at the very least,
-attention must be paid to the choice of levels.\todo{this is a long
-  discussion, how much should be done?}
+attention must be paid to the choice of levels.
 
@@
   set levels?
-\end{enumarate}
+\end{enumerate}
 
@@
 
-\section*{Acknowledgments}
+\paragraph*{Acknowledgments}
 

### 2013-05-10 exp1/iswc2013.tex
reformat .tex
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@572 43ef1190-90fd-4dd7-a769-d0121f02b80e
fixes after paper readthrough
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@573 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{itemize}
-\item As the complexity increases, the number of parameters that
-one may wish to test increases dramatically.
-\item Nothing compels anyone to find microbenchmarks that
-  can refute an assertion, and there is no structured approach to
-  investigate whether a benchmark is flawed.
+\item As the complexity increases, the number of parameters that one
+  may wish to test increases dramatically.
+\item Nothing compels anyone to find microbenchmarks that can refute
+  an assertion, and there is no structured approach to investigate
+  whether a benchmark is flawed.
 \item There is no meaningful summary of the performance as a whole, so
@@
 While this is an improvement which is possibly sufficient for
-industrial uses, it is still flawed as empirical
-research, as it cannot capture the full diversity of the problem: For
-example, if a vendor claims that their implementation is better if
-used on a highly capable hardware platform, then the standardization
-of a hardware platform cannot test this assertion. Moreover, to test
-this assertion, one would need a summary technique that can identify
-the interaction between implementation and hardware platform and
-possibly other parameters.
+industrial uses, it is still flawed as empirical research, as it
+cannot capture the full diversity of the problem: For example, if a
+vendor claims that their implementation is better if used on a highly
+capable hardware platform, then the standardization of a hardware
+platform cannot test this assertion. Moreover, to test this assertion,
+one would need a summary technique that can identify the interaction
+between implementation and hardware platform and possibly other
+parameters.
 
@@
 Thus, attacking the complexity problem by eliminating parameters we
-think influence the outcome for simplicity is untenable as
-scientific practice. We must instead tackle this problem by enhancing
-our capability of testing complex systems and to identify situations
-where our assumptions are flawed. Such constraints can only be made
-when we have very good reason to suspect that a certain parameter does
-not have any influence.
+think influence the outcome for simplicity is untenable as scientific
+practice. We must instead tackle this problem by enhancing our
+capability of testing complex systems and to identify situations where
+our assumptions are flawed. Such constraints can only be made when we
+have very good reason to suspect that a certain parameter does not
+have any influence.
 
@@
 
-A literature survey as not revealed any direct prior art, however, the
+A literature survey has not revealed any direct prior art, however, the
 general approach has been well established, not only in statistics. A
@@
 
-The problematic sides of the benchmarking have been noted in several
+Some problematic sides of benchmarking have been noted in several
 papers, notably \cite{Duan:2011:AOC:1989323.1989340} and
-\cite{MontoyaVCRA12}. We also acknowledge that great
-progress has been made to improve benchmarks, in particular, we are
-indebted to \cite{mxro:Morsey2011DBpedia}.
+\cite{MontoyaVCRA12}. We also acknowledge that great progress has been
+made to improve benchmarks, in particular, we are indebted to
+\cite{mxro:Morsey2011DBpedia}.
 
-We believe that automatic creation of benchmark queries, as
-pioneered by \cite{goerlitz2012splodge} is a critical ingredient for
-the application of DoE to be successful as a methodology.
+We believe that automatic creation of benchmark queries, as pioneered
+by \cite{goerlitz2012splodge} is a critical ingredient for the
+application of DoE to be successful as a methodology.
 
@@
 much simpler than what has been used in benchmarking. We have used the
-data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
-smallest of their data-sets (that we found to be more than 15
-MTriples). Moreover, we have taken subsets of that data-set by using
-the last 1 or 2 MTriples from their file.
+data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but
+only consider the smallest of their data-sets (that we found to be
+more than 15 MTriples). Moreover, we have taken subsets of that
+data-set by using the last 1 or 2 MTriples from their file.
 
@@
 have chosen in this paper to only deal with the relatively
-straightforward aspects of the DoE formalism, and therefore 
-constrain the experiment strictly to 2 levels.
+straightforward aspects of the DoE formalism, and therefore constrain
+the experiment strictly to 2 levels.
 
-We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
-data-set, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
-where one machine runs GNU/Linux Debian Squeeze, has 16 GB RAM, an
-Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
-configuration. The other machine runs Debian Wheezy, has 8 GB RAM, an
-Intel Core i7-3520M CPU and a single SSD on SATA-III.
+We have chosen 8 factors each having 2 levels, ``TripleC'' the number
+of triples in the data-set, 1 or 2 MTriples, ``Machine'', which is the
+software and hardware platform, where one machine runs GNU/Linux
+Debian Squeeze, has 16 GB RAM, an Intel Core2 Duo E8500 CPU and two
+Parallel-ATA disks in a RAID-1 configuration. The other machine runs
+Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
+SSD on SATA-III.
 
-Then, we test some language features: 
-``BGPComp'', which is a Basic Graph Pattern of varying complexity. The
-first level is:
+Then, we test some language features: ``BGPComp'', which is a Basic
+Graph Pattern of varying complexity. The first level is:
 \verbatiminput{experiment/BGPComp-1}
@@
 
-The following factors tests the absence (at level 1) or presence (at level 2) of the following clauses:
-The ``Lang'' factor is this FILTER:
+The following factors tests the absence (at level 1) or presence (at
+level 2) of the following clauses: The ``Lang'' factor is this FILTER:
 \verbatiminput{experiment/Lang-2}
@@
 ``TripleC'' and ``Implement'' specifies a port number. On each host,
-four 4store instances runs independently on different
-ports.
+four 4store instances runs independently on different ports.
 
@@
 performed a simulation that enable us to understand the effect of the
-changes.  
-In a real-world case, there is often the case that an
+changes.  In a real-world case, there is often the case that an
 optimization has negative side-effects, and we would need to simulate
@@
 tools for DoE has been implemented by the R community in packages
-called DoE.base\cite{doebase} and FrF2\cite{frf2}. The experiments are run on a third
-computer in a 1 gigabit Ethernet LAN with the two experiment hosts.  As
-response variable, we have chosen to use the time from the namespace
-server lookup finishes to the data has been transferred. This choice is
-somewhat arbitrary, many other response variables could be chosen, and
-indeed, future work should explore multi-variable responses, but for
-simplicity, we think this reflects the total amount of work done by
-the SPARQL endpoint well. For the measurements, we have chosen to use
-the RCurl package\cite{lang2007r}, which is a thin wrapper around the curl
-library. Curl has well developed facilities for timing requests and
-responses, and so we rely on its measurements.
+called DoE.base\cite{doebase} and FrF2\cite{frf2}. The experiments are
+run on a third computer in a 1 gigabit Ethernet LAN with the two
+experiment hosts.  As response variable, we have chosen to use the
+time from the namespace server lookup finishes to the data has been
+transferred. This choice is somewhat arbitrary, many other response
+variables could be chosen, and indeed, future work should explore
+multi-variable responses, but for simplicity, we think this reflects
+the total amount of work done by the SPARQL endpoint well. For the
+measurements, we have chosen to use the RCurl package\cite{lang2007r},
+which is a thin wrapper around the curl library. Curl has well
+developed facilities for timing requests and responses, and so we rely
+on its measurements.
 
@@
 consisting of the factors with a column containing the corresponding
-curl measurements. To analyze the experiment, we first generate a
-normal plot, see figure~\ref{fig:fullnormal}. In this plot, if the
-null hypothesis $H_0$ holds, all estimated effects will fall on a
-straight line, as their means would be zero. Any significant departure
-from a straight line can be interpreted as a significant effect (see
-Section~4.8 of \cite{wu2009experiments} for a detailed
+curl measurements. To analyze the experiment, we start by fitting a
+linear model to the data with all factors and their interactions. It
+is shown in \cite{wu2009experiments} that the effect, i.e. the change
+from level 1 to 2 is twice the coefficients of the fit. Then, we
+generate a normal plot, see figure~\ref{fig:fullnormal}. In this plot,
+if the null hypothesis $H_0$ holds, all estimated effects will fall on
+a straight line, as their means would be zero. Any significant
+departure from a straight line can be interpreted as a significant
+effect (see Section~4.8 of \cite{wu2009experiments} for a detailed
 explanation). In our case, it is most important to note that any
-effect that is negative means the effect
-enhances the performance of the SPARQL endpoint, the runtime
-decreases. If ``Implement'' and its interactions are negative conditional on that
-the factors involved with its interactions are on their high level,
-then it supports the alternative hypothesis, i.e. we have successfully
-improved the performance.
+effect that is negative means the effect enhances the performance of
+the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
+interactions are negative conditional on that the factors involved
+with its interactions are on their high level, then it supports the
+alternative hypothesis, i.e. we have successfully improved the
+performance.
 
@@
 significant effects. We see that we have many very significant effects
-(unfortunately so many some labels are illegible). We also note that
+(unfortunately so many that some labels are illegible). We also note that
 ``Implement'' itself is significantly negative. We also note that the
@@
 performance. ``Machine'' is marginally significant, but the effect is
-so little, it will be of greater use for us in the hypothesis test. In
+so little, it will be of greater use for us in the following procedure. In
 practical situations, the five factors ``BGPComp'', ``Lang'',
@@
 environmental factors. This leaves us with 4 values for each of the
-two levels of ``Implement'' and a two-sample, t-test can be
+two levels of ``Implement'' and a two-sample t-test can be
 performed. Effectively, we treat our experiment as a $2^6$ experiment
@@
 We perform the test as a one-sided hypothesis test where $H_0$ is that
-the vectors are equal and $H_1$ is that the vectors representing the
-new implementation is smaller. In this example, we find that the
-available data supports the assertion that the new implementation is
-better with a high probability, $p=1.16 \cdot 10^{-07}$.
+the mean of the vectors are equal and $H_1$ is that mean of the
+vectors representing the new implementation is lower. In this example,
+we find that the available data supports the assertion that the new
+implementation is better with a high probability, $p=1.16 \cdot
+10^{-07}$.
 
@@
 As mentioned, the full factorial experiment goes as $2^n$ and becomes
-prohibitively expensive when $n$ is a large number of
-factors. As the evaluation of SPARQL endpoints is an inherently complex
-problem, a large number of factors are needed, and so full experiments
-cannot scale. However, we may reduce the size of the experiment
-significantly by sacrificing some explanatory power, in the form of
-fractional factorial experiments.
+prohibitively expensive when $n$ is a large number of factors. As the
+evaluation of SPARQL endpoints is an inherently complex problem, a
+large number of factors are needed, and so full experiments cannot
+scale. However, we may reduce the size of the experiment significantly
+by sacrificing some explanatory power, in the form of fractional
+factorial experiments.
 
@@
 run time from larger Basic Graph Patterns for large databases, can
-also be explained by a less powerful machine evaluation FILTER clauses
-with ranges. They are indistinguishable. This may or may not cause a
-problem. In many cases, we may not be interested in these effects, we
-may only be interested in ``Implement'' and its
-interactions. Moreover, it is possible (even easy using the FrF2
-package in R) to declare which effects must not be aliased, and
-determine the size of the experiment based on that. Such a main effect
-or two-factor interaction is called \emph{clear} if none of its
-aliases are other main effects or two-factor interactions. Another
-possibility to shrink the size of the experiment is if we \textit{a
-  priori} can say that some effects are negligible. We have not found
-such assumptions to be tenable in our case.
+also be explained by a less powerful machine that is worse at
+evaluating FILTER clauses with ranges. They are
+indistinguishable. This may or may not cause a problem. In many cases,
+we may not be interested in these effects, we may only be interested
+in ``Implement'' and its interactions. Moreover, it is possible (even
+easy using the FrF2 package in R) to declare which effects must not be
+aliased, and determine the size of the experiment based on that. Such
+a main effect or two-factor interaction is called \emph{clear} if none
+of its aliases are other main effects or two-factor
+interactions. Another possibility to shrink the size of the experiment
+is if we \textit{a priori} can say that some effects are
+negligible. We have not found such assumptions to be tenable in our
+case.
 
@@
 effects, and that they correspond well to those we saw in the full
-factorial experiment. They also all have an intuitive
-explanation as above. However, we do not see the effects that we found
-worrisome in the full factorial design. The ``Implement:TripleC''
-interaction emerges only for a level $\alpha=0.15$. To ensure that we discover such
-cases will be a key challenge in further studies.
+factorial experiment. They also all have an intuitive explanation as
+above. However, we do not see the effects that we found worrisome in
+the full factorial design. The ``Implement:TripleC'' interaction
+emerges only for a level $\alpha=0.15$. To ensure that we discover
+such cases will be a key challenge in further studies.
 
@@
 Since the 64-run experiment has 3 inactive factors, ``Machine'',
-``TripleC'' and ``Range'', we can again treat it as a $2^5$ experiment
+``TripleC'' and ``Range'', we can treat it as a $2^5$ experiment
 replicated twice. Using the same procedure as above, we find that
 again, the hypothesis test indicates that the new implementation is an
-improvement at $p = 5.9 \cdot 10^{-6}$. Again, the comparatively higher p-value
-is due to the lower number of runs. It is also possible to perform the
-per-environmental test.
+improvement at $p = 5.9 \cdot 10^{-6}$. Again, the comparatively
+higher p-value is due to the lower number of runs. It is also possible
+to perform the per-environmental factors test.
 
@@
 aware that caching may be important in 4store, as a preliminary
-experiment gave the results in \ref{tab:cacheexp}. We see that the
+experiment gave the results in table~\ref{tab:cacheexp}. We see that the
 response time was significantly lower when an OPTIONAL clause was
@@
 
-It is \emph{randomization} that makes this permissible, i.e. the order
-of execution is random, so the benefit of caching is likely to apply
-randomly to different runs, for example, had the run order been
-reversed in the previous example, it could have been the query without
-OPTIONAL that would have benefitted. Unless the effect of caching is
-cumulative throughout the experiment (which is possible), the
-randomization will have the effect of reversing the runs randomly. The
-end result is that the effect of so-called ``lurking variables'' such
-as caching, warm-up effects, etc, contributes to total unexplained
-variance of the experiment.
+We did this to illustrate \emph{randomization} as this is what that
+makes it permissible, i.e. the order of execution is random, so the
+benefit of caching is likely to apply randomly to different runs, for
+example, had the run order been reversed in the previous example, it
+could have been the query without OPTIONAL that would have
+benefitted. Unless the effect of caching is cumulative throughout the
+experiment (which is possible), the randomization will have the effect
+of reversing the runs randomly. The end result is that the effect of
+so-called ``lurking variables'' such as caching, warm-up effects, etc,
+contributes to total unexplained variance of the experiment, but
+should not skew the results to invalidate the estimated effects.
 
@@
 
-Finally, we have only employed the simplest parts of the DoE
-theory, what has been presented here is part of a much more general
-formalism known as Orthogonal Arrays. The use of orthogonal arrays
-allow for different numbers of levels, for much greater flexibility in
-total run size, and for non-regular designs that can provide a good fit
-for the complex problem of SPARQL endpoint evaluations.
+Finally, we have only employed the simplest parts of the DoE theory,
+what has been presented here is part of a much more general formalism
+known as Orthogonal Arrays. The use of orthogonal arrays allow for
+different numbers of levels, for much greater flexibility in total run
+size, and for non-regular designs that can provide a good fit for the
+complex problem of SPARQL endpoint evaluations.
 
@@
 model, as exemplified by failure to identify the possibly non-linear
-effect of the P-ATA disks. This issue would be fixed by using an
-orthogonal array design.
+effect of the P-ATA disks in the first experiments. This issue can be
+addressed by using an orthogonal array design.
 
@@
 Kjetil Kjernsmo would like to thank his main supervisor Martin Giese
-for kind assistance. He would also like to
-thank Steve Harris for his generous support in degrading his excellent
-work on 4store.
+for kind assistance. He would also like to thank Steve Harris for his
+generous support in degrading his excellent work on 4store.
 

### 2013-05-10 exp1/iswc2013.tex
spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@574 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 as design matrices, but they can also be described sufficiently for
-reproducability in terms of \emph{design generators}. These also
+reproducibility in terms of \emph{design generators}. These also
 declare aliasing relations for some of the main effects. The design
@@
 must be parameterized with many more factors than just the number of
-triples; data heterogenity \cite{Duan:2011:AOC:1989323.1989340} is one
+triples; data heterogeneity \cite{Duan:2011:AOC:1989323.1989340} is one
 example, but also queries over highly skewed data is important. A more
@@
 their SPLODGE system is a key ingredient. They parameterized SPARQL
-queries in such a way they could be autogenerated. One suggestion is
+queries in such a way they could be auto-generated. One suggestion is
 to parameterize queries based on the grammar, so that one factor
@@
 that the experiment was flawed, and we have cautioned that failure to
-control unexplained variance may compromize our ability to do so. We
+control unexplained variance may compromise our ability to do so. We
 have also pointed out how this direction can provide more rigorous
@@
 Even if the complexity is great, with properly tuned endpoints, it is
-feasible to do millions of runs, and with orthogonoal arrays, this
+feasible to do millions of runs, and with orthogonal arrays, this
 formalism can be extended to many different evaluation problems.

### 2013-05-10 exp1/generate-figures.R
Figures could be a little larger
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@575 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -17,13 +17,13 @@ par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac32r1.lm, main="", cex.fac=0.8)
 dev.off()
 
-pdf(file="frac64normal.pdf", height=5)
+pdf(file="frac64normal.pdf", height=6)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64r3.lm, alpha=c(0.05, 0.15), main="", cex.fac=0.8)
 dev.off()
 
 
-pdf(file="frac64hugenormal.pdf", height=5.5)
+pdf(file="frac64hugenormal.pdf", height=6)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64huger1.lm, main="", cex.fac=0.8)
 dev.off()

### 2013-05-10 exp1/iswc2013.tex
wrong university name
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@576 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 N-0316 Oslo, Norway \email{kjekje@ifi.uio.no} \and Department of Mathematical Sciences,
-The Norwegian Institute of Technology,
+Norwegian University of Science and Technology,
 N-7491 Trondheim,

### 2013-05-21 exp1/generate-figures.R
post-submission figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@579 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -7,7 +7,7 @@ source("FrF2-mod/remodel.R")
 source("FrF2-mod/check.R")
 source("FrF2-mod/DanielPlot.R")
 
-pdf(file="fullnormal.pdf")
+pdf(file="fullnormal.pdf", height=8)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(runbgplangf1.lm, main="", cex.fac=0.8)
 dev.off()
@@ -17,13 +17,13 @@ par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac32r1.lm, main="", cex.fac=0.8)
 dev.off()
 
-pdf(file="frac64normal.pdf", height=6)
+pdf(file="frac64normal.pdf", height=5.7)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64r3.lm, alpha=c(0.05, 0.15), main="", cex.fac=0.8)
 dev.off()
 
 
-pdf(file="frac64hugenormal.pdf", height=6)
+pdf(file="frac64hugenormal.pdf", height=5)
 par(mai=c(0.6,0.6,0.05,0.05))
 DanielPlot(frac64huger1.lm, main="", cex.fac=0.8)
 dev.off()

### 2013-05-21 exp1/iswc2013.tex
typo from Arild
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@578 43ef1190-90fd-4dd7-a769-d0121f02b80e
Johns final comment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@581 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 number of runs the experiment needs, and we discuss the trade-offs
-involved. We then continue to discuss techniques that shows our
+involved. We then continue to discuss techniques that show our
 simplistic experiment to be flawed, yet scientifically better than the
@@
 in fact, it hints that our optimization may not work as well for
-larger databases. This is a clear indication that the separation
-between levels for the sizes of the data-set is too small and should
-be investigated further. Such reasoning should be applied to all
-significant effects.
+larger databases. By carefully inspecting table~\ref{tab:pvaluesfull},
+we see that we see that whenever ``TripleC'' is at the low lever, the
+new implementation is always best, only at the higher level, the old
+implementation may be better. This is a clear indication that the
+separation between levels for the sizes of the data-set is too small
+and should be investigated further. Such reasoning should be applied
+to all significant effects.
 

### 2013-06-06 dagstuhl-2013/position.tex
Start of Dagstuhl position paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@584 43ef1190-90fd-4dd7-a769-d0121f02b80e

%\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{verbatim}
\usepackage{graphicx}

\title{Merits of hypermedia systems}
\author{Kjetil Kjernsmo}

\begin{document}

\maketitle

\section{Research interests}

My primary research interest is the optimization of SPARQL queries in
a federated regime, as we have noted that this is not practical
because the federation engine has insufficient information to
optimize, or the information is so large that it defeats the purpose
of optimizations to begin with. I plan to help remedy this problem by
computing very compact digests and expose them in the service
description. I have not yet published any articles on this topic, but
the research is in the immediate extension of
SPLENDID\cite{splendid}. My secondary research interest is replacing
the practice of benchmarking by using statistical design of
experiment.

However, coming from an industry background in software development,
experience suggests that the above research interests does not
adequately address many immediate needs when developing information
systems to process the rapidly increasing amount of available data. I
believe that large SPARQL-driven systems would be the ``right tool for
the job'' in only a limited, and currently unclear, set of
cases. Further exposure to new ideas in the developer community lead
me to develop a third interest, namely hypermedia RDF.

### 2013-06-06 dagstuhl-2013/position.tex
Some details about hypermedia
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@585 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-%\documentclass{article}
+\documentclass{article}
 \usepackage[utf8]{inputenc}
+\usepackage{hyperref}
 \usepackage{cite}
@@
 me to develop a third interest, namely hypermedia RDF.
+
+\section{Hypermedia RDF}
+
+In \cite{kjernsmo_lapis_2012}, examined some practical implications of
+the HATEOAS constraint of the REST architectural style, see
+\cite{Fielding_2000_Architectural-Styles}~Chapter~5., and in that
+light argued why hypermedia RDF is a practical necessity.
+
+Mike Amundsen defines hypermedia types\cite{hypermediatypes} as 
+\begin{quote}
+Hypermedia Types are MIME media types that contain native
+hyper-linking semantics that induce application flow. For example,
+HTML is a hypermedia type; XML is not.
+\end{quote}
+
+We continued to derive a powerful hypermedia type based on RDF within
+a classification suggested by Mike Amundsen. Since this publication, I
+have also noted that the embedded links factor can be achieved by
+using \texttt{data} URIs, thus satisfying all but one of the factors
+proposed by Mike Amundsen.
+
+Further, we noted that other interesting factors is the
+self-description, which is a important characteristic of the RDF
+model, and other voiced other minor concerns.
+
+To bring forward a concrete example of how to make a serialized RDF
+graph into a hypermedia type, we suggest adding some triples to every
+resource (where prefixes are omitted for brevity):
+
+\begin{verbatim}
+<> hm:canBe hm:mergedInto, hm:replaced, hm:deleted ;
+   hm:inCollection <../> ;
+   void:inDataset [ void:sparqlEndpoint </sparql> . ] .
+\end{verbatim}
+
+\bibliographystyle{plain}
+%\bibliographystyle{jbact}
+%\bibliographystyle{splncs03}
+\bibliography{federation,webarch,vocabs,specs,hypermedia,egne,selectivity}
+
+\end{document}

### 2013-06-07 dagstuhl-2013/position.tex
Use case
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@587 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Possible uses}
+
+Over time, I believe that interfaces that require developers to read
+external documentation will loose to interfaces where ``View Source''
+is sufficient to learn everything that is needed. This is the essence
+of hypermedia systems, in the RDF case, everything needed to program
+is available in the RDF. It tells application what it may do next.
+
+In the above, we have included mostly create, add and delete
+primitives, but these could be refined for application scenarios if
+needed.
+
+For example a pizza baker publishes Linked Data about pizzas they
+sell, including data sufficient to create a sophisticated search and
+sales application. The Linked Data will then include triples that
+state explicitly how the order should be placed, i.e. what resources
+need updating. Moreover, the may not only want to sell pizzas, but
+also drinks. Thus, the Linked Data presented to the application should
+not only be a Symetric Concise Bounded Description, as common today,
+but careful designed to provide exactly the data needed to optimize
+sales.
+
 \bibliographystyle{plain}

### 2013-06-07 dagstuhl-2013/position.tex
spell check
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@588 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-My primary research interest is the optimization of SPARQL queries in
+My primary research interest is the optimisation of SPARQL queries in
 a federated regime, as we have noted that this is not practical
 because the federation engine has insufficient information to
-optimize, or the information is so large that it defeats the purpose
-of optimizations to begin with. I plan to help remedy this problem by
+optimise, or the information is so large that it defeats the purpose
+of optimisations to begin with. I plan to help remedy this problem by
 computing very compact digests and expose them in the service
@@
 
-To bring forward a concrete example of how to make a serialized RDF
+To bring forward a concrete example of how to make a serialised RDF
 graph into a hypermedia type, we suggest adding some triples to every
@@
 also drinks. Thus, the Linked Data presented to the application should
-not only be a Symetric Concise Bounded Description, as common today,
-but careful designed to provide exactly the data needed to optimize
+not only be a Symmetric Concise Bounded Description, as common today,
+but careful designed to provide exactly the data needed to optimise
 sales.

### 2013-06-07 dagstuhl-2013/position.tex
Problems with SPARQL
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@589 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Problems with SPARQL interfaces are many: They require extensive
+training of developers; it is not immediately clear what data are
+available and what may done with the data; it is easy to formulate
+queries that will cause the endpoint to become overloaded and hard to
+protect against them without also rejecting legitimate queries; it
+places heavier systems in the execution path of an application, etc.
+
 \section{Hypermedia RDF}
 
-In \cite{kjernsmo_lapis_2012}, examined some practical implications of
+In \cite{kjernsmo_lapis_2012}, I examined some practical implications of
 the HATEOAS constraint of the REST architectural style, see
@@
 self-description, which is a important characteristic of the RDF
-model, and other voiced other minor concerns.
+model, and other minor concerns.
 

### 2013-06-07 dagstuhl-2013/position.tex
Trim to 2 pages
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@590 43ef1190-90fd-4dd7-a769-d0121f02b80e
Final proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@591 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the research is in the immediate extension of
-SPLENDID\cite{splendid}. My secondary research interest is replacing
-the practice of benchmarking by using statistical design of
-experiment.
+SPLENDID\cite{splendid}. My secondary research interest is using statistical design of
+experiment in software performance evaluation.
 
@@
 the HATEOAS constraint of the REST architectural style, see
-\cite{Fielding_2000_Architectural-Styles}~Chapter~5., and in that
+\cite{Fielding_2000_Architectural-Styles}~Chapter~5, and in that
 light argued why hypermedia RDF is a practical necessity.
@@
 a classification suggested by Mike Amundsen. Since this publication, I
-have also noted that the embedded links factor can be achieved by
+have also noted that the ``embedded links'' factor can be achieved by
 using \texttt{data} URIs, thus satisfying all but one of the factors

### 2013-06-20 exp1/rebuttal.txt
rebuttal
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@592 43ef1190-90fd-4dd7-a769-d0121f02b80e

We would like to thank the anonymous reviewers for their thoughtful
reviews. Your points are well taken and we shall work to incorporate
your advices to the final version.

First, benchmarking and designed experiments serve to answer different
questions. Since a designed experiment requires parameterization of
language and data "features", natural benchmarks such as the DBPedia
Benchmark, cited in the paper, will remain a better judge of the
performance of typical queries in some typical situations. However, it
cannot, as we noted in the three points on the first page, answer
scientifically relevant questions on e.g. falsifiability of more general
assertions about performance.

One reviewer notes that it will be tricky to extend the experiment to
real federated SPARQL querying. Indeed, we believe this to be the work
of more than one Ph.D. Thus, this paper is to encourage a collective
effort of the community. It is not unlikely that to attain this goal,
progress beyond the state-of-the-art in statistics is also required.

Two reviewers point out that the introduction to DoE should be
improved. This point is very valid. We envision this paper to be a
"first contact" with DoE for many readers, so the presentation is
especially important, and your suggestions for improvements are very
welcome.

Obviously, we also acknowledge that the "toy example" nature of the
paper is a clear drawback. We are quite confident that this
methodology will be a clear improvement over "constructed" benchmarks,
like for example the now classical Berlin SPARQL Benchmark with
relatively modest means, but we acknowledge that we have not attempted
to prove this notion. We shall attempt to provide at least anecdotal
evidence based on the current example for the camera ready
version. Yet, we are of the opinion that if we had attempted a full
replacement of the larger benchmarks in this paper, the discussion
would have been significantly harder to follow.

As an expansion of the introduction DoE requires more space, one
reviewer's suggestion to drop Table 3 is welcome. However, Table 3 was
not intended as a preliminary investigation, but rather an
illustration of the concept of "lurking variables", which is
conceptually important. We would be interested in hearing the
reviewer's opinion on the importance of this point.

In conclusion, there are no major confusion points to be cleared, the
reviews are helpful and if the paper is accepted, we shall seek to
incorporate their suggestions to the extent possible within the page
limit.

### 2013-07-22 exp1/iswc2013.tex
Reformat to table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@601 43ef1190-90fd-4dd7-a769-d0121f02b80e
Reformat the rest to table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@602 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Then, we test some language features: ``BGPComp'', which is a Basic
-Graph Pattern of varying complexity. The first level is:
-\verbatiminput{experiment/BGPComp-1}
-and the second level is 
-\verbatiminput{experiment/BGPComp-2}
+Graph Pattern of varying complexity.
+
+\begin{tabular}{p{0.45\textwidth}p{0.45\textwidth}}
+Level 1 & Level 2 \\ 
+\verbatiminput{experiment/BGPComp-1} & \verbatiminput{experiment/BGPComp-2} \\
+\end{tabular}
 
 The following factors tests the absence (at level 1) or presence (at
-level 2) of the following clauses: The ``Lang'' factor is this FILTER:
-\verbatiminput{experiment/Lang-2}
-and similarly,  ``Range'' is this FILTER:
-\verbatiminput{experiment/Range-2}
-``Union'' tests the following:
-\verbatiminput{experiment/Union-2}
-and ``Optional''
+level 2) of the following clauses:
+
+\begin{tabular}{p{0.40\textwidth}p{0.55\textwidth}}
+
+``Union'' & ``Lang'' \\
+\verbatiminput{experiment/Union-2} &
+\verbatiminput{experiment/Lang-2} \\
+
+``Range'' & ``Optional'' \\
+\verbatiminput{experiment/Range-2} & 
 \verbatiminput{experiment/Optional-2}
+\end{tabular}
 

### 2013-07-22 exp1/iswc2013.tex
typos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@603 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The following factors tests the absence (at level 1) or presence (at
+The following factors test the absence (at level 1) or presence (at
 level 2) of the following clauses:
@@
 A real optimization is beyond the scope of this paper. Instead we have
-performed a simulation that enable us to understand the effect of the
+performed a simulation that enables us to understand the effect of the
 changes.  In a real-world case, there is often the case that an
@@
 both the optimization and the side-effects. To do that, we degraded
-the performance of a the 4store SPARQL implementation by inserting
+the performance of the 4store SPARQL implementation by inserting
 \texttt{sleep} statements. Specifically we inserted the C statement
@@
 larger databases. By carefully inspecting table~\ref{tab:pvaluesfull},
-we see that we see that whenever ``TripleC'' is at the low lever, the
+we see that whenever ``TripleC'' is at the low level, the
 new implementation is always best, only at the higher level, the old
@@
 
-We loose explanatory power due to \emph{aliasing}, for example, the
+We lose explanatory power due to \emph{aliasing}, for example, the
 ``TripleC:BGPComp'' interaction may be aliased with the
@@
 understanding is needed. To some extent, ``BGPComp'' is also a broad
-factor, since it encompasses many structures that is normally seen in
+factor, since it encompasses many structures that are normally seen in
 a Basic Graph Pattern. In this case, the cause of the performance
@@
 purposes, which consists of degrading it in several ways. The small
-data-sets used here does not imply that the approach will not work for
+data-sets used here do not imply that the approach will not work for
 large data-sets; properly configured, actual optimization will give
@@
 
-We saw that while 2 level experiments can perform well in determining
+We saw that while two-level experiments can perform well in determining
 significant effects, they do not necessarily work well to find a

### 2013-07-22 exp1/iswc2013.tex
Martins suggestions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@604 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 One important aspect of science is the structured approach to falsify
-hypotheses, and thus is important that assertions about performance is
-formulated as hypotheses that can be falsified.
+hypotheses, and thus is important that assertions about performance
+are formulated as hypotheses that can be falsified.
 
@@
 our assumptions are flawed. Such constraints can only be made when we
-have very good reason to suspect that a certain parameter does not
-have any influence.
+have highly confident that a certain parameter does not have any
+influence.
 
@@
 not found the parts discussed in the present paper to be widely
-adopted. Recently, a comprehensive text on experimental methods have
+adopted. Recently, a comprehensive text on experimental methods has
 been published in
@@
 industrial benchmarking rather than scientific evaluations.
+% TODO: skrive om til å snakke om use-case driven
 
@@
 Software ones that can be modified by anyone, and the other is to
-constrain the scope of the study to comparing two different things.
+constrain the scope of the study to comparing just two different
+things.
 
@@
 four 4store instances runs independently on different ports. 
+% TODO: process waiting.
 

### 2013-07-22 exp1/iswc2013.tex
Upcase Figure
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@605 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 from level 1 to 2 is twice the coefficients of the fit. Then, we
-generate a normal plot, see figure~\ref{fig:fullnormal}. In this plot,
+generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,
 if the null hypothesis $H_0$ holds, all estimated effects will fall on
@@
 table~\ref{tab:effects}, the most important effects are listed.  In
-figure~\ref{fig:fullnormal}, Lenth's method is used to label the
+Figure~\ref{fig:fullnormal}, Lenth's method is used to label the
 significant effects. We see that we have many very significant effects
@@
 hypothesis test. By inspecting the normal plot in
-figure~\ref{fig:fullnormal} and table~\ref{tab:effects}, we see that
+Figure~\ref{fig:fullnormal} and table~\ref{tab:effects}, we see that
 six factors are highly significant by either being a significant main
@@
 
-The resulting normal plot is in figure~\ref{fig:frac32normal}. We see
+The resulting normal plot is in Figure~\ref{fig:frac32normal}. We see
 that only two main effects are significant on a $\alpha = 0.05$ level
@@
 We turn to the 64-run experiment and its normal plot in
-figure~\ref{fig:frac64normal}. We see that the we get more significant
+Figure~\ref{fig:frac64normal}. We see that the we get more significant
 effects, and that they correspond well to those we saw in the full
@@
 The importance of finding flaws in the levels is highlighted by the
-normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
+normal plot in Figure~\ref{fig:frac64hugenormal}. The smaller
 experiment made it feasible to run with a larger data-set, this

### 2013-07-22 exp1/iswc2013.tex
More minor fix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@606 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Thus, attacking the complexity problem by eliminating parameters we
-think influence the outcome for simplicity is untenable as scientific
-practice. We must instead tackle this problem by enhancing our
-capability of testing complex systems and to identify situations where
-our assumptions are flawed. Such constraints can only be made when we
-have highly confident that a certain parameter does not have any
-influence.
+think influence the outcome (e.g. caching) for simplicity is untenable
+as scientific practice. We must instead tackle this problem by
+enhancing our capability of testing complex systems and to identify
+situations where our assumptions are flawed. Such constraints can only
+be made when we are highly confident that a certain parameter does not
+have any influence.
 
@@
 Finally, we acknowledge the efforts of the Linked Data Benchmark
-Council and the SEALS project. However, they appear to focus on
-industrial benchmarking rather than scientific evaluations.
-% TODO: skrive om til å snakke om use-case driven
+Council and the SEALS project. However, they appear to focus on cases
+where the workload is assumed to be well characterized, rather than
+scientific evaluations.
 
@@
 ``TripleC'' and ``Implement'' specifies a port number. On each host,
-four 4store instances runs independently on different ports. 
-% TODO: process waiting.
+four 4store instances runs sequentially and independently on different
+ports, where processes that are not actively answering a query are
+idle.
 
@@
 All the code to reproduce this study as well as detailed instructions
-is published on Github: \url{https://github.com/kjetilk/doe-sparql}.
+have been published on Github: \url{https://github.com/kjetilk/doe-sparql}.
 
@@
 
-To investigate whether the detrimental side-effects cancels the
+To investigate whether the detrimental side-effects cancel the
 positive effect of our simulated optimization, we have to formulate a
@@
 \begin{center}
-\caption{p-values for different parts of the experiment}\label{tab:pvaluesfull}
+\caption{$p$-values for different parts of the experiment}\label{tab:pvaluesfull}
 \begin{tabular}{cccccl}
@@
 For future larger experiments, one must investigate whether it is
-necessary to adjust the p-value due to the larger numbers of
+necessary to adjust the $p$-value due to the larger numbers of
 hypotheses tested.
@@
 We lose explanatory power due to \emph{aliasing}, for example, the
-``TripleC:BGPComp'' interaction may be aliased with the
+interaction labelled ``TripleC:BGPComp'' may be aliased with the
 ``Machine:Range'' interaction. That is to say, a detected increase in
@@
 improvement at $p = 5.9 \cdot 10^{-6}$. Again, the comparatively
-higher p-value is due to the lower number of runs. It is also possible
+higher $p$-value is due to the lower number of runs. It is also possible
 to perform the per-environmental factors test.
@@
 in this simulation, to the contrary, we chose 4store because of its
-relative simplicity to modify it to suite our needs for illustrative
+relative simplicity to modify it to suit our needs for illustrative
 purposes, which consists of degrading it in several ways. The small
@@
 minimum. Thus, lurking variables should be turned into factors
-whenever possible, is they degrade the quality of the experiment, but
+whenever possible, as they degrade the quality of the experiment, but
 in many cases, we can live with them if their contribution to the

### 2013-07-22 exp1/iswc2013.tex
Upcase Table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@607 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 estimating significance in an experiment with no replicates. In
-table~\ref{tab:effects}, the most important effects are listed.  In
+Table~\ref{tab:effects}, the most important effects are listed.  In
 Figure~\ref{fig:fullnormal}, Lenth's method is used to label the
@@
 hypothesis test. By inspecting the normal plot in
-Figure~\ref{fig:fullnormal} and table~\ref{tab:effects}, we see that
+Figure~\ref{fig:fullnormal} and Table~\ref{tab:effects}, we see that
 six factors are highly significant by either being a significant main
@@
 of the replications, as above. The results are tabulated in
-table~\ref{tab:pvaluesfull}. We see that there are significant
+Table~\ref{tab:pvaluesfull}. We see that there are significant
 improvements in most cases, one may want to further investigate the
@@
 in fact, it hints that our optimization may not work as well for
-larger databases. By carefully inspecting table~\ref{tab:pvaluesfull},
+larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
 we see that whenever ``TripleC'' is at the low level, the
@@
 aware that caching may be important in 4store, as a preliminary
-experiment gave the results in table~\ref{tab:cacheexp}. We see that the
+experiment gave the results in Table~\ref{tab:cacheexp}. We see that the
 response time was significantly lower when an OPTIONAL clause was

### 2013-07-23 exp1/iswc2013.tex
smaller font in code snippets
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@608 43ef1190-90fd-4dd7-a769-d0121f02b80e
figure placement and sizing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@609 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{graphicx}
+\usepackage{wrapfig}
 
@@
 
-\begin{tabular}{p{0.45\textwidth}p{0.45\textwidth}}
-Level 1 & Level 2 \\ 
-\verbatiminput{experiment/BGPComp-1} & \verbatiminput{experiment/BGPComp-2} \\
+  \begin{center}\small
+    \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
+      \hline
+      \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
+      \verbatiminput{experiment/BGPComp-1}
+      & \verbatiminput{experiment/BGPComp-2}\\
+      \hline
     \end{tabular}
+  \end{center}
 
@@
 
-\begin{tabular}{p{0.40\textwidth}p{0.55\textwidth}}
-
-``Union'' & ``Lang'' \\
+\begin{center}\small
+  \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
+    \hline
+    \textbf{``Union''} & \textbf{``Lang''} \\[-2ex]
     \verbatiminput{experiment/Union-2} &
     \verbatiminput{experiment/Lang-2} \\
-
-``Range'' & ``Optional'' \\
+    \hline
+    \textbf{``Range''} & \textbf{``Optional''} \\[-2ex]
     \verbatiminput{experiment/Range-2} & 
-\verbatiminput{experiment/Optional-2}
+    \verbatiminput{experiment/Optional-2}\\
+    \hline
   \end{tabular}
-
+\end{center}
 These fragments has been carefully designed for illustrative utility,
@@
 
-\begin{figure}[h!]
-  \includegraphics[width=\textwidth]{fullnormal.pdf}
+\begin{figure}[t]
+  \centerline{%
+  \includegraphics[width=.9\textwidth]{fullnormal.pdf}}
   \caption{A normal plot of the Full Factorial Experiment. The
@@
 % Modified by hand to remove 1's
-\begin{table}[hb]
-\begin{center}
+\begin{wraptable}{l}{0.5\textwidth}
 \caption{The magnitude of effects for some important main effects and interactions.}\label{tab:effects}
+\medskip
 \begin{tabular}{rr}
@@
   \hline
-Implement & -21.27 \\ 
-  Implement:Optional & -11.49 \\ 
-  Implement:Union & -6.21 \\ 
+Implement & $-$21.27 \\ 
+  Implement:Optional & $-$11.49 \\ 
+  Implement:Union & $-$6.21 \\ 
   Implement:TripleC:Lang1 & 4.12 \\ 
@@
 \end{tabular}
-\end{center}
-\end{table}
+\end{wraptable}
 
@@
 
-\begin{table}[ht]
+\begin{table}[h!]
 \begin{center}
@@
 \begin{figure}[h!]
-  \includegraphics[width=\textwidth]{frac32normal.pdf}
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{frac32normal.pdf}}
   \caption{A normal plot of the Fractional Factorial Experiment with
@@
     the Lenth criterion at a level
-    $\alpha=0.05$.}\label{fig:frac32normal}
+    $\alpha=0.05$.}
+  \label{fig:frac32normal}
 \end{figure}
@@
 
-\begin{figure}[ht]
-  \includegraphics[width=\textwidth]{frac64normal.pdf}
+\begin{figure}[h!]
+  \centerline{%
+    \includegraphics[width=0.9\textwidth]{frac64normal.pdf}}
   \caption{A normal plot of the Fractional Factorial Experiment with
@@
 \begin{figure}[h!]
-  \includegraphics[width=\textwidth]{frac64hugenormal.pdf}
+  \centerline{%
+  \includegraphics[width=.9\textwidth]{frac64hugenormal.pdf}}
   \caption{A normal plot of the Fractional Factorial Experiment with

### 2013-07-23 exp1/iswc2013.tex
Introductory material
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@610 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Design of Experiments Concepts}
+
+We expect DoE to be new to most readers, but some familiarity with
+hypothesis testing is assumed.
+
+First, note that the concept of parameters as considered above is
+known as \emph{factors} in DoE. For each factor, one may make a
+measurement of a \emph{response variable} at certain \emph{levels},
+e.g. one may measure the response time of a SPARQL endpoint by
+measuring it when it contains 1 or 2 million triples. Then, the
+response time is the response variable, the factor is the number of
+triples and there are two levels, 1 or 2 million triples. We have chosen in
+this paper to only deal with the relatively straightforward aspects of
+the DoE formalism, and therefore constrain the experiment strictly to
+2 levels.
+
+% TODO: 
+% 1) what are effects
+%% 2) what are levels
+%% 3) what are factors 
+%% 4) Can run 2^8 experiments, call full factorial
+%% 5) What are normal plots
+%% 6) explain interaction effects
+
 \section{Related work}
@@
 data-set by using the last 1 or 2 MTriples from their file.
-
-First, note that the concept of parameters as considered in the
-introduction is known as \emph{factors} in DoE. For each factor, we
-have chosen in this paper to only deal with the relatively
-straightforward aspects of the DoE formalism, and therefore constrain
-the experiment strictly to 2 levels.
 

### 2013-07-23 exp1/iswc2013.tex
Describe fact designs; normal plots
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@611 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 direction that promises to address these problems. First, we design a
-simple 8-parameter experiment, and then we demonstrate how a full
-factorial experiment can meaningfully provide a summary of the
-statistics.
+simple 8-parameter experiment, and then we detail an experiment that
+can meaningfully provide a summary of the statistics.
 
-Then, we demonstrate how a fractional factorial experiment can be used
-to maintain a structured approach to the problem while reducing the
-number of runs the experiment needs, and we discuss the trade-offs
-involved. We then continue to discuss techniques that show our
-simplistic experiment to be flawed, yet scientifically better than the
-practice of benchmarking. Finally, we outline a road-map to establish
-the use of these techniques in SPARQL endpoint evaluations both as a
-practical tool and as a scientifically more rigorous methodology.
+Then, we demonstrate how the number of runs required in the experiment
+can be reduced while maintaining a structured approach to the problem,
+and we discuss the trade-offs involved. We then continue to discuss
+techniques that show our simplistic experiment to be flawed, yet
+scientifically better than the practice of benchmarking. Finally, we
+outline a road-map to establish the use of these techniques in SPARQL
+endpoint evaluations both as a practical tool and as a scientifically
+more rigorous methodology.
 
-\subsection{Design of Experiments Concepts}
+\subsection{Key concepts of Design of Experiments}
 
@@
 
-% TODO: 
-% 1) what are effects
-%% 2) what are levels
-%% 3) what are factors 
-%% 4) Can run 2^8 experiments, call full factorial
-%% 5) What are normal plots
-%% 6) explain interaction effects
+In an experiment, there are several factors, and one may run the
+experiment by measuring the response variable for every combination of
+levels. In a two-level experiment, this will result in $2^n$ runs,
+where $n$ is the number of factors. This is called a \emph{full
+  factorial experiment}. This is done in Section~\ref{sec:full}.
+
+In many fields of science, experimental economy is extremely
+important, and so, DoE offers extensive methodology to run a certain
+fraction of the runs at the price of explanatory power. Such
+experiments are called \emph{fractional factorial experiments} and is
+covered in Section~\ref{sec:frac}.
+
+After measurements are done, we seek to estimate \emph{effects}. To
+see how the \emph{main effects} can be estimated, consider a factor
+$A$. We proceed to find the averages $a_1$ and $a_2$ of all measurements at level~1 and
+on level~2 respectively. Then the main effect of $A$ is $a_2 - a_1$.
+
+Similarly, \emph{interaction effects} for 2 factors $AB$ are defined
+by comparing averages for equal versus non-equal levels of $A$ and
+$B$.
+
+This can be generalized to larger combinations of factors, but in
+practice, this analysis is done by fitting a linear model to the data
+with all factors and their interactions. It is shown in
+\cite{wu2009experiments} that the effect is twice the coefficients of
+the fit.
+
+The next step is to understand which factors are important or
+significant.  One approach is to plot the sorted effects against the
+normal distribution. To understand why, note that if there is nothing
+of interest, the measurements have a certain normal distribution
+purely due to noise. When plotted against a normal distribution, the
+plot would be a straight line. If there are any deviations from the
+straight line, it implies that they may be significant. Such plots can
+be found in
+figures~\ref{fig:fullnormal},~\ref{fig:frac32normal}~and~\ref{fig:frac64normal}. The
+range in the $y$-axis is dictated by the number of runs.
 
@@
 consisting of the factors with a column containing the corresponding
-curl measurements. To analyze the experiment, we start by fitting a
-linear model to the data with all factors and their interactions. It
-is shown in \cite{wu2009experiments} that the effect, i.e. the change
-from level 1 to 2 is twice the coefficients of the fit. Then, we
+curl measurements. Then, we
 generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,

### 2013-07-23 exp1/iswc2013.tex
Some minor fixes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@612 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 known as \emph{factors} in DoE. For each factor, one may make a
-measurement of a \emph{response variable} at certain \emph{levels},
-e.g. one may measure the response time of a SPARQL endpoint by
-measuring it when it contains 1 or 2 million triples. Then, the
-response time is the response variable, the factor is the number of
-triples and there are two levels, 1 or 2 million triples. We have chosen in
-this paper to only deal with the relatively straightforward aspects of
-the DoE formalism, and therefore constrain the experiment strictly to
-2 levels.
+measurement of a \emph{response variable} at certain
+\emph{levels}. For example, one may measure the response time of a
+SPARQL endpoint by measuring it when it contains 1 or 2 million
+triples. Then, the response time is the response variable, the factor
+is the number of triples and there are two levels, 1 or 2 million
+triples. We have chosen in this paper to only deal with the relatively
+straightforward aspects of the DoE formalism, and therefore constrain
+the experiment strictly to 2 levels.
 
@@
 be found in
-figures~\ref{fig:fullnormal},~\ref{fig:frac32normal}~and~\ref{fig:frac64normal}. The
+figures~\ref{fig:fullnormal},~\ref{fig:frac32normal},~\ref{fig:frac64normal}~and~\ref{fig:frac64hugenormal}. The
 range in the $y$-axis is dictated by the number of runs.
@@
 are executed, in total 256 runs. This is called a $2^8$ factorial
-experiment, and we can see that the number of runs grows exponentially
-with the number of factors, thus making large experiments
-prohibitively expensive. We shall nevertheless find the following
-small example instructive. Note that each combination is executed only
-once (i.e. it is \emph{unreplicated}), however, so even a full
-factorial experiment compares well to a typical benchmark in which
-each microbenchmark must be executed several times to obtain enough
-data to compute a mean with reasonable certainty.
+experiment, and while it quickly becomes infeasible for many factors,
+we shall nevertheless find the following small example
+instructive. Note that each combination is executed only once (i.e. it
+is \emph{unreplicated}), however, so even a full factorial experiment
+compares well to a typical benchmark in which each microbenchmark must
+be executed several times to obtain enough data to compute a mean with
+reasonable certainty.
 
@@
 consisting of the factors with a column containing the corresponding
-curl measurements. Then, we
-generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,
-if the null hypothesis $H_0$ holds, all estimated effects will fall on
-a straight line, as their means would be zero. Any significant
-departure from a straight line can be interpreted as a significant
-effect (see Section~4.8 of \cite{wu2009experiments} for a detailed
+curl measurements. With that, we do as described in the introduction
+and fit a linear model with all factors. Then, we generate a normal
+plot, see Figure~\ref{fig:fullnormal}.  Any significant departure from
+a straight line can be interpreted as a significant effect (see
+Section~4.8 of \cite{wu2009experiments} for a detailed
 explanation). In our case, it is most important to note that any

### 2013-07-23 exp1/iswc2013.tex
Review 2 suggestions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@613 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 SPARQL endpoint by measuring it when it contains 1 or 2 million
-triples. Then, the response time is the response variable, the factor
-is the number of triples and there are two levels, 1 or 2 million
-triples. We have chosen in this paper to only deal with the relatively
-straightforward aspects of the DoE formalism, and therefore constrain
-the experiment strictly to 2 levels.
+triples. Levels are not constrained to be continous variables, they
+can be discrete, or even two different instances of a class. The
+ordering is not essential, but it may contribute to how easy it is to
+understand the results, as we shall see. Then, the response time is
+the response variable, the factor is the number of triples and there
+are two levels, 1 or 2 million triples. We have chosen in this paper
+to only deal with the relatively straightforward aspects of the DoE
+formalism, and therefore constrain the experiment strictly to 2
+levels.
 
@@
 
-A literature survey has not revealed any direct prior art, however, the
+A literature survey has not revealed any direct prior art, neither in
+the Semantic Web field nor more generally in database research. However, the
 general approach has been well established, not only in statistics. A

### 2013-07-23 exp1/iswc2013.tex
Move host+port number stuff to Review 4
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@614 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 itself set up using 4store (see \cite{harris20094store}) version
-1.1.5. The ``Machine'' factor specifies a hostname, whereas
-``TripleC'' and ``Implement'' specifies a port number. On each host,
-four 4store instances runs sequentially and independently on different
-ports, where processes that are not actively answering a query are
-idle.
+1.1.5.
 
@@
 run on a third computer in a 1 gigabit Ethernet LAN with the two
-experiment hosts.  As response variable, we have chosen to use the
-time from the namespace server lookup finishes to the data has been
+experiment hosts. On the experiment hosts, four 4store instances runs
+sequentially and independently on different ports, where processes
+that are not actively answering a query are idle. Practically, the
+``Machine'' factor specifies a hostname, whereas ``TripleC'' and
+``Implement'' specifies a port number. On each host,
+
+As response variable, we have chosen to use the time from the
+namespace server lookup finishes to the data has been
 transferred. This choice is somewhat arbitrary, many other response

### 2013-07-23 exp1/iswc2013.tex
Machine details in footnote
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@615 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 of triples in the data-set, 1 or 2 MTriples, ``Machine'', which is the
-software and hardware platform, where one machine runs GNU/Linux
-Debian Squeeze, has 16 GB RAM, an Intel Core2 Duo E8500 CPU and two
-Parallel-ATA disks in a RAID-1 configuration. The other machine runs
+software and hardware platform, one larger with slower
+disks\footnote{Running GNU/Linux Debian Squeeze, has 16 GB RAM, an
+  Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
+  configuration} and one smaller with faster disk\footnote{Running
   Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
-SSD on SATA-III.
+  SSD on SATA-III}.
 

### 2013-07-23 exp1/iswc2013.tex
add TODO
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@617 43ef1190-90fd-4dd7-a769-d0121f02b80e
Upcase section
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@618 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   factorial experiment}. This is done in Section~\ref{sec:full}.
+%TODO {Simple 8 x 2 overview table?}
 
@@
 for identifying significant effects even though the underlying
-response may be non-linear.  As we saw in section~\ref{sec:full}, the
+response may be non-linear.  As we saw in Section~\ref{sec:full}, the
 choice of levels may be critical, and if non-linearity is expected,
@@
 
-We saw in section~\ref{sec:frac} that with a smaller experiment, fewer
+We saw in Section~\ref{sec:frac} that with a smaller experiment, fewer
 interactions are significant, and so provides us with fewer clues to
@@
 that could invalidate the experiment, and in
-section~\ref{sec:hugefrac} we saw that the experiment did indeed not
+Section~\ref{sec:hugefrac} we saw that the experiment did indeed not
 hold up to scrutiny.

### 2013-07-23 exp1/iswc2013.tex
Remove table 3 and describe it in the text
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@619 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 aware that caching may be important in 4store, as a preliminary
-experiment gave the results in Table~\ref{tab:cacheexp}. We see that the
-response time was significantly lower when an OPTIONAL clause was
-added to the query. As this is not more restrictive query, the most
-likely explanation is that the result of the evaluation of the Basic
-Graph Pattern was cached, so only the OPTIONAL left join was necessary
-to add in the second query.
-
-\begin{table}[ht]
-\begin{center}
-\caption{Two experiments with and without OPTIONAL clause.}\label{tab:cacheexp}
-\begin{tabular}{ccccccccr}
-  \hline
-``Implement'' & ``TripleC'' & ``Machine'' & ``BGPComp'' & ``Lang'' & ``Range'' & ``Union'' & ``Optional'' & Time \\
- \hline
-1   &    1  &     1  &     2 &   1 &    1   &  1  & 1 &  53.918049 \\
-1   &    1  &     1  &     2 &   1 &    1   &  1  & 2 &   1.252312 \\
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
+experiment gave a response time of 53.9~s of a query with ``BGPComp''
+at level~2 but other factors at level~1. When an OPTIONAL clause was
+added, the response time dropped to 1.25~s. As this is not more
+restrictive query, the most likely explanation is that the result of
+the evaluation of the Basic Graph Pattern was cached, so only the
+OPTIONAL left join was necessary to add in the second query.
 

### 2013-07-24 exp1/iswc2013.tex
A change from John
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@620 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 active factors and so we may treat it as a $2^2$ experiment with 8
-replications, but the resulting $p = 0.00098$ must be expected with
-such a small number of runs and thus higher variance.
+replications. The resulting $p = 0.00098$ is still very low, albeit
+larger than in the previous experiment. This also explains why there
+are so few significant effects, and had to be expected with such a
+small number of runs and thus higher variance.
 

### 2013-07-24 exp1/iswc2013.tex
improvements to DoE explanation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@622 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-After measurements are done, we seek to estimate \emph{effects}. To
-see how the \emph{main effects} can be estimated, consider a factor
-$A$. We proceed to find the averages $a_1$ and $a_2$ of all measurements at level~1 and
-on level~2 respectively. Then the main effect of $A$ is $a_2 - a_1$.
+We describe the influence of the choice of levels on the response in
+terms of \emph{effects}.  For a factor $A$ with two levels, we let
+$a_1$ and $a_2$ be the average response of all $2^{n-1}$ measurements
+with $A$ at level 1, and 2, respectively.  The main effect of $A$ is
+then defined as $a_2 - a_1$.
 
@@
 by comparing averages for equal versus non-equal levels of $A$ and
-$B$.
-
-This can be generalized to larger combinations of factors, but in
-practice, this analysis is done by fitting a linear model to the data
-with all factors and their interactions. It is shown in
-\cite{wu2009experiments} that the effect is twice the coefficients of
-the fit.
+$B$.  Details of this theory, and how to compute the effects in
+practice, using linear regression, are found in
+\cite{wu2009experiments}.
 
@@
 be found in
-figures~\ref{fig:fullnormal},~\ref{fig:frac32normal},~\ref{fig:frac64normal}~and~\ref{fig:frac64hugenormal}. The
+Figs.~\ref{fig:fullnormal} to~\ref{fig:frac64hugenormal}. The
 range in the $y$-axis is dictated by the number of runs.
@@
 ``Machine'' factor specifies a hostname, whereas ``TripleC'' and
-``Implement'' specifies a port number. On each host,
+``Implement'' specifies a port number.
 

### 2013-07-24 exp1/iswc2013.tex
move wraptable
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@623 43ef1190-90fd-4dd7-a769-d0121f02b80e
factor/level explanation.  2 -> two
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@624 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-First, note that the concept of parameters as considered above is
-known as \emph{factors} in DoE. For each factor, one may make a
-measurement of a \emph{response variable} at certain
-\emph{levels}. For example, one may measure the response time of a
-SPARQL endpoint by measuring it when it contains 1 or 2 million
-triples. Levels are not constrained to be continous variables, they
-can be discrete, or even two different instances of a class. The
-ordering is not essential, but it may contribute to how easy it is to
-understand the results, as we shall see. Then, the response time is
-the response variable, the factor is the number of triples and there
-are two levels, 1 or 2 million triples. We have chosen in this paper
-to only deal with the relatively straightforward aspects of the DoE
-formalism, and therefore constrain the experiment strictly to 2
-levels.
+In the experiments DoE is concerned with, a \emph{response variable}
+is measured under a various combinations of parameters.  These
+parameters are known as \emph{factors}.  For each factor, a range of
+possible values is fixed.  These values are known as \emph{levels} for
+this factor.  Levels are not constrained to be continuous variables, they
+can be discrete, or even two different instances of a class.
+Experiments are run by choosing a combination of levels
+for the factors.
+
+For example, one may measure the execution time of a SPARQL endpoint
+by measuring it when it contains 1 or 2 million triples.  Then, the
+execution time is the response variable, the factor is the number of
+triples and it has two levels, 1 or 2 million triples. We have chosen
+in this paper to only deal with the relatively straightforward aspects
+of the DoE formalism, and therefore constrain the experiment strictly
+to two levels.
 
@@
 
-Similarly, \emph{interaction effects} for 2 factors $AB$ are defined
+Similarly, \emph{interaction effects} for two factors $AB$ are defined
 by comparing averages for equal versus non-equal levels of $A$ and
@@
 
-% latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Fri May 10 16:31:49 2013
-% Modified by hand to remove 1's
-\begin{wraptable}{l}{0.5\textwidth}
-\caption{The magnitude of effects for some important main effects and interactions.}\label{tab:effects}
-\medskip
-\begin{tabular}{rr}
-  \hline
-Factors & Effect  \\ 
-  \hline
-Implement & $-$21.27 \\ 
-  Implement:Optional & $-$11.49 \\ 
-  Implement:Union & $-$6.21 \\ 
-  Implement:TripleC:Lang1 & 4.12 \\ 
-  TripleC:Lang & 4.20 \\ 
-  Implement:TripleC & 4.32 \\ 
-  Implement:BGPComp & 4.89 \\ 
-  Lang:Union & 5.16 \\ 
-  Implement:BGPComp:Lang & 5.16 \\ 
-  BGPComp:Lang & 5.25 \\ 
-  Implement:Lang:Union & 5.75 \\ 
-  TripleC & 7.06 \\ 
-  Implement:Lang & 7.73 \\ 
-  Lang & 8.62 \\ 
-  Optional & 11.30 \\ 
-  Union & 17.18 \\ 
-   \hline
-\end{tabular}
-\end{wraptable}
-
 To investigate whether the detrimental side-effects cancel the
@@
 replicated 4 times.
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Fri May 10 16:31:49 2013
+% Modified by hand to remove 1's
+\begin{wraptable}{l}{0.48\textwidth}
+\caption{The magnitude of effects for some important main effects and interactions.}\label{tab:effects}
+\bigskip\bigskip
+\begin{tabular}{rr}
+  \hline
+Factors & Effect  \\ 
+  \hline
+Implement & $-$21.27 \\ 
+  Implement:Optional & $-$11.49 \\ 
+  Implement:Union & $-$6.21 \\ 
+  Implement:TripleC:Lang1 & 4.12 \\ 
+  TripleC:Lang & 4.20 \\ 
+  Implement:TripleC & 4.32 \\ 
+  Implement:BGPComp & 4.89 \\ 
+  Lang:Union & 5.16 \\ 
+  Implement:BGPComp:Lang & 5.16 \\ 
+  BGPComp:Lang & 5.25 \\ 
+  Implement:Lang:Union & 5.75 \\ 
+  TripleC & 7.06 \\ 
+  Implement:Lang & 7.73 \\ 
+  Lang & 8.62 \\ 
+  Optional & 11.30 \\ 
+  Union & 17.18 \\ 
+   \hline
+\end{tabular}
+\end{wraptable}
 

### 2013-07-24 exp1/iswc2013.tex
typos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@625 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 The practice of benchmarking is widespread, both in industry to select
-the most appropriate implementation amongst several choices, and in
+the most appropriate implementation among several choices, and in
 academia as means to verify or refute an assertion on the performance
@@
 could have been the query without OPTIONAL that would have
-benefitted. Unless the effect of caching is cumulative throughout the
+benefited. Unless the effect of caching is cumulative throughout the
 experiment (which is possible), the randomization will have the effect

### 2013-07-25 exp1/iswc2013.tex
table and figure placement
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@626 43ef1190-90fd-4dd7-a769-d0121f02b80e
table and figure placement
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@627 43ef1190-90fd-4dd7-a769-d0121f02b80e
Under/overfull boxes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@628 43ef1190-90fd-4dd7-a769-d0121f02b80e
draft subtitle
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@629 43ef1190-90fd-4dd7-a769-d0121f02b80e
Comment out draft
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@630 43ef1190-90fd-4dd7-a769-d0121f02b8

@@
 
+%\subtitle{---Draft submitted to ISWC 2013---}
+
 \begin{document}
-
 \maketitle
@@
   factorial experiment}. This is done in Section~\ref{sec:full}.
-%TODO {Simple 8 x 2 overview table?}
 
@@
 
-Common experiments include comparing several different
+Common experiments include the comparison of several different
 implementations. However, we focus on a slightly different problem: We
@@
       \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
-      \verbatiminput{experiment/BGPComp-1}
-      & \verbatiminput{experiment/BGPComp-2}\\
-      \hline
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+      \end{verbatim}
+      & 
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+?s foaf:page ?o2 ;
+   dbpprop:subdivisionName ?o3 .
+?o3 skos:subject ?o4 ;
+   dbpprop:seat ?o5 ;
+   a ?c1 .\end{verbatim} 
+\\      \hline
     \end{tabular}
@@
     \textbf{``Union''} & \textbf{``Lang''} \\[-2ex]
-    \verbatiminput{experiment/Union-2} &
-    \verbatiminput{experiment/Lang-2} \\
-    \hline
+    \begin{verbatim}
+{
+  ?o1 dbpprop:longd ?long ;
+      dbpprop:latd  ?lat .
+} UNION {
+  ?o1 geo:long ?long ;
+      geo:lat ?lat .
+}\end{verbatim} 
+&
+\begin{verbatim}
+FILTER langMatches( lang(?l1), "fr" ) 
+\end{verbatim} 
+\\    \hline
     \textbf{``Range''} & \textbf{``Optional''} \\[-2ex]
-    \verbatiminput{experiment/Range-2} & 
-    \verbatiminput{experiment/Optional-2}\\
-    \hline
+    \begin{verbatim}
+FILTER (?l2 > 800000)\end{verbatim} 
+ & 
+    \begin{verbatim}
+OPTIONAL { ?o1 foaf:homepage ?o6 . }\end{verbatim} 
+\\    \hline
   \end{tabular}
@@
 
-\begin{table}[h!]
+\begin{table}[ht!]
 \begin{center}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{%
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{%
-    \includegraphics[width=0.9\textwidth]{frac64normal.pdf}}
+    \includegraphics[width=0.96\textwidth]{frac64normal.pdf}}
   \caption{A normal plot of the Fractional Factorial Experiment with
@@
 %\bibliographystyle{jbact}
-\bibliographystyle{splncs03}
-\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance,specs}
-
+%\bibliographystyle{splncs03}
+%\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance,specs}
+\input{iswc2013.bbl}
 \end{document}

### 2013-07-25 exp1/iswc2013.tex
Final proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@634 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
-  configuration} and one smaller with faster disk\footnote{Running
+  configuration} and one smaller with a faster disk\footnote{Running
   Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
@@
 \end{center}
-These fragments has been carefully designed for illustrative utility,
+These fragments have been carefully designed for illustrative utility,
 as well as suitable selectivity. They are not themselves important,
@@
 (unfortunately so many that some labels are illegible). We also note that
-``Implement'' itself is significantly negative. We also note that the
+``Implement'' itself is significantly negative and that the
 ``Implement:Lang'' interaction is significantly positive, as expected
@@
 optimization is significant, it is merely a simulation. It is to
-establish a critical practice of evaluations. It is of critical
-importance that we understand why some effects are significant. That
+establish a critical practice of evaluations. We must therefore
+understand why some effects are significant. That
 ``Implement'' is significant is hardly a surprise, that is what we
@@
 
-We lose explanatory power due to \emph{aliasing}, for example, the
+We lose explanatory power due to \emph{aliasing}; for example, the
 interaction labelled ``TripleC:BGPComp'' may be aliased with the

### 2013-08-05 exp1/iswc2013.tex
Helen's feedback
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@637 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-One important aspect of science is the structured approach to falsify
-hypotheses, and thus is important that assertions about performance
+One important aspect of science is the structured approach to falsifying
+hypotheses, and therefore is important that assertions about performance
 are formulated as hypotheses that can be falsified.
 
-Thus, attacking the complexity problem by eliminating parameters we
+Therefore, attacking the complexity problem by eliminating parameters we
 think influence the outcome (e.g. caching) for simplicity is untenable
@@
 in this paper to only deal with the relatively straightforward aspects
-of the DoE formalism, and therefore constrain the experiment strictly
+of the DoE formalism, and therefore constrained the experiment strictly
 to two levels.
@@
 fraction of the runs at the price of explanatory power. Such
-experiments are called \emph{fractional factorial experiments} and is
+experiments are called \emph{fractional factorial experiments} and are
 covered in Section~\ref{sec:frac}.
@@
 As the focus of this paper is didactical, the actual measurements are
-much simpler than what has been used in benchmarking. We have used the
+much simpler than those that have been used in benchmarking. We have used the
 data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but
@@
 performed a simulation that enables us to understand the effect of the
-changes.  In a real-world case, there is often the case that an
+changes.  In a real-world case, it is often the case that an
 optimization has negative side-effects, and we would need to simulate
@@
 
-Finally note that there are two common issues in benchmarking we do
+Finally, note that there are two common issues in benchmarking we do
 not consider: we do not allow a warm-up run, nor do we take into
@@
 We perform the test as a one-sided hypothesis test where $H_0$ is that
-the mean of the vectors are equal and $H_1$ is that mean of the
+the mean of the vectors are equal and $H_1$ is that mean sof the
 vectors representing the new implementation is lower. In this example,
@@
 implementation is an improvement, it is also interesting to know if
-there are cases that it fails. Again, we treat the experiment as a
+there are cases where it fails. Again, we treat the experiment as a
 $2^6$ experiment replicated 4 times. Now, for each of the 32 resulting
@@
 Table~\ref{tab:pvaluesfull}. We see that there are significant
-improvements in most cases, one may want to further investigate the
+improvements in most cases; one may want to further investigate the
 cases where they do not.
@@
 that require joins are strong. The strong ``BGPComp:Lang'' and
-``Lang:Union'' interactions might be due to that the many more triples
-needs to be searched for a language tag in one case. However, the
+``Lang:Union'' interactions might be due to that many more triples
+need to be searched for a language tag in one of the levels of the
+interacting factors. However, the
 positive ``Implement:TripleC'' interaction evades such explanations;
@@
 we see that whenever ``TripleC'' is at the low level, the
-new implementation is always best, only at the higher level, the old
+new implementation is always best. Only at the higher level, the old
 implementation may be better. This is a clear indication that the
@@
 We turn to the 64-run experiment and its normal plot in
-Figure~\ref{fig:frac64normal}. We see that the we get more significant
+Figure~\ref{fig:frac64normal}. We see that then we get more significant
 effects, and that they correspond well to those we saw in the full
@@
     the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
-    and crosses at $\alpha=0.05$. The use of different symbols have
+    and crosses at $\alpha=0.05$. The use of different symbols has
     been added to FrF2 by the authors.}\label{fig:frac64normal}
@@
 see that the interactions ``Implement:Lang'' and ``Implement:TripleC''
-is highly significant. It is clear what happened: For larger
+are highly significant. It is clear what happened: For larger
 ``TripleC'', the delay we introduced in the langMatches function has a
@@
 a good example of usage that also helps us manage complexity: If
-possible, one may group many factors into one if one is in doubt of
-its relevance and break it up in subsequent experiments if deeper
+possible, one may group many factors into one if in doubt of
+their relevance and break it up in subsequent experiments if deeper
 understanding is needed. To some extent, ``BGPComp'' is also a broad
@@
 at level~2 but other factors at level~1. When an OPTIONAL clause was
-added, the response time dropped to 1.25~s. As this is not more
+added, the response time dropped to 1.25~s. As this is not a more
 restrictive query, the most likely explanation is that the result of
@@
 
-We did this to illustrate \emph{randomization} as this is what that
+We did this to illustrate \emph{randomization} as this is what
 makes it permissible, i.e. the order of execution is random, so the
-benefit of caching is likely to apply randomly to different runs, for
+benefit of caching is likely to apply randomly to different runs. For
 example, had the run order been reversed in the previous example, it
@@
 so-called ``lurking variables'' such as caching, warm-up effects, etc,
-contributes to total unexplained variance of the experiment, but
+contribute to the total unexplained variance of the experiment, but
 should not skew the results to invalidate the estimated effects.
@@
 We saw in Section~\ref{sec:frac} that with a smaller experiment, fewer
-interactions are significant, and so provides us with fewer clues to
+interactions are significant, and so provide us with fewer clues to
 assess the soundness of the experiment. This should be a focus of
@@
 
-Finally, we have only employed the simplest parts of the DoE theory,
-what has been presented here is part of a much more general formalism
+Finally, we have only employed the simplest parts of the DoE theory.
+What has been presented here is part of a much more general formalism
 known as Orthogonal Arrays. The use of orthogonal arrays allow for
@@
 
-As evaluating SPARQL endpoints is inherently difficult a simplistic
+As evaluating SPARQL endpoints is inherently difficult, a simplistic
 experiment such as the one we designed should not hold up to
@@
 
-Also, experiments as small as the 32-run is not useful in estimating
-any detailed characteristics and is thus of little use for science,
+Also, experiments as small as the 32-run are not useful in estimating
+any detailed characteristics and are thus of little use for science,
 but could be useful in engineering: If the experiment has been
@@
 We have seen that we can perform a proper hypothesis test based on
-summary statistics, and by using our expertise, we showed how to show
+summary statistics, and by using our expertise, we showed how to reveal
 that the experiment was flawed, and we have cautioned that failure to

### 2013-08-05 exp1/iswc2013.tex
Comment on DoE SotA
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@638 43ef1190-90fd-4dd7-a769-d0121f02b80e
Slight reformulations
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@639 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Finally, we have only employed the simplest parts of the DoE theory.
+In this paper, we have only employed the simplest parts of the DoE theory.
 What has been presented here is part of a much more general formalism
@@
 complex problem of SPARQL endpoint evaluations.
+
+Finally, as each run is cheap, and the experiment can usually run
+without human intervention, we believe it is interesting as a case for
+advancing the state of the art of DoE.
 

### 2013-08-13 background/70percent.tex
Letter for 70% position
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@640 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass[norsk]{uioletter}
\usepackage[utf8]{inputenc}

\signature{Kjetil Kjernsmo}

\ouremail{kjekje@ifi.uio.no}

\begin{document}

\begin{letter}{Institutt for Informatikk}

\opening{Søknad om redusert stilling}

Grunnet et ønske om å tilbringe mer tid med egne barn søkes det herved
om å deltidsstilling fra dagens dato og ut høstsemesteret 2013. Det
ønskes å sette ned stillingsprosenten til 70\%, og at stipendet
forlenges tilsvarende.

\closing{Med vennlig hilsen,}

\end{letter}
\end{document}

### 2013-09-30 exp1/present2013.tex
Stub presentation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@643 43ef1190-90fd-4dd7-a769-d0121f02b80e

%\documentclass[english,aspectratio=169]]{ifislide}
\documentclass[english,print,aspectratio=169]{ifislide}
%\documentclass[english,handout,aspectratio=169]{ifislide}

\usepackage[utf8]{inputenc}
% Packages used specific for this lecture:
\usepackage{tikz}
%\usepackage{bussproofs}
%\usepackage{textpos}

\EnableBpAbbreviations

\usetikzlibrary{arrows,shapes,mindmap}
\usetikzlibrary{decorations.pathmorphing}
\tikzstyle{resource}=[ellipse,draw=black!40,fill=blue!20,thick, font=\ttfamily]
\tikzstyle{blank}=[ellipse,draw=black!40,fill=white!40,thick, font=\ttfamily]
\tikzstyle{literal}=[rectangle,text centered,draw=black!40,fill=black!20,thick]
\tikzstyle{predicate}=[sloped,above,text centered,midway,font={\scriptsize\ttfamily}]
\tikzstyle{predicateb}=[left,midway,font={\scriptsize\ttfamily}]
\tikzstyle{class}=[rectangle,text centered,draw=black!40,fill=white,thick]
\usetikzlibrary{arrows,automata}

\newcommand{\typed}{\char`\^\char`\^}
\newcommand{\blank}{\char`\_}

\newcommand{\tuple}[1]{\left<#1\right>}
\newcommand{\inter}{\mathcal{I}}
\newcommand{\domain}{\mathop{\mathsf{dom}}}
\newcommand{\range}{\mathop{\mathsf{rg}}}

% set lecture number, date, title:
%\LectureNumber{13}
\LectureDate{23.10.2013}
\LectureTitle{Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation}
% set name of lecturer:
\author{Kjetil Kjernsmo}
\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\IfiTitleSlide     % inserts title frame

\IfiTOCSlide       % inserts today's plan frame

\section{Introduction}

\begin{frame}{Experimental tradition}

Teh Tycho

\end{frame}

\end{document}

### 2013-09-30 exp1/present2013.tex
Comment out some stuff
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@644 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\documentclass[english,aspectratio=169]]{ifislide}
-\documentclass[english,print,aspectratio=169]{ifislide}
-%\documentclass[english,handout,aspectratio=169]{ifislide}
+%\documentclass[english,print,aspectratio=169]{ifislide}
+\documentclass[english,handout,aspectratio=169]{ifislide}
 
@@
 % Packages used specific for this lecture:
-\usepackage{tikz}
+
 %\usepackage{bussproofs}
@@
 
-\EnableBpAbbreviations
+%\EnableBpAbbreviations
+%% \usepackage{tikz}
 
-\usetikzlibrary{arrows,shapes,mindmap}
-\usetikzlibrary{decorations.pathmorphing}
-\tikzstyle{resource}=[ellipse,draw=black!40,fill=blue!20,thick, font=\ttfamily]
-\tikzstyle{blank}=[ellipse,draw=black!40,fill=white!40,thick, font=\ttfamily]
-\tikzstyle{literal}=[rectangle,text centered,draw=black!40,fill=black!20,thick]
-\tikzstyle{predicate}=[sloped,above,text centered,midway,font={\scriptsize\ttfamily}]
-\tikzstyle{predicateb}=[left,midway,font={\scriptsize\ttfamily}]
-\tikzstyle{class}=[rectangle,text centered,draw=black!40,fill=white,thick]
-\usetikzlibrary{arrows,automata}
+%% \usetikzlibrary{arrows,shapes,mindmap}
+%% \usetikzlibrary{decorations.pathmorphing}
+%% \tikzstyle{resource}=[ellipse,draw=black!40,fill=blue!20,thick, font=\ttfamily]
+%% \tikzstyle{blank}=[ellipse,draw=black!40,fill=white!40,thick, font=\ttfamily]
+%% \tikzstyle{literal}=[rectangle,text centered,draw=black!40,fill=black!20,thick]
+%% \tikzstyle{predicate}=[sloped,above,text centered,midway,font={\scriptsize\ttfamily}]
+%% \tikzstyle{predicateb}=[left,midway,font={\scriptsize\ttfamily}]
+%% \tikzstyle{class}=[rectangle,text centered,draw=black!40,fill=white,thick]
+%% \usetikzlibrary{arrows,automata}
 
@@
 
-\IfiTOCSlide       % inserts today's plan frame
+%\IfiTOCSlide       % inserts today's plan frame
 

### 2013-10-03 exp1/present2013.tex
Add a figure and a table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@645 43ef1190-90fd-4dd7-a769-d0121f02b80e
A start
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@646 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\begin{frame}{What's a benchmark?}
+\end{frame}
+\note{If your queries are all fully known, by all means, run them}
+\note{In most scientific relevant cases, no such assumptions can be
+  made}
+
+\begin{frame}{Problems with benchmarking}
+\end{frame}
+\note{If you standardize hardware, you can't test claims about
+  improved hardware. You'd need interactions too}
+\note{Cache-buster is evil}
+
+\begin{frame}{Design of Experiments}
+\end{frame}
+
+\begin{frame}{Full normal plot}
+\begin{figure}[t]
+  \centerline{%
+  \includegraphics[width=.9\textwidth]{fullnormal.pdf}}
+  \caption{A normal plot of the Full Factorial Experiment. The
+    labelled points are considered significant by using the Lenth
+    criterion at a level $\alpha=0.05$.}\label{fig:fullnormal}
+\end{figure}
+
+\end{frame}
+\note{Don't worry about the axes}
+
+\begin{frame}
+\begin{table}[ht!]
+\begin{center}
+\caption{$p$-values for different parts of the experiment}\label{tab:pvaluesfull}
+\begin{tabular}{cccccl}
+  \hline
+``TripleC'' & ``BGPComp'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
+  \hline
+  1 & 1 & 1 & 1 & 1 & 0.012 \\ 
+  1 & 1 & 1 & 1 & 2 & $1.4 \cdot 10^{-09}$ \\ 
+  1 & 1 & 1 & 2 & 1 & $3.1 \cdot 10^{-09}$ \\ 
+  1 & 1 & 1 & 2 & 2 & $6.1 \cdot 10^{-11}$ \\ 
+  1 & 1 & 2 & 1 & 1 & $3.3 \cdot 10^{-06}$ \\ 
+  1 & 1 & 2 & 1 & 2 & $2.7 \cdot 10^{-09}$ \\ 
+  1 & 1 & 2 & 2 & 1 & $2 \cdot 10^{-06}$ \\ 
+  1 & 1 & 2 & 2 & 2 & $3.6 \cdot 10^{-10}$ \\ 
+  1 & 2 & 1 & 1 & 1 & 0.014 \\ 
+  1 & 2 & 1 & 1 & 2 & $1.2 \cdot 10^{-10}$ \\ 
+  1 & 2 & 1 & 2 & 1 & $2.8 \cdot 10^{-14}$ \\ 
+  1 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
+  1 & 2 & 2 & 1 & 1 & $2.1 \cdot 10^{-05}$ \\ 
+  1 & 2 & 2 & 1 & 2 & $2.7 \cdot 10^{-07}$ \\ 
+  1 & 2 & 2 & 2 & 1 & 0.0072 \\ 
+  1 & 2 & 2 & 2 & 2 & $1.6 \cdot 10^{-05}$ \\ 
+  2 & 1 & 1 & 1 & 1 & 0.28 \\ 
+  2 & 1 & 1 & 1 & 2 & $3 \cdot 10^{-07}$ \\ 
+  2 & 1 & 1 & 2 & 1 & $3.3 \cdot 10^{-07}$ \\ 
+  2 & 1 & 1 & 2 & 2 & $1.7 \cdot 10^{-08}$ \\ 
+  2 & 1 & 2 & 1 & 1 & 0.0023 \\ 
+  2 & 1 & 2 & 1 & 2 & $6.5 \cdot 10^{-07}$ \\ 
+  2 & 1 & 2 & 2 & 1 & 0.00032 \\ 
+  2 & 1 & 2 & 2 & 2 & $1.3 \cdot 10^{-06}$ \\ 
+  2 & 2 & 1 & 1 & 1 & 0.013 \\ 
+  2 & 2 & 1 & 1 & 2 & $1 \cdot 10^{-11}$ \\ 
+  2 & 2 & 1 & 2 & 1 & $3.8 \cdot 10^{-11}$ \\ 
+  2 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
+  2 & 2 & 2 & 1 & 1 &   1 \\ 
+  2 & 2 & 2 & 1 & 2 & $2.3 \cdot 10^{-05}$ \\ 
+  2 & 2 & 2 & 2 & 1 &   1 \\ 
+  2 & 2 & 2 & 2 & 2 & 0.99 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\end{frame}
+
 \end{document}

### 2013-10-04 exp1/present2013.tex
Key concepts
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@647 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{frame}{Problems with benchmarking}
+
+\begin{itemize}
+\item Large number of parameters to test for complex scenarios.
+\item No structured approach to investigate flaws.
+\item No meaningful summary of the overall performance.
+\item Standardized benchmarks cannot test assertions outside of the
+  standard.
+\item Attempting to neutralize the effect of certain optimization
+  techniques oversimplifies the test.
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Design of Experiments}
+
+  \begin{itemize}
+  \item Pioneered by Fischer in the 1930-ties.
+  \item Well-established in many fields of engineering, agriculture,
+    medicine.
+  \item Well-suited to manage very complex experiments.
+  \item Requires parameterization of experiments.
+  \end{itemize}
+\end{frame}
+\note{How many is familiar with DoE?}
+
+\subsection{Key Concepts Design of Experiments}
+
+\begin{frame}{Response variable}
+
+% TODO Format well
+   A \emph{response variable} is measured under various
+    combinations of parameters.
+
+For example:
+\begin{itemize}
+\item Throughput
+\item Query execution time
+\item Response time
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Factor}
+
+Such parameters are called \emph{factor}s.
+
+For example:
+\begin{itemize}
+\item Hardware platform
+\item Data heterogeneity
+\item Number of triples in the store
+\item Absence or presence of certain language features
+\item Order of experiments
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Levels}
+
+For each factor, a range of possible values are fixed. These values
+are called \emph{levels}.
+
+For example:
+\begin{itemize}
+\item For number of triples, the levels could be 1 or 2 MTriples.
+\item For language features, the levels could be absence or presence
+  of an \texttt{OPTIONAL} pattern. 
+\end{itemize}
+
+Levels may be continuous, discrete, two different instances of a
+class, etc.
+
+\end{frame}
+
+\begin{frame}{Design Matrix}
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Fri Oct  4 22:47:01 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{r|rrrrrrrr}
+  \hline
+ & ``Implement'' & ``TripleC'' & ``Machine'' & ``BGPComp'' & ``Lang'' & ``Range'' & ``Union'' & ``Optional'' \\ 
+  \hline
+1 & 2 & 2 & 1 & 2 & 1 & 2 & 2 & 2 \\ 
+  2 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\ 
+  3 & 1 & 2 & 1 & 2 & 2 & 2 & 1 & 2 \\ 
+  4 & 1 & 1 & 2 & 1 & 2 & 2 & 1 & 1 \\ 
+  5 & 1 & 1 & 1 & 2 & 1 & 2 & 2 & 2 \\ 
+  6 & 1 & 2 & 1 & 2 & 2 & 1 & 1 & 1 \\ 
+  7 & 1 & 1 & 2 & 2 & 1 & 2 & 2 & 1 \\ 
+  8 & 2 & 1 & 2 & 2 & 2 & 2 & 1 & 2 \\ 
+ ...  \\
+  256 & 1 & 1 & 2 & 2 & 1 & 1 & 1 & 1 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 \end{frame}

### 2013-10-05 exp1/present2013.tex
Effects and simpler design matrix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@648 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{itemize}
+\item A concrete implementation
 \item Hardware platform
@@
 \begin{itemize}
+\item For implementation, 4store or Virtuoso.
 \item For number of triples, the levels could be 1 or 2 MTriples.
-\item For language features, the levels could be absence or presence
-  of an \texttt{OPTIONAL} pattern. 
+\item For language features, the levels could be \textsf{SELECT} and \textsf{CONSTRUCT}
 \end{itemize}
 
-Levels may be continuous, discrete, two different instances of a
+Levels may be continuous, discrete, different instances of a
 class, etc.
@@
 \begin{frame}{Design Matrix}
+
 % latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Fri Oct  4 22:47:01 2013
+% Sat Oct  5 22:16:51 2013
 \begin{table}[ht]
 \begin{center}
-\begin{tabular}{r|rrrrrrrr}
+\begin{tabular}{r|ccc}
   \hline
- & ``Implement'' & ``TripleC'' & ``Machine'' & ``BGPComp'' & ``Lang'' & ``Range'' & ``Union'' & ``Optional'' \\ 
+ & Implement & TripleC & Union \\ 
   \hline
-1 & 2 & 2 & 1 & 2 & 1 & 2 & 2 & 2 \\ 
-  2 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\ 
-  3 & 1 & 2 & 1 & 2 & 2 & 2 & 1 & 2 \\ 
-  4 & 1 & 1 & 2 & 1 & 2 & 2 & 1 & 1 \\ 
-  5 & 1 & 1 & 1 & 2 & 1 & 2 & 2 & 2 \\ 
-  6 & 1 & 2 & 1 & 2 & 2 & 1 & 1 & 1 \\ 
-  7 & 1 & 1 & 2 & 2 & 1 & 2 & 2 & 1 \\ 
-  8 & 2 & 1 & 2 & 2 & 2 & 2 & 1 & 2 \\ 
- ...  \\
-  256 & 1 & 1 & 2 & 2 & 1 & 1 & 1 & 1 \\ 
+1 & 2 & 1 & 2 \\ 
+  2 & 2 & 2 & 2 \\ 
+  3 & 2 & 1 & 1 \\ 
+  4 & 1 & 1 & 1 \\ 
+  5 & 2 & 2 & 1 \\ 
+  6 & 1 & 2 & 1 \\ 
+  7 & 1 & 2 & 2 \\ 
+  8 & 1 & 1 & 2 \\ 
    \hline
@@
 \end{table}
+
 \end{frame}
+\note{There are 3 factors, 8 experiments, which is $2^3$}
+
+\begin{frame}{Factorial experiments}
+
+  \begin{itemize}
+  \item Full factorial experiments:
+    \begin{itemize}
+    \item 2 levels give $2^n$ combinations (\emph{runs}), where $n$ is
+      the number of factors.
+    \item 3 levels give $3^n$ runs, etc.
+    \end{itemize}
+  \item Fractional factorial experiments:
+
+    \begin{itemize}
+    \item When experimental economy is important,
+    \item DoE theory offers extensive facilities for running a
+      fraction of the runs ($2^{n-1}$ etc).
+    \item The price is of smaller experiments is explanatory power.
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Main effects}
+
+\emph{Effects} describe the influence of the levels on the response.
+
+The the \emph{main effect}: For a factor $A$ with two levels, we let
+$a_1$ and $a_2$ be the average response of all $2^{n-1}$ measurements
+with $A$ at level 1, and 2, respectively.  The main effect of $A$ is
+then defined as $a_2 - a_1$.
+
+For example:
+
+For each implementation, average all the responses. Then, the main
+effect of the implementation is the difference between them.
+
+\end{frame}
+
+\begin{frame}{Interaction effects}
+
+\emph{Interaction effects} for two factors $AB$ are defined
+by comparing averages for equal versus non-equal levels of $A$ and
+$B$.
+
+For example:
+
+For each implementation, find the averages for each dataset size, the
+interaction effect is % TODO: Clear example.
+
+In practice, Quite Easily Done using linear regression.
+\end{frame}
+\note{Interaction effects may be complex, like one implementation is
+  better on CONSTRUCT queries, on small datasets}
+\note{We still need to understand which effects are important and
+  significant, we do that in the discussion of our experiment]
 

### 2013-10-05 exp1/present2013.tex
The factors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@649 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \section{Introduction}
+
+\begin{frame}{Objective}
+
+To introduce a path to critical practice of evaluations, that makes
+use of contemporary statistical techniques, to establish a practice
+that can be used to refute assertions on performance.
+
+The focus is didactical, and the experiment we present is a toy
+example.
+
+\end{frame}
 
@@
 
+% SECTION
+
+\section{Experiments}
+
+\begin{frame}{Constraints on scope of experiment}
+
+Above all: Easy to understand for newcomers to DoE!
+
+\begin{itemize}
+\item Just 2 levels.
+\item Handful of illustrative factors.
+\item Analysis that illustrates key concepts.
+\item Straightforward to program and reproduce.
+\item No actual optimization under study.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Choices to meet these constraints}
+
+  \begin{itemize}
+  \item Response variable: Time from DNS lookup finishes to endpoint
+    has delivered a full response, measured by \texttt{curl}.
+  \item We don't compare different implementions, we compare the
+    performance before and after some change.
+  \item Query engine: \texttt{4store} where we insert \textsf{sleep}
+    statements in the join function on one level and language matching
+    function on the other. 
+  \end{itemize}
+
+\end{frame}
+
+\begin{frame}{Factors and levels}
+  \begin{itemize}
+  \item [``Implement''] The implementation under evaluation.
+  \item [``TripleC''] 1 or 2 MTriples from the DBPedia SPARQL
+    Benchmark.
+  \item [``Machine''] Software and hardware platform, one smaller
+    machine with fast SSD and one larger with slower disks in RAID1.
+  \item [``BGPComp''] Two different basic graph patterns with varying
+    complexity.
+  \item [``Union''] Presence or absence of a \textsf{UNION} pattern.
+  \item [``Lang''] Presence or absence of a \textsf{langMatches}
+    filter.
+  \item [``Range''] Presence or absence of a filter with a larger-than
+    operator.
+  \item [``Optional''] Presence or absence of an \textsf{OPTIONAL} pattern.
+  \end{itemize}
+\end{frame}
+
 \begin{frame}{Full normal plot}

### 2013-10-08 exp1/present2013.tex
Small change
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@650 43ef1190-90fd-4dd7-a769-d0121f02b80e
queries and github mention
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@651 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \documentclass[english,handout,aspectratio=169]{ifislide}
+\usepackage{verbatim}
 
@@
 \note{We still need to understand which effects are important and
-  significant, we do that in the discussion of our experiment]
+  significant, we do that in the discussion of our experiment}
 
@@
 \item Straightforward to program and reproduce.
-\item No actual optimization under study.
+\item Artificial performance differences.
 \end{itemize}
@@
     machine with fast SSD and one larger with slower disks in RAID1.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{The Basic Graph pattern factor}
+  \begin{itemize}
+
   \item [``BGPComp''] Two different basic graph patterns with varying
     complexity.
-  \item [``Union''] Presence or absence of a \textsf{UNION} pattern.
-  \item [``Lang''] Presence or absence of a \textsf{langMatches}
+
+  \begin{center}\small
+    \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.40\textwidth}|}
+      \hline
+      \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+      \end{verbatim}
+      & 
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+?s foaf:page ?o2 ;
+   dbpprop:subdivisionName ?o3 .
+?o3 skos:subject ?o4 ;
+   dbpprop:seat ?o5 ;
+   a ?c1 .
+\end{verbatim} 
+\\      \hline
+    \end{tabular}
+  \end{center}
+
+  \end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]{Absence or presence factors}
+  \begin{itemize}
+  \item [``Union''] Absence or presence of a \textsf{UNION} pattern.
+
+    \begin{verbatim}
+{ ?o1 dbpprop:longd ?long ;
+      dbpprop:latd  ?lat .
+} UNION {
+  ?o1 geo:long ?long ;
+      geo:lat ?lat . }
+\end{verbatim}
+
+  \item [``Lang''] Absence or presence of a \textsf{langMatches}
     filter.
-  \item [``Range''] Presence or absence of a filter with a larger-than
+
+\begin{verbatim}
+FILTER langMatches( lang(?l1), "fr" ) 
+\end{verbatim}
+
+  \item [``Range''] Absence or presence of a filter with a larger-than
     operator.
-  \item [``Optional''] Presence or absence of an \textsf{OPTIONAL} pattern.
+
+    \begin{verbatim}
+FILTER (?l2 > 800000)
+\end{verbatim}
+
+  \item [``Optional''] Absence or presence of an \textsf{OPTIONAL}
+    pattern.
+    \begin{verbatim}
+OPTIONAL { ?o1 foaf:homepage ?o6 . }
+\end{verbatim}
+
   \end{itemize}
 \end{frame}
+\note{Implemented in R, using packages \texttt{DoE.base} and \texttt{FrF2}.}
+\note{Two experiment machines, one machine running queries}
+\note{No warm-up, no cache busting}
 
@@
 
+\begin{frame}{Thank you!}
+
+  \begin{itemize}
+  \item Code and instructions at Github:
+    \url{https://github.com/kjetilk/doe-sparql}.
+  \item Happy to help getting it to run!
+  \end{itemize}
+
+\end{frame}
+
 \end{document}

### 2013-10-08 exp1/present2013.tex
add effects table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@652 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{verbatim}
+\usepackage{wrapfig}
 
@@
   \centerline{%
-  \includegraphics[width=.9\textwidth]{fullnormal.pdf}}
-  \caption{A normal plot of the Full Factorial Experiment. The
-    labelled points are considered significant by using the Lenth
-    criterion at a level $\alpha=0.05$.}\label{fig:fullnormal}
+  \includegraphics[height=0.85\textheight]{fullnormal.pdf}}
 \end{figure}
+% TODO side by side
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Fri May 10 16:31:49 2013
+% Modified by hand to remove 1's
+\begin{wraptable}{l}{0.2\textwidth}
+\begin{tabular}{rr}
+  \hline
+Factors & Effect  \\ 
+  \hline
+Implement & $-$21.27 \\ 
+  Implement:Optional & $-$11.49 \\ 
+  Implement:Union & $-$6.21 \\ 
+  Implement:TripleC:Lang1 & 4.12 \\ 
+  TripleC:Lang & 4.20 \\ 
+  Implement:TripleC & 4.32 \\ 
+  Implement:BGPComp & 4.89 \\ 
+  Lang:Union & 5.16 \\ 
+  Implement:BGPComp:Lang & 5.16 \\ 
+  BGPComp:Lang & 5.25 \\ 
+  Implement:Lang:Union & 5.75 \\ 
+  TripleC & 7.06 \\ 
+  Implement:Lang & 7.73 \\ 
+  Lang & 8.62 \\ 
+  Optional & 11.30 \\ 
+  Union & 17.18 \\ 
+   \hline
+\end{tabular}
+\end{wraptable}
 
 \end{frame}
+% TODO explain Lenth
 \note{Don't worry about the axes}
+\note{Yes, the Implementation is better, we knew that, joins are
+  naturally strong}
+\note{``Union'', ``Optional'' and ``Lang'' demanding things}
 

### 2013-10-08 exp1/iswc2013.tex
Correct post-publication typo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@653 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 We perform the test as a one-sided hypothesis test where $H_0$ is that
-the mean of the vectors are equal and $H_1$ is that mean sof the
+the mean of the vectors are equal and $H_1$ is that means of the
 vectors representing the new implementation is lower. In this example,

### 2013-10-08 exp1/present2013.tex
full expirment now ok
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@654 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \note{No warm-up, no cache busting}
+\note{256 experiments in full factorial, no replications, compare
+  favorably to benchmarks}
+
+\subsection{Full Factorial Experiment}
 
@@
 \note{``Union'', ``Optional'' and ``Lang'' demanding things}
+\note{``BGPComp:Lang'' and ``Lang:Union'' interactions might be due to
+  that many more triples need to be searched for a language tag in one
+  of the levels of the interacting factors.}
+\note{But what about ``Implement::TripleC''?}
+\note{We want to do a hypothesis test}
+\note{Two factors do not contribute significantly to the variation
+  ``Range'' and ``Machine''}
 
-\begin{frame}
-\begin{table}[ht!]
+\begin{frame}{Hypothesis tests}
+
+Hypothesis formulation:
+  \begin{itemize}
+  \item $H_0$: The new implementation is no better than the old.
+  \item $H_1$: The new implementation is better than the old.
+  \end{itemize}
+
+Classifying factors:
+  \begin{itemize}
+  \item ``Range'' and ``Machine'' are called \emph{inactive factors}.
+  \item ``BGPComp'', ``Lang'', ``Optional'', ``Union'' and ``TripleC''
+    are called \emph{environmental factors}.
+  \item ``Implement'' is called a \emph{control factor}.
+  \end{itemize}
+
+\end{frame}
+\begin{frame}{Hypothesis tests}
+
+\begin{itemize}
+\item Average over all the 32 level-combinations of the environmental
+  factors.
+
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Tue Oct  8 14:51:28 2013
+\begin{table}[ht]
 \begin{center}
-\caption{$p$-values for different parts of the experiment}\label{tab:pvaluesfull}
-\begin{tabular}{cccccl}
+\begin{tabular}{rlll|r}
   \hline
-``TripleC'' & ``BGPComp'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
+ & Implement & Machine & Range & Response \\ 
   \hline
-  1 & 1 & 1 & 1 & 1 & 0.012 \\ 
-  1 & 1 & 1 & 1 & 2 & $1.4 \cdot 10^{-09}$ \\ 
-  1 & 1 & 1 & 2 & 1 & $3.1 \cdot 10^{-09}$ \\ 
-  1 & 1 & 1 & 2 & 2 & $6.1 \cdot 10^{-11}$ \\ 
-  1 & 1 & 2 & 1 & 1 & $3.3 \cdot 10^{-06}$ \\ 
-  1 & 1 & 2 & 1 & 2 & $2.7 \cdot 10^{-09}$ \\ 
-  1 & 1 & 2 & 2 & 1 & $2 \cdot 10^{-06}$ \\ 
-  1 & 1 & 2 & 2 & 2 & $3.6 \cdot 10^{-10}$ \\ 
-  1 & 2 & 1 & 1 & 1 & 0.014 \\ 
-  1 & 2 & 1 & 1 & 2 & $1.2 \cdot 10^{-10}$ \\ 
-  1 & 2 & 1 & 2 & 1 & $2.8 \cdot 10^{-14}$ \\ 
-  1 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
-  1 & 2 & 2 & 1 & 1 & $2.1 \cdot 10^{-05}$ \\ 
-  1 & 2 & 2 & 1 & 2 & $2.7 \cdot 10^{-07}$ \\ 
-  1 & 2 & 2 & 2 & 1 & 0.0072 \\ 
-  1 & 2 & 2 & 2 & 2 & $1.6 \cdot 10^{-05}$ \\ 
-  2 & 1 & 1 & 1 & 1 & 0.28 \\ 
-  2 & 1 & 1 & 1 & 2 & $3 \cdot 10^{-07}$ \\ 
-  2 & 1 & 1 & 2 & 1 & $3.3 \cdot 10^{-07}$ \\ 
-  2 & 1 & 1 & 2 & 2 & $1.7 \cdot 10^{-08}$ \\ 
-  2 & 1 & 2 & 1 & 1 & 0.0023 \\ 
-  2 & 1 & 2 & 1 & 2 & $6.5 \cdot 10^{-07}$ \\ 
-  2 & 1 & 2 & 2 & 1 & 0.00032 \\ 
-  2 & 1 & 2 & 2 & 2 & $1.3 \cdot 10^{-06}$ \\ 
-  2 & 2 & 1 & 1 & 1 & 0.013 \\ 
-  2 & 2 & 1 & 1 & 2 & $1 \cdot 10^{-11}$ \\ 
-  2 & 2 & 1 & 2 & 1 & $3.8 \cdot 10^{-11}$ \\ 
-  2 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
-  2 & 2 & 2 & 1 & 1 &   1 \\ 
-  2 & 2 & 2 & 1 & 2 & $2.3 \cdot 10^{-05}$ \\ 
-  2 & 2 & 2 & 2 & 1 &   1 \\ 
-  2 & 2 & 2 & 2 & 2 & 0.99 \\ 
+1 & 1 & 1 & 1 & 33.47 \\ 
+  2 & 2 & 1 & 1 & 11.82 \\ 
+  3 & 1 & 2 & 1 & 32.16 \\ 
+  4 & 2 & 2 & 1 & 11.56 \\ 
+  5 & 1 & 1 & 2 & 33.97 \\ 
+  6 & 2 & 1 & 2 & 12.46 \\ 
+  7 & 1 & 2 & 2 & 32.90 \\ 
+  8 & 2 & 2 & 2 & 11.60 \\ 
    \hline
@@
 
+\item May create a one-sided two-sample t-test with 4 values for each of the levels.
+\item Lends support to $H_1$ with a high probability, $p=1.16 \cdot
+10^{-07}$.
+
+\end{itemize}
+
 \end{frame}
+\note{The paper shows how similar testing can be done to investigate
+  when the new implementation fails.} % TODO: Add more on this?
 

### 2013-10-08 exp1/present2013.tex
fractional experiments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@655 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Fractional Factorial Experiments}
+
+\begin{frame}{Experimental Economy}
+
+  \begin{itemize}
+  \item Full factorial experimental goes as $2^n$.
+  \item SPARQL Endpoint evaluation is inherently complex, with many
+    possible factors.
+  \item May reduce the size of the experiment with Fractional
+    Factorial Experiments.
+  \item The cost is explanatory power due to \emph{aliasing}.
+  \item E.g. we cannot tell the difference of an effect of
+    ``TripleC::BGPComp'' from ``Machine::Range''
+  \item Easy in R to declare which effects must not be aliased.
+  \item We have made two designs: One 32-run and one 64-run.
+  \end{itemize}
+
+\end{frame}
+\note{ That is to say, a detected increase in
+run time from larger Basic Graph Patterns for large databases, can
+also be explained by a less powerful machine that is worse at
+evaluating FILTER clauses with ranges. }
+
+\begin{frame}{32-run experiment}
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[height=0.85\textheight,width=.9\textwidth]{frac32normal.pdf}}
+  \label{fig:frac32normal}
+\end{figure}
+
+\end{frame}
+\note{The total variance is great, due to few runs, so few conclusions
+  can be drawn}
+\note{Implementation is good, and Union is hard}
+\note{In the paper, we also do a hypothesis test that returns  $p =
+  0.00098$. Still very low}
+\note{Very rough conclusions; useful in a CI system}
+
+\begin{frame}{64-run experiment}
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[height=0.85\textheight,width=0.96\textwidth]{frac64normal.pdf}}
+\end{figure}
+
+\end{frame}
+\note{Have two Lenth criterions, $\alpha=0.15$ and crosses at
+  $\alpha=0.05$. ($\alpha$ is the probability of rejecting $H_0$ when
+  it is true)}
+
+\begin{frame}{Analysis of 64-run experiment}
+
+  \begin{itemize}
+  \item Find most of the same significant effects as the full experiment.
+  \item However, we see that ``Implement:TripleC'' emerges only for
+  $\alpha=0.15$.
+  \item ``Machine'', ``TripleC'' and ``Range'' are inactive, we can
+    perform hypothesis test as above, and find $p = 5.9 \cdot 10^{-6}$.
+  \end{itemize}
+\end{frame}
+\note{We see that ``Implement:TripleC'' emerges only for
+  $\alpha=0.15$, which is important to better in real studies.}
+\note{We have shown that how the experiment could be set up, we have
+  shown the analysis and a hypothesis test.}
+\note{Should this simplistic experiment stand up to scrutiny?}
+
+\subsection{Tearing it all down}
+
+\begin{frame}{Fractional experiment with more data}
+\begin{figure}[ht!]
+  \centerline{%
+  \includegraphics[height=0.85\textheight,width=.9\textwidth]{frac64hugenormal.pdf}}
+\end{figure}
+
+\end{frame}
+\note{``Machine'' is a broad factor, good idea to group to manage
+  complexity, can be broken down later}
+\note{``TripleC:Machine'' and ``TripleC'' highly significant, implies
+  the factors were wrongly set.}
+\note{This invalidates the three preceeding experiments. We got strong
+indications early on.}
+
+\section{Conclusion}
+
 \begin{frame}{Thank you!}

### 2013-10-08 exp1/present2013.tex
randomization
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@656 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Discussion}
+
+\begin{frame}{Linearity and choice of levels}
+  \begin{itemize}
+  \item 2-level experiments can only model linear effects.
+  \item Choice of levels may be critical.
+  \item If response is known to be non-linear, using more levels must be considered.
+  \end{itemize}
+
+\end{frame}
+\note{2-level good enough for identifying effects, insufficient as
+  statistical model}
+
+\begin{frame}{The role of randomization}
+  \begin{itemize}
+  \item We have allowed us to disregard the effect of caching,
+    warm-up-runs, etc.
+  \item This is different from neutralizing the effect!
+  \item They are now ``lurking variables''.
+  \item They contribute to the overall unexplained variance.
+  \item Unexplained variance should be kept to a minimum.
+  \item Time-dependent or order-dependent factors may be required.
+  \end{itemize}
+\end{frame}
+\note{Even though we know caching is important}
+\note{Caching is good practice, shouldn't be neutralized. Warm-up may
+  not be permissible in real-time scenarios}
+\note{Variance can obscure problems like ``Implement:TripleC''}
+\note{most problematic cases are those that are
+covered neither by the broad factors, nor the lurking variables, or
+any of the specific factors. Invalidated by absence of language
+features, such as solution modifiers.}
+
+\section{Future Work}
+
 \section{Conclusion}
+
+\begin{frame}{Rules of Thumb}
+
+\begin{enumerate}
+\item Are there factors that cover all realistic features?
+\item If not, are they adequately covered by randomization?
+\item If so, would the variance resulting from randomization obscure
+  factors that could provide clues that the levels are wrongly set?
+\item By carefully examining interactions with ``Implement'', are
+  there any that are unaccounted for, and that could point out wrongly
+  set levels?
+\end{enumerate}
+
+\end{frame}
 

### 2013-10-08 exp1/present2013.tex
future work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@657 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \author{Kjetil Kjernsmo}
-\AtBeginSection[]
-{
-  \begin{frame}
-    \frametitle{Outline}
-    \tableofcontents[currentsection]
-  \end{frame}
-}
 
@@
 
+\begin{frame}{Choice of factors}
+
+  \begin{itemize}
+  \item Different strategies must be tried, open field.
+  \item Parameterization is very important.
+    \begin{itemize}
+    \item Data heterogeneity, skewed distributions, etc.
+    \item Parameterizing SPARQL queries, see e.g. SPLODGE by Görlitz
+      et al.
+    \item Parameterizing SPARQL queries using the grammar pragmatically?
+    \end{itemize}
+  \end{itemize}
+
+\end{frame}
+
+\begin{frame}{Other important issues}
+
+  \begin{itemize}
+  \item Ensuring that small and efficient experiments provide enough
+    details to assess the soundness of the experiment.
+  \item This paper scratches the surface of DoE, Orthogonal Arrays is
+    a more general formalism.
+  \item This topic is also interesting for progress beyond the
+    state-of-the-art in statistics.
+  \end{itemize}
+
+\end{frame}
+
 \section{Conclusion}

### 2013-10-09 exp1/present2013.tex
conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@658 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 example.
-
-\end{frame}
-
-\begin{frame}{Experimental tradition}
-
-Teh Tycho
 
@@
 \item Order of experiments
+\item Tuning parameters
+\item Concurrency
 \end{itemize}
@@
 \end{frame}
+\note{Optimization parameters, hardware characteristics}
 
@@
   \begin{itemize}
+  \item Comparing completely different SPARQL implementations.
   \item Ensuring that small and efficient experiments provide enough
@@
 
-\begin{frame}{Rules of Thumb}
+\begin{frame}
+We saw
+  \begin{itemize}
+  \item ... an experimental setup using DoE.
+  \item ... how the analysis pointed out the important effects.
+  \item ... how we got a comprehensive view of the experiment.
+  \item ... how hypothesis tests could be formulated.
+  \item ... how 2-level experiments can determine significant effects,
+    even though it is not good enough for modelling.
+  \item ... how more economic experiments can be designed, and noted
+    their limitations.
+  \item ... finally how the experiment could be shown to be flawed.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Questions to ask}
 
@@
 \end{enumerate}
+
+\end{frame}
+
+\begin{frame}{What's a worthy result?}
+  \begin{itemize}
+  \item I have not made this a primary focus of my work.
+  \item Even though I think I should.
+  \item 3000 full-time researchers to construct the ATLAS experiment
+    at CERN.
+  \item Improvement in experimental methodology should be a worthy result.
+  \end{itemize}
 

### 2013-10-09 exp1/present2013.tex
about benchmarks
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@659 43ef1190-90fd-4dd7-a769-d0121f02b80e
use descripiton instead of itemize
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@660 43ef1190-90fd-4dd7-a769-d0121f02b80e
show notes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@661 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-%\documentclass[english,aspectratio=169]]{ifislide}
+%\documentclass[english,aspectratio=196]]{ifislide}
 %\documentclass[english,print,aspectratio=169]{ifislide}
@@
 \usepackage{wrapfig}
-
+\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
 \begin{frame}{What's a benchmark?}
+
+  \begin{itemize}
+  \item Natural benchmarks.
+  \item Benchmarking with broad and well-known workloads. 
+  \item Use-case driven benchmarks.
+  \end{itemize}
+
 \end{frame}
+\note{Natural benchmarks, query log-derived, DBPedia SPARQL Benchmark}
 \note{If your queries are all fully known, by all means, run them}
@@
   made}
+\note{What I will present will likely create artifical workloads}
+\note{Use-case driven will seem artificial for everything else}
 
@@
 \begin{frame}{Factors and levels}
-  \begin{itemize}
+  \begin{description}
   \item [``Implement''] The implementation under evaluation.
@@
     machine with fast SSD and one larger with slower disks in RAID1.
-  \end{itemize}
+  \end{description}
 \end{frame}
@@
 \begin{frame}[fragile]{The Basic Graph pattern factor}
-  \begin{itemize}
+  \begin{description}
 
@@
 
-  \end{itemize}
+  \end{description}
 
@@
 \begin{frame}[fragile]{Absence or presence factors}
-  \begin{itemize}
+  \begin{description}
   \item [``Union''] Absence or presence of a \textsf{UNION} pattern.
@@
 
-  \end{itemize}
+  \end{description}
 \end{frame}
@@
 \note{The paper shows how similar testing can be done to investigate
-  when the new implementation fails.} % TODO: Add more on this?
+  when the new implementation fails.}
 

### 2013-10-11 exp1/present2013.tex
Lenths method
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@662 43ef1190-90fd-4dd7-a769-d0121f02b80e
Martin's comments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@663 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{frame}{Objective}
-
-To introduce a path to critical practice of evaluations, that makes
+  \begin{itemize}
+  \item To introduce a path to critical practice of evaluations, that makes
 use of contemporary statistical techniques, to establish a practice
 that can be used to refute assertions on performance.
+\item The focus is didactical, and the experiment we present is a toy
+example.
+\item To apply a well established method in statistics, rarely used in
+  Computer Science, to SPARQL endpoint evaluation.
 
-The focus is didactical, and the experiment we present is a toy
-example.
+  \end{itemize}
 
@@
 \end{frame}
-% TODO explain Lenth
+\note{Unreplicated, so we don't have a standard error, so Lenth used
+  a restricted subset the median of the effects to create a Pseudo
+  Standard Error}
 \note{Don't worry about the axes}

### 2013-10-14 exp1/present2013.tex
title slide
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@664 43ef1190-90fd-4dd7-a769-d0121f02b80e
Full normal plot 1 page
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@665 43ef1190-90fd-4dd7-a769-d0121f02b80e
Full normal plot 1 page
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@666 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{wrapfig}
-\setbeameroption{show notes}
+%\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
 \LectureDate{23.10.2013}
-\LectureTitle{Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation}
-% set name of lecturer:
-\author{Kjetil Kjernsmo}
 
@@
 
-\IfiTitleSlide     % inserts title frame
+%\IfiTitleSlide     % inserts title frame
 
 %\IfiTOCSlide       % inserts today's plan frame
+
+\title[Statistical Design]{Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation}
+\subtitle{}
+\author{Kjetil Kjernsmo and John Tyssedal}
+\maketitle
 
@@
   \begin{itemize}
-  \item Pioneered by Fischer in the 1930-ties.
+  \item Pioneered by Fischer in the 1930s.
   \item Well-established in many fields of engineering, agriculture,
@@
 
+\begin{block}
   For example:
@@
   \end{itemize}
-
+\end{block}
 Levels may be continuous, discrete, different instances of a
@@
     complexity.
+  \end{description}
 
@@
   \end{center}
-
-  \end{description}
 
@@
 \begin{frame}{Full normal plot}
-\begin{figure}[t]
-  \centerline{%
-  \includegraphics[height=0.85\textheight]{fullnormal.pdf}}
-\end{figure}
-% TODO side by side
-% latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Fri May 10 16:31:49 2013
-% Modified by hand to remove 1's
-\begin{wraptable}{l}{0.2\textwidth}
+  \begin{minipage}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{fullnormal.pdf}
+  \end{minipage}
+  %
+  \begin{minipage}{0.49\textwidth}\small
+    \vspace{-2ex}\qquad
     \begin{tabular}{rr}
       \hline
-Factors & Effect  \\ 
+      \textbf{Factors} & \textbf{Effect}  \\ 
       \hline
@@
     \end{tabular}
-\end{wraptable}
+  \end{minipage}
+\end{frame}
 
-\end{frame}
 \note{Unreplicated, so we don't have a standard error, so Lenth used

### 2013-10-14 exp1/present2013.tex
example -> block
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@667 43ef1190-90fd-4dd7-a769-d0121f02b80e
Fix date and shorttitle
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@668 43ef1190-90fd-4dd7-a769-d0121f02b80e
do some reformatting
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@669 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\LectureNumber{13}
-\LectureDate{23.10.2013}
+\LectureDate{24.10.2013}
 
@@
 
-\title[Statistical Design]{Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation}
+\title[DoE in SPARQL Evaluations]{Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation}
 \subtitle{}
@@
 
-\begin{frame}{What's a benchmark?}
+\begin{frame}{Not all benchmarks are created equal}
 
@@
   \item Benchmarking with broad and well-known workloads. 
-  \item Use-case driven benchmarks.
+  \item Artificial use-case driven benchmarks.
   \end{itemize}
@@
 
-% TODO Format well
+ \Large
    A \emph{response variable} is measured under various
     combinations of parameters.
+ \normalsize
 
-For example:
+    \begin{block}{Examples}
       \begin{itemize}
@@
       \end{itemize}
+    \end{block}
 
@@
 
+\Large
 Such parameters are called \emph{factor}s. 
+\normalsize
 
-For example:
+\begin{block}{Examples}
   \begin{itemize}
@@
   \end{itemize}
+\end{block}
 
@@
 
+\Large
 For each factor, a range of possible values are fixed. These values
 are called \emph{levels}.
+\normalsize
 
-\begin{block}
-  For example:
+\begin{block}{Examples}
   \begin{itemize}

### 2013-10-14 exp1/present2013.tex
fix the effects slide
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@670 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{frame}{Main effects}
+\begin{frame}{Effects}
 
+\Larger 
 \emph{Effects} describe the influence of the levels on the response.
+\normalsize
 
-The the \emph{main effect}: For a factor $A$ with two levels, we let
-$a_1$ and $a_2$ be the average response of all $2^{n-1}$ measurements
-with $A$ at level 1, and 2, respectively.  The main effect of $A$ is
-then defined as $a_2 - a_1$.
+\begin{itemize}
+\item Similarities to regression.
+\item \emph{Main effects} consider the factors on their own.
+\item \emph{Interaction effects} consider factors given levels of
+  other factors.
+\item In practice done with linear regression routines.
+\end{itemize}
 
-For example:
-
-For each implementation, average all the responses. Then, the main
-effect of the implementation is the difference between them.
-
-\end{frame}
-
-\begin{frame}{Interaction effects}
-
-\emph{Interaction effects} for two factors $AB$ are defined
-by comparing averages for equal versus non-equal levels of $A$ and
-$B$.
-
-For example:
-
-For each implementation, find the averages for each dataset size, the
-interaction effect is % TODO: Clear example.
-
-In practice, Quite Easily Done using linear regression.
 \end{frame}

### 2013-10-14 exp1/present2013.tex
bug
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@671 43ef1190-90fd-4dd7-a769-d0121f02b80e
Fix the notes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@672 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{wrapfig}
-%\setbeameroption{show notes}
+\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
 \end{frame}
-\note{Natural benchmarks, query log-derived, DBPedia SPARQL Benchmark}
-\note{If your queries are all fully known, by all means, run them}
-\note{In most scientific relevant cases, no such assumptions can be
-  made}
-\note{What I will present will likely create artifical workloads}
-\note{Use-case driven will seem artificial for everything else}
+\note{
+  \begin{itemize}
+  \item Natural benchmarks, query log-derived, DBPedia SPARQL
+    Benchmark
+  \item If your queries are all fully known, by all means, run them
+  \item In most scientific relevant cases, no such assumptions can be
+    made
+  \item What I will present will likely create artifical workloads
+  \item Use-case driven will seem artificial for everything else
+  \end{itemize}
+}
 
@@
 \end{frame}
-\note{If you standardize hardware, you can't test claims about
-  improved hardware. You'd need interactions too}
-\note{Cache-buster is evil}
+\note{  
+  \begin{itemize}
+  \item If you standardize hardware, you can't test claims about
+  improved hardware. You'd need interactions too
+  \item Cache-buster is evil
+\end{itemize}
+
+}
 
@@
 
-\Larger 
+\Large
 \emph{Effects} describe the influence of the levels on the response.
@@
 \end{frame}
-\note{Interaction effects may be complex, like one implementation is
-  better on CONSTRUCT queries, on small datasets}
-\note{We still need to understand which effects are important and
-  significant, we do that in the discussion of our experiment}
 
+\note{
+
+  \begin{itemize}
+  \item Interaction effects may be complex, like one implementation is
+    better on CONSTRUCT queries, on small datasets
+\item We still need to understand which effects are important and
+  significant, we do that in the discussion of our experiment
+\end{itemize}
+}
 % SECTION
@@
 \end{frame}
-\note{Implemented in R, using packages \texttt{DoE.base} and \texttt{FrF2}.}
-\note{Two experiment machines, one machine running queries}
-\note{No warm-up, no cache busting}
-\note{256 experiments in full factorial, no replications, compare
-  favorably to benchmarks}
+
+\note{
+
+  \begin{itemize}
+  \item Implemented in R, using packages \texttt{DoE.base} and
+    \texttt{FrF2}.
+  \item Two experiment machines, one machine running queries
+  \item No warm-up, no cache busting
+  \item 256 experiments in full factorial, no replications, compare
+    favorably to benchmarks
+  \end{itemize}
+}
 
@@
 
-\note{Unreplicated, so we don't have a standard error, so Lenth used
+\note{
+  \begin{itemize}
+  \item Unreplicated, so we don't have a standard error, so Lenth used
     a restricted subset the median of the effects to create a Pseudo
-  Standard Error}
-\note{Don't worry about the axes}
-\note{Yes, the Implementation is better, we knew that, joins are
-  naturally strong}
-\note{``Union'', ``Optional'' and ``Lang'' demanding things}
-\note{``BGPComp:Lang'' and ``Lang:Union'' interactions might be due to
-  that many more triples need to be searched for a language tag in one
-  of the levels of the interacting factors.}
-\note{But what about ``Implement::TripleC''?}
-\note{We want to do a hypothesis test}
-\note{Two factors do not contribute significantly to the variation
-  ``Range'' and ``Machine''}
+    Standard Error
+  \item Don't worry about the axes
+  \item Yes, the Implementation is better, we knew that, joins are
+    naturally strong
+  \item ``Union'', ``Optional'' and ``Lang'' demanding things
+  \item ``BGPComp:Lang'' and ``Lang:Union'' interactions might be due
+    to that many more triples need to be searched for a language tag
+    in one of the levels of the interacting factors.
+  \item But what about ``Implement::TripleC''?
+  \item We want to do a hypothesis test
+  \item Two factors do not contribute significantly to the variation
+    ``Range'' and ``Machine''
+  \end{itemize}
+}
 
@@
 \end{frame}
-\note{The total variance is great, due to few runs, so few conclusions
-  can be drawn}
-\note{Implementation is good, and Union is hard}
-\note{In the paper, we also do a hypothesis test that returns  $p =
-  0.00098$. Still very low}
-\note{Very rough conclusions; useful in a CI system}
+\note{
+
+  \begin{itemize}
+  \item The total variance is great, due to few runs, so few
+    conclusions can be drawn
+  \item Implementation is good, and Union is hard
+  \item In the paper, we also do a hypothesis test that returns $p =
+    0.00098$. Still very low
+  \item Very rough conclusions; useful in a CI system
+  \end{itemize}
+}
 
@@
 \end{frame}
-\note{We see that ``Implement:TripleC'' emerges only for
-  $\alpha=0.15$, which is important to better in real studies.}
-\note{We have shown that how the experiment could be set up, we have
-  shown the analysis and a hypothesis test.}
-\note{Should this simplistic experiment stand up to scrutiny?}
+\note{
+  \begin{itemize}
+  \item We see that ``Implement:TripleC'' emerges only for
+    $\alpha=0.15$, which is important to better in real studies.
+  \item We have shown that how the experiment could be set up, we have
+    shown the analysis and a hypothesis test.
+  \item Should this simplistic experiment stand up to scrutiny?
+  \end{itemize}
+}
 
@@
 \end{frame}
-\note{``Machine'' is a broad factor, good idea to group to manage
-  complexity, can be broken down later}
-\note{``TripleC:Machine'' and ``TripleC'' highly significant, implies
-  the factors were wrongly set.}
-\note{This invalidates the three preceeding experiments. We got strong
-indications early on.}
+\note{
+  \begin{itemize}
+  \item ``Machine'' is a broad factor, good idea to group to manage
+    complexity, can be broken down later
+  \item ``TripleC:Machine'' and ``TripleC'' highly significant,
+    implies the factors were wrongly set.
+  \item This invalidates the three preceeding experiments. We got
+    strong indications early on.
+  \end{itemize}
+}
 
@@
 \end{frame}
-\note{Even though we know caching is important}
-\note{Caching is good practice, shouldn't be neutralized. Warm-up may
-  not be permissible in real-time scenarios}
-\note{Variance can obscure problems like ``Implement:TripleC''}
-\note{most problematic cases are those that are
-covered neither by the broad factors, nor the lurking variables, or
-any of the specific factors. Invalidated by absence of language
-features, such as solution modifiers.}
+\note{
+  \begin{itemize}
+  \item Even though we know caching is important
+  \item Caching is good practice, shouldn't be neutralized. Warm-up
+    may not be permissible in real-time scenarios
+  \item Variance can obscure problems like ``Implement:TripleC''
+  \item most problematic cases are those that are covered neither by
+    the broad factors, nor the lurking variables, or any of the
+    specific factors. Invalidated by absence of language features,
+    such as solution modifiers.
+  \end{itemize}
+}
 

### 2013-10-14 exp1/present2013.tex
remove comments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@673 43ef1190-90fd-4dd7-a769-d0121f02b80e
some cuts
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@674 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{verbatim}
-\usepackage{wrapfig}
-\setbeameroption{show notes}
+
+%\setbeameroption{show only notes}
 \usepackage[utf8]{inputenc}
-% Packages used specific for this lecture:
-
-%\usepackage{bussproofs}
-%\usepackage{textpos}
-
-%\EnableBpAbbreviations
-%% \usepackage{tikz}
-
-%% \usetikzlibrary{arrows,shapes,mindmap}
-%% \usetikzlibrary{decorations.pathmorphing}
-%% \tikzstyle{resource}=[ellipse,draw=black!40,fill=blue!20,thick, font=\ttfamily]
-%% \tikzstyle{blank}=[ellipse,draw=black!40,fill=white!40,thick, font=\ttfamily]
-%% \tikzstyle{literal}=[rectangle,text centered,draw=black!40,fill=black!20,thick]
-%% \tikzstyle{predicate}=[sloped,above,text centered,midway,font={\scriptsize\ttfamily}]
-%% \tikzstyle{predicateb}=[left,midway,font={\scriptsize\ttfamily}]
-%% \tikzstyle{class}=[rectangle,text centered,draw=black!40,fill=white,thick]
-%% \usetikzlibrary{arrows,automata}
 
@@
 
-% set lecture number, date, title:
-%\LectureNumber{13}
 \LectureDate{24.10.2013}
@@
 \begin{document}
-
-%\IfiTitleSlide     % inserts title frame
-
-%\IfiTOCSlide       % inserts today's plan frame
 
@@
 
-\begin{frame}{Not all benchmarks are created equal}
-
-  \begin{itemize}
-  \item Natural benchmarks.
-  \item Benchmarking with broad and well-known workloads. 
-  \item Artificial use-case driven benchmarks.
-  \end{itemize}
-
-\end{frame}
-\note{
-  \begin{itemize}
-  \item Natural benchmarks, query log-derived, DBPedia SPARQL
-    Benchmark
-  \item If your queries are all fully known, by all means, run them
-  \item In most scientific relevant cases, no such assumptions can be
-    made
-  \item What I will present will likely create artifical workloads
-  \item Use-case driven will seem artificial for everything else
-  \end{itemize}
-}
-
 \begin{frame}{Problems with benchmarking}
@@
   improved hardware. You'd need interactions too
-  \item Cache-buster is evil
+  \item Caching is a good practice, and its effect need to be a part
+    of the test, not gotten rid of.
+  \item That said benchmarking is still nice to have if you know what
+    queries you will run, it can't replace that.
 \end{itemize}
@@
   \begin{itemize}
-  \item Pioneered by Fischer in the 1930s.
+  \item Pioneered by Fischer in the 1920s.
   \item Well-established in many fields of engineering, agriculture,
@@
 
-\begin{frame}{Factorial experiments}
-
-  \begin{itemize}
-  \item Full factorial experiments:
-    \begin{itemize}
-    \item 2 levels give $2^n$ combinations (\emph{runs}), where $n$ is
-      the number of factors.
-    \item 3 levels give $3^n$ runs, etc.
-    \end{itemize}
-  \item Fractional factorial experiments:
-
-    \begin{itemize}
-    \item When experimental economy is important,
-    \item DoE theory offers extensive facilities for running a
-      fraction of the runs ($2^{n-1}$ etc).
-    \item The price is of smaller experiments is explanatory power.
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
 \begin{frame}{Effects}
@@
 }
+
+\begin{frame}{Factorial experiments}
+
+  \begin{itemize}
+  \item Full factorial experiments:
+    \begin{itemize}
+    \item 2 levels give $2^n$ combinations (\emph{runs}), where $n$ is
+      the number of factors.
+    \item 3 levels give $3^n$ runs, etc.
+    \end{itemize}
+  \item Fractional factorial experiments:
+
+    \begin{itemize}
+    \item When experimental economy is important,
+    \item DoE theory offers extensive facilities for running a
+      fraction of the runs ($2^{n-1}$ etc).
+    \item The price is of smaller experiments is explanatory power.
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
 % SECTION
@@
 \end{frame}
+\note{Go through very quickly}
 
@@
 
-\note{
-
-  \begin{itemize}
-  \item Implemented in R, using packages \texttt{DoE.base} and
-    \texttt{FrF2}.
-  \item Two experiment machines, one machine running queries
-  \item No warm-up, no cache busting
-  \item 256 experiments in full factorial, no replications, compare
-    favorably to benchmarks
-  \end{itemize}
-}
+\note{Implemented in R, system for statistics.}
 
@@
   \item Don't worry about the axes
-  \item Yes, the Implementation is better, we knew that, joins are
+  \item Yes, the Implementation is better, we knew that, 
+  \item Interactions with Implement with joins are
     naturally strong
@@
     in one of the levels of the interacting factors.
-  \item But what about ``Implement::TripleC''?
+  \item But what about ``Implement::TripleC''? Leave that hanging for
+    a bit.
   \item We want to do a hypothesis test
@@
 
-\begin{frame}{Hypothesis tests}
+\begin{frame}{Preparing hypothesis tests}
 
@@
 
-\item May create a one-sided two-sample t-test with 4 values for each of the levels.
+\item May create a one-sided two-sample t-test with 4 values for each
+  of the levels of ``Implement''.
 \item Lends support to $H_1$ with a high probability, $p=1.16 \cdot
-10^{-07}$.
+10^{-7}$.
 
@@
   \item We see that ``Implement:TripleC'' emerges only for
-    $\alpha=0.15$, which is important to better in real studies.
+    $\alpha=0.15$, which is important to improve when designing a
+    study for actual tests.
   \item We have shown that how the experiment could be set up, we have
@@
 \section{Discussion}
-
-\begin{frame}{Linearity and choice of levels}
-  \begin{itemize}
-  \item 2-level experiments can only model linear effects.
-  \item Choice of levels may be critical.
-  \item If response is known to be non-linear, using more levels must be considered.
-  \end{itemize}
-
-\end{frame}
-\note{2-level good enough for identifying effects, insufficient as
-  statistical model}
 

### 2013-10-15 exp1/present2013.tex
introduce style Martin uses
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@675 43ef1190-90fd-4dd7-a769-d0121f02b80e
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@676 43ef1190-90fd-4dd7-a769-d0121f02b80e
More math
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@677 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\documentclass[english,print,aspectratio=169]{ifislide}
-\documentclass[english,handout,aspectratio=169]{ifislide}
+%\documentclass[english,handout,aspectratio=169]{ifislide}
+\documentclass[english,usenames,dvipsnames,aspectratio=169]{beamer}
 \usepackage{verbatim}
+\usepackage{config}
+
+\usetheme{ifi_en}
 
@@
 \usepackage[utf8]{inputenc}
-
-\newcommand{\typed}{\char`\^\char`\^}
-\newcommand{\blank}{\char`\_}
-
-\newcommand{\tuple}[1]{\left<#1\right>}
-\newcommand{\inter}{\mathcal{I}}
-\newcommand{\domain}{\mathop{\mathsf{dom}}}
-\newcommand{\range}{\mathop{\mathsf{rg}}}
-
-\LectureDate{24.10.2013}
 
@@
 \author{Kjetil Kjernsmo and John Tyssedal}
+\date{24.10.2013}
 \maketitle
@@
 \end{frame}
-\note{How many is familiar with DoE?}
+\note{
+
+  \begin{itemize}
+  \item There are problems, but time doesn't allow discussion, more in paper.
+  \item How many are familiar with DoE?
+  \end{itemize}
+}
 
@@
 
+    \begin{block}{For the mathematically inclined}
+      The response variable is a stochastic variable $R$
+    \end{block}
+
 \end{frame}
@@
   \begin{itemize}
-  \item A concrete implementation
-  \item Hardware platform
-  \item Data heterogeneity
+  \item Which concrete implementation
+  \item Which Hardware platform
   \item Number of triples in the store
@@
   \item Order of experiments
-  \item Tuning parameters
-  \item Concurrency
+  \item Level of concurrency
   \end{itemize}
+\end{block}
+
+\begin{block}{For the mathematically inclined}
+  The response variable is itself a function $R \( a, b, c, \ldots
+  \)$, where $a$, $b$, $c$ are factors.
 \end{block}
@@
   \item Happy to help getting it to run!
+  \item Happy to discuss collaborations or workshops on this topic!
   \end{itemize}

### 2013-10-15 exp1/present2013.tex
some math additions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@678 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   \end{itemize}
+
+  \begin{block}{Message for the takeaway}
+    If you develop techniques for performance enhancements of complex
+    systems, you need to think carefully about how you prove that 
+    you've achieved something of general validity. 
+  \end{block}
 
@@
   \item Which concrete implementation
-  \item Which Hardware platform
+  \item Which hardware platform
   \item Number of triples in the store
@@
 \begin{block}{For the mathematically inclined}
-  The response variable is itself a function $R \( a, b, c, \ldots
-  \)$, where $a$, $b$, $c$ are factors.
+  The response variable is itself a function 
+  $R ( A, B, C, \ldots ) $, 
+  where $A$, $B$, $C$ etc. are factors.
 \end{block}
@@
 \item \emph{Main effects} consider the factors on their own.
+\begin{block}{For the mathematically inclined}
+  $ME(A) = \bar{z}(A_2) - \bar{z}(A_1)$ where $\bar{z}(A_n)$ is the
+  average of all values observed for factor $A$ at level $n$.
+\end{block}
+
 \item \emph{Interaction effects} consider factors given levels of
@@
   \item Different strategies must be tried, open field.
+  \item Formal methods from AI?
   \item Parameterization is very important.

### 2013-10-15 exp1/present2013.tex
cut fractionals
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@679 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     systems, you need to think carefully about how you prove that 
-    you've achieved something of general validity. 
+    you've achieved something of general validity. This is my proposal
+    in that direction.
   \end{block}
@@
   techniques oversimplifies the test.
+\item \emph{A benchmark is an engineering tool, not a scientific experiment.}
 \end{itemize}
@@
   \begin{itemize}
-  \item Unreplicated, so we don't have a standard error, so Lenth used
-    a restricted subset the median of the effects to create a Pseudo
-    Standard Error
+  \item Those that are deemed significant are labelled.
   \item Don't worry about the axes
@@
 \end{frame}
-\note{The paper shows how similar testing can be done to investigate
-  when the new implementation fails.}
-
-\subsection{Fractional Factorial Experiments}
-
-\begin{frame}{Experimental Economy}
-
-  \begin{itemize}
-  \item Full factorial experimental goes as $2^n$.
-  \item SPARQL Endpoint evaluation is inherently complex, with many
-    possible factors.
-  \item May reduce the size of the experiment with Fractional
-    Factorial Experiments.
-  \item The cost is explanatory power due to \emph{aliasing}.
-  \item E.g. we cannot tell the difference of an effect of
-    ``TripleC::BGPComp'' from ``Machine::Range''
-  \item Easy in R to declare which effects must not be aliased.
-  \item We have made two designs: One 32-run and one 64-run.
-  \end{itemize}
-
-\end{frame}
-\note{ That is to say, a detected increase in
-run time from larger Basic Graph Patterns for large databases, can
-also be explained by a less powerful machine that is worse at
-evaluating FILTER clauses with ranges. }
-
-\begin{frame}{32-run experiment}
-\begin{figure}[ht!]
-  \centerline{%
-    \includegraphics[height=0.85\textheight,width=.9\textwidth]{frac32normal.pdf}}
-  \label{fig:frac32normal}
-\end{figure}
-
-\end{frame}
-\note{
-
-  \begin{itemize}
-  \item The total variance is great, due to few runs, so few
-    conclusions can be drawn
-  \item Implementation is good, and Union is hard
-  \item In the paper, we also do a hypothesis test that returns $p =
-    0.00098$. Still very low
-  \item Very rough conclusions; useful in a CI system
-  \end{itemize}
-}
-
-\begin{frame}{64-run experiment}
-\begin{figure}[ht!]
-  \centerline{%
-    \includegraphics[height=0.85\textheight,width=0.96\textwidth]{frac64normal.pdf}}
-\end{figure}
-
-\end{frame}
-\note{Have two Lenth criterions, $\alpha=0.15$ and crosses at
-  $\alpha=0.05$. ($\alpha$ is the probability of rejecting $H_0$ when
-  it is true)}
-
-\begin{frame}{Analysis of 64-run experiment}
-
-  \begin{itemize}
-  \item Find most of the same significant effects as the full experiment.
-  \item However, we see that ``Implement:TripleC'' emerges only for
-  $\alpha=0.15$.
-  \item ``Machine'', ``TripleC'' and ``Range'' are inactive, we can
-    perform hypothesis test as above, and find $p = 5.9 \cdot 10^{-6}$.
-  \end{itemize}
-\end{frame}
 \note{  
   \begin{itemize}
-  \item We see that ``Implement:TripleC'' emerges only for
-    $\alpha=0.15$, which is important to improve when designing a
-    study for actual tests.
-  \item We have shown that how the experiment could be set up, we have
-    shown the analysis and a hypothesis test.
+  \item The paper shows how similar testing can be done to investigate
+  when the new implementation fails.} 
+\item The paper has two fractional factorial experiments.
   \item Should this simplistic experiment stand up to scrutiny?

### 2013-10-15 exp1/present2013.tex
cut factors
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@680 43ef1190-90fd-4dd7-a769-d0121f02b80e
cut a bit from hypothesis test
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@681 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     machine with fast SSD and one larger with slower disks in RAID1.
-  \end{description}
-\end{frame}
-
-\begin{frame}[fragile]{The Basic Graph pattern factor}
-  \begin{description}
-
   \item [``BGPComp''] Two different basic graph patterns with varying
     complexity.
-  \end{description}
-
-  \begin{center}\small
-    \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.40\textwidth}|}
-      \hline
-      \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
-      \begin{verbatim}
-?s  rdfs:label ?l1 ;
-    ?p1 ?o1 .
-?o1 dbo:populationTotal ?l2 .
-      \end{verbatim}
-      & 
-      \begin{verbatim}
-?s  rdfs:label ?l1 ;
-    ?p1 ?o1 .
-?o1 dbo:populationTotal ?l2 .
-?s foaf:page ?o2 ;
-   dbpprop:subdivisionName ?o3 .
-?o3 skos:subject ?o4 ;
-   dbpprop:seat ?o5 ;
-   a ?c1 .
-\end{verbatim} 
-\\      \hline
-    \end{tabular}
-  \end{center}
-
-\end{frame}
-
-\begin{frame}[fragile]{Absence or presence factors}
-  \begin{description}
   \item [``Union''] Absence or presence of a \textsf{UNION} pattern.
-
-    \begin{verbatim}
-{ ?o1 dbpprop:longd ?long ;
-      dbpprop:latd  ?lat .
-} UNION {
-  ?o1 geo:long ?long ;
-      geo:lat ?lat . }
-\end{verbatim}
-
   \item [``Lang''] Absence or presence of a \textsf{langMatches}
     filter.
-
-\begin{verbatim}
-FILTER langMatches( lang(?l1), "fr" ) 
-\end{verbatim}
-
   \item [``Range''] Absence or presence of a filter with a larger-than
     operator.
-
-    \begin{verbatim}
-FILTER (?l2 > 800000)
-\end{verbatim}
-
   \item [``Optional''] Absence or presence of an \textsf{OPTIONAL}
     pattern.
-    \begin{verbatim}
-OPTIONAL { ?o1 foaf:homepage ?o6 . }
-\end{verbatim}
 
@@
 
-\begin{frame}{Preparing hypothesis tests}
+\begin{frame}{Hypothesis tests}
 
@@
 
-\end{frame}
-\begin{frame}{Hypothesis tests}
+Tests to run:
+\begin{itemize}
 
-\begin{itemize}
 \item Average over all the 32 level-combinations of the environmental
   factors.
-
-% latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Tue Oct  8 14:51:28 2013
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{rlll|r}
-  \hline
- & Implement & Machine & Range & Response \\ 
-  \hline
-1 & 1 & 1 & 1 & 33.47 \\ 
-  2 & 2 & 1 & 1 & 11.82 \\ 
-  3 & 1 & 2 & 1 & 32.16 \\ 
-  4 & 2 & 2 & 1 & 11.56 \\ 
-  5 & 1 & 1 & 2 & 33.97 \\ 
-  6 & 2 & 1 & 2 & 12.46 \\ 
-  7 & 1 & 2 & 2 & 32.90 \\ 
-  8 & 2 & 2 & 2 & 11.60 \\ 
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
 
@@
   \item The paper shows how similar testing can be done to investigate
-  when the new implementation fails.} 
+  when the new implementation fails.
 \item The paper has two fractional factorial experiments.
@@
     implies the factors were wrongly set.
-  \item This invalidates the three preceeding experiments. We got
+  \item This invalidates the preceeding experiments. We got
     strong indications early on.

### 2013-10-15 exp1/present2013.tex
remove classifying remove questions remove worthy result
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@682 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \end{frame}
+\note{We first study a full factorial, in the paper we have three
+  fractional, we shall see the final fractional}
 
@@
 
-Classifying factors:
-  \begin{itemize}
-  \item ``Range'' and ``Machine'' are called \emph{inactive factors}.
-  \item ``BGPComp'', ``Lang'', ``Optional'', ``Union'' and ``TripleC''
-    are called \emph{environmental factors}.
-  \item ``Implement'' is called a \emph{control factor}.
-  \end{itemize}
-
 Tests to run:
@@
   when the new implementation fails.
-\item The paper has two fractional factorial experiments.
+
   \item Should this simplistic experiment stand up to scrutiny?
@@
   \centerline{%
-  \includegraphics[height=0.85\textheight,width=.9\textwidth]{frac64hugenormal.pdf}}
+  \includegraphics[height=0.8\textheight,width=.9\textwidth]{frac64hugenormal.pdf}}
 \end{figure}
@@
 
-\section{Conclusion}
-
-\begin{frame}
+\begin{frame}{Conclusion}
 We saw
@@
 
-\begin{frame}{Questions to ask}
-
-\begin{enumerate}
-\item Are there factors that cover all realistic features?
-\item If not, are they adequately covered by randomization?
-\item If so, would the variance resulting from randomization obscure
-  factors that could provide clues that the levels are wrongly set?
-\item By carefully examining interactions with ``Implement'', are
-  there any that are unaccounted for, and that could point out wrongly
-  set levels?
-\end{enumerate}
-
-\end{frame}
-
-\begin{frame}{What's a worthy result?}
-  \begin{itemize}
-  \item I have not made this a primary focus of my work.
-  \item Even though I think I should.
-  \item 3000 full-time researchers to construct the ATLAS experiment
-    at CERN.
-  \item Improvement in experimental methodology should be a worthy result.
-  \end{itemize}
-
-\end{frame}
-
 \begin{frame}{Thank you!}

### 2013-10-15 exp1/present2013.tex
simplify experiment description
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@683 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{frame}{Constraints on scope of experiment}
+\begin{frame}{The experiments}
 
-Above all: Easy to understand for newcomers to DoE!
+\textbf{Main goal}: Easy to understand for newcomers to DoE!
 
-\begin{itemize}
-\item Just 2 levels.
-\item Handful of illustrative factors.
-\item Analysis that illustrates key concepts.
-\item Straightforward to program and reproduce.
-\item Artificial performance differences.
-\end{itemize}
+\begin{block}{Description}
+  Experiment with 8 factors, 2 levels where we compare a single SPARQL
+  implementation before and after some change. We used \texttt{4store}
+  where we insert \textsf{sleep} statements in the join function on
+  one level and language matching function on the other, to simulate
+  an optimization that may have detrimental side-effects.
 
-\end{frame}
-
-\begin{frame}{Choices to meet these constraints}
-
-  \begin{itemize}
-  \item Response variable: Time from DNS lookup finishes to endpoint
-    has delivered a full response, measured by \texttt{curl}.
-  \item We don't compare different implementions, we compare the
-    performance before and after some change.
-  \item Query engine: \texttt{4store} where we insert \textsf{sleep}
-    statements in the join function on one level and language matching
-    function on the other. 
-  \end{itemize}
+\end{block}
 
@@
 
-\item Average over all the 32 level-combinations of the environmental
-  factors.
-
 \item May create a one-sided two-sample t-test with 4 values for each
-  of the levels of ``Implement''.
+  of the levels of ``Implement'' based on averages of other factors.
 \item Lends support to $H_1$ with a high probability, $p=1.16 \cdot

### 2013-10-16 exp1/present2013.tex
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@684 43ef1190-90fd-4dd7-a769-d0121f02b80e
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@685 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   \item Caching is a good practice, and its effect need to be a part
-    of the test, not gotten rid of.
+    of the test, not gotten rid of. If something is too complex, we
+    need to enhance our ability to deal with complexity, not oversimplify.
   \item That said benchmarking is still nice to have if you know what
@@
   \begin{itemize}
-  \item There are problems, but time doesn't allow discussion, more in paper.
+  \item The query can be thought of as a parameter, problem is that it
+    is actually an infinite amount of parameters. An important
+    challenge is to break it down to a constrained set of parameters.
   \item How many are familiar with DoE?
@@
     better on CONSTRUCT queries, on small datasets
-\item We still need to understand which effects are important and
-  significant, we do that in the discussion of our experiment
 \end{itemize}
@@
 \end{frame}
-\note{Go through very quickly}
 
@@
   \item Two factors do not contribute significantly to the variation
-    ``Range'' and ``Machine''
+    ``Range'' and ``Machine'', in the paper we describe how this
+    enables us to average over certain factors and look upon the rest
+    as a replicated experiment.
   \end{itemize}
@@
 
-\begin{frame}{Choice of factors}
+\begin{frame}{Future work: Choice of factors}
 
@@
 
-\begin{frame}{Other important issues}
+\begin{frame}{Future work: Other important issues}
 
@@
     details to assess the soundness of the experiment.
+  \item Multiresponse experiments (e.g. both response time and throughput)
   \item This paper scratches the surface of DoE, Orthogonal Arrays is
@@
 \end{frame}
+
+\section{Conclusion}
 

### 2013-10-16 exp1/present2013.tex
objkective statement
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@686 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   \begin{block}{Message for the takeaway}
-    If you develop techniques for performance enhancements of complex
+    If you make assertions about performance enhancements of complex
     systems, you need to think carefully about how you prove that 

### 2013-10-16 exp1/present2013.tex
reintroduce questions, just in case
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@687 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\begin{frame}{Questions to ask}
+
+\begin{enumerate}
+\item Are there factors that cover all realistic features?
+\item If not, are they adequately covered by randomization?
+\item If so, would the variance resulting from randomization obscure
+  factors that could provide clues that the levels are wrongly set?
+\item By carefully examining interactions with ``Implement'', are
+  there any that are unaccounted for, and that could point out wrongly
+  set levels?
+\end{enumerate}
+
 \begin{frame}{Thank you!}

### 2013-10-16 exp1/present2013.tex
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@688 43ef1190-90fd-4dd7-a769-d0121f02b80e
final tweaking
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@689 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\documentclass[english,handout,aspectratio=169]{ifislide}
-\documentclass[english,usenames,dvipsnames,aspectratio=169]{beamer}
+\documentclass[english,usenames,dvipsnames,aspectratio=169,handout]{beamer}
 \usepackage{verbatim}
 \usepackage{config}
+\usepackage{pgf}
+\usepackage{pgfpages}
+
+%
+%  Load other packages you may need here
+% 
+% \pgfpagesuselayout{4 on 1}[a4paper,landscape,border shrink=5mm
+\pgfpagesuselayout{2 on 1}[a4paper,border shrink=5mm]
 
@@
 
-%\setbeameroption{show only notes}
+\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
 \end{frame}
-\note{There are 3 factors, 8 experiments, which is $2^3$}
 
@@
   \begin{itemize}
+  \item The main effect of a 2-level factor is the average of the
+    observed values on the second level minus the corresponding
+    average of the first.
   \item Interaction effects may be complex, like one implementation is
@@
 \end{frame}
-\note{We first study a full factorial, in the paper we have three
-  fractional, we shall see the final fractional}
+\note{
+  \begin{itemize}
+  \item These are called fractional experiments, and come in two
+    shapes, full, as we have seen a design matrix of, and fractional. 
+  \item With fractional, we trade explanatory power for economy.
+  \item We first study a full factorial, in the paper we have three
+    fractional, we shall see the final fractional
+  \end{itemize}
+}
 
@@
   \item This invalidates the preceeding experiments. We got
-    strong indications early on.
+    strong indications early on. This is one of the strategies for
+    falsification we outline.
+  \item This experiment is also invalidated by the absence of certain
+    features, such as language modifiers.
 \end{itemize}
@@
   \item Different strategies must be tried, open field.
-  \item Formal methods from AI?
   \item Parameterization is very important.
@@
     \item Parameterizing SPARQL queries using the grammar pragmatically?
+    \item Formal methods from AI?
     \end{itemize}
@@
 \end{frame}
-\note{Optimization parameters, hardware characteristics}
 
@@
 
-\section{Conclusion}
+\section{Conclusions}
 
 \begin{frame}{Conclusion}
-We saw
+We saw (and the paper details):
   \begin{itemize}
@@
 
+\end{frame}
+
 \begin{frame}{Thank you!}

### 2013-10-23 exp1/present2013.tex
Comment out things to shorten
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@690 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\documentclass[english,handout,aspectratio=169]{ifislide}
-\documentclass[english,usenames,dvipsnames,aspectratio=169,handout]{beamer}
+%\documentclass[english,usenames,dvipsnames,aspectratio=169,handout]{beamer}
+\documentclass[english,usenames,dvipsnames,aspectratio=169]{beamer}
 \usepackage{verbatim}
@@
 % \pgfpagesuselayout{4 on 1}[a4paper,landscape,border shrink=5mm
-\pgfpagesuselayout{2 on 1}[a4paper,border shrink=5mm]
+%\pgfpagesuselayout{2 on 1}[a4paper,border shrink=5mm]
 
@@
 
-\setbeameroption{show notes}
+%\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
   \item ``Union'', ``Optional'' and ``Lang'' demanding things
-  \item ``BGPComp:Lang'' and ``Lang:Union'' interactions might be due
-    to that many more triples need to be searched for a language tag
-    in one of the levels of the interacting factors.
+%  \item ``BGPComp:Lang'' and ``Lang:Union'' interactions might be due
+%    to that many more triples need to be searched for a language tag
+%    in one of the levels of the interacting factors.
   \item But what about ``Implement::TripleC''? Leave that hanging for
@@
     complexity, can be broken down later
-  \item ``TripleC:Machine'' and ``TripleC'' highly significant,
+  \item ``Implement:TripleC'' and ``TripleC'' highly significant,
     implies the factors were wrongly set.

### 2014-02-01 background/70percent2.tex
Extend 70% position
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@691 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2013-08-13 background/70percent.tex
@@
-\documentclass[norsk]{uioletter}
+\documentclass{letter}
 \usepackage[utf8]{inputenc}
 
-\signature{Kjetil Kjernsmo}
+\name{Kjetil Kjernsmo}
 
-\ouremail{kjekje@ifi.uio.no}
+\address{Kjetil Kjernsmo\\Olaf Bulls vei 72\\0765 Oslo\\kjekje@ifi.uio.no}
+%\countersign{Martin Giese}
+%\countersigntitle{Veileder}
+
+\telephone{98648234}
 
@@
 Grunnet et ønske om å tilbringe mer tid med egne barn søkes det herved
-om å deltidsstilling fra dagens dato og ut høstsemesteret 2013. Det
-ønskes å sette ned stillingsprosenten til 70\%, og at stipendet
+om å forlenge omsorgspermisjonen jeg nå har ut vårsemesteret 2014. Det
+ønskes å fortsatt ha stillingsprosent på 70\%, og at stipendet
 forlenges tilsvarende.
+
+Det bør også nevnes at min kone nå er gravid med vårt tredje barn, og
+at vi i løpet av vinteren flytter til nytt hus, noe som gjør det meget
+vanskelig å jobbe 100\%.
 

### 2014-02-03 dagstuhl-2013/position.tex
Fix refs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@693 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\bibliographystyle{plain}
-%\bibliographystyle{jbact}
-%\bibliographystyle{splncs03}
-\bibliography{federation,webarch,vocabs,specs,hypermedia,egne,selectivity}
+\begin{thebibliography}{1}
+
+\bibitem{hypermediatypes}
+Mike Amundsen.
+\newblock {Hypermedia Types}.
+\newblock \url{http://amundsen.com/hypermedia/}, 2010.
+
+\bibitem{Fielding_2000_Architectural-Styles}
+Roy~Thomas Fielding.
+\newblock {\em Architectural Styles and the Design of Network-based Software
+  Architectures}.
+\newblock PhD thesis, {University of California, Irvine}, 2000.
+
+\bibitem{splendid}
+Olaf G\"{o}rlitz and Steffen Staab.
+\newblock {SPLENDID: SPARQL Endpoint Federation Exploiting VOID Descriptions}.
+\newblock In {\em Proceedings of the 2nd International Workshop on Consuming
+  Linked Data}, Bonn, Germany, 2011.
+
+\bibitem{kjernsmo_lapis_2012}
+Kjetil Kjernsmo.
+\newblock The necessity of hypermedia rdf and an approach to achieve it.
+\newblock In {\em Proceedings of the First Linked APIs workshop at the Ninth
+  Extended Semantic Web Conference}, May 2012.
+
+\end{thebibliography}
 

### 2014-07-01 cold2014/magnus.tex
Adding stub paper
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@698 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2013-08-05 exp1/iswc2013.tex
@@
 
-\title{Introducing Statistical Design of Experiments to SPARQL
-  Endpoint Evaluation}
-\author{Kjetil Kjernsmo\inst{1} \and John S. Tyssedal\inst{2}}
-\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
+\title{}
+\author{Magn\'{u}s D\ae hlen\inst{1} \and Kjetil Kjernsmo\inst{1}
 \institute{Department of Informatics,
 Postboks 1080 Blindern,
-N-0316 Oslo, Norway \email{kjekje@ifi.uio.no} \and Department of Mathematical Sciences,
-Norwegian University of Science and Technology,
-N-7491 Trondheim,
-Norway
-\email{john.tyssedal@math.ntnu.no}}
+N-0316 Oslo, Norway \email{\magnudae,kjekje\}@ifi.uio.no}
 
-%\subtitle{---Draft submitted to ISWC 2013---}
+\subtitle{---Draft submitted to COLD 2014---}
 
@@
 \begin{abstract}
-This paper argues that the common practice of benchmarking is
-inadequate as a scientific evaluation methodology. It further attempts
-to introduce the empirical tradition of the physical sciences by using
-techniques from Statistical Design of Experiments applied to the
-example of SPARQL endpoint performance evaluation. It does so by
-studying full as well as fractional factorial experiments designed to
-evaluate an assertion that some change introduced in a system has
-improved performance. This paper does not present a finished
-experimental design, rather its main focus is didactical, to shift the
-focus of the community away from benchmarking towards higher
-scientific rigor.
+
 \end{abstract}
 
-\section{Introduction}
-
-The practice of benchmarking is widespread, both in industry to select
-the most appropriate implementation among several choices, and in
-academia as means to verify or refute an assertion on the performance
-of a certain practice or system. While we could contribute to the
-industrial cases, our focus is on the latter. A benchmark typically
-consists of several microbenchmarks, which are procedures for placing
-loads on the system to measure its performance as a function of the
-load. For SPARQL benchmarks, usually the performance is measured as
-the number of queries the SPARQL implementation is able to answer in a
-certain time (the throughput) or the time required for it to respond
-to a query.
-
-However, there are many problems with this approach when used to
-evaluate assertions about the overall performance of a system or
-several systems:
-\begin{itemize}
-\item As the complexity increases, the number of parameters that one
-  may wish to test increases dramatically.
-\item Nothing compels anyone to find microbenchmarks that can refute
-  an assertion, and there is no structured approach to investigate
-  whether a benchmark is flawed.
-\item There is no meaningful summary of the performance as a whole, so
-  that an assertion about performance cannot be refuted as an
-  optimization may have detrimental side effects that cannot be
-  identified, or cannot be compared to possibly positive effects in a
-  different microbenchmark.
-\end{itemize}
-
-To address some of these problems, one may try to standardize
-benchmarks, to eliminate some parameters and identify others as
-believed to be important. Also, social pressure should then prompt
-developers to use a standard benchmark so that assertions may be
-refuted.
-
-While this is an improvement which is possibly sufficient for
-industrial uses, it is still flawed as empirical research, as it
-cannot capture the full diversity of the problem: For example, if a
-vendor claims that their implementation is better if used on a highly
-capable hardware platform, then the standardization of a hardware
-platform cannot test this assertion. Moreover, to test this assertion,
-one would need a summary technique that can identify the interaction
-between implementation and hardware platform and possibly other
-parameters.
-
-One important aspect of science is the structured approach to falsifying
-hypotheses, and therefore is important that assertions about performance
-are formulated as hypotheses that can be falsified.
-
-Therefore, attacking the complexity problem by eliminating parameters we
-think influence the outcome (e.g. caching) for simplicity is untenable
-as scientific practice. We must instead tackle this problem by
-enhancing our capability of testing complex systems and to identify
-situations where our assumptions are flawed. Such constraints can only
-be made when we are highly confident that a certain parameter does not
-have any influence.
-
-In this paper, we employ techniques from the well-established field in
-statistics known as Design of Experiments (DoE) to point out a
-direction that promises to address these problems. First, we design a
-simple 8-parameter experiment, and then we detail an experiment that
-can meaningfully provide a summary of the statistics.
-
-Then, we demonstrate how the number of runs required in the experiment
-can be reduced while maintaining a structured approach to the problem,
-and we discuss the trade-offs involved. We then continue to discuss
-techniques that show our simplistic experiment to be flawed, yet
-scientifically better than the practice of benchmarking. Finally, we
-outline a road-map to establish the use of these techniques in SPARQL
-endpoint evaluations both as a practical tool and as a scientifically
-more rigorous methodology.
-
-\subsection{Key concepts of Design of Experiments}
-
-We expect DoE to be new to most readers, but some familiarity with
-hypothesis testing is assumed.
-
-In the experiments DoE is concerned with, a \emph{response variable}
-is measured under a various combinations of parameters.  These
-parameters are known as \emph{factors}.  For each factor, a range of
-possible values is fixed.  These values are known as \emph{levels} for
-this factor.  Levels are not constrained to be continuous variables, they
-can be discrete, or even two different instances of a class.
-Experiments are run by choosing a combination of levels
-for the factors.
-
-For example, one may measure the execution time of a SPARQL endpoint
-by measuring it when it contains 1 or 2 million triples.  Then, the
-execution time is the response variable, the factor is the number of
-triples and it has two levels, 1 or 2 million triples. We have chosen
-in this paper to only deal with the relatively straightforward aspects
-of the DoE formalism, and therefore constrained the experiment strictly
-to two levels.
-
-In an experiment, there are several factors, and one may run the
-experiment by measuring the response variable for every combination of
-levels. In a two-level experiment, this will result in $2^n$ runs,
-where $n$ is the number of factors. This is called a \emph{full
-  factorial experiment}. This is done in Section~\ref{sec:full}.
-
-In many fields of science, experimental economy is extremely
-important, and so, DoE offers extensive methodology to run a certain
-fraction of the runs at the price of explanatory power. Such
-experiments are called \emph{fractional factorial experiments} and are
-covered in Section~\ref{sec:frac}.
-
-We describe the influence of the choice of levels on the response in
-terms of \emph{effects}.  For a factor $A$ with two levels, we let
-$a_1$ and $a_2$ be the average response of all $2^{n-1}$ measurements
-with $A$ at level 1, and 2, respectively.  The main effect of $A$ is
-then defined as $a_2 - a_1$.
-
-Similarly, \emph{interaction effects} for two factors $AB$ are defined
-by comparing averages for equal versus non-equal levels of $A$ and
-$B$.  Details of this theory, and how to compute the effects in
-practice, using linear regression, are found in
-\cite{wu2009experiments}.
-
-The next step is to understand which factors are important or
-significant.  One approach is to plot the sorted effects against the
-normal distribution. To understand why, note that if there is nothing
-of interest, the measurements have a certain normal distribution
-purely due to noise. When plotted against a normal distribution, the
-plot would be a straight line. If there are any deviations from the
-straight line, it implies that they may be significant. Such plots can
-be found in
-Figs.~\ref{fig:fullnormal} to~\ref{fig:frac64hugenormal}. The
-range in the $y$-axis is dictated by the number of runs.
-
-\section{Related work}
-
-A literature survey has not revealed any direct prior art, neither in
-the Semantic Web field nor more generally in database research. However, the
-general approach has been well established, not only in statistics. A
-relatively well cited textbook is \cite{citeulike:5190414} but we have
-not found the parts discussed in the present paper to be widely
-adopted. Recently, a comprehensive text on experimental methods has
-been published in
-\cite{Springer-2010-Experimental-Methods-for-the-Analysis-of-Optimization-Algorithms}
-demonstrating the broad applicability of the methodology we employ.
-We have chosen SPARQL Endpoint evaluation as an example in this study
-since SPARQL Federation is our main interest, and to turn to
-statistical standard texts \cite{wu2009experiments} for this study.
-
-Some problematic sides of benchmarking have been noted in several
-papers, notably \cite{Duan:2011:AOC:1989323.1989340} and
-\cite{MontoyaVCRA12}. We also acknowledge that great progress has been
-made to improve benchmarks, in particular, we are indebted to
-\cite{mxro:Morsey2011DBpedia}.
-
-We believe that automatic creation of benchmark queries, as pioneered
-by \cite{goerlitz2012splodge} is a critical ingredient for the
-application of DoE to be successful as a methodology.
-
-Finally, we acknowledge the efforts of the Linked Data Benchmark
-Council and the SEALS project. However, they appear to focus on cases
-where the workload is assumed to be well characterized, rather than
-scientific evaluations.
-
-\section{Experiments}
-
-Common experiments include the comparison of several different
-implementations. However, we focus on a slightly different problem: We
-would like to compare the same implementation before and after some
-change has been made, typically with the intention to enhance the
-performance of the system. There are two reasons for this: One is that
-there are currently many SPARQL implementations available, also Free
-Software ones that can be modified by anyone, and the other is to
-constrain the scope of the study to comparing just two different
-things.
-
-\subsection{Setup}
-
-As the focus of this paper is didactical, the actual measurements are
-much simpler than those that have been used in benchmarking. We have used the
-data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but
-only consider the smallest of their data-sets (that we found to be
-more than 15 MTriples). Moreover, we have taken subsets of that
-data-set by using the last 1 or 2 MTriples from their file.
-
-We have chosen 8 factors each having 2 levels, ``TripleC'' the number
-of triples in the data-set, 1 or 2 MTriples, ``Machine'', which is the
-software and hardware platform, one larger with slower
-disks\footnote{Running GNU/Linux Debian Squeeze, has 16 GB RAM, an
-  Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
-  configuration} and one smaller with a faster disk\footnote{Running
-  Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
-  SSD on SATA-III}.
-
-Then, we test some language features: ``BGPComp'', which is a Basic
-Graph Pattern of varying complexity.
-
-  \begin{center}\small
-    \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
-      \hline
-      \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
-      \begin{verbatim}
-?s  rdfs:label ?l1 ;
-    ?p1 ?o1 .
-?o1 dbo:populationTotal ?l2 .
-      \end{verbatim}
-      & 
-      \begin{verbatim}
-?s  rdfs:label ?l1 ;
-    ?p1 ?o1 .
-?o1 dbo:populationTotal ?l2 .
-?s foaf:page ?o2 ;
-   dbpprop:subdivisionName ?o3 .
-?o3 skos:subject ?o4 ;
-   dbpprop:seat ?o5 ;
-   a ?c1 .\end{verbatim} 
-\\      \hline
-    \end{tabular}
-  \end{center}
-
-The following factors test the absence (at level 1) or presence (at
-level 2) of the following clauses:
-
-\begin{center}\small
-  \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
-    \hline
-    \textbf{``Union''} & \textbf{``Lang''} \\[-2ex]
-    \begin{verbatim}
-{
-  ?o1 dbpprop:longd ?long ;
-      dbpprop:latd  ?lat .
-} UNION {
-  ?o1 geo:long ?long ;
-      geo:lat ?lat .
-}\end{verbatim} 
-&
-\begin{verbatim}
-FILTER langMatches( lang(?l1), "fr" ) 
-\end{verbatim} 
-\\    \hline
-    \textbf{``Range''} & \textbf{``Optional''} \\[-2ex]
-    \begin{verbatim}
-FILTER (?l2 > 800000)\end{verbatim} 
- & 
-    \begin{verbatim}
-OPTIONAL { ?o1 foaf:homepage ?o6 . }\end{verbatim} 
-\\    \hline
-  \end{tabular}
-\end{center}
-These fragments have been carefully designed for illustrative utility,
-as well as suitable selectivity. They are not themselves important,
-they only serve to illustrate certain examples of possible
-factors. When the experiment is run, they are combined to yield a
-query, which is then sent to a SPARQL endpoint. The SPARQL endpoint is
-itself set up using 4store (see \cite{harris20094store}) version
-1.1.5.
-
-Finally, ``Implement'' is the implementation undergoing
-evaluation. Level 1 is running the new implementation, whereas Level 2
-is the old implementation. The null hypothesis $H_0$ is that the old
-implementation is as good as the new, and the alternative hypothesis
-$H_1$ is that the new implementation has overall improved the
-performance of the system.
-
-A real optimization is beyond the scope of this paper. Instead we have
-performed a simulation that enables us to understand the effect of the
-changes.  In a real-world case, it is often the case that an
-optimization has negative side-effects, and we would need to simulate
-both the optimization and the side-effects. To do that, we degraded
-the performance of the 4store SPARQL implementation by inserting
-\texttt{sleep} statements. Specifically we inserted the C statement
-\texttt{sleep(2)} on line~920 in \texttt{src/frontend/query.c} on
-level~1, to simulate the optimization. This has the effect of delaying
-execution for 2~seconds for every block for all kinds of joins. On
-level 2, we inserted \texttt{usleep(2000)} on line 987 in
-\texttt{src/frontend/filter.c} to simulate the negative
-side-effect. This delays execution for 2 milliseconds every time the
-\texttt{langMatches} SPARQL function is called.
-
-The experimentation is implemented in R\cite{rmanual}, which is a free
-software environment for statistical computing and graphics. Necessary
-tools for DoE has been implemented by the R community in packages
-called DoE.base\cite{doebase} and FrF2\cite{frf2}. The experiments are
-run on a third computer in a 1 gigabit Ethernet LAN with the two
-experiment hosts. On the experiment hosts, four 4store instances runs
-sequentially and independently on different ports, where processes
-that are not actively answering a query are idle. Practically, the
-``Machine'' factor specifies a hostname, whereas ``TripleC'' and
-``Implement'' specifies a port number.
-
-As response variable, we have chosen to use the time from the
-namespace server lookup finishes to the data has been
-transferred. This choice is somewhat arbitrary, many other response
-variables could be chosen, and indeed, future work should explore
-multi-variable responses, but for simplicity, we think this reflects
-the total amount of work done by the SPARQL endpoint well. For the
-measurements, we have chosen to use the RCurl package\cite{lang2007r},
-which is a thin wrapper around the curl library. Curl has well
-developed facilities for timing requests and responses, and so we rely
-on its measurements.
-
-Finally, note that there are two common issues in benchmarking we do
-not consider: we do not allow a warm-up run, nor do we take into
-account that the server may cache all or parts of the result set. The
-reasons for this choice will be discussed later.
-
-All the code to reproduce this study as well as detailed instructions
-have been published on Github: \url{https://github.com/kjetilk/doe-sparql}.
-
-\subsection{Full factorial experiment}\label{sec:full}
-
-In the full factorial experiment, all combinations of the 8 factors
-are executed, in total 256 runs. This is called a $2^8$ factorial
-experiment, and while it quickly becomes infeasible for many factors,
-we shall nevertheless find the following small example
-instructive. Note that each combination is executed only once (i.e. it
-is \emph{unreplicated}), however, so even a full factorial experiment
-compares well to a typical benchmark in which each microbenchmark must
-be executed several times to obtain enough data to compute a mean with
-reasonable certainty.
-
-The above experimental setup is executed across a design matrix
-generated with the DoE.base package, which returns a data frame
-consisting of the factors with a column containing the corresponding
-curl measurements. With that, we do as described in the introduction
-and fit a linear model with all factors. Then, we generate a normal
-plot, see Figure~\ref{fig:fullnormal}.  Any significant departure from
-a straight line can be interpreted as a significant effect (see
-Section~4.8 of \cite{wu2009experiments} for a detailed
-explanation). In our case, it is most important to note that any
-effect that is negative means the effect enhances the performance of
-the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
-interactions are negative conditional on that the factors involved
-with its interactions are on their high level, then it supports the
-alternative hypothesis, i.e. we have successfully improved the
-performance.
-
-\begin{figure}[t]
-  \centerline{%
-  \includegraphics[width=.9\textwidth]{fullnormal.pdf}}
-  \caption{A normal plot of the Full Factorial Experiment. The
-    labelled points are considered significant by using the Lenth
-    criterion at a level $\alpha=0.05$.}\label{fig:fullnormal}
-\end{figure}
-
-To proceed beyond a visual judgment of significance, we may use
-Lenth's method\cite{lenthmethod}, which is a simple method for
-estimating significance in an experiment with no replicates. In
-Table~\ref{tab:effects}, the most important effects are listed.  In
-Figure~\ref{fig:fullnormal}, Lenth's method is used to label the
-significant effects. We see that we have many very significant effects
-(unfortunately so many that some labels are illegible). We also note that
-``Implement'' itself is significantly negative and that the
-``Implement:Lang'' interaction is significantly positive, as expected
-from our setup, where ``Lang'' was the detrimental side-effect we
-simulated.
-
-To investigate whether the detrimental side-effects cancel the
-positive effect of our simulated optimization, we have to formulate a
-hypothesis test. By inspecting the normal plot in
-Figure~\ref{fig:fullnormal} and Table~\ref{tab:effects}, we see that
-six factors are highly significant by either being a significant main
-effect or participate in a highly significant interaction. We say that
-the two factors ``Range'' and ``Machine'' are \emph{inactive} since
-they do not contribute significantly to the observed variation in
-performance. ``Machine'' is marginally significant, but the effect is
-so little, it will be of greater use for us in the following procedure. In
-practical situations, the five factors ``BGPComp'', ``Lang'',
-``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
-i.e. we cannot change the size of the data-set without loss of
-information, or change the query, since it gives a different
-answer. In practice, we can only control ``Implement'', i.e. we can
-only change the implementation. In this context, we call ``Implement''
-a \emph{control} factor and the other five active factors
-\emph{environmental}.
-
-We would like an overall test to see whether the new implementation is
-better than the old, and the presence of inactive factors makes it
-possible to average the performance of the new and the old
-implementations into two distinct vectors. We do this by creating an
-average over all the 32 level-combinations of the five active
-environmental factors. This leaves us with 4 values for each of the
-two levels of ``Implement'' and a two-sample t-test can be
-performed. Effectively, we treat our experiment as a $2^6$ experiment
-replicated 4 times.
-% latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Fri May 10 16:31:49 2013
-% Modified by hand to remove 1's
-\begin{wraptable}{l}{0.48\textwidth}
-\caption{The magnitude of effects for some important main effects and interactions.}\label{tab:effects}
-\bigskip\bigskip
-\begin{tabular}{rr}
-  \hline
-Factors & Effect  \\ 
-  \hline
-Implement & $-$21.27 \\ 
-  Implement:Optional & $-$11.49 \\ 
-  Implement:Union & $-$6.21 \\ 
-  Implement:TripleC:Lang1 & 4.12 \\ 
-  TripleC:Lang & 4.20 \\ 
-  Implement:TripleC & 4.32 \\ 
-  Implement:BGPComp & 4.89 \\ 
-  Lang:Union & 5.16 \\ 
-  Implement:BGPComp:Lang & 5.16 \\ 
-  BGPComp:Lang & 5.25 \\ 
-  Implement:Lang:Union & 5.75 \\ 
-  TripleC & 7.06 \\ 
-  Implement:Lang & 7.73 \\ 
-  Lang & 8.62 \\ 
-  Optional & 11.30 \\ 
-  Union & 17.18 \\ 
-   \hline
-\end{tabular}
-\end{wraptable}
-
-We perform the test as a one-sided hypothesis test where $H_0$ is that
-the mean of the vectors are equal and $H_1$ is that mean sof the
-vectors representing the new implementation is lower. In this example,
-we find that the available data supports the assertion that the new
-implementation is better with a high probability, $p=1.16 \cdot
-10^{-07}$.
-
-While it is interesting to test the overall hypothesis that the new
-implementation is an improvement, it is also interesting to know if
-there are cases where it fails. Again, we treat the experiment as a
-$2^6$ experiment replicated 4 times. Now, for each of the 32 resulting
-combinations of the environmental factors for each level of
-``Implement'', we extract 4 values. We may now run a t-test for each
-of the replications, as above. The results are tabulated in
-Table~\ref{tab:pvaluesfull}. We see that there are significant
-improvements in most cases; one may want to further investigate the
-cases where they do not.
-
-\begin{table}[ht!]
-\begin{center}
-\caption{$p$-values for different parts of the experiment}\label{tab:pvaluesfull}
-\begin{tabular}{cccccl}
-  \hline
-``TripleC'' & ``BGPComp'' & ``Lang'' & ``Union'' & ``Optional'' & $p$ \\ 
-  \hline
-  1 & 1 & 1 & 1 & 1 & 0.012 \\ 
-  1 & 1 & 1 & 1 & 2 & $1.4 \cdot 10^{-09}$ \\ 
-  1 & 1 & 1 & 2 & 1 & $3.1 \cdot 10^{-09}$ \\ 
-  1 & 1 & 1 & 2 & 2 & $6.1 \cdot 10^{-11}$ \\ 
-  1 & 1 & 2 & 1 & 1 & $3.3 \cdot 10^{-06}$ \\ 
-  1 & 1 & 2 & 1 & 2 & $2.7 \cdot 10^{-09}$ \\ 
-  1 & 1 & 2 & 2 & 1 & $2 \cdot 10^{-06}$ \\ 
-  1 & 1 & 2 & 2 & 2 & $3.6 \cdot 10^{-10}$ \\ 
-  1 & 2 & 1 & 1 & 1 & 0.014 \\ 
-  1 & 2 & 1 & 1 & 2 & $1.2 \cdot 10^{-10}$ \\ 
-  1 & 2 & 1 & 2 & 1 & $2.8 \cdot 10^{-14}$ \\ 
-  1 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
-  1 & 2 & 2 & 1 & 1 & $2.1 \cdot 10^{-05}$ \\ 
-  1 & 2 & 2 & 1 & 2 & $2.7 \cdot 10^{-07}$ \\ 
-  1 & 2 & 2 & 2 & 1 & 0.0072 \\ 
-  1 & 2 & 2 & 2 & 2 & $1.6 \cdot 10^{-05}$ \\ 
-  2 & 1 & 1 & 1 & 1 & 0.28 \\ 
-  2 & 1 & 1 & 1 & 2 & $3 \cdot 10^{-07}$ \\ 
-  2 & 1 & 1 & 2 & 1 & $3.3 \cdot 10^{-07}$ \\ 
-  2 & 1 & 1 & 2 & 2 & $1.7 \cdot 10^{-08}$ \\ 
-  2 & 1 & 2 & 1 & 1 & 0.0023 \\ 
-  2 & 1 & 2 & 1 & 2 & $6.5 \cdot 10^{-07}$ \\ 
-  2 & 1 & 2 & 2 & 1 & 0.00032 \\ 
-  2 & 1 & 2 & 2 & 2 & $1.3 \cdot 10^{-06}$ \\ 
-  2 & 2 & 1 & 1 & 1 & 0.013 \\ 
-  2 & 2 & 1 & 1 & 2 & $1 \cdot 10^{-11}$ \\ 
-  2 & 2 & 1 & 2 & 1 & $3.8 \cdot 10^{-11}$ \\ 
-  2 & 2 & 1 & 2 & 2 & $4.1 \cdot 10^{-15}$ \\ 
-  2 & 2 & 2 & 1 & 1 &   1 \\ 
-  2 & 2 & 2 & 1 & 2 & $2.3 \cdot 10^{-05}$ \\ 
-  2 & 2 & 2 & 2 & 1 &   1 \\ 
-  2 & 2 & 2 & 2 & 2 & 0.99 \\ 
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
-
-For future larger experiments, one must investigate whether it is
-necessary to adjust the $p$-value due to the larger numbers of
-hypotheses tested.
-
-The main objective of this paper is not to establish that the
-optimization is significant, it is merely a simulation. It is to
-establish a critical practice of evaluations. We must therefore
-understand why some effects are significant. That
-``Implement'' is significant is hardly a surprise, that is what we
-worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
-are strong is also due to that these are fairly demanding things to
-evaluate. Since the delay we inserted affects all kinds of joins, it
-is also intuitive that interactions between ``Implement'' and those
-that require joins are strong. The strong ``BGPComp:Lang'' and
-``Lang:Union'' interactions might be due to that many more triples
-need to be searched for a language tag in one of the levels of the
-interacting factors. However, the
-positive ``Implement:TripleC'' interaction evades such explanations;
-in fact, it hints that our optimization may not work as well for
-larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
-we see that whenever ``TripleC'' is at the low level, the
-new implementation is always best. Only at the higher level, the old
-implementation may be better. This is a clear indication that the
-separation between levels for the sizes of the data-set is too small
-and should be investigated further. Such reasoning should be applied
-to all significant effects.
-
-\subsection{Fractional factorial experiment}\label{sec:frac}
-
-As mentioned, the full factorial experiment goes as $2^n$ and becomes
-prohibitively expensive when $n$ is a large number of factors. As the
-evaluation of SPARQL endpoints is an inherently complex problem, a
-large number of factors are needed, and so full experiments cannot
-scale. However, we may reduce the size of the experiment significantly
-by sacrificing some explanatory power, in the form of fractional
-factorial experiments.
-
-We lose explanatory power due to \emph{aliasing}; for example, the
-interaction labelled ``TripleC:BGPComp'' may be aliased with the
-``Machine:Range'' interaction. That is to say, a detected increase in
-run time from larger Basic Graph Patterns for large databases, can
-also be explained by a less powerful machine that is worse at
-evaluating FILTER clauses with ranges. They are
-indistinguishable. This may or may not cause a problem. In many cases,
-we may not be interested in these effects, we may only be interested
-in ``Implement'' and its interactions. Moreover, it is possible (even
-easy using the FrF2 package in R) to declare which effects must not be
-aliased, and determine the size of the experiment based on that. Such
-a main effect or two-factor interaction is called \emph{clear} if none
-of its aliases are other main effects or two-factor
-interactions. Another possibility to shrink the size of the experiment
-is if we \textit{a priori} can say that some effects are
-negligible. We have not found such assumptions to be tenable in our
-case.
-
-We have made two fractional factorial designs: One with 32 runs and
-one with 64 runs. Again, the number of runs is in powers of 2, but
-much smaller. In both cases, we have specified that all the factors
-must be clear and also all two-factor interactions where ``Implement''
-is involved must be clear. The resulting experiments is returned to us
-as design matrices, but they can also be described sufficiently for
-reproducibility in terms of \emph{design generators}. These also
-declare aliasing relations for some of the main effects. The design
-generators are\\
-\begin{tabular}{ccl}
-``Range'' &=& ``TripleC'' ``Machine'' ``BGPComp'' \\  
-``Union'' &=& ``TripleC'' ``Machine'' ``Lang'' \\  
-``Optional'' &=& ``TripleC'' ``BGPComp'' ``Lang'' ``Implement'' 
-\end{tabular}
-\\and\\
-\begin{tabular}{ccl}
-``Union'' &=& ``Implement''  ``TripleC''  ``Machine'' ``BGPComp'' \\
-``Optional'' &=& ``Implement''  ``TripleC'' ``Lang''  ``Range'' 
-\end{tabular}
-\\for the 32 and 64 run experiments respectively.
-
-\begin{figure}[ht!]
-  \centerline{%
-    \includegraphics[width=.9\textwidth]{frac32normal.pdf}}
-  \caption{A normal plot of the Fractional Factorial Experiment with
-    32 runs. The labelled points are considered significant by using
-    the Lenth criterion at a level
-    $\alpha=0.05$.}
-  \label{fig:frac32normal}
-\end{figure}
-
-The resulting normal plot is in Figure~\ref{fig:frac32normal}. We see
-that only two main effects are significant on a $\alpha = 0.05$ level
-according to the Lenth criterion, ``Implement'' is again deemed a
-significant improvement, and ``Union'' is deemed significantly the
-hardest operation for the query engine. We see that some other points
-also deviate from the straight line, but with as few runs as this, the
-total variance is great, so no other conclusions can be drawn. We may
-use the t-test we used in the previous section, as we have only two
-active factors and so we may treat it as a $2^2$ experiment with 8
-replications. The resulting $p = 0.00098$ is still very low, albeit
-larger than in the previous experiment. This also explains why there
-are so few significant effects, and had to be expected with such a
-small number of runs and thus higher variance.
-
-For any other purpose than a rough idea of some key influences on the
-overall performance of the endpoint for the given levels, this little
-experiment is insufficient.
-
-We turn to the 64-run experiment and its normal plot in
-Figure~\ref{fig:frac64normal}. We see that then we get more significant
-effects, and that they correspond well to those we saw in the full
-factorial experiment. They also all have an intuitive explanation as
-above. However, we do not see the effects that we found worrisome in
-the full factorial design. The ``Implement:TripleC'' interaction
-emerges only for a level $\alpha=0.15$. To ensure that we discover
-such cases will be a key challenge in further studies.
-
-\begin{figure}[ht!]
-  \centerline{%
-    \includegraphics[width=0.96\textwidth]{frac64normal.pdf}}
-  \caption{A normal plot of the Fractional Factorial Experiment with
-    64 runs. The labelled points are considered significant by using
-    the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
-    and crosses at $\alpha=0.05$. The use of different symbols has
-    been added to FrF2 by the authors.}\label{fig:frac64normal}
-\end{figure}
-
-Since the 64-run experiment has 3 inactive factors, ``Machine'',
-``TripleC'' and ``Range'', we can treat it as a $2^5$ experiment
-replicated twice. Using the same procedure as above, we find that
-again, the hypothesis test indicates that the new implementation is an
-improvement at $p = 5.9 \cdot 10^{-6}$. Again, the comparatively
-higher $p$-value is due to the lower number of runs. It is also possible
-to perform the per-environmental factors test.
-
-\subsection{Fractional factorial experiment with more data}\label{sec:hugefrac}
-
-The importance of finding flaws in the levels is highlighted by the
-normal plot in Figure~\ref{fig:frac64hugenormal}. The smaller
-experiment made it feasible to run with a larger data-set, this
-experiment uses the original 15~MTriples data-set of
-\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
-previous discussion, for larger ``TripleC'', our simulated performance
-is actually worse, as evident from the positive ``Implement''. We also
-see that the interactions ``Implement:Lang'' and ``Implement:TripleC''
-are highly significant. It is clear what happened: For larger
-``TripleC'', the delay we introduced in the langMatches function has a
-devastating effect, since it delays the execution for every triple
-that the query engine applies the filter to, and thereby completely
-dominates the execution time.
-
-\begin{figure}[ht!]
-  \centerline{%
-  \includegraphics[width=.9\textwidth]{frac64hugenormal.pdf}}
-  \caption{A normal plot of the Fractional Factorial Experiment with
-    64 runs where the level 2 of ``TripleC'' is 15 MTriples. The
-    labelled points are considered significant by using the Lenth
-    criterion at a level $\alpha=0.05$.}\label{fig:frac64hugenormal}
-\end{figure}
-
-Moreover, we note that the ``Machine'' factor has become highly
-significant in the other end along with its ``Machine:TripleC''
-interaction. One should take careful notice that the ``Machine''
-factor encompasses many different things, that should be different
-factors; CPU, disks, available RAM, installed software, tuning
-parameters and so on. It is advantageous to have such broad factors in
-the design in early experiments, because they may point out unexpected
-effects that should be further scrutinized. The ``Machine'' factor is
-a good example of usage that also helps us manage complexity: If
-possible, one may group many factors into one if in doubt of
-their relevance and break it up in subsequent experiments if deeper
-understanding is needed. To some extent, ``BGPComp'' is also a broad
-factor, since it encompasses many structures that are normally seen in
-a Basic Graph Pattern. In this case, the cause of the performance
-problem was highly audible: It was the old P-ATA disks of the larger
-machine, but in general, such clues are not available to the
-experimenter, therefore more factors are needed.
-
-We are forced to conclude that the three preceding experiments are
-severely flawed, as they failed to take into account the effects from
-larger data sizes on slow hardware.
-
-Note that we do not take advantage of the usual performance of 4store
-in this simulation, to the contrary, we chose 4store because of its
-relative simplicity to modify it to suit our needs for illustrative
-purposes, which consists of degrading it in several ways. The small
-data-sets used here do not imply that the approach will not work for
-large data-sets; properly configured, actual optimization will give
-very different runtimes, and the small number of runs will make it
-possible to evaluate a very large number of factors.
-
-\section{Discussion}
-
-It is a fundamental property of two-level experiments that they can
-only model linear effects, however, they are usually quite sufficient
-for identifying significant effects even though the underlying
-response may be non-linear.  As we saw in Section~\ref{sec:full}, the
-choice of levels may be critical, and if non-linearity is expected,
-more than two levels must be considered, or at the very least,
-attention must be paid to the choice of levels.
-
-We have argued that fixing certain parameters is untenable, but we
-have allowed ourselves to disregard caching and warm-up runs. We are
-aware that caching may be important in 4store, as a preliminary
-experiment gave a response time of 53.9~s of a query with ``BGPComp''
-at level~2 but other factors at level~1. When an OPTIONAL clause was
-added, the response time dropped to 1.25~s. As this is not a more
-restrictive query, the most likely explanation is that the result of
-the evaluation of the Basic Graph Pattern was cached, so only the
-OPTIONAL left join was necessary to add in the second query.
-
-We did this to illustrate \emph{randomization} as this is what
-makes it permissible, i.e. the order of execution is random, so the
-benefit of caching is likely to apply randomly to different runs. For
-example, had the run order been reversed in the previous example, it
-could have been the query without OPTIONAL that would have
-benefited. Unless the effect of caching is cumulative throughout the
-experiment (which is possible), the randomization will have the effect
-of reversing the runs randomly. The end result is that the effect of
-so-called ``lurking variables'' such as caching, warm-up effects, etc,
-contribute to the total unexplained variance of the experiment, but
-should not skew the results to invalidate the estimated effects.
-
-As the unexplained variance of the experiment may cloud important
-effects, notably effects that indicate flaws, it should be kept to a
-minimum. Thus, lurking variables should be turned into factors
-whenever possible, as they degrade the quality of the experiment, but
-in many cases, we can live with them if their contribution to the
-variance is small.
-
-Finally, we note that the most problematic cases are those that are
-covered neither by the broad factors, nor the lurking variables, or
-any of the specific factors. We cannot in the general case be sure
-that they are accounted for, so reviewers must remain vigilant that
-some factors are simply ignored. What finally invalidates the present
-experiment is the complete absence of certain obvious language
-features, such as solution modifiers.
-
-\section{Future Work}
-
-The present experiment is very primitive in its choice of factors, and
-we believe that the strategy employed to construct suitable
-experiments will be the main determinant for the long-term feasibility
-of this direction. One strategy will be to start with broad factors
-such as ``Machine'' and refine them with experience. Also, the data
-must be parameterized with many more factors than just the number of
-triples; data heterogeneity \cite{Duan:2011:AOC:1989323.1989340} is one
-example, but also queries over highly skewed data is important. A more
-difficult issue is how to address the problem of testing the whole
-SPARQL language\cite{sparql11query}. We believe that the
-parameterization work initiated by \cite{goerlitz2012splodge} with
-their SPLODGE system is a key ingredient. They parameterized SPARQL
-queries in such a way they could be auto-generated. One suggestion is
-to parameterize queries based on the grammar, so that one factor
-becomes the number of e.g. UNIONs, and the levels are chosen. One
-problem one will soon run into is that parts of the language is
-recursive, e.g. a GroupGraphPattern can consist of several
-GroupGraphPatterns. However, we believe that pragmatic limits can be
-set, it is for example not of practical interest to nest OPTIONALs to
-any great depth even if it is allowed by the language.
-
-We saw in Section~\ref{sec:frac} that with a smaller experiment, fewer
-interactions are significant, and so provide us with fewer clues to
-assess the soundness of the experiment. This should be a focus of
-further research.
-
-Finally, we have only employed the simplest parts of the DoE theory.
-What has been presented here is part of a much more general formalism
-known as Orthogonal Arrays. The use of orthogonal arrays allow for
-different numbers of levels, for much greater flexibility in total run
-size, and for non-regular designs that can provide a good fit for the
-complex problem of SPARQL endpoint evaluations.
-
-\section{Conclusions}
-
-As evaluating SPARQL endpoints is inherently difficult, a simplistic
-experiment such as the one we designed should not hold up to
-scrutiny. We set out to demonstrate how an experiment could be set up
-using DoE, and show how it can be analyzed. In
-sections~\ref{sec:full}~and~\ref{sec:frac}, we first saw how the
-analysis correctly pointed out the most important effects, under
-assumptions dictated by the factors and levels that were given. We saw
-how the formalism provided a comprehensive view of the
-experiment. Then, we saw that the formalism pointed out weaknesses
-that could invalidate the experiment, and in
-Section~\ref{sec:hugefrac} we saw that the experiment did indeed not
-hold up to scrutiny.
-
-We saw that while two-level experiments can perform well in determining
-significant effects, they do not necessarily work well to find a
-model, as exemplified by failure to identify the possibly non-linear
-effect of the P-ATA disks in the first experiments. This issue can be
-addressed by using an orthogonal array design.
-
-Also, experiments as small as the 32-run are not useful in estimating
-any detailed characteristics and are thus of little use for science,
-but could be useful in engineering: If the experiment has been
-validated by a larger experiment, it could serve well in a Continuous
-Integration system.
-
-We have seen that we can perform a proper hypothesis test based on
-summary statistics, and by using our expertise, we showed how to reveal
-that the experiment was flawed, and we have cautioned that failure to
-control unexplained variance may compromise our ability to do so. We
-have also pointed out how this direction can provide more rigorous
-evaluation practices.
-
-To this end, the following questions must be asked:
-\begin{enumerate}
-\item Are there factors that cover all realistic features?
-\item If not, are they adequately covered by randomization?
-\item If so, would the variance resulting from randomization obscure
-  factors that could provide clues that the levels are wrongly set?
-\item By carefully examining interactions with ``Implement'', are
-  there any that are unaccounted for, and that could point out wrongly
-  set levels?
-\end{enumerate}
-
-Even if the complexity is great, with properly tuned endpoints, it is
-feasible to do millions of runs, and with orthogonal arrays, this
-formalism can be extended to many different evaluation problems.
-
-\paragraph*{Acknowledgments}
-
-Kjetil Kjernsmo would like to thank his main supervisor Martin Giese
-for kind assistance. He would also like to thank Steve Harris for his
-generous support in degrading his excellent work on 4store.
-
-%\bibliographystyle{plain}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{jbact}
-%\bibliographystyle{splncs03}
-%\bibliography{practical,selectivity,federation,benchmarks,optimization,experimentsperformance,specs}
-\input{iswc2013.bbl}
 \end{document}

### 2014-07-01 cold2014/magnus.tex
first stab at an abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@699 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\title{}
+\title{Performance of two different ontological and data access methods}
 \author{Magn\'{u}s D\ae hlen\inst{1} \and Kjetil Kjernsmo\inst{1}
@@
 \begin{abstract}
+  We evaluate two different approaches that have been used in the
+  literature and in practice by different groups publishing metadata
+  about cars when consumed by a prototype application. These
+  approaches are characterized by a generic vs. domain specific
+  ontology; by a constrained API vs. openly queryable data % TODO:
+                                % Magnus, what else
+  We have implemented a prototype application where a potential user
+  may query selected properties of cars, and we investigate how the
+  different choices made by the original designers influence the
+  performance of the application. For the evaluation, we employ
+  the statistical disipline of Design of Experiments.
 

### 2014-07-01 cold2014/magnus.tex
Introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@700 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Introduction}
+
+The retrieval of information from different sources and then combine
+them to allow a user to find a certain combination of properties that
+suit their purpose is an archetypical Semantic Web use case. We have
+chosen to focus on a product selection use case, specifically on cars,
+since some car makers have embraced the Semantic Web vision and chosen
+to share detailed data on their products.
+
+The findings, we assert, have a broader validity, given the generic
+nature of the problem they are trying to solve, and so our
+recommendations should have broad relevance to similar use cases.
+
+We present two different approaches, one promoted by Renault, see
+\cite{SemWebAppRes} and \cite{ren1}, and another made by Martin~Hepp
+in collaboration with Volkswagen resulting in amongst other things,
+the Car Options Ontology~\cite{COO}. We shall compare them to find
+their strengths and weaknesses. Each approach contains an ontology and
+has its own way of representing data.  To make this comparison we have
+used data about the same domain, data about car models and their
+component constraints.  The first approach is a generic ontology. By
+generic it means that it can be used to represent any product model
+with component constraints. The second approach is a domain specific
+ontology which as the name implies, is only applicable with one
+particular domain. In this thesis that domain is about car models.
+
+We will also show how to create a viable web application which
+utilizes such complex data, mainly for the purpose of conducting
+performance tests to determine the weaknesses of each approach. This
+will be done with complex data found on the web today from different
+car manufacturers.  With performance we mean the response time between
+a HTTP~\cite{http}} post operation against the application and when
+the application presents the user with an answer. The application will
+contain the possibility to do HTTP posts against both approaches.
+
+There will be several options on how to query the data because of all
+the different specifications.  That is why we have chosen to use the
+testing approach \textit{Design of Experiments} (DoE). This approach
+will be further explained alongside the results in
+Section~\ref{Results}. The evaluation will be based on four
+experiments testing several aspects of the approaches. They will help
+us determine what kind of factors are significant to the performance.
+
+\section{Related work}
+Complex products and specifying configurations has been a research
+topic for over a decade.  The possibility to personalize more and more
+products is why several research articles have proposed different
+approaches on how to handle these configurations. Most of the research
+are around finding the ultimate solution with a specification
+system. In 1999, M. Aldanondo et. al proposed how to structure a
+system to handle configurations and their constraints. This included
+proposed definitions for products, configurations and
+configurators. The paper focused on making a generic solution to fit
+several manufacturers.~\cite{OldConf} In a newer article,
+H. Afsarmanesh and M. Shafahi (2013) proposed a complex product
+specification system.~\cite{NewConf} This include object modelling and
+a user interface. They have focused on supporting stakeholders in the
+specification process.
+
+Unfortunately these articles do not present any research done with
+semantic technologies. The use of semantic technologies on complex
+products is a young field of research. The only thing done here is
+what Renault and Volkswagen have presented. Both of these car
+manufacturers have presented the public with two different
+solutions. They have also shown how to present the data on the web and
+the complexity around their solutions.
+
 \end{document}

### 2014-07-01 cold2014/magnus.tex
fix title and author add something on evaluation platform Add results
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@701 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \title{Performance of two different ontological and data access methods}
-\author{Magn\'{u}s D\ae hlen\inst{1} \and Kjetil Kjernsmo\inst{1}
+\author{Magn\'{u}s D\ae hlen \and Kjetil Kjernsmo}
 \institute{Department of Informatics,
 Postboks 1080 Blindern,
-N-0316 Oslo, Norway \email{\magnudae,kjekje\}@ifi.uio.no}
+N-0316 Oslo, Norway \email{\{magnudae,kjekje\}@ifi.uio.no} }
 
@@
 
+The Volkswagen solution builds upon GoodRelations, which is a web
+vocabulary for e-commerce and was launched in 2008 and is now
+widespread. GoodRelations.~\cite{GR} GoodRelations can be used for
+detailed information about products to be sold online.
+
+% TODO: Har litt på følelsen at dette blir for tynt for et paper,
+% mulig det burde vært mer generelt om Goodrelations, etc. Vi kan jo
+% bare se det an.
+
+\section{Evaluation platform}
+
+We set out to evaluate different approaches based on a real use case
+where a user is seeking to buy a new car based on detailed data
+published by car manufacturers. Today only Renault, of all the car
+manufacturers, has opened their data using semantic
+technologies. Volkswagen had their data published until the start of
+2013, but unfortunately they canceled their Semantic Web project and
+the data were removed.  Fortunately, their ontologies are still open
+for the community to use. Eventually, we found Daimler very
+forthcoming, as they offered data about their A- and B-class cars.
+The Daimler data used a custom XML-based format, so we lifted the data
+using a Python script into the ontologies developed by Martin Hepp and
+Volkswagen.
+
+To be able to put large stress on the application, which again would
+allow us to bring out key strengths and weaknesses of the two
+different approaches, we chose to create an application where a user
+choose several or all car options specifications.
+
+The application was implemented using Apache Maven, Spring and Apache
+Tomcat as well as Jena.
+
+The application present users with a web interface where they can
+input values for any specification they might want to search for.  In
+Figure \vref{initialView} we can see how the application initially
+looks for a user. Here the user can input values for the different
+specifications, for instance fuel type or CO$_2$ emission. After a
+user is done choosing specifications he can execute the actual car
+model search. The application extracts the data from the form and
+comprises it into another format that can easily be used later on. The
+next step begins in the search module. Here it starts a thread for
+each possible car model to execute each individual search. In the end
+the user is presented with a set of valid models to choose from, which
+contain the chosen specifications. In Figure \vref{afterSearch}, we
+can see what the user is presented with after a successful query
+execution, in this case seventeen different models. In this run, the
+user have searched for a model with diesel fuel, automatic
+transmission, four or more gears and a maximum of six seats. We also
+have the values and URIs for each of these valid models, which means
+that the application could present the user with more information than
+it does today.
+
+\begin{figure}
+  \centering
+      \includegraphics[width=12cm]{initialView.png}
+  \caption{Initial view for a user}\label{initialView}
+\end{figure}
+
+\begin{figure}
+  \centering
+      \includegraphics[width=12cm]{afterSearch.png}
+  \caption{The view after a search}\label{afterSearch}
+\end{figure}
+There are done some pre-computations to make the application a little
+more effective. This includes extracting all the model names from both
+ontologies, initiate all possible partially defined products as a
+PartialCar object and creating some internal structure. These
+pre-computations were done so that it would be possible to present the
+user with the name of each car model, not just the URI. We will also
+describe more in depth how the car model search is done later on.
+
+% TODO differences between Daimler and Renault
+
+\section{Experiments}\label{Results}
+
+To perform a statistically sound experiment, we turn to techniques
+from statistics known as Design of Experiments (DoE). For an
+introduction to the use of DoE in Semantic Web research, see~\cite{Kjern}.
+
+\subsection{Planning the experiment}
+As other approaches, there are several steps in doing experiments with DoE. 
+First one has to state an objective which is important to give the experiment a purpose. 
+The next step is to choose a response. The response is the experimental outcome or 
+observation. There may be multiple responses in an experiment. In this experiment 
+there will be one response and that is the response time on a HTTP post against the application. 
+These first two steps are the same for all the experiments in this thesis.
+
+The next two steps of planning an experiment is specific to each experiment itself. This means 
+that these steps will be described more in depth in each experiment because it may vary depending on the actual 
+experiment. 
+The third step in planning an experiment is choosing factors and levels. This is a key part of 
+DoE. A factor is a variable that is studied in the experiment, and to be able to study this factor 
+it is needed to use two or more values of this factor. These values are referred to as levels. 
+This varies from benchmarking because often these levels are not required to do proper benchmarking. 
+The levels will allow us the check each factor for its significance. It is important to 
+identify the key factors in the planning stage. This is to get the maximum effect out of the experiment. 
+Factors may be \textit{quantitative} or \textit{qualitative}. Quantitative factors are often 
+numerical values that represent an interval. For instance the weight of a car is considered a 
+Quantitative value. Qualitative factors are predefined values within a known set of values. 
+In our case a Qualitative factor can be the type of fuel, for instance Diesel.
+
+The fourth step in planning is to choose the experimental plan. Here we will use one specific approach. The approach 
+is called \textit{full factorial experiment} which we will go into more detail about in the next sub-section. There are other 
+approaches like fractional factorial experiment, but they were not applicable or relevant to this thesis. 
+The last three steps of planning the experiment are performing the experiment, analyzing the output 
+and in the end drawing conclusions. More about these steps will be taken individually at each experiment.~\cite{PlanExp}
+
+\subsection{Full factorial experiment}
+A full factorial experiment is when one got $k$ factors and $n$ amount of levels. This means 
+that there is $n^k$ \textit{factorial designs} which result in $n^k$ designs to execute. 
+In the thesis every experiment will only have two levels, but the factors may vary. This means 
+that we will use a $2^k$ full factorial design for every experiment. The experiments consist of $2^k$ combinations 
+with $k$ factors. For instance if we have three factors where every factor got two levels 
+we get $2^3$ designs, also called \textit{runs}. In Table \ref{designs} we can see all 
+the possible runs we get from 3 factors. The + and - represent the level for each factor. 
+When running the actual experiment one should use a \textit{planning matrix} to display the actual runs in the 
+experiment. In Table \ref{designsspec} we can see the same table as before, just that now it is filled out with 
+actual factors and their different levels.
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{ | l  l  l |}
+    \hline
+    {\bf Factor 1} & {\bf Factor 2} & {\bf Factor 3} \\ \hline
+	 + & + & + \\ \hline
+	  - & + & + \\ \hline
+	 + & - & + \\ \hline
+	  - & - & + \\ \hline
+	 + & + & - \\ \hline
+	  - & + & - \\ \hline
+	 + & - & - \\ \hline
+	  - & - &  - \\ \hline
+    \end{tabular}
+\end{center}
+\caption{$2^3$ possible runs}\label{designs}
+\end{table}
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{ | l | l  l  l |}
+    \hline
+    {\bf Run} &  {\bf Fuel type} & {\bf Transmission} & {\bf Speed} \\ \hline
+	 1 & Diesel & Automatic & 100 \\ \hline
+	 2 & Petrol & Automatic & 100 \\ \hline
+	 3 & Diesel & Manual & 100 \\ \hline
+	 4 & Petrol & Manual & 100 \\ \hline
+	 5 & Diesel & Automatic & 200 \\ \hline
+	 6 & Petrol & Automatic & 200 \\ \hline
+	 7 & Diesel & Manual & 200 \\ \hline
+	 8 & Petrol & Manual &  200 \\ \hline
+    \end{tabular}
+\end{center}
+\caption{A filled planning matrix with 3 factors}\label{designsspec}
+\end{table}
+
+Two key properties of full factorial designs are \textit{balance} and \textit{orthogonality}. 
+Balance means that each factor level appears the same amount of times in all the runs. For instance that 
+the factor Diesel appears the same amount of runs as Petrol. Orthogonal means that two factors level combinations 
+will appear in the same number of runs. As an example will Diesel fuel combined with Automatic transmission 
+appear in the same number of runs as Petrol fuel with manual transmission. We can see in Table \ref{designsspec} 
+that both these level combinations appear two times. Diesel and automatic appears in run 1 and 5, while Petrol and 
+manual appears in run 4 and 8.
+
+Experiments like this can also be replicated, which just means that the experimental plan is run more than one time. 
+It is not always efficient to replicate an experiment. Sometimes it can be too expensive to run it more than one time. 
+According to Wu and Hamada(2009) replication is unnecessary for computer experiments because repeated computer runs 
+with the same input give the same output.
+If it is replicated the plan should be randomized to get the best result. One reason for randomizing is to 
+identify and get rid of \textit{lurking variables}. A lurking variable is a factor that is not taken into 
+account when doing an experiment and that might affect the outcome of the experiment. For instance 
+if one is doing an experiment outside. The outdoor temperature might be a affecting the experiment if not taken 
+into account.~\cite{FullFac}
+
+\section{Experiments}
+In this section we will present several experiments with different amount of factors.
+The objective is to find out how the ontologies will perform in comparison and identifying 
+which factors affect the outcome of each experiment. We did the ontology alignment and 
+development of the application to prevent the loss of precision with either ontology.
+
+In the experiments the definition for VSO/COO is \textit{-vso} and for CO is \textit{-co}.
+These two definitions stand for which ontology are tested in a particular run. In practice this means that when 
+we send in the keyword \textit{-vso} we only run the specification search against the car models represented by VSO/COO, 
+in our case the data from Daimler. This is referred to 
+as the ontology factor in each experiment and is a factor in all the experiments which are presented in 
+this thesis. The ontology factor represents the difference in how the data are represented and how the ontologies are 
+structures, and with this factor we want to determine how this affects the overall performance. 
+We have programmed the application so that the querying of the ontologies are done as similar as possible on both 
+levels. This was done to eliminate any interference from the application. There are some 
+differences in the programming due that the ontologies were different and they will be discussed later on. 
+In Chapter \ref{prototype}, the application and its differences are explained in depth.
+
+Before starting on the experiment we had to find a way to do the actual experiment. We wanted to 
+first create the planning matrix and then execute each run as a HTTP post. To do this we needed to 
+make a small script to calculate the $2^k$ runs and perform the HTTP posts. We chose to use 
+the programming language \textit{python} to do this. That was because python was a familiar language  
+and it also had a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the DoE package we just declared 
+the amount of factors and the matrix was created. At that point the matrix looked more like Table 
+\ref{designs}, which meant that we had to substitute the + and - with the levels of each factor. 
+Further in the script we used the package \textit{request}\footnote{http://requests.readthedocs.org/en/latest/} 
+to do the HTTP posts in python. Below one can see the line of code executing the request.
+\scriptsize\begin{lstlisting}
+ r = requests.post('http://localhost:8080/linkedopendata-magnudae/index', 
+		  data=payload)
+\end{lstlisting}
+\normalsize
+The \textit{payload} is a dictionary of all the values to post to the application. 
+After the request we saved the response which contained the response 
+time of a run. After each request we saved the run plus the response time and wrote 
+it to file on CSV format. To evaluate the result we used the programming language 
+R with the packages \textit{DoE.base}~\cite{DoEBase}, \textit{FrF2}~\cite{FrF2} 
+and \textit{BsMD}~\cite{BsMD}. R allowed us to parse 
+the CSV files created from the experiments and output them in two types of graphs. 
+A similarity with all the experiments were that they all contained at least the factors 
+ontology, transmission and fuel type.
+
+To evaluate the graphs we have used a method in the \textit{FrF2} R package called \textit{DanielPlot}. This method assumes that the estimated effects 
+are normally distributed with means equal to the effects. The means of all estimated effects are zero. Resulting 
+in a plot where the estimated effect would end up on a straight line. This plot is then testing whether all the estimated 
+effects have the same distribution. All deviations of the straight line indicates a significant factor. In our results we have 
+used the absolute value of the effects, also called a half-normal plot. The advantages of using the half-normal plot is that 
+all large estimated effects will end up in the top right corner. This means that the highest performance significance will end up in 
+the top right corner.~\cite{Plotting} \\
+Here is a quick explanation of the axes shown in each graph. 
+The X axis will show the absolute effect of each factor, shown in time. The values on the X axis 
+is represented in seconds which means that the a plot will have a significance measured in seconds. 
+The Y axis shows the half-normal scores. There are two way of representing the graph and 
+that is with either half-normal plots or normal plots. With half-normal the effects on the X 
+axis are shown with the absolute effect. This means that all factors are placed with a positive 
+half-normal score. This is to avoid visual misleading scores because with the normal scores, deviations from 
+the line on negative scores might confuse the reader who are evaluating the charts. 
+The scores them selves are a measure for which factor that is the most significant. 
+We have also set the \textit{alpha} to $0.05$ which means that we will tolerate a false percentage of $5\%$.
+
+All of the planning matrices, graphs and results will be presented with each experiment later on in this section.
+The discussion and interpretation of the results will be presented in Chapter \ref{Discussion}. 
+Before starting the experiments we will present abbreviations for each factor because their full name 
+made the graphs totally illegible. These abbreviations will be used in the graphs and tables further on in this section. 
+The results from all the experiments can be found in appendix \ref{appD}.
+\begin{itemize}
+  \item Ontology = O
+  \item Fuel Type = FT
+  \item Transmission = T
+  \item Weight = W
+  \item Total Weight = WT
+  \item Emission = E
+  \item Nr. of Gears = G
+  \item Seating Capacity = SC
+  \item Fuel Consumption = FC
+  \item Doors = D
+\end{itemize}
+
+\subsection{Experiment with three factors}
+The first experiment that was done, was the experiment with only three factors. The goal here 
+was to see how the ontologies behave with a small amount of querying and factors. 
+Table \ref{facandlevExp1} show the different factors and their levels. We chose to start with a 
+small experiment using only simple factors which had few possible values. That is why we chose transmission 
+and fuel type as factors as well as the ontology factor. With these three factors we got $2^3$ number of runs. 
+This meant that we had 8 runs to execute against our application. This experiment was also replicated 
+four times to get the most accurate results. The experiment was small so the replication did not cause any issues. 
+The results were calculated by finding the average response time for each run between all four replications.
+
+\begin{table}[H]
+\begin{center}
+    \begin{tabular}{ | l | l l |}
+    \hline
+    {\bf Factor} & {\bf Level 1} & {\bf Level 2} \\ \hline
+	Ontology & -vso & -co \\ \hline 
+	Fuel type & Diesel & Unleaded Petrol \\ \hline 
+	Transmission & Manual Gearbox & Automatic Gearbox \\ \hline 
+    \end{tabular}
+\end{center}
+\caption{Factors and levels for experiment 1}\label{facandlevExp1}
+\end{table}
+
+The next step was to create the planning matrix for this experiment. As mention in the beginning of 
+this section we used a package in python to create and fill out the planning matrix. Table \ref{3factor} 
+shows all the 8 runs for the experiment with three factors. We also ran a randomized version the experiment to check 
+whether the randomization would influence the experiment. The results were almost equal which eliminated randomization as a factor. 
+Both results can be found in appendix \ref{appD}. After 
+each run the response time were logged to that particular run. The response time were then used to 
+check for significant effect when the experiment were evaluated
+\begin{table}
+\begin{center}
+    \begin{tabular}{ | l | l l l |}
+    \hline
+    {\bf Run} & {\bf Ontology} & {\bf Fuel type} & {\bf Transmission} \\ \hline
+	1 & -vso & Diesel & Automatic Gearbox \\ \hline 
+	2 & -co & Diesel & Automatic Gearbox \\ \hline 
+	3 & -vso & Unleaded Petrol & Automatic Gearbox \\ \hline 
+	4 & -co & Unleaded Petrol & Automatic Gearbox \\ \hline 
+	5 & -vso & Diesel & Manual Gearbox \\ \hline 
+	6 & -co & Diesel & Manual Gearbox \\ \hline 
+	7 & -vso & Unleaded Petrol & Manual Gearbox \\ \hline 
+	8 & -co & Unleaded Petrol & Manual Gearbox \\ \hline 
+    \end{tabular}
+\end{center}
+\caption{Planning matrix for experiment with three factors}\label{3factor}
+\end{table}
+
+Like the rest of the experiments we used DanielPlot to evaluate and 
+a package in R to visualize the result. The Figure \vref{3factorGraph} shows the results. 
+Here we see no significant factors. Still there are small effects 
+from several factors. In Table \vref{3factorEffect} we can see the time effects from largest to smallest. 
+We see that the ontology factor has the largest effect on approximately 2.2 seconds which means that one of the 
+ontologies overall is 2.2 seconds faster. By reading the results from the linear model it is the factor level 
+\textit{-vso} which has an overall faster response time.
+
+\begin{figure}
+ \includegraphics[width=12cm]{2factorfinal.pdf}
+  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
+\end{figure}
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{ | l l |}
+    \hline
+    {\bf Factors} & {\bf Absolute Effect}  \\ \hline
+	  O     & 2.2463476\\ \hline
+	  O:T    &0.4536135 \\ \hline
+	  O:FT:T &0.4534723 \\ \hline
+	  T      &0.4513712 \\ \hline
+	  FT:T   &0.4499820\\ \hline
+	  O:FT   &0.1240831 \\ \hline
+	  FT     &0.1189629 \\ \hline
+    \end{tabular}
+\end{center}
+\caption{Table of effects for the three factor experiment}\label{3factorEffect}
+\end{table}
+
+\subsection{Experiment with ten factors}
+The last experiment we did was a ten factor experiment. The goal 
+here was to see which factors would be deemed significant in a complete 
+car model search. With the results from this experiment we could discuss 
+several aspects of using the different ontologies. 
+Here we added 4 new factors in comparison to the six factor experiment. The newly 
+added factors were \textit{weight}, \textit{nr. of doors}, \textit{nr. of seats} and 
+\textit{fuel consumption}. We used the same approach here as in the previous experiments 
+to calculate the average value for the new factors. The average values can be seen in the 
+planning matrix in appendix \ref{appB}. With 10 factors the amount of runs were 
+$2^10$. This results in 1024 number of runs which made the planning matrix too big to show here.
+
+In the graph \vref{10factorGraph} we can see the results from running all the 1024 runs.
+We can see that there are a lot of significant effects present here. There are more than 
+any of the previous experiments, but it is also more factors and factor interactions.
+Unfortunately there are so many significant factors that there are some of them that are 
+illegible, but most of them have a low impact on the significance graph. Like the six factor experiment we also see 
+here that there are three groups of significant factors. The first group where most of them 
+are placed range from around 2 seconds effect to 11. Here there are a lot of factor interactions 
+and why they are there will be explained in the discussion. 
+The second group is very like the second group we saw in the six factor experiment, the only difference is 
+that the total weight (WT) factor is substituted with fuel consumption (FC). The last group is 
+the same for all the experiments, except the two factor experiment. Here we find the ontology factor. 
+In Table \vref{10factorEffect} we can see the top twenty absolute effects. This includes all the 
+significant factors from the last group and the second one, but also 5 from the first group. 
+We can see that emission also here are present in a lot of the entries, but so is ontology 
+as well. The new thing here is the fuel consumption has entered the most significant factors, and 
+is by itself higher than emission. The linear model again tells us that the option \textit{-vso} has a significantly lower 
+response time than \textit{-co}
+
+When we accumulated the results from all the runs we ended up with the total 
+amount of response time to be approximately 30095 seconds, which is 8 hours, 21 minutes and 35 seconds. 
+This were the total response time after running all the runs. This did not take into account all the other 
+steps during the experiments which took time, like writing results to file and initiating the application. 
+This discouraged the replication of the experiment due to the size and was also a reason why we did not 
+do experiments with more factors.
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{ | l l |}
+    \hline
+    {\bf Factors} & {\bf Absolute Effect}  \\ \hline
+      O & 58.1785542324229 \\ \hline
+      E:FC & 45.6971032050781 \\ \hline
+      O:E:FC & 45.5502538925781\\ \hline
+      FC & 38.6140886777344\\ \hline
+      O:FC & 38.5571969433593\\ \hline
+      O:E & 35.3967871113281\\ \hline
+      E & 35.3827983613282\\ \hline
+      O:T & 33.1249425527344\\ \hline
+      T & 33.1154579746093\\ \hline
+      O:T:E:FC & 29.3077075019531\\ \hline
+      T:E:FC & 29.2986405957031\\ \hline
+      O:T:FC & 28.2420862089844\\ \hline
+      T:FC & 28.2325333808593\\ \hline
+      O:T:E & 25.3121093378906\\ \hline
+      T:E & 25.3039236347656\\ \hline
+      WT:E & 11.9191989902344\\ \hline
+      O:WT:E & 11.8908780371094\\ \hline
+      O:WT:FC & 11.7234715332032\\ \hline
+      WT:FC & 11.7140515488281\\ \hline
+      O:D:E & 10.1241599902344\\ \hline
+    \end{tabular}
+\end{center}
+\caption{The top twenty significant effects of the ten factor graph}\label{10factorEffect}
+\end{table}
+
+\begin{figure}
+ \includegraphics[width=14cm]{10factorAvarageDanielPlot.pdf}
+  \caption{Resulting graph for the ten factor experiment}\label{10factorGraph}
+\end{figure}
+
+% TODO: In conclusion, discuss external validity.
+
 \end{document}

### 2014-07-01 cold2014/magnus.tex
just reformat
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@702 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \subsection{Planning the experiment}
-As other approaches, there are several steps in doing experiments with DoE. 
-First one has to state an objective which is important to give the experiment a purpose. 
-The next step is to choose a response. The response is the experimental outcome or 
-observation. There may be multiple responses in an experiment. In this experiment 
-there will be one response and that is the response time on a HTTP post against the application. 
-These first two steps are the same for all the experiments in this thesis.
+As other approaches, there are several steps in doing experiments with
+DoE.  First one has to state an objective which is important to give
+the experiment a purpose.  The next step is to choose a response. The
+response is the experimental outcome or observation. There may be
+multiple responses in an experiment. In this experiment there will be
+one response and that is the response time on a HTTP post against the
+application.  These first two steps are the same for all the
+experiments in this thesis.
 
-The next two steps of planning an experiment is specific to each experiment itself. This means 
-that these steps will be described more in depth in each experiment because it may vary depending on the actual 
-experiment. 
-The third step in planning an experiment is choosing factors and levels. This is a key part of 
-DoE. A factor is a variable that is studied in the experiment, and to be able to study this factor 
-it is needed to use two or more values of this factor. These values are referred to as levels. 
-This varies from benchmarking because often these levels are not required to do proper benchmarking. 
-The levels will allow us the check each factor for its significance. It is important to 
-identify the key factors in the planning stage. This is to get the maximum effect out of the experiment. 
-Factors may be \textit{quantitative} or \textit{qualitative}. Quantitative factors are often 
-numerical values that represent an interval. For instance the weight of a car is considered a 
-Quantitative value. Qualitative factors are predefined values within a known set of values. 
-In our case a Qualitative factor can be the type of fuel, for instance Diesel.
+The next two steps of planning an experiment is specific to each
+experiment itself. This means that these steps will be described more
+in depth in each experiment because it may vary depending on the
+actual experiment.  The third step in planning an experiment is
+choosing factors and levels. This is a key part of DoE. A factor is a
+variable that is studied in the experiment, and to be able to study
+this factor it is needed to use two or more values of this
+factor. These values are referred to as levels.  This varies from
+benchmarking because often these levels are not required to do proper
+benchmarking.  The levels will allow us the check each factor for its
+significance. It is important to identify the key factors in the
+planning stage. This is to get the maximum effect out of the
+experiment.  Factors may be \textit{quantitative} or
+\textit{qualitative}. Quantitative factors are often numerical values
+that represent an interval. For instance the weight of a car is
+considered a Quantitative value. Qualitative factors are predefined
+values within a known set of values.  In our case a Qualitative factor
+can be the type of fuel, for instance Diesel.
 
-The fourth step in planning is to choose the experimental plan. Here we will use one specific approach. The approach 
-is called \textit{full factorial experiment} which we will go into more detail about in the next sub-section. There are other 
-approaches like fractional factorial experiment, but they were not applicable or relevant to this thesis. 
-The last three steps of planning the experiment are performing the experiment, analyzing the output 
-and in the end drawing conclusions. More about these steps will be taken individually at each experiment.~\cite{PlanExp}
+The fourth step in planning is to choose the experimental plan. Here
+we will use one specific approach. The approach is called \textit{full
+  factorial experiment} which we will go into more detail about in the
+next sub-section. There are other approaches like fractional factorial
+experiment, but they were not applicable or relevant to this thesis.
+The last three steps of planning the experiment are performing the
+experiment, analyzing the output and in the end drawing
+conclusions. More about these steps will be taken individually at each
+experiment.~\cite{PlanExp}
 
 \subsection{Full factorial experiment}
-A full factorial experiment is when one got $k$ factors and $n$ amount of levels. This means 
-that there is $n^k$ \textit{factorial designs} which result in $n^k$ designs to execute. 
-In the thesis every experiment will only have two levels, but the factors may vary. This means 
-that we will use a $2^k$ full factorial design for every experiment. The experiments consist of $2^k$ combinations 
-with $k$ factors. For instance if we have three factors where every factor got two levels 
-we get $2^3$ designs, also called \textit{runs}. In Table \ref{designs} we can see all 
-the possible runs we get from 3 factors. The + and - represent the level for each factor. 
-When running the actual experiment one should use a \textit{planning matrix} to display the actual runs in the 
-experiment. In Table \ref{designsspec} we can see the same table as before, just that now it is filled out with 
-actual factors and their different levels.
+A full factorial experiment is when one got $k$ factors and $n$ amount
+of levels. This means that there is $n^k$ \textit{factorial designs}
+which result in $n^k$ designs to execute.  In the thesis every
+experiment will only have two levels, but the factors may vary. This
+means that we will use a $2^k$ full factorial design for every
+experiment. The experiments consist of $2^k$ combinations with $k$
+factors. For instance if we have three factors where every factor got
+two levels we get $2^3$ designs, also called \textit{runs}. In Table
+\ref{designs} we can see all the possible runs we get from 3
+factors. The + and - represent the level for each factor.  When
+running the actual experiment one should use a \textit{planning
+  matrix} to display the actual runs in the experiment. In Table
+\ref{designsspec} we can see the same table as before, just that now
+it is filled out with actual factors and their different levels.
 
@@
 
-Two key properties of full factorial designs are \textit{balance} and \textit{orthogonality}. 
-Balance means that each factor level appears the same amount of times in all the runs. For instance that 
-the factor Diesel appears the same amount of runs as Petrol. Orthogonal means that two factors level combinations 
-will appear in the same number of runs. As an example will Diesel fuel combined with Automatic transmission 
-appear in the same number of runs as Petrol fuel with manual transmission. We can see in Table \ref{designsspec} 
-that both these level combinations appear two times. Diesel and automatic appears in run 1 and 5, while Petrol and 
-manual appears in run 4 and 8.
+Two key properties of full factorial designs are \textit{balance} and
+\textit{orthogonality}.  Balance means that each factor level appears
+the same amount of times in all the runs. For instance that the factor
+Diesel appears the same amount of runs as Petrol. Orthogonal means
+that two factors level combinations will appear in the same number of
+runs. As an example will Diesel fuel combined with Automatic
+transmission appear in the same number of runs as Petrol fuel with
+manual transmission. We can see in Table \ref{designsspec} that both
+these level combinations appear two times. Diesel and automatic
+appears in run 1 and 5, while Petrol and manual appears in run 4 and
+8.
 
-Experiments like this can also be replicated, which just means that the experimental plan is run more than one time. 
-It is not always efficient to replicate an experiment. Sometimes it can be too expensive to run it more than one time. 
-According to Wu and Hamada(2009) replication is unnecessary for computer experiments because repeated computer runs 
-with the same input give the same output.
-If it is replicated the plan should be randomized to get the best result. One reason for randomizing is to 
-identify and get rid of \textit{lurking variables}. A lurking variable is a factor that is not taken into 
-account when doing an experiment and that might affect the outcome of the experiment. For instance 
-if one is doing an experiment outside. The outdoor temperature might be a affecting the experiment if not taken 
-into account.~\cite{FullFac}
+Experiments like this can also be replicated, which just means that
+the experimental plan is run more than one time.  It is not always
+efficient to replicate an experiment. Sometimes it can be too
+expensive to run it more than one time.  According to Wu and
+Hamada(2009) replication is unnecessary for computer experiments
+because repeated computer runs with the same input give the same
+output.  If it is replicated the plan should be randomized to get the
+best result. One reason for randomizing is to identify and get rid of
+\textit{lurking variables}. A lurking variable is a factor that is not
+taken into account when doing an experiment and that might affect the
+outcome of the experiment. For instance if one is doing an experiment
+outside. The outdoor temperature might be a affecting the experiment
+if not taken into account.~\cite{FullFac}
 
 \section{Experiments}
-In this section we will present several experiments with different amount of factors.
-The objective is to find out how the ontologies will perform in comparison and identifying 
-which factors affect the outcome of each experiment. We did the ontology alignment and 
-development of the application to prevent the loss of precision with either ontology.
+In this section we will present several experiments with different
+amount of factors.  The objective is to find out how the ontologies
+will perform in comparison and identifying which factors affect the
+outcome of each experiment. We did the ontology alignment and
+development of the application to prevent the loss of precision with
+either ontology.
 
-In the experiments the definition for VSO/COO is \textit{-vso} and for CO is \textit{-co}.
-These two definitions stand for which ontology are tested in a particular run. In practice this means that when 
-we send in the keyword \textit{-vso} we only run the specification search against the car models represented by VSO/COO, 
-in our case the data from Daimler. This is referred to 
-as the ontology factor in each experiment and is a factor in all the experiments which are presented in 
-this thesis. The ontology factor represents the difference in how the data are represented and how the ontologies are 
-structures, and with this factor we want to determine how this affects the overall performance. 
-We have programmed the application so that the querying of the ontologies are done as similar as possible on both 
-levels. This was done to eliminate any interference from the application. There are some 
-differences in the programming due that the ontologies were different and they will be discussed later on. 
-In Chapter \ref{prototype}, the application and its differences are explained in depth.
+In the experiments the definition for VSO/COO is \textit{-vso} and for
+CO is \textit{-co}.  These two definitions stand for which ontology
+are tested in a particular run. In practice this means that when we
+send in the keyword \textit{-vso} we only run the specification search
+against the car models represented by VSO/COO, in our case the data
+from Daimler. This is referred to as the ontology factor in each
+experiment and is a factor in all the experiments which are presented
+in this thesis. The ontology factor represents the difference in how
+the data are represented and how the ontologies are structures, and
+with this factor we want to determine how this affects the overall
+performance.  We have programmed the application so that the querying
+of the ontologies are done as similar as possible on both levels. This
+was done to eliminate any interference from the application. There are
+some differences in the programming due that the ontologies were
+different and they will be discussed later on.  In Chapter
+\ref{prototype}, the application and its differences are explained in
+depth.
 
-Before starting on the experiment we had to find a way to do the actual experiment. We wanted to 
-first create the planning matrix and then execute each run as a HTTP post. To do this we needed to 
-make a small script to calculate the $2^k$ runs and perform the HTTP posts. We chose to use 
-the programming language \textit{python} to do this. That was because python was a familiar language  
-and it also had a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the DoE package we just declared 
-the amount of factors and the matrix was created. At that point the matrix looked more like Table 
-\ref{designs}, which meant that we had to substitute the + and - with the levels of each factor. 
-Further in the script we used the package \textit{request}\footnote{http://requests.readthedocs.org/en/latest/} 
-to do the HTTP posts in python. Below one can see the line of code executing the request.
+Before starting on the experiment we had to find a way to do the
+actual experiment. We wanted to first create the planning matrix and
+then execute each run as a HTTP post. To do this we needed to make a
+small script to calculate the $2^k$ runs and perform the HTTP
+posts. We chose to use the programming language \textit{python} to do
+this. That was because python was a familiar language and it also had
+a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the
+DoE package we just declared the amount of factors and the matrix was
+created. At that point the matrix looked more like Table
+\ref{designs}, which meant that we had to substitute the + and - with
+the levels of each factor.  Further in the script we used the package
+\textit{request}\footnote{http://requests.readthedocs.org/en/latest/}
+to do the HTTP posts in python. Below one can see the line of code
+executing the request.
+
 \scriptsize\begin{lstlisting}
@@
 \normalsize
-The \textit{payload} is a dictionary of all the values to post to the application. 
-After the request we saved the response which contained the response 
-time of a run. After each request we saved the run plus the response time and wrote 
-it to file on CSV format. To evaluate the result we used the programming language 
-R with the packages \textit{DoE.base}~\cite{DoEBase}, \textit{FrF2}~\cite{FrF2} 
-and \textit{BsMD}~\cite{BsMD}. R allowed us to parse 
-the CSV files created from the experiments and output them in two types of graphs. 
-A similarity with all the experiments were that they all contained at least the factors 
-ontology, transmission and fuel type.
 
-To evaluate the graphs we have used a method in the \textit{FrF2} R package called \textit{DanielPlot}. This method assumes that the estimated effects 
-are normally distributed with means equal to the effects. The means of all estimated effects are zero. Resulting 
-in a plot where the estimated effect would end up on a straight line. This plot is then testing whether all the estimated 
-effects have the same distribution. All deviations of the straight line indicates a significant factor. In our results we have 
-used the absolute value of the effects, also called a half-normal plot. The advantages of using the half-normal plot is that 
-all large estimated effects will end up in the top right corner. This means that the highest performance significance will end up in 
+The \textit{payload} is a dictionary of all the values to post to the
+application.  After the request we saved the response which contained
+the response time of a run. After each request we saved the run plus
+the response time and wrote it to file on CSV format. To evaluate the
+result we used the programming language R with the packages
+\textit{DoE.base}~\cite{DoEBase}, \textit{FrF2}~\cite{FrF2} and
+\textit{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
+from the experiments and output them in two types of graphs.  A
+similarity with all the experiments were that they all contained at
+least the factors ontology, transmission and fuel type.
+
+To evaluate the graphs we have used a method in the \textit{FrF2} R
+package called \textit{DanielPlot}. This method assumes that the
+estimated effects are normally distributed with means equal to the
+effects. The means of all estimated effects are zero. Resulting in a
+plot where the estimated effect would end up on a straight line. This
+plot is then testing whether all the estimated effects have the same
+distribution. All deviations of the straight line indicates a
+significant factor. In our results we have used the absolute value of
+the effects, also called a half-normal plot. The advantages of using
+the half-normal plot is that all large estimated effects will end up
+in the top right corner. This means that the highest performance
+significance will end up in
 the top right corner.~\cite{Plotting} \\
-Here is a quick explanation of the axes shown in each graph. 
-The X axis will show the absolute effect of each factor, shown in time. The values on the X axis 
-is represented in seconds which means that the a plot will have a significance measured in seconds. 
-The Y axis shows the half-normal scores. There are two way of representing the graph and 
-that is with either half-normal plots or normal plots. With half-normal the effects on the X 
-axis are shown with the absolute effect. This means that all factors are placed with a positive 
-half-normal score. This is to avoid visual misleading scores because with the normal scores, deviations from 
-the line on negative scores might confuse the reader who are evaluating the charts. 
-The scores them selves are a measure for which factor that is the most significant. 
-We have also set the \textit{alpha} to $0.05$ which means that we will tolerate a false percentage of $5\%$.
+Here is a quick explanation of the axes shown in each graph.  The X
+axis will show the absolute effect of each factor, shown in time. The
+values on the X axis is represented in seconds which means that the a
+plot will have a significance measured in seconds.  The Y axis shows
+the half-normal scores. There are two way of representing the graph
+and that is with either half-normal plots or normal plots. With
+half-normal the effects on the X axis are shown with the absolute
+effect. This means that all factors are placed with a positive
+half-normal score. This is to avoid visual misleading scores because
+with the normal scores, deviations from the line on negative scores
+might confuse the reader who are evaluating the charts.  The scores
+them selves are a measure for which factor that is the most
+significant.  We have also set the \textit{alpha} to $0.05$ which
+means that we will tolerate a false percentage of $5\%$.
 
-All of the planning matrices, graphs and results will be presented with each experiment later on in this section.
-The discussion and interpretation of the results will be presented in Chapter \ref{Discussion}. 
-Before starting the experiments we will present abbreviations for each factor because their full name 
-made the graphs totally illegible. These abbreviations will be used in the graphs and tables further on in this section. 
-The results from all the experiments can be found in appendix \ref{appD}.
+All of the planning matrices, graphs and results will be presented
+with each experiment later on in this section.  The discussion and
+interpretation of the results will be presented in Chapter
+\ref{Discussion}.  Before starting the experiments we will present
+abbreviations for each factor because their full name made the graphs
+totally illegible. These abbreviations will be used in the graphs and
+tables further on in this section.  The results from all the
+experiments can be found in appendix \ref{appD}.
 \begin{itemize}
@@
 \subsection{Experiment with three factors}
-The first experiment that was done, was the experiment with only three factors. The goal here 
-was to see how the ontologies behave with a small amount of querying and factors. 
-Table \ref{facandlevExp1} show the different factors and their levels. We chose to start with a 
-small experiment using only simple factors which had few possible values. That is why we chose transmission 
-and fuel type as factors as well as the ontology factor. With these three factors we got $2^3$ number of runs. 
-This meant that we had 8 runs to execute against our application. This experiment was also replicated 
-four times to get the most accurate results. The experiment was small so the replication did not cause any issues. 
-The results were calculated by finding the average response time for each run between all four replications.
+The first experiment that was done, was the experiment with only three
+factors. The goal here was to see how the ontologies behave with a
+small amount of querying and factors.  Table \ref{facandlevExp1} show
+the different factors and their levels. We chose to start with a small
+experiment using only simple factors which had few possible
+values. That is why we chose transmission and fuel type as factors as
+well as the ontology factor. With these three factors we got $2^3$
+number of runs.  This meant that we had 8 runs to execute against our
+application. This experiment was also replicated four times to get the
+most accurate results. The experiment was small so the replication did
+not cause any issues.  The results were calculated by finding the
+average response time for each run between all four replications.
 
@@
 
-The next step was to create the planning matrix for this experiment. As mention in the beginning of 
-this section we used a package in python to create and fill out the planning matrix. Table \ref{3factor} 
-shows all the 8 runs for the experiment with three factors. We also ran a randomized version the experiment to check 
-whether the randomization would influence the experiment. The results were almost equal which eliminated randomization as a factor. 
-Both results can be found in appendix \ref{appD}. After 
-each run the response time were logged to that particular run. The response time were then used to 
-check for significant effect when the experiment were evaluated
+The next step was to create the planning matrix for this
+experiment. As mention in the beginning of this section we used a
+package in python to create and fill out the planning matrix. Table
+\ref{3factor} shows all the 8 runs for the experiment with three
+factors. We also ran a randomized version the experiment to check
+whether the randomization would influence the experiment. The results
+were almost equal which eliminated randomization as a factor.  Both
+results can be found in appendix \ref{appD}. After each run the
+response time were logged to that particular run. The response time
+were then used to check for significant effect when the experiment
+were evaluated
 \begin{table}
@@
 
-Like the rest of the experiments we used DanielPlot to evaluate and 
-a package in R to visualize the result. The Figure \vref{3factorGraph} shows the results. 
-Here we see no significant factors. Still there are small effects 
-from several factors. In Table \vref{3factorEffect} we can see the time effects from largest to smallest. 
-We see that the ontology factor has the largest effect on approximately 2.2 seconds which means that one of the 
-ontologies overall is 2.2 seconds faster. By reading the results from the linear model it is the factor level 
-\textit{-vso} which has an overall faster response time.
+Like the rest of the experiments we used DanielPlot to evaluate and a
+package in R to visualize the result. The Figure \vref{3factorGraph}
+shows the results.  Here we see no significant factors. Still there
+are small effects from several factors. In Table \vref{3factorEffect}
+we can see the time effects from largest to smallest.  We see that the
+ontology factor has the largest effect on approximately 2.2 seconds
+which means that one of the ontologies overall is 2.2 seconds
+faster. By reading the results from the linear model it is the factor
+level \textit{-vso} which has an overall faster response time.
 
@@
 \subsection{Experiment with ten factors}
-The last experiment we did was a ten factor experiment. The goal 
-here was to see which factors would be deemed significant in a complete 
-car model search. With the results from this experiment we could discuss 
-several aspects of using the different ontologies. 
-Here we added 4 new factors in comparison to the six factor experiment. The newly 
-added factors were \textit{weight}, \textit{nr. of doors}, \textit{nr. of seats} and 
-\textit{fuel consumption}. We used the same approach here as in the previous experiments 
-to calculate the average value for the new factors. The average values can be seen in the 
-planning matrix in appendix \ref{appB}. With 10 factors the amount of runs were 
-$2^10$. This results in 1024 number of runs which made the planning matrix too big to show here.
+The last experiment we did was a ten factor experiment. The goal here
+was to see which factors would be deemed significant in a complete car
+model search. With the results from this experiment we could discuss
+several aspects of using the different ontologies.  Here we added 4
+new factors in comparison to the six factor experiment. The newly
+added factors were \textit{weight}, \textit{nr. of doors},
+\textit{nr. of seats} and \textit{fuel consumption}. We used the same
+approach here as in the previous experiments to calculate the average
+value for the new factors. The average values can be seen in the
+planning matrix in appendix \ref{appB}. With 10 factors the amount of
+runs were $2^10$. This results in 1024 number of runs which made the
+planning matrix too big to show here.
 
-In the graph \vref{10factorGraph} we can see the results from running all the 1024 runs.
-We can see that there are a lot of significant effects present here. There are more than 
-any of the previous experiments, but it is also more factors and factor interactions.
-Unfortunately there are so many significant factors that there are some of them that are 
-illegible, but most of them have a low impact on the significance graph. Like the six factor experiment we also see 
-here that there are three groups of significant factors. The first group where most of them 
-are placed range from around 2 seconds effect to 11. Here there are a lot of factor interactions 
-and why they are there will be explained in the discussion. 
-The second group is very like the second group we saw in the six factor experiment, the only difference is 
-that the total weight (WT) factor is substituted with fuel consumption (FC). The last group is 
-the same for all the experiments, except the two factor experiment. Here we find the ontology factor. 
-In Table \vref{10factorEffect} we can see the top twenty absolute effects. This includes all the 
-significant factors from the last group and the second one, but also 5 from the first group. 
-We can see that emission also here are present in a lot of the entries, but so is ontology 
-as well. The new thing here is the fuel consumption has entered the most significant factors, and 
-is by itself higher than emission. The linear model again tells us that the option \textit{-vso} has a significantly lower 
-response time than \textit{-co}
+In the graph \vref{10factorGraph} we can see the results from running
+all the 1024 runs.  We can see that there are a lot of significant
+effects present here. There are more than any of the previous
+experiments, but it is also more factors and factor interactions.
+Unfortunately there are so many significant factors that there are
+some of them that are illegible, but most of them have a low impact on
+the significance graph. Like the six factor experiment we also see
+here that there are three groups of significant factors. The first
+group where most of them are placed range from around 2 seconds effect
+to 11. Here there are a lot of factor interactions and why they are
+there will be explained in the discussion.  The second group is very
+like the second group we saw in the six factor experiment, the only
+difference is that the total weight (WT) factor is substituted with
+fuel consumption (FC). The last group is the same for all the
+experiments, except the two factor experiment. Here we find the
+ontology factor.  In Table \vref{10factorEffect} we can see the top
+twenty absolute effects. This includes all the significant factors
+from the last group and the second one, but also 5 from the first
+group.  We can see that emission also here are present in a lot of the
+entries, but so is ontology as well. The new thing here is the fuel
+consumption has entered the most significant factors, and is by itself
+higher than emission. The linear model again tells us that the option
+\textit{-vso} has a significantly lower response time than
+\textit{-co}
 
-When we accumulated the results from all the runs we ended up with the total 
-amount of response time to be approximately 30095 seconds, which is 8 hours, 21 minutes and 35 seconds. 
-This were the total response time after running all the runs. This did not take into account all the other 
-steps during the experiments which took time, like writing results to file and initiating the application. 
-This discouraged the replication of the experiment due to the size and was also a reason why we did not 
-do experiments with more factors.
+When we accumulated the results from all the runs we ended up with the
+total amount of response time to be approximately 30095 seconds, which
+is 8 hours, 21 minutes and 35 seconds.  This were the total response
+time after running all the runs. This did not take into account all
+the other steps during the experiments which took time, like writing
+results to file and initiating the application.  This discouraged the
+replication of the experiment due to the size and was also a reason
+why we did not do experiments with more factors.
 

### 2014-07-01 cold2014/magnus.tex
trimming update abstract textit -> emph
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@703 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   approaches are characterized by a generic vs. domain specific
-  ontology; by a constrained API vs. openly queryable data % TODO:
-                                % Magnus, what else
-  We have implemented a prototype application where a potential user
-  may query selected properties of cars, and we investigate how the
-  different choices made by the original designers influence the
-  performance of the application. For the evaluation, we employ
-  the statistical disipline of Design of Experiments.
+  ontology; by a constrained tree-oriented API vs. openly queryable
+  data; encouraging vs discouraging reasoning, etc.  We have
+  implemented a prototype application where a potential user may query
+  selected properties of cars, and we investigate how the different
+  choices made by the original designers influence the performance of
+  the application. For the evaluation, we employ the statistical
+  disipline of Design of Experiments.
 
@@
 the different specifications.  That is why we have chosen to use the
-testing approach \textit{Design of Experiments} (DoE). This approach
+testing approach \emph{Design of Experiments} (DoE). This approach
 will be further explained alongside the results in
@@
 % TODO differences between Daimler and Renault
+% TODO Alignment
 
@@
 \subsection{Planning the experiment}
-As other approaches, there are several steps in doing experiments with
+Like other approaches, there are several steps in doing experiments with
 DoE.  First one has to state an objective which is important to give
@@
 this factor it is needed to use two or more values of this
-factor. These values are referred to as levels.  This varies from
-benchmarking because often these levels are not required to do proper
-benchmarking.  The levels will allow us the check each factor for its
-significance. It is important to identify the key factors in the
-planning stage. This is to get the maximum effect out of the
-experiment.  Factors may be \textit{quantitative} or
-\textit{qualitative}. Quantitative factors are often numerical values
-that represent an interval. For instance the weight of a car is
-considered a Quantitative value. Qualitative factors are predefined
-values within a known set of values.  In our case a Qualitative factor
-can be the type of fuel, for instance Diesel.
+factor. These values are referred to as levels.  The levels will allow
+us to examine the influence of each factor. It is important to
+identify the key factors in the planning stage. This is to get the
+maximum effect out of the experiment.  Factors may be
+\emph{quantitative} or \emph{qualitative}. Quantitative factors
+are often numerical values that represent an interval. For instance
+the weight of a car is considered a Quantitative value. Qualitative
+factors are predefined values within a known set of values.  In our
+case a Qualitative factor can be the type of fuel, for instance
+Diesel.
 
 The fourth step in planning is to choose the experimental plan. Here
-we will use one specific approach. The approach is called \textit{full
+we will use one specific approach. The approach is called \emph{full
   factorial experiment} which we will go into more detail about in the
@@
 A full factorial experiment is when one got $k$ factors and $n$ amount
-of levels. This means that there is $n^k$ \textit{factorial designs}
+of levels. This means that there is $n^k$ \emph{factorial designs}
 which result in $n^k$ designs to execute.  In the thesis every
@@
 factors. For instance if we have three factors where every factor got
-two levels we get $2^3$ designs, also called \textit{runs}. In Table
-\ref{designs} we can see all the possible runs we get from 3
-factors. The + and - represent the level for each factor.  When
-running the actual experiment one should use a \textit{planning
-  matrix} to display the actual runs in the experiment. In Table
-\ref{designsspec} we can see the same table as before, just that now
-it is filled out with actual factors and their different levels.
-
-\begin{table}
-\begin{center}
-    \begin{tabular}{ | l  l  l |}
-    \hline
-    {\bf Factor 1} & {\bf Factor 2} & {\bf Factor 3} \\ \hline
-	 + & + & + \\ \hline
-	  - & + & + \\ \hline
-	 + & - & + \\ \hline
-	  - & - & + \\ \hline
-	 + & + & - \\ \hline
-	  - & + & - \\ \hline
-	 + & - & - \\ \hline
-	  - & - &  - \\ \hline
-    \end{tabular}
-\end{center}
-\caption{$2^3$ possible runs}\label{designs}
-\end{table}
+two levels we get $2^3$ designs, also called \emph{runs}. This results
+in a \emph{planning matrix} that shows the actual runs in the
+experiment as tabulated in Table~\ref{designsspec}. This contains the
+chosen factors and their levels.
 
@@
 
-Two key properties of full factorial designs are \textit{balance} and
-\textit{orthogonality}.  Balance means that each factor level appears
-the same amount of times in all the runs. For instance that the factor
-Diesel appears the same amount of runs as Petrol. Orthogonal means
-that two factors level combinations will appear in the same number of
-runs. As an example will Diesel fuel combined with Automatic
-transmission appear in the same number of runs as Petrol fuel with
-manual transmission. We can see in Table \ref{designsspec} that both
-these level combinations appear two times. Diesel and automatic
-appears in run 1 and 5, while Petrol and manual appears in run 4 and
-8.
-
-Experiments like this can also be replicated, which just means that
-the experimental plan is run more than one time.  It is not always
-efficient to replicate an experiment. Sometimes it can be too
-expensive to run it more than one time.  According to Wu and
-Hamada(2009) replication is unnecessary for computer experiments
-because repeated computer runs with the same input give the same
-output.  If it is replicated the plan should be randomized to get the
-best result. One reason for randomizing is to identify and get rid of
-\textit{lurking variables}. A lurking variable is a factor that is not
-taken into account when doing an experiment and that might affect the
-outcome of the experiment. For instance if one is doing an experiment
-outside. The outdoor temperature might be a affecting the experiment
-if not taken into account.~\cite{FullFac}
-
-\section{Experiments}
+\subsection{Running the experiments}
 In this section we will present several experiments with different
@@
 
-In the experiments the definition for VSO/COO is \textit{-vso} and for
-CO is \textit{-co}.  These two definitions stand for which ontology
+In the experiments the definition for VSO/COO is \emph{-vso} and for
+CO is \emph{-co}.  These two definitions stand for which ontology
 are tested in a particular run. In practice this means that when we
-send in the keyword \textit{-vso} we only run the specification search
+send in the keyword \emph{-vso} we only run the specification search
 against the car models represented by VSO/COO, in our case the data
@@
 small script to calculate the $2^k$ runs and perform the HTTP
-posts. We chose to use the programming language \textit{python} to do
+posts. We chose to use the programming language \emph{python} to do
 this. That was because python was a familiar language and it also had
@@
 the levels of each factor.  Further in the script we used the package
-\textit{request}\footnote{http://requests.readthedocs.org/en/latest/}
+\emph{request}\footnote{http://requests.readthedocs.org/en/latest/}
 to do the HTTP posts in python. Below one can see the line of code
@@
 
-The \textit{payload} is a dictionary of all the values to post to the
+The \emph{payload} is a dictionary of all the values to post to the
 application.  After the request we saved the response which contained
@@
 result we used the programming language R with the packages
-\textit{DoE.base}~\cite{DoEBase}, \textit{FrF2}~\cite{FrF2} and
-\textit{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
+\emph{DoE.base}~\cite{DoEBase}, \emph{FrF2}~\cite{FrF2} and
+\emph{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
 from the experiments and output them in two types of graphs.  A
@@
 
-To evaluate the graphs we have used a method in the \textit{FrF2} R
-package called \textit{DanielPlot}. This method assumes that the
+To evaluate the graphs we have used a method in the \emph{FrF2} R
+package called \emph{DanielPlot}. This method assumes that the
 estimated effects are normally distributed with means equal to the
@@
 them selves are a measure for which factor that is the most
-significant.  We have also set the \textit{alpha} to $0.05$ which
+significant.  We have also set the \emph{alpha} to $0.05$ which
 means that we will tolerate a false percentage of $5\%$.
@@
 faster. By reading the results from the linear model it is the factor
-level \textit{-vso} which has an overall faster response time.
+level \emph{-vso} which has an overall faster response time.
 
@@
 new factors in comparison to the six factor experiment. The newly
-added factors were \textit{weight}, \textit{nr. of doors},
-\textit{nr. of seats} and \textit{fuel consumption}. We used the same
+added factors were \emph{weight}, \emph{nr. of doors},
+\emph{nr. of seats} and \emph{fuel consumption}. We used the same
 approach here as in the previous experiments to calculate the average
@@
 higher than emission. The linear model again tells us that the option
-\textit{-vso} has a significantly lower response time than
-\textit{-co}
+\emph{-vso} has a significantly lower response time than
+\emph{-co}
 

### 2014-07-02 cold2014/magnus.tex
add conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@704 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 planning matrix in appendix \ref{appB}. With 10 factors the amount of
-runs were $2^10$. This results in 1024 number of runs which made the
+runs were $2^{10}$. This results in 1024 number of runs which made the
 planning matrix too big to show here.
@@
 
+\section{Conclusions}
+
+We recommend that both alternatives should be used together for
+representing complex products. % TODO: Spiser ikke helt denne
+Using only CO there should be done some alterations to reduce
+bottleneck issues. We would not recommend using the data represented
+by Renault today, because it is not robust enough and has a too long
+response time for bigger operations.  It is also important that the
+data representation is done properly to ensure readability for both
+humans and machines.  Today, the Renault representation is neither
+very understandable for machine nor for human. This is reflected in
+the discussion about the application development.  The results from
+the experiments done against the application shows that the current
+representation is not feasible with today's performance standards and
+user expectations. This is the area where most of the workload is on
+the use of the ontologies.  This means that for a third party user VSO
+in conjunction with COO are the best solution to work with product
+models. However, this approach is limited to only represent data about
+car models. It would also require some reasoning to be effective, but
+with the technology and knowledge today that should not be a to
+difficult issue. For other users wanting to represent constraints
+between product models, CO can be a viable options if the proper
+alterations is done. It can be very useful when working with several
+data domains, but then the compatibility issue has to be solved.  One
+solution could be to use the compatibility approach that COO
+offers. This would result in adding several new triples in each
+lexicon to avoid the configuration traversals.  It would also add the
+need for reasoning to be effective, which is opposite of what Renault
+proposed. If the compatibility issue is solved we believe the
+performance can be feasible for almost every scenario. It will demand
+more from the third party user, but can utilize more than just data
+about car models. To make it even more effective, there should be done
+several other alterations to both the representation and the ontology
+as discussed in Section \ref{Alterations}.
+
+Our recommendation is based on giving the user freedom to query the
+data in several ways.  A future goal would be to do a more thorough
+test of the ontologies for all possible scenarios, since the
+versatility of the ontologies are huge. We believe that ultimately the
+data from manufacturers and others can be profitable for several
+online vendors and third party sites. Other manufactures of complex
+product could profit from using semantic
+technologies to represent their data.
+
+Based on these investigations we conclude:
+\begin{itemize}
+ \item Both ontology approaches can be useful, but the generic needs several alterations to be more robust.
+ \item The representation of the data is just as important as the ontology structure.
+ \item Data should be as machine readable as possible, and properties denote more information, rather than classes if it is possible.
+ \item To speed up the performance of the generic approach, the compatibility issue has to be resolved. One solution is to use the approach proposed in COO.
+ \item Reasoning over the product models should be applied to increase efficiency and versatility.
+\end{itemize}
+
 % TODO: In conclusion, discuss external validity.

### 2014-07-02 cold2014/magnus.tex
stop using vref
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@705 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 input values for any specification they might want to search for.  In
-Figure \vref{initialView} we can see how the application initially
+Figure \ref{initialView} we can see how the application initially
 looks for a user. Here the user can input values for the different
@@
 the user is presented with a set of valid models to choose from, which
-contain the chosen specifications. In Figure \vref{afterSearch}, we
+contain the chosen specifications. In Figure \ref{afterSearch}, we
 can see what the user is presented with after a successful query
@@
 Like the rest of the experiments we used DanielPlot to evaluate and a
-package in R to visualize the result. The Figure \vref{3factorGraph}
+package in R to visualize the result. The Figure \ref{3factorGraph}
 shows the results.  Here we see no significant factors. Still there
-are small effects from several factors. In Table \vref{3factorEffect}
+are small effects from several factors. In Table \ref{3factorEffect}
 we can see the time effects from largest to smallest.  We see that the
@@
 
-In the graph \vref{10factorGraph} we can see the results from running
+In the graph \ref{10factorGraph} we can see the results from running
 all the 1024 runs.  We can see that there are a lot of significant
@@
 experiments, except the two factor experiment. Here we find the
-ontology factor.  In Table \vref{10factorEffect} we can see the top
+ontology factor.  In Table \ref{10factorEffect} we can see the top
 twenty absolute effects. This includes all the significant factors

### 2014-07-02 cold2014/magnus.tex
subsection experiments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@706 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsection{Experiment with three factors}
+We ran several experiments, both randomized and unrandomized
+experiments, with 3, 4, 6 and 10 factors. Here, we report on the
+unrandomized 3 and 10 full factorial experiments:
+
+\subsubsection{Experiment with three factors}
 The first experiment that was done, was the experiment with only three
@@
 
-\subsection{Experiment with ten factors}
+\subsubsection{Experiment with ten factors}
 The last experiment we did was a ten factor experiment. The goal here
@@
 We recommend that both alternatives should be used together for
-representing complex products. % TODO: Spiser ikke helt denne
+representing complex products. 
 Using only CO there should be done some alterations to reduce

### 2014-07-02 cold2014/magnus.tex
introducing the ontologies
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@707 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We present two different approaches, one promoted by Renault, see
-\cite{SemWebAppRes} and \cite{ren1}, and another made by Martin~Hepp
-in collaboration with Volkswagen resulting in amongst other things,
-the Car Options Ontology~\cite{COO}. We shall compare them to find
-their strengths and weaknesses. Each approach contains an ontology and
-has its own way of representing data.  To make this comparison we have
-used data about the same domain, data about car models and their
-component constraints.  The first approach is a generic ontology. By
-generic it means that it can be used to represent any product model
-with component constraints. The second approach is a domain specific
+We present two different approaches. One is promoted by Renault, see
+\cite{SemWebAppRes} and \cite{ren1}. They have developed the
+Configuration Ontology (CO) \cite{confOnt}, and developed an API to allow
+users to traverse the RDF graph to find a smaller result set of cars.
+The ontology focuses on building an RDF graph to represent all valid
+configurations for a product model with component constraints.  It is
+also made so that it is not needed to do any reasoning to get a valid
+result set.  They call it a traversal of \textit{Configurations} which
+will end up with a \textit{Partially Defined Products}
+(PDP)\footnote{Partially Defined Product is a way of defining a
+  product without using all of the features. For instance a car which
+  may have many features, but we describe it as a car with mp3 and
+  sunroof.}.  Their method insists there is a starting point, which is
+a car model. From there, there is a \emph{lexicon} which can be used
+to choose different specifications which may lead to a valid
+configuration. There are also stated if some specifications are
+impossible in the current configuration.  These are represented by
+either the \textit{co:possible} or \textit{co:impossible}
+property. Furthermore, string comparison must be used to check if a
+certain specification is representing for instance fuel type.
+
+The other approach was made by Martin~Hepp in collaboration with
+Volkswagen resulting in the Car Options Ontology (COO) \cite{COO} and
+the Vehicle Sales Ontology (VSO). They are designed to be used in
+combination with GoodRelations, also designed by Hepp. When using COO,
+one defines a base car model, and a \emph{trim} is defined on the
+base, while a derivative car model should specify compatibilities and
+constraints between configurations, e.g. an electrical car does not
+have a conventional gear box.
+
+We shall compare them to find their strengths and
+weaknesses. Each approach contains an ontology and has its own way of
+representing data.  To make this comparison we have used data about
+the same domain, data about car models and their component
+constraints.  The first approach is a generic ontology. By generic it
+means that it can be used to represent any product model with
+component constraints. The second approach is a domain specific
 ontology which as the name implies, is only applicable with one
@@
 where a user is seeking to buy a new car based on detailed data
-published by car manufacturers. Today only Renault, of all the car
-manufacturers, has opened their data using semantic
-technologies. Volkswagen had their data published until the start of
-2013, but unfortunately they canceled their Semantic Web project and
-the data were removed.  Fortunately, their ontologies are still open
-for the community to use. Eventually, we found Daimler very
-forthcoming, as they offered data about their A- and B-class cars.
-The Daimler data used a custom XML-based format, so we lifted the data
-using a Python script into the ontologies developed by Martin Hepp and
-Volkswagen.
+published by car manufacturers. Note that the implementation is not
+itself an important contribution, it is the evaluation that is the key
+contribution. Today only Renault, of all the car manufacturers, has
+opened their data using semantic technologies. Volkswagen had their
+data published until the start of 2013, but unfortunately they
+canceled their Semantic Web project and the data were removed.
+Fortunately, their ontologies are still open for the community to
+use. Eventually, we found Daimler very forthcoming, as they offered
+data about their A- and B-class cars.  The Daimler data used a custom
+XML-based format, so we lifted the data using a Python script into the
+ontologies developed by Martin Hepp and Volkswagen.
 

### 2014-07-02 cold2014/magnus.tex
classes vs properties
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@708 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+A central difference with these ontologies is that CO allows the
+configuration to be defined in terms of classes, whereas VSO
+represents them with properties, for example:
+\begin{lstlisting}[basicstyle=\tiny, frame=single]
+#CO representation of fuel type
+:var_PT1628  a           co:ConfigurationVariable , owl:Class ;
+        rdfs:label       "Fuel Type"@en ;
+        co:confVarId   "PT1628" ;
+        co:hasValue    <http://uk.co.rplug.renault.com/product/gen/spec/PT1628_diesel/-#this> .
+
+#VSO representation of fuel type
+daim:A200CDI vso:fuelType daim:A200CDI_fuelType_Diesel_fuel .
+\end{lstlisting}
+
 We shall compare them to find their strengths and
@@
 car manufacturers.  With performance we mean the response time between
-a HTTP~\cite{http}} post operation against the application and when
+a HTTP post operation against the application and when
 the application presents the user with an answer. The application will

### 2014-07-02 cold2014/magnus.tex
mention alignment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@709 43ef1190-90fd-4dd7-a769-d0121f02b80e
remove todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@710 43ef1190-90fd-4dd7-a769-d0121f02b80e
add discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@711 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+To align the above ontologies, we used SPARQL CONSTRUCT statements
+from the lexicon to each specification.
+
 The application present users with a web interface where they can
@@
 describe more in depth how the car model search is done later on.
-
-% TODO differences between Daimler and Renault
-% TODO Alignment
 
@@
 
+\section{Discussion}
+
+We will discuss the results in terms of \textit{representation} and
+\textit{structure}. The structure is the ontology definitions from
+classes to properties.  The representation is how the actual data is
+represented with each ontology. For instance, how the fuel type
+\textit{Diesel} is represented.
+
+\subsection{The small scale experiments}
+The first experiment, using only three factors, was done to simulate a
+small workload on the ontologies. This can represent an application
+that present the user for a specification one at the time. The
+configurator Renault has on their web page does exactly
+that.\footnote{\url{http://www.semanlink.net/2012/cold/configurator.html}}
+Our application made the workload on both ontologies to be as equal as
+possible.  The only difference from a programming perspective was that
+with CO we had to move to a new configuration URI after each
+specification choice.
+
+In the Figure \vref{3factorGraph} we can see that only one significant
+effects, the O factor. In the results data found in appendix
+\ref{appD} we can see that the Configuration Ontology was almost two
+seconds slower than the VSO/COO runs.  Three out of four CO runs kept
+the response time around two seconds, but one run had to use
+approximately 3.5 seconds which resulted in an average run time of
+$\approx2.3$ seconds.  In a study done to research the amount of time
+a user would be willing to wait, they found out that the tolerable
+waiting time for any information retrieval was approximately 2
+seconds.~\cite{waitTime} This means that with three factors we were
+still within the time frame for a good user experience.
+
+One reason that CO was slower than VSO/COO for this experiment is that
+all the car models represented with CO is placed in a separate RDF
+graph. Each model had their own lexicon with all the information.
+This means that an application has to retrieve in all 22 RDF graphs,
+the amount of cars Renault has present in their API. These 22
+retrievals are done right after a HTTP post is detected. This differs
+from how the cars in VSO/COO are represented. There it is one RDF
+graph per base model, which means that there were only two RDF graphs
+to retrieve.  In our application this results in 20 more RDF graph
+retrievals with CO than VSO/COO and because the three factor
+experiment was so small, it had a higher impact than in the later
+experiments.
+
+Another factor that would influence the results is the amount of
+values the search has to iterate over.  In Table \vref{averages}, we
+presented all the average values and the average number of values
+present in the Daimler and Renault data. We can see that CO have
+$0.54$ more fuel types on average to iterate over and $0.3$ more
+transmissions. This is not a lot, but can account for some delay in
+the response time.
+
+\subsection{The large scale experiments}
+The last two experiments we define as large scale experiments. This is
+because the workload and time consumption were significantly larger
+than the two previous experiments. Both of these experiments were done
+to simulate a workload for an application were one could define all
+the specifications at once. Defining all specifications means that a
+user can query for all specifications that may be present in a product
+model.
+
+The last experiment done was the one with ten factors. This experiment
+had four more factors than the six factor experiment.  We chose to add
+factors which represented specifications that were present in every
+car model in each RDF graph. That was because some specifications like
+height was not always declared in a car model and then would skew the
+results. These incidents were present in CO's RDF graphs.
+
+In Figure \vref{10factorGraph} we can see the results from this
+experiment. There are unfortunately a lot of illegible significant
+factors, but these are the less significant ones. In Table
+\vref{10factorEffect} we see the top twenty absolute effects and
+similar to all the other experiments, O is on top of the list. The
+average response time of a run done against CO was $\approx58.5$
+seconds while a run against VSO/COO had an average response time of
+$\approx0.3$ seconds.  This corresponds with the absolute effect of
+the O factor to be 58 seconds.
+
+If we take a look at the other factors, Fuel Consumption (FC) has
+taken the top single factor spot on the list.  This can be explained
+with the same reason as when emission was topping the list in the six
+factor experiment. Each car has a lot of possible FC values to iterate
+over. In CO, each car has an average amount of values of approximately
+5.8. For VSO/COO it is 3.75.  This is the second highest overall after
+emission. This by itself does not explain why FC is on the top of the
+list, but when we take a look at when FC is queried it becomes more
+clear. The order each specification is queried is shown below.  We can
+see that FC comes right before E which means that it will have a
+greater impact on the amount of values checked later.
+
+ \begin{enumerate}
+
+		\item Gears (G)
+		\item Transmission (T)
+		\item Fuel Consumption (FC)
+		\item Emission (E)
+		\item Doors (D)
+		\item Total Weight (WT)
+		\item Fuel Type (FT)
+		\item Weight (W)
+		\item Seating Capacity (SC)
+\end{enumerate}
+
+If we take another look at Table \ref{10factorEffect} we see that
+there are only three single factors except O that are in the top 15
+significant factors.  After the 15 significant factor the absolute
+effect drops significantly. The three factors are E, FC and T. E and T
+are present because of the same reasons as in the six factor
+experiment. We also see in the top 15 that there are only these three
+factors present in all the factor interactions as well, which
+indicates that T, FC and E affects each other. This is logical because
+if a car has a high fuel consumption it also has a high emission
+rate. There are some possible explanations for why there are so many
+significant factor interactions in some of the experiments. The order
+will affect this as mention above. Another explanation is that we
+chose an average value to be an upper and lower limit of a factor. We
+used these as factor levels to avoid an unrealistic high number of
+levels. This could have opened for more new specifications to check on
+one level of the factor.  As we see in the Table
+\vref{10factorEffect}, \textit{E:WT} is higher than \textit{WT}
+itself. This might indicate that one of the levels of E will open up
+for more values to traverse in one of the levels of WT. This is a
+complex matter which will depend on each product model as well as the
+interaction between the specifications.
+
+\section{Pros and cons}
+There are several ways of making an application handle PDPs. As
+mentioned there can be differences in collecting data, presenting data
+and using data. This means that a recommended ontology and data
+representation needs to be robust. This means that it has to be able
+to perform well under different scenarios.  The ideal semantic
+situation would not rely on string comparison, but with both
+ontologies one have to deal with string literals eventually. A string
+literal is still the value a user is presented with.  It depends on
+how much of the data one wants machine accessible and how much is for
+internal human reading.
+
+The ontologies in this thesis are not made for tackling every possible
+problem, but can perform well within their domain. The ontology
+structure of CO in conjunction of the data representation, is made for
+one particular scenario. That is to specify one specification after
+another, not all at once. Here CO has several tools at disposal to
+make it easier trimming down the result sets. The property
+\textit{co:impossible} was to no use in our application due that our
+application specified all specifications before searching. This
+property lets the application remove possibilities if a user have
+chosen some specifications. One would still has to use the lexicon as
+reference, but could eliminate further possibilities. The property
+\textit{co:impliedSpec} does something similar and gives the user a
+list of specifications that are implied and do not need to be searched
+for.  COO have some of the same functions and possibilities, but some
+of these has to be reasoned over to be useful. One can produce the
+same functionality by using \textit{coo:incompatibleWith} with
+reasoning.
+
+From the performance tests and programming, the product specific
+ontology came out as the best.  This, as discussed, is not solely due
+to the ontology structure. In the generic ontology, many of the issues
+lies within the representation.  The generic ontology can easily be
+used to represent different PDPs which will let the user stick to one
+ontology.  Some alterations to the representation is needed to make it
+more robust than what Renault have done.  They mention that reasoning
+is something a third party user should be shielded from. We disagree
+because if the user can choose when the reasoning is done they can
+avoid possible bottlenecks. This means that for the representation to
+work properly we recommend using a similar compatibility system to
+COO. Here one can add the triples to each RDF graph of the product
+model and do reasoning here. This gives the user freedom to choose
+whatever way of reasoning and storing they might find most applicable,
+which could be to pre-compute any big reasoning operations that may be
+unsuited for real time. This will remove all the extra queries which
+has to be done to move through Renault's large graph of
+configurations. This representation puts more workload on the third
+party user, but it will most likely decrease the response time
+drastically.
+
 \section{Conclusions}

### 2014-07-02 cold2014/magnus.tex
challenges to validity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@712 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Pros and cons}
+\subsection{Pros and cons}
 There are several ways of making an application handle PDPs. As
@@
 drastically.
+
+\subsection{Challenges to validity of methodology}
+
+There are several challenges to the validity of present
+study. Following the discussion of \cite{publication-9417}, we
+think that the most important problem to the internal validity is
+related to the ordering of which the specifications are queried. We
+have argued that the order is important, but we have not investigated
+whether a change of the order will affect the overall performance.
+
+The main challenge to construct validity is the choice of factors in
+the large scale experiments: Even though a 10 factor experiment is a
+large full factorial experiment, and although the most significant
+factors are the same in all experiments, we have not investigated
+whether other factors could influence the result as much.
+
+Since the present study represents a common use case for Semantic Web
+application, we are of the opinion that there is a strong case for its
+external validity, but it does obviously rely on that the study itself
+is valid.
 

### 2014-07-02 cold2014/magnus.tex
minor clean
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@713 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 factors present in all the factor interactions as well, which
-indicates that T, FC and E affects each other. This is logical because
-if a car has a high fuel consumption it also has a high emission
-rate. There are some possible explanations for why there are so many
-significant factor interactions in some of the experiments. The order
-will affect this as mention above. Another explanation is that we
-chose an average value to be an upper and lower limit of a factor. We
-used these as factor levels to avoid an unrealistic high number of
-levels. This could have opened for more new specifications to check on
-one level of the factor.  As we see in the Table
-\vref{10factorEffect}, \textit{E:WT} is higher than \textit{WT}
+indicates that T, FC and E affects each other. This is to be expected
+because if a car has a high fuel consumption it also has a high
+emission rate. There are some possible explanations for why there are
+so many significant factor interactions in some of the
+experiments. The order will affect this as mention above. Another
+explanation is that we chose an average value to be an upper and lower
+limit of a factor. We used these as factor levels to avoid an
+unrealistic high number of levels. This could have opened for more new
+specifications to check on one level of the factor.  As we see in the
+Table \vref{10factorEffect}, \textit{E:WT} is higher than \textit{WT}
 itself. This might indicate that one of the levels of E will open up
@@
 
-% TODO: In conclusion, discuss external validity.
-
 \end{document}

### 2014-07-02 cold2014/magnus.tex
vref to ref textit to emph
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@714 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{wrapfig}
+\usepackage{listings}
 
@@
 also made so that it is not needed to do any reasoning to get a valid
-result set.  They call it a traversal of \textit{Configurations} which
-will end up with a \textit{Partially Defined Products}
+result set.  They call it a traversal of \emph{Configurations} which
+will end up with a \emph{Partially Defined Products}
 (PDP)\footnote{Partially Defined Product is a way of defining a
@@
 impossible in the current configuration.  These are represented by
-either the \textit{co:possible} or \textit{co:impossible}
+either the \emph{co:possible} or \emph{co:impossible}
 property. Furthermore, string comparison must be used to check if a
@@
 
-We will discuss the results in terms of \textit{representation} and
-\textit{structure}. The structure is the ontology definitions from
+We will discuss the results in terms of \emph{representation} and
+\emph{structure}. The structure is the ontology definitions from
 classes to properties.  The representation is how the actual data is
 represented with each ontology. For instance, how the fuel type
-\textit{Diesel} is represented.
+\emph{Diesel} is represented.
 
@@
 
-In the Figure \vref{3factorGraph} we can see that only one significant
+In the Figure \ref{3factorGraph} we can see that only one significant
 effects, the O factor. In the results data found in appendix
@@
 Another factor that would influence the results is the amount of
-values the search has to iterate over.  In Table \vref{averages}, we
+values the search has to iterate over.  In Table \ref{averages}, we
 presented all the average values and the average number of values
@@
 
-In Figure \vref{10factorGraph} we can see the results from this
+In Figure \ref{10factorGraph} we can see the results from this
 experiment. There are unfortunately a lot of illegible significant
 factors, but these are the less significant ones. In Table
-\vref{10factorEffect} we see the top twenty absolute effects and
+\ref{10factorEffect} we see the top twenty absolute effects and
 similar to all the other experiments, O is on top of the list. The
@@
 specifications to check on one level of the factor.  As we see in the
-Table \vref{10factorEffect}, \textit{E:WT} is higher than \textit{WT}
+Table \ref{10factorEffect}, \emph{E:WT} is higher than \emph{WT}
 itself. This might indicate that one of the levels of E will open up
@@
 make it easier trimming down the result sets. The property
-\textit{co:impossible} was to no use in our application due that our
+\emph{co:impossible} was to no use in our application due that our
 application specified all specifications before searching. This
@@
 reference, but could eliminate further possibilities. The property
-\textit{co:impliedSpec} does something similar and gives the user a
+\emph{co:impliedSpec} does something similar and gives the user a
 list of specifications that are implied and do not need to be searched
@@
 of these has to be reasoned over to be useful. One can produce the
-same functionality by using \textit{coo:incompatibleWith} with
+same functionality by using \emph{coo:incompatibleWith} with
 reasoning.

### 2014-07-02 cold2014/magnus.tex
mention Lenths method
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@715 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 impossible in the current configuration.  These are represented by
-either the \emph{co:possible} or \emph{co:impossible}
+either the \textsf{co:possible} or \textsf{co:impossible}
 property. Furthermore, string comparison must be used to check if a
@@
 
-In the experiments the definition for VSO/COO is \emph{-vso} and for
-CO is \emph{-co}.  These two definitions stand for which ontology
+In the experiments the definition for VSO/COO is \textsf{-vso} and for
+CO is \textsf{-co}.  These two definitions stand for which ontology
 are tested in a particular run. In practice this means that when we
-send in the keyword \emph{-vso} we only run the specification search
+send in the keyword \textsf{-vso} we only run the specification search
 against the car models represented by VSO/COO, in our case the data
@@
 the levels of each factor.  Further in the script we used the package
-\emph{request}\footnote{http://requests.readthedocs.org/en/latest/}
+\textsf{request}\footnote{http://requests.readthedocs.org/en/latest/}
 to do the HTTP posts in python. Below one can see the line of code
@@
 result we used the programming language R with the packages
-\emph{DoE.base}~\cite{DoEBase}, \emph{FrF2}~\cite{FrF2} and
-\emph{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
+\textsf{DoE.base}~\cite{DoEBase}, \textsf{FrF2}~\cite{FrF2} and
+\textsf{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
 from the experiments and output them in two types of graphs.  A
@@
 
-To evaluate the graphs we have used a method in the \emph{FrF2} R
-package called \emph{DanielPlot}. This method assumes that the
+To evaluate the graphs we have used a method in the \textsf{FrF2} R
+package called \textsf{DanielPlot}. This method assumes that the
 estimated effects are normally distributed with means equal to the
@@
 them selves are a measure for which factor that is the most
-significant.  We have also set the \emph{alpha} to $0.05$ which
-means that we will tolerate a false percentage of $5\%$.
+significant.  We have also set  the significance level at $\alpha =
+0.05$, where \textsf{DanielPlot} is using Lenth's method for the
+analysis of unreplicated factorials. % TODO: Need a citation here.
 
@@
 faster. By reading the results from the linear model it is the factor
-level \emph{-vso} which has an overall faster response time.
+level \textsf{-vso} which has an overall faster response time.
 
@@
 higher than emission. The linear model again tells us that the option
-\emph{-vso} has a significantly lower response time than
-\emph{-co}
+\textsf{-vso} has a significantly lower response time than
+\textsf{-co}
 
@@
 specifications to check on one level of the factor.  As we see in the
-Table \ref{10factorEffect}, \emph{E:WT} is higher than \emph{WT}
+Table \ref{10factorEffect}, \textsf{E:WT} is higher than \textsf{WT}
 itself. This might indicate that one of the levels of E will open up
@@
 make it easier trimming down the result sets. The property
-\emph{co:impossible} was to no use in our application due that our
+\textsf{co:impossible} was to no use in our application due that our
 application specified all specifications before searching. This
@@
 reference, but could eliminate further possibilities. The property
-\emph{co:impliedSpec} does something similar and gives the user a
+\textsf{co:impliedSpec} does something similar and gives the user a
 list of specifications that are implied and do not need to be searched
@@
 of these has to be reasoned over to be useful. One can produce the
-same functionality by using \emph{coo:incompatibleWith} with
+same functionality by using \textsf{coo:incompatibleWith} with
 reasoning.

### 2014-07-03 cold2014/magnus.tex
change the formulation of external validity a bit
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@716 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Since the present study represents a common use case for Semantic Web
-application, we are of the opinion that there is a strong case for its
-external validity, but it does obviously rely on that the study itself
-is valid.
+Since the present study uses actually deployed ontologies and data,
+and also addresses a common use case for Semantic Web applications, we
+are of the opinion that there is a strong case for its external
+validity, but it does obviously rely on that the study itself is
+valid.
 

### 2014-07-03 cold2014/magnus.tex
Fjernet overflodig tekst
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@717 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 chosen to focus on a product selection use case, specifically on cars,
-since some car makers have embraced the Semantic Web vision and chosen
+since some car manufacturers have embraced the Semantic Web vision and chosen
 to share detailed data on their products.
@@
         co:confVarId   "PT1628" ;
-        co:hasValue    <http://uk.co.rplug.renault.com/product/gen/spec/PT1628_diesel/-#this> .
+        co:hasValue    
+	    <http://uk.co.rplug.renault.com/product/gen/spec/PT1628_diesel/-#this> .
 
@@
 data about their A- and B-class cars.  The Daimler data used a custom
-XML-based format, so we lifted the data using a Python script into the
+XML-based format, so we lifted the data using a \textit{python} script into the
 ontologies developed by Martin Hepp and Volkswagen.
@@
 well as the ontology factor. With these three factors we got $2^3$
-number of runs.  This meant that we had 8 runs to execute against our
-application. This experiment was also replicated four times to get the
-most accurate results. The experiment was small so the replication did
-not cause any issues.  The results were calculated by finding the
+number of runs. This experiment was also replicated four times to get the
+most accurate results.  The results were calculated by finding the
 average response time for each run between all four replications.
@@
 The next step was to create the planning matrix for this
-experiment. As mention in the beginning of this section we used a
-package in python to create and fill out the planning matrix. Table
+experiment. Table
 \ref{3factor} shows all the 8 runs for the experiment with three
@@
 planning matrix in appendix \ref{appB}. With 10 factors the amount of
-runs were $2^{10}$. This results in 1024 number of runs which made the
-planning matrix too big to show here.
+runs were $2^{10}$, which made the planning matrix too big to show here.
 
@@
 some of them that are illegible, but most of them have a low impact on
-the significance graph. Like the six factor experiment we also see
-here that there are three groups of significant factors. The first
-group where most of them are placed range from around 2 seconds effect
-to 11. Here there are a lot of factor interactions and why they are
-there will be explained in the discussion.  The second group is very
-like the second group we saw in the six factor experiment, the only
-difference is that the total weight (WT) factor is substituted with
-fuel consumption (FC). The last group is the same for all the
-experiments, except the two factor experiment. Here we find the
-ontology factor.  In Table \ref{10factorEffect} we can see the top
-twenty absolute effects. This includes all the significant factors
-from the last group and the second one, but also 5 from the first
-group.  We can see that emission also here are present in a lot of the
+the significance graph.  In Table \ref{10factorEffect} we can see the top
+twenty absolute effects. We can see that emission also here are present in a lot of the
 entries, but so is ontology as well. The new thing here is the fuel
@@
 \subsection{The large scale experiments}
-The last two experiments we define as large scale experiments. This is
+The last experiment we define as a large scale experiment. This is
 because the workload and time consumption were significantly larger
-than the two previous experiments. Both of these experiments were done
+than the previous experiment. This experiment were done
 to simulate a workload for an application were one could define all
@@
 
-The last experiment done was the one with ten factors. This experiment
-had four more factors than the six factor experiment.  We chose to add
+This experiment had seven more factors than the three factor experiment.  We chose to add
 factors which represented specifications that were present in every
@@
 more from the third party user, but can utilize more than just data
-about car models. To make it even more effective, there should be done
-several other alterations to both the representation and the ontology
-as discussed in Section \ref{Alterations}.
+about car models.
 

### 2014-07-03 cold2014/magnus.tex
Running the experiment er trimmet
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@718 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \documentclass{llncs}
-%\documentclass{article}
+\documentclass{article}
 \usepackage[utf8]{inputenc}
@@
 
-\begin{table}
-\begin{center}
-    \begin{tabular}{ | l | l  l  l |}
-    \hline
-    {\bf Run} &  {\bf Fuel type} & {\bf Transmission} & {\bf Speed} \\ \hline
-	 1 & Diesel & Automatic & 100 \\ \hline
-	 2 & Petrol & Automatic & 100 \\ \hline
-	 3 & Diesel & Manual & 100 \\ \hline
-	 4 & Petrol & Manual & 100 \\ \hline
-	 5 & Diesel & Automatic & 200 \\ \hline
-	 6 & Petrol & Automatic & 200 \\ \hline
-	 7 & Diesel & Manual & 200 \\ \hline
-	 8 & Petrol & Manual &  200 \\ \hline
-    \end{tabular}
-\end{center}
-\caption{A filled planning matrix with 3 factors}\label{designsspec}
-\end{table}
+%\begin{table}
+%\begin{center}
+%    \begin{tabular}{ | l | l  l  l |}
+%    \hline
+%    {\bf Run} &  {\bf Fuel type} & {\bf Transmission} & {\bf Speed} \\ \hline
+%	 1 & Diesel & Automatic & 100 \\ \hline
+%	 2 & Petrol & Automatic & 100 \\ \hline
+%	 3 & Diesel & Manual & 100 \\ \hline
+%	 4 & Petrol & Manual & 100 \\ \hline
+%	 5 & Diesel & Automatic & 200 \\ \hline
+%	 6 & Petrol & Automatic & 200 \\ \hline
+%	 7 & Diesel & Manual & 200 \\ \hline
+%	 8 & Petrol & Manual &  200 \\ \hline
+%    \end{tabular}
+%\end{center}
+%\caption{A filled planning matrix with 3 factors}\label{designsspec}
+%\end{table}
 
@@
 of the ontologies are done as similar as possible on both levels. This
-was done to eliminate any interference from the application. There are
-some differences in the programming due that the ontologies were
-different and they will be discussed later on.  In Chapter
-\ref{prototype}, the application and its differences are explained in
-depth.
+was done to eliminate any interference from the application.
 
-Before starting on the experiment we had to find a way to do the
-actual experiment. We wanted to first create the planning matrix and
+First we wanted to create the planning matrix and
 then execute each run as a HTTP post. To do this we needed to make a
@@
 DoE package we just declared the amount of factors and the matrix was
-created. At that point the matrix looked more like Table
-\ref{designs}, which meant that we had to substitute the + and - with
-the levels of each factor.  Further in the script we used the package
+created.  Further in the script we used the package
 \textsf{request}\footnote{http://requests.readthedocs.org/en/latest/}
@@
 
-All of the planning matrices, graphs and results will be presented
-with each experiment later on in this section.  The discussion and
-interpretation of the results will be presented in Chapter
-\ref{Discussion}.  Before starting the experiments we will present
+Before starting the experiments we will present
 abbreviations for each factor because their full name made the graphs
-totally illegible. These abbreviations will be used in the graphs and
-tables further on in this section.  The results from all the
-experiments can be found in appendix \ref{appD}.
+totally illegible. The results from all the
+experiments can be found in appendix \ref{appD}. %TODO: What to do with the appendix?
 \begin{itemize}

### 2014-07-03 cold2014/magnus.tex
wrapfigging
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@719 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \documentclass{llncs}
-\documentclass{article}
+%\documentclass{article}
 \usepackage[utf8]{inputenc}
@@
   \centering
-      \includegraphics[width=12cm]{initialView.png}
-  \caption{Initial view for a user}\label{initialView}
-\end{figure}
-
-\begin{figure}
-  \centering
       \includegraphics[width=12cm]{afterSearch.png}
@@
 
-%\begin{table}
+%\begin{wraptable}
 %\begin{center}
@@
 %\caption{A filled planning matrix with 3 factors}\label{designsspec}
-%\end{table}
+%\end{wraptable}
 
@@
 
-\begin{table}[H]
-\begin{center}
+\begin{wraptable}{l}{0.67\textwidth}
     \begin{tabular}{ | l | l l |}
@@
     \end{tabular}
-\end{center}
 \caption{Factors and levels for experiment 1}\label{facandlevExp1}
-\end{table}
+\end{wraptable}
 
@@
 were evaluated
-\begin{table}
-\begin{center}
+\begin{wraptable}{l}{0.67\textwidth}
     \begin{tabular}{ | l | l l l |}
@@
     \end{tabular}
-\end{center}
 \caption{Planning matrix for experiment with three factors}\label{3factor}
-\end{table}
+\end{wraptable}
 
 Like the rest of the experiments we used DanielPlot to evaluate and a
-package in R to visualize the result. The Figure \ref{3factorGraph}
-shows the results.  Here we see no significant factors. Still there
-are small effects from several factors. In Table \ref{3factorEffect}
+package in R to visualize the result. Figure \ref{3factorGraph}
+shows only one significant factor, namely the \textsf{O} factor in the
+upper right corner. In Table \ref{3factorEffect}
 we can see the time effects from largest to smallest.  We see that the
@@
 
-\begin{figure}
- \includegraphics[width=12cm]{2factorfinal.pdf}
-  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
-\end{figure}
-
-\begin{table}
-\begin{center}
+\begin{wraptable}{l}{0.35\textwidth}
+\vspace{-25pt}
     \begin{tabular}{ | l l |}
@@
     \end{tabular}
-\end{center}
 \caption{Table of effects for the three factor experiment}\label{3factorEffect}
-\end{table}
+\vspace{-65pt}
+\end{wraptable}
+
+\begin{figure}
+ \includegraphics[width=12cm, height=8cm]{2factorfinal.pdf}
+  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
+\end{figure}
 
@@
 
-\begin{table}
-\begin{center}
+\begin{wraptable}{l}{0.4\textwidth}
+\vspace{-25pt}
     \begin{tabular}{ | l l |}
@@
     \end{tabular}
-\end{center}
 \caption{The top twenty significant effects of the ten factor graph}\label{10factorEffect}
-\end{table}
+      \vspace{-25pt}
+\end{wraptable}
 

### 2014-07-03 cold2014/magnus.tex
remove screenshot entirely
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@720 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the user is presented with a set of valid models to choose from, which
-contain the chosen specifications. In Figure \ref{afterSearch}, we
-can see what the user is presented with after a successful query
-execution, in this case seventeen different models. In this run, the
-user have searched for a model with diesel fuel, automatic
-transmission, four or more gears and a maximum of six seats. We also
-have the values and URIs for each of these valid models, which means
-that the application could present the user with more information than
-it does today.
+contain the chosen specifications, e.g. the user have searched for a
+model with diesel fuel, automatic transmission, four or more gears and
+a maximum of six seats.
 
-\begin{figure}
-  \centering
-      \includegraphics[width=12cm]{afterSearch.png}
-  \caption{The view after a search}\label{afterSearch}
-\end{figure}
 There are done some pre-computations to make the application a little

### 2014-07-03 cold2014/magnus.tex
make itemize a wrappable table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@721 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-First we wanted to create the planning matrix and
-then execute each run as a HTTP post. To do this we needed to make a
-small script to calculate the $2^k$ runs and perform the HTTP
-posts. We chose to use the programming language \emph{python} to do
-this. That was because python was a familiar language and it also had
-a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the
-DoE package we just declared the amount of factors and the matrix was
-created.  Further in the script we used the package
-\textsf{request}\footnote{http://requests.readthedocs.org/en/latest/}
-to do the HTTP posts in python. Below one can see the line of code
-executing the request.
-
-\scriptsize\begin{lstlisting}
- r = requests.post('http://localhost:8080/linkedopendata-magnudae/index', 
-		  data=payload)
-\end{lstlisting}
-\normalsize
+First, we created the planning matrix and then executed each run as a
+HTTP post. To do this we needed to make a small script to calculate
+the $2^k$ runs and perform the HTTP posts. We used the programming
+language \emph{Python} with the
+DoE\footnote{http://pythonhosted.org/pyDOE/} package.
 
@@
 abbreviations for each factor because their full name made the graphs
-totally illegible. The results from all the
-experiments can be found in appendix \ref{appD}. %TODO: What to do with the appendix?
-\begin{itemize}
-  \item Ontology = O
-  \item Fuel Type = FT
-  \item Transmission = T
-  \item Weight = W
-  \item Total Weight = WT
-  \item Emission = E
-  \item Nr. of Gears = G
-  \item Seating Capacity = SC
-  \item Fuel Consumption = FC
-  \item Doors = D
-\end{itemize}
+totally illegible.
+
+\begin{wraptable}{l}{0.35\textwidth}
+    \begin{tabular}{ | l | l |}
+    \hline
+    {\bf Factor} & {\bf Abbrev} \\ \hline
+   Ontology & \textsf{O} \\
+   Fuel Type & \textsf{FT} \\
+   Transmission & \textsf{T} \\
+   Weight & \textsf{W} \\
+   Total Weight & \textsf{WT} \\
+   Emission & \textsf{E} \\
+   Nr. of Gears & \textsf{G} \\
+   Seating Capacity & \textsf{SC} \\
+   Fuel Consumption & \textsf{FC} \\
+   Doors & \textsf{D} \\
+\hline
+    \end{tabular}
+\caption{Factors and their abbreviation}\label{facandabbrev}
+\end{wraptable}
 
@@
 
-\begin{wraptable}{l}{0.35\textwidth}
+\begin{wraptable}{r}{0.35\textwidth}
 \vspace{-25pt}
@@
 \caption{Table of effects for the three factor experiment}\label{3factorEffect}
-\vspace{-65pt}
+\vspace{-25pt}
 \end{wraptable}
 
-\begin{figure}
- \includegraphics[width=12cm, height=8cm]{2factorfinal.pdf}
+\begin{wrapfigure}{l}{0.65\textwidth}
+ \includegraphics[width=8.5cm, height=7cm]{2factorfinal.pdf}
   \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
-\end{figure}
+\end{wrapfigure}
 

### 2014-07-03 cold2014/magnus.tex
reduce precision
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@722 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     {\bf Factors} & {\bf Absolute Effect}  \\ \hline
-	  O     & 2.2463476\\ \hline
-	  O:T    &0.4536135 \\ \hline
-	  O:FT:T &0.4534723 \\ \hline
-	  T      &0.4513712 \\ \hline
-	  FT:T   &0.4499820\\ \hline
-	  O:FT   &0.1240831 \\ \hline
-	  FT     &0.1189629 \\ \hline
+	  O     & 2.246\\ \hline
+	  O:T    &0.454 \\ \hline
+	  O:FT:T &0.453 \\ \hline
+	  T      &0.451 \\ \hline
+	  FT:T   &0.450 \\ \hline
+	  O:FT   &0.124 \\ \hline
+	  FT     &0.118 \\ \hline
     \end{tabular}
@@
     {\bf Factors} & {\bf Absolute Effect}  \\ \hline
-      O & 58.1785542324229 \\ \hline
-      E:FC & 45.6971032050781 \\ \hline
-      O:E:FC & 45.5502538925781\\ \hline
-      FC & 38.6140886777344\\ \hline
-      O:FC & 38.5571969433593\\ \hline
-      O:E & 35.3967871113281\\ \hline
-      E & 35.3827983613282\\ \hline
-      O:T & 33.1249425527344\\ \hline
-      T & 33.1154579746093\\ \hline
-      O:T:E:FC & 29.3077075019531\\ \hline
-      T:E:FC & 29.2986405957031\\ \hline
-      O:T:FC & 28.2420862089844\\ \hline
-      T:FC & 28.2325333808593\\ \hline
-      O:T:E & 25.3121093378906\\ \hline
-      T:E & 25.3039236347656\\ \hline
-      WT:E & 11.9191989902344\\ \hline
-      O:WT:E & 11.8908780371094\\ \hline
-      O:WT:FC & 11.7234715332032\\ \hline
-      WT:FC & 11.7140515488281\\ \hline
-      O:D:E & 10.1241599902344\\ \hline
+      O & 58.17 \\ \hline
+      E:FC & 45.69 \\ \hline
+      O:E:FC & 45.55\\ \hline
+      FC & 38.61\\ \hline
+      O:FC & 38.56\\ \hline
+      O:E & 35.40\\ \hline
+      E & 35.38\\ \hline
+      O:T & 33.12\\ \hline
+      T & 33.12\\ \hline
+      O:T:E:FC & 29.31\\ \hline
+      T:E:FC & 29.30\\ \hline
+      O:T:FC & 28.24\\ \hline
+      T:FC & 28.23\\ \hline
+      O:T:E & 25.31\\ \hline
+      T:E & 25.30\\ \hline
+      WT:E & 11.92\\ \hline
+      O:WT:E & 11.89\\ \hline
+      O:WT:FC & 11.72\\ \hline
+      WT:FC & 11.71\\ \hline
+      O:D:E & 10.12\\ \hline
     \end{tabular}

### 2014-07-03 cold2014/magnus.tex
small fixes in spacing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@723 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{wraptable}{r}{0.35\textwidth}
-\vspace{-25pt}
+\begin{wraptable}{l}{0.25\textwidth}
+%\vspace{-25pt}
     \begin{tabular}{ | l l |}
     \hline
-    {\bf Factors} & {\bf Absolute Effect}  \\ \hline
+    {\bf Factors} & {\bf Effect}  \\ \hline
 	  O     & 2.246\\ \hline
@@
     \end{tabular}
-\caption{Table of effects for the three factor experiment}\label{3factorEffect}
-\vspace{-25pt}
+\caption{Table of absolute effects for the three factor experiment}\label{3factorEffect}
+%\vspace{-25pt}
 \end{wraptable}
@@
 
-\begin{wraptable}{l}{0.4\textwidth}
+\begin{wraptable}{l}{0.25\textwidth}
 \vspace{-25pt}
@@
     \hline
-    {\bf Factors} & {\bf Absolute Effect}  \\ \hline
+    {\bf Factors} & {\bf Effect}  \\ \hline
       O & 58.17 \\ \hline
@@
     \end{tabular}
-\caption{The top twenty significant effects of the ten factor graph}\label{10factorEffect}
+\caption{The top twenty significant absolute effects of the ten factor graph}\label{10factorEffect}
       \vspace{-25pt}

### 2014-07-03 cold2014/magnus.tex
remove levels table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@724 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \subsubsection{Experiment with three factors}
-The first experiment that was done, was the experiment with only three
-factors. The goal here was to see how the ontologies behave with a
-small amount of querying and factors.  Table \ref{facandlevExp1} show
-the different factors and their levels. We chose to start with a small
-experiment using only simple factors which had few possible
-values. That is why we chose transmission and fuel type as factors as
-well as the ontology factor. With these three factors we got $2^3$
-number of runs. This experiment was also replicated four times to get the
-most accurate results.  The results were calculated by finding the
-average response time for each run between all four replications.
-
-\begin{wraptable}{l}{0.67\textwidth}
-    \begin{tabular}{ | l | l l |}
-    \hline
-    {\bf Factor} & {\bf Level 1} & {\bf Level 2} \\ \hline
-	Ontology & -vso & -co \\ \hline 
-	Fuel type & Diesel & Unleaded Petrol \\ \hline 
-	Transmission & Manual Gearbox & Automatic Gearbox \\ \hline 
-    \end{tabular}
-\caption{Factors and levels for experiment 1}\label{facandlevExp1}
-\end{wraptable}
+We first made a very simple three factor experiment. The main goal here was
+to see how the ontologies behave with a small amount of querying and
+factors. We chose to start with a small experiment using only
+simple factors which had few possible values. That is why we chose
+transmission and fuel type as factors as well as the ontology
+factor. With these three factors we got $2^3$ number of runs. This
+experiment was also replicated four times to get the most accurate
+results.  The results were calculated by finding the average response
+time for each run between all four replications.
 
 The next step was to create the planning matrix for this
-experiment. Table
-\ref{3factor} shows all the 8 runs for the experiment with three
-factors. We also ran a randomized version the experiment to check
-whether the randomization would influence the experiment. The results
-were almost equal which eliminated randomization as a factor.  Both
-results can be found in appendix \ref{appD}. After each run the
-response time were logged to that particular run. The response time
-were then used to check for significant effect when the experiment
-were evaluated
+experiment. Table \ref{3factor} shows all the 8 runs for the
+experiment with three factors.  After each run the response time were
+logged to that particular run. The response time were then used to
+check for significant effect when the experiment were evaluated
 \begin{wraptable}{l}{0.67\textwidth}
@@
     \end{tabular}
-\caption{Planning matrix for experiment with three factors}\label{3factor}
+    \caption{Planning matrix for experiment with three factors and
+      their levels.}\label{3factor}
 \end{wraptable}

### 2014-07-03 cold2014/magnus.tex
use a tilde to avoid breakage
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@725 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 input values for any specification they might want to search for.  In
-Figure \ref{initialView} we can see how the application initially
+Figure~\ref{initialView} we can see how the application initially
 looks for a user. Here the user can input values for the different
@@
 
-Before starting the experiments we will present
-abbreviations for each factor because their full name made the graphs
-totally illegible.
+For brevity, we will use abbreviations given in Figure~\ref{facandabbrev}.
 
@@
 The next step was to create the planning matrix for this
-experiment. Table \ref{3factor} shows all the 8 runs for the
+experiment. Table~\ref{3factor} shows all the 8 runs for the
 experiment with three factors.  After each run the response time were
@@
 Like the rest of the experiments we used DanielPlot to evaluate and a
-package in R to visualize the result. Figure \ref{3factorGraph}
+package in R to visualize the result. Figure~\ref{3factorGraph}
 shows only one significant factor, namely the \textsf{O} factor in the
-upper right corner. In Table \ref{3factorEffect}
+upper right corner. In Table~\ref{3factorEffect}
 we can see the time effects from largest to smallest.  We see that the
@@
 value for the new factors. The average values can be seen in the
-planning matrix in appendix \ref{appB}. With 10 factors the amount of
+planning matrix in appendix~\ref{appB}. With 10 factors the amount of
 runs were $2^{10}$, which made the planning matrix too big to show here.
 
-In the graph \ref{10factorGraph} we can see the results from running
+In the graph~\ref{10factorGraph} we can see the results from running
 all the 1024 runs.  We can see that there are a lot of significant
@@
 some of them that are illegible, but most of them have a low impact on
-the significance graph.  In Table \ref{10factorEffect} we can see the top
+the significance graph.  In Table~\ref{10factorEffect} we can see the top
 twenty absolute effects. We can see that emission also here are present in a lot of the
@@
 
-In the Figure \ref{3factorGraph} we can see that only one significant
+In the Figure~\ref{3factorGraph} we can see that only one significant
 effects, the O factor. In the results data found in appendix
@@
 Another factor that would influence the results is the amount of
-values the search has to iterate over.  In Table \ref{averages}, we
+values the search has to iterate over.  In Table~\ref{averages}, we
 presented all the average values and the average number of values
@@
 
-In Figure \ref{10factorGraph} we can see the results from this
+In Figure~\ref{10factorGraph} we can see the results from this
 experiment. There are unfortunately a lot of illegible significant
@@
 
-If we take another look at Table \ref{10factorEffect} we see that
+If we take another look at Table~\ref{10factorEffect} we see that
 there are only three single factors except O that are in the top 15
@@
 specifications to check on one level of the factor.  As we see in the
-Table \ref{10factorEffect}, \textsf{E:WT} is higher than \textsf{WT}
+Table~\ref{10factorEffect}, \textsf{E:WT} is higher than \textsf{WT}
 itself. This might indicate that one of the levels of E will open up

### 2014-07-03 cold2014/magnus.tex
cleaning up references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@726 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 The application present users with a web interface where they can
-input values for any specification they might want to search for.  In
-Figure~\ref{initialView} we can see how the application initially
-looks for a user. Here the user can input values for the different
-specifications, for instance fuel type or CO$_2$ emission. After a
-user is done choosing specifications he can execute the actual car
-model search. The application extracts the data from the form and
-comprises it into another format that can easily be used later on. The
-next step begins in the search module. Here it starts a thread for
-each possible car model to execute each individual search. In the end
-the user is presented with a set of valid models to choose from, which
-contain the chosen specifications, e.g. the user have searched for a
-model with diesel fuel, automatic transmission, four or more gears and
-a maximum of six seats.
+input values for any specification they might want to search
+for. There the user can input values for the different specifications,
+for instance fuel type or CO$_2$ emission. After a user is done
+choosing specifications he can execute the actual car model
+search. The application extracts the data from the form and comprises
+it into another format that can easily be used later on. The next step
+begins in the search module. Here it starts a thread for each possible
+car model to execute each individual search. In the end the user is
+presented with a set of valid models to choose from, which contain the
+chosen specifications, e.g. the user have searched for a model with
+diesel fuel, automatic transmission, four or more gears and a maximum
+of six seats.
 
-There are done some pre-computations to make the application a little
+Some pre-computations are done to make the application a little
 more effective. This includes extracting all the model names from both
@@
 of levels. This means that there is $n^k$ \emph{factorial designs}
-which result in $n^k$ designs to execute.  In the thesis every
-experiment will only have two levels, but the factors may vary. This
+which result in $n^k$ designs to execute.  We have chosen only
+two-level experiments for this research, but the factors may vary. This
 means that we will use a $2^k$ full factorial design for every
@@
 in a \emph{planning matrix} that shows the actual runs in the
-experiment as tabulated in Table~\ref{designsspec}. This contains the
-chosen factors and their levels.
-
-%\begin{wraptable}
-%\begin{center}
-%    \begin{tabular}{ | l | l  l  l |}
-%    \hline
-%    {\bf Run} &  {\bf Fuel type} & {\bf Transmission} & {\bf Speed} \\ \hline
-%	 1 & Diesel & Automatic & 100 \\ \hline
-%	 2 & Petrol & Automatic & 100 \\ \hline
-%	 3 & Diesel & Manual & 100 \\ \hline
-%	 4 & Petrol & Manual & 100 \\ \hline
-%	 5 & Diesel & Automatic & 200 \\ \hline
-%	 6 & Petrol & Automatic & 200 \\ \hline
-%	 7 & Diesel & Manual & 200 \\ \hline
-%	 8 & Petrol & Manual &  200 \\ \hline
-%    \end{tabular}
-%\end{center}
-%\caption{A filled planning matrix with 3 factors}\label{designsspec}
-%\end{wraptable}
+experiment.
 
@@
 approach here as in the previous experiments to calculate the average
-value for the new factors. The average values can be seen in the
-planning matrix in appendix~\ref{appB}. With 10 factors the amount of
+value for the new factors. % TODO: Explain more how the levels were set?
+With 10 factors the amount of
 runs were $2^{10}$, which made the planning matrix too big to show here.
@@
 In the Figure~\ref{3factorGraph} we can see that only one significant
-effects, the O factor. In the results data found in appendix
-\ref{appD} we can see that the Configuration Ontology was almost two
-seconds slower than the VSO/COO runs.  Three out of four CO runs kept
-the response time around two seconds, but one run had to use
-approximately 3.5 seconds which resulted in an average run time of
-$\approx2.3$ seconds.  In a study done to research the amount of time
-a user would be willing to wait, they found out that the tolerable
-waiting time for any information retrieval was approximately 2
-seconds.~\cite{waitTime} This means that with three factors we were
-still within the time frame for a good user experience.
+effects, the O factor. We saw in Table~\ref{3factorEffect} that the
+Configuration Ontology was almost two seconds slower than the VSO/COO
+runs.  Three out of four CO runs kept the response time around two
+seconds, but one run had to use approximately 3.5 seconds which
+resulted in an average run time of $\approx2.3$ seconds.  In a study
+done to research the amount of time a user would be willing to wait,
+they found out that the tolerable waiting time for any information
+retrieval was approximately 2 seconds.~\cite{waitTime} This means that
+with three factors we were still within the time frame for a good user
+experience.
 
@@
 Another factor that would influence the results is the amount of
-values the search has to iterate over.  In Table~\ref{averages}, we
-presented all the average values and the average number of values
-present in the Daimler and Renault data. We can see that CO have
+values the search has to iterate over.  We have found that CO have
 $0.54$ more fuel types on average to iterate over and $0.3$ more

### 2014-07-03 cold2014/magnus.tex
comment vspace for now
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@727 43ef1190-90fd-4dd7-a769-d0121f02b80e
run spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@728 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   the application. For the evaluation, we employ the statistical
-  disipline of Design of Experiments.
+  discipline of Design of Experiments.
 
@@
 them to allow a user to find a certain combination of properties that
-suit their purpose is an archetypical Semantic Web use case. We have
+suit their purpose is an archetypal Semantic Web use case. We have
 chosen to focus on a product selection use case, specifically on cars,
@@
 H. Afsarmanesh and M. Shafahi (2013) proposed a complex product
-specification system.~\cite{NewConf} This include object modelling and
+specification system.~\cite{NewConf} This include object modeling and
 a user interface. They have focused on supporting stakeholders in the
@@
    Emission & \textsf{E} \\
-   Nr. of Gears & \textsf{G} \\
+   # of Gears & \textsf{G} \\
    Seating Capacity & \textsf{SC} \\
@@
 new factors in comparison to the six factor experiment. The newly
-added factors were \emph{weight}, \emph{nr. of doors},
-\emph{nr. of seats} and \emph{fuel consumption}. We used the same
+added factors were \emph{weight}, \emph{# of doors},
+\emph{# of seats} and \emph{fuel consumption}. We used the same
 approach here as in the previous experiments to calculate the average
@@
 \begin{wraptable}{l}{0.25\textwidth}
-\vspace{-25pt}
+%\vspace{-25pt}
     \begin{tabular}{ | l l |}
@@
 \caption{The top twenty significant absolute effects of the ten factor graph}\label{10factorEffect}
-      \vspace{-25pt}
+%      \vspace{-25pt}
 \end{wraptable}

### 2014-07-03 cold2014/magnus.tex
ooops # is special
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@729 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
    Emission & \textsf{E} \\
-   # of Gears & \textsf{G} \\
+   \# of Gears & \textsf{G} \\
    Seating Capacity & \textsf{SC} \\
@@
 new factors in comparison to the six factor experiment. The newly
-added factors were \emph{weight}, \emph{# of doors},
-\emph{# of seats} and \emph{fuel consumption}. We used the same
+added factors were \emph{weight}, \emph{\# of doors},
+\emph{\# of seats} and \emph{fuel consumption}. We used the same
 approach here as in the previous experiments to calculate the average

### 2014-07-03 cold2014/magnus.tex
reference spec for GR
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@730 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The Volkswagen solution builds upon GoodRelations, which is a web
-vocabulary for e-commerce and was launched in 2008 and is now
-widespread. GoodRelations.~\cite{GR} GoodRelations can be used for
-detailed information about products to be sold online.
+The Volkswagen solution builds upon GoodRelations\cite{GRQual}, which
+is a web vocabulary for e-commerce and was launched in 2008 and is now
+widespread.  GoodRelations can be used for detailed information about
+products to be sold online.
 
@@
 \subsection{Planning the experiment}
-Like other approaches, there are several steps in doing experiments with
-DoE.  First one has to state an objective which is important to give
-the experiment a purpose.  The next step is to choose a response. The
-response is the experimental outcome or observation. There may be
-multiple responses in an experiment. In this experiment there will be
-one response and that is the response time on a HTTP post against the
-application.  These first two steps are the same for all the
-experiments in this thesis.
+Like other approaches, there are several steps in doing experiments
+with DoE. A good textbook on the subject is \cite{PlanExp}. First one
+has to state an objective which is important to give the experiment a
+purpose.  The next step is to choose a response. The response is the
+experimental outcome or observation. There may be multiple responses
+in an experiment. In this experiment there will be one response and
+that is the response time on a HTTP post against the application.
+These first two steps are the same for all the experiments in this
+thesis.
 
@@
 conclusions. More about these steps will be taken individually at each
-experiment.~\cite{PlanExp}
+experiment.
 
@@
 in the top right corner. This means that the highest performance
-significance will end up in
-the top right corner.~\cite{Plotting} \\
-Here is a quick explanation of the axes shown in each graph.  The X
-axis will show the absolute effect of each factor, shown in time. The
-values on the X axis is represented in seconds which means that the a
-plot will have a significance measured in seconds.  The Y axis shows
-the half-normal scores. There are two way of representing the graph
-and that is with either half-normal plots or normal plots. With
-half-normal the effects on the X axis are shown with the absolute
-effect. This means that all factors are placed with a positive
-half-normal score. This is to avoid visual misleading scores because
-with the normal scores, deviations from the line on negative scores
-might confuse the reader who are evaluating the charts.  The scores
-them selves are a measure for which factor that is the most
-significant.  We have also set  the significance level at $\alpha =
+significance will end up in the top right corner. The
+axes themselves are not important to understand the plot nor the
+results, refer to \cite{PlanExp} for details.
+We have also set  the significance level at $\alpha =
 0.05$, where \textsf{DanielPlot} is using Lenth's method for the
-analysis of unreplicated factorials. % TODO: Need a citation here.
+analysis of unreplicated factorials.
 
@@
 
+\bibliographystyle{plain}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{jbact}
+%\bibliographystyle{splncs03}
+\bibliography{bibliography}
+
 \end{document}

### 2014-07-05 cold2014/magnus.tex
add references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@731 43ef1190-90fd-4dd7-a769-d0121f02b80e
Removed some new lines
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@732 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 of six seats.
-
-Some pre-computations are done to make the application a little
-more effective. This includes extracting all the model names from both
-ontologies, initiate all possible partially defined products as a
-PartialCar object and creating some internal structure. These
-pre-computations were done so that it would be possible to present the
-user with the name of each car model, not just the URI. We will also
-describe more in depth how the car model search is done later on.
 
@@
 %\bibliographystyle{splncs03}
-\bibliography{bibliography}
+\bibliography{bibliography,science}
 

### 2014-07-05 cold2014/magnus.tex
Removed some new lines in discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@733 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 model search. With the results from this experiment we could discuss
-several aspects of using the different ontologies.  Here we added 4
-new factors in comparison to the six factor experiment. The newly
-added factors were \emph{weight}, \emph{\# of doors},
-\emph{\# of seats} and \emph{fuel consumption}. We used the same
-approach here as in the previous experiments to calculate the average
-value for the new factors. % TODO: Explain more how the levels were set?
+several aspects of using the different ontologies.   % TODO: Explain more how the levels were set?
 With 10 factors the amount of
 runs were $2^{10}$, which made the planning matrix too big to show here. 
+To determine the levels of each factor, we calculated the average value 
+of each factor present in the data sets.
 
@@
 all the 1024 runs.  We can see that there are a lot of significant
-effects present here. There are more than any of the previous
-experiments, but it is also more factors and factor interactions.
+effects present here. 
 Unfortunately there are so many significant factors that there are
@@
 the significance graph.  In Table~\ref{10factorEffect} we can see the top
-twenty absolute effects. We can see that emission also here are present in a lot of the
-entries, but so is ontology as well. The new thing here is the fuel
-consumption has entered the most significant factors, and is by itself
-higher than emission. The linear model again tells us that the option
+fifteen absolute effects. We can see that emission and fuel consumption are 
+the most significant factors besides the ontology factor.
+The linear model again tells us that the option
 \textsf{-vso} has a significantly lower response time than
@@
       T:E & 25.30\\ \hline
-      WT:E & 11.92\\ \hline
-      O:WT:E & 11.89\\ \hline
-      O:WT:FC & 11.72\\ \hline
-      WT:FC & 11.71\\ \hline
-      O:D:E & 10.12\\ \hline
     \end{tabular}
-\caption{The top twenty significant absolute effects of the ten factor graph}\label{10factorEffect}
+\caption{The top fifteen significant absolute effects of the ten factor graph}\label{10factorEffect}
 %      \vspace{-25pt}
@@
 small workload on the ontologies. This can represent an application
-that present the user for a specification one at the time. The
-configurator Renault has on their web page does exactly
-that.\footnote{\url{http://www.semanlink.net/2012/cold/configurator.html}}
+that present the user for a specification one at the time. 
 Our application made the workload on both ontologies to be as equal as
-possible.  The only difference from a programming perspective was that
-with CO we had to move to a new configuration URI after each
-specification choice.
+possible.
 
@@
 
-This experiment had seven more factors than the three factor experiment.  We chose to add
-factors which represented specifications that were present in every
-car model in each RDF graph. That was because some specifications like
-height was not always declared in a car model and then would skew the
-results. These incidents were present in CO's RDF graphs.
-
 In Figure~\ref{10factorGraph} we can see the results from this
-experiment. There are unfortunately a lot of illegible significant
-factors, but these are the less significant ones. In Table
+experiment. In Table
 \ref{10factorEffect} we see the top twenty absolute effects and

### 2014-07-05 cold2014/magnus.tex
Small change in conluscion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@734 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 If we take a look at the other factors, Fuel Consumption (FC) has
-taken the top single factor spot on the list.  This can be explained
-with the same reason as when emission was topping the list in the six
-factor experiment. Each car has a lot of possible FC values to iterate
+taken the top single factor spot on the list.  Each car has a lot of 
+possible FC values to iterate
 over. In CO, each car has an average amount of values of approximately
@@
  \begin{enumerate}
-
 		\item Gears (G)
@@
 there are only three single factors except O that are in the top 15
-significant factors.  After the 15 significant factor the absolute
-effect drops significantly. The three factors are E, FC and T. E and T
+significant factors. The three factors are E, FC and T. E and T
 are present because of the same reasons as in the six factor
@@
 representation is not feasible with today's performance standards and
-user expectations. This is the area where most of the workload is on
-the use of the ontologies.  This means that for a third party user VSO
+user expectations. This means that for a third party user VSO
 in conjunction with COO are the best solution to work with product
@@
 with the technology and knowledge today that should not be a to
-difficult issue. For other users wanting to represent constraints
-between product models, CO can be a viable options if the proper
-alterations is done. It can be very useful when working with several
+difficult issue. For other scenarios, CO can be a viable option with 
+some alterations. It can be very useful when working with several
 data domains, but then the compatibility issue has to be solved.  One

### 2014-07-05 cold2014/magnus.tex
Big change in conluscion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@735 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Our recommendation is based on giving the user freedom to query the
-data in several ways.  A future goal would be to do a more thorough
-test of the ontologies for all possible scenarios, since the
-versatility of the ontologies are huge. We believe that ultimately the
-data from manufacturers and others can be profitable for several
-online vendors and third party sites. Other manufactures of complex
-product could profit from using semantic
-technologies to represent their data.
-
 Based on these investigations we conclude:

### 2014-07-06 cold2014/magnus.tex
Removed order list
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@736 43ef1190-90fd-4dd7-a769-d0121f02b80e
Small running experiment change
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@737 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The \emph{payload} is a dictionary of all the values to post to the
-application.  After the request we saved the response which contained
-the response time of a run. After each request we saved the run plus
+After each request we saved the run plus
 the response time and wrote it to file on CSV format. To evaluate the
@@
 \textsf{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
-from the experiments and output them in two types of graphs.  A
-similarity with all the experiments were that they all contained at
-least the factors ontology, transmission and fuel type.
+from the experiments and output them in two types of graphs.
 
@@
 list, but when we take a look at when FC is queried it becomes more
-clear. The order each specification is queried is shown below.  We can
+clear. The order each specification is queried in this experiment 
+from first to last is: G - T - FC - E - D - WT - FT - W - SC.  We can
 see that FC comes right before E which means that it will have a
 greater impact on the amount of values checked later.
-
- \begin{enumerate}
-		\item Gears (G)
-		\item Transmission (T)
-		\item Fuel Consumption (FC)
-		\item Emission (E)
-		\item Doors (D)
-		\item Total Weight (WT)
-		\item Fuel Type (FT)
-		\item Weight (W)
-		\item Seating Capacity (SC)
-\end{enumerate}
 

### 2014-07-06 cold2014/magnus.tex
Removed chunk in ten factor experiment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@738 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Like the rest of the experiments we used DanielPlot to evaluate and a
-package in R to visualize the result. Figure~\ref{3factorGraph}
+Figure~\ref{3factorGraph}
 shows only one significant factor, namely the \textsf{O} factor in the
@@
 model search. With the results from this experiment we could discuss
-several aspects of using the different ontologies.   % TODO: Explain more how the levels were set?
+several aspects of using the different ontologies.  
 With 10 factors the amount of
@@
 \textsf{-vso} has a significantly lower response time than
-\textsf{-co}
-
-When we accumulated the results from all the runs we ended up with the
-total amount of response time to be approximately 30095 seconds, which
-is 8 hours, 21 minutes and 35 seconds.  This were the total response
-time after running all the runs. This did not take into account all
-the other steps during the experiments which took time, like writing
-results to file and initiating the application.  This discouraged the
-replication of the experiment due to the size and was also a reason
-why we did not do experiments with more factors.
+\textsf{-co}.
 

### 2014-07-08 cold2014/magnus.tex
some low-hanging fruit for shortening
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@739 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subtitle{---Draft submitted to COLD 2014---}
+%\subtitle{---Draft submitted to COLD 2014---}
 
@@
 chosen to focus on a product selection use case, specifically on cars,
-since some car manufacturers have embraced the Semantic Web vision and chosen
+since some car-makers have embraced the Semantic Web vision and chosen
 to share detailed data on their products.
@@
 
-A central difference with these ontologies is that CO allows the
-configuration to be defined in terms of classes, whereas VSO
-represents them with properties, for example:
+A key difference with these ontologies is that CO allows the
+configuration to be defined in terms of classes, while VSO
+represents them with properties, e.g:
 \begin{lstlisting}[basicstyle=\tiny, frame=single]
@@
 
-The Volkswagen solution builds upon GoodRelations\cite{GRQual}, which
-is a web vocabulary for e-commerce and was launched in 2008 and is now
+Volkswagen builds upon GoodRelations\cite{GRQual}, which
+is a web vocabulary for e-commerce that was launched in 2008 and is now
 widespread.  GoodRelations can be used for detailed information about
@@
 data about their A- and B-class cars.  The Daimler data used a custom
-XML-based format, so we lifted the data using a \textit{python} script into the
+XML-based format, so we lifted the data using a \textit{Python} script into the
 ontologies developed by Martin Hepp and Volkswagen.
@@
 different approaches, we chose to create an application where a user
-choose several or all car options specifications.
-
-The application was implemented using Apache Maven, Spring and Apache
-Tomcat as well as Jena.
+choose several or all car options specifications. The application was
+implemented using Apache Maven, Spring and Apache Tomcat as well as
+Jena.
 
@@
 which result in $n^k$ designs to execute.  We have chosen only
-two-level experiments for this research, but the factors may vary. This
-means that we will use a $2^k$ full factorial design for every
-experiment. The experiments consist of $2^k$ combinations with $k$
-factors. For instance if we have three factors where every factor got
-two levels we get $2^3$ designs, also called \emph{runs}. This results
-in a \emph{planning matrix} that shows the actual runs in the
-experiment.
+two-level experiments for this research, but the factors may
+vary. This means that we will use a $2^k$ full factorial design for
+every experiment. The experiments consist of $2^k$ combinations with
+$k$ factors. This results in a \emph{planning matrix} that shows the
+actual runs in the experiment.
 

### 2014-07-08 cold2014/magnus.tex
typo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@740 43ef1190-90fd-4dd7-a769-d0121f02b80e
smaller changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@741 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the same domain, data about car models and their component
-constraints.  The first approach is a generic ontology. By generic it
-means that it can be used to represent any product model with
+constraints.  The first approach is a generic ontology. By generic we
+mean that it can be used to represent any product model with
 component constraints. The second approach is a domain specific
@@
 published by car manufacturers. Note that the implementation is not
-itself an important contribution, it is the evaluation that is the key
+itself important, it is the evaluation that is the key
 contribution. Today only Renault, of all the car manufacturers, has
@@
 
-To be able to put large stress on the application, which again would
+To be able to put a major stress on the application, which again would
 allow us to bring out key strengths and weaknesses of the two
@@
 
-The application present users with a web interface where they can
+The application presents users with a web interface where they can
 input values for any specification they might want to search
-for. There the user can input values for the different specifications,
+for. There, the user can input values for the different specifications,
 for instance fuel type or CO$_2$ emission. After a user is done
@@
 In this section we will present several experiments with different
-amount of factors.  The objective is to find out how the ontologies
+number of factors.  The objective is to find out how the ontologies
 will perform in comparison and identifying which factors affect the
@@
 
-For brevity, we will use abbreviations given in Figure~\ref{facandabbrev}.
+For brevity, we will use abbreviations given in Table~\ref{facandabbrev}.
 

### 2014-07-08 cold2014/magnus.tex
explain ontology approach
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@742 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In this section we will present several experiments with different
-number of factors.  The objective is to find out how the ontologies
+number of factors.  The objective is to find out how the two approaches
 will perform in comparison and identifying which factors affect the
-outcome of each experiment. We did the ontology alignment and
-development of the application to prevent the loss of precision with
-either ontology.
+outcome of each experiment. Although the ontology is not the
+only issue that distinguishes these approaches, it is the key feature,
+and therefore, we will shall refer to that as a factor.
 

### 2014-07-08 cold2014/magnus.tex
remove "thesis"
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@743 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We shall compare them to find their strengths and
-weaknesses. Each approach contains an ontology and has its own way of
-representing data.  To make this comparison we have used data about
-the same domain, data about car models and their component
-constraints.  The first approach is a generic ontology. By generic we
-mean that it can be used to represent any product model with
-component constraints. The second approach is a domain specific
-ontology which as the name implies, is only applicable with one
-particular domain. In this thesis that domain is about car models.
+We shall compare them to find their strengths and weaknesses. Each
+approach contains an ontology and has its own way of representing
+data.  To make this comparison we have used data about the same
+domain, data about car models and their component constraints.  The
+first approach is a generic ontology, that is, it can be used to
+represent any product model with component constraints. The second
+approach is a domain specific ontology which as the name implies, is
+only applicable with one particular domain.
 
@@
 Unfortunately these articles do not present any research done with
-semantic technologies. The use of semantic technologies on complex
-products is a young field of research. The only thing done here is
-what Renault and Volkswagen have presented. Both of these car
-manufacturers have presented the public with two different
-solutions. They have also shown how to present the data on the web and
-the complexity around their solutions.
+semantic technologies. To our knowledge, the only research is what
+Renault and Volkswagen have presented. They have also shown how to
+present the data on the web and the complexity around their solutions.
 
@@
 These first two steps are the same for all the experiments in this
-thesis.
+paper.
 
@@
 next sub-section. There are other approaches like fractional factorial
-experiment, but they were not applicable or relevant to this thesis.
+experiment, which may be used for even larger experiments in future work.
 The last three steps of planning the experiment are performing the
@@
 experiment and is a factor in all the experiments which are presented
-in this thesis. The ontology factor represents the difference in how
+in this paper. The ontology factor represents the difference in how
 the data are represented and how the ontologies are structures, and
@@
 
-The ontologies in this thesis are not made for tackling every possible
+These ontologies are not made for tackling every possible
 problem, but can perform well within their domain. The ontology

### 2014-07-08 cold2014/magnus.tex
smaller changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@744 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We will also show how to create a viable web application which
-utilizes such complex data, mainly for the purpose of conducting
-performance tests to determine the weaknesses of each approach. This
-will be done with complex data found on the web today from different
-car manufacturers.  With performance we mean the response time between
-a HTTP post operation against the application and when
-the application presents the user with an answer. The application will
-contain the possibility to do HTTP posts against both approaches.
+We have created a web application prototype which uses such complex
+data, mainly for the purpose of conducting performance tests to
+determine the weaknesses of each approach. This will be done with
+complex data found on the web today from different car manufacturers.
+With performance we mean the response time between a HTTP post
+operation against the application and when the application presents
+the user with an answer. The application will contain the possibility
+to do HTTP posts against both approaches.
 
@@
 contribution. Today only Renault, of all the car manufacturers, has
-opened their data using semantic technologies. Volkswagen had their
+opened their data using Semantic Web technologies. Volkswagen had their
 data published until the start of 2013, but unfortunately they
@@
 \subsection{Full factorial experiment}
-A full factorial experiment is when one got $k$ factors and $n$ amount
-of levels. This means that there is $n^k$ \emph{factorial designs}
+A full factorial experiment is when one got $k$ factors and the number
+of levels is $n$. This means that there is $n^k$ \emph{factorial designs}
 which result in $n^k$ designs to execute.  We have chosen only
@@
 will perform in comparison and identifying which factors affect the
-outcome of each experiment. Although the ontology is not the
-only issue that distinguishes these approaches, it is the key feature,
-and therefore, we will shall refer to that as a factor.
+outcome of each experiment.
 
@@
 from Daimler. This is referred to as the ontology factor in each
-experiment and is a factor in all the experiments which are presented
+experiment, and is a factor in all the experiments which are presented
 in this paper. The ontology factor represents the difference in how
@@
 To evaluate the graphs we have used a method in the \textsf{FrF2} R
-package called \textsf{DanielPlot}. This method assumes that the
+package called \textsf{DanielPlot}. This method assumes that the % TODO: Dette må forbedres
 estimated effects are normally distributed with means equal to the

### 2014-07-10 cold2014/magnus.tex
rewrite the DanielPlot section a bit
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@745 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 To evaluate the graphs we have used a method in the \textsf{FrF2} R
-package called \textsf{DanielPlot}. This method assumes that the % TODO: Dette må forbedres
-estimated effects are normally distributed with means equal to the
-effects. The means of all estimated effects are zero. Resulting in a
-plot where the estimated effect would end up on a straight line. This
-plot is then testing whether all the estimated effects have the same
-distribution. All deviations of the straight line indicates a
-significant factor. In our results we have used the absolute value of
-the effects, also called a half-normal plot. The advantages of using
-the half-normal plot is that all large estimated effects will end up
-in the top right corner. This means that the highest performance
-significance will end up in the top right corner. The
-axes themselves are not important to understand the plot nor the
-results, refer to \cite{PlanExp} for details.
-We have also set  the significance level at $\alpha =
-0.05$, where \textsf{DanielPlot} is using Lenth's method for the
-analysis of unreplicated factorials.
+package called \textsf{DanielPlot} to create half-normal plots. The
+intuition behind this is simple: If errors are normally distributed,
+which is a conservative assumption, and there are no significant
+effects they will also be normally distributed, and so the estimated
+effects would end up on a straight line. Any significant deviations
+from a straight line indicates a significant factor. That we use a
+half-normal plot (rather than a normal plot) means that
+the most significant effects will end up in the top right
+corner. The axes themselves are not important to understand the plot
+nor the results, refer to \cite{PlanExp} for details.  We have also
+set the significance level at $\alpha = 0.05$, where
+\textsf{DanielPlot} is using Lenth's method for the analysis of
+unreplicated factorials.
 

### 2014-07-10 cold2014/magnus.tex
move table 1
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@746 43ef1190-90fd-4dd7-a769-d0121f02b80e
move table 2
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@747 43ef1190-90fd-4dd7-a769-d0121f02b80e
move table 3
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@748 43ef1190-90fd-4dd7-a769-d0121f02b80e
see if we can tolerate figure 1 as is
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@749 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the most significant effects will end up in the top right
-corner. The axes themselves are not important to understand the plot
-nor the results, refer to \cite{PlanExp} for details.  We have also
-set the significance level at $\alpha = 0.05$, where
-\textsf{DanielPlot} is using Lenth's method for the analysis of
-unreplicated factorials.
-
-For brevity, we will use abbreviations given in Table~\ref{facandabbrev}.
-
-\begin{wraptable}{l}{0.35\textwidth}
+corner. 
+%
+\begin{wraptable}{r}{0.35\textwidth}
+\vspace{-35pt}
     \begin{tabular}{ | l | l |}
@@
 \caption{Factors and their abbreviation}\label{facandabbrev}
+\vspace{-45pt}
 \end{wraptable}
+%
+The axes themselves are not important to understand the plot
+nor the results, refer to \cite{PlanExp} for details.  We have also
+set the significance level at $\alpha = 0.05$, where
+\textsf{DanielPlot} is using Lenth's method for the analysis of
+unreplicated factorials.
+
+For brevity, we will use abbreviations given in Table~\ref{facandabbrev}.
 
@@
 logged to that particular run. The response time were then used to
-check for significant effect when the experiment were evaluated
-\begin{wraptable}{l}{0.67\textwidth}
+check for significant effect when the experiment was evaluated.
+
+\begin{wraptable}{r}{0.67\textwidth}
+\vspace{-20pt}
     \begin{tabular}{ | l | l l l |}
@@
       their levels.}\label{3factor}
+\vspace{-20pt}
 \end{wraptable}
 
-Figure~\ref{3factorGraph}
-shows only one significant factor, namely the \textsf{O} factor in the
-upper right corner. In Table~\ref{3factorEffect}
-we can see the time effects from largest to smallest.  We see that the
-ontology factor has the largest effect on approximately 2.2 seconds
-which means that one of the ontologies overall is 2.2 seconds
-faster. By reading the results from the linear model it is the factor
-level \textsf{-vso} which has an overall faster response time.
+Figure~\ref{3factorGraph} shows only one significant factor, namely
+the \textsf{O} factor in the upper right corner. In
+Table~\ref{3factorEffect} we can see the time effects from largest to
+smallest.  We see that the ontology factor has the largest effect on
+approximately 2.2 seconds which means that one of the ontologies
+overall is 2.2 seconds faster. By reading the results from the linear
+model it is the factor level \textsf{-vso} which has an overall faster
+response time.
 
 \begin{wraptable}{l}{0.25\textwidth}
-%\vspace{-25pt}
+\vspace{-35pt}
     \begin{tabular}{ | l l |}
@@
 \caption{Table of absolute effects for the three factor experiment}\label{3factorEffect}
-%\vspace{-25pt}
+\vspace{-15pt}
 \end{wraptable}
-
-\begin{wrapfigure}{l}{0.65\textwidth}
- \includegraphics[width=8.5cm, height=7cm]{2factorfinal.pdf}
-  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
-\end{wrapfigure}
 
@@
 
-In the graph~\ref{10factorGraph} we can see the results from running
+In Figure~\ref{10factorGraph} we can see the results from running
 all the 1024 runs.  We can see that there are a lot of significant
@@
 \textsf{-co}.
+
+\begin{figure}%{r}{0.65\textwidth}
+\begin{center}
+\vspace{-15pt}
+ \includegraphics[width=8.5cm, height=7cm]{2factorfinal.pdf}
+  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
+\vspace{-10pt}
+\end{center}
+\end{figure}
 

### 2014-07-11 cold2014/magnus.tex
fixing the final figure
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@750 43ef1190-90fd-4dd7-a769-d0121f02b80e
remove too big to show
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@753 43ef1190-90fd-4dd7-a769-d0121f02b80e
mention mercedes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@754 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 contribution. Today only Renault, of all the car manufacturers, has
-opened their data using Semantic Web technologies. Volkswagen had their
-data published until the start of 2013, but unfortunately they
+opened their data using Semantic Web technologies. Volkswagen had
+their data published until the start of 2013, but unfortunately they
 canceled their Semantic Web project and the data were removed.
@@
 use. Eventually, we found Daimler very forthcoming, as they offered
-data about their A- and B-class cars.  The Daimler data used a custom
-XML-based format, so we lifted the data using a \textit{Python} script into the
-ontologies developed by Martin Hepp and Volkswagen.
+data about their Mercedes A- and B-class cars.  The Daimler data used
+a custom XML-based format, so we lifted the data using a
+\textit{Python} script into the ontologies developed by Martin Hepp
+and Volkswagen.
 
@@
 several aspects of using the different ontologies.  
-With 10 factors the amount of
-runs were $2^{10}$, which made the planning matrix too big to show here. 
+With 10 factors, we have $2^{10}$, which is quite large for a full
+factorial experiment.
 To determine the levels of each factor, we calculated the average value 
-of each factor present in the data sets.
+of each factor present in the data sets. % TODO: more on this.
 
@@
 \begin{wraptable}{l}{0.25\textwidth}
-%\vspace{-25pt}
+\vspace{-25pt}
     \begin{tabular}{ | l l |}
@@
 \caption{The top fifteen significant absolute effects of the ten factor graph}\label{10factorEffect}
-%      \vspace{-25pt}
+      \vspace{-75pt}
 \end{wraptable}
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 %\bibliographystyle{abbrv}
@@
 %\bibliographystyle{splncs03}
-\bibliography{bibliography,science}
+%\bibliography{bibliography,science}
+\input{magnus.bbl}
 

### 2014-07-11 cold2014/magnus.tex
final spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@755 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   We evaluate two different approaches that have been used in the
-  literature and in practice by different groups publishing metadata
+  literature and in practice by different groups publishing meta-data
   about cars when consumed by a prototype application. These

### 2014-08-20 background/67percent.tex
Omsorgsperm
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@758 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2013-08-13 background/70percent.tex
@@
-\documentclass[norsk]{uioletter}
+\documentclass{letter}
 \usepackage[utf8]{inputenc}
 
-\signature{Kjetil Kjernsmo}
+\name{Kjetil Kjernsmo}
 
-\ouremail{kjekje@ifi.uio.no}
+\address{Kjetil Kjernsmo\\Solfjellveien 22\\1389 Heggedal\\kjekje@ifi.uio.no}
+%\countersign{Martin Giese}
+%\countersigntitle{Veileder}
+
+\telephone{98648234}
 
@@
 
-Grunnet et ønske om å tilbringe mer tid med egne barn søkes det herved
-om å deltidsstilling fra dagens dato og ut høstsemesteret 2013. Det
-ønskes å sette ned stillingsprosenten til 70\%, og at stipendet
-forlenges tilsvarende.
+Grunnet nyankommet tredje barn ønsker undertegnede å ha delvis
+stilling som begynnes snarest mulig og varer t.o.m 15. januar
+2015. Det er allerede i dag sikkert at det ikke kommer til å søkes om
+ytterligere redusert stilling etter denne dato. Det søkes om en
+stillingsbrøk på $2/3$ og at stipendet forlenges tilsvarende.
 

### 2014-10-01 exp1/present2013.tex
mention Empirical add TODOs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@760 43ef1190-90fd-4dd7-a769-d0121f02b80e
ISWC 2013 was a 20 minute talk Extending from here Details on the experiment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@761 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+% TODO: My factors
+
 \note{
@@
 \textbf{Main goal}: Easy to understand for newcomers to DoE!
+
+\begin{itemize}
+\item Just 2 levels.
+\item Handful of illustrative factors.
+\item Analysis that illustrates key concepts.
+\item Straightforward to program and reproduce.
+\item Artificial performance differences.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Choices to meet these constraints}
+
+  \begin{itemize}
+  \item Response variable: Time from DNS lookup finishes to endpoint
+    has delivered a full response, measured by \texttt{curl}.
+  \item We don't compare different implementions, we compare the
+    performance before and after some change.
+  \item Query engine: \texttt{4store} where we insert \textsf{sleep}
+    statements in the join function on one level and language matching
+    function on the other. 
+  \end{itemize}
 
@@
 
+% TODO: Much around the hypothesis test
+
 Tests to run:
@@
 }
+
+% Experimental economy
 
@@
   \item Happy to help getting it to run!
-  \item Happy to discuss collaborations or workshops on this topic!
+  \item Happy to discuss collaborations!
+  \item The ESWC Empirical workshop is a venue for this study: http://empirical-ws.org/
   \end{itemize}

### 2014-10-01 exp1/present2013.tex
bump date to the Vienna talk
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@762 43ef1190-90fd-4dd7-a769-d0121f02b80e
reintroduce details on the experiment
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@763 43ef1190-90fd-4dd7-a769-d0121f02b80e
bugfix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@764 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \author{Kjetil Kjernsmo and John Tyssedal}
-\date{24.10.2013}
+\date{07.10.2014}
 \maketitle
@@
     machine with fast SSD and one larger with slower disks in RAID1.
+  \end{description}
+\end{frame}
+
+\begin{frame}[fragile]{The Basic Graph pattern factor}
+  \begin{description}
+
   \item [``BGPComp''] Two different basic graph patterns with varying
     complexity.
+  \end{description}
+
+  \begin{center}\small
+    \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.40\textwidth}|}
+      \hline
+      \textbf{Level 1} & \textbf{Level 2} \\[-2ex] 
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+      \end{verbatim}
+      & 
+      \begin{verbatim}
+?s  rdfs:label ?l1 ;
+    ?p1 ?o1 .
+?o1 dbo:populationTotal ?l2 .
+?s foaf:page ?o2 ;
+   dbpprop:subdivisionName ?o3 .
+?o3 skos:subject ?o4 ;
+   dbpprop:seat ?o5 ;
+   a ?c1 .
+\end{verbatim} 
+\\      \hline
+    \end{tabular}
+  \end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]{Absence or presence factors}
+  \begin{description}
   \item [``Union''] Absence or presence of a \textsf{UNION} pattern.
+
+    \begin{verbatim}
+{ ?o1 dbpprop:longd ?long ;
+      dbpprop:latd  ?lat .
+} UNION {
+  ?o1 geo:long ?long ;
+      geo:lat ?lat . }
+\end{verbatim}
+
   \item [``Lang''] Absence or presence of a \textsf{langMatches}
     filter.
+
+\begin{verbatim}
+FILTER langMatches( lang(?l1), "fr" ) 
+\end{verbatim}
+
   \item [``Range''] Absence or presence of a filter with a larger-than
     operator.
+
+    \begin{verbatim}
+FILTER (?l2 > 800000)
+\end{verbatim}
+
   \item [``Optional''] Absence or presence of an \textsf{OPTIONAL}
     pattern.
+    \begin{verbatim}
+OPTIONAL { ?o1 foaf:homepage ?o6 . }
+\end{verbatim}
 
@@
   \begin{itemize}
+% TODO: Lenth
   \item Those that are deemed significant are labelled.
@@
 
-% TODO: Much around the hypothesis test
+Classifying factors:
+  \begin{itemize}
+  \item ``Range'' and ``Machine'' are called \emph{inactive factors}.
+  \item ``BGPComp'', ``Lang'', ``Optional'', ``Union'' and ``TripleC''
+    are called \emph{environmental factors}.
+  \item ``Implement'' is called a \emph{control factor}.
+  \end{itemize}
 
-Tests to run:
+\end{frame}
+\begin{frame}{Hypothesis tests}
+
+\begin{itemize}
+\item Average over all the 32 level-combinations of the environmental
+  factors. 
+\end{itemize}
+
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Tue Oct  8 14:51:28 2013
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{rlll|r}
+  \hline
+ & Implement & Machine & Range & Response \\ 
+  \hline
+1 & 1 & 1 & 1 & 33.47 \\ 
+  2 & 2 & 1 & 1 & 11.82 \\ 
+  3 & 1 & 2 & 1 & 32.16 \\ 
+  4 & 2 & 2 & 1 & 11.56 \\ 
+  5 & 1 & 1 & 2 & 33.97 \\ 
+  6 & 2 & 1 & 2 & 12.46 \\ 
+  7 & 1 & 2 & 2 & 32.90 \\ 
+  8 & 2 & 2 & 2 & 11.60 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \begin{itemize}

### 2014-10-01 exp1/present2013.tex
reintroduce fractional experiments
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@765 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   \item ``Union'', ``Optional'' and ``Lang'' demanding things
-%  \item ``BGPComp:Lang'' and ``Lang:Union'' interactions might be due
-%    to that many more triples need to be searched for a language tag
-%    in one of the levels of the interacting factors.
+  \item ``BGPComp:Lang'' and ``Lang:Union'' interactions might be due
+    to that many more triples need to be searched for a language tag
+    in one of the levels of the interacting factors.
   \item But what about ``Implement::TripleC''? Leave that hanging for
@@
 \end{frame}
+\note{The paper shows how similar testing can be done to investigate
+  when the new implementation fails.}
+
+\subsection{Fractional Factorial Experiments}
+
+\begin{frame}{Experimental Economy}
+
+  \begin{itemize}
+  \item Full factorial experimental goes as $2^n$.
+  \item SPARQL Endpoint evaluation is inherently complex, with many
+    possible factors.
+  \item May reduce the size of the experiment with Fractional
+    Factorial Experiments.
+  \item The cost is explanatory power due to \emph{aliasing}.
+  \item E.g. we cannot tell the difference of an effect of
+    ``TripleC::BGPComp'' from ``Machine::Range''
+  \item Easy in R to declare which effects must not be aliased.
+  \item We have made two designs: One 32-run and one 64-run.
+  \end{itemize}
+
+\end{frame}
+\note{ That is to say, a detected increase in
+run time from larger Basic Graph Patterns for large databases, can
+also be explained by a less powerful machine that is worse at
+evaluating FILTER clauses with ranges. }
+
+\begin{frame}{32-run experiment}
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[height=0.85\textheight,width=.9\textwidth]{frac32normal.pdf}}
+  \label{fig:frac32normal}
+\end{figure}
+
+\end{frame}
+\note{
+
+  \begin{itemize}
+  \item The total variance is great, due to few runs, so few
+    conclusions can be drawn
+  \item Implementation is good, and Union is hard
+  \item In the paper, we also do a hypothesis test that returns $p =
+    0.00098$. Still very low
+  \item Very rough conclusions; useful in a CI system
+  \end{itemize}
+}
+
+\begin{frame}{64-run experiment}
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[height=0.85\textheight,width=0.96\textwidth]{frac64normal.pdf}}
+\end{figure}
+
+\end{frame}
+\note{Have two Lenth criterions, $\alpha=0.15$ and crosses at
+  $\alpha=0.05$. ($\alpha$ is the probability of rejecting $H_0$ when
+  it is true)}
+
+\begin{frame}{Analysis of 64-run experiment}
+
+  \begin{itemize}
+  \item Find most of the same significant effects as the full experiment.
+  \item However, we see that ``Implement:TripleC'' emerges only for
+  $\alpha=0.15$.
+  \item ``Machine'', ``TripleC'' and ``Range'' are inactive, we can
+    perform hypothesis test as above, and find $p = 5.9 \cdot 10^{-6}$.
+  \end{itemize}
+\end{frame}
 \note{
   \begin{itemize}
-  \item The paper shows how similar testing can be done to investigate
-  when the new implementation fails.
-
+  \item We see that ``Implement:TripleC'' emerges only for
+    $\alpha=0.15$, which is important to improve when designing a
+    study for actual tests.
+  \item We have shown that how the experiment could be set up, we have
+    shown the analysis and a hypothesis test.
   \item Should this simplistic experiment stand up to scrutiny?
@@
 
+\begin{frame}{What's a worthy result?}
+  \begin{itemize}
+  \item I have not made this a primary focus of my work.
+  \item Even though I think I should.
+  \item 3000 full-time researchers to construct the ATLAS experiment
+    at CERN.
+  \item Improvement in experimental methodology should be a worthy result.
+  \end{itemize}
+
+\end{frame}
+
 \begin{frame}{Thank you!}

### 2014-10-01 exp1/present2013.tex
reintroduce smaller things
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@766 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     better on CONSTRUCT queries, on small datasets
+  \item We still need to understand which effects are important and
+    significant, we do that in the discussion of our experiment
+
 \end{itemize}
@@
 
-\begin{frame}{The experiments}
+\begin{frame}{Constraints on scope of experiment}
 
@@
   \item Two factors do not contribute significantly to the variation
-    ``Range'' and ``Machine'', in the paper we describe how this
-    enables us to average over certain factors and look upon the rest
-    as a replicated experiment.
+    ``Range'' and ``Machine''.
   \end{itemize}

### 2014-10-01 exp1/present2013.tex
describe Lenth
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@767 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-% TODO: My factors
-
 \note{
@@
   \item With fractional, we trade explanatory power for economy.
-  \item We first study a full factorial, in the paper we have three
-    fractional, we shall see the final fractional
+  \item We first study a full factorial, then three fractional 
   \end{itemize}
@@
   \begin{itemize}
-% TODO: Lenth
+  \item Significance is based on Lenth's method, where he computes a
+    pseudo-standard error, since we have only one replicate. This is
+    based on computing the median of the absolute effects, and then
+    again the median of the smallest effects.
   \item Those that are deemed significant are labelled.
@@
 
-% Experimental economy
-
 \subsection{Tearing it all down}

### 2014-10-01 exp1/present2013.tex
take out the experiment description
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@768 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{block}{Description}
-  Experiment with 8 factors, 2 levels where we compare a single SPARQL
-  implementation before and after some change. We used \texttt{4store}
-  where we insert \textsf{sleep} statements in the join function on
-  one level and language matching function on the other, to simulate
-  an optimization that may have detrimental side-effects.
+\end{frame}
 
-\end{block}
+\note{
 
-\end{frame}
+  \begin{itemize}
+  \item Experiment with 8 factors, 2 levels
+  \item sleep to simulate an optimization that may have detrimental
+    side-effects
+  \end{itemize}
+
+}
 

### 2014-10-01 exp1/present2013.tex
add vskip
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@769 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-%\setbeameroption{show notes}
+\setbeameroption{show notes}
 \usepackage[utf8]{inputenc}
@@
 \note{
+  \vskip -18pt
   \begin{itemize}
-  \item Significance is based on Lenth's method, where he computes a
-    pseudo-standard error, since we have only one replicate. This is
-    based on computing the median of the absolute effects, and then
-    again the median of the smallest effects.
+  \item Lenth's method, one replicate: computes a pseudo-standard
+    error, median of the absolute effects, and then median of the
+    smallest effects.
   \item Those that are deemed significant are labelled.
@@
   \item But what about ``Implement::TripleC''? Leave that hanging for
-    a bit.
-  \item We want to do a hypothesis test
+    a bit. Do a hypothesis test
   \item Two factors do not contribute significantly to the variation

### 2014-10-01 exp1/analysis.R
emph myself as presenter
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@771 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -15,6 +15,7 @@ robust.simple <- function(model, control = "Implement",
                    pairwise=FALSE, ...) {
   fm <- as.formula(paste("experiments ~", control, " * ", paste(inactive, collapse=" * ")))
   allmeans <- aggregate(fm, data=model, mean)
+  browser()
   t.test(allmeans[allmeans[control] == 2,]$experiments,
          allmeans[allmeans[control] == 1,]$experiments,
          alternative="less")

### 2014-12-16 sanity/fetcher.pl
Add links to the request statements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@890 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -10,12 +10,12 @@ use Progress::Any;
 
 use RDF::Trine qw(iri statement literal);
 use RDF::Trine::Store::File::Quad;
-use RDF::Generator::HTTP;
+use RDF::Generator::HTTP 0.003;
 
 my $prfetch = Progress::Any->get_indicator(
         task => "fetching");
 my $prparse = Progress::Any->get_indicator(
-        task => "fetching");
+        task => "parsing");
 
 $prparse->update(message => "Setting up");
 
@@ -175,6 +175,7 @@ foreach my $host (@hosts) {
 																		],
 													  graph => $context);
 	  $hhg->generate($model);
+	  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $hhg->request_subject, $context);
 	  my $prevresponse = $firstresponse;
 
 	  if ($firstresponse->is_success) {
@@ -191,6 +192,7 @@ foreach my $host (@hosts) {
 																					  'Date'],
 																	graph => $context);
 			  $condhhg->generate($model);
+			  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context);
 			  $prevresponse = $condresponse;
 		  }
 
@@ -229,6 +231,7 @@ foreach my $host (@hosts) {
 																						'Date'],
 																	 graph => $context);
 			  $cond2hhg->generate($model);
+			  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $cond2hhg->request_subject, $context);
 			  if (($cond2response->code == 200) &&
 					($prevresponse->header('ETag') eq $cond2response->header('ETag')) &&
 					($prevresponse->header('Last-Modified') eq $cond2response->header('Last-Modified'))) {
@@ -303,6 +306,7 @@ foreach my $host (@hosts) {
 																								 ],
 																				graph => iri($endpoint));
 							  $ehhg->generate($model);
+							  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $ehhg->request_subject, $context);
 							  if ($eresponse->is_success) {
 								  my $anyres = has_sparql_results($eresponse->decoded_content, $eresponse->content_type) ? "Has results" : "No results";
 								  $model->add_statement(statement(iri($uri), iri('urn:app:endpoint'), literal($anyres), iri($endpoint)));
@@ -343,6 +347,7 @@ foreach my $host (@hosts) {
 																				 ],
[…]

### 2014-12-16 sanity/fetcher.pl
Increase sleep
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@891 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -183,7 +183,7 @@ foreach my $host (@hosts) {
 		  # What to do if data for conditional requests was available in BTC
 		  if ($firstresponse->code == 304) {
 			  my $condrequest = HTTP::Request->new(GET => $uri);
-			  sleep 5;
+			  sleep 10;
 			  my $condresponse = $ua->request( $condrequest );
 			  my $condhhg = RDF::Generator::HTTP->new(message => $condresponse,
 																	whitelist => ['Warning',
@@ -220,7 +220,7 @@ foreach my $host (@hosts) {
 		  }
 		  if ($ya) {
 			  no warnings 'uninitialized';
-			  sleep 5;
+			  sleep 10;
 			  my $cond2response = $ua->request( $cond2request );
 			  my $cond2hhg = RDF::Generator::HTTP->new(message => $cond2response,
 																	 whitelist => ['Warning',
@@ -350,7 +350,7 @@ foreach my $host (@hosts) {
 			  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $ahhg->request_subject, $context);
 		  }
 	  }
-	  sleep 5 if ($uricount > 1);
+	  sleep 10 if ($uricount > 1);
   }
   $pm->finish; # Terminates the child process
 }

### 2014-12-16 sanity/fetcher.pl
Don't let the process die if parser fails
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@893 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -276,7 +276,6 @@ foreach my $host (@hosts) {
 										  });
 				  } catch {
 					  $model->add_statement(statement(iri($uri), iri('urn:app:parseerror'), literal($_), $context));
-					  $pm->finish;
 				  };
 
 				  unless ($details->{type} eq 'vocabulary') {

### 2014-12-16 sanity/fetcher.pl
Ready for run3
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@894 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -23,7 +23,7 @@ $prparse->update(message => "Setting up");
 my $dct = RDF::Trine::Namespace->new('http://purl.org/dc/terms/');
 
 #my $basedir = '/home/kjetil/data/sanity/';
-my $basedir = '/mnt/ssdstore/data/btc-processed/';
+my $basedir = '/mnt/ssdstore/data/btc-processed/run3';
 
 my $writedir = $basedir . 'crawl/';
 

### 2014-12-16 sanity/fetcher.pl
Ready for run3
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@895 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -23,7 +23,7 @@ $prparse->update(message => "Setting up");
 my $dct = RDF::Trine::Namespace->new('http://purl.org/dc/terms/');
 
 #my $basedir = '/home/kjetil/data/sanity/';
-my $basedir = '/mnt/ssdstore/data/btc-processed/run3';
+my $basedir = '/mnt/ssdstore/data/btc-processed/run3/';
 
 my $writedir = $basedir . 'crawl/';
 

### 2014-12-16 sanity/fetcher.pl
Forgot statement wrappers
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@896 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -23,7 +23,7 @@ $prparse->update(message => "Setting up");
 my $dct = RDF::Trine::Namespace->new('http://purl.org/dc/terms/');
 
 #my $basedir = '/home/kjetil/data/sanity/';
-my $basedir = '/mnt/ssdstore/data/btc-processed/run3/';
+my $basedir = '/mnt/ssdstore/data/btc-processed/';
 
 my $writedir = $basedir . 'crawl/';
 
@@ -175,7 +175,8 @@ foreach my $host (@hosts) {
 																		],
 													  graph => $context);
 	  $hhg->generate($model);
-	  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $hhg->request_subject, $context);
+	  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $hhg->request_subject, $context));
+
 	  my $prevresponse = $firstresponse;
 
 	  if ($firstresponse->is_success) {
@@ -192,7 +193,7 @@ foreach my $host (@hosts) {
 																					  'Date'],
 																	graph => $context);
 			  $condhhg->generate($model);
-			  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context);
+			  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context));
 			  $prevresponse = $condresponse;
 		  }
 
@@ -231,7 +232,7 @@ foreach my $host (@hosts) {
 																						'Date'],
 																	 graph => $context);
 			  $cond2hhg->generate($model);
-			  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $cond2hhg->request_subject, $context);
+			  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $cond2hhg->request_subject, $context));
 			  if (($cond2response->code == 200) &&
 					($prevresponse->header('ETag') eq $cond2response->header('ETag')) &&
 					($prevresponse->header('Last-Modified') eq $cond2response->header('Last-Modified'))) {
@@ -305,7 +306,7 @@ foreach my $host (@hosts) {
 																								 ],
 																				graph => iri($endpoint));
 							  $ehhg->generate($model);
-							  $model->add_statement(iri($uri), iri('urn:app:hasrequest'), $ehhg->request_subject, $context);
[…]

### 2014-12-16 sanity/fetcher.pl
Ready  fro run4
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@897 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -92,7 +92,7 @@ $prparse->finish;
 $prfetch->update(message => "Initializing UA");
 
 use Parallel::ForkManager;
-my $pm = Parallel::ForkManager->new(25);
+my $pm = Parallel::ForkManager->new(30);
 
 use LWP::UserAgent;
 use HTTP::Request;

### 2014-12-16 sanity/paper1.tex
Start of introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@900 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass{llncs}
%\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{wrapfig}

\title{A survey of HTTP caching options on the open Web}
\author{Kjetil Kjernsmo\inst{1}}
\institute{Department of Informatics,
Postboks 1080 Blindern,
N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}

\subtitle{---Unpublished Working Draft, do not circulate---}

\begin{document}
\maketitle

\begin{abstract}
\end{abstract}

\section{Introduction}

Caching has been given a prominent place in the foundational documents
of the World Wide Web. Out of the 6 documents that make up the HTTP
1.1 standard, RFC7234 \todo{ref} is entirely devoted to the topic, and
RFC7232 defines conditional requests, which is also important when
constructing caches. As the former notes:
\begin{quote} 
  The goal of caching in HTTP/1.1 is to significantly improve
  performance by reusing a prior response message to satisfy a current
  request.
\end{quote}
Furthermore, the Architecture of the World Wide Web \todo{ref}
discusses caching throughout, and the definition of the
Representational State Transfer (REST) architectural style \todo{ref}
is partly motivated from the requirement to implement efficient
caching. We also note that caching in the Internet infrastructure,
through so-called Content Delivery Networks, is both a large business
area and something that could provide great value to the Semantic Web
community.

In spite of this, we have not seen it in widespread use, and therefore
we decided to conduct a survey to investigate the actual compliance to
RFC7234 and RFC7232. Our reasons to do this include
\begin{enumerate}
\item Understand the actual usage rather than rely on anecdotal
  conceptions.
\item Encourage the implementation of these mechanisms in Semantic Web
  infrastructure.
\item Point out future research directions.
\end{enumerate}

We note that caching is not only useful for long-living information
resources, even though that may be the most important. If a resource
is frequently requested, it may make sense to cache it even though it
may be fresh for only a very short period.

Caching may be deployed at several different levels: A HTTP cache may
be in a reverse proxy close to the server, in which case it may have
much in common with a conventional database cache. It may also be
anywhere between a server and a client, in which case it may be
shared, i.e. it may cache responses from several servers served to
several clients. Another example is an institutional forward proxy,
which are close to several users. Finally, the User Agent may
implement a private cache for its user at the client side.

Caching, as defined in RFC7234, requires that the server declares how
long into the future it expects the information resource in the
response to remain valid. Based on this, the client need not contact
the server at all to reuse a cached response, but this requires a
commitment from the content provider. The standard also gives the
client an opportunity to heuristically determine a time to
live. RFC7232, on the other hand, defines a protocol for asking the
server if the cached response is still fresh using conditional
requests. This doesn't burden the content provider with the task of
estimating the time to live beforehand, but then, it must be able to
answer if the resource has changed less expensively than serving the
entire response. These two approaches can be combined, as a client may
ask if a response is fresh after it has expired.

\end{document}

### 2014-12-16 sanity/paper1.tex
some related work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@901 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\section{Related work}
+
+We are not aware of any surveys of this type. Although the database
+literature is rich with query cache literature, it is mostly relevant
+to what would happen within the server or between the server and a
+reverse proxy, which is opaque to the Internet, and therefore not of
+our concern.
+
+LDF\todo{ref paper1} claims to take advantage of caching and contrasts
+this with the inavailabilty of SPARQL query caches, and claims this is
+an architectural problem.
+
+The implications of HTTP for query caches has been discussed by
+\todo{ref kasei}. In \todo{marin, Auer} the authors implemented a
+reverse proxy that controlled the changes to the dataset, and
+therefore could
+
+A general caching approach is given in \todo{Caching intermediate
+  result of sparql queries}.
+
+\todo{fast vs. fresh} didn't consider caching in the sense we do, they
+rather prefetch an entire dataset to a local store and based on
+heuristics tried to determine which parts of the query should be
+evaluated remotely and locally.
+
+\todo{cache or not} explored when caching had a positive effect on
+complex SPARQL queries.
+
+In the broader Web literature, \todo{Revisiting Cacheability
+in Times of User Generated Content} analysed the value of caching
+based anonymized traces of actual Web usage at a major Internet
+Service Provider.
+
 \end{document}

### 2014-12-16 sanity/paper1.tex
some on methodology
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@902 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 long into the future it expects the information resource in the
-response to remain valid. Based on this, the client need not contact
-the server at all to reuse a cached response, but this requires a
-commitment from the content provider. The standard also gives the
-client an opportunity to heuristically determine a time to
-live. RFC7232, on the other hand, defines a protocol for asking the
-server if the cached response is still fresh using conditional
-requests. This doesn't burden the content provider with the task of
-estimating the time to live beforehand, but then, it must be able to
-answer if the resource has changed less expensively than serving the
-entire response. These two approaches can be combined, as a client may
-ask if a response is fresh after it has expired.
+response to remain fresh, i.e. how long it is permissible to use
+it. Based on this, the client need not contact the server at all to
+reuse a cached response, but this requires a commitment from the
+content provider. The standard also gives the client an opportunity to
+heuristically determine a time to live. RFC7232, on the other hand,
+defines a protocol for asking the server if the cached response is
+still fresh using conditional requests. This doesn't burden the
+content provider with the task of estimating the time to live
+beforehand, but then, it must be able to answer if the resource has
+changed less expensively than serving the entire response. These two
+approaches can be combined, as a client may ask if a response is fresh
+after it has expired.
 
@@
 
+\section{Methodology}
+
+We want to find information resources on the Web, and examine HTTP
+headers that may allow caching. We record headers recommended by
+current standards, as well as obsoleted and non-standard
+headers. Additionally, we examine the triples in the returned
+information resources to see if there is information there that may be
+used to calculate heuristic freshness.
+
+However, as the community has both spidered the Web as well as
+recorded lists of relevant resources, we need not do that ourselves.
+
+The SPARQLES survey\todo{ref} has recorded the availability of SPARQL
+endpoints registered in the datahub.io portal, and found that many of
+them have ceased to operate. We used their list of SPARQL endpoints as
+of 2014-11-17, and filtered out those deemed unresponsive by
+them. This resulted in a list of 312 endpoints.
+
+To compile a list of Web ontologies, we first queried the Linked Open
+Vocabularies \todo{ref} SPARQL endpoint to retrieve all registered
+ontologies. Then, we took the full list of URIs registered with
+prefix.cc. The assumption that they all identify an ontology is
+clearly false, as we found that for example Tim Berners-Lee's FOAF
+profile has a registered URI, but it is unlikely to introduce
+significant bias.
+
+Then, we retrieved the dataset of the Billion Triple Challenge
+2014\todo{ref}, a 4~GTriple corpus of spidered Web data. To compile a
+list of interesting candidates to further examine, we performed a
+series of data reduction steps. Due to the presence of invalid RDF, we
+iterated through the NQuad files on a line-by-line basis. First, we
+matched each line against a regular expression were lines matching
+\texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
+the Perl framework \text{RDF::Trine}\todo{footnote} was used to parse
+the line. Lines that failed to parse was discarded. We have not
+investigated whether this could introduce biases. Statements were then
+accepted into a new NQuad file if they had a predicate that matched
+the \texttt{sd:endpoint}\todo{footnote} or matched a case-insensitive
+regular expression \texttt{sparql} if the subject and objects were
+both resources, or the predicates \texttt{void:vocabulary},
+\texttt{rdfa:vocabulary} or \texttt{api:vocabulary} \todo{footnotes
+  for all}, as well as having an resource as object. Finally,
+statements with the the classes \texttt{cogs:Endpoint},
+\texttt{owl:Ontology} and \texttt{voaf:Vocabulary} \todo{footnotes for
+  all} in the object position were also accepted. More classes and
+properties were considered, but not used in the data reduction if they
+did not occur in the original data.
+
+In the next step, we filtered out statements with URIs that were
+invalid or irrelevant, e.g. URIs that didn't have a scheme or where
+the scheme wasn't HTTP(S), or they were referring to private IP addresses.
+
 \end{document}

### 2014-12-17 sanity/paper1.tex
describe hitlist headers operation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@903 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{wrapfig}
+
+\newcommand{\rdfterm}[1]{\texttt{#1}}
+\newcommand{\httph}[1]{\texttt{#1}}
+\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 
+LDF\todo{ref} asserted that SPARQL query caching is only possible on a
+per-query basis, but future research should challenge this
+assumption. With this in mind, if reasoning is not involved, we note
+that ontologies are not relevant to query answering since they usually
+occur in bound terms. It is interesting to understand caching when
+this is not the case. Moreover, many ontologies are maintained by a
+small group of humans and change seldomly.
+
 \section{Related work}
@@
 list of interesting candidates to further examine, we performed a
-series of data reduction steps. Due to the presence of invalid RDF, we
+series of data reduction steps, manually inspecting the result between
+each step. Due to the presence of invalid RDF, we
 iterated through the NQuad files on a line-by-line basis. First, we
@@
 accepted into a new NQuad file if they had a predicate that matched
-the \texttt{sd:endpoint}\todo{footnote} or matched a case-insensitive
+the \rdfterm{sd:endpoint}\todo{footnote} or matched a case-insensitive
 regular expression \texttt{sparql} if the subject and objects were
-both resources, or the predicates \texttt{void:vocabulary},
-\texttt{rdfa:vocabulary} or \texttt{api:vocabulary} \todo{footnotes
+both resources, or the predicates \rdfterm{void:vocabulary},
+\rdfterm{rdfa:vocabulary} or \rdfterm{api:vocabulary} \todo{footnotes
   for all}, as well as having an resource as object. Finally,
-statements with the the classes \texttt{cogs:Endpoint},
-\texttt{owl:Ontology} and \texttt{voaf:Vocabulary} \todo{footnotes for
+statements with the the classes \rdfterm{cogs:Endpoint},
+\rdfterm{owl:Ontology} and \rdfterm{voaf:Vocabulary} \todo{footnotes for
   all} in the object position were also accepted. More classes and
@@
 
+We then sought to classify resources into the categories ``endpoint''
+for SPARQL endpoints, ``vocabulary'' for ontologies, ``dataset'' for
+datasets that may contain further descriptions of several resources,
+or simply ``inforesources'' for those that did not fit in the above
+classes. To do so, we classified based on certain predicates and
+classes, see table\todo{create table}. Additionally, URIs derived from
+prefix.cc were classified as ``vocabulary'' and those from SPARQLES as
+``endpoint''.
+
+Since we blatantly violated URI opacity with our regular expression
+matching in the first step, we needed to further filter candidates for
+SPARQL endpoints. This step therefore included filtering as well as
+classification.
+
+We found in the data a large number of ontologies that consist of many
+information resources with just a few triples in each. Since they
+appear to be produced by the same software, usually Semantic
+Mediawiki, we assume that they are configured with a single setup, and
+thus we will merely sample these resources.
+
+We continued to also sample the HTTP headers gathered in the BTC2014
+dataset. 
+%as they had recorded some of the relevant headers,
+%\httph{Expires}, \httph{Last-Modified} and \httph{ETag}. 
+First, we traversed the files with a simple UNIX \texttt{grep} to find
+the resources that had reported one of the RDF serializations as
+content type. We then traversed this list, first discarded the
+resources that did not have a valid IRI (this amounted to just 3273
+resources). For the resources we found, as well as for the resources
+that was of \rdfterm{rdf:type owl:Ontology} above, we kept one
+resource per hostname, with the exception of the popular blogging
+platforms Livejournal and SAPO, where each blog has their own host and
+they expose FOAF data. For those, we only kept one hostname.
+
 \end{document}

### 2014-12-17 sanity/paper1.tex
Description of the fetching process fairly complete
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@904 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\section{Methodology}
+\section{Implementation}
 
@@
 
+We then developed a fairly elaborate paralell spider to get the
+resources found on 3117 unique hosts in the previous steps using the
+Perl frameworks RDF::Trine and libwww. The spider operated with a
+timeout of 20 seconds and a maximum message size it would accept of 1
+MB.
+
+The paralell spider would then launch a process per host, but each
+request to one host would be delayed by 10 seconds. For each host, the
+spider would go through the list of URIs. Since the BTC2014 recorded
+the \httph{Expires} (an RFC7234 caching header), \httph{Last-Modified}
+and \httph{ETag} (RFC7232 conditional request) headers where they
+existed, we first examined if any of the resources were still fresh,
+but none were. Whereever the last two headers existed, we added the
+corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
+headers for a conditional initial request.
+
+For endpoints, we made the following SPARQL query:
+\begin{verbatim}
+SELECT REDUCED ?Concept WHERE {[] a ?Concept} LIMIT 2
+\end{verbatim}
+which should be quite light, yet likely yield results.
+
+Then, the first request would be made, and a selection of the
+resulting HTTP headers recorded in an per-host NQuads file. For this
+purpose, we developed and released a module
+RDF::Generator::HTTP\todo{footnote} to the Comprehensive Perl Archive
+Network. We recorded whether the conditional request showed that the
+BTC data were still fresh, and if it was, we retrieved the current
+data, as we had not couple the headers to the body in our original
+retrieval.
+
+Based on the resulting headers, we first let libwww calculate the
+freshness lifetime in seconds if it was possible, and recorded
+that. Then, we also let it calculate a \emph{heuristic} freshness
+time based on a suggestion in RFC7234, section~4.2.2, and record that.
+
+If the initial response had RFC7232 headers, we make another
+request to see if the server just included the headers but does not
+actually support conditional requests. The heuristic we employed is
+that if the headers remain the same, but the result was returned,
+rather than just a response code \httph{304} (which indicates that the
+previous result can be reused), then the server does most probably not
+support it.
+
+For endpoints, we examine the response message, to see if there are
+any results to our query, and record that if there are.
+
+For all others, we parsed the response, and recorded any errors if the
+parser concluded the content were invalid. We noted that the Dublin Core
+Metadata terms vocabulary has a number of predicates that may become
+useful in determining heuristic freshness in the future, so we
+recorded any statements containing the predicates \rdfterm{dct:date},
+\rdfterm{dct:accrualPeriodicity}, \rdfterm{dct:created},
+\rdfterm{dct:issued}, \rdfterm{dct:modified} or \rdfterm{dct:valid}.
+
+For resource types other than ``vocabulary'', we look for SPARQL
+endpoints in the response, using the predicates \rdfterm{sd:endpoint}
+and \rdfterm{void:sparqlEndpoint}. We then do the same query as above
+and record the relevant headers.
+
+Finally, if the LOV SPARQL endpoint used a URI for the vocabulary that
+was different from the namespace URI (after a normalization step),
+another request would be made to record the selected HTTP headers from
+that as well.
+
 \end{document}

### 2014-12-17 sanity/paper1.tex
smaller changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@905 43ef1190-90fd-4dd7-a769-d0121f02b80e
spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@906 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-\documentclass{llncs}
-%\documentclass{article}
+%\documentclass{llncs}
+\documentclass{article}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{graphicx}
-\usepackage{wrapfig}
+%\usepackage{wrapfig}
 
@@
 \title{A survey of HTTP caching options on the open Web}
-\author{Kjetil Kjernsmo\inst{1}}
-\institute{Department of Informatics,
-Postboks 1080 Blindern,
-N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}
+\author{Kjetil Kjernsmo}%\inst{1}}
+%\institute{Department of Informatics,
+%Postboks 1080 Blindern,
+%N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}
 
-\subtitle{---Unpublished Working Draft, do not circulate---}
+%\subtitle{---Unpublished Working Draft, do not circulate---}
 
@@
 LDF\todo{ref paper1} claims to take advantage of caching and contrasts
-this with the inavailabilty of SPARQL query caches, and claims this is
+this with the unavailability of SPARQL query caches, and claims this is
 an architectural problem.
@@
 \texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
-the Perl framework \text{RDF::Trine}\todo{footnote} was used to parse
+the Perl framework RDF::Trine\todo{footnote} was used to parse
 the line. Lines that failed to parse was discarded. We have not
@@
 
-We then developed a fairly elaborate paralell spider to get the
+We then developed a fairly elaborate parallel spider to get the
 resources found on 3117 unique hosts in the previous steps using the
@@
 
-The paralell spider would then launch a process per host, but each
+The parallel spider would then launch a process per host, but each
 request to one host would be delayed by 10 seconds. For each host, the
-spider would go through the list of URIs. Since the BTC2014 recorded
+spider would go through the list of URLs found by previous steps for
+that host. Since the BTC2014 recorded
 the \httph{Expires} (an RFC7234 caching header), \httph{Last-Modified}
@@
 existed, we first examined if any of the resources were still fresh,
-but none were. Whereever the last two headers existed, we added the
+but none were. Wherever the last two headers existed, we added the
 corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
 headers for a conditional initial request.
+
+\begin{table}
+\caption{Recorded HTTP headers}
+ \begin{tabular}{ | l |  p{3cm} | p{5cm} |}
+    \hline
+    Header & Reference & Description \\ \hline
+\httph{Age} & RFC7234 & When obtaining response from a cache, the number of
+seconds since validation \\ \hline
+\httph{Cache-Control} & RFC7234 & Header used for a variety of directives \\ \hline
+\httph{Expires} & RFC7234  & Gives the date/time after which the
+   response is considered stale. \\ \hline
+\httph{Pragma} & RFC7234 & Archaic HTTP 1.0 header  \\ \hline
+\httph{Warning} & RFC7234  & For additional information about possible incorrectness \\ \hline
+\httph{Content-Type} & RFC7231 & To select the correct parser \\ \hline
+\httph{If-None-Match} & RFC7232  & Request header to check if
+                                   \httph{ETag} has changed   \\ \hline
+\httph{If-Modified-Since} & RFC7232  & Request header to check if
+                                   \httph{Last-Modified} has changed    \\ \hline
+\httph{Last-Modified} & RFC7232 & When the resource was last modified \\ \hline
+\httph{ETag} & RFC7232 & An opaque validator to check if the resource
+has changed  \\ \hline
+\httph{X-Cache} &  & Inserted by some caches to indicate cache status \\ \hline
+\httph{Date} & RFC7231 & The time of the message. Used in conditional requests
+and heuristics \\ \hline
+\httph{Surrogates} & Edge Architecture Specification\todo{ref} & Draft
+to allow more fine-grained control for proxies. \\ \hline
+\httph{Client-Aborted} & libwww\todo{ref}  & Header inserted by User
+Agent to indicate that it aborted the download \\ \hline
+\httph{Client-Warning} & libwww\todo{ref}  & Header inserted by User
+Agent to give details about problems with the download \\ \hline
+
+    \hline
+    \end{tabular}
+\end{table}
 
@@
 that. Then, we also let it calculate a \emph{heuristic} freshness
-time based on a suggestion in RFC7234, section~4.2.2, and record that.
+time based on a suggestion in RFC7234, section~4.2.2, and recorded that.
 
@@
 
-For endpoints, we examine the response message, to see if there are
-any results to our query, and record that if there are.
+For endpoints, we examined the response message, to see if there are
+any results to our query, and recorded that if there are.
 

### 2014-12-17 sanity/paper1.tex
small change
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@907 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In spite of this, we have not seen it in widespread use, and therefore
-we decided to conduct a survey to investigate the actual compliance to
-RFC7234 and RFC7232. Our reasons to do this include
+In spite of this, we have not seen it in widespread use in the
+Semantic Web, and therefore we decided to conduct a survey to
+investigate the actual compliance to RFC7234 and RFC7232. Our reasons
+to do this include
 \begin{enumerate}

### 2014-12-17 sanity/fetcher.pl
Fix bug that didn't run conditional
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@908 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -179,23 +179,28 @@ foreach my $host (@hosts) {
 
 	  my $prevresponse = $firstresponse;
 
-	  if ($firstresponse->is_success) {
-
-		  # What to do if data for conditional requests was available in BTC
-		  if ($firstresponse->code == 304) {
-			  my $condrequest = HTTP::Request->new(GET => $uri);
-			  sleep 10;
-			  my $condresponse = $ua->request( $condrequest );
-			  my $condhhg = RDF::Generator::HTTP->new(message => $condresponse,
-																	whitelist => ['Warning',
-																					  'Last-Modified',
-																					  'ETag',
-																					  'Date'],
-																	graph => $context);
-			  $condhhg->generate($model);
-			  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context));
-			  $prevresponse = $condresponse;
+	  # What to do if data for conditional requests was available in BTC
+	  if ($firstresponse->code == 304) {
+		  my $condrequest = HTTP::Request->new(GET => $uri);
+		  sleep 5;
+		  my $condresponse = $ua->request( $condrequest );
+		  my $condhhg = RDF::Generator::HTTP->new(message => $condresponse,
+																whitelist => ['Warning',
+																				  'Last-Modified',
+																				  'ETag',
+																				  'Date'],
+																graph => $context);
+		  $condhhg->generate($model);
+		  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context));
+		  if ($condrequest->is_success) {
+			  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
+		  } else {
+			  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('Second conditional failed'), $context));
 		  }
+
+		  $prevresponse = $condresponse;
+
+	  } elsif ($firstresponse->is_success) {
 
 		  # Add freshness triples
 		  if ($prevresponse->freshness_lifetime(heuristic_expiry => 0)) {
@@ -323,10 +328,8 @@ foreach my $host (@hosts) {
 			  }
 		  }
 	  } else {
-		  unless ($firstresponse->code == 304) {
-			  $model->add_statement(statement(iri($uri), iri('http://www.w3.org/2007/ont/http#status_line'), literal($prevresponse->status_line), $context));
[…]

### 2014-12-17 sanity/fetcher.pl
Fix minor bug
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@909 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -192,7 +192,7 @@ foreach my $host (@hosts) {
 																graph => $context);
 		  $condhhg->generate($model);
 		  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $condhhg->request_subject, $context));
-		  if ($condrequest->is_success) {
+		  if ($condresponse->is_success) {
 			  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
 		  } else {
 			  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('Second conditional failed'), $context));

### 2014-12-18 sanity/paper1.tex
Move details into appendix
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@911 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Then, we retrieved the dataset of the Billion Triple Challenge
-2014\todo{ref}, a 4~GTriple corpus of spidered Web data. To compile a
+To examine as many different implementations and hosts as possible, we
+noted that the Billion Triple Challenge
+2014\todo{ref} dataset consisted of a 4~GTriple corpus of spidered Web data. To compile a
 list of interesting candidates to further examine, we performed a
 series of data reduction steps, manually inspecting the result between
-each step. Due to the presence of invalid RDF, we
-iterated through the NQuad files on a line-by-line basis. First, we
-matched each line against a regular expression were lines matching
+each step. The details of this process is given in Appendix~\ref{app:reduction}.
+
+\section{Implementation details\todo{proper formatting}}\label{app:reduction}
+
+The Billion Triple Challenge 2014 dataset provided data in the form of
+NQuad files. Due to the presence of invalid RDF, we iterated through
+the NQuad files on a line-by-line basis. First, we matched each line
+against a regular expression were lines matching
 \texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
-the Perl framework RDF::Trine\todo{footnote} was used to parse
-the line. Lines that failed to parse was discarded. We have not
+the Perl framework RDF::Trine\todo{footnote} was used to parse the
+line. Lines that failed to parse was discarded. We have not
 investigated whether this could introduce biases. Statements were then
@@
 statements with the the classes \rdfterm{cogs:Endpoint},
-\rdfterm{owl:Ontology} and \rdfterm{voaf:Vocabulary} \todo{footnotes for
-  all} in the object position were also accepted. More classes and
+\rdfterm{owl:Ontology} and \rdfterm{voaf:Vocabulary} \todo{footnotes
+  for all} in the object position were also accepted. More classes and
 properties were considered, but not used in the data reduction if they

### 2014-12-18 sanity/fetcher.pl

@@ -157,7 +157,7 @@ foreach my $host (@hosts) {
 	  my $firstresponse = $ua->request( $request );
 		  # Get the relevant headers
 	  my $hhg = RDF::Generator::HTTP->new(message => $firstresponse,
-													  whitelist => ['Age',
+													  whitelist => ['Age', # TODO : Server
 																		 'Cache-Control',
 																		 'Expires',
 																		 'Pragma',

### 2014-12-18 sanity/paper1.tex
fix appendices more reorg
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@912 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 %\usepackage{wrapfig}
+\usepackage{appendix}
 
@@
 
-\section{Implementation}
+\section{Methodology}
 
@@
 To examine as many different implementations and hosts as possible, we
-noted that the Billion Triple Challenge
-2014\todo{ref} dataset consisted of a 4~GTriple corpus of spidered Web data. To compile a
+noted that the Billion Triple Challenge 2014\todo{ref} dataset
+consisted of a 4~GTriple corpus of spidered Web data. To compile a
 list of interesting candidates to further examine, we performed a
 series of data reduction steps, manually inspecting the result between
-each step. The details of this process is given in Appendix~\ref{app:reduction}.
+each step. The details of this process is given in
+Appendix~\ref{app:reduction}.
 
-\section{Implementation details\todo{proper formatting}}\label{app:reduction}
+The end result of this process is a list of 3117 unique hosts, for
+each several resources would be visited, some several times, as they
+may host SPARQL Endpoints, ontologies, or other information resources,
+by a spider detailed in Appendix~\ref{app:fetcher}
+
+This results in an NQuads file per host, which is then loaded into a
+SPARQL Endpoint\todo{which} for analysis by using the statistics
+system R.
+
+\begin{appendices}
+
+\section{Implementation of data reduction}\label{app:reduction}
 
@@
 
-We then developed a fairly elaborate parallel spider to get the
-resources found on 3117 unique hosts in the previous steps using the
-Perl frameworks RDF::Trine and libwww. The spider operated with a
-timeout of 20 seconds and a maximum message size it would accept of 1
-MB.
+\section{Implementation of spider}\label{app:fetcher}
+
+We then developed a fairly elaborate parallel spider to examine the
+resources found on hosts that the previous steps deemed interesting
+using the Perl frameworks RDF::Trine and libwww. The spider operated
+with a timeout of 20 seconds and a maximum message size it would
+accept of 1 MB.
 
@@
 that as well.
+\end{appendices}
 

### 2014-12-18 sanity/fetcher.pl
Add Server header
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@914 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -157,7 +157,8 @@ foreach my $host (@hosts) {
 	  my $firstresponse = $ua->request( $request );
 		  # Get the relevant headers
 	  my $hhg = RDF::Generator::HTTP->new(message => $firstresponse,
-													  whitelist => ['Age', # TODO : Server
+													  whitelist => ['Age',
+																		 'Server',
 																		 'Cache-Control',
 																		 'Expires',
 																		 'Pragma',
@@ -186,6 +187,7 @@ foreach my $host (@hosts) {
 		  my $condresponse = $ua->request( $condrequest );
 		  my $condhhg = RDF::Generator::HTTP->new(message => $condresponse,
 																whitelist => ['Warning',
+																				  'Server',
 																				  'Last-Modified',
 																				  'ETag',
 																				  'Date'],
@@ -230,6 +232,7 @@ foreach my $host (@hosts) {
 			  my $cond2response = $ua->request( $cond2request );
 			  my $cond2hhg = RDF::Generator::HTTP->new(message => $cond2response,
 																	 whitelist => ['Warning',
+																						'Server',
 																						'If-None-Match',
 																						'If-Modified-Since',
 																						'Last-Modified',
@@ -296,6 +299,7 @@ foreach my $host (@hosts) {
 							  # Get the relevant headers
 							  my $ehhg = RDF::Generator::HTTP->new(message => $eresponse,
 																				whitelist => ['Age',
+																								  'Server',
 																								  'Cache-Control',
 																								  'Expires',
 																								  'Pragma',
@@ -335,6 +339,7 @@ foreach my $host (@hosts) {
 			  # Get the relevant headers
 			  my $ahhg = RDF::Generator::HTTP->new(message => $aresponse,
 																whitelist => ['Age',
+																				  'Server',
 																				  'Cache-Control',
 																				  'Expires',
 																				  'Pragma',

### 2014-12-19 sanity/fetcher.pl
Track which request is answered
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@915 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -155,6 +155,7 @@ foreach my $host (@hosts) {
 		  }
 	  }
 	  my $firstresponse = $ua->request( $request );
+	  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('firstresponse'), $context));
 		  # Get the relevant headers
 	  my $hhg = RDF::Generator::HTTP->new(message => $firstresponse,
 													  whitelist => ['Age',
@@ -185,6 +186,7 @@ foreach my $host (@hosts) {
 		  my $condrequest = HTTP::Request->new(GET => $uri);
 		  sleep 5;
 		  my $condresponse = $ua->request( $condrequest );
+		  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('condresponse'), $context));
 		  my $condhhg = RDF::Generator::HTTP->new(message => $condresponse,
 																whitelist => ['Warning',
 																				  'Server',
@@ -230,6 +232,7 @@ foreach my $host (@hosts) {
 			  no warnings 'uninitialized';
 			  sleep 10;
 			  my $cond2response = $ua->request( $cond2request );
+			  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('cond2response'), $context));
 			  my $cond2hhg = RDF::Generator::HTTP->new(message => $cond2response,
 																	 whitelist => ['Warning',
 																						'Server',
@@ -296,6 +299,7 @@ foreach my $host (@hosts) {
 							  my $erequest = HTTP::Request->new(GET => $uri);
 							  $erequest->header( Accept => 'application/sparql-results+xml,application/sparql-results+json;q=0.9' );
 							  my $eresponse = $ua->request( $erequest );
+							  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('eresponse'), $context));
 							  # Get the relevant headers
 							  my $ehhg = RDF::Generator::HTTP->new(message => $eresponse,
 																				whitelist => ['Age',
@@ -336,6 +340,7 @@ foreach my $host (@hosts) {
 		  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('Unsuccessful response'), $context));
 		  if ($details->{alternate}) {
 			  my $aresponse = $ua->get($details->{alternate}, Accept => $accept_header);
+			  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('aresponse'), $context));
[…]

### 2015-01-02 sanity/fetcher.pl
Move status messages around
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@920 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -265,6 +265,7 @@ foreach my $host (@hosts) {
 				  my $parser = $parsertype->new;
 				  # Then it is likely we get RDF
 				  my $size = 0;
+				  my $parseerror;
 
 				  try {
 					  no warnings;
@@ -288,6 +289,7 @@ foreach my $host (@hosts) {
 										  });
 				  } catch {
 					  $model->add_statement(statement(iri($uri), iri('urn:app:parseerror'), literal($_), $context));
+					  $parseerror = 1;
 				  };
 
 				  unless ($details->{type} eq 'vocabulary') {
@@ -327,10 +329,12 @@ foreach my $host (@hosts) {
 						  }
 					  }
 				  }
-				  if ($promise) {
-					  $model->add_statement(statement(iri($uri), iri('urn:app:promising'), literal($promise), $context));
+
+				  if ($parseerror) {
+					  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('parseerror'), $context));
+				  } else {
+					  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
 				  }
-				  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
 			  } else {
 				  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('Invalid media type'), $context));
 			  }
@@ -362,6 +366,9 @@ foreach my $host (@hosts) {
 			  $ahhg->generate($model);
 			  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $ahhg->request_subject, $context));
 		  }
+	  }
+	  if ($promise) {
+		  $model->add_statement(statement(iri($uri), iri('urn:app:promising'), literal($promise), $context));
 	  }
 	  sleep 10 if ($uricount > 1);
   }

### 2015-01-02 sanity/fetcher.pl
Do expiry for follow-up requests
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@921 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -325,6 +325,15 @@ foreach my $host (@hosts) {
 							  if ($eresponse->is_success) {
 								  my $anyres = has_sparql_results($eresponse->decoded_content, $eresponse->content_type) ? "Has results" : "No results";
 								  $model->add_statement(statement(iri($uri), iri('urn:app:endpoint'), literal($anyres), iri($endpoint)));
+								  # Add freshness triples
+								  if ($eresponse->freshness_lifetime(heuristic_expiry => 0)) {
+									  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:hard'), literal($eresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
+									  $promise = 'hard';
+								  } elsif ($eresponse->headers->last_modified) {
+									  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:heuristic'), literal($eresponse->freshness_lifetime(h_min => 1, h_max => 31536001, h_default =>0)), $context));
+									  $promise = 'heuristic';
+		  }
+
 							  }
 						  }
 					  }
@@ -365,6 +374,15 @@ foreach my $host (@hosts) {
 																graph => $context);
 			  $ahhg->generate($model);
 			  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $ahhg->request_subject, $context));
+			  # Add freshness triples
+			  if ($aresponse->freshness_lifetime(heuristic_expiry => 0)) {
+				  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:hard'), literal($aresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
+				  $promise = 'hard';
+			  } elsif ($aresponse->headers->last_modified) {
+				  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:heuristic'), literal($aresponse->freshness_lifetime(h_min => 1, h_max => 31536001, h_default =>0)), $context));
+				  $promise = 'heuristic';
+		  }
+
 		  }
 	  }
 	  if ($promise) {

### 2015-01-02 sanity/fetcher.pl
Ready for run 7
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@922 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,6 +1,7 @@
 #!/usr/bin/perl
 
 # Rev 877 is first attempt to prod
+# Rev 922 is analysis run
 use strict;
 use warnings;
 use Data::Dumper;

### 2015-01-02 sanity/fetcher.pl
Say Ok to alternates
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@923 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -93,7 +93,7 @@ $prparse->finish;
 $prfetch->update(message => "Initializing UA");
 
 use Parallel::ForkManager;
-my $pm = Parallel::ForkManager->new(30);
+my $pm = Parallel::ForkManager->new(50);
 
 use LWP::UserAgent;
 use HTTP::Request;
@@ -326,6 +326,8 @@ foreach my $host (@hosts) {
 							  if ($eresponse->is_success) {
 								  my $anyres = has_sparql_results($eresponse->decoded_content, $eresponse->content_type) ? "Has results" : "No results";
 								  $model->add_statement(statement(iri($uri), iri('urn:app:endpoint'), literal($anyres), iri($endpoint)));
+								  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
+
 								  # Add freshness triples
 								  if ($eresponse->freshness_lifetime(heuristic_expiry => 0)) {
 									  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:hard'), literal($eresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
@@ -333,7 +335,7 @@ foreach my $host (@hosts) {
 								  } elsif ($eresponse->headers->last_modified) {
 									  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:heuristic'), literal($eresponse->freshness_lifetime(h_min => 1, h_max => 31536001, h_default =>0)), $context));
 									  $promise = 'heuristic';
-		  }
+								  }
 
 							  }
 						  }
@@ -354,7 +356,8 @@ foreach my $host (@hosts) {
 		  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('Unsuccessful response'), $context));
 		  if ($details->{alternate}) {
 			  my $aresponse = $ua->get($details->{alternate}, Accept => $accept_header);
-			  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('aresponse'), $context));
+			  my $alturi = iri($details->{alternate});
+			  $model->add_statement(statement($alturi, iri('urn:app:whichrequest'), literal('aresponse'), $context));
 			  # Get the relevant headers
 			  my $ahhg = RDF::Generator::HTTP->new(message => $aresponse,
 																whitelist => ['Age',
@@ -374,16 +377,21 @@ foreach my $host (@hosts) {
 																				 ],
 																graph => $context);
[…]

### 2015-01-02 sanity/fetcher.pl
Remove block checking extra endpoints; returned garbage anyway
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@924 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -293,55 +293,6 @@ foreach my $host (@hosts) {
 					  $parseerror = 1;
 				  };
 
-				  unless ($details->{type} eq 'vocabulary') {
-					  # Look for endpoints, if so, do as in endpoint
-					  foreach my $endpoint (@endpoints) {
-						  my $euri = URI->new($endpoint);
-						  unless ($data->{$euri->host}->{$endpoint}) {
-							  $uri = $endpoint . '?query=' . uri_escape('select reduced ?Concept where {[] a ?Concept} LIMIT 2');	  
-							  my $erequest = HTTP::Request->new(GET => $uri);
-							  $erequest->header( Accept => 'application/sparql-results+xml,application/sparql-results+json;q=0.9' );
-							  my $eresponse = $ua->request( $erequest );
-							  $model->add_statement(statement(iri($uri), iri('urn:app:whichrequest'), literal('eresponse'), $context));
-							  # Get the relevant headers
-							  my $ehhg = RDF::Generator::HTTP->new(message => $eresponse,
-																				whitelist => ['Age',
-																								  'Server',
-																								  'Cache-Control',
-																								  'Expires',
-																								  'Pragma',
-																								  'Warning',
-																								  'Content-Type',
-																								  'Last-Modified',
-																								  'ETag',
-																								  'X-Cache',
-																								  'Date',
-																								  'Surrogates',
-																								  'Client-Aborted',
-																								  'Client-Warning'
-																								 ],
-																				graph => iri($endpoint));
-							  $ehhg->generate($model);
-							  $model->add_statement(statement(iri($uri), iri('urn:app:hasrequest'), $ehhg->request_subject, $context));
-							  if ($eresponse->is_success) {
-								  my $anyres = has_sparql_results($eresponse->decoded_content, $eresponse->content_type) ? "Has results" : "No results";
-								  $model->add_statement(statement(iri($uri), iri('urn:app:endpoint'), literal($anyres), iri($endpoint)));
-								  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
-
-								  # Add freshness triples
-								  if ($eresponse->freshness_lifetime(heuristic_expiry => 0)) {
-									  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:hard'), literal($eresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
-									  $promise = 'hard';
[…]

### 2015-01-02 sanity/paper1.tex
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@913 43ef1190-90fd-4dd7-a769-d0121f02b80e
comment on the removal of eresponse
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@925 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 SPARQL Endpoint\todo{which} for analysis by using the statistics
-system R.
+system R\todo{ref}.
 
@@
 and \rdfterm{void:sparqlEndpoint}. We then do the same query as above
-and record the relevant headers.
+and record the relevant headers. Unfortunately, we found early that
+this only turns up misconfigured endpoints that point to localhost,
+and was removed from the spider for the final analysis.
 

### 2015-01-02 sanity/fetcher.pl
Use DISTINCT in the query since VU has a problem
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@926 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -143,7 +143,7 @@ foreach my $host (@hosts) {
 
 	  my $request = HTTP::Request->new(GET => $uri);
 	  if ($details->{type} eq 'endpoint') {
-		  $uri .= '?query=' . uri_escape('select reduced ?Concept where {[] a ?Concept} LIMIT 2');
+		  $uri .= '?query=' . uri_escape('SELECT DISTINCT ?Concept WHERE {[] a ?Concept} LIMIT 2');
 		  $request->uri( $uri );
 		  $request->header( Accept => 'application/sparql-results+xml,application/sparql-results+json;q=0.9' );
 	  } else {

### 2015-01-02 sanity/fetcher.pl
Have a successful status for endpoints
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@927 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -259,6 +259,7 @@ foreach my $host (@hosts) {
 		  if ($details->{type} eq 'endpoint') {
 			  # Check if we got any results
 			  my $anyres = has_sparql_results($content, $prevresponse->content_type) ? "Has results" : "No results";
+			  $model->add_statement(statement(iri($uri), iri('urn:app:status'), literal('OK'), $context));
 			  $model->add_statement(statement(iri($uri), iri('urn:app:endpoint'), literal($anyres), $context));
 		  } else { # All RDF resources
 			  my $parsertype = RDF::Trine::Parser->parser_by_media_type($prevresponse->content_type);

### 2015-01-02 sanity/fetcher.pl
Record more promises
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@928 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -210,10 +210,10 @@ foreach my $host (@hosts) {
 		  # Add freshness triples
 		  if ($prevresponse->freshness_lifetime(heuristic_expiry => 0)) {
 			  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:hard'), literal($prevresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
-			  $promise = 'hard';
+			  $promise .= ' hard';
 		  } elsif ($prevresponse->headers->last_modified) {
 			  $model->add_statement(statement(iri($uri), iri('urn:app:freshtime:heuristic'), literal($prevresponse->freshness_lifetime(h_min => 1, h_max => 31536001, h_default =>0)), $context));
-			  $promise = 'heuristic';
+			  $promise .= ' heuristic';
 		  }
 
 		  my $content = $prevresponse->decoded_content;
@@ -251,7 +251,7 @@ foreach my $host (@hosts) {
 				  # We should have gotten 304
 				  $model->add_statement(statement(iri($uri), iri('urn:app:conditional'), literal("Got all"), $context));
 			  } else {
-				  $promise = 'not-modified';
+				  $promise .= ' not-modified';
 			  }
 		  }
 		  my @endpoints;
@@ -283,7 +283,7 @@ foreach my $host (@hosts) {
 													$st->predicate->equal($dct->modified) ||
 													$st->predicate->equal($dct->valid)) {
 												  $model->add_statement(statement($st->subject, $st->predicate, $st->object, $context));
-												  $promise = 'predicate';
+												  $promise .= ' predicate';
 											  } elsif ($st->predicate->equal(iri('http://www.w3.org/ns/sparql-service-description#endpoint')) ||
 														  $st->predicate->equal(iri(('http://rdfs.org/ns/void#sparqlEndpoint')))) {
 												  push(@endpoints, $st->object->uri_value);
@@ -336,10 +336,10 @@ foreach my $host (@hosts) {
 				  # Add freshness triples
 				  if ($aresponse->freshness_lifetime(heuristic_expiry => 0)) {
 					  $model->add_statement(statement($alturi, iri('urn:app:freshtime:hard'), literal($aresponse->freshness_lifetime(heuristic_expiry => 0)), $context));
-					  $promise = 'hard';
+					  $promise .= ' hard';
 				  } elsif ($aresponse->headers->last_modified) {
[…]

### 2015-01-02 sanity/queries.R
First analysis of Server header with Hard
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@929 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -1,1 +1,5 @@
 checkserver <- sparqlfile("check-server.rq")
+reshardbin <- as.integer(checkserver$results$hard)
+reshardbin[! is.na(reshardbin)] <- 1
+reshardbin[is.na(reshardbin)] <- 0
+chisq.test(checkserver$results$server, reshardbin, simulate.p.value=T, B = 10000)

### 2015-01-02 sanity/queries.R
Test for Server with promise
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@930 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -3,3 +3,8 @@ reshardbin <- as.integer(checkserver$results$hard)
 reshardbin[! is.na(reshardbin)] <- 1
 reshardbin[is.na(reshardbin)] <- 0
 chisq.test(checkserver$results$server, reshardbin, simulate.p.value=T, B = 10000)
+
+respromisebin <- checkserver$results$promise
+respromisebin[! is.na(respromisebin)] <- 1
+respromisebin[is.na(respromisebin)] <- 0
+chisq.test(checkserver$results$server, respromisebin, simulate.p.value=T, B = 10000)

### 2015-01-04 sanity/paper1.tex
some additional todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@931 43ef1190-90fd-4dd7-a769-d0121f02b80e
writeup about servers
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@932 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\rdfterm}[1]{\texttt{#1}}
+\newcommand{\pvalue}{\textit{p}-value}
 \newcommand{\httph}[1]{\texttt{#1}}
@@
 
+Key contributions of this paper are:\todo{write}
+
 \section{Related work}
@@
 system R\todo{ref}.
+
+\section{Analytics}
+
+\subsection{Different server implementations}
+
+First, we investigated whether certain servers provided better support
+for caching than others. To do this, we formulated SPARQL queries to
+examine the \httph{Server} headers of successful responses, with
+optional clauses matching the standard-compliant computed freshness
+time (which is the ideal) as well as whether the response had other
+promising features that could assist caching, such as \httph{ETag}s,
+modification time, certain predicates, etc.
+
+For each unique \httph{Server} header, we found the ones where
+\emph{all} responses had a freshness time or a promising feature. For
+the former, this amounted to 22 servers, tabulated in~\todo{include
+  table}. 70 servers always responded with promising features. As we
+can see, apart from well-known Virtuoso and Callimachus servers, as
+well as the Perl modules RDF::LinkedData and RDF::Endpoint, which is
+partly developed by and runs on a server operated by this author, most
+headers tells very little about the RDF-specific parts of the
+underlying server implementation.
+
+For a more systematic approach, we wish to test the hypothesis that
+some servers are better configured to support caching than others. To
+test this hypothesis, we create two so-called \emph{contigency
+  tables}\todo{ref}. This formalism is suited to see if the
+distribution of  \httph{Server} headers are different for those implementations that
+offer caching headers from those that don't. Intuitively, we expect
+these distributions to be similar, ``long-tail'' distributions, i.e. a
+handful of servers are used by a large number of projects, and then it
+falls off rapidly, and so, some servers are used only by very
+few. Likewise, it is only to be expected that only a few projects have
+given caching enough attention, but that they account for the majority
+of the support. The question is if the presence of caching headers is
+merely a matter of proportion, or if there are some that have given it
+more attention, but still is in relatively little use.
+
+Using the statistics system R\todo{ref}, we use a statistical test,
+namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000
+replicates). The simulation is done using a Monte Carlo method to
+compensate for the fact that many servers will not expose caching
+headers at all.
+
+In both the cases of standard-compliant freshness time and for the
+other promising features, the test reports \pvalue = 0.0001. We can
+conclude that it is highly likely that some servers are better at
+exposing cache headers than others. Unfortunately, since most
+\httph{Server} headers only report generic values, little can be
+learnt about these implementations. We note, however, that DBPedia
+exposes standards-compliant freshness time of 604800 seconds (i.e. 1
+week) for both LOD and SPARQL endpoints. \todo{do this specifically
+  for SPARQLES endpoints?}
+
+\section{Recommendations}\todo{write}
 

### 2015-01-04 sanity/queries.R
Get the proportions of cached
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@933 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -8,3 +8,7 @@ respromisebin <- checkserver$results$promise
 respromisebin[! is.na(respromisebin)] <- 1
 respromisebin[is.na(respromisebin)] <- 0
 chisq.test(checkserver$results$server, respromisebin, simulate.p.value=T, B = 10000)
+
+checkservercount <- sparqlfile("check-server-count.rq")
+serverratio <- data.frame(name=checkservercount$results$server, promise=checkservercount$results$pc/checkservercount$results$sc, hard=checkservercount$results$hc/checkservercount$results$sc, pc=checkservercount$results$pc, hc=checkservercount$results$hc)
+xtable(serverratio[which(serverratio$hard == 1), ]$name)

### 2015-01-05 sanity/paper1.tex
few details
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@934 43ef1190-90fd-4dd7-a769-d0121f02b80e
stub further work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@935 43ef1190-90fd-4dd7-a769-d0121f02b80e
minor tweaks
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@936 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 headers tells very little about the RDF-specific parts of the
-underlying server implementation.
+underlying server implementation. A quick inspection of all
+\httph{Server} headers confirm that few reveal any further detail.
 
@@
 compensate for the fact that many servers will not expose caching
-headers at all.
+headers at all, an issue that would otherwise violate the underlying
+assumptions of the test.
 
@@
   for SPARQLES endpoints?}
+
+\section{Future work}
+
+Identify implementations that expose suitable caching headers.
 
@@
 \begin{verbatim}
-SELECT REDUCED ?Concept WHERE {[] a ?Concept} LIMIT 2
+SELECT DISTINCT ?Concept WHERE {[] a ?Concept} LIMIT 2
 \end{verbatim}
@@
 For endpoints, we examined the response message, to see if there are
-any results to our query, and recorded that if there are.
+any results to our query, and recorded that if there are. In addition
+to the endpoints registered in the SPARQLES survey, our process found
+18~endpoints that responded with results.
 

### 2015-01-05 sanity/paper1.tex
Add server table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@938 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \emph{all} responses had a freshness time or a promising feature. For
-the former, this amounted to 22 servers, tabulated in~\todo{include
-  table}. 70 servers always responded with promising features. As we
-can see, apart from well-known Virtuoso and Callimachus servers, as
-well as the Perl modules RDF::LinkedData and RDF::Endpoint, which is
-partly developed by and runs on a server operated by this author, most
-headers tells very little about the RDF-specific parts of the
-underlying server implementation. A quick inspection of all
-\httph{Server} headers confirm that few reveal any further detail.
+the former, this amounted to 22 servers, tabulated
+in~\ref{tab:servers}. 70 servers always responded with promising
+features. As we can see, apart from well-known Virtuoso and
+Callimachus servers, as well as the Perl modules RDF::LinkedData and
+RDF::Endpoint, which is partly developed by and runs on a server
+operated by this author, most headers tells very little about the
+RDF-specific parts of the underlying server implementation. A quick
+inspection of all \httph{Server} headers confirmed that few reveal any
+further detail.
+
+% latex table generated in R 2.15.1 by xtable 1.5-6 package
+% Mon Jan  5 11:27:35 2015
+\begin{table}[ht]
+  \caption{\httph{Server} headers for hosts that enabled a freshness
+    lifetime to be computed for all requests.}\label{tab:servers}
+\begin{center}
+\begin{tabular}{rl}
+  \hline
+ & x \\ 
+  \hline
+1 & DFE/largefile \\ 
+  2 & git\_frontend \\ 
+  3 & nginx/1.3.9 \\ 
+  4 & thin 1.6.0 codename Greek Yogurt \\ 
+  5 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server OracleAS-Web-Cache-10g/10.1.2.3.1 (N;ecid=72058133508068653,0) \\ 
+  6 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server OracleAS-Web-Cache-10g/10.1.2.3.1 (N;ecid=72057948824487974,0) \\ 
+  7 & TwistedWeb/8.2.0 \\ 
+  8 & RDF::Endpoint/0.07 \\ 
+  9 & Jetty(6.1.26) \\ 
+  10 & nginx/1.6.1 \\ 
+  11 & Jigsaw/2.3.0-beta3 \\ 
+  12 & Apache/2.2.9 (Win32) PHP/5.2.6 \\ 
+  13 & Apache/2.4.10 (Unix) mod\_fcgid/2.3.9 \\ 
+  14 &  \\ 
+  15 & GFE/2.0 \\ 
+  16 & RDF::LinkedData/0.70 \\ 
+  17 & Apache/2.2.17 (Unix) mod\_wsgi/3.3 Python/2.6.6 \\ 
+  18 & Virtuoso/07.10.3211 (Linux) i686-generic-linux-glibc212-64  VDB \\ 
+  19 & Apache/2.2.24 (Unix) mod\_ssl/2.2.24 OpenSSL/0.9.8y \\ 
+  20 & Apache/2.2.22 (Fedora) \\ 
+  21 & INSEE \\ 
+  22 & GitHub.com \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 

### 2015-01-05 sanity/queries.R
use matrix for xtable
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@939 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -11,4 +11,5 @@ chisq.test(checkserver$results$server, respromisebin, simulate.p.value=T, B = 10
 
 checkservercount <- sparqlfile("check-server-count.rq")
 serverratio <- data.frame(name=checkservercount$results$server, promise=checkservercount$results$pc/checkservercount$results$sc, hard=checkservercount$results$hc/checkservercount$results$sc, pc=checkservercount$results$pc, hc=checkservercount$results$hc)
-xtable(serverratio[which(serverratio$hard == 1), ]$name)
+xtable(as.matrix(serverratio[which(serverratio$hard == 1), ]$name))
+

### 2015-01-05 sanity/paper1.tex
distribution of freshness time
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@940 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+A\todo{Do I need details on how to compute the freshness lifetime?}
+
 \subsection{Different server implementations}
@@
   for SPARQLES endpoints?}
+
+\subsection{Distribution of freshness time}
+
+We next examine the freshness lifetime of the 2965 information
+resources that returned a successful response, either with SPARQL
+results or RDF data. 405 returned valid headers, but 114 did so to
+\emph{prohibit} caching of the response, and 3 contained conflicting
+headers, i.e. set an freshness lifetime, but also prohibited
+caching. Additionally, 269 resources had a \httph{Cache-Control}
+header to control other aspects of caching than lifetime, i.e. to say
+that only private caches may use the response, that the cache must be
+revalidated, or to prohibit caching.
+
+Caching may prohibited by setting a non-positive freshness
+lifetime. Common practice is to set the \httph{Expires} header to a
+date far back in time, which has also been done in large parts of the
+present dataset. For the following, we have set any non-positive
+lifetime to 0, as that is semantically equivalent. With that, we
+obtain the summary statistics in Table~\ref{tab:summaryhard}. In
+figure\todo{add figure}, we have included a histogram with the counts
+of the resources that had 0 lifetime, and then boxes of varying length.
+
+\begin{table}[ht]
+\begin{center}
+\caption{Summary statistics for standards-compliant freshness lifetime}\label{tab:summaryhard}
+\begin{tabular}{lr}
+Min.   &       0   \\ 
+1st Qu.&       0   \\ 
+Median &     600   \\ 
+Mean   &  583003   \\ 
+3rd Qu.&  172800   \\ 
+Max.   & 52702521   \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 

### 2015-01-05 sanity/queries.R
Stats for all fresh
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@941 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -13,3 +13,14 @@ checkservercount <- sparqlfile("check-server-count.rq")
 serverratio <- data.frame(name=checkservercount$results$server, promise=checkservercount$results$pc/checkservercount$results$sc, hard=checkservercount$results$hc/checkservercount$results$sc, pc=checkservercount$results$pc, hc=checkservercount$results$hc)
 xtable(as.matrix(serverratio[which(serverratio$hard == 1), ]$name))
 
+timebuckets <- function(data) {
+    list(points = c(0,1,60,60*60,24*60*60,7*24*60*60,30*24*60*60,365*24*60*60,max(data)),
+         names = c("Off","Seconds","Minutes","Hours","Days","Weeks","Months","Years"))
+}
+
+allhard <- sparqlfile("all-hard.rq")
+fresh <-  allhard$results$fresh
+fresh[fresh <= 0] <- 0
+hardhist <- hist(fresh, breaks=timebuckets(fresh)$points, plot=F)
+bphard <- barplot(hardhist$count, col="white", names.arg=timebuckets(fresh)$name, xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
+text(bphard, hardhist$counts, labels=hardhist$counts, pos=1)

### 2015-01-08 sanity/queries.R
Queries for all types
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@942 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -24,3 +24,13 @@ fresh[fresh <= 0] <- 0
 hardhist <- hist(fresh, breaks=timebuckets(fresh)$points, plot=F)
 bphard <- barplot(hardhist$count, col="white", names.arg=timebuckets(fresh)$name, xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
 text(bphard, hardhist$counts, labels=hardhist$counts, pos=1)
+
+otherhard <- sparqlfile("other-hard.rq")
+
+lapply(unique(otherhard$results$type), function(type) {
+    times <- otherhard$results[which(otherhard$results$type == type),3]
+    times[times <= 0] <- 0
+    myhist <- hist(times, breaks=timebuckets(times)$points, plot=F)
+    myhist$xname <- type
+    myhist
+})

### 2015-01-08 sanity/queries.R
Generate table and barplot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@943 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -27,10 +27,22 @@ text(bphard, hardhist$counts, labels=hardhist$counts, pos=1)
 
 otherhard <- sparqlfile("other-hard.rq")
 
-lapply(unique(otherhard$results$type), function(type) {
+mybreaks <- timebuckets(otherhard$results$fresh)$points
+types <- unique(otherhard$results$type)
+
+allhists <- lapply(types, function(type) {
     times <- otherhard$results[which(otherhard$results$type == type),3]
     times[times <= 0] <- 0
-    myhist <- hist(times, breaks=timebuckets(times)$points, plot=F)
+    myhist <- hist(times, breaks=mybreaks, plot=F)
     myhist$xname <- type
     myhist
 })
+
+tmp <- NULL # It feels wrong to do this this way...
+tmpm <- sapply(allhists, function(thishist) {
+    tmp <- c(tmp, thishist$counts)
+})
+colnames(tmpm) <- types
+rownames(tmpm) <- timebuckets(otherhard$results$fresh)$names
+alltable <- as.table(tmpm)
+rm(tmp, tmpm)

### 2015-01-08 sanity/paper1.tex
write about the breakup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@944 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Other caching headers}
+
+The two most important headers to control caching are \httph{Expires}
+and \httph{Cache-Control}. As shown in Table~\ref{tab:headers}, there
+are also others. We found \httph{Pragma} in 287 responses, but except
+for two hosts, where they were superflous, they were only used to
+prohibit caching. \httph{Surrogates} were not observed.
+
 \subsection{Distribution of freshness time}
@@
 headers, i.e. set an freshness lifetime, but also prohibited
-caching. Additionally, 269 resources had a \httph{Cache-Control}
-header to control other aspects of caching than lifetime, i.e. to say
-that only private caches may use the response, that the cache must be
-revalidated, or to prohibit caching.
+caching. In most cases, \httph{Cache-Control} and \httph{Expires} both
+occured, but the former is more common than the latter in the cases
+where only one of them occur.  Additionally, 269 resources had a
+\httph{Cache-Control} header to control other aspects of caching than
+lifetime, i.e. to say that only private caches may use the response,
+that the cache must be revalidated, or to prohibit caching.
 
-Caching may prohibited by setting a non-positive freshness
+Caching may also be prohibited by setting a non-positive freshness
 lifetime. Common practice is to set the \httph{Expires} header to a
@@
 \end{table}
+
+In Figure\todo{figure with over freshness}, there is a barplot where
+the freshness lifetime is grouped in bins, where the first bin are the
+cases where caching is explicitly prohibited. The next bins are for
+lifetimes, where the values are grouped if they are on the order of
+seconds, minutes, hours, etc, i.e. the second bin counts the lifetimes
+in the interval \[1,59\] seconds, etc. We see that in these
+categories, the most common is to prohibit caching. Nevertheless, many
+also declare a standards compliant freshness lifetime in minutes to
+days.
+
+In Figure\todo{figure with freshness breakdown}, we have broken this
+up by the type of resource that was accessed, i.e. SPARQL endpoints,
+vocabularies, dataset descriptions or unclassified information
+resources. First, we note that it seems like the distribution of
+freshness lifetime is quite different for the different types, an
+observation that is also supported by a similar hypothesis test as
+above, with a \pvalue = 0.00001. We also note that it is often
+prohibited to cache dataset descriptions. This is odd, since
+statistics about datasets is usually costly to compute and should be
+cached. The VoID specification\todo{ref} also notes that the
+statistics are considered estimates.
+
+We also note that prohibition of SPARQL results caching is rare,
+amongst the servers that expose caching headers, it is common
+that the result may be cached for some minutes.\todo{more to point out?}
 
@@
 \begin{table}
-\caption{Recorded HTTP headers}
+\caption{Recorded HTTP headers}\label{tab:headers}
  \begin{tabular}{ | l |  p{3cm} | p{5cm} |}
@@
 another request would be made to record the selected HTTP headers from
-that as well.
+that as well.\todo{say how many requests failed? It was a lot}
 \end{appendices}

### 2015-01-08 sanity/queries.R
Refactor analysis
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@946 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -18,31 +18,35 @@ timebuckets <- function(data) {
          names = c("Off","Seconds","Minutes","Hours","Days","Weeks","Months","Years"))
 }
 
-allhard <- sparqlfile("all-hard.rq")
-fresh <-  allhard$results$fresh
-fresh[fresh <= 0] <- 0
 hardhist <- hist(fresh, breaks=timebuckets(fresh)$points, plot=F)
-bphard <- barplot(hardhist$count, col="white", names.arg=timebuckets(fresh)$name, xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
-text(bphard, hardhist$counts, labels=hardhist$counts, pos=1)
 
-otherhard <- sparqlfile("other-hard.rq")
 
-mybreaks <- timebuckets(otherhard$results$fresh)$points
-types <- unique(otherhard$results$type)
+lifetimetable <- function(filename) {
+    data <- sparqlfile(filename)
+    
+    mybreaks <- timebuckets(data$results$fresh)$points
+    types <- unique(data$results$type)
+    
+    allhists <- lapply(types, function(type) {
+        times <- data$results[which(data$results$type == type),3]
+        times[times <= 0] <- 0
+        myhist <- hist(times, breaks=mybreaks, plot=F)
+        myhist$xname <- type
+        myhist
+    })
+    
+    tmp <- NULL # It feels wrong to do this this way...
+    tmpm <- sapply(allhists, function(thishist) {
+        tmp <- c(tmp, thishist$counts)
+    })
+    colnames(tmpm) <- types
+    rownames(tmpm) <- timebuckets(data$results$fresh)$names
+    as.table(tmpm)
+}
 
-allhists <- lapply(types, function(type) {
-    times <- otherhard$results[which(otherhard$results$type == type),3]
-    times[times <= 0] <- 0
-    myhist <- hist(times, breaks=mybreaks, plot=F)
-    myhist$xname <- type
-    myhist
-})
+hardtable <- lifetimetable("other-hard.rq")
+hardall <- apply(hardtable, 1, sum)
+bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
+text(bphard, hardall, labels=hardall, pos=1)
 
-tmp <- NULL # It feels wrong to do this this way...
-tmpm <- sapply(allhists, function(thishist) {
[…]

### 2015-01-08 sanity/queries.R
Add heuristics
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@947 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -50,3 +50,10 @@ bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness life
 text(bphard, hardall, labels=hardall, pos=1)
 
 mosaicplot(hardtable)
+
+heuristictable <- lifetimetable("other-heuristic.rq")
+heuristicall <- apply(heuristictable, 1, sum)
+bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Frequency", main='')
+text(bpheuristic, heuristicall, labels=heuristicall, pos=1)
+
+mosaicplot(heuristictable)

### 2015-01-08 sanity/queries.R
Cleanup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@948 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -18,9 +18,6 @@ timebuckets <- function(data) {
          names = c("Off","Seconds","Minutes","Hours","Days","Weeks","Months","Years"))
 }
 
-hardhist <- hist(fresh, breaks=timebuckets(fresh)$points, plot=F)
-
-
 lifetimetable <- function(filename) {
     data <- sparqlfile(filename)
     

### 2015-01-08 sanity/paper1.tex
monitor to test brokenness
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@945 43ef1190-90fd-4dd7-a769-d0121f02b80e
heuristics 1
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@949 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+We next consider the simple heuristic freshness lifetime. This is
+implemented as suggested in RFC7234, section~4.2.2. This is based on a
+fraction of the time lapsed between the current time and the
+modification time given in the \httph{Last-Modified} header. This
+approach therefore still requires the Web server to be cooperative to
+be successful, but commonly, Web servers can track this, based e.g. on
+the file modification time of the underlying file system.
+
+We were able to compute a heuristic lifetime for 554 resources, a
+larger number than standards-compliant resources. In Figure\todo{add all
+  heuristics figure}, we see that the distribution of lifetimes are
+radically different from the case in Figure\todo{ref fig:allhard}. In
+this case, we may cache many resources for months. Only a handful of
+resources had changed in the last minutes. Since this is based
+on actual times since last modifications, this suggests that many
+resources should have had explicit cache headers with very long
+lifetimes. \todo{compare Jürgen and LD observatory}
+
+Moreover, by inspecting Figure\todo{add heuristic table figure}, we
+note that the difference between different types of resources is much
+smaller. This is confirmed by a $\chi^2$ test that yields \pvalue =
+0.02.
+
+We find that only one SPARQL Endpoint yields a heuristic lifetime, on
+closer inspection, we find this to be hosted by Dydra\todo{ref?}. We
+speculate that this is due to that few underlying DBMS systems help
+track modification times in a way that can be used on a SPARQL result
+basis.
+
 \section{Future work}
@@
 
+Monitor the change frequency and use a contingency table to assess the
+correctness of caching headers.
+
 \section{Recommendations}\todo{write}
+
+\todo{ref kasei on mtime}.
 

### 2015-01-09 sanity/paper1.tex
DCT discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@950 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 resources that returned a successful response, either with SPARQL
-results or RDF data. 405 returned valid headers, but 114 did so to
+results or RDF data.
+
+\subsubsection{Standards-compliant caching headers}
+
+Of these resources, 405 returned valid headers, but 114 did so to
 \emph{prohibit} caching of the response, and 3 contained conflicting
@@
 
+\subsubsection{Simple heuristic freshness estimates}
+
 We next consider the simple heuristic freshness lifetime. This is
@@
 
+\subsubsection{Heuristic freshness from Dublin Core properties}
+
+We noted that the Dublin Core Metadata terms vocabulary has a number
+of predicates that may become useful in determining heuristic
+freshness in the future, so we recorded any statements containing the
+predicates \rdfterm{dct:date}, \rdfterm{dct:accrualPeriodicity},
+\rdfterm{dct:created}, \rdfterm{dct:issued}, \rdfterm{dct:modified} or
+\rdfterm{dct:valid}.
+
+First, we compare dates given in \rdfterm{dct:modified} to dates given
+in \httph{Last-Modified} when both are available for a resource. They
+are often not the same, but it appears that dates in the former are
+further back in time than the latter. We speculate that this may be
+due to that the webserver tracks semantically insignificant changes
+through the filesystem, while authors of RDF only update timestamp
+when significant changes are made, or it may be that authors forget to
+update their timestamps.
+
+\rdfterm{dct:modified} occurs in 2687 triples, but 2487 of these does
+not have the Request-URI of the information resource as their subject,
+i.e. it gives the last modification time of some subgraph of the
+returned RDF. Nevertheless, given its prevalence, it is highly likely
+that the presence of \rdfterm{dct:modified} will be useful in
+determining a heuristic freshness lifetime, as the latest date may be
+used.\todo{Actually calculate lifetimes based on this and compare}
+
+\rdfterm{dct:valid} occurred 21 times, and could be used in lieu of an
+\httph{Expires} header, but none of these occurences had a date in the
+future.
+
+\rdfterm{dct:accrualPeriodicity} occurred only twice, and in neither
+case contained machine readable data.
+
+\rdfterm{dct:date}, \rdfterm{dct:created} and \rdfterm{dct:issued}
+were present in 36, 389 and 1475 triples respectively. They correspond
+roughly to the \httph{Date} header, which is present in all requests,
+and they are therefore not important, but given their prevalence they
+could be useful in further analysis.
+
 \section{Future work}
@@
 
+Further investigate the correctness of the \rdfterm{dct:modified} property.
+
 \section{Recommendations}\todo{write}
@@
 \todo{ref kasei on mtime}.
+
+A change periodicity predicate should be standardized in Void.
 
@@
 For all others, we parsed the response, and recorded any errors if the
-parser concluded the content were invalid. We noted that the Dublin Core
-Metadata terms vocabulary has a number of predicates that may become
-useful in determining heuristic freshness in the future, so we
-recorded any statements containing the predicates \rdfterm{dct:date},
-\rdfterm{dct:accrualPeriodicity}, \rdfterm{dct:created},
-\rdfterm{dct:issued}, \rdfterm{dct:modified} or \rdfterm{dct:valid}.
+parser concluded the content were invalid.
 

### 2015-01-09 sanity/queries.R
Create QQplot for KM
ok vs error

git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@951 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -54,3 +54,9 @@ bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshne
 text(bpheuristic, heuristicall, labels=heuristicall, pos=1)
 
 mosaicplot(heuristictable)
+
+harderrordata <- sparqlfile("failed-hard.rq")
+harderrordata$results$fresh[harderrordata$results$fresh <0] <- 0
+errorok <- harderrordata$results$fresh[which(harderrordata$results$status == "OK")]
+errorparse <- harderrordata$results$fresh[which(harderrordata$results$status == "parseerror")]
+qqplot(errorok,errorparse)

### 2015-01-09 sanity/paper1.tex
Future work on collections
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@952 43ef1190-90fd-4dd7-a769-d0121f02b80e
mention qqplot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@953 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 may host SPARQL Endpoints, ontologies, or other information resources,
-by a spider detailed in Appendix~\ref{app:fetcher}
+by a spider detailed in Appendix~\ref{app:fetcher}, resulting in 7745
+requests.
 
@@
 
-We next examine the freshness lifetime of the 2965 information
-resources that returned a successful response, either with SPARQL
-results or RDF data.
+We obtained a successful response from a total 2965 information
+resources, either with SPARQL results or RDF data. A successful
+response is rather tightly defined, not only must there be a
+successful HTTP response after redirects are resolved, the response must also
+return a valid RDF media type (unless it is a SPARQL result) and the
+RDF must parse into an RDF model.
+
+Since web servers may be configured to instruct clients and proxies to
+cache errors differently from successfully served requests, we should
+first investigate whether we can simply use all cache headers in the
+analysis, or if we should remove unsuccessful responses to avoid
+introducing bias to the analysis of valid responses.
+
+For a visual inspection, we may use a quantile-quantile plot to plot
+the freshness lifetimes of successful responses versus requests that
+failed for some reason. If the distribution of both variables are the
+same, these points will draw a straight line. In Figure\todo{include
+  qqplot}, we see successful responses vs. the most common parse
+error, and it is clear that these are so different that no further
+formal analysis is necessary, we must filter unsuccessful requests in
+the following analysis.
 
@@
 amongst the servers that expose caching headers, it is common
-that the result may be cached for some minutes.\todo{more to point out?}
+that the result may be cached for some minutes.\todo{more to point
+  out?}
 
@@
 Identify implementations that expose suitable caching headers.
+
+Investigate whether curated collections such as LOV or SPARQLES
+contain resources that has different characteristics. Since the
+different sources we surveyed overlap, this cannot be done with
+Pearson's $\chi^2$ test, but requires more sophisticated statistics.
 

### 2015-01-09 sanity/paper1.tex
small changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@954 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 This results in an NQuads file per host, which is then loaded into a
-SPARQL Endpoint\todo{which} for analysis by using the statistics
+Virtuoso-based SPARQL Endpoint for analysis by using the statistics
 system R\todo{ref}.
@@
 time (which is the ideal) as well as whether the response had other
-promising features that could assist caching, such as \httph{ETag}s,
-modification time, certain predicates, etc.
+promising features that could assist caching, modification time,
+certain predicates, etc.
 
@@
 seconds, minutes, hours, etc, i.e. the second bin counts the lifetimes
-in the interval \[1,59\] seconds, etc. We see that in these
+in the interval [1,59] seconds, etc. We see that in these
 categories, the most common is to prohibit caching. Nevertheless, many

### 2015-01-09 sanity/paper1.tex
key-value caches
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@955 43ef1190-90fd-4dd7-a769-d0121f02b80e
mention DyLDO
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@956 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 our concern.
+
+The \todo{ref dyldo} Dynamic Linked Data Observatory (DyLDO)
+performed, and continue to do so, of parts of the Linked Open Data
+Cloud to determine dynamicity characteristics of Linked Data. Caching
+is one of their motivations, but they have not published statistics on
+HTTP headers.
 
@@
 
+All Web cache implementations we have studied have cached responses in
+the form of a key that identifies a serialized object. For short-term
+impact, future work should accept this as an architectural constraint.
+
 \begin{appendices}

### 2015-01-09 sanity/queries.R
Use % in labels
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@957 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -44,14 +44,14 @@ lifetimetable <- function(filename) {
 hardtable <- lifetimetable("other-hard.rq")
 hardall <- apply(hardtable, 1, sum)
 bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
-text(bphard, hardall, labels=hardall, pos=1)
+text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
 
 mosaicplot(hardtable)
 
 heuristictable <- lifetimetable("other-heuristic.rq")
 heuristicall <- apply(heuristictable, 1, sum)
 bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Frequency", main='')
-text(bpheuristic, heuristicall, labels=heuristicall, pos=1)
+text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=1)
 
 mosaicplot(heuristictable)
 

### 2015-01-09 sanity/paper1.tex
comment on 60%
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@958 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 resources should have had explicit cache headers with very long
-lifetimes. \todo{compare Jürgen and LD observatory}
+lifetimes. This is supported by DyLDO\todo{ref}, which concludes that:
+\begin{quote}
+[...] We found that 62.2\% of documents didn’t change over the six
+months and found that 51.9\% of domains were considered static.
+\end{quote}
+This agrees well with that 60\% of the simple heuristic lifetimes are
+in the month range.
 

### 2015-01-09 sanity/paper1.tex
some recs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@959 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\todo{ref kasei on mtime}.
+Based on the results of DyLDO, it is highly likely that most cache
+prohibitions are misguided, and server administrators are adviced to
+turn them off unless they are sure they are required.
+
+Data owners should seek to understand their change frequency, and
+adjust cache headers accordingly. Framework authors should make it
+easy to set the expected lifetime, and manage changes to the
+publishing schedule.
+
+To allow generation and validation of \httph{Last-Modified} and
+\httph{ETag} headers, DBMS authors should make sure it is much cheaper
+to retrieve the modification time of any subgraph, than to retrieve
+the subgraph itself. This would be a great improvement for
+RFC7232-based caching, when revalidation is required. It would also
+help simple heuristic based caching.\todo{ref kasei on mtime}.
 

### 2015-01-12 sanity/paper1.tex
start of conditionals
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@960 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\subsection{Cache validation}
+
+Hitherto, we have considered the case where the client or proxy does
+not send a HTTP request to the server when a resource that is present
+in the cache is requested. This is desireable if the client can be
+sufficiently confident that the response is fresh, either by having a
+standards-compliant freshness lifetime or a heuristic to determine
+it. At the end of this period, or if the previous response had no
+lifetime, due to lack of information, or to that the server has stated
+so in the \httph{Cache-Control} header, responses must be
+revalidated. In this case, RFC7232 defines the behaviour, with
+\httph{ETag} and \httph{Last-Modified} as the relevant response
+headers.
+
 \section{Future work}

### 2015-01-12 sanity/paper1.tex
conditionals
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@961 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subsubsection{Simple heuristic freshness estimates}
+\subsubsection{Simple heuristic freshness estimates}\label{sec:simplefresh}
 
@@
 headers.
+
+The BTC had recorded these headers in their data, where 1733 had the
+\httph{ETag} header and 690 had \httph{Last-Modified} with a great
+overlap. For the resources where either or both were available, we
+made our initial request conditional, and 911 responses were verified
+as still fresh.
+
+In total, 1260 successful initial responses contained
+an \httph{ETag}, 606 for vocabularies, 117 for datasets, just 12 for
+endpoints and 525 for unclassified information resources.
+
+To see if the server actually supported conditional requests, and not
+just merely set the response headers, we made another 1822 requests to
+the resources that had these headers. Then, we checked if the response
+code was 200, and the conditional headers had not changed since our
+initial requests. In 85 cases, conditional requests were not supported
+according to the standard, no cases for endpoints, 3 for datasets and
+23 for vocabularies, 59 for generic information resources.\todo{table instead?}
+
+We already used \httph{Last-Modified} to compute heuristic lifetime
+in Section~\ref{sec:simplefresh} for 554 resources.\todo{really needed?}
 

### 2015-01-12 sanity/paper1.tex
small
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@963 43ef1190-90fd-4dd7-a769-d0121f02b80e
some references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@964 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-%\documentclass{llncs}
-\documentclass{article}
+\documentclass{llncs}
+%\documentclass{article}
 \usepackage[utf8]{inputenc}
@@
 \title{A survey of HTTP caching options on the open Web}
-\author{Kjetil Kjernsmo}%\inst{1}}
-%\institute{Department of Informatics,
-%Postboks 1080 Blindern,
-%N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}
+\author{Kjetil Kjernsmo\inst{1}}
+\institute{Department of Informatics,
+Postboks 1080 Blindern,
+N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}
 
-%\subtitle{---Unpublished Working Draft, do not circulate---}
+\subtitle{---Unpublished Working Draft, do not circulate---}
 
@@
 of the World Wide Web. Out of the 6 documents that make up the HTTP
-1.1 standard, RFC7234 \todo{ref} is entirely devoted to the topic, and
-RFC7232 defines conditional requests, which is also important when
-constructing caches. As the former notes:
+1.1 standard, RFC7234 \cite{rfc7234} is entirely devoted to the topic,
+and RFC7232 \cite{rfc7232} defines conditional requests, which is also
+important when constructing caches. As the former notes:
 \begin{quote} 
@@
 \end{quote}
-Furthermore, the Architecture of the World Wide Web \todo{ref}
-discusses caching throughout, and the definition of the
-Representational State Transfer (REST) architectural style \todo{ref}
-is partly motivated from the requirement to implement efficient
-caching. We also note that caching in the Internet infrastructure,
-through so-called Content Delivery Networks, is both a large business
-area and something that could provide great value to the Semantic Web
-community.
+Furthermore, the Architecture of the World Wide Web
+\cite{Jacobs:04:AWW} discusses caching throughout, and the definition
+of the Representational State Transfer (REST) architectural style
+\cite{Fielding_2000_Architectural-Styles} is partly motivated from the
+requirement to implement efficient caching. We also note that caching
+in the Internet infrastructure, through so-called Content Delivery
+Networks, is both a large business area and something that could
+provide great value to the Semantic Web community.
 
@@
 investigate the actual compliance to RFC7234 and RFC7232. Our reasons
-to do this include
+to do this include\todo{formulate as contributions?}
 \begin{enumerate}
@@
 
-LDF\todo{ref} asserted that SPARQL query caching is only possible on a
-per-query basis, but future research should challenge this
-assumption. With this in mind, if reasoning is not involved, we note
-that ontologies are not relevant to query answering since they usually
-occur in bound terms. It is interesting to understand caching when
-this is not the case. Moreover, many ontologies are maintained by a
-small group of humans and change seldomly.
+In \cite{ldf1}, it was asserted that SPARQL query caching is only
+possible on a per-query basis, but future research should challenge
+this assumption. With this in mind, if reasoning is not involved, we
+note that ontologies are not relevant to query answering since they
+usually occur in bound terms. It is interesting to understand caching
+when this is not the case. Moreover, many ontologies are maintained by
+a small group of humans and change seldomly.\todo{not particularly clear}
 
@@
 
-The \todo{ref dyldo} Dynamic Linked Data Observatory (DyLDO)
+The \cite{dyldo2} Dynamic Linked Data Observatory (DyLDO)
 performed, and continue to do so, of parts of the Linked Open Data
@@
 
-LDF\todo{ref paper1} claims to take advantage of caching and contrasts
-this with the unavailability of SPARQL query caches, and claims this is
-an architectural problem.
+Linked Data Fragments \cite{ldf1} claims to take advantage of caching
+and contrasts this with the unavailability of SPARQL query caches, and
+claims this is an architectural problem.
 
 The implications of HTTP for query caches has been discussed by
-\todo{ref kasei}. In \todo{marin, Auer} the authors implemented a
+\cite{kaseicache}. In \cite{sparqlproxy} the authors implemented a
 reverse proxy that controlled the changes to the dataset, and
-therefore could
+therefore could make sure the proxy had all the information needed to
+determine freshness. We are interested in the situation where the
+changes cannot be controlled.
 
-A general caching approach is given in \todo{Caching intermediate
-  result of sparql queries}.
+A general caching approach is given in \cite{yang2011caching}.  In
+\cite{umbrich2012hybrid}, the term caching was used in a different
+sense than we do, they rather prefetch an entire dataset to a local
+store and based on heuristics tried to determine which parts of the
+query should be evaluated remotely and locally. \cite{lampo2011cache}
+explored when caching had a positive effect on complex SPARQL queries.
 
-\todo{fast vs. fresh} didn't consider caching in the sense we do, they
-rather prefetch an entire dataset to a local store and based on
-heuristics tried to determine which parts of the query should be
-evaluated remotely and locally.
-
-\todo{cache or not} explored when caching had a positive effect on
-complex SPARQL queries.
-
-In the broader Web literature, \todo{Revisiting Cacheability
-in Times of User Generated Content} analysed the value of caching
-based anonymized traces of actual Web usage at a major Internet
-Service Provider.
+In the broader Web literature, \cite{ager2010revisiting} analysed the
+value of caching based anonymized traces of actual Web usage at a
+major Internet Service Provider.
 
@@
 test this hypothesis, we create two so-called \emph{contigency
-  tables}\todo{ref}. This formalism is suited to see if the
+  tables}\cite{kn:bj}. This formalism is suited to see if the
 distribution of  \httph{Server} headers are different for those implementations that
@@
 
-Using the statistics system R\todo{ref}, we use a statistical test,
+Using the statistics system R\cite{kn:r}, we use a statistical test,
 namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000

### 2015-01-12 sanity/paper1.tex
fix all cites
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@966 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The SPARQLES survey\todo{ref} has recorded the availability of SPARQL
-endpoints registered in the datahub.io portal, and found that many of
-them have ceased to operate. We used their list of SPARQL endpoints as
-of 2014-11-17, and filtered out those deemed unresponsive by
-them. This resulted in a list of 312 endpoints.
+The SPARQLES survey\cite{buil2013sparql} has recorded the availability
+of SPARQL endpoints registered in the datahub.io portal, and found
+that many of them have ceased to operate. We used their list of SPARQL
+endpoints as of 2014-11-17, and filtered out those deemed unresponsive
+by them. This resulted in a list of 312 endpoints.
 
@@
 To examine as many different implementations and hosts as possible, we
-noted that the Billion Triple Challenge 2014\todo{ref} dataset
+noted that the Billion Triple Challenge 2014 \cite{btc-2014} dataset
 consisted of a 4~GTriple corpus of spidered Web data. To compile a
@@
 Virtuoso-based SPARQL Endpoint for analysis by using the statistics
-system R\todo{ref}.
+system R\cite{kn:r}.
 
@@
 statistics about datasets is usually costly to compute and should be
-cached. The VoID specification\todo{ref} also notes that the
+cached. The VoID specification \cite{voidnote} also notes that the
 statistics are considered estimates.
@@
 resources should have had explicit cache headers with very long
-lifetimes. This is supported by DyLDO\todo{ref}, which concludes that:
+lifetimes. This is supported by DyLDO\cite{dyldo2}, which concludes that:
 \begin{quote}
@@
 RFC7232-based caching, when revalidation is required. It would also
-help simple heuristic based caching.\todo{ref kasei on mtime}.
+help simple heuristic based caching. Research in that direction has
+been published in \cite{kaseicache}.
 
@@
 impact, future work should accept this as an architectural constraint.
+
+\bibliographystyle{plain}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{jbact}
+%\bibliographystyle{splncs03}
+\bibliography{webarch,data,rfc,hypermedia,dynamicity}
 
@@
 and heuristics \\ \hline
-\httph{Surrogates} & Edge Architecture Specification\todo{ref} & Draft
+\httph{Surrogates} & Edge Architecture Specification\cite{edgearch} & Draft
 to allow more fine-grained control for proxies. \\ \hline

### 2015-01-13 sanity/paper1.tex
All references
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@967 43ef1190-90fd-4dd7-a769-d0121f02b80e
LOV ref
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@968 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Postboks 1080 Blindern,
-N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}
+N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}}
 
@@
 To compile a list of Web ontologies, we first queried the Linked Open
-Vocabularies \todo{ref} SPARQL endpoint to retrieve all registered
-ontologies. Then, we took the full list of URIs registered with
-prefix.cc. The assumption that they all identify an ontology is
-clearly false, as we found that for example Tim Berners-Lee's FOAF
-profile has a registered URI, but it is unlikely to introduce
-significant bias.
+Vocabularies (LOV) \cite{doi:10.1108/LHT-03-2013-0027} SPARQL endpoint
+to retrieve all registered ontologies. Then, we took the full list of
+URIs registered with prefix.cc. The assumption that they all identify
+an ontology is clearly false, as we found that for example Tim
+Berners-Lee's FOAF profile has a registered URI, but it is unlikely to
+introduce significant bias.
 
@@
 %\bibliographystyle{splncs03}
-\bibliography{webarch,data,rfc,hypermedia,dynamicity}
+\bibliography{webarch,data,rfc,hypermedia,dynamicity,stat1,specs}
 
-\begin{appendices}
+\begin{subappendices}
+\renewcommand{\thesection}{\Alph{section}}%
 
@@
 that as well.\todo{say how many requests failed? It was a lot}
-\end{appendices}
+\end{subappendices}
 

### 2015-01-13 sanity/paper1.tex
Add footnotes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@969 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 We find that only one SPARQL Endpoint yields a heuristic lifetime, on
-closer inspection, we find this to be hosted by Dydra\todo{ref?}. We
-speculate that this is due to that few underlying DBMS systems help
-track modification times in a way that can be used on a SPARQL result
-basis.
+closer inspection, we find this to be hosted by
+Dydra\footnote{http://dydra.com/}. We speculate that this is due to
+that few underlying DBMS systems help track modification times in a
+way that can be used on a SPARQL result basis.
 
@@
 \texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
-the Perl framework RDF::Trine\todo{footnote} was used to parse the
-line. Lines that failed to parse was discarded. We have not
-investigated whether this could introduce biases. Statements were then
-accepted into a new NQuad file if they had a predicate that matched
-the \rdfterm{sd:endpoint}\todo{footnote} or matched a case-insensitive
-regular expression \texttt{sparql} if the subject and objects were
-both resources, or the predicates \rdfterm{void:vocabulary},
-\rdfterm{rdfa:vocabulary} or \rdfterm{api:vocabulary} \todo{footnotes
-  for all}, as well as having an resource as object. Finally,
-statements with the the classes \rdfterm{cogs:Endpoint},
-\rdfterm{owl:Ontology} and \rdfterm{voaf:Vocabulary} \todo{footnotes
-  for all} in the object position were also accepted. More classes and
-properties were considered, but not used in the data reduction if they
-did not occur in the original data.
+the Perl framework
+RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011} was used
+to parse the line. Lines that failed to parse was discarded. We have
+not investigated whether this could introduce biases. Statements were
+then accepted into a new NQuad file if they had a predicate that
+matched the
+\rdfterm{sd:endpoint}\footnote{http://www.w3.org/TR/sparql11-service-description/#sd-endpoint}
+or matched a case-insensitive regular expression \texttt{sparql} if
+the subject and objects were both resources, or the predicates
+\rdfterm{void:vocabulary}\footnote{http://www.w3.org/TR/void/#vocabularies},
+\rdfterm{rdfa:vocabulary}\footnote{http://www.w3.org/TR/rdfa-core/#term-prefix-definitions-vocabulary}
+or
+\rdfterm{api:vocabulary}\footnote{http://purl.org/linked-data/api/vocab#vocabulary},
+as well as having an resource as object. Finally, statements with the
+the classes
+\rdfterm{cogs:Endpoint}\footnote{http://vocab.deri.ie/cogs#Endpoint},
+\rdfterm{owl:Ontology} and
+\rdfterm{voaf:Vocabulary}\footnote{http://lov.okfn.org/vocab/voaf/v2.3/index.html#Vocabulary}
+in the object position were also accepted. More classes and properties
+were considered, but not used in the data reduction if they did not
+occur in the original data.
 
@@
 to allow more fine-grained control for proxies. \\ \hline
-\httph{Client-Aborted} & libwww\todo{ref}  & Header inserted by User
+\httph{Client-Aborted} & libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}  & Header inserted by User
 Agent to indicate that it aborted the download \\ \hline
-\httph{Client-Warning} & libwww\todo{ref}  & Header inserted by User
+\httph{Client-Warning} & libwww  & Header inserted by User
 Agent to give details about problems with the download \\ \hline

### 2015-01-13 sanity/paper1.tex
Footnotes and final reference
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@970 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 To compile a list of Web ontologies, we first queried the Linked Open
-Vocabularies (LOV) \cite{doi:10.1108/LHT-03-2013-0027} SPARQL endpoint
+Vocabularies (LOV) \cite{lov2} SPARQL endpoint
 to retrieve all registered ontologies. Then, we took the full list of
@@
 %\bibliographystyle{splncs03}
-\bibliography{webarch,data,rfc,hypermedia,dynamicity,stat1,specs}
+\bibliography{webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs}
 
@@
 matched the
-\rdfterm{sd:endpoint}\footnote{http://www.w3.org/TR/sparql11-service-description/#sd-endpoint}
+\rdfterm{sd:endpoint}\footnote{http://www.w3.org/TR/sparql11-service-description/\#sd-endpoint}
 or matched a case-insensitive regular expression \texttt{sparql} if
 the subject and objects were both resources, or the predicates
-\rdfterm{void:vocabulary}\footnote{http://www.w3.org/TR/void/#vocabularies},
-\rdfterm{rdfa:vocabulary}\footnote{http://www.w3.org/TR/rdfa-core/#term-prefix-definitions-vocabulary}
+\rdfterm{void:vocabulary}\footnote{http://www.w3.org/TR/void/\#vocabularies},
+\rdfterm{rdfa:vocabulary}\footnote{http://www.w3.org/TR/rdfa-core/\#term-prefix-definitions-vocabulary}
 or
-\rdfterm{api:vocabulary}\footnote{http://purl.org/linked-data/api/vocab#vocabulary},
+\rdfterm{api:vocabulary}\footnote{http://purl.org/linked-data/api/vocab\#vocabulary},
 as well as having an resource as object. Finally, statements with the
 the classes
-\rdfterm{cogs:Endpoint}\footnote{http://vocab.deri.ie/cogs#Endpoint},
+\rdfterm{cogs:Endpoint}\footnote{http://vocab.deri.ie/cogs\#Endpoint},
 \rdfterm{owl:Ontology} and
-\rdfterm{voaf:Vocabulary}\footnote{http://lov.okfn.org/vocab/voaf/v2.3/index.html#Vocabulary}
+\rdfterm{voaf:Vocabulary}\footnote{http://lov.okfn.org/vocab/voaf/v2.3/index.html\#Vocabulary}
 in the object position were also accepted. More classes and properties

### 2015-01-13 sanity/paper1.tex
R:G:HTTP footnote
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@973 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 purpose, we developed and released a module
-RDF::Generator::HTTP\todo{footnote} to the Comprehensive Perl Archive
-Network. We recorded whether the conditional request showed that the
-BTC data were still fresh, and if it was, we retrieved the current
-data, as we had not couple the headers to the body in our original
-retrieval.
+RDF::Generator::HTTP\footnote{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}
+to the Comprehensive Perl Archive Network. We recorded whether the
+conditional request showed that the BTC data were still fresh, and if
+it was, we retrieved the current data, as we had not couple the
+headers to the body in our original retrieval.
 

### 2015-01-13 sanity/paper1.tex
qqplot fig
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@974 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 exposes standards-compliant freshness time of 604800 seconds (i.e. 1
-week) for both LOD and SPARQL endpoints. \todo{do this specifically
-  for SPARQLES endpoints?}
+week) for both LOD and SPARQL endpoints.
 
@@
 failed for some reason. If the distribution of both variables are the
-same, these points will draw a straight line. In Figure\todo{include
-  qqplot}, we see successful responses vs. the most common parse
-error, and it is clear that these are so different that no further
-formal analysis is necessary, we must filter unsuccessful requests in
-the following analysis.
+same, these points will lie on a straight line. In
+Figure~\ref{fig:errorsqq}, we see successful responses vs. the most
+common parse error, and it is clear that these are so different that
+no further formal analysis is necessary, we must filter unsuccessful
+requests in the following analysis.
+
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[width=.5\textwidth]{errorsqq.pdf}}
+  \caption{A quantile-quantile plot of freshness lifetimes for
+    standard compliant headers with and without errors.}
+  \label{fig:errorsqq}
+\end{figure}
 

### 2015-01-13 sanity/queries.R

@@ -59,4 +59,8 @@ harderrordata <- sparqlfile("failed-hard.rq")
 harderrordata$results$fresh[harderrordata$results$fresh <0] <- 0
 errorok <- harderrordata$results$fresh[which(harderrordata$results$status == "OK")]
 errorparse <- harderrordata$results$fresh[which(harderrordata$results$status == "parseerror")]
-qqplot(errorok,errorparse)
+pdf(file="errorsqq.pdf", height=4,width=4)
+par(mai=c(1,1,0.2,0.05))
+qqplot(errorok,errorparse, xlab="Lifetime when no errors", ylab="Lifetime with parse errors")
+dev.off()
+

### 2015-01-13 sanity/paper1.tex
hardall fig
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@975 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 lifetime to 0, as that is semantically equivalent. With that, we
-obtain the summary statistics in Table~\ref{tab:summaryhard}. In
-figure\todo{add figure}, we have included a histogram with the counts
-of the resources that had 0 lifetime, and then boxes of varying length.
+obtain the summary statistics in Table~\ref{tab:summaryhard}.
 
@@
 
-In Figure\todo{figure with over freshness}, there is a barplot where
-the freshness lifetime is grouped in bins, where the first bin are the
-cases where caching is explicitly prohibited. The next bins are for
-lifetimes, where the values are grouped if they are on the order of
-seconds, minutes, hours, etc, i.e. the second bin counts the lifetimes
-in the interval [1,59] seconds, etc. We see that in these
+In Figure~\ref{fig:hardall}, there is a barplot where
+the freshness lifetime is grouped in bins. We see that in these
 categories, the most common is to prohibit caching. Nevertheless, many
@@
 days.
+
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{hardall.pdf}}
+  \caption{Barplot of all standard-compliant freshness lifetimes
+    found. The first bin are the cases where caching is explicitly
+    prohibited. The next bins are for lifetimes, where the values are
+    grouped if they are on the order of seconds, minutes, hours, etc,
+    i.e. the second bin counts the lifetimes in the interval [1,59]
+    seconds, etc.}
+  \label{fig:hardall}
+\end{figure}
 

### 2015-01-13 sanity/queries.R

@@ -43,8 +43,11 @@ lifetimetable <- function(filename) {
 
 hardtable <- lifetimetable("other-hard.rq")
 hardall <- apply(hardtable, 1, sum)
+pdf(file="hardall.pdf", height=5.3,width=8)
+par(mai=c(1,1,0.2,0.05))
 bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
 text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
+dev.off()
 
 mosaicplot(hardtable)
 

### 2015-01-13 sanity/paper1.tex
hardtable fig
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@976 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In Figure\todo{figure with freshness breakdown}, we have broken this
+In Figure~\ref{fig:hardtable}, we have broken this
 up by the type of resource that was accessed, i.e. SPARQL endpoints,
@@
 statistics are considered estimates.
+
+\begin{figure}[ht!]
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{hardtable.pdf}}
+  \caption{Mosaic Plot. Divided by type of resource on the vertical
+    axis and by the same bins as in Figure~\ref{fig:hardall} on the
+    horizontal axis. }
+  \label{fig:hardtable}
+\end{figure}
 

### 2015-01-13 sanity/queries.R

@@ -43,13 +43,16 @@ lifetimetable <- function(filename) {
 
 hardtable <- lifetimetable("other-hard.rq")
 hardall <- apply(hardtable, 1, sum)
-pdf(file="hardall.pdf", height=5.3,width=8)
+pdf(file="hardall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
 bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
 text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
 dev.off()
 
-mosaicplot(hardtable)
+pdf(file="hardtable.pdf", height=5.2,width=8)
+par(mai=c(0.6,0.6,0.05,0.05))
+mosaicplot(hardtable, main='', xlab="Standards-compliant freshness lifetime", ylab="Frequency of resource types")
+dev.off()
 
 heuristictable <- lifetimetable("other-heuristic.rq")
 heuristicall <- apply(heuristictable, 1, sum)

### 2015-01-13 sanity/paper1.tex
Heuristic plots
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@978 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\begin{figure}[ht!]
+\begin{figure}[ht]
   \centerline{%
@@
 
+\begin{figure}[th!]
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{heuristicall.pdf}}
+  \caption{Barplot of all simple heuristic freshness lifetimes
+    found. Bins as in Figure~\ref{fig:hardall}.}
+  \label{fig:heuristicall}
+\end{figure}
+
 We were able to compute a heuristic lifetime for 554 resources, a
-larger number than standards-compliant resources. In Figure\todo{add all
-  heuristics figure}, we see that the distribution of lifetimes are
-radically different from the case in Figure\todo{ref fig:allhard}. In
-this case, we may cache many resources for months. Only a handful of
-resources had changed in the last minutes. Since this is based
-on actual times since last modifications, this suggests that many
-resources should have had explicit cache headers with very long
-lifetimes. This is supported by DyLDO\cite{dyldo2}, which concludes that:
+larger number than standards-compliant resources. In
+Figure~\ref{fig:heuristicall}, we see that the distribution of
+lifetimes are radically different from the case in
+Figure~\ref{fig:hardall}. In this case, we may cache many resources
+for months. Only a handful of resources had changed in the last
+minutes. Since this is based on actual times since last modifications,
+this suggests that many resources should have had explicit cache
+headers with very long lifetimes. This is supported by
+DyLDO\cite{dyldo2}, which concludes that:
 \begin{quote}
@@
 
-Moreover, by inspecting Figure\todo{add heuristic table figure}, we
+Moreover, by inspecting Figure~\ref{fig:heuristictable}, we
 note that the difference between different types of resources is much
@@
 0.02.
+
+\begin{figure}[hb!]
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{heuristictable.pdf}}
+  \caption{Mosaic Plot. Divided by type of resource on the vertical
+    axis and by the same bins as in Figure~\ref{fig:hardall} on the
+    horizontal axis. }
+  \label{fig:heuristictable}
+\end{figure}
 

### 2015-01-13 sanity/queries.R

@@ -56,10 +56,16 @@ dev.off()
 
 heuristictable <- lifetimetable("other-heuristic.rq")
 heuristicall <- apply(heuristictable, 1, sum)
+pdf(file="heuristicall.pdf", height=5.2,width=8)
+par(mai=c(1,1,0.2,0.05))
 bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Frequency", main='')
 text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=1)
+dev.off()
 
-mosaicplot(heuristictable)
+pdf(file="heuristictable.pdf", height=5.2,width=8)
+par(mai=c(0.6,0.6,0.05,0.05))
+mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Frequency of resource types")
+dev.off()
 
 harderrordata <- sparqlfile("failed-hard.rq")
 harderrordata$results$fresh[harderrordata$results$fresh <0] <- 0

### 2015-01-14 sanity/paper1.tex
Some proofreading; some more on related work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@979 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Semantic Web, and therefore we decided to conduct a survey to
-investigate the actual compliance to RFC7234 and RFC7232. Our reasons
-to do this include\todo{formulate as contributions?}
+investigate the actual compliance to RFC7234 and RFC7232. 
+With this paper, we hope to achieve the following:
 \begin{enumerate}
@@
 response to remain fresh, i.e. how long it is permissible to use
-it. Based on this, the client need not contact the server at all to
-reuse a cached response, but this requires a commitment from the
-content provider. The standard also gives the client an opportunity to
-heuristically determine a time to live. RFC7232, on the other hand,
-defines a protocol for asking the server if the cached response is
-still fresh using conditional requests. This doesn't burden the
-content provider with the task of estimating the time to live
-beforehand, but then, it must be able to answer if the resource has
-changed less expensively than serving the entire response. These two
-approaches can be combined, as a client may ask if a response is fresh
-after it has expired.
+it. Based on this, the client doesn't need to contact the server at
+all to reuse a cached response, but this requires a commitment from
+the content provider. The standard also gives the client an
+opportunity to heuristically determine a time to live. RFC7232, on the
+other hand, defines a protocol for asking the server if the cached
+response is still fresh using conditional requests. This doesn't
+burden the content provider with the task of estimating the time to
+live beforehand, but then, it must be able to answer if the resource
+has changed less expensively than serving the entire response. These
+two approaches can be combined, as a client may ask if a response is
+fresh after it has expired. A number of headers are relevant to the
+problem, we recorded those in Table~\ref{tab:headers} in
+Appendix~\ref{app:fetcher} but they were not all used in this analysis.
 
@@
 
-Key contributions of this paper are:\todo{write}
-
 \section{Related work}
@@
 
-The \cite{dyldo2} Dynamic Linked Data Observatory (DyLDO)
-performed, and continue to do so, of parts of the Linked Open Data
+The \cite{dyldo2} Dynamic Linked Data Observatory (DyLDO) performed,
+and continue to do so, monitoring of parts of the Linked Open Data
 Cloud to determine dynamicity characteristics of Linked Data. Caching
@@
 query should be evaluated remotely and locally. \cite{lampo2011cache}
-explored when caching had a positive effect on complex SPARQL queries.
+explored when caching had a positive effect on complex SPARQL
+queries.
 
@@
 value of caching based anonymized traces of actual Web usage at a
-major Internet Service Provider.
+major Internet Service Provider. They find that while caching often
+yields little benefit when content is user-generated, they still find
+some potential.
+
+While these studies have little overlap with the present paper, they
+underline the importance of understanding the current deployment and
+future potential. In some of the related work, it is shown that
+caching does not necessarily give tanglible benefits. Yet, we shall
+assume that sharing the metadata required for caching outside of the
+server is desireable, and that it is possible in most cases. We shall
+see that it most likely will be beneficial in cases that do not
+benefit from caching today.
 

### 2015-01-14 sanity/paper1.tex
Smaller proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@980 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\rdfterm}[1]{\texttt{#1}}
-\newcommand{\pvalue}{\textit{p}-value}
+\newcommand{\pvalue}{\textit{p}-value\ }
 \newcommand{\httph}[1]{\texttt{#1}}
@@
 \emph{all} responses had a freshness time or a promising feature. For
-the former, this amounted to 22 servers, tabulated
-in~\ref{tab:servers}. 70 servers always responded with promising
+the former, this amounted to 22 servers, see 
+Table~\ref{tab:servers}. 70 servers always responded with promising
 features. As we can see, apart from well-known Virtuoso and
@@
   \hline
- & x \\ 
-  \hline
 1 & DFE/largefile \\ 
@@
   4 & thin 1.6.0 codename Greek Yogurt \\ 
-  5 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server OracleAS-Web-Cache-10g/10.1.2.3.1 (N;ecid=72058133508068653,0) \\ 
-  6 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server OracleAS-Web-Cache-10g/10.1.2.3.1 (N;ecid=72057948824487974,0) \\ 
+  5 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
+  6 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
   7 & TwistedWeb/8.2.0 \\ 
@@
 exposing cache headers than others. Unfortunately, since most
-\httph{Server} headers only report generic values, little can be
+\httph{Server} headers only contain generic values, little can be
 learnt about these implementations. We note, however, that DBPedia
 exposes standards-compliant freshness time of 604800 seconds (i.e. 1
-week) for both LOD and SPARQL endpoints.
+week) for both LOD and SPARQL endpoints.\todo{hall of fame?}
 
@@
 
-\subsection{Distribution of freshness time}
+\subsection{Distribution of freshness lifetime}
 
@@
 resources, either with SPARQL results or RDF data. A successful
-response is rather tightly defined, not only must there be a
-successful HTTP response after redirects are resolved, the response must also
-return a valid RDF media type (unless it is a SPARQL result) and the
-RDF must parse into an RDF model.
+response is rather strictly defined, not only must there be a
+successful HTTP response after redirects are resolved, the response
+must also return a valid RDF media type (unless it is a SPARQL result)
+and the response must parse into an RDF model.
 
@@
 We next consider the simple heuristic freshness lifetime. This is
-implemented as suggested in RFC7234, section~4.2.2. This is based on a
+implemented as suggested in RFC7234, Section~4.2.2. This is based on a
 fraction of the time lapsed between the current time and the
@@
 \begin{quote}
-[...] We found that 62.2\% of documents didn’t change over the six
+[\ldots] We found that 62.2\% of documents didn’t change over the six
 months and found that 51.9\% of domains were considered static.
@@
 that. Then, we also let it calculate a \emph{heuristic} freshness
-time based on a suggestion in RFC7234, section~4.2.2, and recorded that.
+time based on a suggestion in RFC7234, Section~4.2.2, and recorded that.
 

### 2015-01-14 sanity/paper1.tex
Smaller proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@981 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 by a spider detailed in Appendix~\ref{app:fetcher}, resulting in 7745
-requests.
+requests, done on 2015-01-02.
 
@@
 
-\section{Analytics}
+\section{Analysis}
 
@@
 approach therefore still requires the Web server to be cooperative to
-be successful, but commonly, Web servers can track this, based e.g. on
-the file modification time of the underlying file system.
+be successful, but commonly, Web servers can track this, for example
+if RDF is served from a file system, the file modification time is used.
 
@@
 RFC7232-based caching, when revalidation is required. It would also
-help simple heuristic based caching. Research in that direction has
+help simple heuristics based caching. Research in that direction has
 been published in \cite{kaseicache}.

### 2015-01-14 sanity/paper1.tex
Conclusions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@982 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 future potential. In some of the related work, it is shown that
-caching does not necessarily give tanglible benefits. Yet, we shall
-assume that sharing the metadata required for caching outside of the
-server is desireable, and that it is possible in most cases. We shall
+caching does not necessarily give tangible benefits. Yet, we shall
+assume that sharing the meta data required for caching outside of the
+server is desirable, and that it is possible in most cases. We shall
 see that it most likely will be beneficial in cases that do not
@@
 are also others. We found \httph{Pragma} in 287 responses, but except
-for two hosts, where they were superflous, they were only used to
+for two hosts, where they were superfluous, they were only used to
 prohibit caching. \httph{Surrogates} were not observed.
@@
 caching. In most cases, \httph{Cache-Control} and \httph{Expires} both
-occured, but the former is more common than the latter in the cases
+occurred, but the former is more common than the latter in the cases
 where only one of them occur.  Additionally, 269 resources had a
@@
 further back in time than the latter. We speculate that this may be
-due to that the webserver tracks semantically insignificant changes
-through the filesystem, while authors of RDF only update timestamp
+due to that the web-server tracks semantically insignificant changes
+through the file system, while authors of RDF only update timestamp
 when significant changes are made, or it may be that authors forget to
@@
 \rdfterm{dct:valid} occurred 21 times, and could be used in lieu of an
-\httph{Expires} header, but none of these occurences had a date in the
+\httph{Expires} header, but none of these occurrences had a date in the
 future.
@@
 not send a HTTP request to the server when a resource that is present
-in the cache is requested. This is desireable if the client can be
+in the cache is requested. This is desirable if the client can be
 sufficiently confident that the response is fresh, either by having a
@@
 
-\section{Future work}
+\section{Conclusions and outlook}
+
+We found that support for HTTP caching and conditional requests are
+occurring regularly in the Semantic Web, but they are not
+commonplace. We found, in agreement with DyLDO\cite{dyldo2}, that many
+resources change at a very slow pace, but also that this is not
+reflected in the standards-compliant freshness lifetimes advertized by
+servers.
+
+We found that errors are commonplace, but that they does not usually
+pertain to the caching headers. We found a small number of
+self-contradictory \httph{Cache-Control} headers, and some servers
+that set conditional request response headers, that could not support
+conditional requests.
+
+It is possible in a substantial number of cases to compute a heuristic
+freshness lifetime, either from the \httph{Last-Modified} header, or
+from Dublin Core properties.
+
+For SPARQL endpoints, we found that conditional requests are 
+seldomly supported, but standards-compliant freshness lifetimes have
+been seen, and since it is supported by DBPedia, benefits from caching
+may be realized already.
+
+In spite of this, most of the Semantic Web is unhelpful in spite of
+being slowly changing.
+
+\subsection{Future work}
 
@@
 
-\section{Recommendations}\todo{write}
+\subsection{Recommendations}\todo{write}
 
 Based on the results of DyLDO, it is highly likely that most cache
-prohibitions are misguided, and server administrators are adviced to
+prohibitions are misguided, and server administrators are advised to
 turn them off unless they are sure they are required.

### 2015-01-14 sanity/paper1.tex
Future work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@983 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Identify implementations that expose suitable caching headers.
+In this work, we have only explored the cases where the server is
+cooperative, in the sense that message data or metadata provides at
+least a hint of a resource's cacheability. We also noted that the
+majority of the Semantic Web is unhelpful. Therefore, an interesting
+direction is to learn the change frequency of resources to use in a
+heuristic freshness lifetime estimation. However, such work should
+operate within the loose constraints of RFC7234, Section~4.2.2, and
+should find it niche when the other techniques described in this paper
+are unavailable. Once this is done, the correctness of caching headers
+should be assessed, possibly using contingency tables.
 
@@
 
-Monitor the change frequency and use a contingency table to assess the
-correctness of caching headers.
+We found that some implementations are likely better than others, but
+further understanding of these differences are impeded by the fact
+that most \httph{Server} headers contained very little
+information. Future work could seek more sophisticated fingerprinting
+and understanding of the nature of these differences.
 
-Further investigate the correctness of the \rdfterm{dct:modified} property.
+Further investigate the suitability of the \rdfterm{dct:modified}
+property for estimating heuristic freshness.
+
+For SPARQL Endpoints, it is interesting to examine the queries in a
+shared proxy, e.g. an institutional cache or a Content
+Delivery Network for frequent subpatterns, and prefetch and cache such
+intermediate results so that subsequent queries can be answered in
+part by the proxy.
 

### 2015-01-14 sanity/paper1.tex
Promote lifetime problems to future work
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@984 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Estimating the freshness lifetime is a challenging problem for data
+owners. It must necessarily involve the human users involved in the
+publishing cycle since they are making a committment about future
+changes. Designing user support systems as well as interfaces that fit
+the publisher's workflow, is an important problem.
+
 For SPARQL Endpoints, it is interesting to examine the queries in a
@@
 
-Data owners should seek to understand their change frequency, and
-adjust cache headers accordingly. Framework authors should make it
-easy to set the expected lifetime, and manage changes to the
-publishing schedule.
+In many cases, setting a reasonable cache headers is straightforward,
+and should be done by data owners. Framework authors should make it
+easy to set the expected lifetime.
 

### 2015-01-14 sanity/paper1.tex
cap void
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@985 43ef1190-90fd-4dd7-a769-d0121f02b80e
Acknowledgements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@986 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-A change periodicity predicate should be standardized in Void.
+A change periodicity predicate should be standardized in VoID.
 
@@
 impact, future work should accept this as an architectural constraint.
+
+\paragraph*{Acknowledgements} The author would like to thank Jürgen
+Umbrich and Axel Polleres for inspiring me to start this study, to
+Martin Giese for careful review and critical comments, and Gregory
+Todd Williams for promptly solving issues in the underlying libraries
+used in this study.
 

### 2015-01-15 sanity/paper1.tex
remove todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@987 43ef1190-90fd-4dd7-a769-d0121f02b80e
comment on the small number of resources
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@988 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \title{A survey of HTTP caching options on the open Web}
-\author{Kjetil Kjernsmo\inst{1}}
+\author{Kjetil Kjernsmo}
 \institute{Department of Informatics,
@@
 
-Linked Data Fragments \cite{ldf1} claims to take advantage of caching
+Linked Data Fragments \cite{ldf1} is said to take advantage of caching
 and contrasts this with the unavailability of SPARQL query caches, and
-claims this is an architectural problem.
+it is claimed that this is an architectural problem.
 
@@
 RDF::Endpoint, which is partly developed by and runs on a server
-operated by this author, most headers tells very little about the
+operated by this author, most headers tell very little about the
 RDF-specific parts of the underlying server implementation. A quick
@@
 % Mon Jan  5 11:27:35 2015
-\begin{table}[ht]
+\begin{table}[hb]
   \caption{\httph{Server} headers for hosts that enabled a freshness
@@
 must also return a valid RDF media type (unless it is a SPARQL result)
-and the response must parse into an RDF model.
+and the response must parse into an RDF model. We have given priority
+to survey many hosts since configuration usually doesn't differ much
+across a host, especially since it also captures different types of
+resources. It is therefore acceptable that the number of resources is
+relatively small.
 
@@
 are unavailable. Once this is done, the correctness of caching headers
-should be assessed, possibly using contingency tables.
+should be assessed, possibly using contigency tables.
 
@@
 
-\subsection{Recommendations}\todo{write}
+\subsection{Recommendations}
 

### 2015-01-15 sanity/paper1.tex
An abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@989 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \begin{abstract}
+We have surveyed RDF resources on the Semantic Web for the use of HTTP
+cache and conditional request headers. We find that while they are
+prevalent enough to be used in some cases, they are not commonly used
+even on relatively static data, and when they are given, they are very
+conservatively set. We point out future directions and give
+recommendations for the enhanced use of caching in the Semantic Web.
 \end{abstract}

### 2015-01-16 sanity/paper1.tex
Minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@990 43ef1190-90fd-4dd7-a769-d0121f02b80e
Martins abstract
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@992 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2015-01-15 sanity/paper1.tex
@@
 
-\title{A survey of HTTP caching options on the open Web}
+\title{A survey of HTTP caching options on the open Semantic Web}
 \author{Kjetil Kjernsmo}
@@
 \begin{abstract}
+Scalability of the data access architecture in the Semantic Web is
+dependent on the establishment of caching mechanisms to take the load
+off of servers.  Unfortunately, there is a chicken and egg problem
+here: research, implementation, and evaluation of caching
+infrastructure is uninteresting as long as data providers do not
+publish relevant metadata.  And publishing metadata is useless as long
+as there is no infrastructure that uses it.
+
+We show by means of a survey of live RDF data sources that caching
+metadata is prevalent enough already to be used in some cases.  On the
+other hand, they are not commonly used even on relatively static data,
+and when they are given, they are very conservatively set. We point
+out future directions and give recommendations for the enhanced use of
+caching in the Semantic Web.
 \end{abstract}
@@
 % Mon Jan  5 11:27:35 2015
-\begin{table}[hb]
+\begin{table}[h]
   \caption{\httph{Server} headers for hosts that enabled a freshness

### 2015-01-16 sanity/paper1.tex
Now start trimming
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@993 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 off of servers.  Unfortunately, there is a chicken and egg problem
-here: research, implementation, and evaluation of caching
+here: Research, implementation, and evaluation of caching
 infrastructure is uninteresting as long as data providers do not

### 2015-01-16 sanity/paper1.tex
Shrink table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@994 43ef1190-90fd-4dd7-a769-d0121f02b80e
small
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@995 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 test this hypothesis, we create two so-called \emph{contigency
-  tables}\cite{kn:bj}. This formalism is suited to see if the
+  tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the
 distribution of  \httph{Server} headers are different for those implementations that
@@
 \caption{Recorded HTTP headers}\label{tab:headers}
- \begin{tabular}{ | l |  p{3cm} | p{5cm} |}
+ \begin{tabular}{ | l |  p{2.3cm} | p{6.8cm} |}
     \hline
@@
 and heuristics \\ \hline
-\httph{Surrogates} & Edge Architecture Specification\cite{edgearch} & Draft
-to allow more fine-grained control for proxies. \\ \hline
+\httph{Surrogates} & Edge \cite{edgearch} & Draft
+to allow fine-grained control for proxies. \\ \hline
 \httph{Client-Aborted} & libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}  & Header inserted by User

### 2015-01-16 sanity/paper1.tex
Clear cuts
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@996 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Appendix~\ref{app:fetcher} but they were not all used in this analysis.
-
-In \cite{ldf1}, it was asserted that SPARQL query caching is only
-possible on a per-query basis, but future research should challenge
-this assumption. With this in mind, if reasoning is not involved, we
-note that ontologies are not relevant to query answering since they
-usually occur in bound terms. It is interesting to understand caching
-when this is not the case. Moreover, many ontologies are maintained by
-a small group of humans and change seldomly.\todo{not particularly clear}
 
@@
 present dataset. For the following, we have set any non-positive
-lifetime to 0, as that is semantically equivalent. With that, we
-obtain the summary statistics in Table~\ref{tab:summaryhard}.
-
-\begin{table}[ht]
-\begin{center}
-\caption{Summary statistics for standards-compliant freshness lifetime}\label{tab:summaryhard}
-\begin{tabular}{lr}
-Min.   &       0   \\ 
-1st Qu.&       0   \\ 
-Median &     600   \\ 
-Mean   &  583003   \\ 
-3rd Qu.&  172800   \\ 
-Max.   & 52702521   \\ 
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
+lifetime to 0, as that is semantically equivalent.
 
@@
 \caption{Recorded HTTP headers}\label{tab:headers}
- \begin{tabular}{ | l |  p{2.3cm} | p{6.8cm} |}
+ \begin{tabular}{ | l |  p{1.4cm} | p{6.8cm} |}
     \hline

### 2015-01-16 sanity/paper1.tex
Martin proofread
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@997 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Networks, is both a large business area and something that could
-provide great value to the Semantic Web community.
+provide great value to the Semantic Web.
 
@@
 We note that caching is not only useful for long-living information
-resources, even though that may be the most important. If a resource
+resources, even though that may be the most important use. If a resource
 is frequently requested, it may make sense to cache it even though it
@@
 
-Caching may be deployed at several different levels: A HTTP cache may
+Caching may be deployed at several different levels: An HTTP cache may
 be in a reverse proxy close to the server, in which case it may have
@@
 The \cite{dyldo2} Dynamic Linked Data Observatory (DyLDO) performed,
-and continue to do so, monitoring of parts of the Linked Open Data
+and continues to do so, monitoring of parts of the Linked Open Data
 Cloud to determine dynamicity characteristics of Linked Data. Caching
@@
 
-Linked Data Fragments \cite{ldf1} is said to take advantage of caching
-and contrasts this with the unavailability of SPARQL query caches, and
-it is claimed that this is an architectural problem.
+Linked Data Fragments is claimed in \cite{ldf1} to take advantage of
+caching and contrasts this with the unavailability of SPARQL query
+caches, and it is claimed that this is an architectural problem.
 
@@
 \cite{umbrich2012hybrid}, the term caching was used in a different
-sense than we do, they rather prefetch an entire dataset to a local
+sense than we do, they rather prefetched an entire dataset to a local
 store and based on heuristics tried to determine which parts of the
@@
 
-First, we investigated whether certain servers provided better support
+First, we investigated whether certain server implementations provided better support
 for caching than others. To do this, we formulated SPARQL queries to
@@
 falls off rapidly, and so, some servers are used only by very
-few. Likewise, it is only to be expected that only a few projects have
+few. Likewise, it is to be expected that only a few projects have
 given caching enough attention, but that they account for the majority
@@
 
-We found that support for HTTP caching and conditional requests are
-occurring regularly in the Semantic Web, but they are not
-commonplace. We found, in agreement with DyLDO\cite{dyldo2}, that many
-resources change at a very slow pace, but also that this is not
+We found moderate uptake for HTTP caching and conditional requests in
+the Semantic Web. We found, in agreement with DyLDO\cite{dyldo2}, that
+many resources change at a very slow pace, but also that this is not
 reflected in the standards-compliant freshness lifetimes advertized by
@@
 \rdfterm{api:vocabulary}\footnote{http://purl.org/linked-data/api/vocab\#vocabulary},
-as well as having an resource as object. Finally, statements with the
+as well as having a resource as object. Finally, statements with the
 the classes

### 2015-01-16 sanity/paper1.tex
Martin todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@998 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 problem, we recorded those in Table~\ref{tab:headers} in
-Appendix~\ref{app:fetcher} but they were not all used in this analysis.
+Appendix~\ref{app:fetcher} but they were not all used in this
+analysis.\todo{Structured walkthrough}
 
@@
 
-A\todo{Do I need details on how to compute the freshness lifetime?}
-
 \subsection{Different server implementations}
@@
 
-For a more systematic approach, we wish to test the hypothesis that
+For a more systematic approach\todo{Move most to appendix}, we wish to test the hypothesis that
 some servers are better configured to support caching than others. To
@@
 Since web servers may be configured to instruct clients and proxies to
-cache errors differently from successfully served requests, we should
-first investigate whether we can simply use all cache headers in the
-analysis, or if we should remove unsuccessful responses to avoid
-introducing bias to the analysis of valid responses.
+cache errors differently\todo{cut most to just state stuff} from
+successfully served requests, we should first investigate whether we
+can simply use all cache headers in the analysis, or if we should
+remove unsuccessful responses to avoid introducing bias to the
+analysis of valid responses.
 
@@
 
-In Figure~\ref{fig:hardtable}, we have broken this
+In Figure~\ref{fig:hardtable}\todo{look for others or explain, colors}, we have broken this
 up by the type of resource that was accessed, i.e. SPARQL endpoints,

### 2015-01-17 sanity/paper1.tex
Jurgen todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@999 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Networks, is both a large business area and something that could
-provide great value to the Semantic Web.
+provide great value to the Semantic Web.\todo{pitch it}
 
@@
 Appendix~\ref{app:fetcher} but they were not all used in this
-analysis.\todo{Structured walkthrough}
+analysis.\todo{Structured walkthrough}\todo{Impact}
 
@@
 
-A general caching approach is given in \cite{yang2011caching}.  In
+A general caching approach is given in \cite{yang2011caching}\todo{explain} . In
 \cite{umbrich2012hybrid}, the term caching was used in a different
@@
 explored when caching had a positive effect on complex SPARQL
-queries.
+queries\todo{explain} .
 
@@
 \section{Methodology}
-
+\todo{Improve, big}
 We want to find information resources on the Web, and examine HTTP
@@
 exposes standards-compliant freshness time of 604800 seconds (i.e. 1
-week) for both LOD and SPARQL endpoints.\todo{hall of fame?}
+week) for both LOD and SPARQL endpoints.\todo{Where and how}
 
@@
 The two most important headers to control caching are \httph{Expires}
-and \httph{Cache-Control}. As shown in Table~\ref{tab:headers}, there
-are also others. We found \httph{Pragma} in 287 responses, but except
-for two hosts, where they were superfluous, they were only used to
-prohibit caching. \httph{Surrogates} were not observed.
+and \httph{Cache-Control}\todo{move to intro}. As shown in Table~\ref{tab:headers}, there
+are also others. We found \httph{Pragma} (archaic HTTP 1.0 header) in
+287 responses, but except for two hosts, where they were superfluous,
+they were only used to prohibit caching. \httph{Surrogates} were not
+observed.
 
@@
 
-In Figure~\ref{fig:hardall}, there is a barplot where
+In Figure~\ref{fig:hardall}\todo{axis and labels}, there is a barplot where
 the freshness lifetime is grouped in bins. We see that in these

### 2015-01-17 sanity/paper1.tex
Pitch caching
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1000 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Networks, is both a large business area and something that could
-provide great value to the Semantic Web.\todo{pitch it}
+provide great value to the Semantic Web.
+
+If used correctly, caching mechanisms will reduce the need to make
+HTTP requests, reduce lookups to the backend systems, reduce the need
+to make repetitive computations, enable sharing of responses in
+Internet infrastructure and reduce latency since requests may be
+answered closer to the client.
 

### 2015-01-18 sanity/paper1.tex
Explain rfcs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1001 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 to make repetitive computations, enable sharing of responses in
-Internet infrastructure and reduce latency since requests may be
-answered closer to the client.
+Internet infrastructure, improve uptime and reduce latency since
+requests may be answered closer to the client.
 
@@
 
-Caching, as defined in RFC7234, requires that the server declares how
-long into the future it expects the information resource in the
-response to remain fresh, i.e. how long it is permissible to use
-it. Based on this, the client doesn't need to contact the server at
-all to reuse a cached response, but this requires a commitment from
-the content provider. The standard also gives the client an
-opportunity to heuristically determine a time to live. RFC7232, on the
-other hand, defines a protocol for asking the server if the cached
-response is still fresh using conditional requests. This doesn't
-burden the content provider with the task of estimating the time to
-live beforehand, but then, it must be able to answer if the resource
-has changed less expensively than serving the entire response. These
-two approaches can be combined, as a client may ask if a response is
-fresh after it has expired. A number of headers are relevant to the
-problem, we recorded those in Table~\ref{tab:headers} in
-Appendix~\ref{app:fetcher} but they were not all used in this
+As mentioned, two standards are relevant for this study, RFC7234,
+which is named ``Caching'' and RFC7232, named ``Conditional
+Requests''. The main difference is that the caching standard defines
+when a response may be reused without contacting the origin server at
+all, whereas the conditional requests define how to validate a
+response by contacting the origin server. The two can be combined:
+Clients and proxies may use the latter to revalidate a response that
+has been cached based on the former.
+
+RFC7234 defines two important headers: \httph{Expires}, whose value is
+a date and time of when the response is considered stale, and
+therefore should not be used; and \httph{Cache-Control}, which allows
+detailed control of the cache, including a \httph{max-age} field,
+which gives the time in seconds for how long the the response may be
+used from the time of the request. \httph{max-age} takes precedence
+over \httph{Expires}. In the following \emph{freshness lifetime} is
+understood as the number of seconds that the response may be used
+without contacting the origin server. Ideally, the calculation of the
+freshness lifetime should be based of the above, but it can also be
+based on heuristics, see Section~\ref{sec:simplefresh}.
+
+RFC7232, on the other hand, defines a protocol for asking the server
+if the cached response is still fresh using conditional requests. This
+doesn't burden the content provider with the task of estimating the
+time to live beforehand, but then, it must be able to answer if the
+resource has changed less expensively than serving the entire
+response. Either of two headers must be set by the server to achieve
+this: \httph{ETag}, which sets an opaque identifier for the response,
+or \httph{Last-Modified} which gives the time and date of the last
+modification of the resource. Clients may use these values to retrieve
+the updated response only if the resource has changed. Other relevant
+headers are described in Table~\ref{tab:headers}
+in Appendix~\ref{app:fetcher} but they were not all used in this
 analysis.\todo{Structured walkthrough}\todo{Impact}
+
+In this paper, we study to what extent SPARQL Endpoints, vocabulary
+and data publishers support these standards.
 

### 2015-01-19 sanity/paper1.tex
Explain refs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1002 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 The implications of HTTP for query caches has been discussed by
-\cite{kaseicache}. In \cite{sparqlproxy} the authors implemented a
-reverse proxy that controlled the changes to the dataset, and
-therefore could make sure the proxy had all the information needed to
-determine freshness. We are interested in the situation where the
-changes cannot be controlled.
+\cite{kaseicache}, where it is shown how changes can be tracked
+efficiently using different data structures. In \cite{sparqlproxy} the
+authors implemented a reverse proxy that controlled the changes to the
+dataset, and therefore could make sure the proxy had all the
+information needed to determine freshness. We are interested in the
+situation where the changes cannot be controlled.
 
-A general caching approach is given in \cite{yang2011caching}\todo{explain} . In
-\cite{umbrich2012hybrid}, the term caching was used in a different
+In \cite{umbrich2012hybrid}, the term caching was used in a different
 sense than we do, they rather prefetched an entire dataset to a local
@@
 explored when caching had a positive effect on complex SPARQL
-queries\todo{explain} .
+queries.
 

### 2015-01-19 sanity/paper1.tex
move more explanations to intro
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1003 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 without contacting the origin server. Ideally, the calculation of the
-freshness lifetime should be based of the above, but it can also be
-based on heuristics, see Section~\ref{sec:simplefresh}.
+freshness lifetime should be based of the above, we therefore shall
+refer to this as ``standards-compliant caching''. However it can also
+be based on heuristics, and RFC7234, Section~4.2.2 provides some loose
+constraints for such practice, as well as a suggestion for a useful
+heuristic. This heuristic is based on a fraction of the time lapsed
+between the current time and the modification time given in the
+\httph{Last-Modified} header. This approach therefore still requires
+the Web server to be cooperative to be successful, but commonly, Web
+servers can track this, for example if RDF is served from a file
+system, the file modification time is used.
 
@@
 in Appendix~\ref{app:fetcher} but they were not all used in this
-analysis.\todo{Structured walkthrough}\todo{Impact}
+analysis.
+
+RFC7234 provides detailed control of caching, and caching may also be
+prohibited by the server, either by setting a non-positive freshness
+lifetime or explicitly using a \httph{no-store} control field.
 
@@
 
-The two most important headers to control caching are \httph{Expires}
-and \httph{Cache-Control}\todo{move to intro}. As shown in Table~\ref{tab:headers}, there
-are also others. We found \httph{Pragma} (archaic HTTP 1.0 header) in
-287 responses, but except for two hosts, where they were superfluous,
-they were only used to prohibit caching. \httph{Surrogates} were not
-observed.
+As shown in Table~\ref{tab:headers}, there are also other caching
+headers than those mentioned in the introduction. We found
+\httph{Pragma} (archaic HTTP 1.0 header) in 287 responses, but except
+for two hosts, where they were superfluous, they were only used to
+prohibit caching. \httph{Surrogates} were not observed.
 
@@
 Of these resources, 405 returned valid headers, but 114 did so to
-\emph{prohibit} caching of the response, and 3 contained conflicting
+prohibit caching of the response, and 3 contained conflicting
 headers, i.e. set an freshness lifetime, but also prohibited
@@
 
-Caching may also be prohibited by setting a non-positive freshness
-lifetime. Common practice is to set the \httph{Expires} header to a
-date far back in time, which has also been done in large parts of the
-present dataset. For the following, we have set any non-positive
-lifetime to 0, as that is semantically equivalent.
+Common practice when prohibiting caching is to set the \httph{Expires}
+header to a date far back in time, which has also been done in large
+parts of the present dataset. For the following, we have set any
+non-positive lifetime to 0, as that is semantically equivalent.
 
@@
 
-We next consider the simple heuristic freshness lifetime. This is
-implemented as suggested in RFC7234, Section~4.2.2. This is based on a
-fraction of the time lapsed between the current time and the
-modification time given in the \httph{Last-Modified} header. This
-approach therefore still requires the Web server to be cooperative to
-be successful, but commonly, Web servers can track this, for example
-if RDF is served from a file system, the file modification time is used.
+We next consider the simple heuristic freshness lifetime as suggested
+RFC7234, Section~4.2.2 and detailed in the introduction.
 
@@
 determining a heuristic freshness lifetime, as the latest date may be
-used.\todo{Actually calculate lifetimes based on this and compare}
+used.
 

### 2015-01-19 sanity/paper1.tex
Future work and recommendation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1004 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+Even though it is clear that other parts of the Web benefit greatly
+from caching, and that the potential for HTTP metadata to better
+reflect actual update practices in the Semantic Web is great,
+estimating the actual impact of doing so should be done.
+
 \subsection{Recommendations}
@@
 and should be done by data owners. Framework authors should make it
-easy to set the expected lifetime.
+easy to set the expected lifetime, heeding the metadata association
+good practice recommendation of \cite{Jacobs:04:AWW}. If the author is
+unable to influence HTTP headers, they should set a
+\rdfterm{dct:valid} time into the future and make use of
+\rdfterm{dct:modified}.
 

### 2015-01-19 sanity/paper1.tex
Trim todos were not doing
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1005 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 exposes standards-compliant freshness time of 604800 seconds (i.e. 1
-week) for both LOD and SPARQL endpoints.\todo{Where and how}
+week) for both LOD and SPARQL endpoints.
 
@@
 amongst the servers that expose caching headers, it is common
-that the result may be cached for some minutes.\todo{more to point
-  out?}
+that the result may be cached for some minutes.
 
@@
 23 for vocabularies, 59 for generic information resources.\todo{table instead?}
-
-We already used \httph{Last-Modified} to compute heuristic lifetime
-in Section~\ref{sec:simplefresh} for 554 resources.\todo{really needed?}
 
@@
 classes. To do so, we classified based on certain predicates and
-classes, see table\todo{create table}. Additionally, URIs derived from
-prefix.cc were classified as ``vocabulary'' and those from SPARQLES as
-``endpoint''.
+classes. Additionally, URIs derived from prefix.cc were classified as
+``vocabulary'' and those from SPARQLES as ``endpoint''.
 
@@
 another request would be made to record the selected HTTP headers from
-that as well.\todo{say how many requests failed? It was a lot}
+that as well.
 \end{subappendices}

### 2015-01-19 sanity/paper1.tex
minor trim
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1006 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 to allow fine-grained control for proxies. \\ \hline
-\httph{Client-Aborted} & libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}  & Header inserted by User
+\httph{Client-Aborted} & libwww  & Header inserted by User
 Agent to indicate that it aborted the download \\ \hline
@@
 
-Based on the resulting headers, we first let libwww calculate the
-freshness lifetime in seconds if it was possible, and recorded
-that. Then, we also let it calculate a \emph{heuristic} freshness
-time based on a suggestion in RFC7234, Section~4.2.2, and recorded that.
+Based on the resulting headers, we let
+libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}
+calculate both standards-compliant and simple heuristic freshness
+lifetime.
 

### 2015-01-19 sanity/paper1.tex
Cut the qqplot
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1007 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
   \caption{\httph{Server} headers for hosts that enabled a freshness
-    lifetime to be computed for all requests.}\label{tab:servers}
+    lifetime to be computed for \emph{all} requests.}\label{tab:servers}
 \begin{center}
@@
 
-For a more systematic approach\todo{Move most to appendix}, we wish to test the hypothesis that
-some servers are better configured to support caching than others. To
-test this hypothesis, we create two so-called \emph{contigency
-  tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the
-distribution of  \httph{Server} headers are different for those implementations that
-offer caching headers from those that don't. Intuitively, we expect
-these distributions to be similar, ``long-tail'' distributions, i.e. a
-handful of servers are used by a large number of projects, and then it
-falls off rapidly, and so, some servers are used only by very
-few. Likewise, it is to be expected that only a few projects have
-given caching enough attention, but that they account for the majority
-of the support. The question is if the presence of caching headers is
-merely a matter of proportion, or if there are some that have given it
-more attention, but still is in relatively little use.
-
-Using the statistics system R\cite{kn:r}, we use a statistical test,
-namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000
-replicates). The simulation is done using a Monte Carlo method to
-compensate for the fact that many servers will not expose caching
-headers at all, an issue that would otherwise violate the underlying
-assumptions of the test.
-
-In both the cases of standard-compliant freshness time and for the
-other promising features, the test reports \pvalue = 0.0001. We can
-conclude that it is highly likely that some servers are better at
-exposing cache headers than others. Unfortunately, since most
-\httph{Server} headers only contain generic values, little can be
-learnt about these implementations. We note, however, that DBPedia
-exposes standards-compliant freshness time of 604800 seconds (i.e. 1
-week) for both LOD and SPARQL endpoints.
+For a more systematic approach, we wish to test the hypothesis that
+some servers are better configured to support caching than others.
+Using the methodology given in Appendix~\ref{app:stats}, we find in
+both the cases of standard-compliant freshness time and for the other
+promising features, the test reports \pvalue = 0.0001. We can conclude
+that it is highly likely that some servers are better at exposing
+cache headers than others. Unfortunately, since most \httph{Server}
+headers only contain generic values, little can be learnt about these
+implementations. We note, however, that DBPedia exposes
+standards-compliant freshness time of 604800 seconds (i.e. 1 week) for
+both LOD and SPARQL endpoints.
 
@@
 
-Since web servers may be configured to instruct clients and proxies to
-cache errors differently\todo{cut most to just state stuff} from
-successfully served requests, we should first investigate whether we
-can simply use all cache headers in the analysis, or if we should
-remove unsuccessful responses to avoid introducing bias to the
-analysis of valid responses.
-
-For a visual inspection, we may use a quantile-quantile plot to plot
-the freshness lifetimes of successful responses versus requests that
-failed for some reason. If the distribution of both variables are the
-same, these points will lie on a straight line. In
-Figure~\ref{fig:errorsqq}, we see successful responses vs. the most
-common parse error, and it is clear that these are so different that
-no further formal analysis is necessary, we must filter unsuccessful
-requests in the following analysis.
-
-\begin{figure}[ht!]
-  \centerline{%
-    \includegraphics[width=.5\textwidth]{errorsqq.pdf}}
-  \caption{A quantile-quantile plot of freshness lifetimes for
-    standard compliant headers with and without errors.}
-  \label{fig:errorsqq}
-\end{figure}
+Since we are interested in the properties of valid response, including
+examining some of the RDF contained in them, and that web servers may
+be configured to instruct clients and proxies to cache errors
+differently, we will study the statistical properties of valid
+responses.
 
@@
 that as well.
+
+\section{Statistical method}\label{app:stats}
+
+The hypothesis tests in this paper is implemented using \emph{contigency
+  tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the
+distribution of  \httph{Server} headers are different for those implementations that
+offer caching headers from those that don't. Intuitively, we expect
+these distributions to be similar, ``long-tail'' distributions, i.e. a
+handful of servers are used by a large number of projects, and then it
+falls off rapidly, and so, some servers are used only by very
+few. Likewise, it is to be expected that only a few projects have
+given caching enough attention, but that they account for the majority
+of the support. The question is if the presence of caching headers is
+merely a matter of proportion, or if there are some that have given it
+more attention, but still is in relatively little use.
+
+Using the statistics system R\cite{kn:r}, we use a statistical test,
+namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000
+replicates). The simulation is done using a Monte Carlo method to
+compensate for the fact that many servers will not expose caching
+headers at all, an issue that would otherwise violate the underlying
+assumptions of the test.
+
 \end{subappendices}

### 2015-01-19 sanity/paper1.tex
Acknowledgements
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1008 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 according to the standard, no cases for endpoints, 3 for datasets and
-23 for vocabularies, 59 for generic information resources.\todo{table instead?}
+23 for vocabularies, 59 for generic information resources.
 
@@
 \paragraph*{Acknowledgements} The author would like to thank Jürgen
-Umbrich and Axel Polleres for inspiring me to start this study, to
-Martin Giese for careful review and critical comments, and Gregory
-Todd Williams for promptly solving issues in the underlying libraries
-used in this study.
+Umbrich and Martin Giese for careful review and critical comments, and
+Gregory Todd Williams for promptly solving issues in the underlying
+libraries used in this study, and Axel Polleres for encouraging comments.
 

### 2015-01-19 sanity/paper1.tex
Remove URIs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1009 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\subtitle{---Unpublished Working Draft, do not circulate---}
+%\subtitle{---Unpublished Working Draft, do not circulate---}
 
@@
 the Perl framework
-RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011} was used
-to parse the line. Lines that failed to parse was discarded. We have
-not investigated whether this could introduce biases. Statements were
-then accepted into a new NQuad file if they had a predicate that
-matched the
-\rdfterm{sd:endpoint}\footnote{http://www.w3.org/TR/sparql11-service-description/\#sd-endpoint}
-or matched a case-insensitive regular expression \texttt{sparql} if
-the subject and objects were both resources, or the predicates
-\rdfterm{void:vocabulary}\footnote{http://www.w3.org/TR/void/\#vocabularies},
-\rdfterm{rdfa:vocabulary}\footnote{http://www.w3.org/TR/rdfa-core/\#term-prefix-definitions-vocabulary}
-or
-\rdfterm{api:vocabulary}\footnote{http://purl.org/linked-data/api/vocab\#vocabulary},
-as well as having a resource as object. Finally, statements with the
-the classes
-\rdfterm{cogs:Endpoint}\footnote{http://vocab.deri.ie/cogs\#Endpoint},
-\rdfterm{owl:Ontology} and
-\rdfterm{voaf:Vocabulary}\footnote{http://lov.okfn.org/vocab/voaf/v2.3/index.html\#Vocabulary}
-in the object position were also accepted. More classes and properties
-were considered, but not used in the data reduction if they did not
-occur in the original data.
+RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011}
+was used to parse the line. Lines that failed to parse were
+discarded. We have not investigated whether this could introduce
+biases. Statements were then accepted into a new NQuad file if they
+had a predicate that matched the \rdfterm{sd:endpoint} or matched a
+case-insensitive regular expression \texttt{sparql} if the subject and
+objects were both resources, or the predicates
+\rdfterm{void:vocabulary}, \rdfterm{rdfa:vocabulary}, or
+\rdfterm{api:vocabulary}, as well as having a resource as
+object. Finally, statements with the the classes
+\rdfterm{cogs:Endpoint}, \rdfterm{owl:Ontology} and
+\rdfterm{voaf:Vocabulary} in the object position were also
+accepted\footnote{LOV may be used to resolve these prefixed
+  names}. More classes and properties were considered, but not used in
+the data reduction if they did not occur in the original data.
 

### 2015-01-19 sanity/paper1.tex
Appendix pass
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1010 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 spider would go through the list of URLs found by previous steps for
-that host. Since the BTC2014 recorded
-the \httph{Expires} (an RFC7234 caching header), \httph{Last-Modified}
-and \httph{ETag} (RFC7232 conditional request) headers where they
-existed, we first examined if any of the resources were still fresh,
-but none were. Wherever the last two headers existed, we added the
+that host. Since the BTC2014 recorded the \httph{Expires},
+\httph{Last-Modified} and \httph{ETag} headers where they existed, we
+first examined if any of the resources were still fresh, but none
+were. Wherever the last two headers existed, we added the
 corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
@@
 conditional request showed that the BTC data were still fresh, and if
-it was, we retrieved the current data, as we had not couple the
+it was, we retrieved the current data, as we had not coupled the
 headers to the body in our original retrieval.

### 2015-01-19 sanity/paper1.tex
Trim pass
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1011 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The implications of HTTP for query caches has been discussed by
-\cite{kaseicache}, where it is shown how changes can be tracked
-efficiently using different data structures. In \cite{sparqlproxy} the
-authors implemented a reverse proxy that controlled the changes to the
-dataset, and therefore could make sure the proxy had all the
-information needed to determine freshness. We are interested in the
-situation where the changes cannot be controlled.
+In \cite{sparqlproxy} the authors implemented a reverse proxy that
+controlled the changes to the dataset, and therefore could make sure
+the proxy had all the information needed to determine freshness. We
+are interested in the situation where the changes cannot be
+controlled.
 
@@
 We next consider the simple heuristic freshness lifetime as suggested
-RFC7234, Section~4.2.2 and detailed in the introduction.
+RFC7234, Section~4.2.2 and mentioned in the introduction.
 

### 2015-01-19 sanity/paper1.tex
Remove server table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1012 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 For each unique \httph{Server} header, we found the ones where
-\emph{all} responses had a freshness time or a promising feature. For
-the former, this amounted to 22 servers, see 
-Table~\ref{tab:servers}. 70 servers always responded with promising
-features. As we can see, apart from well-known Virtuoso and
-Callimachus servers, as well as the Perl modules RDF::LinkedData and
-RDF::Endpoint, which is partly developed by and runs on a server
-operated by this author, most headers tell very little about the
-RDF-specific parts of the underlying server implementation. A quick
-inspection of all \httph{Server} headers confirmed that few reveal any
-further detail.
-
-% latex table generated in R 2.15.1 by xtable 1.5-6 package
-% Mon Jan  5 11:27:35 2015
-\begin{table}[h]
-  \caption{\httph{Server} headers for hosts that enabled a freshness
-    lifetime to be computed for \emph{all} requests.}\label{tab:servers}
-\begin{center}
-\begin{tabular}{rl}
-  \hline
-1 & DFE/largefile \\ 
-  2 & git\_frontend \\ 
-  3 & nginx/1.3.9 \\ 
-  4 & thin 1.6.0 codename Greek Yogurt \\ 
-  5 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
-  6 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
-  7 & TwistedWeb/8.2.0 \\ 
-  8 & RDF::Endpoint/0.07 \\ 
-  9 & Jetty(6.1.26) \\ 
-  10 & nginx/1.6.1 \\ 
-  11 & Jigsaw/2.3.0-beta3 \\ 
-  12 & Apache/2.2.9 (Win32) PHP/5.2.6 \\ 
-  13 & Apache/2.4.10 (Unix) mod\_fcgid/2.3.9 \\ 
-  14 &  \\ 
-  15 & GFE/2.0 \\ 
-  16 & RDF::LinkedData/0.70 \\ 
-  17 & Apache/2.2.17 (Unix) mod\_wsgi/3.3 Python/2.6.6 \\ 
-  18 & Virtuoso/07.10.3211 (Linux) i686-generic-linux-glibc212-64  VDB \\ 
-  19 & Apache/2.2.24 (Unix) mod\_ssl/2.2.24 OpenSSL/0.9.8y \\ 
-  20 & Apache/2.2.22 (Fedora) \\ 
-  21 & INSEE \\ 
-  22 & GitHub.com \\ 
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
+\emph{all} responses had a freshness time or a promising\todo{what are
+  they} feature. For the former, this amounted to 22 servers. 70
+servers always responded with promising features. Inspecting the
+values we find the well-known Virtuoso and Callimachus servers, as
+well as the Perl modules RDF::LinkedData and RDF::Endpoint, which is
+partly developed by and runs on a server operated by this
+author. Apart from those, a typical server string is
+e.g. \httph{Apache/2.2.22 (Fedora)}, which reveals very little about
+the RDF-specific parts of the underlying server implementation. A
+quick inspection of all \httph{Server} headers confirmed that few
+reveal any further detail.
 

### 2015-01-19 sanity/paper1.tex
Improve methodology; promising voes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1013 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 headers are described in Table~\ref{tab:headers}
-in Appendix~\ref{app:fetcher} but they were not all used in this
-analysis.
+in Appendix~\ref{app:fetcher}.
 
@@
 \section{Methodology}
-\todo{Improve, big}
 We want to find information resources on the Web, and examine HTTP
-headers that may allow caching. We record headers recommended by
-current standards, as well as obsoleted and non-standard
-headers. Additionally, we examine the triples in the returned
-information resources to see if there is information there that may be
-used to calculate heuristic freshness.
+headers that may allow caching. To do this, we perform GET requests on
+SPARQL endpoints, vocabularies, dataset descriptions and other
+resources and record headers recommended by current standards, as well
+as obsoleted and non-standard headers. Additionally, we examine the
+triples in the returned information resources to see if there is
+information there that may be used to calculate heuristic freshness.
 
-However, as the community has both spidered the Web as well as
-recorded lists of relevant resources, we need not do that ourselves.
+We make several approaches to ensure that we visit a large and
+representative section of the open Semantic Web. We take SPARQL
+Endpoints from the SPARQLES survey \cite{buil2013sparql}, vocabularies
+from Linked Open Vocabularies (LOV) \cite{lov2} and prefix.cc, and
+we augment these data with spidered data from the Billion Triple
+Challenge 2014 \cite{btc-2014} dataset.
 
-The SPARQLES survey\cite{buil2013sparql} has recorded the availability
-of SPARQL endpoints registered in the datahub.io portal, and found
-that many of them have ceased to operate. We used their list of SPARQL
-endpoints as of 2014-11-17, and filtered out those deemed unresponsive
-by them. This resulted in a list of 312 endpoints.
-
-To compile a list of Web ontologies, we first queried the Linked Open
-Vocabularies (LOV) \cite{lov2} SPARQL endpoint
-to retrieve all registered ontologies. Then, we took the full list of
-URIs registered with prefix.cc. The assumption that they all identify
-an ontology is clearly false, as we found that for example Tim
-Berners-Lee's FOAF profile has a registered URI, but it is unlikely to
-introduce significant bias.
+We used SPARQLES survey list of SPARQL endpoints as of 2014-11-17, and
+filtered out those deemed unresponsive by them. This resulted in a
+list of 312 endpoints.
 
@@
 consisted of a 4~GTriple corpus of spidered Web data. To compile a
-list of interesting candidates to further examine, we performed a
+list of candidates to further examine, we performed a
 series of data reduction steps, manually inspecting the result between
@@
 each several resources would be visited, some several times, as they
-may host SPARQL Endpoints, ontologies, or other information resources,
-by a spider detailed in Appendix~\ref{app:fetcher}, resulting in 7745
-requests, done on 2015-01-02.
+may host SPARQL Endpoints, vocabularies, or other information
+resources, by a spider detailed in Appendix~\ref{app:fetcher},
+resulting in 7745 requests, done on 2015-01-02. Although this is not a
+very large number, we assume that caching headers are usually set on a
+per-host basis, and so the number of hosts is more important than the
+number of resources.
 
@@
 Virtuoso-based SPARQL Endpoint for analysis by using the statistics
-system R\cite{kn:r}.
+system R\cite{kn:r} in the following section.
 
@@
 optional clauses matching the standard-compliant computed freshness
-time (which is the ideal) as well as whether the response had other
-promising features that could assist caching, modification time,
-certain predicates, etc.
+lifetime (which is the ideal) as well as whether the response had other
+features that may assist caching such as modification time,
+certain predicates, etc.\todo{promising}
 
 For each unique \httph{Server} header, we found the ones where
-\emph{all} responses had a freshness time or a promising\todo{what are
-  they} feature. For the former, this amounted to 22 servers. 70
-servers always responded with promising features. Inspecting the
-values we find the well-known Virtuoso and Callimachus servers, as
-well as the Perl modules RDF::LinkedData and RDF::Endpoint, which is
-partly developed by and runs on a server operated by this
-author. Apart from those, a typical server string is
-e.g. \httph{Apache/2.2.22 (Fedora)}, which reveals very little about
-the RDF-specific parts of the underlying server implementation. A
-quick inspection of all \httph{Server} headers confirmed that few
-reveal any further detail.
+\emph{all} responses had a freshness time or a promising feature. For
+the former, this amounted to 22 servers. 70 servers always responded
+with promising features. Inspecting the values we find the well-known
+Virtuoso and Callimachus servers, as well as the Perl modules
+RDF::LinkedData and RDF::Endpoint, which is partly developed by and
+runs on a server operated by this author. Apart from those, a typical
+server string is e.g. \httph{Apache/2.2.22 (Fedora)}, which reveals
+very little about the RDF-specific parts of the underlying server
+implementation. A quick inspection of all \httph{Server} headers
+confirmed that few reveal any further detail.
 
@@
 
-Of these resources, 405 returned valid headers, but 114 did so to
+Of the 2965 resources, 405 returned valid headers, but 114 did so to
 prohibit caching of the response, and 3 contained conflicting
@@
 
+% todo: kill this?
 Common practice when prohibiting caching is to set the \httph{Expires}
@@
 We then sought to classify resources into the categories ``endpoint''
-for SPARQL endpoints, ``vocabulary'' for ontologies, ``dataset'' for
+for SPARQL endpoints, ``vocabulary'' for vocabularies, ``dataset'' for
 datasets that may contain further descriptions of several resources,
@@
 classes. Additionally, URIs derived from prefix.cc were classified as
-``vocabulary'' and those from SPARQLES as ``endpoint''.
+``vocabulary'' (even if we found several that were not) and those from
+SPARQLES as ``endpoint''.
 

### 2015-01-19 sanity/paper1.tex
Prohibtion is 0
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1014 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 lifetime, i.e. to say that only private caches may use the response,
-that the cache must be revalidated, or to prohibit caching.
-
-% todo: kill this?
-Common practice when prohibiting caching is to set the \httph{Expires}
-header to a date far back in time, which has also been done in large
-parts of the present dataset. For the following, we have set any
-non-positive lifetime to 0, as that is semantically equivalent.
+that the cache must be revalidated, or to prohibit caching. Note that the
+freshness lifetime is 0 whenever caching is prohibited.
 

### 2015-01-19 sanity/paper1.tex
Prohibtion is 0
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1015 43ef1190-90fd-4dd7-a769-d0121f02b80e
Improve figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1016 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 lifetime, i.e. to say that only private caches may use the response,
-that the cache must be revalidated, or to prohibit caching. Note that the
-freshness lifetime is 0 whenever caching is prohibited.
+that the cache must be revalidated, or to prohibit caching. 
+Note that the freshness lifetime is 0 whenever caching is prohibited.
 
-In Figure~\ref{fig:hardall}\todo{axis and labels}, there is a barplot where
+In Figure~\ref{fig:hardall}, there is a barplot where
 the freshness lifetime is grouped in bins. We see that in these
@@
     \includegraphics[width=.9\textwidth]{hardall.pdf}}
-  \caption{Barplot of all standard-compliant freshness lifetimes
-    found. The first bin are the cases where caching is explicitly
+  \caption{Barplot counting all standard-compliant freshness lifetimes
+    found, with the percentage of occurances indicated on the
+    bars. The first bin are the cases where caching is explicitly
     prohibited. The next bins are for lifetimes, where the values are
@@
     \includegraphics[width=.9\textwidth]{heuristicall.pdf}}
-  \caption{Barplot of all simple heuristic freshness lifetimes
+  \caption{Barplot counting all simple heuristic freshness lifetimes
     found. Bins as in Figure~\ref{fig:hardall}.}

### 2015-01-19 sanity/queries.R

@@ -45,7 +45,7 @@ hardtable <- lifetimetable("other-hard.rq")
 hardall <- apply(hardtable, 1, sum)
 pdf(file="hardall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
-bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Frequency", main='')
+bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Number of occurences", main='')
 text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
 dev.off()
 
@@ -58,7 +58,7 @@ heuristictable <- lifetimetable("other-heuristic.rq")
 heuristicall <- apply(heuristictable, 1, sum)
 pdf(file="heuristicall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
-bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Frequency", main='')
+bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Number of occurances", main='')
 text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=1)
 dev.off()
 

### 2015-01-19 sanity/paper1.tex
Colors iin the mosaic plots; reordered table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1017 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
     \includegraphics[width=.9\textwidth]{hardtable.pdf}}
-  \caption{Mosaic Plot. Divided by type of resource on the vertical
-    axis and by the same bins as in Figure~\ref{fig:hardall} on the
-    horizontal axis. }
+  \caption{Mosaic Plot. On the vertical axis, the size of the boxes
+    are determined by the fraction of the types of resources. On the
+    horizontal axis the width of the boxes is proportional to the
+    total counts, using the same bins as in
+    Figure~\ref{fig:hardall}. Yellow boxes denote SPARQL Endpoints,
+    green dataset descriptions, blue generic information resources and
+    red vocabularies.}
   \label{fig:hardtable}
@@
     \includegraphics[width=.9\textwidth]{heuristictable.pdf}}
-  \caption{Mosaic Plot. Divided by type of resource on the vertical
-    axis and by the same bins as in Figure~\ref{fig:hardall} on the
-    horizontal axis. }
+  \caption{Mosaic Plot for heuristic freshness lifetime. See caption
+    of Figure~\ref{fig:hardtable} for description.  }
   \label{fig:heuristictable}

### 2015-01-19 sanity/queries.R

@@ -49,12 +49,15 @@ bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness life
 text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
 dev.off()
 
+heuristictable <- lifetimetable("other-heuristic.rq")
+hardtable2 <- cbind(hardtable[,3], hardtable[,1], hardtable[,4], hardtable[,2])
+colnames(hardtable2) <- colnames(heuristictable)
+
 pdf(file="hardtable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(hardtable, main='', xlab="Standards-compliant freshness lifetime", ylab="Frequency of resource types")
+mosaicplot(hardtable2, main='', xlab="Standards-compliant freshness lifetime", ylab="Frequency of resource types", color=c("red", "blue", "green", "yellow"))
 dev.off()
 
-heuristictable <- lifetimetable("other-heuristic.rq")
 heuristicall <- apply(heuristictable, 1, sum)
 pdf(file="heuristicall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
@@ -64,7 +67,7 @@ dev.off()
 
 pdf(file="heuristictable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Frequency of resource types")
+mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Frequency of resource types", color=c("red", "blue", "green", "yellow"))
 dev.off()
 
 harderrordata <- sparqlfile("failed-hard.rq")

### 2015-01-19 sanity/paper1.tex
remove todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1018 43ef1190-90fd-4dd7-a769-d0121f02b80e
15 pages now
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1019 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 We used SPARQLES survey list of SPARQL endpoints as of 2014-11-17, and
-filtered out those deemed unresponsive by them. This resulted in a
+filtered out those deemed unresponsive. This resulted in a
 list of 312 endpoints.
@@
 must also return a valid RDF media type (unless it is a SPARQL result)
-and the response must parse into an RDF model. We have given priority
-to survey many hosts since configuration usually doesn't differ much
-across a host, especially since it also captures different types of
-resources. It is therefore acceptable that the number of resources is
-relatively small.
+and the response must parse into an RDF model.
 
@@
 
-In Figure~\ref{fig:hardtable}\todo{look for others or explain, colors}, we have broken this
+In Figure~\ref{fig:hardtable}, we have broken this
 up by the type of resource that was accessed, i.e. SPARQL endpoints,
@@
 RDF::Generator::HTTP\footnote{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}
-to the Comprehensive Perl Archive Network. We recorded whether the
-conditional request showed that the BTC data were still fresh, and if
-it was, we retrieved the current data, as we had not coupled the
-headers to the body in our original retrieval.
-
-Based on the resulting headers, we let
+to CPAN. We recorded whether the conditional request showed that the
+BTC data were still fresh, and if it was, we retrieved the current
+data, as we had not coupled the headers to the body in our original
+retrieval.  Based on the resulting headers, we let
 libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}
@@
 
-If the initial response had RFC7232 headers, we make another
-request to see if the server just included the headers but does not
-actually support conditional requests. The heuristic we employed is
+If the initial response had RFC7232 headers, we made another
+request to see if the server included the headers but does not
+support conditional requests. The heuristic we employed is
 that if the headers remain the same, but the result was returned,
 rather than just a response code \httph{304} (which indicates that the
-previous result can be reused), then the server does most probably not
+previous result can be reused), the server does most probably not
 support it.
@@
 to the endpoints registered in the SPARQLES survey, our process found
-18~endpoints that responded with results.
-
-For all others, we parsed the response, and recorded any errors if the
-parser concluded the content were invalid.
+18~endpoints that responded with results. For all others, we parsed
+the response, and recorded any errors if the parser concluded the
+content were invalid.
 

### 2015-01-19 sanity/paper1.tex
change headline
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1020 43ef1190-90fd-4dd7-a769-d0121f02b80e
First version submitted, spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1021 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\title{A survey of HTTP caching options on the open Semantic Web}
+\title{A survey of HTTP caching implementations on the open Semantic Web}
 \author{Kjetil Kjernsmo}
@@
 features that may assist caching such as modification time,
-certain predicates, etc.\todo{promising}
+certain predicates, etc.%\todo{promising}
 
@@
   \caption{Barplot counting all standard-compliant freshness lifetimes
-    found, with the percentage of occurances indicated on the
+    found, with the percentage of occurrences indicated on the
     bars. The first bin are the cases where caching is explicitly
@@
 many resources change at a very slow pace, but also that this is not
-reflected in the standards-compliant freshness lifetimes advertized by
+reflected in the standards-compliant freshness lifetimes advertised by
 servers.
@@
 been seen, and since it is supported by DBPedia, benefits from caching
-may be realized already.
+may be realised already.
 
@@
 owners. It must necessarily involve the human users involved in the
-publishing cycle since they are making a committment about future
+publishing cycle since they are making a commitment about future
 changes. Designing user support systems as well as interfaces that fit

### 2015-01-19 sanity/paper1.tex
Final proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1022 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 very large number, we assume that caching headers are usually set on a
-per-host basis, and so the number of hosts is more important than the
+per-host basis, if any differences exist, they are likely on different
+types, and so the number of hosts is more important than the
 number of resources.
@@
 
-First, we investigated whether certain server implementations provided better support
-for caching than others. To do this, we formulated SPARQL queries to
-examine the \httph{Server} headers of successful responses, with
-optional clauses matching the standard-compliant computed freshness
-lifetime (which is the ideal) as well as whether the response had other
-features that may assist caching such as modification time,
-certain predicates, etc.%\todo{promising}
+First, we investigated whether certain server implementations provided
+better support for caching than others. To do this, we formulated
+SPARQL queries to examine the \httph{Server} headers of successful
+responses, with optional clauses matching the standard-compliant
+computed freshness lifetime (which is the ideal) as well as whether
+the response had other promising features that may assist caching such
+as modification time, certain predicates, etc.%\todo{promising}
 
 For each unique \httph{Server} header, we found the ones where
-\emph{all} responses had a freshness time or a promising feature. For
+\emph{all} responses had a freshness lifetime or a promising feature. For
 the former, this amounted to 22 servers. 70 servers always responded
@@
 Using the methodology given in Appendix~\ref{app:stats}, we find in
-both the cases of standard-compliant freshness time and for the other
+both the cases of standard-compliant freshness lifetime and for the other
 promising features, the test reports \pvalue = 0.0001. We can conclude
@@
 implementations. We note, however, that DBPedia exposes
-standards-compliant freshness time of 604800 seconds (i.e. 1 week) for
+standards-compliant freshness lifetime of 604800 seconds (i.e. 1 week) for
 both LOD and SPARQL endpoints.
@@
 note that the difference between different types of resources is much
-smaller. This is confirmed by a $\chi^2$ test that yields \pvalue =
+smaller. This is confirmed by a hypothesis test that yields \pvalue =
 0.02.
@@
 operate within the loose constraints of RFC7234, Section~4.2.2, and
-should find it niche when the other techniques described in this paper
+should find its niche when the other techniques described in this paper
 are unavailable. Once this is done, the correctness of caching headers
-should be assessed, possibly using contigency tables.
+should be assessed, possibly using contigency tables similar to those
+in Appendix~\ref{app:stats}.
 
@@
 contain resources that has different characteristics. Since the
-different sources we surveyed overlap, this cannot be done with
-Pearson's $\chi^2$ test, but requires more sophisticated statistics.
+different sources we surveyed overlap, this cannot be done with the
+simple hypothesis test in this paper, but requires more sophisticated
+statistics.
 
@@
 Further investigate the suitability of the \rdfterm{dct:modified}
-property for estimating heuristic freshness.
+property for estimating heuristic freshness lifetime.
 
@@
 
-A change periodicity predicate should be standardized in VoID.
+A change periodicity predicate should be standardized in VoID \cite{voidnote}.
 

### 2015-01-19 sanity/queries.R

@@ -55,7 +55,7 @@ colnames(hardtable2) <- colnames(heuristictable)
 
 pdf(file="hardtable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(hardtable2, main='', xlab="Standards-compliant freshness lifetime", ylab="Frequency of resource types", color=c("red", "blue", "green", "yellow"))
+mosaicplot(hardtable2, main='', xlab="Standards-compliant freshness lifetime", ylab="Relative abundance of resource types", color=c("red", "blue", "green", "yellow"))
 dev.off()
 
 heuristicall <- apply(heuristictable, 1, sum)
@@ -67,7 +67,7 @@ dev.off()
 
 pdf(file="heuristictable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Frequency of resource types", color=c("red", "blue", "green", "yellow"))
+mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Relative abundance of resource types", color=c("red", "blue", "green", "yellow"))
 dev.off()
 
 harderrordata <- sparqlfile("failed-hard.rq")

### 2015-01-20 sanity/paper1.tex
Jonas proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1023 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \documentclass{llncs}
+\usepackage[english,british]{babel}
 %\documentclass{article}
@@
 \usepackage{graphicx}
-%\usepackage{wrapfig}
+\usepackage{hyperref}
 \usepackage{appendix}
@@
 \httph{Last-Modified} header. This approach therefore still requires
-the Web server to be cooperative to be successful, but commonly, Web
+the Web server to be cooperative to be successful. Commonly, Web
 servers can track this, for example if RDF is served from a file
-system, the file modification time is used.
+system the file modification time is used.
 
@@
 caching does not necessarily give tangible benefits. Yet, we shall
-assume that sharing the meta data required for caching outside of the
+assume that sharing the metadata required for caching outside of the
 server is desirable, and that it is possible in most cases. We shall
@@
 we augment these data with spidered data from the Billion Triple
-Challenge 2014 \cite{btc-2014} dataset.
+Challenge (BTC) 2014 \cite{btc-2014} dataset.
 
@@
 SPARQL queries to examine the \httph{Server} headers of successful
-responses, with optional clauses matching the standard-compliant
+responses, with optional clauses matching the standards-compliant
 computed freshness lifetime (which is the ideal) as well as whether
@@
 Using the methodology given in Appendix~\ref{app:stats}, we find in
-both the cases of standard-compliant freshness lifetime and for the other
+both the cases of standards-compliant freshness lifetime and for the other
 promising features, the test reports \pvalue = 0.0001. We can conclude
@@
 prohibit caching of the response, and 3 contained conflicting
-headers, i.e. set an freshness lifetime, but also prohibited
+headers, i.e. set a freshness lifetime, but also prohibited
 caching. In most cases, \httph{Cache-Control} and \httph{Expires} both
@@
     \includegraphics[width=.9\textwidth]{hardall.pdf}}
-  \caption{Barplot counting all standard-compliant freshness lifetimes
+  \caption{Barplot counting all standards-compliant freshness lifetimes
     found, with the percentage of occurrences indicated on the
@@
 
-We also note that prohibition of SPARQL results caching is rare,
-amongst the servers that expose caching headers, it is common
+We also note that prohibition of SPARQL results caching is rare.
+Amongst the servers that expose caching headers it is common
 that the result may be cached for some minutes.
@@
 
-We next consider the simple heuristic freshness lifetime as suggested
+We next consider the simple heuristic freshness lifetime as suggested in
 RFC7234, Section~4.2.2 and mentioned in the introduction.
@@
 
-We found that errors are commonplace, but that they does not usually
+We found that errors are commonplace, but that they do not usually
 pertain to the caching headers. We found a small number of
 self-contradictory \httph{Cache-Control} headers, and some servers
-that set conditional request response headers, that could not support
+that set conditional request response headers, but could not support
 conditional requests.
@@
 
-In spite of this, most of the Semantic Web is unhelpful in spite of
-being slowly changing.
+In spite of this, most of the Semantic Web is unhelpful in even
+though it is changing slowly.
 
@@
 are unavailable. Once this is done, the correctness of caching headers
-should be assessed, possibly using contigency tables similar to those
+should be assessed, possibly using contingency tables similar to those
 in Appendix~\ref{app:stats}.
@@
 
-In many cases, setting a reasonable cache headers is straightforward,
+In many cases, setting reasonable cache headers is straightforward,
 and should be done by data owners. Framework authors should make it
@@
 Gregory Todd Williams for promptly solving issues in the underlying
-libraries used in this study, and Axel Polleres for encouraging comments.
+libraries used in this study, Axel Polleres for encouraging comments
+and Jonas Smedegaard for proofreading.
 
@@
 Mediawiki, we assume that they are configured with a single setup, and
-thus we will merely sample these resources.
+thus we merely sampled these resources.
 
@@
 platforms Livejournal and SAPO, where each blog has their own host and
-they expose FOAF data. For those, we only kept one hostname.
+they expose FOAF data. For those, we only kept one hostname, since
+they are likely configured similarly.
 
@@
 to CPAN. We recorded whether the conditional request showed that the
-BTC data were still fresh, and if it was, we retrieved the current
+BTC2014 data were still fresh, and if it was, we retrieved the current
 data, as we had not coupled the headers to the body in our original
@@
 
-The hypothesis tests in this paper is implemented using \emph{contigency
+The hypothesis tests in this paper is implemented using \emph{contingency
   tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the

### 2015-01-20 sanity/queries.R
Typo, ready to submit
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1024 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -61,7 +61,7 @@ dev.off()
 heuristicall <- apply(heuristictable, 1, sum)
 pdf(file="heuristicall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
-bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Number of occurances", main='')
+bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Number of occurrences", main='')
 text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=1)
 dev.off()
 

### 2015-02-18 sanity/rebuttal.txt
A start on the rebuttals
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1026 43ef1190-90fd-4dd7-a769-d0121f02b80e

I would like to thank the reviewers for their extensive reviews.

I would first like to address the main point of critisism voiced
mainly by Review 4 but also by Review 2, that the survey is of a very
limited "Semantic Web" and so is unlikely to be generalizeable.

This critisism is very valid and taken to heart. Indeed, this survey
is not meant to be generalizeable beyond the current state of the art,
as I find it quite likely that the current state must change in order
for the Semantic Web to be successful. The scope of the survey is a
snapshot, confined to what this community can reasonably do something
about, and currently, we are such a tiny fraction of the Web, that we
have to understand the Web better to succeed.

This survey is not intended as means to understand what the rest of
the Web is doing right and wrong, that wisdom is available elsewhere
in the broader literature, but as I tried not to pass judgement,
except in a small number of cases where current practice is clearly
severely flawed, it was not within my scope, nor the pagelimit, to
cite.

The intention is mostly to have empirical research to counter
statements like "it is no point in using caches when evaluating SPARQL
queries, since no endpoints are exposing them".

### 2015-02-18 sanity/rebuttal.txt
Discuss generalization
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1027 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-This critisism is very valid and taken to heart. Indeed, this survey
-is not meant to be generalizeable beyond the current state of the art,
-as I find it quite likely that the current state must change in order
-for the Semantic Web to be successful. The scope of the survey is a
-snapshot, confined to what this community can reasonably do something
-about, and currently, we are such a tiny fraction of the Web, that we
-have to understand the Web better to succeed.
+This critisism is very valid and taken to heart. To understand the
+constraints of the study, it is first important to note that it is
+directed at the community that convenes at ESWC. Also, such a paper
+could address what this community does, what it can do, and what it
+should do. We have consciously avoided strong "should do"
+recommendations, except in the case where current practice is clearly
+misguided. Most recommendations are therefore of the much softer "can
+do" type, but the overall focus of the paper is on what the community
+does.
+
+Consequently, this survey is not meant to be generalizeable beyond the
+current state of the art. I find it quite likely that the current
+state must change in order for the Semantic Web to be successful, and
+if the Semantic Web succeeds, the current conclusions will most
+certainly not hold. The scope of the survey is a snapshot, confined to
+what this community can reasonably do something about, and currently,
+we are such a tiny fraction of the Web, that we have to understand the
+Web better to succeed.
 
@@
 the Web is doing right and wrong, that wisdom is available elsewhere
-in the broader literature, but as I tried not to pass judgement,
-except in a small number of cases where current practice is clearly
-severely flawed, it was not within my scope, nor the pagelimit, to
-cite.
+in the broader literature, but since I believe a survey paper should
+focus mostly on "does", it was not within my scope, nor the pagelimit,
+to cite. This wisdom was therefore only cited from the standards.
 
-The intention is mostly to have empirical research to counter
+The intention is mostly to have empirical research to see if 
 statements like "it is no point in using caches when evaluating SPARQL
-queries, since no endpoints are exposing them".
+queries, since no endpoints are exposing them" have merit.
+
+Review 1 asked us to discuss the importance of caching metadata for
+query processing. We believe this is inappropriate for two reasons,
+one is the "does" focus of the paper, but more importantly, this is
+largely unknown at present. The study by Lampo et al, cited among
+related work, provides a sobering account, and cautions us to not
+oversell the importance of caching for query processing.
+
+Review 4 pointed out a flaw in that the Surrogates header was
+wrong. We acknowledge the error, and reran the spider recording the
+Surrogate-Capability header instead, but it was not found, and thus
+does not influence the conclusions.

### 2015-02-19 sanity/rebuttal.txt
we -> I
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1028 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 could address what this community does, what it can do, and what it
-should do. We have consciously avoided strong "should do"
+should do. I have consciously avoided strong "should do"
 recommendations, except in the case where current practice is clearly
@@
 Review 1 asked us to discuss the importance of caching metadata for
-query processing. We believe this is inappropriate for two reasons,
+query processing. I believe this is inappropriate for two reasons,
 one is the "does" focus of the paper, but more importantly, this is
@@
 
+As this is an broad area of critisism from several reviews, I
+shall attempt to condense the above discussion into a concise
+clarification in the paper.
+
 Review 4 pointed out a flaw in that the Surrogates header was
-wrong. We acknowledge the error, and reran the spider recording the
+wrong. I acknowledge the error, and reran the spider recording the
 Surrogate-Capability header instead, but it was not found, and thus
-does not influence the conclusions.
+does not influence the conclusions, but I'll fix this.

### 2015-02-19 sanity/rebuttal.txt
some review 6 things
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1029 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 does not influence the conclusions, but I'll fix this.
+
+I note that the paper is very close to the page limit, so I would have
+to remove parts of the paper to be able to expand on other parts. In
+fact, I removed discussions relevant to the questions raised by Review
+3 from the initial draft as they felt to deviate from the focus too
+much.
+
+Another example of things I removed was a list of the contributions,
+as requested by Review 6, as I felt it was merely a reiteration of the
+objectives that didn't not serve to clarify the paper, for any other
+reason than convention.
+
+Review 6 suggests that I use a logarithmic scale in some of the
+figures. I initially did this, but felt that they didn't effectively
+convey the essence, and that the problem is that our time and calendar
+doesn't follow linear progression either. Therefore, one is left with
+the choice of a classification that doesn't make intuitively sense or
+one that doesn't make sense mathematically. I'll try out the
+suggestion of combining the two.
+
+I'll discuss these issues with local faculty to find a good solution.
+
+Review 6 also suggests that I restructure the paper, and focus the
+discussion. With a focused discussion, it may be room to do some
+expansions of other points. I'll try my best, but it does not seem
+straightforward to include all the elaborations requested by the
+reviewers within the page limit.
+
+Details:
+
+Review 6 requests data published. I shall publish the data and the
+code in a tar archive.

### 2015-02-19 sanity/rebuttal.txt
minor
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1030 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Review 1 asked us to discuss the importance of caching metadata for
-query processing. I believe this is inappropriate for two reasons,
-one is the "does" focus of the paper, but more importantly, this is
-largely unknown at present. The study by Lampo et al, cited among
-related work, provides a sobering account, and cautions us to not
-oversell the importance of caching for query processing.
+query processing. I believe this is inappropriate for two reasons, one
+is the "does" focus of the paper, but more importantly, this is
+largely unknown at present, and it cannot be inferred from a survey
+alone. The study by Lampo et al, cited among related work, provides a
+sobering account, and cautions us to not oversell the importance of
+caching for query processing.
 
@@
 
-Review 4 pointed out a flaw in that the Surrogates header was
-wrong. I acknowledge the error, and reran the spider recording the
-Surrogate-Capability header instead, but it was not found, and thus
-does not influence the conclusions, but I'll fix this.
+Review 4 pointed out a flaw in that the Surrogates header was wrong. I
+acknowledge the error, and reran the spider recording the
+Surrogate-Capability header instead. It was not found, and therefore
+the flaw does not influence the conclusions, but I'll fix this.
 

### 2015-02-19 sanity/rebuttal.txt
details
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1031 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Details:
+I'm not sure how to address the comment of Review 1 that the survey is
+weak because it is based on only two documents. I assume they refer to
+the two RFCs. First, it is not entirely true, it also references other
+documents, only to show it is not widely used. Moreover, the two
+documents defines the entire standards-based caching infrastructure of
+the Web, as such, the survey is exhaustive.
 
-Review 6 requests data published. I shall publish the data and the
-code in a tar archive.
+Review 6 requests data to be published. I shall publish the data and
+the code in a tar archive.
+
+The reviewers provided detailed editorial comments that I shall take
+into account when creating a camera ready version. I'll try to find a
+native English writer to review the paper.

### 2015-02-21 sanity/rebuttal.txt
mention federated SPARQL
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1032 43ef1190-90fd-4dd7-a769-d0121f02b80e
improve English
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1033 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 if the Semantic Web succeeds, the current conclusions will most
-certainly not hold. The scope of the survey is a snapshot, confined to
-what this community can reasonably do something about, and currently,
-we are such a tiny fraction of the Web, that we have to understand the
-Web better to succeed.
+certainly not hold. In particular, there is nothing on the Web today
+that supports query evaluation in the way that a large-scale
+deployment of federated SPARQL would. The scope of the survey is a
+snapshot, confined to what this community can reasonably do something
+about, and currently, we are such a tiny fraction of the Web, that we
+have to understand the Web better to succeed.
 
@@
 the two RFCs. First, it is not entirely true, it also references other
-documents, only to show it is not widely used. Moreover, the two
-documents defines the entire standards-based caching infrastructure of
-the Web, as such, the survey is exhaustive.
+documents, only to show that they are not widely used. Moreover, the
+two documents defines the entire standards-based caching
+infrastructure of the Web, as such, the survey is exhaustive.
 

### 2015-02-21 sanity/rebuttal.txt
Shorten
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1034 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-This critisism is very valid and taken to heart. To understand the
-constraints of the study, it is first important to note that it is
-directed at the community that convenes at ESWC. Also, such a paper
-could address what this community does, what it can do, and what it
-should do. I have consciously avoided strong "should do"
-recommendations, except in the case where current practice is clearly
-misguided. Most recommendations are therefore of the much softer "can
-do" type, but the overall focus of the paper is on what the community
-does.
-
-Consequently, this survey is not meant to be generalizeable beyond the
-current state of the art. I find it quite likely that the current
-state must change in order for the Semantic Web to be successful, and
-if the Semantic Web succeeds, the current conclusions will most
-certainly not hold. In particular, there is nothing on the Web today
-that supports query evaluation in the way that a large-scale
-deployment of federated SPARQL would. The scope of the survey is a
-snapshot, confined to what this community can reasonably do something
-about, and currently, we are such a tiny fraction of the Web, that we
-have to understand the Web better to succeed.
-
-This survey is not intended as means to understand what the rest of
-the Web is doing right and wrong, that wisdom is available elsewhere
-in the broader literature, but since I believe a survey paper should
-focus mostly on "does", it was not within my scope, nor the pagelimit,
-to cite. This wisdom was therefore only cited from the standards.
-
-The intention is mostly to have empirical research to see if 
-statements like "it is no point in using caches when evaluating SPARQL
-queries, since no endpoints are exposing them" have merit.
+This critisism is very valid and taken to heart, and I should strive
+to make it clearer why this is the scope I chose to work within.
+Indeed, this Semantic Web is a mere fraction of the open data web,
+where e.g. JSON APIs are common. Coffee break discussions around
+caching have often stranded on the lack of data around what the
+community as a whole actually does, and therefore, that is the main
+focus of this paper. What it can do is given some space, but a
+discussion of what it should do is consciously avoided, except in the
+case where current practice is clearly misguided. Since the paper
+already fills the pagelimit, I have only cited the foundational
+documents for wisdom on what is thought to be important for the Web, a
+further discussion would be beyond the available space, and probably
+also be better suited for a different conference.
 
 Review 1 asked us to discuss the importance of caching metadata for
-query processing. I believe this is inappropriate for two reasons, one
-is the "does" focus of the paper, but more importantly, this is
+query processing. This is the focus our research, but this is
 largely unknown at present, and it cannot be inferred from a survey
@@
 
-As this is an broad area of critisism from several reviews, I
-shall attempt to condense the above discussion into a concise
-clarification in the paper.
+Review 4 pointed out a flaw in that the Surrogates header was wrong. I
+acknowledge the error, and will rerun the spider.
 
-Review 4 pointed out a flaw in that the Surrogates header was wrong. I
-acknowledge the error, and reran the spider recording the
-Surrogate-Capability header instead. It was not found, and therefore
-the flaw does not influence the conclusions, but I'll fix this.
+I note that the paper is very close to the page limit, and amongst the
+things that were removed from the initial draft are discussions of the
+questions raised by Review 3 and a list of the contributions, as
+requested by Review 6. Review 6 suggest that I focus the discussion,
+which could reduce the length somewhat, but I would welcome
+suggestions on what else could be trimmed, to make room for
+elaborations suggested in the reviews.
 
-I note that the paper is very close to the page limit, so I would have
-to remove parts of the paper to be able to expand on other parts. In
-fact, I removed discussions relevant to the questions raised by Review
-3 from the initial draft as they felt to deviate from the focus too
-much.
-
-Another example of things I removed was a list of the contributions,
-as requested by Review 6, as I felt it was merely a reiteration of the
-objectives that didn't not serve to clarify the paper, for any other
-reason than convention.
-
-Review 6 suggests that I use a logarithmic scale in some of the
-figures. I initially did this, but felt that they didn't effectively
-convey the essence, and that the problem is that our time and calendar
-doesn't follow linear progression either. Therefore, one is left with
-the choice of a classification that doesn't make intuitively sense or
-one that doesn't make sense mathematically. I'll try out the
-suggestion of combining the two.
-
-I'll discuss these issues with local faculty to find a good solution.
-
-Review 6 also suggests that I restructure the paper, and focus the
-discussion. With a focused discussion, it may be room to do some
-expansions of other points. I'll try my best, but it does not seem
-straightforward to include all the elaborations requested by the
-reviewers within the page limit.
+I initially used logarithmic scale as suggested by Review 6, for the
+reasons they mention. I found the intuitive interpretation of the
+figures to be very difficult, since the time and calendar scale is as
+it is. The suggestion to combine the two is interesting, and I will
+see if provides any benefit if the paper is accepted.
 

### 2015-02-21 sanity/rebuttal.txt
shorten
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1035 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
-I would like to thank the reviewers for their extensive reviews.
+I would like to thank the reviewers for their extensive reviews. The
+many editorial advices will certainly be taken into account without
+further comment.
 
-I would first like to address the main point of critisism voiced
-mainly by Review 4 but also by Review 2, that the survey is of a very
-limited "Semantic Web" and so is unlikely to be generalizeable.
-
-This critisism is very valid and taken to heart, and I should strive
-to make it clearer why this is the scope I chose to work within.
-Indeed, this Semantic Web is a mere fraction of the open data web,
-where e.g. JSON APIs are common. Coffee break discussions around
-caching have often stranded on the lack of data around what the
-community as a whole actually does, and therefore, that is the main
-focus of this paper. What it can do is given some space, but a
-discussion of what it should do is consciously avoided, except in the
-case where current practice is clearly misguided. Since the paper
-already fills the pagelimit, I have only cited the foundational
-documents for wisdom on what is thought to be important for the Web, a
-further discussion would be beyond the available space, and probably
-also be better suited for a different conference.
+The critisism that I only surveyed a small part of the Web and so, the
+conclusions are not generalizeable is valid, and I should strive to
+make it clearer why this is the scope I chose to work within. Indeed,
+this Semantic Web is a mere fraction of the open data web, where
+e.g. JSON APIs are common. Coffee break discussions around caching
+have often stranded on the lack of data around what the community as a
+whole actually does, and therefore, that is the main focus of this
+paper. What it can do is given some space, but a discussion of what it
+should do is consciously avoided, except in the case where current
+practice is clearly misguided. Since the paper already fills the
+pagelimit, I have only cited the foundational documents for wisdom on
+what is thought to be important for the Web, a further discussion
+would be beyond the available space.
 
@@
 Review 4 pointed out a flaw in that the Surrogates header was wrong. I
-acknowledge the error, and will rerun the spider.
+acknowledge the error, and have rerun the spider.
 
@@
 it is. The suggestion to combine the two is interesting, and I will
-see if provides any benefit if the paper is accepted.
+see if it provides any benefit if the paper is accepted.
 
@@
 weak because it is based on only two documents. I assume they refer to
-the two RFCs. First, it is not entirely true, it also references other
-documents, only to show that they are not widely used. Moreover, the
-two documents defines the entire standards-based caching
+the two RFCs. First, it is not entirely true, see Table 1. Moreover,
+the two documents define the entire standards-based caching
 infrastructure of the Web, as such, the survey is exhaustive.
-
-Review 6 requests data to be published. I shall publish the data and
-the code in a tar archive.
-
-The reviewers provided detailed editorial comments that I shall take
-into account when creating a camera ready version. I'll try to find a
-native English writer to review the paper.

### 2015-02-21 sanity/rebuttal.txt
shorten
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1036 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 this Semantic Web is a mere fraction of the open data web, where
-e.g. JSON APIs are common. Coffee break discussions around caching
-have often stranded on the lack of data around what the community as a
-whole actually does, and therefore, that is the main focus of this
-paper. What it can do is given some space, but a discussion of what it
-should do is consciously avoided, except in the case where current
-practice is clearly misguided. Since the paper already fills the
-pagelimit, I have only cited the foundational documents for wisdom on
-what is thought to be important for the Web, a further discussion
-would be beyond the available space.
+e.g. JSON APIs are common. ISWC/ESWC coffee break discussions around
+caching have often stranded on the lack of data around what the
+community as a whole actually does, and therefore, that is the main
+focus of this paper. The recommendations are meant as suggestions on
+what the community may do, but a discussion of what it should do is
+consciously avoided, except in the case where current practice is
+clearly misguided. Since the paper already fills the pagelimit, I have
+only cited the foundational documents for wisdom on what is thought to
+be important for the Web, a further discussion would be beyond the
+available space.
 
@@
 I initially used logarithmic scale as suggested by Review 6, for the
-reasons they mention. I found the intuitive interpretation of the
-figures to be very difficult, since the time and calendar scale is as
-it is. The suggestion to combine the two is interesting, and I will
+reasons they mention. Then, I found the intuitive interpretation of
+the figures to be very difficult, since the time and calendar scale is
+as it is. The suggestion to combine the two is interesting, and I will
 see if it provides any benefit if the paper is accepted.

### 2015-02-21 sanity/fetcher.pl
Surrogates were wrong
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1037 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -172,7 +172,7 @@ foreach my $host (@hosts) {
 																		 'ETag',
 																		 'X-Cache',
 																		 'Date',
-																		 'Surrogates',
+																		 'Surrogate-Capability',
 																		 'Client-Aborted',
 																		 'Client-Warning'
 																		],
@@ -323,7 +323,7 @@ foreach my $host (@hosts) {
 																				  'ETag',
 																				  'X-Cache',
 																				  'Date',
-																				  'Surrogates',
+																				  'Surrogate-Capability',
 																				  'Client-Aborted',
 																				  'Client-Warning'
 																				 ],

### 2015-02-21 sanity/paper1.tex
Useable metadata; reorder Section refs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1038 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 refer to this as ``standards-compliant caching''. However it can also
-be based on heuristics, and RFC7234, Section~4.2.2 provides some loose
+be based on heuristics, and Section~4.2.2 in RFC7234 provides some loose
 constraints for such practice, as well as a suggestion for a useful
@@
 computed freshness lifetime (which is the ideal) as well as whether
-the response had other promising features that may assist caching such
-as modification time, certain predicates, etc.%\todo{promising}
+the response had other indications of caching-related metadata that
+may assist caching such as modification time, certain predicates, etc.
 
 For each unique \httph{Server} header, we found the ones where
-\emph{all} responses had a freshness lifetime or a promising feature. For
+\emph{all} responses had a freshness lifetime or other usable metadata. For
 the former, this amounted to 22 servers. 70 servers always responded
-with promising features. Inspecting the values we find the well-known
+with usable metadata. Inspecting the values we find the well-known
 Virtuoso and Callimachus servers, as well as the Perl modules
@@
 both the cases of standards-compliant freshness lifetime and for the other
-promising features, the test reports \pvalue = 0.0001. We can conclude
+usable metadata, the test reports \pvalue = 0.0001. We can conclude
 that it is highly likely that some servers are better at exposing
@@
 We next consider the simple heuristic freshness lifetime as suggested in
-RFC7234, Section~4.2.2 and mentioned in the introduction.
+Section~4.2.2 in RFC7234 and mentioned in the introduction.
 
@@
 heuristic freshness lifetime estimation. However, such work should
-operate within the loose constraints of RFC7234, Section~4.2.2, and
+operate within the loose constraints of Section~4.2.2 in RFC7234, and
 should find its niche when the other techniques described in this paper

### 2015-02-21 sanity/paper1.tex
lc endpoint
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1039 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In this paper, we study to what extent SPARQL Endpoints, vocabulary
+In this paper, we study to what extent SPARQL endpoints, vocabulary
 and data publishers support these standards.
@@
 each several resources would be visited, some several times, as they
-may host SPARQL Endpoints, vocabularies, or other information
+may host SPARQL endpoints, vocabularies, or other information
 resources, by a spider detailed in Appendix~\ref{app:fetcher},
@@
 This results in an NQuads file per host, which is then loaded into a
-Virtuoso-based SPARQL Endpoint for analysis by using the statistics
+Virtuoso-based SPARQL endpoint for analysis by using the statistics
 system R\cite{kn:r} in the following section.
@@
     total counts, using the same bins as in
-    Figure~\ref{fig:hardall}. Yellow boxes denote SPARQL Endpoints,
+    Figure~\ref{fig:hardall}. Yellow boxes denote SPARQL endpoints,
     green dataset descriptions, blue generic information resources and
@@
 
-We find that only one SPARQL Endpoint yields a heuristic lifetime, on
+We find that only one SPARQL endpoint yields a heuristic lifetime, on
 closer inspection, we find this to be hosted by
@@
 
-For SPARQL Endpoints, it is interesting to examine the queries in a
+For SPARQL endpoints, it is interesting to examine the queries in a
 shared proxy, e.g. an institutional cache or a Content

### 2015-02-21 sanity/paper1.tex
Move citation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1040 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The \cite{dyldo2} Dynamic Linked Data Observatory (DyLDO) performed,
+The Dynamic Linked Data Observatory (DyLDO) \cite{dyldo2} performed,
 and continues to do so, monitoring of parts of the Linked Open Data

### 2015-02-21 sanity/paper1.tex
Some typo correction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1041 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In spite of this, most of the Semantic Web is unhelpful in even
-though it is changing slowly.
+In spite of this, most of the Semantic Web is unhelpful even though it
+is changing slowly.
 
@@
 Investigate whether curated collections such as LOV or SPARQLES
-contain resources that has different characteristics. Since the
+contain resources that have different characteristics. Since the
 different sources we surveyed overlap, this cannot be done with the
@@
 invalid or irrelevant, e.g. URIs that didn't have a scheme or where
-the scheme wasn't HTTP(S), or they were referring to private IP addresses.
+the scheme were't HTTP(S), or they were referring to private IP addresses.
 

### 2015-03-05 cold2014/magnus.tex
Added more on how the levels are set
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@756 43ef1190-90fd-4dd7-a769-d0121f02b80e
Added more on how the levels are set
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@757 43ef1190-90fd-4dd7-a769-d0121f02b80e
Remove figures
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1042 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 To determine the levels of each factor, we calculated the average value 
-of each factor present in the data sets. % TODO: more on this.
+of each factor present in the data sets. The two levels of each factor would then 
+represent either values over or under the average. This was done to hit 
+as many values in the data sets as possible. % TODO: more on this.
 
@@
 \textsf{-co}.
-
-\begin{figure}%{r}{0.65\textwidth}
-\begin{center}
-\vspace{-15pt}
- \includegraphics[width=8.5cm, height=7cm]{2factorfinal.pdf}
-  \caption{Resulting graph for the three factor experiment}\label{3factorGraph}
-\vspace{-10pt}
-\end{center}
-\end{figure}
 
@@
 \end{wraptable}
-
-\begin{figure}
- \includegraphics[width=14cm]{10factorAvarageDanielPlot.pdf}
-  \caption{Resulting graph for the ten factor experiment}\label{10factorGraph}
-\end{figure}
 

### 2015-03-05 cold2014/magnus.tex
Rewrite the text around figures in experiment setup
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1043 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \textsf{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
-from the experiments and output them in two types of graphs.
+from the experiments for further study.
 
-To evaluate the graphs we have used a method in the \textsf{FrF2} R
-package called \textsf{DanielPlot} to create half-normal plots. The
-intuition behind this is simple: If errors are normally distributed,
-which is a conservative assumption, and there are no significant
-effects they will also be normally distributed, and so the estimated
-effects would end up on a straight line. Any significant deviations
-from a straight line indicates a significant factor.  That we use a
-half-normal plot (rather than a normal plot) means that
-the most significant effects will end up in the top right
-corner. 
+The \textsf{FrF2} R package implements a criterion of significance
+known as Lenth's method, which we have used with a significance level at $\alpha =
+0.05$. % TODO: Kan slenge med referanse
 %
@@
 %
-The axes themselves are not important to understand the plot
-nor the results, refer to \cite{PlanExp} for details.  We have also
-set the significance level at $\alpha = 0.05$, where
-\textsf{DanielPlot} is using Lenth's method for the analysis of
-unreplicated factorials.
-
 For brevity, we will use abbreviations given in Table~\ref{facandabbrev}.
@@
 
-Figure~\ref{3factorGraph} shows only one significant factor, namely
-the \textsf{O} factor in the upper right corner. In
-Table~\ref{3factorEffect} we can see the time effects from largest to
-smallest.  We see that the ontology factor has the largest effect on
-approximately 2.2 seconds which means that one of the ontologies
-overall is 2.2 seconds faster. By reading the results from the linear
-model it is the factor level \textsf{-vso} which has an overall faster
-response time.
+Using the Lenth criterion, we find only one significant factor, namely
+the \textsf{O} factor. In Table~\ref{3factorEffect} we can see the
+time effects from largest to smallest.  We see that the ontology
+factor has the largest effect on approximately 2.2 seconds which means
+that one of the ontologies overall is 2.2 seconds faster. By reading
+the results from the linear model it is the factor level \textsf{-vso}
+which has an overall faster response time.
 
@@
 
-In Figure~\ref{10factorGraph} we can see the results from running
-all the 1024 runs.  We can see that there are a lot of significant
-effects present here. 
-Unfortunately there are so many significant factors that there are
-some of them that are illegible, but most of them have a low impact on
-the significance graph.  In Table~\ref{10factorEffect} we can see the top
-fifteen absolute effects. We can see that emission and fuel consumption are 
-the most significant factors besides the ontology factor.
-The linear model again tells us that the option
+There is a large number of significant effects when we employ Lenth's
+method on this large experiment. In Table~\ref{10factorEffect} we have
+included the top fifteen absolute effects. We can see that emission
+and fuel consumption are the most significant factors besides the
+ontology factor.  The linear model again tells us that the option
 \textsf{-vso} has a significantly lower response time than

### 2015-03-05 cold2014/magnus.tex
Rewrite the text around figures in discussion
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1044 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-In the Figure~\ref{3factorGraph} we can see that only one significant
-effects, the O factor. We saw in Table~\ref{3factorEffect} that the
+We noted that the O factor was the only significant factor. 
+We saw in Table~\ref{3factorEffect} that the
 Configuration Ontology was almost two seconds slower than the VSO/COO
@@
 
-In Figure~\ref{10factorGraph} we can see the results from this
-experiment. In Table
-\ref{10factorEffect} we see the top twenty absolute effects and
-similar to all the other experiments, O is on top of the list. The
+In Table \ref{10factorEffect} we see the top twenty absolute effects
+and similar to all the other experiments, O is on top of the list. The
 average response time of a run done against CO was $\approx58.5$

### 2015-03-10 sanity/paper1.tex
Address comment 4; on LOD cloud coverage
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1045 43ef1190-90fd-4dd7-a769-d0121f02b80e
Move appendices to report
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1046 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \documentclass{llncs}
+%\documentclass{sig-alternate}
 \usepackage[english,british]{babel}
@@
 noted that the Billion Triple Challenge 2014 \cite{btc-2014} dataset
-consisted of a 4~GTriple corpus of spidered Web data. To compile a
-list of candidates to further examine, we performed a
-series of data reduction steps, manually inspecting the result between
-each step. The details of this process is given in
-Appendix~\ref{app:reduction}.
+consisted of a 4~GTriple corpus of spidered Web data. This was seeded
+from datahub.io (aka CKAN), which is also the definition of the LOD
+cloud, as well as other sources. To compile a list of candidates to
+further examine, we performed a series of data reduction steps,
+manually inspecting the result between each step. The details of this
+process is given in Appendix~\ref{app:reduction}.
 
@@
 
-\begin{subappendices}
-\renewcommand{\thesection}{\Alph{section}}%
-
-\section{Implementation of data reduction}\label{app:reduction}
-
-The Billion Triple Challenge 2014 dataset provided data in the form of
-NQuad files. Due to the presence of invalid RDF, we iterated through
-the NQuad files on a line-by-line basis. First, we matched each line
-against a regular expression were lines matching
-\texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
-the Perl framework
-RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011}
-was used to parse the line. Lines that failed to parse were
-discarded. We have not investigated whether this could introduce
-biases. Statements were then accepted into a new NQuad file if they
-had a predicate that matched the \rdfterm{sd:endpoint} or matched a
-case-insensitive regular expression \texttt{sparql} if the subject and
-objects were both resources, or the predicates
-\rdfterm{void:vocabulary}, \rdfterm{rdfa:vocabulary}, or
-\rdfterm{api:vocabulary}, as well as having a resource as
-object. Finally, statements with the the classes
-\rdfterm{cogs:Endpoint}, \rdfterm{owl:Ontology} and
-\rdfterm{voaf:Vocabulary} in the object position were also
-accepted\footnote{LOV may be used to resolve these prefixed
-  names}. More classes and properties were considered, but not used in
-the data reduction if they did not occur in the original data.
-
-In the next step, we filtered out statements with URIs that were
-invalid or irrelevant, e.g. URIs that didn't have a scheme or where
-the scheme were't HTTP(S), or they were referring to private IP addresses.
-
-We then sought to classify resources into the categories ``endpoint''
-for SPARQL endpoints, ``vocabulary'' for vocabularies, ``dataset'' for
-datasets that may contain further descriptions of several resources,
-or simply ``inforesources'' for those that did not fit in the above
-classes. To do so, we classified based on certain predicates and
-classes. Additionally, URIs derived from prefix.cc were classified as
-``vocabulary'' (even if we found several that were not) and those from
-SPARQLES as ``endpoint''.
-
-Since we blatantly violated URI opacity with our regular expression
-matching in the first step, we needed to further filter candidates for
-SPARQL endpoints. This step therefore included filtering as well as
-classification.
-
-We found in the data a large number of ontologies that consist of many
-information resources with just a few triples in each. Since they
-appear to be produced by the same software, usually Semantic
-Mediawiki, we assume that they are configured with a single setup, and
-thus we merely sampled these resources.
-
-We continued to also sample the HTTP headers gathered in the BTC2014
-dataset. 
-%as they had recorded some of the relevant headers,
-%\httph{Expires}, \httph{Last-Modified} and \httph{ETag}. 
-First, we traversed the files with a simple UNIX \texttt{grep} to find
-the resources that had reported one of the RDF serializations as
-content type. We then traversed this list, first discarded the
-resources that did not have a valid IRI (this amounted to just 3273
-resources). For the resources we found, as well as for the resources
-that was of \rdfterm{rdf:type owl:Ontology} above, we kept one
-resource per hostname, with the exception of the popular blogging
-platforms Livejournal and SAPO, where each blog has their own host and
-they expose FOAF data. For those, we only kept one hostname, since
-they are likely configured similarly.
-
-\section{Implementation of spider}\label{app:fetcher}
-
-We then developed a fairly elaborate parallel spider to examine the
-resources found on hosts that the previous steps deemed interesting
-using the Perl frameworks RDF::Trine and libwww. The spider operated
-with a timeout of 20 seconds and a maximum message size it would
-accept of 1 MB.
-
-The parallel spider would then launch a process per host, but each
-request to one host would be delayed by 10 seconds. For each host, the
-spider would go through the list of URLs found by previous steps for
-that host. Since the BTC2014 recorded the \httph{Expires},
-\httph{Last-Modified} and \httph{ETag} headers where they existed, we
-first examined if any of the resources were still fresh, but none
-were. Wherever the last two headers existed, we added the
-corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
-headers for a conditional initial request.
-
-\begin{table}
-\caption{Recorded HTTP headers}\label{tab:headers}
- \begin{tabular}{ | l |  p{1.4cm} | p{6.8cm} |}
-    \hline
-    Header & Reference & Description \\ \hline
-\httph{Age} & RFC7234 & When obtaining response from a cache, the number of
-seconds since validation \\ \hline
-\httph{Cache-Control} & RFC7234 & Header used for a variety of directives \\ \hline
-\httph{Expires} & RFC7234  & Gives the date/time after which the
-   response is considered stale. \\ \hline
-\httph{Pragma} & RFC7234 & Archaic HTTP 1.0 header  \\ \hline
-\httph{Warning} & RFC7234  & For additional information about possible incorrectness \\ \hline
-\httph{Content-Type} & RFC7231 & To select the correct parser \\ \hline
-\httph{If-None-Match} & RFC7232  & Request header to check if
-                                   \httph{ETag} has changed   \\ \hline
-\httph{If-Modified-Since} & RFC7232  & Request header to check if
-                                   \httph{Last-Modified} has changed    \\ \hline
-\httph{Last-Modified} & RFC7232 & When the resource was last modified \\ \hline
-\httph{ETag} & RFC7232 & An opaque validator to check if the resource
-has changed  \\ \hline
-\httph{X-Cache} &  & Inserted by some caches to indicate cache status \\ \hline
-\httph{Date} & RFC7231 & The time of the message. Used in conditional requests
-and heuristics \\ \hline
-\httph{Surrogates} & Edge \cite{edgearch} & Draft
-to allow fine-grained control for proxies. \\ \hline
-\httph{Client-Aborted} & libwww  & Header inserted by User
-Agent to indicate that it aborted the download \\ \hline
-\httph{Client-Warning} & libwww  & Header inserted by User
-Agent to give details about problems with the download \\ \hline
-
-    \hline
-    \end{tabular}
-\end{table}
-
-For endpoints, we made the following SPARQL query:
-\begin{verbatim}
-SELECT DISTINCT ?Concept WHERE {[] a ?Concept} LIMIT 2
-\end{verbatim}
-which should be quite light, yet likely yield results.
-
-Then, the first request would be made, and a selection of the
-resulting HTTP headers recorded in an per-host NQuads file. For this
-purpose, we developed and released a module
-RDF::Generator::HTTP\footnote{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}
-to CPAN. We recorded whether the conditional request showed that the
-BTC2014 data were still fresh, and if it was, we retrieved the current
-data, as we had not coupled the headers to the body in our original
-retrieval.  Based on the resulting headers, we let
-libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}
-calculate both standards-compliant and simple heuristic freshness
-lifetime.
-
-If the initial response had RFC7232 headers, we made another
-request to see if the server included the headers but does not
-support conditional requests. The heuristic we employed is
-that if the headers remain the same, but the result was returned,
-rather than just a response code \httph{304} (which indicates that the
-previous result can be reused), the server does most probably not
-support it.
-
-For endpoints, we examined the response message, to see if there are
-any results to our query, and recorded that if there are. In addition
-to the endpoints registered in the SPARQLES survey, our process found
-18~endpoints that responded with results. For all others, we parsed
-the response, and recorded any errors if the parser concluded the
-content were invalid.
-
-For resource types other than ``vocabulary'', we look for SPARQL
-endpoints in the response, using the predicates \rdfterm{sd:endpoint}
-and \rdfterm{void:sparqlEndpoint}. We then do the same query as above
-and record the relevant headers. Unfortunately, we found early that
-this only turns up misconfigured endpoints that point to localhost,
-and was removed from the spider for the final analysis.
-
-Finally, if the LOV SPARQL endpoint used a URI for the vocabulary that
-was different from the namespace URI (after a normalization step),
-another request would be made to record the selected HTTP headers from
-that as well.
-
-\section{Statistical method}\label{app:stats}
-
-The hypothesis tests in this paper is implemented using \emph{contingency
-  tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the
-distribution of  \httph{Server} headers are different for those implementations that
-offer caching headers from those that don't. Intuitively, we expect
-these distributions to be similar, ``long-tail'' distributions, i.e. a
-handful of servers are used by a large number of projects, and then it
-falls off rapidly, and so, some servers are used only by very
-few. Likewise, it is to be expected that only a few projects have
-given caching enough attention, but that they account for the majority
-of the support. The question is if the presence of caching headers is
-merely a matter of proportion, or if there are some that have given it
-more attention, but still is in relatively little use.
-
-Using the statistics system R\cite{kn:r}, we use a statistical test,
-namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000
-replicates). The simulation is done using a Monte Carlo method to
-compensate for the fact that many servers will not expose caching
-headers at all, an issue that would otherwise violate the underlying
-assumptions of the test.
-
-\end{subappendices}
-
 \end{document}

### 2015-03-10 sanity/report.tex
Move appendices to report
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1046 43ef1190-90fd-4dd7-a769-d0121f02b80e

\usepackage[english,british]{babel}
%\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{appendix}

\newcommand{\rdfterm}[1]{\texttt{#1}}
\newcommand{\pvalue}{\textit{p}-value\ }
\newcommand{\httph}[1]{\texttt{#1}}
\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}

\title{A survey of HTTP caching implementations on the open Semantic Web}
\author{Kjetil Kjernsmo}
\institute{Department of Informatics,
Postboks 1080 Blindern,
N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}}

%\subtitle{---Unpublished Working Draft, do not circulate---}

\begin{document}
\maketitle

\section{Introduction}

\section{Implementation of data reduction}\label{app:reduction}

The Billion Triple Challenge 2014 dataset provided data in the form of
NQuad files. Due to the presence of invalid RDF, we iterated through
the NQuad files on a line-by-line basis. First, we matched each line
against a regular expression were lines matching
\texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
the Perl framework
RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011}
was used to parse the line. Lines that failed to parse were
discarded. We have not investigated whether this could introduce
biases. Statements were then accepted into a new NQuad file if they
had a predicate that matched the \rdfterm{sd:endpoint} or matched a
case-insensitive regular expression \texttt{sparql} if the subject and
objects were both resources, or the predicates
\rdfterm{void:vocabulary}, \rdfterm{rdfa:vocabulary}, or
\rdfterm{api:vocabulary}, as well as having a resource as
object. Finally, statements with the the classes
\rdfterm{cogs:Endpoint}, \rdfterm{owl:Ontology} and
\rdfterm{voaf:Vocabulary} in the object position were also
accepted\footnote{LOV may be used to resolve these prefixed
  names}. More classes and properties were considered, but not used in
the data reduction if they did not occur in the original data.

In the next step, we filtered out statements with URIs that were
invalid or irrelevant, e.g. URIs that didn't have a scheme or where
the scheme were't HTTP(S), or they were referring to private IP addresses.

We then sought to classify resources into the categories ``endpoint''
for SPARQL endpoints, ``vocabulary'' for vocabularies, ``dataset'' for
datasets that may contain further descriptions of several resources,
or simply ``inforesources'' for those that did not fit in the above
classes. To do so, we classified based on certain predicates and
classes. Additionally, URIs derived from prefix.cc were classified as
``vocabulary'' (even if we found several that were not) and those from
SPARQLES as ``endpoint''.

Since we blatantly violated URI opacity with our regular expression
matching in the first step, we needed to further filter candidates for
SPARQL endpoints. This step therefore included filtering as well as
classification.

We found in the data a large number of ontologies that consist of many
information resources with just a few triples in each. Since they
appear to be produced by the same software, usually Semantic
Mediawiki, we assume that they are configured with a single setup, and
thus we merely sampled these resources.

We continued to also sample the HTTP headers gathered in the BTC2014
dataset. 
%as they had recorded some of the relevant headers,
%\httph{Expires}, \httph{Last-Modified} and \httph{ETag}. 
First, we traversed the files with a simple UNIX \texttt{grep} to find
the resources that had reported one of the RDF serializations as
content type. We then traversed this list, first discarded the
resources that did not have a valid IRI (this amounted to just 3273
resources). For the resources we found, as well as for the resources
that was of \rdfterm{rdf:type owl:Ontology} above, we kept one
resource per hostname, with the exception of the popular blogging
platforms Livejournal and SAPO, where each blog has their own host and
they expose FOAF data. For those, we only kept one hostname, since
they are likely configured similarly.

\section{Implementation of spider}\label{app:fetcher}

We then developed a fairly elaborate parallel spider to examine the
resources found on hosts that the previous steps deemed interesting
using the Perl frameworks RDF::Trine and libwww. The spider operated
with a timeout of 20 seconds and a maximum message size it would
accept of 1 MB.

The parallel spider would then launch a process per host, but each
request to one host would be delayed by 10 seconds. For each host, the
spider would go through the list of URLs found by previous steps for
that host. Since the BTC2014 recorded the \httph{Expires},
\httph{Last-Modified} and \httph{ETag} headers where they existed, we
first examined if any of the resources were still fresh, but none
were. Wherever the last two headers existed, we added the
corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
headers for a conditional initial request.

\begin{table}
\caption{Recorded HTTP headers}\label{tab:headers}
 \begin{tabular}{ | l |  p{1.4cm} | p{6.8cm} |}
    \hline
    Header & Reference & Description \\ \hline
\httph{Age} & RFC7234 & When obtaining response from a cache, the number of
seconds since validation \\ \hline
\httph{Cache-Control} & RFC7234 & Header used for a variety of directives \\ \hline
\httph{Expires} & RFC7234  & Gives the date/time after which the
   response is considered stale. \\ \hline
\httph{Pragma} & RFC7234 & Archaic HTTP 1.0 header  \\ \hline
\httph{Warning} & RFC7234  & For additional information about possible incorrectness \\ \hline
\httph{Content-Type} & RFC7231 & To select the correct parser \\ \hline
\httph{If-None-Match} & RFC7232  & Request header to check if
                                   \httph{ETag} has changed   \\ \hline
\httph{If-Modified-Since} & RFC7232  & Request header to check if
                                   \httph{Last-Modified} has changed    \\ \hline
\httph{Last-Modified} & RFC7232 & When the resource was last modified \\ \hline
\httph{ETag} & RFC7232 & An opaque validator to check if the resource
has changed  \\ \hline
\httph{X-Cache} &  & Inserted by some caches to indicate cache status \\ \hline
\httph{Date} & RFC7231 & The time of the message. Used in conditional requests
and heuristics \\ \hline
\httph{Surrogates} & Edge \cite{edgearch} & Draft
to allow fine-grained control for proxies. \\ \hline
\httph{Client-Aborted} & libwww  & Header inserted by User
Agent to indicate that it aborted the download \\ \hline
\httph{Client-Warning} & libwww  & Header inserted by User
Agent to give details about problems with the download \\ \hline

    \hline
    \end{tabular}
\end{table}

For endpoints, we made the following SPARQL query:
\begin{verbatim}
SELECT DISTINCT ?Concept WHERE {[] a ?Concept} LIMIT 2
\end{verbatim}
which should be quite light, yet likely yield results.

Then, the first request would be made, and a selection of the
resulting HTTP headers recorded in an per-host NQuads file. For this
purpose, we developed and released a module
RDF::Generator::HTTP\footnote{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}
to CPAN. We recorded whether the conditional request showed that the
BTC2014 data were still fresh, and if it was, we retrieved the current
data, as we had not coupled the headers to the body in our original
retrieval.  Based on the resulting headers, we let
libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}
calculate both standards-compliant and simple heuristic freshness
lifetime.

If the initial response had RFC7232 headers, we made another
request to see if the server included the headers but does not
support conditional requests. The heuristic we employed is
that if the headers remain the same, but the result was returned,
rather than just a response code \httph{304} (which indicates that the
previous result can be reused), the server does most probably not
support it.

For endpoints, we examined the response message, to see if there are
any results to our query, and recorded that if there are. In addition
to the endpoints registered in the SPARQLES survey, our process found
18~endpoints that responded with results. For all others, we parsed
the response, and recorded any errors if the parser concluded the
content were invalid.

For resource types other than ``vocabulary'', we look for SPARQL
endpoints in the response, using the predicates \rdfterm{sd:endpoint}
and \rdfterm{void:sparqlEndpoint}. We then do the same query as above
and record the relevant headers. Unfortunately, we found early that
this only turns up misconfigured endpoints that point to localhost,
and was removed from the spider for the final analysis.

Finally, if the LOV SPARQL endpoint used a URI for the vocabulary that
was different from the namespace URI (after a normalization step),
another request would be made to record the selected HTTP headers from
that as well.

\section{Statistical method}\label{app:stats}

The hypothesis tests in this paper is implemented using \emph{contingency
  tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the
distribution of  \httph{Server} headers are different for those implementations that
offer caching headers from those that don't. Intuitively, we expect
these distributions to be similar, ``long-tail'' distributions, i.e. a
handful of servers are used by a large number of projects, and then it
falls off rapidly, and so, some servers are used only by very
few. Likewise, it is to be expected that only a few projects have
given caching enough attention, but that they account for the majority
of the support. The question is if the presence of caching headers is
merely a matter of proportion, or if there are some that have given it
more attention, but still is in relatively little use.

Using the statistics system R\cite{kn:r}, we use a statistical test,
namely Pearson's $\chi^2$ test with simulated \pvalue (based on 10000
replicates). The simulation is done using a Monte Carlo method to
compensate for the fact that many servers will not expose caching
headers at all, an issue that would otherwise violate the underlying
assumptions of the test.

\bibliographystyle{plain}
%\bibliographystyle{abbrv}
%\bibliographystyle{jbact}
%\bibliographystyle{splncs03}
\bibliography{webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs}

\end{document}

### 2015-03-10 sanity/paper1.tex
Add some todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1047 43ef1190-90fd-4dd7-a769-d0121f02b80e
More todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1048 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 investigate the actual compliance to RFC7234 and RFC7232. 
-With this paper, we hope to achieve the following:
+With this paper, we hope to achieve the following:\todo{formulate as contributions}
 \begin{enumerate}
@@
 which are close to several users. Finally, the User Agent may
-implement a private cache for its user at the client side.
+implement a private cache for its user at the client side.\todo{split here}
 
@@
 headers are described in Table~\ref{tab:headers}
-in Appendix~\ref{app:fetcher}.
+in Appendix~\ref{app:fetcher}.\todo{ref report}\todo{More on HTTP}
 
@@
 In this paper, we study to what extent SPARQL endpoints, vocabulary
-and data publishers support these standards.
+and data publishers support these standards.\todo{link to dump}
 
@@
 reverse proxy, which is opaque to the Internet, and therefore not of
-our concern.
+our concern. For the same reason, caching that happens within the
+SPARQL engine is not our concern.
 
@@
 explored when caching had a positive effect on complex SPARQL
-queries.
+queries.\todo{discuss SPARQL cache complexity}
 
@@
 some potential.
+
+\todo{reintroduce refs} \todo{add Aidan's desiderata}
 
@@
 consisted of a 4~GTriple corpus of spidered Web data. This was seeded
-from datahub.io (aka CKAN), which is also the definition of the LOD
-cloud, as well as other sources. To compile a list of candidates to
-further examine, we performed a series of data reduction steps,
-manually inspecting the result between each step. The details of this
-process is given in Appendix~\ref{app:reduction}.
+from datahub.io (aka CKAN, which is also used to manage the datasets
+in the ``LOD cloud'')\todo{expand on the coverage}, as well as other sources. To compile a list of
+candidates to further examine, we performed a series of data reduction
+steps, manually inspecting the result between each step. The details
+of this process is given in Appendix~\ref{app:reduction}.\todo{more
+  detail on filtering}
 
@@
 Virtuoso-based SPARQL endpoint for analysis by using the statistics
-system R\cite{kn:r} in the following section.
+system R\cite{kn:r} in the following section.\todo{Reorg, rev 6}
 
@@
 \emph{all} responses had a freshness lifetime or other usable metadata. For
-the former, this amounted to 22 servers. 70 servers always responded
+the former, this amounted to 22 servers\todo{reintroduce table}.  70 servers always responded
 with usable metadata. Inspecting the values we find the well-known
@@
 standards-compliant freshness lifetime of 604800 seconds (i.e. 1 week) for
-both LOD and SPARQL endpoints.
+both LOD and SPARQL endpoints.\todo{elaborate}
 
@@
 for two hosts, where they were superfluous, they were only used to
-prohibit caching. \httph{Surrogates} were not observed.
+prohibit caching. \httph{Surrogates} were not observed.\todo{fix}
 
@@
 also declare a standards compliant freshness lifetime in minutes to
-days.
+days\todo{explain number of occur; fix labels}.
 
@@
 information. Future work could seek more sophisticated fingerprinting
-and understanding of the nature of these differences.
+and understanding of the nature of these differences.\todo{comment on
+  conscious setting}
 
@@
 intermediate results so that subsequent queries can be answered in
-part by the proxy.
+part by the proxy.\todo{more on this difficulty}
 
@@
 prohibitions are misguided, and server administrators are advised to
-turn them off unless they are sure they are required.
+turn them off unless they are sure they are required.\todo{split this
+  in the stuff that follows}
 

### 2015-03-10 sanity/report.tex
More todos
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1048 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 corresponding \httph{If-Modified-Since} and \httph{If-None-Match}
-headers for a conditional initial request.
+headers for a conditional initial request.\todo{fix surrogates}
 

### 2015-03-11 sanity/paper1.tex
Split HTTP into subsection
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1049 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 which are close to several users. Finally, the User Agent may
-implement a private cache for its user at the client side.\todo{split here}
+implement a private cache for its user at the client side.
 
-As mentioned, two standards are relevant for this study, RFC7234,
-which is named ``Caching'' and RFC7232, named ``Conditional
-Requests''. The main difference is that the caching standard defines
-when a response may be reused without contacting the origin server at
-all, whereas the conditional requests define how to validate a
-response by contacting the origin server. The two can be combined:
-Clients and proxies may use the latter to revalidate a response that
-has been cached based on the former.
+\subsection{HTTP Caching standards}
+
+As mentioned, the two documents from the HTTP 1.1 standards suite that
+are relevant for this study are RFC7234, which is named ``Caching''
+and RFC7232, named ``Conditional Requests''. The main difference is
+that the caching standard defines when a response may be reused
+without contacting the origin server at all, whereas the conditional
+requests define how to validate a response by contacting the origin
+server. The two can be combined: Clients and proxies may use the
+latter to revalidate a response that has been cached based on the
+former.
 

### 2015-03-11 sanity/paper1.tex
Site Aidan's desiderata
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1050 43ef1190-90fd-4dd7-a769-d0121f02b80e
Reintroduce cuts
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1051 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 caching and contrasts this with the unavailability of SPARQL query
-caches, and it is claimed that this is an architectural problem.
+caches, and it is claimed that this is an architectural problem. In
+\cite{hogan2014paths}, the authors examine cacheable as one of the
+desiderata for sustainable data access. They claim, without further
+justification, that SPARQL isn't cacheable.
 
@@
 some potential.
-
-\todo{reintroduce refs} \todo{add Aidan's desiderata}
 
@@
 also declare a standards compliant freshness lifetime in minutes to
-days\todo{explain number of occur; fix labels}.
+days\todo{explain number of occur; fix labels}\todo{qqplot in report}.
 

### 2015-03-11 sanity/paper1.tex
Fix surrogates
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1052 43ef1190-90fd-4dd7-a769-d0121f02b80e
Reintroduce Server table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1053 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 For each unique \httph{Server} header, we found the ones where
-\emph{all} responses had a freshness lifetime or other usable metadata. For
-the former, this amounted to 22 servers\todo{reintroduce table}.  70 servers always responded
-with usable metadata. Inspecting the values we find the well-known
-Virtuoso and Callimachus servers, as well as the Perl modules
-RDF::LinkedData and RDF::Endpoint, which is partly developed by and
-runs on a server operated by this author. Apart from those, a typical
-server string is e.g. \httph{Apache/2.2.22 (Fedora)}, which reveals
-very little about the RDF-specific parts of the underlying server
-implementation. A quick inspection of all \httph{Server} headers
-confirmed that few reveal any further detail.
+\emph{all} responses had a freshness lifetime or other usable
+metadata. For the former, this amounted to 22 servers, which are
+listed in Table~\ref{tab:servers}.  70 servers always responded with
+usable metadata. Inspecting the values we find the well-known Virtuoso
+and Callimachus servers, as well as the Perl module
+RDF::LinkedData\footnote{See
+  \url{https://metacpan.org/pod/RDF::LinkedData}. It is also present
+  in Debian/Ubuntu as \texttt{librdf-linkeddata-perl}}, which is
+developed by and runs on a server operated by this author. Apart from
+those, we see that the they reveal very little about the RDF-specific
+parts of the underlying server implementation. A quick inspection of
+all \httph{Server} headers confirmed that few reveal any further
+detail.
+
+\begin{table}[h]
+  \caption{\httph{Server} headers for hosts that enabled a freshness
+    lifetime to be computed for \emph{all} requests.}\label{tab:servers}
+\begin{center}
+\begin{tabular}{rl}
+  \hline
+  1 & DFE/largefile \\ 
+  2 & git\_frontend \\ 
+  3 & nginx/1.3.9 \\ 
+  4 & thin 1.6.0 codename Greek Yogurt \\ 
+  5 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
+  6 & Oracle-Application-Server-10g/10.1.3.4.0 Oracle-HTTP-Server [\ldots]\\
+  7 & TwistedWeb/8.2.0 \\ 
+  8 & RDF::Endpoint/0.07 \\ 
+  9 & Jetty(6.1.26) \\ 
+  10 & nginx/1.6.1 \\ 
+  11 & Jigsaw/2.3.0-beta3 \\ 
+  12 & Apache/2.2.9 (Win32) PHP/5.2.6 \\ 
+  13 & Apache/2.4.10 (Unix) mod\_fcgid/2.3.9 \\ 
+  14 &  \\ 
+  15 & GFE/2.0 \\ 
+  16 & RDF::LinkedData/0.70 \\ 
+  17 & Apache/2.2.17 (Unix) mod\_wsgi/3.3 Python/2.6.6 \\ 
+  18 & Virtuoso/07.10.3211 (Linux) i686-generic-linux-glibc212-64  VDB \\ 
+  19 & Apache/2.2.24 (Unix) mod\_ssl/2.2.24 OpenSSL/0.9.8y \\ 
+  20 & Apache/2.2.22 (Fedora) \\ 
+  21 & INSEE \\ 
+  22 & GitHub.com \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 for two hosts, where they were superfluous, they were only used to
-prohibit caching. \httph{Surrogates} were not observed.\todo{fix}
+prohibit caching. \httph{Surrogate-Capability} were not observed.
 

### 2015-03-11 sanity/paper1.tex
\url for links
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1054 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 closer inspection, we find this to be hosted by
-Dydra\footnote{http://dydra.com/}. We speculate that this is due to
+Dydra\footnote{\url{http://dydra.com/}}. We speculate that this is due to
 that few underlying DBMS systems help track modification times in a

### 2015-03-11 sanity/paper1.tex
Similarity ref
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1055 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 explored when caching had a positive effect on complex SPARQL
-queries.\todo{discuss SPARQL cache complexity}
+queries.
 
@@
 changes. Designing user support systems as well as interfaces that fit
-the publisher's workflow, is an important problem.
+the publisher's workflow is an important problem.
 
@@
 intermediate results so that subsequent queries can be answered in
-part by the proxy.\todo{more on this difficulty}
+part by the proxy.
+
+We believe that  \cite{ldf1} and \cite{hogan2014paths} prematurely
+reject caching in connection to SPARQL. They are correct that
+currently, query results can only be cached on a per-query
+basis. Moreover, semantically insignificant changes to a query, such
+as whitespace or the order of triple patterns in a basic graph
+pattern, currently causes problems for caches. The latter problem can
+probably be fixed by developing digest algorithms. Such algorithms
+exist, but are focused on cryptographical strength, and much simpler
+algorithms could be used for this problem.
+
+Furthermore, by using similarity measures, like those discussed in
 

### 2015-03-11 sanity/paper1.tex
SPARQL evaluation with cache
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1056 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Furthermore, by using similarity measures, like those discussed in
+\cite{dividino2013following}, proxies can proactively prefetch data
+into a cache to help answer queries. A cost model that takes into
+account several data access methods, those described by
+\cite{hogan2014paths} as well as  the novel work in \cite{ldf1} may be
+a key ingredient in enabling efficient SPARQL evaluation on the Web.
 

### 2015-03-11 sanity/paper1.tex
Recommendations
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1057 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-Based on the results of DyLDO, it is highly likely that most cache
-prohibitions are misguided, and server administrators are advised to
-turn them off unless they are sure they are required.\todo{split this
-  in the stuff that follows}
+It is mainly the objective of this paper to be descriptive, but from
+the results of this study and on the results of DyLDO, we note that it
+is highly likely that most cache prohibitions are misguided, and
+server administrators are advised to turn them off unless they are
+sure they are required.
+
+Additionally, based in part on the survey, but also on other practical
+experience, we suggest the following:
 
@@
 libraries used in this study, Axel Polleres for encouraging comments
-and Jonas Smedegaard for proofreading.
+and Jonas Smedegaard for proofreading. We also thank the anonymous
+reviewers for extensive reviews.
 

### 2015-03-11 sanity/paper1.tex
Conscious settings
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1058 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 information. Future work could seek more sophisticated fingerprinting
-and understanding of the nature of these differences.\todo{comment on
-  conscious setting}
+and understanding of the nature of these differences. With this, it
+may also be possible to investigate whether expiry times have been set
+consciously.
 

### 2015-03-11 sanity/paper1.tex
Make refs to tech report
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1059 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \usepackage{hyperref}
-\usepackage{appendix}
 
@@
 the updated response only if the resource has changed. Other relevant
-headers are described in Table~\ref{tab:headers}
-in Appendix~\ref{app:fetcher}.\todo{ref report}\todo{More on HTTP}
+headers are described in a companion technical report
+\cite{kjernsmo_add_survey_2015}.
+% Table~\ref{tab:headers} in Appendix~\ref{app:fetcher}.
+\todo{More on HTTP}
 
@@
 steps, manually inspecting the result between each step. The details
-of this process is given in Appendix~\ref{app:reduction}.\todo{more
-  detail on filtering}
+of this process is given in the companion technical report
+\cite{kjernsmo_add_survey_2015}.% Appendix~\ref{app:reduction}.
+\todo{more detail on filtering}
 
@@
 may host SPARQL endpoints, vocabularies, or other information
-resources, by a spider detailed in Appendix~\ref{app:fetcher},
-resulting in 7745 requests, done on 2015-01-02. Although this is not a
-very large number, we assume that caching headers are usually set on a
-per-host basis, if any differences exist, they are likely on different
-types, and so the number of hosts is more important than the
-number of resources.
+resources, by a spider detailed in the companion technical report
+\cite{kjernsmo_add_survey_2015}, resulting in 7745 requests, done on
+2015-01-02. Although this is not a very large number, we assume that
+caching headers are usually set on a per-host basis, if any
+differences exist, they are likely on different types, and so the
+number of hosts is more important than the number of resources.
 
@@
 some servers are better configured to support caching than others.
-Using the methodology given in Appendix~\ref{app:stats}, we find in
+Using the methodology given in the companion technical report
+\cite{kjernsmo_add_survey_2015}, we find in
 both the cases of standards-compliant freshness lifetime and for the other
@@
 
-As shown in Table~\ref{tab:headers}, there are also other caching
+The companion technical report \cite{kjernsmo_add_survey_2015}, lists other caching
 headers than those mentioned in the introduction. We found
@@
 should be assessed, possibly using contingency tables similar to those
-in Appendix~\ref{app:stats}.
+in the companion technical report \cite{kjernsmo_add_survey_2015}.
 
@@
 %\bibliographystyle{splncs03}
-\bibliography{webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs}
+\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}
 

### 2015-03-11 sanity/paper1.tex
contributions
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1060 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 investigate the actual compliance to RFC7234 and RFC7232. 
-With this paper, we hope to achieve the following:\todo{formulate as contributions}
+The objectives of this paper are:
 \begin{enumerate}
@@
 \end{enumerate}
+The contributions of this paper are to meet these objective by means
+of a survey that shows that while the uptake has been moderate,
+practical benefits may be realized already. Based on this survey as
+well as practical experience, we point out future research directions
+as well as some recommendations for deployed implementations.
 

### 2015-03-11 sanity/paper1.tex
Long comment on coverage and reduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1061 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 consisted of a 4~GTriple corpus of spidered Web data. This was seeded
-from datahub.io (aka CKAN, which is also used to manage the datasets
-in the ``LOD cloud'')\todo{expand on the coverage}, as well as other sources. To compile a list of
+from datahub.io (aka CKAN), as well as other sources. To compile a list of
 candidates to further examine, we performed a series of data reduction
@@
 \cite{kjernsmo_add_survey_2015}.% Appendix~\ref{app:reduction}.
-\todo{more detail on filtering}
 
@@
 \cite{kjernsmo_add_survey_2015}, resulting in 7745 requests, done on
-2015-01-02. Although this is not a very large number, we assume that
-caching headers are usually set on a per-host basis, if any
-differences exist, they are likely on different types, and so the
-number of hosts is more important than the number of resources.
+2015-01-02.
 
@@
 system R\cite{kn:r} in the following section.\todo{Reorg, rev 6}
+
+A key challenge to the validity of the survey is biases that may be
+introduced by the coverage and then the data reduction. The breadth of
+the Semantic Web is derived mainly from the BTC2014 crawl. While
+LODstats\footnote{\url{http://stats.lod2.eu/}} has presently seen an
+order of magnitude more triples, the number of error-free datasets
+were at the time of this writing 4442. We work under the assumption
+that cache headers are set mostly on a per-host basis, and if this
+assumption is valid, sampling a URL per host is sufficient. LODstats
+do not report per-host statistics, but often one host will host
+several datasets. Another recent crawl was reported by
+\cite{stateoflod2014}. It is not clear how many hosts were crawled, but the
+number of triples in the crawl is much smaller than that of
+BTC2014. It is therefore a fair assumption that BTC2014 fairly well
+represent the breadth of the Semantic Web, momentarily at least.
+
+As for the coverage of vocabularies, we realize that LODstats might
+have been a better source, but we have been able to select the most
+important, and not far inferior in number. The number of SPARQL
+endpoints found in SPARQLES is larger than LODstats, and we also
+looked for further endpoints both in the BTC2014 and our own crawl,
+finding only a few. If endpoints went underdiscovered, then there is a
+discovery problem that is beyond this survey to rectify.
+
+The data reduction that was subsequently done was mainly done to
+eliminate errors. We have not investigated biases that may be
+introduced by discarding momentarily dysfunctional parts of the
+Semantic Web, but we investigated whether the freshness lifetimes
+reported in the case of certain errors were distributed differently
+from those that returned a valid response, the companion technical
+report \cite{kjernsmo_add_survey_2015}. We found that they were, but
+we have assumed that this is due to that errors are configured to be
+cached differently, which we know from experience is common
+practice. The following analysis is based on valid responses.
 

### 2015-03-11 sanity/paper1.tex
Make a section of validity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1062 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-A key challenge to the validity of the survey is biases that may be
+\subsection{Challenges to validity}
+
+Key challenges to the validity of the survey are biases that may be
 introduced by the coverage and then the data reduction. The breadth of
@@
 
-As for the coverage of vocabularies, we realize that LODstats might
-have been a better source, but we have been able to select the most
-important, and not far inferior in number. The number of SPARQL
-endpoints found in SPARQLES is larger than LODstats, and we also
-looked for further endpoints both in the BTC2014 and our own crawl,
-finding only a few. If endpoints went underdiscovered, then there is a
-discovery problem that is beyond this survey to rectify.
+As for the coverage of vocabularies, we have verified that all
+resolveable vocabularies in LODstats that are found in more than 10
+datasets are visited, and it is not far inferior in number to
+LODstats. The number of SPARQL endpoints found in SPARQLES is larger
+than LODstats, and we also looked for further endpoints both in the
+BTC2014 and our own crawl, finding only 18. If endpoints went
+underdiscovered, then there is a discovery problem that is beyond this
+survey to rectify.
 
@@
 reported in the case of certain errors were distributed differently
-from those that returned a valid response, the companion technical
+from those that returned a valid response, see the companion technical
 report \cite{kjernsmo_add_survey_2015}. We found that they were, but

### 2015-03-12 sanity/paper1.tex
size is beside the point, Comment 3
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1063 43ef1190-90fd-4dd7-a769-d0121f02b80e
Comment on RKBexplorer
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1064 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Endpoints from the SPARQLES survey \cite{buil2013sparql}, vocabularies
-from Linked Open Vocabularies (LOV) \cite{lov2} and prefix.cc, and
-we augment these data with spidered data from the Billion Triple
-Challenge (BTC) 2014 \cite{btc-2014} dataset.
+from Linked Open Vocabularies (LOV) \cite{lov2} and prefix.cc, and we
+augment these data with spidered data from the Billion Triple
+Challenge (BTC) 2014 \cite{btc-2014} dataset. Of these, BTC2014 is by
+far the largest, the others are small, curated and targeted
+datasets. However, the size is besides the point, we are only
+interested in examining as many hosts as possible, and they are still
+few.
 
@@
 Amongst the servers that expose caching headers it is common
-that the result may be cached for some minutes.
+that the result may be cached for some minutes, and closer inspection
+of the dataset reveals that many of these are due to that
+RBKExplorer\footnote{\url{http://www.rkbexplorer.com/}} sets an
+freshness lifetime of 300~s for many endpoints.
 

### 2015-03-12 sanity/paper1.tex
Comment on RKBexplorer
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1065 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 of the dataset reveals that many of these are due to that
-RBKExplorer\footnote{\url{http://www.rkbexplorer.com/}} sets an
+RBKExplorer\footnote{\url{http://www.rkbexplorer.com/about/}} sets an
 freshness lifetime of 300~s for many endpoints.

### 2015-03-12 sanity/paper1.tex
Elaborate on DBpedia
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1066 43ef1190-90fd-4dd7-a769-d0121f02b80e
RDF::Endpoint is in there
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1067 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
+\todo{overview of characteristics and why}
+
 \subsection{Different server implementations}
@@
 usable metadata. Inspecting the values we find the well-known Virtuoso
-and Callimachus servers, as well as the Perl module
-RDF::LinkedData\footnote{See
-  \url{https://metacpan.org/pod/RDF::LinkedData}. It is also present
-  in Debian/Ubuntu as \texttt{librdf-linkeddata-perl}}, which is
+and Callimachus servers, as well as the Perl modules
+RDF::LinkedData and RDF::Endpoint, which are partly
 developed by and runs on a server operated by this author. Apart from
@@
 standards-compliant freshness lifetime of 604800 seconds (i.e. 1 week) for
-both LOD and SPARQL endpoints.\todo{elaborate}
+both LOD and SPARQL endpoints. DBPedia has historically been updated
+only a few times a year, but this was probably chosen so as to not
+make a commitment long into the future. It may provide considerable benefits.
 

### 2015-03-12 sanity/queries.R
Make a figure with log scale
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1069 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -79,3 +79,15 @@ par(mai=c(1,1,0.2,0.05))
 qqplot(errorok,errorparse, xlab="Lifetime when no errors", ylab="Lifetime with parse errors")
 dev.off()
 
+# To generate logarithmic as request by review 6
+data <- sparqlfile("other-hard.rq")
+hardvec <- data$results$fresh
+hardvec[hardvec <= 0] <- 0
+hardveclog <- log10(hardvec)
+hardveclog[hardveclog <= 0] <- -1
+hardvecloghist <- hist(hardveclog, axes=F, main=NULL)
+axis(side=2, at=seq(0, max(hardvecloghist$counts), 20))
+axis(side=1, at=hardvecloghist$breaks, label=c(0, 10^(hardvecloghist$breaks[-1])) )
+tblog <- log10(timebuckets(hardvec)$points)
+tblog[tblog < 0] <- -1
+axis(3, at=tblog, label=c(timebuckets(hardvec)$names, "Max"))

### 2015-03-12 sanity/queries.R
Use colors that flatten fine to grayscale
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1070 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@ -53,9 +53,11 @@ heuristictable <- lifetimetable("other-heuristic.rq")
 hardtable2 <- cbind(hardtable[,3], hardtable[,1], hardtable[,4], hardtable[,2])
 colnames(hardtable2) <- colnames(heuristictable)
 
+colors <- c("#fdb863", "#e66101", "#5e3c99", "#b2abd2")
+
 pdf(file="hardtable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(hardtable2, main='', xlab="Standards-compliant freshness lifetime", ylab="Relative abundance of resource types", color=c("red", "blue", "green", "yellow"))
+mosaicplot(hardtable2, main='', xlab="Standards-compliant freshness lifetime", ylab="Relative abundance of resource types", color=colors)
 dev.off()
 
 heuristicall <- apply(heuristictable, 1, sum)
@@ -67,7 +69,7 @@ dev.off()
 
 pdf(file="heuristictable.pdf", height=5.2,width=8)
 par(mai=c(0.6,0.6,0.05,0.05))
-mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Relative abundance of resource types", color=c("red", "blue", "green", "yellow"))
+mosaicplot(heuristictable, main='', xlab="Heuristic freshness lifetime", ylab="Relative abundance of resource types", color=colors)
 dev.off()
 
 harderrordata <- sparqlfile("failed-hard.rq")

### 2015-03-12 sanity/paper1.tex
Update my report bib
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1068 43ef1190-90fd-4dd7-a769-d0121f02b80e
Use colors that flatten fine to grayscale
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1070 43ef1190-90fd-4dd7-a769-d0121f02b80e
Fix the percentage labels
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1071 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 also declare a standards compliant freshness lifetime in minutes to
-days\todo{explain number of occur; fix labels}\todo{qqplot in report}.
+days\todo{explain number of occur}\todo{qqplot in report}.
 
@@
     total counts, using the same bins as in
-    Figure~\ref{fig:hardall}. Yellow boxes denote SPARQL endpoints,
-    green dataset descriptions, blue generic information resources and
-    red vocabularies.}
+    Figure~\ref{fig:hardall}. From bottom to top, light blue boxes
+    denote SPARQL endpoints, dark violet dataset descriptions, orange
+    generic information resources and light orange vocabularies.}
   \label{fig:hardtable}
@@
 
-\bibliographystyle{plain}
-%\bibliographystyle{abbrv}
+%\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}

### 2015-03-12 sanity/queries.R

@@ -46,7 +46,8 @@ hardall <- apply(hardtable, 1, sum)
 pdf(file="hardall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
 bphard <- barplot(hardall, col="white", xlab="Standards-compliant freshness lifetime", ylab="Number of occurences", main='')
-text(bphard, hardall, labels=paste0(signif(hardall*100/sum(hardall),2),'%'), pos=1)
+text(bphard, c(0, hardall[-1]), labels=c('',paste0(signif(hardall[-1]*100/sum(hardall),2),'%')), pos=3)
+text(bphard, hardall[1], labels=c(paste0(signif(hardall[1]*100/sum(hardall),2),'%'), rep('', 7)), pos=1)
 dev.off()
 
 heuristictable <- lifetimetable("other-heuristic.rq")
@@ -64,7 +65,10 @@ heuristicall <- apply(heuristictable, 1, sum)
 pdf(file="heuristicall.pdf", height=5.2,width=8)
 par(mai=c(1,1,0.2,0.05))
 bpheuristic <- barplot(heuristicall, col="white", xlab="Simple heuristic freshness lifetime", ylab="Number of occurrences", main='')
-text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=1)
+text(bpheuristic, heuristicall, labels=paste0(signif(heuristicall*100/sum(heuristicall),2),'%'), pos=3)
+text(bpheuristic, c(rep(0, 6), heuristicall[7], 0),
+     labels=c(rep('', 6), paste0(signif(heuristicall[7]*100/sum(heuristicall),2),'%'), ''),
+         pos=1)
 dev.off()
 
 pdf(file="heuristictable.pdf", height=5.2,width=8)

### 2015-03-12 sanity/paper1.tex
Explain the axes better
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1072 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 also declare a standards compliant freshness lifetime in minutes to
-days\todo{explain number of occur}\todo{qqplot in report}.
+days\todo{qqplot in report}.
 
@@
     \includegraphics[width=.9\textwidth]{hardall.pdf}}
-  \caption{Barplot counting all standards-compliant freshness lifetimes
-    found, with the percentage of occurrences indicated on the
-    bars. The first bin are the cases where caching is explicitly
-    prohibited. The next bins are for lifetimes, where the values are
-    grouped if they are on the order of seconds, minutes, hours, etc,
-    i.e. the second bin counts the lifetimes in the interval [1,59]
-    seconds, etc.}
+  \caption{Barplot counting all standards-compliant freshness
+    lifetimes found, with the percentage of occurrences indicated on
+    the bars. On the horizontal axis, the first bin are the cases
+    where caching is explicitly prohibited. The next bins are for
+    lifetimes, where the values are grouped if they are on the order
+    of seconds, minutes, hours, etc, i.e. the second bin counts the
+    lifetimes in the interval [1,59] seconds, etc. On the vertical
+    axis, the number of times a certain freshness lifetime was found.}
   \label{fig:hardall}
@@
   \caption{Barplot counting all simple heuristic freshness lifetimes
-    found. Bins as in Figure~\ref{fig:hardall}.}
+    found. Axes as in Figure~\ref{fig:hardall}.}
   \label{fig:heuristicall}

### 2015-03-12 sanity/paper1.tex
smaller things
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1073 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-The companion technical report \cite{kjernsmo_add_survey_2015}, lists other caching
+The companion technical report \cite{kjernsmo_add_survey_2015} lists other caching
 headers than those mentioned in the introduction. We found
@@
 observation that is also supported by a similar hypothesis test as
-above, with a \pvalue = 0.00001. We also note that it is often
-prohibited to cache dataset descriptions. This is odd, since
-statistics about datasets is usually costly to compute and should be
-cached. The VoID specification \cite{voidnote} also notes that the
-statistics are considered estimates.
+above, with a \pvalue = 0.00001 (note, however, it is more contrived
+than above, since the bins are like in Figure~\ref{fig:hardtable},
+which is chosen for intuitive interpretation rather than statistical
+rigor). Second, we note that it is often prohibited to cache dataset
+descriptions. This is odd, since statistics about datasets is usually
+costly to compute and should be cached. The VoID specification
+\cite{voidnote} also notes that the statistics are considered
+estimates.
 

### 2015-03-12 sanity/paper1.tex
Analysis characteristics
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1074 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\todo{overview of characteristics and why}
+The analysis is focused on finding descriptive statistics to
+understand how different servers support caching, for how long
+resources hosted with those that do support caching may be considered
+fresh, if it is possible to easily compute a heuristic freshness
+lifetime, and revalidate the response on expiry. Apart from quoting
+the numbers we aggregated, we do this by presenting summarized data
+distribution visualizations, to allow for an intuitive understanding
+of the data. Where appropriate, we also do statistical hypothesis
+tests.
 

### 2015-03-12 sanity/paper1.tex
Move TODOs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1075 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 Virtuoso-based SPARQL endpoint for analysis by using the statistics
-system R\cite{kn:r} in the following section.\todo{Reorg, rev 6}
+system R\cite{kn:r} in the following section.
 
@@
 also declare a standards compliant freshness lifetime in minutes to
-days\todo{qqplot in report}.
+days.
 

### 2015-03-12 sanity/paper1.tex
Bring back headers table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1077 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the updated response only if the resource has changed. Other relevant
-headers are described in a companion technical report
-\cite{kjernsmo_add_survey_2015}.
-% Table~\ref{tab:headers} in Appendix~\ref{app:fetcher}.
+headers are described in % a companion technical report
+%\cite{kjernsmo_add_survey_2015}.
+ Table~\ref{tab:headers}.
 \todo{More on HTTP}
+
+\begin{table}[tb]
+\caption{Recorded HTTP headers}\label{tab:headers}
+ \begin{tabular}{ | l |  p{1.4cm} | p{6.8cm} |}
+    \hline
+    Header & Reference & Description \\ \hline
+\httph{Age} & RFC7234 & When obtaining response from a cache, the number of
+seconds since validation \\ \hline
+\httph{Cache-Control} & RFC7234 & Header used for a variety of directives \\ \hline
+\httph{Expires} & RFC7234  & Gives the date/time after which the
+   response is considered stale. \\ \hline
+\httph{Pragma} & RFC7234 & Archaic HTTP 1.0 header  \\ \hline
+\httph{Warning} & RFC7234  & For additional information about possible incorrectness \\ \hline
+\httph{Content-Type} & RFC7231 & To select the correct parser \\ \hline
+\httph{If-None-Match} & RFC7232  & Request header to check if
+                                   \httph{ETag} has changed   \\ \hline
+\httph{If-Modified-Since} & RFC7232  & Request header to check if
+                                   \httph{Last-Modified} has changed    \\ \hline
+\httph{Last-Modified} & RFC7232 & When the resource was last modified \\ \hline
+\httph{ETag} & RFC7232 & An opaque validator to check if the resource
+has changed  \\ \hline
+\httph{X-Cache} &  & Inserted by some caches to indicate cache status \\ \hline
+\httph{Date} & RFC7231 & The time of the message. Used in conditional requests
+and heuristics \\ \hline
+\httph{Surrogate-Capability} & Edge \cite{edgearch} & Draft
+to allow fine-grained control for proxies. \\ \hline
+\httph{Client-Aborted} & libwww  & Header inserted by User
+Agent to indicate that it aborted the download \\ \hline
+\httph{Client-Warning} & libwww  & Header inserted by User
+Agent to give details about problems with the download \\ \hline
+
+    \hline
+    \end{tabular}
+\end{table}
 

### 2015-03-12 sanity/report.tex
Fix surrogates
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1052 43ef1190-90fd-4dd7-a769-d0121f02b80e
Make a section of validity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1062 43ef1190-90fd-4dd7-a769-d0121f02b80e
Move TODOs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1075 43ef1190-90fd-4dd7-a769-d0121f02b80e
Bring back headers table
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1077 43ef1190-90fd-4dd7-a769-d0121f02b80e

--- 2015-03-10 sanity/report.tex
@@
 \section{Introduction}
+
+\todo{qqplot in report}.
 
@@
 dataset. 
-%as they had recorded some of the relevant headers,
-%\httph{Expires}, \httph{Last-Modified} and \httph{ETag}. 
 First, we traversed the files with a simple UNIX \texttt{grep} to find
@@
 headers for a conditional initial request.
-
-\begin{table}
-\caption{Recorded HTTP headers}\label{tab:headers}
- \begin{tabular}{ | l |  p{1.4cm} | p{6.8cm} |}
-    \hline
-    Header & Reference & Description \\ \hline
-\httph{Age} & RFC7234 & When obtaining response from a cache, the number of
-seconds since validation \\ \hline
-\httph{Cache-Control} & RFC7234 & Header used for a variety of directives \\ \hline
-\httph{Expires} & RFC7234  & Gives the date/time after which the
-   response is considered stale. \\ \hline
-\httph{Pragma} & RFC7234 & Archaic HTTP 1.0 header  \\ \hline
-\httph{Warning} & RFC7234  & For additional information about possible incorrectness \\ \hline
-\httph{Content-Type} & RFC7231 & To select the correct parser \\ \hline
-\httph{If-None-Match} & RFC7232  & Request header to check if
-                                   \httph{ETag} has changed   \\ \hline
-\httph{If-Modified-Since} & RFC7232  & Request header to check if
-                                   \httph{Last-Modified} has changed    \\ \hline
-\httph{Last-Modified} & RFC7232 & When the resource was last modified \\ \hline
-\httph{ETag} & RFC7232 & An opaque validator to check if the resource
-has changed  \\ \hline
-\httph{X-Cache} &  & Inserted by some caches to indicate cache status \\ \hline
-\httph{Date} & RFC7231 & The time of the message. Used in conditional requests
-and heuristics \\ \hline
-\httph{Surrogates} & Edge \cite{edgearch} & Draft
-to allow fine-grained control for proxies. \\ \hline
-\httph{Client-Aborted} & libwww  & Header inserted by User
-Agent to indicate that it aborted the download \\ \hline
-\httph{Client-Warning} & libwww  & Header inserted by User
-Agent to give details about problems with the download \\ \hline
-
-    \hline
-    \end{tabular}
-\end{table}
 

### 2015-03-12 sanity/paper1.tex
Link to homepage
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1078 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 In this paper, we study to what extent SPARQL endpoints, vocabulary
-and data publishers support these standards.\todo{link to dump}
+and data publishers support these standards. Data and code to
+reproduce this work will be available at
+\url{http://folk.uio.no/kjekje/#cache-survey}.
 

### 2015-03-12 sanity/paper1.tex
Fix links and trim
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1079 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the updated response only if the resource has changed. Other relevant
-headers are described in % a companion technical report
-%\cite{kjernsmo_add_survey_2015}.
- Table~\ref{tab:headers}.
+headers are described in Table~\ref{tab:headers}.
 \todo{More on HTTP}
@@
 steps, manually inspecting the result between each step. The details
-of this process is given in the companion technical report
-\cite{kjernsmo_add_survey_2015}.% Appendix~\ref{app:reduction}.
+of this process is given in a companion technical report
+\cite{kjernsmo_add_survey_2015}.
 
@@
 may host SPARQL endpoints, vocabularies, or other information
-resources, by a spider detailed in the companion technical report
+resources, by a spider also detailed in
 \cite{kjernsmo_add_survey_2015}, resulting in 7745 requests, done on
@@
 \cite{stateoflod2014}. It is not clear how many hosts were crawled, but the
-number of triples in the crawl is much smaller than that of
+number of triples is much smaller than that of
 BTC2014. It is therefore a fair assumption that BTC2014 fairly well
@@
 of the data. Where appropriate, we also do statistical hypothesis
-tests.
+tests, using so-called contingency tables, see
+\cite{kjernsmo_add_survey_2015} for details.
 
@@
 
-The companion technical report \cite{kjernsmo_add_survey_2015} lists other caching
-headers than those mentioned in the introduction. We found
+We also looked for other headers in \ref{tab:headers}. We found
 \httph{Pragma} (archaic HTTP 1.0 header) in 287 responses, but except

### 2015-03-12 sanity/paper1.tex
Remove todo
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1080 43ef1190-90fd-4dd7-a769-d0121f02b80e
Ready to proofread and review
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1081 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 \newcommand{\httph}[1]{\texttt{#1}}
-\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
+%\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 
-We note that caching is not only useful for long-living information
+We note that caching is not only useful for long-living 
 resources, even though that may be the most important use. If a resource
@@
 freshness lifetime should be based of the above, we therefore shall
-refer to this as ``standards-compliant caching''. However it can also
+refer to this as ``standards-compliant caching''. It can also
 be based on heuristics, and Section~4.2.2 in RFC7234 provides some loose
@@
 or \httph{Last-Modified} which gives the time and date of the last
-modification of the resource. Clients may use these values to retrieve
-the updated response only if the resource has changed. Other relevant
-headers are described in Table~\ref{tab:headers}.
-\todo{More on HTTP}
+modification of the resource. Clients that have obtained these values
+may use them to validate an earlier response by using
+\httph{If-None-Match} and/or \httph{If-Modified-Since} respectively in
+a subsequent request. If the server finds the response has not changed
+based on this, it will respond with a \httph{304} status code and no
+body, otherwise it will return the full response. The other headers we
+recorded are listed in Table~\ref{tab:headers}.
 
-\begin{table}[tb]
+\begin{table}[bt]
 \caption{Recorded HTTP headers}\label{tab:headers}

### 2015-03-13 sanity/letter.tex
Start of response letter
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1082 43ef1190-90fd-4dd7-a769-d0121f02b80e

\documentclass{article}
\usepackage[english,british]{babel}

\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{appendix}

\newcommand{\rdfterm}[1]{\texttt{#1}}
\newcommand{\pvalue}{\textit{p}-value\ }
\newcommand{\httph}[1]{\texttt{#1}}
\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}

\title{Letter concerning changes in ``A survey of HTTP caching implementations on the open Semantic Web''}
\author{Kjetil Kjernsmo}

%\subtitle{---Unpublished Working Draft, do not circulate---}

\begin{document}
\maketitle

I have as requested made extensive changes to the submitted
paper. First, the appendices were moved to a Technical Report, due to
be published by the University of Oslo shortly as
\cite{kjernsmo_add_survey_2015}. This report will also include some
extra material, particularly statistical methods, that were removed
from the paper before submission.

\section{Meta-review items}

\begin{enumerate}

\item We include Table~2, which lists the \httph{Server} headers for
  hosts that enabled a freshness lifetime to be computed for
  \emph{all} requests. It makes the conclusion that not much can be
  found from these headers clear, and shows how they should look. For
  the contingency table test, the reader is referred to
  \cite{kjernsmo_add_survey_2015}.

\item Number of occurences is now explained in the caption. The
  percentage numbers have been hacked to display well.

\item This comment seems to pertain to different aspects of the paper:
  First, I acknowledge that I quote many numbers in the text, outside
  of the descriptive statistics, and I understand that this may seem
  disorganized, and that the desire may emerge to see them nicely
  tabulated. They don't lend themselves well to tabular display, since
  they come from quite different parts of the analysis. For example, I
  don't think it would serve paper well if the figures of the
  \httph{Pragma} headers would be in a table with the
  \httph{Cache-Control} headers, the discussion is too
  different. Also, I note that the majority of the reviewers rated the
  paper well with respect to clarity. The other aspect of the paper
  seems related to coverage. To answer that, I have noted that the
  relative sizes of the databases are not important. See also the next
  item.

\item I have made a separate section entitled ``Challenges to
  validity'' to address some of these concerns. In retrospect, using
  LODstats for a list of vocabularies would probably have spared me
  much trouble, but should not challenge the validity much.

\item First, it is clearly out of the scope of the paper to discuss
  how to cache and get a performance improvement. As this is the focus
  of my current work, I took the opportunity to talk about it dearly
  anyway, and extended the future work section with a discussion. I
  note that the reviewer had changed their review to ask us to
  ``acknowledge the complexity to cache SPARQL'', which I think I had
  done in the related work section where I wrote that ``In some of the
  related work, it is shown that caching does not necessarily give
  tangible benefits.'', but I didn't at that time find space to
  elaborate. With space freed by the technical report, I have now done
  so anyway.

\item A very brief comment in ``Future work'', as doing this as a
  major undertaking. Ultimately, it can probably only be done by an
  extensive survey using questionnaires.

\item I acknowledge the error, and reran the spider recording
  \httph{Surrogate-Capability}, which is the correct header, but that
  were not observed either.

\item First, I changed ``RFC7234, Section~4.2.2'' to ``Section~4.2.2
  in RFC7234'', which I hope should clarify that it pertains to the
  standard. Second, since it is my agenda to spread knowledge about
  foundational Web technologys, I would like to follow this
  suggestion. However, to satisfy all the other requirements, I found
  this hard to prioritize, after all this is about the HTTP standard,
  a standard that is fundamental to all practical contributions to the
  Semantic Web. Since review 5 requested further elaboration on how
  the revalidation process works, I detailed that. I also reintroduced
  the table of all recorded headers, that previously was in the
  appendix.

\item See Section~\ref{rev6}.

\item I have made another pass of the paper, and also sent it to a
  professional English proofreader, but don't know if it will fit in
  their schedule to review it before the deadline.

\end{enumerate}

\section{Reviewer 6's comments}\label{rev6}

\begin{description}
\item [Contribution list] I have rewritten two paragraphs, essentially
  keeping the objectives, but detailing how meeting those objectives
  realize the contributions.
\item [Preliminaries and Fundamentals] I have, as noted about, further
  expanded on the RFCs, and put that in a separate subsection under
  Introduction, as I think the text flows better with that
  arrangement. I also included the table as per their suggestion.
\item [Methodology] I identified the the source of the correlation
  that reviewer mentioned as due to a single host having many SPARQL
  endpoints configured similarly and noted that.
\item [Data] I have included a URL to my homepage, where I will make
  an archive available shortly with data and code (I have to admit
  that the code isn't all that pretty, I didn't intend the survey to
  become as extensive as it that, that's why I didn't just release it
  immediately).
\item [Analysis1] I'm not quite sure what they are getting at, but
  included a paragraph just after the section headline.
\item [Analysis2] See Section~\ref{histogramstuff}.
\item [Discussion] This is really about an extensive restructuring of
  the paper, see Section~\ref{restructure} for my take.
\end{description}

%\bibliographystyle{plain}
\bibliographystyle{abbrv}
%\bibliographystyle{jbact}
%\bibliographystyle{splncs03}
\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}

\end{document}

### 2015-03-13 sanity/letter.tex
Mostly finished letter
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1083 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\item Number of occurences is now explained in the caption. The
+\item Number of occurrences is now explained in the caption. The
   percentage numbers have been hacked to display well.
@@
   of the descriptive statistics, and I understand that this may seem
-  disorganized, and that the desire may emerge to see them nicely
+  disorganised, and that the desire may emerge to see them nicely
   tabulated. They don't lend themselves well to tabular display, since
@@
   LODstats for a list of vocabularies would probably have spared me
-  much trouble, but should not challenge the validity much.
+  much trouble, but should not challenge the validity much. However, I
+  acknowledge that I have not quite understood what
+  \cite{stateoflod2014} did, as the LOD cloud is now quite a lot
+  larger than the BTC2014 dataset, but the website quoted in the
+  review has only crawled 188 MTriples, which isn't at all large.
 
@@
   standard. Second, since it is my agenda to spread knowledge about
-  foundational Web technologys, I would like to follow this
+  foundational Web technologies, I would like to follow this
   suggestion. However, to satisfy all the other requirements, I found
-  this hard to prioritize, after all this is about the HTTP standard,
+  this hard to prioritise, after all this is about the HTTP standard,
   a standard that is fundamental to all practical contributions to the
@@
   keeping the objectives, but detailing how meeting those objectives
-  realize the contributions.
+  realise the contributions.
 \item [Preliminaries and Fundamentals] I have, as noted about, further
@@
 
+\subsection{Histogram breaks}\label{histogramstuff}
+
+The reviewer remarks that the diagrams do not follow clean
+mathematical mechanics. This is a very interesting comment, and one
+that I have devoted much thought to both while writing the paper, and
+after receiving the reviews.
+
+First, it is important to note that histograms, mosaic plots,
+boxplots, density plots are all ways to visualise summarised
+distributions of data. As summarisations, they can be seriously
+misleading, even deceiving, if done wrong. Moreover, I acknowledge
+that violating ``mathematical mechanics'' in histograms is a
+frequently seen technique to ``lie with statistics''.
+
+However, the key goal of the paper is to communicate some essential
+characteristics, and I had already before the paper was submitted used
+a logarithmic scale in those plots, like the reviewer suggested. They
+did not communicate well, since we do not live on a scale of seconds
+when we talk beyond a few minutes.
+
+I had not, however used both axis, as the reviewer suggested, so I
+have now done that, see Figure~\ref{fig:log}.
+
+\begin{figure}[h!]
+  \centerline{%
+    \includegraphics[width=.9\textwidth]{logandtime.pdf}}
+  \caption{Histogram counting all standards-compliant freshness
+    lifetimes found. The data is presented on a logarithmic scale,
+    with the lower x-axis giving the lifetime in seconds and the upper
+    x-axis dividing into ``off'', and then times in usual
+    calendar. Sans polishing.}
+  \label{fig:log}
+\end{figure}
+
+I still opted not to polish and include this into the paper, as
+including the axis on top didn't solve the fundamental problem of
+communication: One doesn't have any everyday feeling for how long
+100~000 seconds is, that is seldomly of interest, rather, one is more
+likely to ask how many months a cache may be valid, and in
+Figure~\ref{fig:log} that question cannot be answered.
+
+Further, if one is interested in the freshness for a resource at that
+level of detail, it is more likely that the primary interest is actual
+change frequency, a topic that is already better covered by DyLDO \cite{dyldo2}.
+
+The only advocacy I derive from these graphs is the suggestion that
+cache prohibition is misguided, which is something that the violation
+of mathematical mechanics doesn't change. Other than that, the
+intention here is purely descriptive, and that is better served with
+the current breaks.
+
+\subsection{Restructure}\label{restructure}
+
+Review 6 asked for a major restructure of the paper. I have noted that
+most of the other reviewers found that the the clarity was good, also
+across expertise levels: Both reviewer 4, who indicated expert
+confidence and reviewer 1, who indicated confidence 3, rather the
+paper with ``very clear''. Reviewer 5 also write that the paper is
+very well presented even though they give it a 3 rating on
+clarity. I'm therefore wary about making changes that may ruin the
+clarity for the majority, to satisfy what seems to be a minority.
+
+I have therefore rather tried to understand what reviewer 6 finds
+problematic about it than take their suggestions literally. I have
+therefore done little to the Analysis section, apart from providing an
+initial overview.
+
+Clearly, reviewer 6 as found the diagrams as not conforming to usual
+form, and they are like reviewer 2, somewhat confused by all the
+numbers that appear in the text, which indeed is somewhat problematic
+but not easily improved.
+
+However, I think the main concern of reviewer 6 is that they put an
+emphasis of the recommendations that I didn't intend. The paper is
+first and foremost meant descriptive, and that's the reason the
+Recommendations is just a tiny section at the end. The only
+recommendation that is derived directly from the paper is the one on
+cache prohibitions. The rest of the recommendations are based in part
+of that I've been working with this for a decade.
+
+Therefore, I have rather than restructured the paper, clarified this
+in the Recommendations, but also in the contributions, and also
+changed wording here and there. I hope this makes it possible to read
+the paper with different eyes, and that it will be clearer to
+everyone.
+
 %\bibliographystyle{plain}

### 2015-03-13 sanity/queries.R

@@ -91,9 +91,11 @@ hardvec <- data$results$fresh
 hardvec[hardvec <= 0] <- 0
 hardveclog <- log10(hardvec)
 hardveclog[hardveclog <= 0] <- -1
+pdf(file="logandtime.pdf", height=5.2,width=8)
 hardvecloghist <- hist(hardveclog, axes=F, main=NULL)
 axis(side=2, at=seq(0, max(hardvecloghist$counts), 20))
 axis(side=1, at=hardvecloghist$breaks, label=c(0, 10^(hardvecloghist$breaks[-1])) )
 tblog <- log10(timebuckets(hardvec)$points)
 tblog[tblog < 0] <- -1
 axis(3, at=tblog, label=c(timebuckets(hardvec)$names, "Max"))
+dev.off()

### 2015-03-18 sanity/paper1.tex
Input the bib
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1100 43ef1190-90fd-4dd7-a769-d0121f02b80e
Final proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1101 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-We also looked for other headers in \ref{tab:headers}. We found
+We also looked for other headers in Table~\ref{tab:headers}. We found
 \httph{Pragma} (archaic HTTP 1.0 header) in 287 responses, but except
@@
 
-For SPARQL endpoints, it is interesting to examine the queries in a
-shared proxy, e.g. an institutional cache or a Content
-Delivery Network for frequent subpatterns, and prefetch and cache such
-intermediate results so that subsequent queries can be answered in
-part by the proxy.
-
 We believe that  \cite{ldf1} and \cite{hogan2014paths} prematurely
@@
 Furthermore, by using similarity measures, like those discussed in
-\cite{dividino2013following}, proxies can proactively prefetch data
-into a cache to help answer queries. A cost model that takes into
-account several data access methods, those described by
-\cite{hogan2014paths} as well as  the novel work in \cite{ldf1} may be
-a key ingredient in enabling efficient SPARQL evaluation on the Web.
+\cite{dividino2013following}, shared proxies, e.g. an institutional
+cache or a Content Delivery Network, can examine queries for frequent
+subpatterns and proactively prefetch data into a cache to help answer
+queries on the proxy. A cost model that takes into account several
+data access methods, those described by \cite{hogan2014paths} as well
+as the novel work in \cite{ldf1} may be a key ingredient in enabling
+efficient SPARQL evaluation on the Web.
 
@@
 %\bibliographystyle{plain}
-\bibliographystyle{abbrv}
+%\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}
 %\bibliographystyle{splncs03}
-\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}
+%\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}
+\input{paper1.bbl}
 

### 2015-03-19 sanity/paper1.tex
Put bibtex back in
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1102 43ef1190-90fd-4dd7-a769-d0121f02b80e
Helen proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1103 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 of the World Wide Web. Out of the 6 documents that make up the HTTP
-1.1 standard, RFC7234 \cite{rfc7234} is entirely devoted to the topic,
-and RFC7232 \cite{rfc7232} defines conditional requests, which is also
-important when constructing caches. As the former notes:
+1.1 standard, RFC7234 \cite{rfc7234} is entirely devoted to the topic.
+RFC7232 \cite{rfc7232} defines conditional requests, and is also
+important when constructing caches. As RFC7234 notes:
 \begin{quote} 
@@
 \end{quote}
-Furthermore, the Architecture of the World Wide Web
-\cite{Jacobs:04:AWW} discusses caching throughout, and the definition
-of the Representational State Transfer (REST) architectural style
+Furthermore, caching is discussed throughout the Architecture of the
+World Wide Web \cite{Jacobs:04:AWW}, and the definition of the
+Representational State Transfer (REST) architectural style
 \cite{Fielding_2000_Architectural-Styles} is partly motivated from the
@@
 in the Internet infrastructure, through so-called Content Delivery
-Networks, is both a large business area and something that could
+Networks, is both a large business area and could
 provide great value to the Semantic Web.
@@
 \end{enumerate}
-The contributions of this paper are to meet these objective by means
+The contributions of this paper are to meet these objectives by means
 of a survey that shows that while the uptake has been moderate,
@@
 anywhere between a server and a client, in which case it may be
-shared, i.e. it may cache responses from several servers served to
-several clients. Another example is an institutional forward proxy,
+shared, i.e. it may cache responses from a number of servers to
+many clients. Another example is an institutional forward proxy,
 which are close to several users. Finally, the User Agent may
@@
 As mentioned, the two documents from the HTTP 1.1 standards suite that
-are relevant for this study are RFC7234, which is named ``Caching''
+are relevant for this study are RFC7234, named ``Caching'',
 and RFC7232, named ``Conditional Requests''. The main difference is
 that the caching standard defines when a response may be reused
-without contacting the origin server at all, whereas the conditional
+without any contact to origin server, whereas the conditional
 requests define how to validate a response by contacting the origin
@@
 
-RFC7234 defines two important headers: \httph{Expires}, whose value is
+RFC7234 defines two important headers. The first of which is \httph{Expires}, whose value is
 a date and time of when the response is considered stale, and
-therefore should not be used; and \httph{Cache-Control}, which allows
+therefore should not be used. The second is \httph{Cache-Control}, which allows
 detailed control of the cache, including a \httph{max-age} field,
@@
 used from the time of the request. \httph{max-age} takes precedence
-over \httph{Expires}. In the following \emph{freshness lifetime} is
+over \httph{Expires}. In this article, \emph{freshness lifetime} is
 understood as the number of seconds that the response may be used
@@
 refer to this as ``standards-compliant caching''. It can also
-be based on heuristics, and Section~4.2.2 in RFC7234 provides some loose
-constraints for such practice, as well as a suggestion for a useful
+be based on heuristics, Section~4.2.2 in RFC7234 provides some loose
+constraints for such practice as well as a suggestion for a useful
 heuristic. This heuristic is based on a fraction of the time lapsed
 between the current time and the modification time given in the
-\httph{Last-Modified} header. This approach therefore still requires
+\httph{Last-Modified} header. This approach still requires
 the Web server to be cooperative to be successful. Commonly, Web
@@
 doesn't burden the content provider with the task of estimating the
-time to live beforehand, but then, it must be able to answer if the
-resource has changed less expensively than serving the entire
+freshness lifetime beforehand. However, the server is then required to be able to answer if the
+resource has changed less expensively than it would be to serve the entire
 response. Either of two headers must be set by the server to achieve
@@
 our concern. For the same reason, caching that happens within the
-SPARQL engine is not our concern.
+SPARQL engine is not relevant.
 
@@
 caching and contrasts this with the unavailability of SPARQL query
-caches, and it is claimed that this is an architectural problem. In
+caches. They assert that this is an architectural problem. In
 \cite{hogan2014paths}, the authors examine cacheable as one of the
@@
 In \cite{umbrich2012hybrid}, the term caching was used in a different
-sense than we do, they rather prefetched an entire dataset to a local
+sense than we use it. They rather prefetched an entire dataset to a local
 store and based on heuristics tried to determine which parts of the
@@
 value of caching based anonymized traces of actual Web usage at a
-major Internet Service Provider. They find that while caching often
-yields little benefit when content is user-generated, they still find
+major Internet Service Provider. They found that while caching often
+yields little benefit when content is user-generated, there is still 
 some potential.
@@
 triples in the returned information resources to see if there is
-information there that may be used to calculate heuristic freshness.
+information that may be used to calculate heuristic freshness.
 
-We make several approaches to ensure that we visit a large and
-representative section of the open Semantic Web. We take SPARQL
+We made several approaches to ensure that we visited a large and
+representative section of the open Semantic Web. We took SPARQL
 Endpoints from the SPARQLES survey \cite{buil2013sparql}, vocabularies
 from Linked Open Vocabularies (LOV) \cite{lov2} and prefix.cc, and we
-augment these data with spidered data from the Billion Triple
+augmented these data with spidered data from the Billion Triple
 Challenge (BTC) 2014 \cite{btc-2014} dataset. Of these, BTC2014 is by
 far the largest, the others are small, curated and targeted
-datasets. However, the size is besides the point, we are only
+datasets. However, the size is besides the point, we were only
 interested in examining as many hosts as possible, and they are still
@@
 from datahub.io (aka CKAN), as well as other sources. To compile a list of
-candidates to further examine, we performed a series of data reduction
+candidates for further examination, we performed a series of data reduction
 steps, manually inspecting the result between each step. The details
-of this process is given in a companion technical report
+of this process are given in a companion technical report
 \cite{kjernsmo_add_survey_2015}.
@@
 BTC2014. It is therefore a fair assumption that BTC2014 fairly well
-represent the breadth of the Semantic Web, momentarily at least.
+represents the breadth of the Semantic Web, momentarily at least.
 
@@
 SPARQL queries to examine the \httph{Server} headers of successful
-responses, with optional clauses matching the standards-compliant
+responses. We used optional clauses matching the standards-compliant
 computed freshness lifetime (which is the ideal) as well as whether
@@
 RDF::LinkedData and RDF::Endpoint, which are partly
-developed by and runs on a server operated by this author. Apart from
+developed by and run on a server operated by this author. Apart from
 those, we see that the they reveal very little about the RDF-specific
@@
 both LOD and SPARQL endpoints. DBPedia has historically been updated
-only a few times a year, but this was probably chosen so as to not
-make a commitment long into the future. It may provide considerable benefits.
+only a few times a year, but this was probably chosen to avoid
+making a commitment far into the future. It may provide considerable benefits.
 
@@
 vocabularies, dataset descriptions or unclassified information
-resources. First, we note that it seems like the distribution of
+resources. Firstly, we note that it seems like the distribution of
 freshness lifetime is quite different for the different types, an
@@
 which is chosen for intuitive interpretation rather than statistical
-rigor). Second, we note that it is often prohibited to cache dataset
+rigor). Secondly, we note that it is often prohibited to cache dataset
 descriptions. This is odd, since statistics about datasets is usually
@@
 Figure~\ref{fig:heuristicall}, we see that the distribution of
-lifetimes are radically different from the case in
+lifetimes is radically different from the case in
 Figure~\ref{fig:hardall}. In this case, we may cache many resources
-for months. Only a handful of resources had changed in the last
+for months. Only a handful of resources changed in the last
 minutes. Since this is based on actual times since last modifications,
@@
 
-First, we compare dates given in \rdfterm{dct:modified} to dates given
+First, we compared dates given in \rdfterm{dct:modified} to dates given
 in \httph{Last-Modified} when both are available for a resource. They
-are often not the same, but it appears that dates in the former are
-further back in time than the latter. We speculate that this may be
+were often not the same, but it appeared that dates in the former are
+further back in time than the latter. We speculated that this may be
 due to that the web-server tracks semantically insignificant changes
@@
 
-\rdfterm{dct:modified} occurs in 2687 triples, but 2487 of these does
+\rdfterm{dct:modified} occured in 2687 triples, but 2487 of these does
 not have the Request-URI of the information resource as their subject,
@@
 
-Hitherto, we have considered the case where the client or proxy does
+So far, we have considered the case where the client or proxy does
 not send a HTTP request to the server when a resource that is present
@@
 standards-compliant freshness lifetime or a heuristic to determine
-it. At the end of this period, or if the previous response had no
+it. At the end of this period, or if the previous response have no
 lifetime, due to lack of information, or to that the server has stated
@@
 Estimating the freshness lifetime is a challenging problem for data
-owners. It must necessarily involve the human users involved in the
+owners. It must necessarily include the human users involved in the
 publishing cycle since they are making a commitment about future
@@
 reflect actual update practices in the Semantic Web is great,
-estimating the actual impact of doing so should be done.
+estimating the actual impact of doing so should be a topic for further research.
 
@@
 and Jonas Smedegaard for proofreading. We also thank the anonymous
-reviewers for extensive reviews.
+reviewers for extensive reviews. Many to Helen Murray for linguistic
+assistance.
 
 %\bibliographystyle{plain}
-%\bibliographystyle{abbrv}
+\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}
 %\bibliographystyle{splncs03}
-%\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}
-\input{paper1.bbl}
+\bibliography{egne,webarch,data,rfc,hypermedia,dynamicity,stat1,specs,vocabs,similarity,optimization}
+%\input{paper1.bbl}
 

### 2015-03-19 sanity/paper1.tex
Final fixes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1104 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 
-\rdfterm{dct:modified} occured in 2687 triples, but 2487 of these does
+\rdfterm{dct:modified} occurred in 2687 triples, but 2487 of these does
 not have the Request-URI of the information resource as their subject,
@@
 and Jonas Smedegaard for proofreading. We also thank the anonymous
-reviewers for extensive reviews. Many to Helen Murray for linguistic
-assistance.
+reviewers for extensive reviews. Finally, many thanks to Helen Murray
+for linguistic assistance.
 

### 2015-03-19 sanity/paper1.tex
Generic web servers
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1105 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 those, we see that the they reveal very little about the RDF-specific
-parts of the underlying server implementation. A quick inspection of
+parts of the underlying server implementation, e.g. Apache
+is a very common generic Web server, the others are also generic. A quick inspection of
 all \httph{Server} headers confirmed that few reveal any further

### 2015-03-20 sanity/report.tex
Introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1106 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
+\documentclass{article}
 \usepackage[english,british]{babel}
-%\documentclass{article}
 \usepackage[utf8]{inputenc}
@@
 
-\title{A survey of HTTP caching implementations on the open Semantic Web}
+\title{Addendum to a survey of HTTP caching on the Semantic Web}
 \author{Kjetil Kjernsmo}
-\institute{Department of Informatics,
-Postboks 1080 Blindern,
-N-0316 Oslo, Norway \email{kjetil@kjernsmo.net}}
 
@@
 
-\todo{qqplot in report}.
+This Technical Report is an addendum to a paper to be published in the
+proceedings of the Extended Semantic Web Conference 2015 titled ``A
+survey of HTTP caching implementations on the open Semantic Web''.
+
+That paper was motivated from the perception that we had that caching
+was much used on the open Semantic Web. This is surprising, since
+caching has been given a very prominent role in the architecture of
+the World Wide Web, and caching based on HTTP headers as defined in
+RFC7234 \cite{rfc7234} and RFC7232 \cite{rfc7232} is prevalent
+elsewhere in the Web.
+
+To understand the actual usage, we conducted an extensive survey of
+actual resources on the Semantic Web. We obtained data from different
+sources, with the breadth of the Semantic Web mainly covered by the
+Billion Triple Challenge (BTC) 2014 dataset  \cite{btc-2014}.
+
+In this addendum, we document in detail several aspects that there
+were not sufficient space for in the conference paper. Since we made
+the assumption that the relevant headers are mostly the same for each
+host, Section~\ref{app:reduction} documents how the large BTC2014 data
+set is reduced to visit a large number hosts, but a small number of
+resources. Then, Section~\ref{app:fetcher} discusses the system we
+developed to visit these resources and extract the information that we
+were interested. Section~\ref{app:stats} documents our main
+statistical method and Section~\ref{app:qqplot} has an analysis of
+valid versus invalid responses.
 
@@
 
-The Billion Triple Challenge 2014 dataset provided data in the form of
-NQuad files. Due to the presence of invalid RDF, we iterated through
-the NQuad files on a line-by-line basis. First, we matched each line
-against a regular expression were lines matching
+The BTC2014 dataset provided data in the form of NQuad files. Due to
+the presence of invalid RDF, we iterated through the NQuad files on a
+line-by-line basis. First, we matched each line against a regular
+expression were lines matching
 \texttt{ontology|endpoint|sparql|vocabular} passed the filter. Then,
@@
 \rdfterm{voaf:Vocabulary} in the object position were also
-accepted\footnote{LOV may be used to resolve these prefixed
-  names}. More classes and properties were considered, but not used in
-the data reduction if they did not occur in the original data.
+accepted\footnote{Linked Open Vocabularies \cite{lov2} may be used to
+  resolve these prefixed names}. More classes and properties were
+considered, but not used in the data reduction if they did not occur
+in the original data.
 
@@
 
-Finally, if the LOV SPARQL endpoint used a URI for the vocabulary that
-was different from the namespace URI (after a normalization step),
-another request would be made to record the selected HTTP headers from
-that as well.
+Finally, if the Linked Open Vocabularies \cite{lov2} SPARQL endpoint
+used a URI for the vocabulary that was different from the namespace
+URI (after a normalization step), another request would be made to
+record the selected HTTP headers from that as well.
 
@@
 
-\bibliographystyle{plain}
-%\bibliographystyle{abbrv}
+\section{Valid versus invalid responses}\label{app:qqplot}
+
+\todo{qqplot in report}.
+
+%\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 %\bibliographystyle{jbact}

### 2015-03-20 sanity/report.tex
Minor changes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1107 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 the Perl framework
-RDF::Trine\footnote{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011}
+RDF::Trine\footnote{\url{https://metacpan.org/release/GWILLIAMS/RDF-Trine-1.011}}
 was used to parse the line. Lines that failed to parse were
@@
 purpose, we developed and released a module
-RDF::Generator::HTTP\footnote{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}
+RDF::Generator::HTTP\footnote{\url{https://metacpan.org/release/KJETILK/RDF-Generator-HTTP-0.003}}
 to CPAN. We recorded whether the conditional request showed that the
@@
 retrieval.  Based on the resulting headers, we let
-libwww\footnote{https://metacpan.org/release/GAAS/libwww-perl-6.04}
+libwww\footnote{\url{https://metacpan.org/release/GAAS/libwww-perl-6.04}}
 calculate both standards-compliant and simple heuristic freshness
@@
 any results to our query, and recorded that if there are. In addition
-to the endpoints registered in the SPARQLES survey, our process found
-18~endpoints that responded with results. For all others, we parsed
-the response, and recorded any errors if the parser concluded the
-content were invalid.
+to the endpoints registered in the SPARQLES survey
+\cite{buil2013sparql}, our process found 18~endpoints that responded
+with results. For all others, we parsed the response, and recorded any
+errors if the parser concluded the content were invalid.
 
@@
 
+Since the same test can be used for binned data in many different
+situations, we used it more informally in other parts of the paper as well.
+
 \section{Valid versus invalid responses}\label{app:qqplot}
 
-\todo{qqplot in report}.
+Since web servers may be configured to instruct clients and proxies to
+cache errors differently from successfully served requests, it is
+interesting to investigate whether we can simply use all cache headers
+in the analysis, or if we should remove unsuccessful responses to
+avoid introducing bias to the analysis of valid responses.
+
+For a visual inspection, we may use a quantile-quantile plot to plot
+the freshness lifetimes of successful responses versus requests that
+failed for some reason. If the distribution of both variables are the
+same, these points will lie on a straight line. In
+Figure~\ref{fig:errorsqq}, we see successful responses vs. the most
+common parse error, and it is clear that these are so different that
+no further formal analysis is necessary. This is one of the reasons
+why we filtered unsuccessful requests before we analyzed the freshness
+lifetimes in the paper.
+
+\begin{figure}[h]
+  \centerline{%
+    \includegraphics[width=.8\textwidth]{errorsqq.pdf}}
+  \caption{A quantile-quantile plot of freshness lifetimes for
+    standard compliant headers with and without errors.}
+  \label{fig:errorsqq}
+\end{figure}
 

### 2015-03-20 sanity/report.tex
Proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@1108 43ef1190-90fd-4dd7-a769-d0121f02b80e

@@
 statistical method and Section~\ref{app:qqplot} has an analysis of
-valid versus invalid responses.
+the distribution of valid versus invalid responses.
 
@@
 \httph{Last-Modified} and \httph{ETag} headers where they existed, we
-first examined if any of the resources were still fresh, but none
+first examined whether any of the resources were still fresh, but none
 were. Wherever the last two headers existed, we added the
@@
 
-The hypothesis tests in this paper is implemented using \emph{contingency
+The hypothesis tests in this paper were implemented using \emph{contingency
   tables} (see e.g. \cite{kn:bj}). This formalism is suited to see if the

### 2015-07-13 drafts/tmpsolutions.tex
Move old dissertation draft out of the way

--- 2012-10-25 dissertation/tmpsolutions.tex
@@
 than the individual parts, it makes sense to develop this theory
-first, and then derive the VoID extensions from that.
+first, and then derive the VoID extensions from that as well as more
+complex results.
 
@@
 the end of the year. If I embark on a study of \SRL{}, this would not
-be possible.
+be possible. There is a risk that it will take more than a year to finish
+a work on this, and then there is little to fall back on.
 

### 2015-07-13 dissertation/introdefhistory.tex
First paragraph

The present work sits in the confluence of several contemporary
efforts in the Semantic Web community. The focus is on query answering
with the SPARQL query language, with emphasis on exploiting the World
Wide Web, but it touches upon query federation, hypermedia, empiricial
methods for evaluating performance, standards compliance and even
philosophy of science. With a scope of this width, the investigation
is shallow, it is hoped that some key insights will provoke further
investigation rather than being the end of any conversations.

\section{Semantic Web -- A definition?}

### 2015-07-14 dissertation/introdefhistory.tex
no definition

@@
 \section{Semantic Web -- A definition?}
+
+The Semantic Web can be described as a machine-readable Web of Data,
+defined by a technology stack defined by the World Wide Web
+Consortium. While individual technologies can be given precise,
+operational definitions, the above definition is naive and can hardly
+be used operationally to derive new knowledge. The Semantic Web is a
+complex, human artefact that must be understood not only in technical
+terms, but also in social, economic and cultural terms. First, the Web
+is not only readable, but also writeable. Whether the essential parts
+of a data integration problem will be solved by e.g. pervasive
+ontology alignment, a linked open vocabularies approach where many
+players adopt a fairly consistent set of vocabularies, or by major
+players are able to force convergence towards e.g. \texttt{Schema.org}
+is an example of the social, economic and cultural mechanisms that
+shape the future of the Semantic Web. Likewise, it may not be the W3C
+Semantic Web stack at all that achieves success, it may be
+microformats\footnote{see e.g. }, microdata\cite{TODO}, or something
+else entirely\footnote{for a thoughtful criticism of Linked Data with
+  a proposed alternative, see \href{http://berjon.com/linked-data/}},
+but again, it is likely that social, economic and cultural factors
+will be as decisive as technical. One may say that e.g. microformats
+cannot help achieve the visions that may be stated for the Semantic
+Web, but that again emphasises visions, not operational definitions.
+
+One such vision is the article by Berners-Lee, Hendler and Lassila in
+Scientific American 2001\cite{TODO}, but it does not attempt such an
+operational definition. Neither \cite{TODO} nor \cite{TODO} contains
+such definitions. However, it is clear that an essential
+characteristic of the Web, and by extension, the Semantic Web is its
+\emph{universality} and therefore, any actor should be as free as
+possible to adapt it to their visions and needs, and so, a pursuit of
+such a definition may not be desirable. Therefore, rather than
+defining it, one should carefully declare one's visions.
+
+\section{A personal history}

### 2015-07-14 dissertation/introdefhistory.tex
First part of personal history

@@
+\section{Overview}
+
+The World Wide Web, or just the Web for short, is a well-known global
+information space invented by Tim Berners-Lee in 1989, further emerged
+in the 1990-ties. It is characterised first and foremost by its
+openness, anyone can set up a computer, connect it to the Internet and
+start serving data or documents from it.
+
+The Semantic Web is an extension of the Web that has been under
+development since 1997\footnote{The first working group draft of the
+  RDF specification is dated 1997-08-01, see
+  \href{http://www.w3.org/TR/WD-rdf-syntax-971002/}.}, to extend the
+Web with languages for expressing information in a machine processable
+form\cite{TODO}.
+
 The present work sits in the confluence of several contemporary
@@
 Semantic Web stack at all that achieves success, it may be
-microformats\footnote{see e.g. }, microdata\cite{TODO}, or something
-else entirely\footnote{for a thoughtful criticism of Linked Data with
-  a proposed alternative, see \href{http://berjon.com/linked-data/}},
-but again, it is likely that social, economic and cultural factors
-will be as decisive as technical. One may say that e.g. microformats
-cannot help achieve the visions that may be stated for the Semantic
-Web, but that again emphasises visions, not operational definitions.
+microformats\footnote{see \href{http://microformats.org/]},
+microdata\cite{TODO}, or something else entirely\footnote{for a
+  thoughtful criticism of Linked Data with a proposed alternative, see
+  \href{http://berjon.com/linked-data/}}, but again, it is likely that
+social, economic and cultural factors will be as decisive as
+technical. One may say that e.g. microformats cannot help achieve the
+visions that may be stated for the Semantic Web, but that again
+emphasises visions, not operational definitions.
 
@@
 such definitions. However, it is clear that an essential
-characteristic of the Web, and by extension, the Semantic Web is its
+characteristic of the Web, and by extension, the Semantic Web, is its
 \emph{universality} and therefore, any actor should be as free as
@@
 \section{A personal history}
+
+I created my first Web page in December 1994, and strongly appreciated
+the value of the fact that I could easily contribute to the Web, and
+therefore, I quickly posted anything that I had of value, for example
+a collection of my best mountaineering pictures, constrained not by
+the platform, but by my own time. Initially, I learnt purely by
+example, i.e. using the ``View Source'' menu item when I encountered
+web pages that I liked. I also had the power to mint identifiers.
+
+With identifiers, and the hypertext language HTML came the fact that I
+could link to anyone, and anyone could link to me was highly
+empowering: Discovering that people I didn't know linked to my
+material was a strong, social reward. After some time, I had tens of
+thousands of visitors every month.
+
+In 1996, I started the Web pages of the Norwegian Skeptics Society
+\texttt{skepsis.no}, and therefore got the control over an entire Web
+server. That was another revelation, as I could actually run code, and
+from that arose the need to not just learn by ``View Source'', but
+start to read the specifications, notably HTML, CSS and
+HTTP. Appreciating the design of these specifications took a while,
+but eventually, the value of orthogonal specifications and separation
+of concerns became apparent.
+
+When the RDF working drafts were first published, I was first
+inundated by the amount of new text to read, but at the same time I
+had a growing realisation that getting information out was not solving
+the problem The Norwegian Skeptics Society needed to solve: We needed
+to push information into closed minds. My initial plan to address that
+was to enable a higher degree of targeting of information, and a Web-wide
+conversation. To do that, I wanted to create a large thesaurus of
+topics of interest, and create annotations of the type ``this article
+is a rebuttal of that article'', and an index that browsers could
+query, so that when a user viewed an article, they would also get a
+critical context. At the same time, Netscape was open-sourced as
+Mozilla, and so, I was quite convinced I could contribute the code
+needed for this to be accepted.
+
+I shared these ideas on some mailing lists in August 1998, and one of
+the persons who responded was Dan Brickley, who would chair the RDF
+working group. He quickly convinced me that RDF was not something to
+be afraid of, quite the contrary, it was exactly what I needed.
+
+However, all the above had happened on my spare time, as the topic of
+my study was something else entirely: Cosmology, and so the interest
+laid dormant for a few years as I finished my Cand. Scient.-degree. 
+After that, I got some limited funding to write what today would be
+called a Semantic Content Management System. The project was far too
+ambitious and largely a failure, but it got me some valuable
+experience with Semantic Web technology.
+
+It was not until I joined Opera Software in 2005 to work on the now
+defunct Opera Community social networking site, that I had a bit more
+time to work on Semantic Web technology. When I joined, they had
+rudimentary support for an RDF/XML-based ``Friend of a Friend'' (FOAF)
+export, which I improved before the initial release. At the time, it
+was not anticipated that everything would have a URI, so Dan Brickley
+was a proponent of the view that most things, including people, would
+be identified mainly by their properties, e.g. their email address. It
+was a certain tension between that view and the ``give everything a
+URI''-movement heralded by Tim Berners-Lee at the time, but I decided
+to place myself in between, by giving everything a URI, but at the
+same time ensure that enough properties would identify people. I also
+included outward links, so that people could integrate the database
+export we provided with their hand-maintained FOAF. I also added
+support for photo gallery metadata and linking tags as given in a
+personal SKOS-ontology to Wordnet to enable assigning clearer meaning
+to tags. This effort was acknowledged in the Linked Data Design
+Issue\cite{TODO} that formed the basis for the more recent Linking
+Open Data project.
+
+However, at this point, I started to realise that it was difficult to
+develop applications around traversing the Linked Data, and that
+prompted me to consider the possibility to offer the ability to
+execute arbitrary queries. On 2005-11-30, I published the first public
+SPARQL endpoint with actual production data\footnote{see
+  \href{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
+triples, but it quickly grew to \~15~million triples.
+
+Unfortunately, it wasn't allotted time to enhance and document the
+SPARQL Endpoint further, and it didn't see much practical use. 
+I was, however, allowed to join several World Wide Web Consortium
+groups. First, the Web Content Label Incubator Group, which was tasked
+with discussing a replacement for the then archaic Platform for
+Internet Content Selection, and resolved to use RDF for this
+purpose, and it transitioned into the POWDER Working Group, which
+provided the specifications.
+
+More importantly, I joined the Semantic Web Education and Outreach
+(SWEO) Interest Group, where I admitted my frustration with the lack
+of practical progress with Semantic Web technologies and the relative
+lack of uptake. This sentiment was echoed by the late Aaron Swartz,
+who was asked by the group on his opinion, and he emailed the
+following statement\footnote{see
+  \href{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
+\begin{quote}
+I'm not sure what SWEO is, but my feeling is and pretty much always
+has been that the Semantic Web people need to start putting together
+Genuinely Useful stuff that can be done Right Now.
+\end{quote}
+
+This was one of the motivations I had for starting the SWEO Community
+Projects, where a questionnaire was posted on the Web, and people from
+the Semantic Web community were challenged to come up with a project
+that would provide practical benefits in the short term. We received
+10 proposals, out of which 3 was selected for backing by the SWEO
+IG. None of them was successful, but a fourth proposal, submitted by
+Chris Bizer and Richard Cyganiak built momentum quickly and was
+therefore also selected for backing, despite some criticism by several
+IG members, myself included. The proposal was titled ``Linking Open
+Data'' and sought to take already abundant open data and model it
+using guidelines of the Linked Data Design
+Issue\cite{TODO}. LODstats\cite{TODO} provides extensive statistics on
+the results of this project, and has of this writing seen nearly
+10~000 data sets.

### 2015-07-14 dissertation/introdefhistory.tex
History

@@
 10~000 data sets.
+
+At this point, I left Opera to work as a consultant on Semantic Web
+technologies. One of the projects that were successfully completed was
+called Sublima\cite{TODO}. It was driven entirely by RDF and SPARQL,
+and featured faceted navigation, navigation in a thesaurus, and full
+text search. Experiences from this project influenced requirements of
+SPARQL 1.1, and I was an editor of the SPARQL 1.1 Features and
+Requirements specification\cite{TODO}. Parts of the property paths
+feature and aggregate queries were used extensively. Write operations
+were also implemented with SPARQL using what was then specified only
+in a member submission to the W3C\cite{TODO}. This was the basis for
+the update language in SPARQL 1.1 but differed substantially in
+surface syntax, partly because our work showed how certain patterns
+could be simplified. Another project, based on much of the same code
+base replaced the SKOS ontology of the Sublima project with an OWL
+ontology, and also used reasoning to aid navigation in multimedia
+libraries.
+
+However, it became clear that I could not have sufficient time in the
+industry to pursue what was developing as my main interest, query
+answering over data on the open Web, for generic application
+development. 16 years of experience thus motivates the research done
+in this thesis.

### 2015-07-14 dissertation/introdefhistory.tex
limit per resource

@@
 surface syntax, partly because our work showed how certain patterns
-could be simplified. Another project, based on much of the same code
+could be simplified.
+
+Two features that arose from Sublima requirements were not accepted
+for SPARQL 1.1, however: Full text index and a feature known as
+``Limit Per Resource''\footnote{see
+  \href{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
+latter feature arises from the fact that usually, the user is
+indifferent to the number of solutions to a query, and therefore
+limiting by the number of solutions is unhelpful. This is especially
+true since other features of SPARQL, such as optional clauses, are
+helpful in dealing with heterogeneous data. In the Sublima case, it
+was interesting to limit by the number of articles that were returned,
+but the number of solutions depended on the number of authors of an
+article and the concepts used to classify an article. To achieve the
+desired effect with SPARQL 1.1, one would have to write a very complex
+subquery, which is undesirable for such an important feature.
+
+Another project, based on much of the same code
 base replaced the SKOS ontology of the Sublima project with an OWL

### 2015-07-14 dissertation/introdefhistory.tex
port web to semweb

@@
 but eventually, the value of orthogonal specifications and separation
-of concerns became apparent.
+of concerns became apparent. Also, at the same time, I realised that I
+could be standing on the shoulders of giants by using Free Software.
 
@@
 experience with Semantic Web technology.
+
+I was convinced that the things that made the Web great to work with
+had to be present in the Semantic Web as well: Everybody is empowered
+to publish, find and quickly make use of Free Software, learn by
+``View Source'' (an idea which becomes even more powerful with
+semantics, as if you can understand what you are viewing simply by
+looking at the message, you can begin processing it), anyone can mint
+identifiers, anybody can reuse those identifiers in their own
+contexts, and eventually graduate to reading technical materials as
+needed. Many of these things are virtues of the Semantic Web as it is
+an extension of the Web, others have not been sufficiently cultivated
+or even appreciated.
 
@@
 10 proposals, out of which 3 was selected for backing by the SWEO
-IG. None of them was successful, but a fourth proposal, submitted by
+IG. None of them were successful, but a fourth proposal, submitted by
 Chris Bizer and Richard Cyganiak built momentum quickly and was

### 2015-07-15 dissertation/introdefhistory.tex
add labels
fix ref title

@@
-\section{Overview}
+\section{Overview}\label{sec:introoverview}
 
@@
 
-\section{Semantic Web -- A definition?}
+\section{Semantic Web -- A definition?}\label{sec:noref}
 
@@
 
-\section{A personal history}
+\section{A personal history}\label{sec:history}
 
@@
 SPARQL 1.1, and I was an editor of the SPARQL 1.1 Features and
-Requirements specification\cite{TODO}. Parts of the property paths
+Rationale specification\cite{TODO}. Parts of the property paths
 feature and aggregate queries were used extensively. Write operations

### 2015-07-15 dissertation/motivation.tex
large scale vision

\section{Motivation}\label{sec:motivation}

There is a class of diverse, yet structurally similar, problems people
face nearly every day: When we book a flight, we may be interested in
a romantic getaway, yet we are forced to select the airport of
destination first. We often end up buying the same groceries because
we don't know what will taste well together. Last time I bought a car,
the two most important features was that it had 4 wheel drive and that
the trunk was long enough to fit a full-size baby stroller
length-wise. To find a property to build a home, I took into account
the sun conditions, that it was close to public transport that would
transport me to campus within a certain time frame, close to
ski tracks, abundance of uranium in the ground, number of children for
my children to play with, and an affordable price level. And once the
house was being built, I had to choose the toilet seat depending on
the frame in the wall and even a toilet seat that would fit.

The overarching problem is that even though each of these problems are
rarely encountered, there are so many such problems, that we encounter
them all the time. I would like to see a Semantic Web where the
plumber delivers his offer in a manner so that the open options can be
distinguished from the constraints, where a query can be formulated to
find the area I'd like to build my house based on such extensive
criteria, where the very rare comparison between a collapsed baby
stroller and car trunk can be done, where I will be able to cook
better food without being a chef and where I can book my travel based
on what I want to experience rather than make me choose a destination
based on lacking information.

Now, all this information exists on the Web, or at the very least, in
some database, and for the most part, the above problems can be
solved by surfing many websites. We are, however, reminded of ``Connolly's
Lament'' (due to Dan Connolly, one of the Semantic Web's early champions):
\begin{quote}
 The bane of my existence is doing things I know the computer could do
 for me.
\end{quote}

### 2015-07-15 dissertation/motivation.tex
most of motivation

@@
 
-The overarching problem is that even though each of these problems are
-rarely encountered, there are so many such problems, that we encounter
-them all the time. I would like to see a Semantic Web where the
-plumber delivers his offer in a manner so that the open options can be
-distinguished from the constraints, where a query can be formulated to
-find the area I'd like to build my house based on such extensive
-criteria, where the very rare comparison between a collapsed baby
-stroller and car trunk can be done, where I will be able to cook
-better food without being a chef and where I can book my travel based
-on what I want to experience rather than make me choose a destination
-based on lacking information.
+Of fundamental importance is that is that even though each of these
+problems are rarely encountered, there are so many such problems that
+we encounter them frequently. I would like to see a Semantic Web
+where the plumber delivers his offer in a manner so that the open
+options can be distinguished from the constraints, where a query can
+be formulated to find the area I'd like to build my house based on
+such extensive criteria, where the very rare comparison between a
+collapsed baby stroller and car trunk can be done, where I will be
+able to cook better food without being a chef and where I can book my
+travel based on what I want to experience rather than make me choose a
+destination based on lacking information.
 
@@
 \end{quote}
+
+Moreover, the problem could be solved by writing programs to integrate
+the variety of data sources for each scenario, but this is clearly not
+sustainable, as each particular integration is of interest to very
+few, but the cost of the integration increases for every integration
+task. Fewer people to pay for a more expensive task quickly becomes
+economically infeasible.
+
+Technically, it might be an easier problem to solve if the data could
+be centralized, and that solution has been argued by
+\cite{DBLP:conf/semweb/BetzGHS12}. Nevertheless, even if all their
+arguments remain valid from a technical viewpoint, the most important
+arguments against centralisation are social: If important components
+of the Semantic Web is under the control of some central authority, it
+is likely that one cannot publish anything, link to anything, or
+develop any application without permission. It may create an unhealthy
+power structure, which may create commercial or intellectual
+hegemonies. Tim Berners-Lee discussed this in \cite{TODO} and an
+interview in Wired Magazine in 2014\footnote{see
+  \href{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}.
+This is the key motivation to still design a distributed Semantic Web,
+regardless of the technical difficulties that may arise.
+
+The naive enthusiast may now forward the claim that Linked Data is the
+solution and the whole solution, and at times of hubris, I may myself
+have been guilty of conjecturing that sitting down to write the code
+based on what we already know would solve the problem.
+
+At times of doubt, the large number of unsolved problems, theoretical
+as well as practical become apparent. In the following, I will try to
+enumarate the problems I have foreseen, discuss their place in current
+research, and what problems I have attempted to tackle in this
+dissertation.
+
+%%  LocalWords:  hegemonies

### 2015-07-16 dissertation/motivation.tex
more or less finish evaluation

@@
 
+Both the idea of a decentralised Web and the idea that SPARQL
+endpoints can be made openly available for any query on the Internet
+seems to go against the conventional wisdom of the database
+community. We are, however, motivated to solve this problem by the
+social requirement of decentralisation and by the use cases mentioned
+at the start of this section.
+
+It is also important to frame the question correctly: We want to
+answer queries, we do not require that any remote server accepts the
+entire burden of evaluating queries. It is becoming quite clear that
+the current practice of evaluating any arbitrary SPARQL query on a
+remote server is economically unsustainable, and it is made clear by
+the instability of SPARQL endpoints on the Web \cite{TODO}.
+
+However, we note that the Internet infrastructure is accommodating for
+solving parts of this problem with the presence of caches at several
+different levels. Some clients also have significant processing
+capacity, and these are among the factors that could make SPARQL query
+answering on the open Web practically feasible. The focus of this work
+is to ease the burden of servers by allowing other parties to take
+part of the query evaluation.
+
 %%  LocalWords:  hegemonies

### 2015-07-16 dissertation/related.tex
Some important related work

\section{Related work}\label{sec:related}

There is a substantial amount of relevant literature and ongoing
research that is well aligned with the present work, either because
the authors share parts of the motivation, or more serendipitous
reuse.

Verborgh et al. \cite{TODO} share many of the observations and goals
that motivates this study. They introduce a hypermedia system that can
be used to answer individual triple patterns (known as Triple Pattern
Fragments (TPF)). Their stated goal is to transfer as much as possible
of the burden of evaluating a query from the server to the
client. This can happen because a simple pattern match for a single
triple pattern need not have a SPARQL parser, nor a query
planner. Moreover, the server may materialise responses to frequent
triple patterns and store them in a file system, which can be managed
by a simple Web server. The results may also be paged, so that each
individual response may be kept small. However, TPF mandates that
every response must contain a cardinality estimate for every triple
pattern, an operation that may be quite expensive for the server, and
indeed, this information may be so expensive to provide that SPARQL
planners may not provide it to keep the cost of planning itself
down. In \cite{TODO}, they demonstrated SPARQL evaluation over TPF,
and further improvements have been made in more recent papers.

Olaf Hartig has a large number of papers that are adjacent to the
present work, including his Ph.D. dissertation \cite{TODO} where he
explores querying Linked Data, i.e. query data that is not contained
in an a-priory defined collection of RDF data, but where a resources
are traversed to compute a result. Hartig provides solid formal
foundations for such query processing, as well as computational
feasibility, soundness and completeness. The most relevant paper of
this work is \cite{TODO}.

Acosta et al. \cite{TODO} has ongoing work named ``SHEPHERD'' to
create a SPARQL processor that takes into account observations done by
the SPARQLES survey \cite{TODO}, to decompose a SPARQL query into
subqueries that is indicated to be easier for the server to evaluate,
and so is well aligned with our goal to ease the load of the
server. It is not quite clear from the brief description in 
\cite{TODO} whether the SHEPHERD system can cache results locally.

Montoya et al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query
federation system that takes advantage of client-side data
replication.

### 2015-07-16 dissertation/related.tex
Most of related work

@@
 
-Verborgh et al. \cite{TODO} share many of the observations and goals
+Verborgh et al. \cite{ldf1} share many of the observations and goals
 that motivates this study. They introduce a hypermedia system that can
@@
 replication.
+
+Umbrich et al. \cite{umbrich2012hybrid} considered a situation where
+parts of the graph changes at different paces. For the slow-changing
+parts, they prefetch an entire dataset to a local store to relieve
+remote endpoints from some processing, and thereby speed up query
+responses.
+
+Martin et al. \cite{sparqlproxy} implemented a reverse caching proxy
+that controlled the changes to the dataset, and so could relieve the
+endpoint of the burden of evaluating all queries.
+
+While the approaches are diverse, these studies pull in the same
+direction: Motivated by the problems of maintaining SPARQL endpoints,
+they all seek to share the burden, and so make query answering
+economically feasible.
+
+For more detailed related work, see the corresponding section in the
+individual papers.

### 2015-07-17 dissertation/problems.tex
some sections

\section{Problems}\label{sec:problems}

\subsection{Performance problems}

\subsection{Stability problems}

\subsection{Problems with evaluations}

\subsection{Development problems}

### 2015-07-18 dissertation/introdefhistory.tex
Add link to timbl's testimony
Some bugfixing

@@
   RDF specification is dated 1997-08-01, see
-  \href{http://www.w3.org/TR/WD-rdf-syntax-971002/}.}, to extend the
+  \url{http://www.w3.org/TR/WD-rdf-syntax-971002/}.}, to extend the
 Web with languages for expressing information in a machine processable
@@
 Semantic Web stack at all that achieves success, it may be
-microformats\footnote{see \href{http://microformats.org/]},
+microformats\footnote{see \url{http://microformats.org/}},
 microdata\cite{TODO}, or something else entirely\footnote{for a
   thoughtful criticism of Linked Data with a proposed alternative, see
-  \href{http://berjon.com/linked-data/}}, but again, it is likely that
+  \url{http://berjon.com/linked-data/}}, but again, it is likely that
 social, economic and cultural factors will be as decisive as
@@
 an extension of the Web, others have not been sufficiently cultivated
-or even appreciated.
+or even appreciated. Moreover, my experience corresponds to the
+virtues that Tim Berners-Lee has attributes the Web's success
+to\footnote{see
+  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}
 
@@
 SPARQL endpoint with actual production data\footnote{see
-  \href{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
+  \url{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
 triples, but it quickly grew to \~15~million triples.
@@
 following statement\footnote{see
-  \href{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
+  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
 \begin{quote}
@@
 ``Limit Per Resource''\footnote{see
-  \href{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
+  \url{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
 latter feature arises from the fact that usually, the user is

### 2015-07-18 dissertation/motivation.tex
More on motivation
Some bugfixing

@@
 interview in Wired Magazine in 2014\footnote{see
-  \href{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}.
+  \url{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}
 This is the key motivation to still design a distributed Semantic Web,
@@
 is to ease the burden of servers by allowing other parties to take
-part of the query evaluation.
+parts of the query evaluation.
 

### 2015-07-18 dissertation/introdefhistory.tex
Add some references

@@
 Web with languages for expressing information in a machine processable
-form\cite{TODO}.
+form\cite{semwebroadmap}.
 
@@
 microformats\footnote{see \url{http://microformats.org/}},
-microdata\cite{TODO}, or something else entirely\footnote{for a
+microdata\cite{microdata}, or something else entirely\footnote{for a
   thoughtful criticism of Linked Data with a proposed alternative, see
@@
 One such vision is the article by Berners-Lee, Hendler and Lassila in
-Scientific American 2001\cite{TODO}, but it does not attempt such an
-operational definition. Neither \cite{TODO} nor \cite{TODO} contains
+Scientific American 2001\cite{berners2001semantic}, but it does not attempt such an
+operational definition. Neither \cite{semwebroadmap} nor \cite{berners2000weaving} contains
 such definitions. However, it is clear that an essential
@@
 to\footnote{see
-  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}
+  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.
 
@@
 to tags. This effort was acknowledged in the Linked Data Design
-Issue\cite{TODO} that formed the basis for the more recent Linking
+Issue\cite{linkeddataissue} that formed the basis for the more recent Linking
 Open Data project.
@@
 using guidelines of the Linked Data Design
-Issue\cite{TODO}. LODstats\cite{TODO} provides extensive statistics on
-the results of this project, and has of this writing seen nearly
-10~000 data sets.
+Issue\cite{linkeddataissue}. LODstats\cite{auer2012lodstats} provides
+extensive statistics on the results of this project, and has of this
+writing seen nearly 10~000 data sets.
 
@@
 technologies. One of the projects that were successfully completed was
-called Sublima\cite{TODO}. It was driven entirely by RDF and SPARQL,
+called Sublima\cite{sublima}. It was driven entirely by RDF and SPARQL,
 and featured faceted navigation, navigation in a thesaurus, and full
 text search. Experiences from this project influenced requirements of
-SPARQL 1.1, and I was an editor of the SPARQL 1.1 Features and
-Rationale specification\cite{TODO}. Parts of the property paths
+SPARQL 1.1 \cite{sparql11query}, and I was an editor of the SPARQL 1.1 Features and
+Rationale specification\cite{sparql11new}. Parts of the property paths
 feature and aggregate queries were used extensively. Write operations

### 2015-07-19 dissertation/motivation.tex
All references so far

@@
 power structure, which may create commercial or intellectual
-hegemonies. Tim Berners-Lee discussed this in \cite{TODO} and an
+hegemonies. Tim Berners-Lee discussed this in \cite{berners2000weaving} and an
 interview in Wired Magazine in 2014\footnote{see
@@
 remote server is economically unsustainable, and it is made clear by
-the instability of SPARQL endpoints on the Web \cite{TODO}.
+the instability of SPARQL endpoints on the Web \cite{buil2013sparql}.
 

### 2015-07-19 dissertation/related.tex
All references so far

@@
 planners may not provide it to keep the cost of planning itself
-down. In \cite{TODO}, they demonstrated SPARQL evaluation over TPF,
+down. In \cite{verborgh2014querying}, they demonstrated SPARQL evaluation over TPF,
 and further improvements have been made in more recent papers.
@@
 Olaf Hartig has a large number of papers that are adjacent to the
-present work, including his Ph.D. dissertation \cite{TODO} where he
+present work, including his Ph.D. dissertation \cite{hartig2014querying} where he
 explores querying Linked Data, i.e. query data that is not contained
@@
 foundations for such query processing, as well as computational
-feasibility, soundness and completeness. The most relevant paper of
-this work is \cite{TODO}.
+feasibility, soundness and completeness. The most relevant paper to
+this work is \cite{hartig2011caching}.
 
-Acosta et al. \cite{TODO} has ongoing work named ``SHEPHERD'' to
+Acosta et al. \cite{acosta2014shepherd} has ongoing work named ``SHEPHERD'' to
 create a SPARQL processor that takes into account observations done by
-the SPARQLES survey \cite{TODO}, to decompose a SPARQL query into
+the SPARQLES survey \cite{buil2013sparql}, to decompose a SPARQL query into
 subqueries that is indicated to be easier for the server to evaluate,
@@
 server. It is not quite clear from the brief description in 
-\cite{TODO} whether the SHEPHERD system can cache results locally.
+\cite{acosta2014shepherd} whether the SHEPHERD system can cache results locally.
 

### 2015-07-19 dissertation/introdefhistory.tex
All references so far
Make darobin a ref

@@
 microformats\footnote{see \url{http://microformats.org/}},
-microdata\cite{microdata}, or something else entirely\footnote{for a
+microdata\cite{microdata}, or something else entirely. For a
 thoughtful criticism of Linked Data with a proposed alternative, see
-  \url{http://berjon.com/linked-data/}}, but again, it is likely that
-social, economic and cultural factors will be as decisive as
-technical. One may say that e.g. microformats cannot help achieve the
-visions that may be stated for the Semantic Web, but that again
-emphasises visions, not operational definitions.
+\cite{darobin1}, but again, it is likely that social, economic and
+cultural factors will be as decisive as technical. One may say that
+e.g. microformats cannot help achieve the visions that may be stated
+for the Semantic Web, but that again emphasises visions, not
+operational definitions.
 
@@
 were also implemented with SPARQL using what was then specified only
-in a member submission to the W3C\cite{TODO}. This was the basis for
+in a member submission to the W3C\cite{seaborne2008sparql}. This was the basis for
 the update language in SPARQL 1.1 but differed substantially in

### 2015-07-19 dissertation/problems.tex
new header
First problem def

@@
 
-\subsection{Performance problems}
+\subsection{Performance problems}\label{sec:perfproblems}
 
-\subsection{Stability problems}
+\subsection{Stability problems}\label{sec:stabproblems}
 
-\subsection{Problems with evaluations}
+\subsection{Problems of shared understanding}\label{sec:semproblems}
 
-\subsection{Development problems}
+\subsection{Problems with evaluations}\label{sec:evalproblems}
+
+\subsection{Development problems}\label{sec:devproblems}
+
+There are several issues that make Semantic Web development difficult,
+that are connected with development in current programming languages
+and current Semantic Web libraries. As pointed out in \cite{darobin1},
+there are few things that are simpler in contemporary programming than
+to parse a string containing a tree-formatted data structure into a
+tree that can be accessed directly. However, RDF makes the assumption
+that the natural form of data is not a tree, but a graph, and in the
+general case, it must be addressed as such.
+
+\begin{problem}\label{prob:graph}
+Addressing graph data in a programming language.
+\end{problem}

### 2015-07-19 dissertation/problems.tex
a bit of restructuring

@@
 
-\subsection{Performance problems}\label{sec:perfproblems}
+\subsection{Performance and stability problems}\label{sec:perfproblems}
 
-\subsection{Stability problems}\label{sec:stabproblems}
+\begin{problem}
+The complexity of SPARQL is high.
+\end{problem}
 
@@
 that are connected with development in current programming languages
-and current Semantic Web libraries. As pointed out in \cite{darobin1},
-there are few things that are simpler in contemporary programming than
-to parse a string containing a tree-formatted data structure into a
-tree that can be accessed directly. However, RDF makes the assumption
-that the natural form of data is not a tree, but a graph, and in the
-general case, it must be addressed as such.
+and current Semantic Web libraries.
+
+This is very important, because even though we have argued that the
+programmer needs to be eliminated from the data integration task,
+there are several functions where they need to be involved, including
+end-user application development, and in the short term, certainly
+also the data integration task.
+
+Following the discussion in Section~\ref{sec:history}, the
+availability of tools and examples so that the vast majority of active
+Web developers can quickly publish and make use of RDF is a key to the
+success of the Semantic Web.
+
+As pointed out in \cite{darobin1}, there are few things that are
+simpler in contemporary programming than to parse a string containing
+a tree-formatted data structure into a tree that can be accessed
+directly. However, RDF makes the assumption that the natural form of
+data is not a tree, but a graph, and in the general case, it must be
+addressed as such.
 
@@
 \end{problem}
+
+The initial learning by example by ``View Source'' of HTML markup can
+be generalised to RDF by considering the term \emph{hypermedia}. The
+most important aspect of hypermedia is, as argued in
+\cite{Fielding_2000_Architectural-Styles}, that hypermedia can be used
+to drive the interaction of applications, and that all the information
+needed to do so is in the messages that are passed between the server
+and the client, no information beyond that is needed. As such, the
+importance of ``View Source'' extends well beyond the original
+motivation of learning by example. More concretely, hypermedia can be
+used to describe simple query interfaces, define how read-write
+operations are to be made, etc. %TODO more here
+
+\begin{problem}\label{prob:lapis}
+Developers who are not well versed in RDF need a readily
+understandable format that details how they can interact with an RDF
+server that offers a read-write interface.
+\end{problem}
+
+\begin{problem}\label{prob:tpf}
+Since the SPARQL language is described in an external specification,
+it cannot be hypermedia.
+\end{problem}

### 2015-07-19 dissertation/problems.tex
More development problems

@@
 \end{problem}
+
+Until recently, a query engine would need to break down a SPARQL query
+to individual triple pattern matches, which would then use for example
+the \jcode{listStatements()} method of Jena, \jcode{filter()} method
+of Sesame or \pcode{get\_statements} of \pmodule{RDF::Trine}, so that only
+individual triple patterns could be evaluated against the underlying
+triplestore.  The latter also had a \pcode{get\_pattern} method that
+could be implemented if the underlying store had a way to optimise
+Basic Graph Patterns. Sesame, on the other hand, allow implementations
+to get a query representation, which it must then evaluate.
+Nevertheless, this resulted in many problems:
+
+\begin{problem}
+By breaking down the query down to individual triple patterns, the
+query engine cannot take advantage of optimisations that involve
+multiple triple patterns or other parts of the query.
+\end{problem}
+
+\begin{problem}
+If methods such as \pcode{get\_pattern} are added for every part of the
+query, it would increase the complexity of the API dramatically.
+\end{problem}
+
+\begin{problem}
+By merely passing the entire query representation to an
+implementation, the burden of evaluating the entire query is also
+transferred, which makes it harder to use default implementations for
+most of the query.
+\end{problem}
+
+The same kind of problems occur in many operations, such as parsing
+and serialisation, where an underlying implementation might have
+relevant information to enhance e.g. the performance of an operation,
+but where the API does not allow it to expose it so that upper layers
+can take advantage of it.
+
+Another type of problems occur when working with dynamic RDF that may
+mix terminological and assertional information with object-oriented
+programming. LITEQ \cite{leinberger2014semantic} has a modern approach
+to a static case, where the developer gets substantial support from
+their Integrated Development Environment when programming, and where
+static typing is seen as a virtue that helps quality assurance of the
+code.
+
+However, a more interesting case is where an application (in a broad
+sense) can adapt to the data it is seeing. For example, say that there
+is an implementation of a Boat and a Car. It may not have been
+anticipated when the application was first developed, but the
+application sees an AmphibiousVehicle. Since it already knows how to
+propel a Car and Boat, it should be able to adapt to the situation
+given the context of whether the vehicle is on the road or in water.
+
+\begin{problem}
+Adapt programming languages to deal with dynamic RDF data containing a
+mix of terminological and assertional information.
+\end{problem}

### 2015-07-19 dissertation/problems.tex
reformulate a bit
some evaluation problems

@@
 \subsection{Problems with evaluations}\label{sec:evalproblems}
+
+\begin{problem}
+Evaluation methodologies are not founded on sound statistics.
+\end{problem}
+
+\begin{problem}
+Few laws of the Semantic Web information space can reasonably be
+expected to have validity far beyond the present.
+\end{problem}
+
+\begin{problem}
+Evaluation methodologies has no foundation in philosophy of science.
+\end{problem}
 
@@
 \begin{problem}
-Adapt programming languages to deal with dynamic RDF data containing a
-mix of terminological and assertional information.
+Current programming paradigm deals poorly with dynamic RDF data
+containing a mix of terminological and assertional information.
 \end{problem}

### 2015-07-20 dissertation/introdefhistory.tex
reference semantic problems

@@
 is an example of the social, economic and cultural mechanisms that
-shape the future of the Semantic Web. Likewise, it may not be the W3C
-Semantic Web stack at all that achieves success, it may be
-microformats\footnote{see \url{http://microformats.org/}},
-microdata\cite{microdata}, or something else entirely. For a
-thoughtful criticism of Linked Data with a proposed alternative, see
-\cite{darobin1}, but again, it is likely that social, economic and
-cultural factors will be as decisive as technical. One may say that
-e.g. microformats cannot help achieve the visions that may be stated
-for the Semantic Web, but that again emphasises visions, not
-operational definitions.
+shape the future of the Semantic Web, see
+Section~\ref{sec:semproblems} for a more elaborate
+discussion. Likewise, it may not be the W3C Semantic Web stack at all
+that achieves success, it may be microformats\footnote{see
+  \url{http://microformats.org/}}, microdata\cite{microdata}, or
+something else entirely. For a thoughtful criticism of Linked Data
+with a proposed alternative, see \cite{darobin1}, but again, it is
+likely that social, economic and cultural factors will be as decisive
+as technical. One may say that e.g. microformats cannot help achieve
+the visions that may be stated for the Semantic Web, but that again
+emphasises visions, not operational definitions.
 

### 2015-07-20 dissertation/problems.tex
constrain the scope of problems
Evaluation problems semantics problems

@@
 \section{Problems}\label{sec:problems}
+
+We seek to formulate concrete problems that should be solved to
+realise the vision formulated in Section~\ref{sec:motivation}. It is
+important to note that while some of these issues may be problematic
+for the whole Semantic Web community, their only intended scope is the
+presented vision.
 
@@
 
+In \cite{berners2000weaving}, Tim Berners-Lee take schema
+heterogeneity for granted and that it would be resolved by using
+inference, saying:
+
+\begin{quote}
+When, eventually, thousands of forms are linked together through the
+field for ``family name'' or ``last name'' or ``surname,'' then anyone
+analyzing the Web would realize that there is an important common
+concept here.
+\end{quote}
+
+He further goes on to explain how linking such terms will help make
+that inference, and how partial understanding is common in everyday
+life and will also be sufficient in my cases on the Web. The practice
+of using inference to find a common understanding is now known as
+ontology matching or ontology alignment, and is a very active area of
+research.
+
+% TODO, take some open problems in ontology alignment
+
+\begin{problem}
+How can suitable inference engines be made available to all users?
+\end{problem}
+
+However, we shall note that there are two directions that do not
+require inference to the same extent:
+
+Schema.org is sponsored by Google, Microsoft, Yahoo and Yandex using a
+community process to create and maintain a single, coordinated
+ontology. With their market shares, they have a reasonable chance of
+getting wide-spread adoption. However, we see also the key problem:
+
+\begin{problem}
+With a centralised ontology approach, how are anybody enabled to say
+anything?
+\end{problem}
+
+The next possibility is that authors converge towards using a single
+vocabulary per domain, but with no central control. The Linked Open
+Vocabularies is a useful resource for work in this direction, see
+\cite{lov2}. This may occur for economical reasons: If this happens to
+be the least expensive way to achieve useful data integration, authors
+may realise that this is a reasonable action to take. While this does
+not eliminate the need for inference, it may greatly reduce it.
+
+\begin{problem}
+How can the adoption of existing vocabularies be encouraged?
+\end{problem}
+
 \subsection{Problems with evaluations}\label{sec:evalproblems}
+
+To understand whether a result from a study on e.g. performance is
+valid, an empirical study must be conducted, as a SPARQL system is
+generally too complex to be evaluated solely by formal
+methods. Conventionally, one or more benchmarks have been used for
+this purpose, and there exists a number such
+benchmarks. Unfortunately, due to this complexity, they have to test a
+very large number of parameters, but are unable to do that, and they
+may also oversimplify the test by attempting to neutralise certain
+optimisation techniques. Further, they have no structured approach to
+investigate flaws in the benchmark itself, and have no meaningful
+summary statistics that can be used to understand the overall
+performance. While one may standardise benchmarks, doing so makes it
+impossible to test assertions that are outside of that standard.
 
@@
 \end{problem}
+
+It is clear that neither the vision presented in
+Section~\ref{sec:motivation} nor the more advanced vision presented in
+\cite{berners2001semantic} exists as of today. In both cases, Semantic
+Web would have to be deployed at a very different scale than what it
+is today. If the Semantic Web is sufficiently successful to realise
+these visions, it is also likely that it will be very different from
+today in terms of data profiles, workloads, etc. Since data profiles
+and workloads are important to device a test to evaluate assertions
+about performance, the fundamental problem arises that we currently
+test against something that does not exist, and this differs markedly
+from the methodology of natural science.
 
@@
 \begin{problem}
-Evaluation methodologies has no foundation in philosophy of science.
+Evaluation methodologies has little foundation in philosophy of science.
 \end{problem}

### 2015-07-20 dissertation/problems.tex
Laws of universe

@@
 
+Frank van Harmelen has encouraged a direction that seeks to study laws
+of the Semantic Web information universe\footnote{see
+  e.g. \url{http://videolectures.net/iswc2011_van_harmelen_universal/}
+  and
+  \url{http://sssw.org/2015/invited-speakers-tutors/frank-vanharmelen/}}.
+If this direction is pursued with success, then it may be possible to
+establish laws that can be used to generate data and workloads that
+can support realistic experiments even though a future Semantic Web
+may be several orders of magnitude larger than today's.
+
 \begin{problem}
-Few laws of the Semantic Web information space can reasonably be
+Establish laws of the Semantic Web information universe can reasonably be
 expected to have validity far beyond the present.

### 2015-07-20 dissertation/problems.tex
smaller things
evaluation problems small changes

@@
 How can the adoption of existing vocabularies be encouraged? 
+% TODO: Find the ref by Frank on the economics
 \end{problem}
@@
 
+Even if laws have been found that can make extrapolations across
+orders of magnitude tenable, evaluations still pose significant
+epistemological problems. For example, falsificationism, as argued by
+Karl Popper and adopted in \cite{avinatguide} will be unsuited, for
+reasons that more recent philosophers have pointed out, see
+\cite{chalmers1999whatis} for a comprehensive discussion: If it is
+insisted that hypotheses must be falsifiable, how can a researcher
+know if their hypothesis should be rejected, or if it is something
+wrong with the law, or perhaps the experiment itself?
+
 \begin{problem}
-Evaluation methodologies has little foundation in philosophy of science.
+Evaluation methodologies have little foundation in philosophy of science.
 \end{problem}
@@
 sense) can adapt to the data it is seeing. For example, say that there
-is an implementation of a Boat and a Car. It may not have been
-anticipated when the application was first developed, but the
-application sees an AmphibiousVehicle. Since it already knows how to
-propel a Car and Boat, it should be able to adapt to the situation
-given the context of whether the vehicle is on the road or in water.
+is an implementation of a \rdfterm{Boat} and a \rdfterm{Car}. It may
+not have been anticipated when the application was first developed,
+but the application sees an \rdfterm{AmphibiousVehicle}. Since it
+already knows how to propel a \rdfterm{Car} and \rdfterm{Boat}, it
+should be able to adapt to the situation given the context of whether
+the vehicle is on the road or in water.
 

### 2015-07-21 dissertation/problems.tex
Some text around stability

@@
 
+It was found in \cite{perez2009semantics} that
+
 \begin{problem}
-The complexity of SPARQL is high.
+SPARQL has some highly complex parts.
+\end{problem}
+
+In particular, nested \sparql{OPTIONAL}s will be difficult to
+evaluate. A Cartesian join is also likely to cause trouble, but even
+seemingly straightforward queries may be heavy. We also note from
+experience that:
+
+\begin{problem}
+Public endpoints may have unpredictable usage patterns.
+\end{problem}
+
+Caching plays an important role in the Web architecture
+\cite{webarch}, because previous responses may be reused to satisfy a
+current request. However, there are many problems that need to be
+solved to use it for SPARQL query evaluation:
+
+\begin{problem}
+SPARQL endpoints, which are based on HTTP, will only enable caching
+syntactically identical queries by  default.
+\end{problem}
+
+\begin{problem}
+Determining that two queries will return the exact same answer is a
+complex problem.
+\end{problem}
+
+\begin{problem}
+It is not known how much of the Semantic Web exposes information that
+could be used in HTTP caching.
 \end{problem}

### 2015-07-21 dissertation/problems.tex
Zipf problems

@@
 
+Moreover, an ambition of the Semantic Web is to deal well with
+heterogeneous data, and it has been known for long \cite{1385469} that:
+
+\begin{problem}
+The distribution of data is often dominated by a power-law.
+\end{problem}
+However, 
+\begin{problem}
+The data distribution is often unknown to the query engine, and so,
+following the principle of indifference, assumes uniform
+distributions.
+\end{problem}
+
 Caching plays an important role in the Web architecture
-\cite{webarch}, because previous responses may be reused to satisfy a
-current request. However, there are many problems that need to be
-solved to use it for SPARQL query evaluation:
+\cite{Jacobs:04:AWW}, because previous responses may be reused to
+satisfy a current request. However, there are many problems that need
+to be solved to use it for SPARQL query evaluation:
 

### 2015-07-21 dissertation/papersincontext.tex
Problems addressed

\section{The papers in context}

\subsection{Read-write hypermedia}\label{sec:conlapis}

\subsection{Design of Experiments}\label{sec:condoe}

including the criticism forwarded by
\cite{Duan:2011:AOC:1989323.1989340} and \cite{MontoyaVCRA12},

\subsection{Pushing complexity down the stack}\label{sec:conpush}

\subsection{Survey of HTTP caching}\label{sec:consanity}

### 2015-07-21 dissertation/problems.tex
Problems addressed

@@
 
-\begin{problem}
+\begin{problem}\label{prob:sparqlcomplex}
 SPARQL has some highly complex parts.
@@
 
-\begin{problem}
+\begin{problem}\label{prob:endpointunpred}
 Public endpoints may have unpredictable usage patterns.
@@
 
-\begin{problem}
+\begin{problem}\label{prob:zipf}
 The distribution of data is often dominated by a power-law.
@@
 However, 
-\begin{problem}
+\begin{problem}\label{prob:uniform}
 The data distribution is often unknown to the query engine, and so,
@@
 
-\begin{problem}
+\begin{problem}\label{prob:syntacticcache}
 SPARQL endpoints, which are based on HTTP, will only enable caching
@@
 
-\begin{problem}
+\begin{problem}\label{prob:queryeq}
 Determining that two queries will return the exact same answer is a
@@
 
-\begin{problem}
+\begin{problem}\label{prob:sanity}
 It is not known how much of the Semantic Web exposes information that
@@
 
-\begin{problem}
+\begin{problem}\label{prob:infgivecode}
 How can suitable inference engines be made available to all users?
@@
 
-\begin{problem}
+\begin{problem}\label{prob:sayanything}
 With a centralised ontology approach, how are anybody enabled to say
@@
 
-\begin{problem}
+\begin{problem}\label{prob:useexist}
 How can the adoption of existing vocabularies be encouraged? 
@@
 
-\begin{problem}
+\begin{problem}\label{prob:badstats}
 Evaluation methodologies are not founded on sound statistics.
@@
 
-\begin{problem}
+\begin{problem}\label{prob:lawsfuture}
 Establish laws of the Semantic Web information universe can reasonably be
@@
 
-\begin{problem}
+\begin{problem}\label{prob:badphil}
 Evaluation methodologies have little foundation in philosophy of science.
@@
 
-\begin{problem}
+\begin{problem}\label{prob:breakdown}
 By breaking down the query down to individual triple patterns, the
@@
 
-\begin{problem}
+\begin{problem}\label{prob:complexapi}
 If methods such as \pcode{get\_pattern} are added for every part of the
@@
 
-\begin{problem}
+\begin{problem}\label{prob:dontjustpass}
 By merely passing the entire query representation to an
@@
 
-\begin{problem}
+\begin{problem}\label{prob:dynaprog}
 Current programming paradigm deals poorly with dynamic RDF data
@@
 \end{problem}
+
+\subsection{Problems addressed by this work}\label{sec:probaddress}
+
+Given the technical experience outlined in Section~\ref{sec:history},
+I first addressed Problem~\ref{prob:lapis}, by suggesting a simple
+vocabulary for the server to tell the client what primitive operations
+it is allowed to perform on a given resource. See
+Section~\ref{sec:conlapis} for more on this paper.
+
+My next concern was the implications of the
+problems~\ref{prob:zipf}~and~\ref{prob:uniform} in a federated SPARQL
+engine. While studying the evaluations done in the relevant literature
+in the field, I decided to start an effort to address
+Problem~\ref{prob:badstats} first by adopting methodology from the
+statistical field of ``Design of Experiments'' (DoE). See
+Section~\ref{sec:condoe} for details on this work.
+
+Then, as \cite{buil2013sparql} confirmed my suspicions that there were
+severe stability problems with SPARQL endpoints,
+Problem~\ref{prob:syntacticcache} became my main focus, but clearly,
+the understanding of the actual usage of caching was inadequate, so I
+decided to solve Problem~\ref{prob:sanity} first. This work is
+discussed in Section~\ref{sec:consanity}.
+
+To further address Problem~\ref{prob:syntacticcache}, I decided to
+write a query planner that could reuse previous results or prefetch
+data the system deemed relevant, but then we realised that the current
+programming paradigm was quite unhelpful when customising a query
+planner for such a purpose. Rather than writing a prototypical query
+engine for illustrative purposes, we wanted to integrate it into a
+fully compliant SPARQL query engine, and address
+problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}. This
+is explained in Section~\ref{sec:conpush}

### 2015-07-21 dissertation/problems.tex
reorganize

@@
 severe stability problems with SPARQL endpoints,
-Problem~\ref{prob:syntacticcache} became my main focus, but clearly,
-the understanding of the actual usage of caching was inadequate, so I
-decided to solve Problem~\ref{prob:sanity} first. This work is
-discussed in Section~\ref{sec:consanity}.
-
-To further address Problem~\ref{prob:syntacticcache}, I decided to
-write a query planner that could reuse previous results or prefetch
+Problem~\ref{prob:syntacticcache} became my main focus, and I decided
+to write a query planner that could reuse previous results or prefetch
 data the system deemed relevant, but then we realised that the current
@@
 is explained in Section~\ref{sec:conpush}
+
+Before addressing Problem~\ref{prob:syntacticcache}, I found that the
+understanding of the actual usage of caching was inadequate, so I
+decided to solve Problem~\ref{prob:sanity} first. This work is
+discussed in Section~\ref{sec:consanity}.

### 2015-07-21 dissertation/problems.tex
Excuse bad philosophy

@@
 discussed in Section~\ref{sec:consanity}.
+
+Problem~\ref{prob:badphil} has been a growing concern throughout this
+work. A proper evaluation should be an important part of the
+dissertation, and yet, the problems discussed in
+Section~\ref{sec:evalproblems} are so severe that the time frame of
+this work does not allow a satisfactory solution to these
+problems. The paper discussed in Section~\ref{sec:conphil} serves to
+excuse this.

### 2015-07-21 dissertation/problems.tex
other problems

@@
 excuse this.
+
+TODO: Paper yet to be done on caching.
+
+\subsection{Future work; Discussion of other problems}
+
+As noted in Section~\ref{sec:motivation} it is clear that evaluating
+arbitrary SPARQL queries are in general unsustainable, and it is
+likely that the problems in Section~\ref{sec:perfproblems} is partly
+to blame for it. Our work, and most of the related work in
+Section~\ref{sec:related} approach this work from the direction of
+moving stress from the server, but it is also clear that another way
+to address Problem~\ref{prob:sparqlcomplex} is for the SPARQL engine
+to identify queries it would be too heavy to answer already when
+parsing or planning, and also expose that in the service description
+so that clients can avoid asking them. The HTTP/2 specification
+\cite{rfc7540} has some generic facilities that may be helpful:
+
+\begin{quote}
+\texttt{ENHANCE\_YOUR\_CALM (0xb)}:  The endpoint detected that its peer is
+      exhibiting a behavior that might be generating excessive load.
+\end{quote}
+
+% TODO: Discuss characteristic sets?
+
+Problem~\ref{prob:queryeq} is best addressed by creating hashing
+algorithms that have a reasonable chance of returning the same digest
+if two queries are equivalent. Some works have created cryptographical
+hashing functions, but they may be overly costly to compute for this
+purpose. % TODO ref.

### 2015-07-21 dissertation/problems.tex
semantic problems laws

@@
 purpose. % TODO ref.
+
+I have not approached any of the problems of
+Section~\ref{sec:semproblems}, mainly because I think it is an open
+question which of these approaches will be adopted in practice, and
+whatever the solution may be, their designers should make sure they
+are orthogonal to the other problems discussed here. Thus, it was
+natural to define them out of scope. 
+% TODO: However, I shall assume that no inference is needed in the
+% following examples.
+
+Clearly, Problem~\ref{prob:lawsfuture} is a very interesting one, with
+important implications for evaluation, but also because it shifts the
+focus of our attention. In this work, the methodology used to study
+Problem~\ref{prob:sanity} could be used to gain such understanding,
+but at present, the field is too immature to attempt the formulation
+of any laws. As noted in Section~\ref{sec:history}, this study is
+partly motivated by my subjective experience of why the Web was so
+successful. Clearly, this should be studied more objectively, but I am
+not aware of any such study. Such study could also be applied to see
+if the experiences from the emergence of the Web can be applied to
+accelerating the Semantic Web, like I subjectively propose.

### 2015-07-21 dissertation/problems.tex
hypermedia

@@
 accelerating the Semantic Web, like I subjectively propose.
+
+Problem~\ref{prob:tpf} was along with
+problems~\ref{prob:sparqlcomplex},~\ref{prob:endpointunpred}~and~\ref{prob:syntacticcache}
+part of the motivation of \cite{ldf1} and \cite{verborgh2014querying},
+and this builds in part on the insight I contributed when addressing
+Problem~\ref{prob:lapis}. Their solution was to define an ontology to
+describe an interface that can answer a single triple pattern, but
+since it is hypermedia, it can be described in the messages of the
+protocol.

### 2015-07-21 dissertation/problems.tex
The MOP for dynamic programming

@@
 protocol.
+
+I believe that the core of Problem~\ref{prob:dynaprog} lies in the
+prevalence of statically typed languages. From an engineering
+perspective, static typing seems like a good idea, since errors should
+surface at compile time. However, it seems like it comes at a high
+cost: Programmers seem incapable of thinking beyond the constraints it
+sets. Nevertheless, the theoretical framework that seems to solve the
+problem was proposed in 1991 in the book ``The art of the metaobject
+protocol'' \cite{kiczales1991art}. Perl has a mature implementation of
+the metaobject protocol, and the topic has been discussed on several
+hackathons, and some code has been written, but the solution seems
+elusive.

### 2015-07-21 dissertation/papersincontext.tex
Excuse bad philosophy
context text

@@
 \section{The papers in context}
+
+This section will put the included papers into context. It also enjoys
+the benefit of hindsight, and will therefore address some of the
+shortcomings that have surfaced since their publication.
 
@@
 
+This work is a paper entitled ``The necessity of hypermedia RDF and an
+approach to achieve it'' that was presented in the First Linked APIs
+workshop at the Ninth Extended Semantic Web Conference. I was the sole
+author. The reference is \cite{kjernsmo_lapis_2012}.
+
+%TODO: not the Dagstuhl?
+
+This paper is relevant to the topic of the dissertation for two
+reasons: First, it shows how hypermedia may be used to drive a
+read-write application, not just read-only, which is the focus of most
+of the discussion. Second, it provides some of the basis for
+\cite{ldf1}, which is important in the query planner.
+
+The main idea in this paper is to look upon RDF as a natural language
+sentence with subject, predicate and object, and that sentence should
+then be sufficiently human-readable to be straightforward to
+understand when a developer does ``View Source''.
+
+Its main weakness is how it deals with authentication: The basic idea
+is that when a client visits a resource, it will be told what
+write operations it can perform. However, write operations will rarely
+be permitted unless the client has authenticated and been authorised
+to do so, and so, it assumes that credentials will be supplied with
+the first read request. While the HTTP standard permits clients to
+supply credentials with such requests, in practice, it rarely if ever
+happens, possibly for privacy reasons. Thus, the server will have to
+challenge the client first, but this should not happen on a resource
+that the client is permitted to read. Thus, the proposed hypermedia
+needs a property so that the client understands what to do if it wants
+to do write operations.
+
 \subsection{Design of Experiments}\label{sec:condoe}
 
-including the criticism forwarded by
-\cite{Duan:2011:AOC:1989323.1989340} and \cite{MontoyaVCRA12},
+This work resulted in a paper that was nominated for Best Paper on the
+Evaluation track on the International Semantic Web Conference 2013,
+entitled ``Introducing Statistical Design of Experiments to SPARQL
+Endpoint Evaluation'', which was co-authored with John
+S. Tyssedal. The reference is \cite{kjernsmo_doe_intro}.
+
+To understand the context of this study, I noted none of the
+benchmarks that was in use for the evaluation of SPARQL engines could
+be used for basic statistical hypothesis testing, and clearly,
+hypothesis testing must be a fundamental requirement. Then, I noted
+substantial criticism forwarded by
+\cite{Duan:2011:AOC:1989323.1989340} and \cite{MontoyaVCRA12}.
+
+However, this study merely scratches the surface. To establish a
+Design of Experiment-based methodology would be a multi-year project
+in its own right, clearly beyond the scope of this
+dissertation. Unfortunately, this also implies that I do not have the
+rigorous evaluation that I would preferred to have in this
+dissertation, it is unattainable.
 
@@
 
+This work resulted in a workshop paper entitled ``Pushing complexity
+down the stack'', presented in the Developers Workshop at the
+International Semantic Web Conference, by the lead author Gregory Todd
+Williams, we were the two authors. The reference is
+\cite{williamspushing}.
+
+% TODO: detailed contribution?
+
+This work was the outcome of several hackathons that I had organised
+in the Perl community. It is not very well known that Gregory Todd
+Williams' SPARQL implementation written in Perl is one of the SPARQL
+1.1 reference implementations, and is known to be fully compliant with
+the specification. Through our work, and the availability of
+traits-based programming in Perl, see \cite{traits}, we had extensive
+discussions on how to leave optimisations to lower levels, e.g. the
+underlying database engine, when they had the potential to provide a
+better plan or cost estimate. While it is certainly possible to do
+this without this new programming paradigm, traits made the
+methodology much more comprehensible, and so helped the subsequent
+work substantially.
+
+This work also resulted in the ``Attean''
+framework\footnote{\url{https://metacpan.org/release/Attean}}, also
+mainly authored by Gregory Todd Williams with some smaller
+contributions from me. It is also available in the Linux distribution
+Debian (and derivatives, such as Ubuntu) as \texttt{libattean-perl}.
+
 \subsection{Survey of HTTP caching}\label{sec:consanity}
+
+\subsection{Philosophy of Science}\label{sec:conphil}

### 2015-07-22 dissertation/papersincontext.tex
survey

@@
 
+I committed to do the survey after Jürgen Umbrich challenged me to see
+if the standards were actually followed. Rather than just survey the
+usage for SPARQL Endpoints, I decided to do it for the breadth of the
+Semantic Web. This resulted in a research track paper on the Extended
+Semantic Web Conference 2015 entitled ``A Survey of HTTP Caching
+Implementations on the Open Semantic Web'' with reference
+\cite{kjernsmo_survey_2015} and a technical report
+\cite{kjernsmo_add_survey_2015} that elaborates on data reduction
+methods and statistical methods.
+
+Although the support for caching headers are not yet widespread, it is
+sufficient to perform experiments on real-world data.
+
 \subsection{Philosophy of Science}\label{sec:conphil}

### 2015-07-22 dissertation/papersincontext.tex
philosophy

@@
 
-Although the support for caching headers are not yet widespread, it is
+Although the support for caching headers is not yet widespread, it is
 sufficient to perform experiments on real-world data.
@@
 \subsection{Philosophy of Science}\label{sec:conphil}
+
+Dissatisfaction with the methodology used in the literature,
+specifically benchmarking, to test assertions of performance lead to
+exploration of statistical methods in referenced in
+Section~\ref{sec:condoe}, and to organise the Empirical 2014 Workshop
+at the Extended Semantic Web Conference (which I was unfortunately
+prevented from attending myself). Still, frustrated by the slow
+progress of the last 16 years, and difficulty with comparing methods
+to the more mature natural sciences, I still felt the need to express
+a philosophical view of science, which I did by submitting an article
+to the ``Negative or inconclusive results in Semantic Web'' at the
+Extended Semantic Web Conference 2015. The essay, see
+\cite{kjernsmo_noise_2015} was tangential to the Call for Papers, but
+the chairs decided that Jacco van Ossenbruggen would interview me in
+front of the audience about the contents of this essay.
+
+Again, the essay serves somewhat as an excuse for why the evaluations
+in this dissertation does not meet the standards that I would
+advocate, it would require many years of concerted effort to meet
+those standards. The intention is then to promote a discussion of
+epistemological questions, and as such, the essay is intentionally
+provocative.  In retrospect, the greatest weakness of the essay is
+that I failed to recognise the impact of Frank van Harmelen's call for
+studies that can establish laws of the Semantic Web information
+universe, as discussed in Section~\ref{sec:evalproblems}. To my
+defence, I was only made aware of this keynote by the
+reviewer.

### 2015-07-22 dissertation/related.tex
Discuss the SIGMOD paper on caching

@@
 
-Montoya et al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query
-federation system that takes advantage of client-side data
-replication.
+Recently, Papailiou et al. \cite{papailiou2015graph} made an extensive
+study of SPARQL caching, that is similar in scope to the present work,
+except for our emphasis on deployment on proxies in the Internet and
+integration with HTTP-based caching.  Montoya et
+al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query federation
+system that takes advantage of client-side data replication. Both
+these studies claim to support the entire SPARQL query language, but
+in both cases, it is not clear that they actually do that.
 

### 2015-07-22 dissertation/related.tex
reformulate sigmod relation

--- 2015-07-19 dissertation/related.tex
@@
 
-Montoya et al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query
-federation system that takes advantage of client-side data
-replication.
+Recently, Papailiou et al. \cite{papailiou2015graph} made an extensive
+study of SPARQL caching, addressing some of the same problems as the
+present work, except for our emphasis on deployment on proxies in the
+Internet and integration with HTTP-based caching.  Montoya et
+al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query federation
+system that takes advantage of client-side data replication. Both
+these studies claim to support the entire SPARQL query language, but
+in both cases, it is not clear that they actually do that.
 

### 2015-12-18 dissertation/introdefhistory.tex
Small fixes

@@
 However, at this point, I started to realise that it was difficult to
-develop applications around traversing the Linked Data, and that
+develop applications around traversing Linked Data, and that
 prompted me to consider the possibility to offer the ability to
@@
 the update language in SPARQL 1.1 but differed substantially in
-surface syntax, partly because our work showed how certain patterns
+surface syntax, partly because our work showed how certain usage patterns
 could be simplified.

### 2015-12-18 dissertation/related.tex
Small fixes

@@
 pattern, an operation that may be quite expensive for the server, and
+%TODO: ref characteristic sets?
 indeed, this information may be so expensive to provide that SPARQL

### 2016-04-28 dissertation/introdefhistory.tex
small things

@@
 methods for evaluating performance, standards compliance and even
-philosophy of science. With a scope of this width, the investigation
+philosophy of science. With a scope of this breadth, the investigation
 is shallow, it is hoped that some key insights will provoke further
@@
 
-\section{Semantic Web -- A definition?}\label{sec:noref}
+\section{Semantic Web -- A definition?}\label{sec:nodef}
 
@@
 discussion. Likewise, it may not be the W3C Semantic Web stack at all
-that achieves success, it may be microformats\footnote{see
+that achieves success, though unlikely at this point, it may be
+something in the extension of microformats\footnote{see
   \url{http://microformats.org/}}, microdata\cite{microdata}, or

### 2016-04-28 dissertation/introdefhistory.tex
breaking of conventions

@@
 \section{A personal history}\label{sec:history}
+
+The following challenges several conventions of scientific
+writing. Passive voiced is generally overused in scientific writing,
+sometimes because the author may think it emphasises the objectivity
+of the investigation. In my opinion it rather attempts to hide the
+fact that science is a human and social endeavour, and only by
+acknowledging that fact, an investigator can properly account for
+their own role. Double blind trials in medicine are an example of why
+it is important to acknowledge the investigators role. Moreover, the
+use of ``we'', when it is clear that the work has been done by a
+single investigator is inappropriate as it is attempting to obscure
+the role of the investigator.
+
+As noted, the Semantic Web is a complex technical, social, economic
+and cultural artefact. The objectivity of an investigator may be
+challenged by a large intellectual investment in this artefact, the
+more you want this artefact to succeed, the higher your risk of not
+being sufficiently objective is. I hereby proclaim that I have a very
+large intellectual investment in the success of the Semantic Web, and
+therefore this section is a reflection on my own role in the
+development of this artefact.
 

### 2016-04-29 dissertation/introdefhistory.tex
small corrections

@@
 use of ``we'', when it is clear that the work has been done by a
-single investigator is inappropriate as it is attempting to obscure
+single investigator is inappropriate as it obscures
 the role of the investigator.
@@
 challenged by a large intellectual investment in this artefact, the
-more you want this artefact to succeed, the higher your risk of not
-being sufficiently objective is. I hereby proclaim that I have a very
+more you want this artefact to succeed, the higher may your risk of not
+being sufficiently objective be. I hereby declare that I have a very
 large intellectual investment in the success of the Semantic Web, and
@@
 answering over data on the open Web, for generic application
-development. 16 years of experience thus motivates the research done
-in this thesis.
+development.
+
+While the motivation and the problem statements of this work
+undoubtedly have been influenced by my intellectual investment in the
+Semantic Web, the papers will show that the evaluation has not.

### 2016-04-29 dissertation/papersincontext.tex
contributions of two first papers

@@
 understand when a developer does ``View Source''.
+
+Main contributions:
+\begin{enumerate}
+\item A thorough discussion of the implications of Hypermedia
+  Types\cite{hypermediatypes} for RDF.
+\item A sketch of a vocabulary for read-write RDF hypermedia.
+\end{enumerate}
 
@@
 dissertation, it is unattainable.
+
+Main contributions:
+\begin{enumerate}
+\item Introduction of a path to critical practice of evaluations, that makes
+  use of contemporary statistical techniques, to establish a practice
+  that can be used to refute assertions on performance.
+\item A didactical experiment to help researchers understand the statistics.
+\item The novel application of a well established method in
+  statistics, rarely used in Computer Science, to SPARQL endpoint
+  evaluation.
+
+\end{enumerate}
 

### 2016-05-02 dissertation/papersincontext.tex
pushing contributions
reference for the interview contributions in philosophy

@@
 
-% TODO: detailed contribution?
+Main contributions:
+\begin{enumerate}
+\item A framework to enable the use of low-level optimisations in
+  databases.
+\item Simplification when implementing experimental features in
+  SPARQL.
+\item Make it possible to compose custom query planners that doesn't
+  require inheritance.
+\end{enumerate}
 
@@
 
+The most important contribution of this paper was an understanding the
+actual usage of caching headers, but it was also intended to raise the
+awareness of these important Web standards in the Semantic Web community.
+
 Although the support for caching headers is not yet widespread, it is
@@
 \cite{kjernsmo_noise_2015} was tangential to the Call for Papers, but
-the chairs decided that Jacco van Ossenbruggen would interview me in
-front of the audience about the contents of this essay.
+the chairs decided to accept it as an interview. With Ruben Verborgh
+in the role of the interviewer, we reformulated it as
+\cite{kjernsmo_noise_2015_interview}. It was performed with  Jacco van
+Ossenbruggen as the interviewer in front of the audience.
 
-Again, the essay serves somewhat as an excuse for why the evaluations
-in this dissertation does not meet the standards that I would
-advocate, it would require many years of concerted effort to meet
-those standards. The intention is then to promote a discussion of
-epistemological questions, and as such, the essay is intentionally
-provocative.  In retrospect, the greatest weakness of the essay is
-that I failed to recognise the impact of Frank van Harmelen's call for
-studies that can establish laws of the Semantic Web information
-universe, as discussed in Section~\ref{sec:evalproblems}. To my
-defence, I was only made aware of this keynote by the
-reviewer.
+As a philosophical essay, the contributions are hard to detail. It is
+written in a provocative style, to encourage discussions of
+epistemological questions. It also points out shortcomings in current
+practice, and it is clear that a proper evaluation of the work
+outlined in Chapter~\ref{sec:tpfcacheplanning} would be a major
+undertaking.
+
+In retrospect, the greatest weakness of the essay is that I failed to
+recognise the impact of Frank van Harmelen's call for studies that can
+establish laws of the Semantic Web information universe, as discussed
+in Section~\ref{sec:evalproblems}. To my defence, I was only made
+aware of this keynote by the reviewer.

### 2016-05-03 dissertation/cache-proxy.tex
start of a chapter on cache

\documentclass[a4paper, 12pt, draft]{article}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
%\usepackage{times}
\usepackage{listings}
\usepackage{supertabular}
\usepackage{cite}
\usepackage{amsthm}

\include{notasjon}
\include{helpers}
\usepackage[pdftex,linktocpage]{hyperref}

\begin{document}
\title{Prefetching SPARQL query cacher}
\author{Kjetil Kjernsmo}
%\institute{Department of Informatics,
%Postboks 1080 Blindern,
%0316 Oslo, Norway}

%\email{kjekje@ifi.uio.no}

\maketitle

\section{Introduction}

This chapter describes an effort to create a proxy that can cache the
results of not only SPARQL queries, but also the results of individual
triple patterns. It will asynchronously analyse executed queries, and
may prefetch the results of certain triple patterns into the cache.

The system is in the convergence of the directions described in this
dissertation: It should relieve the remote endpoint of some of the
burden to evaluated the query and it should add to the robustness of the
open Web SPARQL infrastructure. It uses hypermedia to answer individual
triple patterns, based on Triple Pattern Fragments  \cite{ldf1}. The
system can only answer triple patterns, but other than that, supports
SPARQL 1.1 in its entirety. This was made possible to develop quickly
by the efforts put into the query planning in the Attean framework
described in Section~\ref{sec:conpush}. Caching based on RFC7234
\cite{rfc7234} was shown to be of possible use in the survey we
conducted. Finally, it was planned to evaluate the approach using
Design of Experiments.

Unfortunately, the system has as of this writing insufficient
performance for an evaluation to be meaningful. Nevertheless, it
points out some interesting lessons with varying degrees of
certainty. This chapter will detail the system, show its design and
features and discuss its failures.

\bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}
\bibliographystyle{plain}

\end{document}

### 2016-05-03 dissertation/cache-proxy.tex
minor
start of a cache figure
improve caption

@@
 \usepackage{amsthm}
+\usepackage{tikz}
 
@@
 described in Section~\ref{sec:conpush}. Caching based on RFC7234
-\cite{rfc7234} was shown to be of possible use in the survey we
-conducted. Finally, it was planned to evaluate the approach using
+\cite{rfc7234} was shown to be of possible use in the survey I
+conducted. Finally, it was planned to be evaluated using
 Design of Experiments.
@@
 
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}
+\draw (0,0) --(4,0);
+\node at (5.5,0) { Database cache };
+\draw (0,0.5) --(4,0.5);
+\node at (5.5,0.5) { Reverse proxy };
+\draw (0,2) --(4,2);
+\node at (5.5,2) { Caching proxy };
+\draw (0,3.5) --(4,3.5);
+\node at (5.5,3.5) { Forward proxy };
+\draw (0,4) --(4,4);
+\node at (5.5,4) { Client cache };
+\end{tikzpicture}
+\caption{The possible positions of a cache. A cache may reside at both
+  the server side or client side, or any of the intermediate
+  proxies. The database proxy and reverse proxy is typically
+  controlled by the data provider.}
+\end{center}
+\end{figure}
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-03 dissertation/cache-proxy.tex
more on the arrows

@@
 \draw (0,0) --(4,0);
-\node at (5.5,0) { Database cache };
+\node at (5.5,0) { Database };
 \draw (0,0.5) --(4,0.5);
@@
 \node at (5.5,4) { Client cache };
+
+\draw [->] (0.2,4) --(0.2,0);
+\draw [<-] (0.4,4) --(0.4,0);
+\draw [->] (0.7,2) --(0.7,0);
+\draw [<-] (0.9,2) --(0.9,0);
+\draw [->] (1.2,4) --(1.2,2);
+\draw [dashed, ->] (1.2,2) --(1.2,0);
+\draw [dashed, <-] (1.4,2) --(1.4,0);
+\draw [->] (1.5,2) --(1.5,4);
+\draw [->] (1.7,2) --(1.7,0);
+\draw [<-] (1.9,2) --(1.9,0);
+\draw [->] (2.2,4) --(2.2,2);
+\draw [->] (2.4,2) --(2.4,4);
 \end{tikzpicture}
 \caption{The possible positions of a cache. A cache may reside at both
-  the server side or client side, or any of the intermediate
-  proxies. The database proxy and reverse proxy is typically
-  controlled by the data provider.}
+  the server side (as a database cache) or client side, or any of the
+  intermediate proxies. The database cache and reverse proxy is
+  typically controlled by the data provider. The arrows illustrate
+  requests (when pointing down) and responses (when pointing up) in a
+  deployment scenario for when a the query cache considered in this
+  paper is deployed on a caching proxy in the Internet infrastructure,
+  see the text for details.}
 \end{center}

### 2016-05-03 dissertation/cache-proxy.tex
more on the proxies

@@
 
+Figure~\ref{fig:messaging} illustrates where caches may be in the
+Internet infrastructure, as well as HTTP requests and responses in a
+typical deployment scenario of the system.
+
 \begin{figure}
@@
   the server side (as a database cache) or client side, or any of the
-  intermediate proxies. The database cache and reverse proxy is
-  typically controlled by the data provider. The arrows illustrate
-  requests (when pointing down) and responses (when pointing up) in a
-  deployment scenario for when a the query cache considered in this
-  paper is deployed on a caching proxy in the Internet infrastructure,
-  see the text for details.}
+  intermediate proxies. The arrows illustrate requests (when pointing
+  down) and responses (when pointing up) in a deployment scenario for
+  when a the query cache considered in this paper is deployed on a
+  caching proxy in the Internet infrastructure, see the text for
+  details.}\label{fig:messaging}
 \end{center}
 \end{figure}
+
+The system is based on HTTP, which implies a client-server
+architecture. In this architecture, a cache may be present at
+conceptually five different levels. First, the client may have its own
+cache, which caches only responses made by the client itself. The next
+level is known as a forward proxy. They aren't currently very common,
+but have in the past often been institutional proxies, or proxies
+employed by Internet Service Providers for caching responses that may
+be common to many of their users. The next level, generically referred
+to as a caching proxy, may be in the Internet
+infrastructure. Currently, the most common form of this type of proxy
+is known as a Content Delivery Network (CDN). It is very common to
+have a reverse proxy near the server. They are often known as a ``Web
+Accelerator''. Usually, they will communicate with the server with
+HTTP, but it is conceivable that it may use a different protocol, and
+may employ detailed knowledge of the hosted data to optimise cache
+operations. Finally, the database, in this context a SPARQL Endpoint,
+may use conventional database caching techniques. The database cache
+and reverse proxy is typically controlled by the data provider.
+
+The present system may be deployed at any of these levels. However,
+since the database and the reverse proxy in front of it may have
+detailed knowledge of the data, e.g. a complete data profile with
+statistics to optimise join order, these levels have much in common
+with a conventional database cache, which is extensively discussed in
+the literature.
+
+My interest is the case where the caching proxy has no further
+knowledge of the data than what is exposed through the SPARQL
+Endpoint or hypermedia metadata. In practice, there is very
+little information.
 

### 2016-05-03 dissertation/cache-proxy.tex
number the messages more on the cache

@@
 \usepackage{tikz}
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
 
@@
 \draw (0,4) --(4,4);
-\node at (5.5,4) { Client cache };
+\node at (5.5,4) { Client };
 
@@
 \draw [<-] (0.4,4) --(0.4,0);
+\node at (0.3,4.5) {\circled{1}};
 \draw [->] (0.7,2) --(0.7,0);
 \draw [<-] (0.9,2) --(0.9,0);
+\node at (0.8,4.5) {\circled{2}};
 \draw [->] (1.2,4) --(1.2,2);
@@
 \draw [->] (1.5,2) --(1.5,4);
+\node at (1.4,4.5) {\circled{3}};
 \draw [->] (1.7,2) --(1.7,0);
 \draw [<-] (1.9,2) --(1.9,0);
+\node at (1.8,4.5) {\circled{4}};
 \draw [->] (2.2,4) --(2.2,2);
 \draw [->] (2.4,2) --(2.4,4);
+\node at (2.3,4.5) {\circled{5}};
 \end{tikzpicture}
@@
 My interest is the case where the caching proxy has no further
-knowledge of the data than what is exposed through the SPARQL
-Endpoint or hypermedia metadata. In practice, there is very
-little information.
+knowledge of the data than what is exposed through the SPARQL Endpoint
+or hypermedia metadata. In practice, there is very little
+information. If the cache is to be shared, then it would typically
+reside on the forward proxy, or in a CDN.
+
+Figure~\ref{fig:messaging} illustrates the case where HTTP messages
+are being passed, where the cache is assumed to reside on a
+intermediate caching proxy in the Internet, i.e. a CDN. The figure may
+be viewed as having time on the $x$-axis, with the arrows pointing
+down being requests and the arrows point up being responses. In this
+example, the client first makes a request (, which is passed through the
+proxy because it has not yet cached any responses. The response is
+then also transmitted directly from the SPARQL Endpoint at the
+database back to the client. The entire response may be cached at the
+proxy in case it can be reused in its entirety. However,
 

### 2016-05-03 dissertation/cache-proxy.tex
move the arrows a bit
description of the figure

@@
             \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
+\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
 
@@
 \node at (0.3,4.5) {\circled{1}};
-\draw [->] (0.7,2) --(0.7,0);
-\draw [<-] (0.9,2) --(0.9,0);
-\node at (0.8,4.5) {\circled{2}};
-\draw [->] (1.2,4) --(1.2,2);
-\draw [dashed, ->] (1.2,2) --(1.2,0);
-\draw [dashed, <-] (1.4,2) --(1.4,0);
-\draw [->] (1.5,2) --(1.5,4);
-\node at (1.4,4.5) {\circled{3}};
-\draw [->] (1.7,2) --(1.7,0);
-\draw [<-] (1.9,2) --(1.9,0);
-\node at (1.8,4.5) {\circled{4}};
-\draw [->] (2.2,4) --(2.2,2);
-\draw [->] (2.4,2) --(2.4,4);
-\node at (2.3,4.5) {\circled{5}};
+\draw [->] (0.9,2) --(0.9,0);
+\draw [<-] (1.1,2) --(1.1,0);
+\node at (1.0,4.5) {\circled{2}};
+\draw [->] (1.6,4) --(1.6,2);
+\draw [dashed, ->] (1.6,2) --(1.6,0);
+\draw [dashed, <-] (1.8,2) --(1.8,0);
+\draw [->] (1.9,2) --(1.9,4);
+\node at (1.7,4.5) {\circled{3}};
+\draw [->] (2.4,2) --(2.4,0);
+\draw [<-] (2.6,2) --(2.6,0);
+\node at (2.5,4.5) {\circled{4}};
+\draw [->] (3,4) --(3,2);
+\draw [->] (3.2,2) --(3.2,4);
+\node at (3.1,4.5) {\circled{5}};
 \end{tikzpicture}
@@
 down being requests and the arrows point up being responses. In this
-example, the client first makes a request (, which is passed through the
-proxy because it has not yet cached any responses. The response is
-then also transmitted directly from the SPARQL Endpoint at the
-database back to the client. The entire response may be cached at the
-proxy in case it can be reused in its entirety. However,
+example, the client first makes a request (request \circled{1}), which
+is passed through the proxy because it has not yet cached any
+responses. The response (response \circled{1}) is then also
+transmitted directly from the SPARQL Endpoint at the database back to
+the client. The entire response may be cached at the proxy in case it
+can be reused in its entirety. However, the main point is that the
+request will be analysed by the proxy in parallel to being sent to the
+server, and another request \circled{2} for a triple pattern fragment
+that the analyser thinks will enable the proxy to assist the server
+the best for future requests, is sent. This analysis is discussed in
+\todo{ref}. The response \circled{2} is not sent back to the client,
+but will enter the cache. Lets say that the analyser was successful
+and that the cache can be used to answer request \circled{3}. In that
+case, the proxy will send a rewritten query for the rest of the data
+it needs to answer the query to the origin server, indicated by the
+dashed arrows. This may be a SPARQL query, a TPF request or a
+combination thereof, depending on what the planner finds least
+expensive. The proxy will then evaluate the entire query and send the
+final result back as response \circled{3}. In addition, the analyser
+may have identified another Triple Pattern Fragment that it takes as
+likely useful for future requests, and sends a request and caches the
+response \circled{4}. At some point, it may be able to answer the
+entire query using only cached results, as illustrated by request and
+response \circled{5}.
 

### 2016-05-04 dissertation/cache-proxy.tex
relevant modules to the project

@@
-\documentclass[a4paper, 12pt, draft]{article}
+\documentclass[a4paper, 12pt]{article}
 \usepackage{graphicx}
@@
 \usepackage[pdftex,linktocpage]{hyperref}
+\hyperbaseurl{https://github.com/}
 
@@
 
+\section{Implementation}\label{sec:impl}
+
+The implementation consists of several modules that have been
+published to the Comprehensive Perl Archive Network as Free Software,
+and most of which have become a part the Free Software ecosystem. As
+such, it builds on the code of hundreds of authors, too long to
+enumerate. I shall constrain myself to list the modules that have code
+that has been motivated from this study. Modules that have other
+primary authors than myself have been started prior to this present
+project, but may have substantial contributions from this project. The
+Attean framework was started as part of the effort in
+Section~\ref{sec:conpush}, but even though it has been developed
+further with the requirements that arose from this project and has
+contributions from me, the vast majority of the code has been written
+my Gregory Todd Williams. The main work in this project has gone into
+the module AtteanX::Query::Cache, where I have authored the vast
+majority of the code. The relative difference in authorship can be
+further examined by checking the linked Github repositories, or by
+using git2prov\footnote{see \url{http://git2prov.org/}} to create RDF
+data of the commit history. Table~\ref{tab:modules} shows the modules
+that have been influenced or written as part of this project.
+
+\begin{table}
+\caption{The modules that enable the query cache}\label{tab:modules}
+\begin{tabular}{ | l | p{3cm} | l |}
+  \hline
+  Module & Authors & Github URL \\ \hline
+
+  AtteanX::Query::Cache & Kjetil Kjernsmo, Gregory Todd Williams &
+  \url{kjetilk/p5-atteanx-query-cache} \\
+
+  AtteanX::Store::SPARQL & Kjetil Kjernsmo &
+  \url{kjetilk/p5-atteanx-store-sparql} \\
+
+  AtteanX::Store::LDF & Kjetil Kjernsmo, Patrick Hochstenbach &
+  \url{phochste/AtteanX-Store-LDF} \\
+
+  LWP::UserAgent::CHICaching & Kjetil Kjernsmo &
+  \url{kjetilk/p5-lwp-useragent-chicaching} \\
+
+  Attean & Gregory Todd Williams, Kjetil Kjernsmo &
+  \url{kasei/attean} \\
+
+  AtteanX::Endpoint & Gregory Todd Williams &
+  \url{kasei/atteanx-endpoint} \\
+
+  RDF::LDF &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
+  Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\
+
+  \hline
+\end{tabular}
+\end{table}
+
+The system also depends on modules I have written that is not part of
+the project: RDF::LinkedData, RDF::Generator::Void, Test::RDF,
+URI::NamespaceMap and RDF::NS::Curated.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-04 dissertation/introdefhistory.tex

@@
 The following challenges several conventions of scientific
-writing. Passive voiced is generally overused in scientific writing,
+writing. Passive voice is generally overused in scientific writing,
 sometimes because the author may think it emphasises the objectivity
@@
 their own role. Double blind trials in medicine are an example of why
-it is important to acknowledge the investigators role. Moreover, the
+it is important to acknowledge the investigator's role. Moreover, the
 use of ``we'', when it is clear that the work has been done by a
@@
 challenged by a large intellectual investment in this artefact, the
-more you want this artefact to succeed, the higher may your risk of not
-being sufficiently objective be. I hereby declare that I have a very
-large intellectual investment in the success of the Semantic Web, and
-therefore this section is a reflection on my own role in the
-development of this artefact.
+more you want this artefact to succeed, the higher may your risk of
+not being sufficiently objective be. I hereby declare that I have a
+very large intellectual investment in the success of the Semantic Web,
+and therefore this section is a reflection on my own role in the
+development of this artefact. This serves to clarify \emph{my} vision
+as well as aspects that may influence the objectivity of the
+investigation.
 
@@
 undoubtedly have been influenced by my intellectual investment in the
-Semantic Web, the papers will show that the evaluation has not.
+Semantic Web, the papers should show that the investigation has not.

### 2016-05-04 dissertation/problems.tex
Discuss the SIGMOD paper on caching

@@
 this work does not allow a satisfactory solution to these
-problems. The paper discussed in Section~\ref{sec:conphil} serves to
-excuse this.
+problems. The paper discussed in Section~\ref{sec:conphil} is
+motivated in part from these problems.
 
@@
 algorithms that have a reasonable chance of returning the same digest
-if two queries are equivalent. Some works have created cryptographical
-hashing functions, but they may be overly costly to compute for this
-purpose. % TODO ref.
+if two queries are equivalent. This problem was addressed by
+\cite{papailiou2015graph} and their approach might be portable to be
+used in an HTTP context.
 

### 2016-05-10 dissertation/cache-proxy.tex
mention number of Debian packages

@@
 published to the Comprehensive Perl Archive Network as Free Software,
-and most of which have become a part the Free Software ecosystem. As
-such, it builds on the code of hundreds of authors, too long to
-enumerate. I shall constrain myself to list the modules that have code
-that has been motivated from this study. Modules that have other
-primary authors than myself have been started prior to this present
-project, but may have substantial contributions from this project. The
-Attean framework was started as part of the effort in
-Section~\ref{sec:conpush}, but even though it has been developed
-further with the requirements that arose from this project and has
-contributions from me, the vast majority of the code has been written
-my Gregory Todd Williams. The main work in this project has gone into
-the module AtteanX::Query::Cache, where I have authored the vast
-majority of the code. The relative difference in authorship can be
-further examined by checking the linked Github repositories, or by
-using git2prov\footnote{see \url{http://git2prov.org/}} to create RDF
-data of the commit history. Table~\ref{tab:modules} shows the modules
-that have been influenced or written as part of this project.
+and most of which have become a part the Free Software ecosystem. To
+install the system on the top of a Debian base system requires 146
+Debian packages. As such, it builds on the code of hundreds of
+authors, too long to enumerate. I shall constrain myself to list the
+modules that have code that has been motivated from this
+study. Modules that have other primary authors than myself have been
+started prior to this present project, but may have substantial
+contributions from this project. The Attean framework was started as
+part of the effort in Section~\ref{sec:conpush}, but even though it
+has been developed further with the requirements that arose from this
+project and has contributions from me, the vast majority of the code
+has been written my Gregory Todd Williams. The main work in this
+project has gone into the module AtteanX::Query::Cache, where I have
+authored the vast majority of the code. The relative difference in
+authorship can be further examined by checking the linked Github
+repositories, or by using git2prov\footnote{see
+  \url{http://git2prov.org/}} to create RDF data of the commit
+history. Table~\ref{tab:modules} shows the modules that have been
+influenced or written as part of this project.
 
@@
 
+\subsection{The Attean framework}
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-12 dissertation/cache-proxy.tex
essential debian
mention the test suite
more about the modules

--- 2016-05-04 dissertation/cache-proxy.tex
@@
 published to the Comprehensive Perl Archive Network as Free Software,
-and most of which have become a part the Free Software ecosystem. As
-such, it builds on the code of hundreds of authors, too long to
-enumerate. I shall constrain myself to list the modules that have code
-that has been motivated from this study. Modules that have other
-primary authors than myself have been started prior to this present
-project, but may have substantial contributions from this project. The
-Attean framework was started as part of the effort in
+and most of which have become a part the Free Software ecosystem. To
+install the system on the top of a Debian system with only essential
+packages requires 146 Debian packages. Running the test suite requires
+even more. As such, it builds on the code of hundreds of authors, too
+long to enumerate. I shall constrain myself to list the modules that
+have code that has been motivated from this study. Modules that have
+other primary authors than myself have been started prior to this
+present project, but may have substantial contributions from this
+project. The Attean framework was started as part of the effort in
 Section~\ref{sec:conpush}, but even though it has been developed
@@
 
+To further understand the roles of the different modules, note the following:
+LWP::UserAgent::CHICaching is a traits-based \cite{traits} implementation of the
+majority of RFC7234. AtteanX::Endpoint is an implementation of the
+server side of SPARQL 1.1 Protocol.
+
+\subsection{The Attean framework}
+
+\subsection{SPARQL protocol client}
+
+\subsection{Linked Data Fragment client}
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-12 dissertation/cache-proxy.tex
Detail AtteanX::Store::SPARQL

@@
 my Gregory Todd Williams. The main work in this project has gone into
-the module AtteanX::Query::Cache, where I have authored the vast
+the module \pmodule{AtteanX::Query::Cache}, where I have authored the vast
 majority of the code. The relative difference in authorship can be
@@
 
-  AtteanX::Query::Cache & Kjetil Kjernsmo, Gregory Todd Williams &
+  \pmodule{AtteanX::Query::Cache} & Kjetil Kjernsmo, Gregory Todd Williams &
   \url{kjetilk/p5-atteanx-query-cache} \\
 
-  AtteanX::Store::SPARQL & Kjetil Kjernsmo &
+  \pmodule{AtteanX::Store::SPARQL} & Kjetil Kjernsmo &
   \url{kjetilk/p5-atteanx-store-sparql} \\
 
-  AtteanX::Store::LDF & Kjetil Kjernsmo, Patrick Hochstenbach &
+  \pmodule{AtteanX::Store::LDF} & Kjetil Kjernsmo, Patrick Hochstenbach &
   \url{phochste/AtteanX-Store-LDF} \\
 
-  LWP::UserAgent::CHICaching & Kjetil Kjernsmo &
+  \pmodule{LWP::UserAgent::CHICaching} & Kjetil Kjernsmo &
   \url{kjetilk/p5-lwp-useragent-chicaching} \\
 
-  Attean & Gregory Todd Williams, Kjetil Kjernsmo &
+  \pmodule{Attean} & Gregory Todd Williams, Kjetil Kjernsmo &
   \url{kasei/attean} \\
 
-  AtteanX::Endpoint & Gregory Todd Williams &
+  \pmodule{AtteanX::Endpoint} & Gregory Todd Williams &
   \url{kasei/atteanx-endpoint} \\
 
-  RDF::LDF &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
+  \pmodule{RDF::LDF} &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
   Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\
@@
 The system also depends on modules I have written that is not part of
-the project: RDF::LinkedData, RDF::Generator::Void, Test::RDF,
-URI::NamespaceMap and RDF::NS::Curated.
+the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
+\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.
 
-To further understand the roles of the different modules, note the following:
-LWP::UserAgent::CHICaching is a traits-based \cite{traits} implementation of the
-majority of RFC7234. AtteanX::Endpoint is an implementation of the
+To further understand the roles of the different modules, note the
+following: \pmodule{LWP::UserAgent::CHICaching} is a traits-based \cite{traits} implementation of the
+majority of RFC7234. \pmodule{AtteanX::Endpoint} is an implementation of the
 server side of SPARQL 1.1 Protocol.
@@
 \subsection{SPARQL protocol client}
+
+The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
+Protocol client that implements Attean APIs. The store implementation
+itself composes the TripleStore API and implements methods to retrieve
+the results of a single triple pattern and exact cardinality for a
+given triple pattern, by using an aggregate query. In addition, it can
+generate plans for Basic Graph Pattern algebra objects that have more
+than one triple pattern. If that is the case, it will return an
+instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
+also defined by the module. If it is not the case, the module also has
+an implementation of the \pcode{access\_plan} method in a
+\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
+composed by a query planner to provide a single triple pattern
+\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
+model implementation that composes the Model and CostPlanner
+APIs. This contains a \pcode{cost\_for\_plan} implementation that will
+provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
+proportional to the number of triple patterns in the plan, but
+penalises plans that does not have triple patterns that are connected
+through a variable (i.e. will cause a Cartesian join) with a factor
+10. See Section~\ref{sec:costheuristics} for details on the cost
+heuristics.
 

### 2016-05-12 dissertation/cache-proxy.tex
LDF client

@@
 
+The Linked Data Fragment client work was started prior to this project
+by Patrick Hochstenbach, and consists of the client code for Triple
+Pattern Fragments in \pmodule{RDF::LDF}. It also includes the Basic
+Graph Pattern optimisation from \ref{ldf2}. Included with the client
+code is an implementation of a triple store of the legacy
+\pmodule{RDF::Trine} framework that predates Attean. He also started
+an Attean store implementation \pmodule{AtteanX::Store::LDF}, but that
+module was subsequently adopted by me, and I have written most of the
+functionality. It composes the TripleStore API and implements triple
+pattern queries and cardinality estimates, both provided by the
+underlying client code as they are a part of the core functionality of
+Triple Pattern Fragments, but also a \pcode{cost\_for\_plan}
+implementation used by the cost planner.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-12 dissertation/cache-proxy.tex
minor

--- 2016-05-12 dissertation/cache-proxy.tex
@@
 
+The Linked Data Fragment client work was started prior to this project
+by Patrick Hochstenbach, and consists of the client code for Triple
+Pattern Fragments in \pmodule{RDF::LDF}. It also includes the Basic
+Graph Pattern optimisation from \ref{ldf2}. Included with the client
+code is an implementation of a triple store of the legacy
+\pmodule{RDF::Trine} framework that predates Attean. He also started
+an Attean store implementation \pmodule{AtteanX::Store::LDF} as a
+separate module, but that module was subsequently adopted by me, and I
+have written most of the functionality. It composes the TripleStore
+API and implements triple pattern queries and cardinality estimates,
+both provided by the underlying client code as they are a part of the
+core functionality of Triple Pattern Fragments, but also a
+\pcode{cost\_for\_plan} implementation used by the cost planner.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-13 dissertation/cache-proxy.tex
TPF client has costplanner

@@
 have written most of the functionality. It composes the TripleStore
-API and implements triple pattern queries and cardinality estimates,
-both provided by the underlying client code as they are a part of the
-core functionality of Triple Pattern Fragments, but also a
-\pcode{cost\_for\_plan} implementation used by the cost planner.
+and CostPlanner APIs and implements triple pattern queries and
+cardinality estimates, both provided by the underlying client code as
+they are a part of the core functionality of Triple Pattern Fragments,
+as well as \pcode{cost\_for\_plan}.
 

### 2016-05-13 dissertation/cache-proxy.tex
versions used for analysis
a bit about attean and traits

@@
   \pmodule{AtteanX::Query::Cache} & Kjetil Kjernsmo, Gregory Todd Williams &
-  \url{kjetilk/p5-atteanx-query-cache} \\
+  \url{kjetilk/p5-atteanx-query-cache} \\ % 0.001_04
 
   \pmodule{AtteanX::Store::SPARQL} & Kjetil Kjernsmo &
-  \url{kjetilk/p5-atteanx-store-sparql} \\
+  \url{kjetilk/p5-atteanx-store-sparql} \\ % 0.008
 
   \pmodule{AtteanX::Store::LDF} & Kjetil Kjernsmo, Patrick Hochstenbach &
-  \url{phochste/AtteanX-Store-LDF} \\
+  \url{phochste/AtteanX-Store-LDF} \\ % 0.02
 
   \pmodule{LWP::UserAgent::CHICaching} & Kjetil Kjernsmo &
-  \url{kjetilk/p5-lwp-useragent-chicaching} \\
+  \url{kjetilk/p5-lwp-useragent-chicaching} \\ % 0.04
 
   \pmodule{Attean} & Gregory Todd Williams, Kjetil Kjernsmo &
-  \url{kasei/attean} \\
+  \url{kasei/attean} \\ % 0.015
 
   \pmodule{AtteanX::Endpoint} & Gregory Todd Williams &
-  \url{kasei/atteanx-endpoint} \\
+  \url{kasei/atteanx-endpoint} \\ % 0.001
 
   \pmodule{RDF::LDF} &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
-  Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\
+  Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\ % 0.17
 
@@
 \subsection{The Attean framework}
+
+The Attean framework was born from the experiences the PerlRDF
+community gained with \pmodule{RDF::Trine}, the Perl counterpart to
+the more well-known Jena framework, and the opportunities presented by
+the introduction of traits \cite{traits}, or roles as they are known,
+to the Perl world.
+
+Traits are groups of methods that serve as a primitive unit of code
+reuse. Traits are not constructed to objects themselves and do not
+use inheritance for composition, instead a class is composed by a set of
+traits and possibly the class' own methods, and then instantiated. A
+trait can also require a set of methods, and as such, is similar to
+interfaces, but since its focus is code reuse, it can also provide a
+default implementation.
 

### 2016-05-13 dissertation/cache-proxy.tex
more on attean

@@
 
+The Attean framework packages commonly used classes when using
+Semantic Web data, like RDF statements and terms, parsers,
+serializers, triple/quad stores, iterators, but also features that are
+relevant to query answering, like algebra, query planners and
+plans. Additionally, it has an underlying extensive API consisting of
+roles that can be composed to the above classes.
+
+Perl is an interpreted language and is not known to run fast. Instead,
+the focus is on developer efficiency. However, our motivation for
+Attean, expressed in \cite{williamspushing}, was to enable the use
+of faster implementations by using roles to simplify code reuse.
+
 \subsection{SPARQL protocol client}

### 2016-05-16 dissertation/cache-proxy.tex
about planning and stores

@@
 
+For this work, the usage of roles to simplify query planning is of
+particular importance. While we do not exploit optimisations closer to
+the database, we extend the query planner in several directions with
+several separate add-on modules, with only generic functionality in
+the core Attean framework.
+
+We have found it convenient to use classical inheritance in
+combination with traits-based composition, but only when the base
+class provide only fundamental functionality that is very likely to be
+common to all implementations. For an example of this, see
+Section~\ref{sec:cacher}.
+
+Also of importance are the APIs to compose stores. Stores in Attean
+can be triplestores or quadstores, they can be mutable,
+bulkupdateable, can enable the use of \httph{ETag}s or modification
+times for conditional requests, and makes possible further
+extensions. Each of these primitives is represented by a role that
+implements default functionality or simply requires such functionality
+to be required. For instance, a quadstore is required by
+\pmodule{Attean::API::QuadStore} to implement the \pcode{get\_quads}
+method, which can take an RDF term as one or more of the arguments and
+variables for the rest, and should return an iterator over the quads
+that were matched. Based on an implementation of the
+\pcode{get\_quads} method, the \pmodule{Attean::API::QuadStore} role
+provides default implementations of methods \pcode{count\_quads},
+\pcode{count\_quads\_estimate}, \pcode{get\_graphs} and
+\pcode{size}. These methods may be overridden when composing a
+class. Composing other roles will require further methods to be
+implemented.\todo{Explain Model}
+
 \subsection{SPARQL protocol client}
@@
 
+\subsection{The query proxy}\label{sec:cacher}
+
+For example, the query planner that can use a local cache and a remote
+endpoint, named \pmodule{AtteanX::QueryPlanner::Cache}, extends the
+basic query planner of the Attean framework,
+\pmodule{Attean::QueryPlanner}, but also composes the following roles:
+\pmodule{Attean::API::NaiveJoinPlanner},
+\pmodule{Attean::API::SimpleCostPlanner},
+\pmodule{AtteanX::API::JoinRotatingPlanner},
+\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
+\pmodule{AtteanX::Query::AccessPlan::Cache}.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-16 dissertation/cache-proxy.tex
planning

@@
 
+When a query is parsed, algebra objects are generated. Based on the
+algebra tree, models (which may delegate the task to the store) may
+generate plans for any algebra object of their choosing by
+implementing the \pmodule{Attean::API::CostPlanner} role. This amounts
+to implementing \pcode{plans\_for\_algebra} and
+\pcode{cost\_for\_plan} methods. For a given algebra, Attean's query
+planner will trust that the model's \pcode{plans\_for\_algebra} is
+better than its own by default, a choice that was justified in
+\cite{williamspushing}. Additionally, Attean has a simpler mechanism
+for single triple or quad patterns: By default, triple or quad
+patterns are evaluated by calling the store's \pcode{get\_triples} or
+\pcode{get\_quads}, but this may be augmented by adding a wrapper
+around an \pcode{access\_plans} method used by the query planner to
+produce further plans.
+
+Based on the algebra tree and plans generated by the models or the
+default query planner, as well as the access plans, several
+alternative plans will be generated, and their cost will be
+estimated. How this happens in detail depends on how the planner is
+composed. Attean comes with two roles that implement different
+planners, \pmodule{Attean::API::SimpleCostPlanner} will compute all
+possible plans and then estimate the cost of all plans, and finally
+return the best 5. \pmodule{Attean::API::IDPJoinPlanner}, which
+implements an Iterative Dynamic Programming planner
+\cite{Kossmann:2000:IDP:352958.352982}, which prunes plans
+aggressively based on their estimated cost in each iteration of query
+plan generation. When a final plan tree has been found, the query may
+be executed.
+
 \subsection{SPARQL protocol client}

### 2016-05-16 dissertation/cache-proxy.tex
changes in wording

@@
 can be triplestores or quadstores, they can be mutable,
-bulkupdateable, can enable the use of \httph{ETag}s or modification
-times for conditional requests, and makes possible further
-extensions. Each of these primitives is represented by a role that
-implements default functionality or simply requires such functionality
-to be required. For instance, a quadstore is required by
-\pmodule{Attean::API::QuadStore} to implement the \pcode{get\_quads}
-method, which can take an RDF term as one or more of the arguments and
-variables for the rest, and should return an iterator over the quads
-that were matched. Based on an implementation of the
-\pcode{get\_quads} method, the \pmodule{Attean::API::QuadStore} role
-provides default implementations of methods \pcode{count\_quads},
+bulkupdateable, can enable the use of ETags or modification times for
+conditional requests (as defined in RFC7232 \cite{rfc7234}, and makes
+possible further extensions. Each of these primitives is represented
+by a role that implements default functionality or simply requires
+such functionality to be required. For instance, a quadstore is
+required by \pmodule{Attean::API::QuadStore} to implement the
+\pcode{get\_quads} method, which can take an RDF term as one or more
+of the arguments and variables for the rest, and should return an
+iterator over the quads that were matched. Based on an implementation
+of the \pcode{get\_quads} method, the \pmodule{Attean::API::QuadStore}
+role provides default implementations of methods \pcode{count\_quads},
 \pcode{count\_quads\_estimate}, \pcode{get\_graphs} and
@@
 
+That new plan types can be integrated into the planner with ease, and
+that the  planner itself can be changed with very little effort is a
+key contribution of the Attean system, which I have taken advantage of
+in this work. That this is made easy is to the credit of traits, as
+
 \subsection{SPARQL protocol client}
@@
 Pattern Fragments in \pmodule{RDF::LDF}. It also includes the Basic
-Graph Pattern optimisation from \ref{ldf2}. Included with the client
+Graph Pattern optimisation from \cite{verborgh2014querying}. Included with the client
 code is an implementation of a triple store of the legacy

### 2016-05-16 dissertation/cache-proxy.tex
finish the desc of Attean
start description of caching proxy

@@
 in this work. That this is made easy is to the credit of traits, as
+will become clear as we study the details of the implementation in
+Section~\ref{sec:cacher}.
 
@@
 
-\subsection{The query proxy}\label{sec:cacher}
+\subsection{The caching proxy}\label{sec:cacher}
+
+The caching proxy consists of several components:
+
+\begin{itemize}
+\item An analyser, used to determine whether certain triple patterns
+  should have their results prefetched and cached.
+\item A prefetcher that performs the needed actions.
+\item Two query planners, one that can use a local cache and a remote
+  endpoint, and another that extends this to be able to use Triple
+  Pattern Fragments as well, and corresponding models.
+\item Two roles to generate plans for accessing cache and Triple
+  Pattern Fragments and a corresponding plan class to enter results
+  in the cache.
+\item A custom User Agent for caching.
+\item The actual caching proxy that accepts queries, runs the planner
+  and returns the results.
+\item Scripts to run the whole system.
+\end{itemize}
 

### 2016-05-18 dissertation/introdefhistory.tex
new section on objectivity

@@
 \emph{universality} and therefore, any actor should be as free as
-possible to adapt it to their visions and needs, and so, a pursuit of
-such a definition may not be desirable. Therefore, rather than
-defining it, one should carefully declare one's visions.
+possible to adapt it to their visions and needs. A pursuit of a
+definition of the term ``Semantic Web'' would require a philosophical
+treatment which would be beyond the scope of this thesis, and even
+with that, may not be a fruitful exercise from a technical
+perspective.
+
+Therefore, rather than attempt a definition, understanding is rather
+promoted by following the example of \cite{berners2001semantic} and
+declare a vision.
+
+\section{Objectivity of the study}
+
+A vision has necessarily a strong personal element. This challenges
+the ideals of objectivity in scientific investigations. A longer
+discussion of the desirability and attainability of objectivity in
+science is beyond the scope of this thesis, but objectivity can be
+challenged from many different angles
+\cite{sep-scientific-objectivity}.
+
+The impact of the Web in particular and information technologies in
+general, on society is great, and research projects are often funded
+because of their relevance to society. With this perspective, this
+research is hardly value-free. As individual researchers, we may
+develop a strong relationship to the systems we create, as we have a
+great intellectual investment in them. This may influence our ability
+to reject a flawed hypothesis even if we should. Even though we note
+in the discussion of the contributions in Section~\ref{sec:conphil}
+the weaknesses of Thomas Kuhn's philosophy of science, his insistence
+of understanding science as a social endeavour stands firm.
+
+These challenges are inescapable, and therefore, it is important that
+the researcher acknowledges such problems and explicitly details their
+own personal motivations, visions and possible biases to let the
+reader decide whether the researcher's subjective beliefs has
+influenced the validity of their contribution.
+
+I shall attempt to do so in the following, but also note that this
+influences the writing style. There are several conventions of
+scientific writing that stands in opposition to the goal of letting
+the reader understand the possible influence of subjective
+beliefs. The frequent use of passive voice is one such convention,
+that in my opinion tries to hide subjectivity rather than alleviate
+its harmful effects. If the methodology used in the investigation is
+flawed, it should not be obscured by writing style, the methodology
+will need to be enhanced. Double blind trials in medicine are an
+example of why it is important to acknowledge the investigator's
+role. That is not to say that the use of passive voice is always
+inappropriate, when the role of the investigator cannot be
+misunderstood, it can be used just like any other language construct.
+
+Of less importance is the use of ``we'', when it is clear that the
+work has been done by a single investigator. I hold as a general
+principle that the role of the investigator should never be obscured,
+and therefore, such use is inappropriate. The use of pronouns should
+not differ substantially in scientific writing from other writing, and
+I will use them accordingly. In particular, I will use ``we'' when
+more than one person has been involved, or when the reader is included
+(e.g. ``We note that...'').
 
@@
 
-The following challenges several conventions of scientific
-writing. Passive voice is generally overused in scientific writing,
-sometimes because the author may think it emphasises the objectivity
-of the investigation. In my opinion it rather attempts to hide the
-fact that science is a human and social endeavour, and only by
-acknowledging that fact, an investigator can properly account for
-their own role. Double blind trials in medicine are an example of why
-it is important to acknowledge the investigator's role. Moreover, the
-use of ``we'', when it is clear that the work has been done by a
-single investigator is inappropriate as it obscures
-the role of the investigator.
-
-As noted, the Semantic Web is a complex technical, social, economic
-and cultural artefact. The objectivity of an investigator may be
-challenged by a large intellectual investment in this artefact, the
-more you want this artefact to succeed, the higher may your risk of
-not being sufficiently objective be. I hereby declare that I have a
-very large intellectual investment in the success of the Semantic Web,
-and therefore this section is a reflection on my own role in the
-development of this artefact. This serves to clarify \emph{my} vision
-as well as aspects that may influence the objectivity of the
-investigation.
+As noted in the previous section, objectivity may be challenged by a
+researcher's intellectual investment in a certain field of study. I
+admit that I have a very large intellectual investment in the success
+of the Semantic Web. This section serves two purposes: It should give
+the reader a good understanding of this intellectual investment, and
+enable them to assess the impact of any subjective beliefs. Secondly,
+it serves to clarify \emph{my} vision, which we noted in
+Section~\ref{sec:nodef} is more important than to attempt a
+definition. The vision strongly influences the motivation detailed in
+Section~\ref{sec:motivation}.
 

### 2016-05-18 dissertation/cache-proxy.tex
more analysers
more on analysis

@@
 
-\subsection{Linked Data Fragment client}
+\subsection{Linked Data Fragment client}\label{sec:ldfclient}
 
@@
 
+It also supplies a plan class \pmodule{AtteanX::Plan::LDF::Triple}
+used in the query planning.
+
 \subsection{The caching proxy}\label{sec:cacher}
@@
 \begin{itemize}
-\item An analyser, used to determine whether certain triple patterns
+\item Analysers, used to determine whether certain triple patterns
   should have their results prefetched and cached.
@@
 \end{itemize}
+
+\subsubsection{Analysis and prefetching}
+
+Analysers and prefetchers run asynchronously with query
+evaluation. This is achieved by an addition to the proxy that sends a
+query to a persistent analyser script that subscribes to a channel on
+a Redis\footnote{See \url{http://redis.io/}} data structure store.
+
+The analyser script can be configured to use any number of analysers.
+Two such analysers have been implemented, for now constrained to
+single triple patterns. One will count the predicates in consecutive
+queries it sees, and publish triple patterns to another Redis channel
+with the predicate if the count exceeds a certain configurable
+limit. This is an attempt to capture and cache popular parts of
+queries.
+
+The second analyser will rerun the query planner on the incoming query
+while simulating that the results of a certain triple pattern was in
+the cache. If the cost reduction of the best plan is larger than a
+configurable limit, it will publish that triple pattern to the same
+Redis channel as above.
+
+Another persistent script is subscribed to this latter channel. If the
+Model supports Linked Data Fragments, a retriever will then download
+the results of the triple pattern and enter it into the cache, or use
+a single triple pattern SPARQL query if not.
+
+The cache itself is assumed to only cache triples where at least one
+term is bound. I have used the Redis data structure store for the
+cache as well. This choice is somewhat arbitrary, and has not proved
+very successful. The cache will store the results as an array if two
+RDF terms were bound in the triple pattern, or a hash if only one term
+was bound. The strings put into the cache are serialized N-Triples
+strings.
+
+In addition to caching the prefetched results, the proxy may also
+cache the serialized results of any full SPARQL query, and also enter
+the results of any Linked Data Fragment retrieved when the query
+planner evaluates a query into the above cache, using a trivial
+extension to the plan class in Section~\ref{sec:ldfclient}.
+
+\subsubsection{Query planner}
 

### 2016-05-18 dissertation/cache-proxy.tex
describe technical contribs

@@
 
+In terms of technical contribution, most of these are
+straightforward. The proxy itself, for example, is just a few lines of
+wrapper code around the \pmodule{AtteanX::Endpoint} module. The User
+Agent is a class that composes the roles of
+\pmodule{LWP::UserAgent::CHICaching} and adds a hash of the query as
+the cache key.
+
+The main technical contributions are in the analysis of the query and
+in the query planning.
+
 \subsubsection{Analysis and prefetching}

### 2016-05-18 dissertation/cache-proxy.tex
redo refs

@@
 common to all implementations. For an example of this, see
-Section~\ref{sec:cacher}.
+Section~\ref{sec:implqueryplan}.
 
@@
 will become clear as we study the details of the implementation in
-Section~\ref{sec:cacher}.
+Section~\ref{sec:implqueryplan}.
 
@@
 
-\subsubsection{Query planner}
+\subsubsection{Query planner}\label{sec:implqueryplan}
 

### 2016-05-18 dissertation/cache-proxy.tex
more on planning

@@
 
-\subsubsection{Analysis and prefetching}
+\subsubsection{Analysis and prefetching}\label{sec:analpre}
 
@@
 
-For example, the query planner that can use a local cache and a remote
-endpoint, named \pmodule{AtteanX::QueryPlanner::Cache}, extends the
-basic query planner of the Attean framework,
-\pmodule{Attean::QueryPlanner}, but also composes the following roles:
+The query planner is written as completely stand-alone modules, based
+on the facilities in the Attean framework, but without accessing
+Attean internals. The system has in fact three query planners:
+\pmodule{AtteanX::QueryPlanner::Cache},
+\pmodule{AtteanX::QueryPlanner::Cache::LDF} and
+\pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}. Each is an
+extension of the previous. The first is written to use a remote
+endpoint in combination with a local cache. The second extends this to
+also be able to generate and use Triple Pattern Fragments plans. The
+final is a query planner for the analyser in Section~\ref{sec:analpre}
+that reruns the query planning with all triple patterns in a simulated
+cache.
+
+These query planners are a good example of the virtues of traits-based
+programming: \pmodule{AtteanX::QueryPlanner::Cache} extends the basic query
+planner of the Attean framework, \pmodule{Attean::QueryPlanner}, but
+also composes the following roles:
 \pmodule{Attean::API::NaiveJoinPlanner},
@@
 
+\pmodule{Attean::QueryPlanner} has only minimal functionality,
+everything else is provided by the roles. With conventional single
+inheritance, one would end up with a rather complex base class. While
+multiple inheritance or mix-in could be employed, \cite{traits}
+explains why this is an inferior solution.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-19 dissertation/cache-proxy.tex
more on the use of roles

@@
 These query planners are a good example of the virtues of traits-based
-programming: \pmodule{AtteanX::QueryPlanner::Cache} extends the basic query
-planner of the Attean framework, \pmodule{Attean::QueryPlanner}, but
-also composes the following roles:
+programming, as they are typically customised along different axes:
+access plans, cost estimation, planning of joins within Basic Graph
+Patterns or left joins, rewriting (like gathering quad patterns to
+blocks), etc. In the case of our query planner that combines a remote
+endpoint with a local cache, named
+\pmodule{AtteanX::QueryPlanner::Cache}, the basic query planner of the
+Attean framework, \pmodule{Attean::QueryPlanner} is extended
+(i.e. inherited from), but also composes the following roles:
 \pmodule{Attean::API::NaiveJoinPlanner},
@@
 \pmodule{Attean::QueryPlanner} has only minimal functionality,
-everything else is provided by the roles. With conventional single
-inheritance, one would end up with a rather complex base class. While
-multiple inheritance or mix-in could be employed, \cite{traits}
-explains why this is an inferior solution.
+everything else is provided by the roles. Conventional single
+inheritance is constrained to tree structure, and would result in a
+very complex system. Another option is to use the factory method
+pattern, where a factory object would need to be initialised to return
+a suitable object for each axes of customisation.
 

### 2016-05-19 dissertation/cache-proxy.tex
typo

@@
 pattern, where a factory object would need to be initialised to return
-a suitable object for each axes of customisation.
+a suitable object for each axis of customisation.
 

### 2016-05-19 dissertation/cache-proxy.tex
join rotation

@@
 
-\subsection{SPARQL protocol client}
+\subsection{SPARQL protocol client}\label{sec:sparqlclient}
 
@@
 \pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
-\pmodule{AtteanX::Query::AccessPlan::Cache}.
+\pmodule{AtteanX::Query::AccessPlan::Cache}, out of which the three
+first are a part of the Attean framework, the next was described in
+Section~\ref{sec:sparqlclient} and the last is a part of the
+query cache system itself.
 
@@
 
+It is interesting to detail the function of
+\pmodule{AtteanX::API::JoinRotatingPlanner}. Its main function is to
+produce alternatives for join query plans. For example, if we have
+three quad patterns $A$, $B$ and $C$, plans for both $ ( A \bowtie B)
+\bowtie C $ and $ A \bowtie ( B \bowtie C ) $ will be generated. In
+some cases, it may be possible to reduce a join to a more fundamental
+plan type, and this may be done next. In our case, this is done in
+\pmodule{AtteanX::QueryPlanner::Cache}, where adjacent quad patterns
+are coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
+coalesced into larger BGP plans, so that they can be evaluated by the
+remote SPARQL endpoint as a single unit. This will minimise the number
+of HTTP requests, in much the same way as the exclusive groups of
+\cite{springerlink:10.1007/978-3-642-25073-6-38}. Note, however, that
+the generated plans are subject to cost estimation, so whether a large
+BGP plan will be evaluated remotely, or as a join of local cached
+results and a smaller remote BGP, or in the case of
+\pmodule{AtteanX::QueryPlanner::Cache::LDF}, a remote Triple Pattern
+Fragments server, is up to the cost estimation. This will be discussed
+in detail in Section~\ref{sec:costheuristics}.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-20 dissertation/cache-proxy.tex
cost model details

@@
 
+\subsection{Cost Model}\label{sec:costheuristics}
+
+The final query execution is designed to be entirely cost-based,
+i.e. many alternative plans are generated, but the decision on whether
+to use a certain plan is entirely up to the cost model. While the
+framework is flexible and could support further research into cost
+models, I have considered this to be future work, and only a small
+heuristic cost model is present in the current system.
+
+However, since the combination of remotely executed SPARQL with a
+local cache and Triple Pattern Fragments is rather complex, the cost
+heuristics are fairly elaborate. Moreover, since the objective of the
+study is to take load off of the remote endpoint, the heuristics is
+intended to make the proxy work harder than the remote endpoint.
+
+Throughout the system, the query planner takes the cost from the
+models implementing the \pmodule{Attean::API::CostPlanner} role. A
+key challenge is to balance the costs given by different models in the
+system, and this is done in
+\pmodule{AtteanX::Model::SPARQLCache::LDF}. Let us, however, start
+from the bottom.
+
+The cost of evaluating a Triple Pattern Fragment plan $C_{tpf}$ is estimated to
+be 
+\begin{equation}
+C_{tpf} = 10 + \floor{990 \frac{n_{tpf}}{n_{tot}}} ,
+\end{equation}
+where $n_{tpf}$ is the estimated number of triples expected to be
+matched by the triple pattern and $n_{tot}$ is the size of the
+model. Thus, this cost will be an integer between 10 and 1000. The
+reason we use a floor function is that the Attean cost model only
+supports integers.
+
+The cost of accessing the cache is assumed to be small compared to the
+above, and is estimated as 
+\begin{equation}
+C_c = 2 + \floor{\log_{10}{n_c}} , 
+\end{equation}
+where $n_c$ is the number of triples in the cache that matches the
+given triple pattern.
+
+The cost of evaluating a remote SPARQL query is estimated to be the
+number of subplans (usually quad patterns) incremented by one and
+multiplied by 100 if there is a common variable between all the
+subplans, or 1000 if there is not. This is to prevent the remote
+endpoint from having to compute Cartesian joins as much as possible. A
+remote SPARQL query with only one triple pattern is assumed to always
+be more expensive than to evaluate than a Triple Pattern Fragment
+(recall that we want to relieve the remote endpoint of stress), and so
+is given the cost of 1001.
+
+Next, we detail the cost of a join where one of more children in the
+tree is a remote SPARQL plan. It is usually not wanted to execute
+several remote queries, it is assumed that the number of HTTP requests
+should be small, and that plans where the joins have been rotated and
+coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
+less expensive. Therefore, such plans should be penalised, but by how
+much depends on the type plans. Conventionally, for a nested loop join
+plan, we multiply the cost of the subplans, increment the resulting
+cost by one if the right hand plan is more expensive than the left
+hand plan, and multiply by a factor of 10 if there are no shared
+variables, to mildly penalise Cartesian joins. If the plan is a hash
+join the difference is that, we add rather than multiply the cost of
+the of the subplans, and multiply by a factor of 100 rather than 10 so
+that if it is necessary to evaluate a Cartesian join, it is more
+likely done using a nested loop join. Then, the resulting cost is
+multiplied by the number of remote SPARQL Basic Graph Patterns in the
+tree and the factor 1.2, and floored to the nearest integer.
+
+Under the assumption that it is better to join remotely than to
+possibly retrieve large amounts of data, a similar algorithm is used
+to penalise join plans that contain many Triple Pattern Fragment
+plans, by multiplying the cost of the join plan by the number of the
+Triple Pattern Fragment plans in the tree.
+
+Finally, if a plan has a Triple Pattern Fragment plan that has a
+common variable with a remote SPARQL query, that plan is given an
+added cost to 1000, to allow plans where the triple pattern has been
+rotated into the BGP win, again to reduce the amount of data needed to
+be transferred.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-20 dissertation/cache-proxy.tex
intro to evaluation

@@
 
+\section{Evaluation}
+
+The system was developed in a practically oriented test-driven manner,
+with emphasis on testing cases that I assumed \textit{a priori} to be
+difficult. The interested reader is referred to the published code,
+given in Section~\ref{sec:impl}. Some of the tests consist of a number
+of queries, and some of the tests concerns the problem I found in a
+preliminary study, that the cached triples would often break
+subsequent queries to cause Cartesian joins. Much of the above cost
+model is design to prevent this from having a detrimental impact.
+
+While I am quite confident that the code is producing query plans as
+intended and that analysis and prefetching is able to retrieve and
+cache results, an elaborate evaluation would be needed to further
+explore whether the system actually helps remote endpoints and is able
+to serve clients with adequate performance. Unfortunately, the
+evaluation showed that the latter is not the case with the current
+implementation, and the practical constraints of the project then
+prevented further investigation. Thus, this section consists of an
+account of the evaluation that were performed and the methodology I
+planned to employ.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-20 dissertation/cache-proxy.tex
profiler output

@@
 
+\subsection{Actual evaluation}
+
+I decided to set the entire system up as a proxy server, and run a
+realistic query against DBpedia repeatedly. The intention was to let
+the analyser find all required data and eventually add that to the
+cache. Once the cache was filled with all the data, the query cache
+should be able to answer the query on its own, at least within a time
+period on the order of a magnitude of the original endpoint (which
+took ca. 1~second).
+
+The following query was used:
+
+\begin{verbatim}
+PREFIX foaf: <http://xmlns.com/foaf/0.1/>
+PREFIX dbo: <http://dbpedia.org/ontology/>
+SELECT ?name WHERE {
+  ?s a foaf:Person ;
+     foaf:name ?name ;
+     dbo:wikiPageID  9828878 .
+} ORDER BY ?name
+\end{verbatim}
+
+The query was chosen because of its wide variety in terms of
+selectivity of the individual triple patterns. The retriever managed
+to pull all the data from the Linked Data Fragments version of DBpedia
+within some hours.
+
+The cache was kept in a Redis data structure database on a machine
+with a Intel Core2 Duo E8500 CPU with 16 GB of RAM. The proxy server
+with the query engine ran on a  Intel Xeon CPU E3-1230 v5 CPU with 32
+GB of RAM. Running the query on this query took more than 1000 seconds
+without any performance analysis enabled. This is clearly more than
+two orders of magnitude longer than acceptable. Moreover, it is then
+clearly not possible to draw any conclusions as to whether the system
+would be helpful in achieving its goal of taking load off of the
+remote server, as it could process far fewer queries than the public
+remote endpoint, and any load saved on the remote endpoint would
+likely be less than the uncertainty arising from that.
+
+Nevertheless, it is interesting to further understand this failure. In
+Table~\ref{tab:nytprof}, I have tabulated the result of the subroutine
+calls that take the most time.
+
+\begin{table}{l}
+\caption{Top 15 Subroutines Calls as measured by the profiler \pmodule{Devel::NYTProf}.}\label{tab:nytprof}
+
+\begin{tabular}{r p{1.5cm} p{1.5cm} l}
+  \hline
+Calls & 	Exclusive Time & Inclusive Time & Subroutine \\
+\hline
+13309080	& 244s	& 549s & \pcode{IRI::\_parse\_components} \\
+13309080	& 218s	& 218s & \pcode{IRI::CORE:match (opcode)} \\
+8774329	        & 116s	& 270s & \pcode{Attean::Result::new} \\
+10573035	& 111s	& 1420s & \pcode{AtteanX::Parser::NTuples::\_eat\_node} \\
+6355072	 	& 79.7s & 1856s & \pcode{Attean::CodeIterator::next} \\
+2 & 77.4s & 4540s & \pcode{Attean::Plan::HashJoin::\_\_ANON\_\_[Attean/Plan.pm:362]} \\
+4217970		& 70.0s & 847s & \pcode{Attean::Literal::new} \\
+9091102	 	& 61.8s	& 593s & \pcode{Attean::IRI::new} \\
+2419260	        & 50.3s & 208s & \pcode{Attean::API::Result::join} \\
+4217968	        & 48.7s	& 1463s & \pcode{AtteanX::Query::AccessPlan::Cache::\_\_ANON\_\_[AtteanX/Query/AccessPlan/Cache.pm:72]} \\
+24186545	& 48.1s	& 56.6s & \pcode{Role::Tiny::does\_role} \\
+28128408	& 44.4s	& 53.8s	& \pcode{Attean::Result::value} \\
+16871877	& 44.1s	& 470s & \pcode{Attean::Literal::\_\_ANON\_\_[(eval 234)[Class/Method/Modifiers.pm:93]:1] (merge of 4 subs)}\\
+4217969		& 40.0s & 92.0s & \pcode{Attean::API::Literal::\_\_ANON\_\_[Attean/API/Term.pm:160]} \\
+13309111	& 39.5s & 39.5s & \pcode{IRI::CORE:regcomp (opcode)}
+\end{tabular}
+\end{table}
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-20 dissertation/cache-proxy.tex
no restrictions on placement
actual evaluation

@@
 with the query engine ran on a Intel Xeon CPU E3-1230 v5 CPU with 32
-GB of RAM. Running the query on this query took more than 1000 seconds
+GB of RAM, residing a LAN with a gigabit Ethernet link between
+them. Running the query on this query took more than 1000 seconds
 without any performance analysis enabled. This is clearly more than
@@
 
-\begin{table}{l}
+\begin{table}
 \caption{Top 15 Subroutines Calls as measured by the profiler \pmodule{Devel::NYTProf}.}\label{tab:nytprof}
@@
 
+It is clear that the vast majority of time spent is due to my cache
+implementation. When using the Redis cache, I serialise RDF terms as
+N-Triples, and when they are retrieved, they have to be parsed and
+instantiated as objects for the query engine. Remarkably, parsing of
+Internationalized Resource Identifiers (IRIs) is the single most
+time-consuming task. While each parse does not take much time, it
+happens more than 13~million times. The reason why parsing IRIs
+happens is that a certain amount of parsing is required to compare
+them \cite{rfc3987}, and the \pmodule{IRI} module is a rigorous
+implementation of the standard that does syntax based resolution if
+the if an object is constructed with a base. IRI parsing happens not
+only when an \pmodule{Attean::IRI} object is constructed, but may also
+happen when an \pmodule{Attean::Literal} object is constructed, since
+the data type is a IRI.
+
+While most of the time is spent in parsing and comparison, the hash
+joins also require too much time. A discussion of possible
+improvements will follow in Section~\ref{sec:discussfail}.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-24 dissertation/cache-proxy.tex
start story on hypothesis test

@@
 
-\subsection{The Attean framework}
+\subsection{The Attean framework}\label{sec:attean}
 
@@
 
+\subsection{Planned evaluation}
+
+This section outlines the evaluation I had planned to employ on the
+query cacher. The essence is to perform a ``query log replay, while
+recording remote endpoint load under different circumstances,
+determined by a experiment designed with statistical Design of
+Experiments formalism''.
+
+In more detail, query logs can be obtained by from
+e.g. USEWOD~\cite{eps385344} or the Linked SPARQL SPARQL Queries
+Dataset \cite{Saleem2015}. An extraction of one of those datasets
+could be performed, to create a representative progression of
+queries. The data normally residing on a remote endpoint
+(i.e. ``server'') would then be loaded into a single database system
+under my control. A proxy system would be set up using our
+system. Finally, a client would submit the queries to the server, via
+the proxy.
+
+On the server, some metric of system load would recorded. In
+statistical experimentation parlance, this is known as the ``response
+variable''. I had not entirely decided which metric to use, but
+\cite{Ferrari:1987:EIL:647412.725013} indicated the load average may
+be a good choice. It could be trivially recorded by polling the
+virtual file \texttt{/proc/loadavg} on the host system. Possibly a
+correction for any extra time spent by the proxy could be applied.
+
+As for the overall design of the experiment, the reader is referred to
+the paper discussed in Section~\ref{sec:condoe} for an introduction to
+the type of evaluation intended to be used. As noted, a $2^n$
+factorial experiment is the simplest experiment to set up, and I would
+therefore go to great lengths to ensure that such an experiment could
+be used.
+
+It would also be advantageous to design the experiment so that a
+combination of factors would correspond to the null hypothesis,
+i.e. that the proxy was not in use and the remote endpoint would
+tackle the whole load. To achieve this, a fractional factorial
+experiment would likely result where certain combinations would be
+banned. I did not think this complete through to a concrete
+experiment.
+
+However, the factors that may be interesting to investigate include
+both factors that evaluate the analyser and the query planner. These
+factors might be:
+
+\begin{enumerate}
+\item Use of the cache and remote endpoints where the levels are on
+  and off.
+\item Use of a caching Linked Data Fragments client, levels on and
+  off.
+\item Use the predicate count analyser, levels on and off.
+\item Use the best cost improvement analyser, levels on and off.
+\item The order of the application of the different methods for
+  analysis.
+\item Which cost planner to use, with levels being the simple cost
+  planner and the Iterative Dynamic Programming planner, see
+  Section~\ref{sec:attean}.
+\end{enumerate}
+
+While this design would achieve the goal that a combination of levels
+would correspond to the null hypothesis, certain levels are somewhat
+problematic. For example, the Linked Data Fragments client in point~2
+might be an implementation of \cite{verborgh2014querying}, but that
+would be very different from the combination of point~1~and~2, which
+correspond to the usage of
+\pmodule{AtteanX::QueryPlanner::Cache::LDF}. Additionally, I noted
+that downloading the triple pattern fragment \sparql{?s foaf:name
+  ?name .}  resulted in more than 40~000 HTTP requests due to
+paging. Another relevant factor might be to deploy a Triple Pattern
+Fragments server that does not page, and have levels with paging on
+and off. Integrating this factor in the experimental design would
+further complicate it.
+
+If the experiment outlined here could come to fruition, and we take
+the factor in the first point with level~1 as the cache being enabled,
+then we can examine the experiment similarly to what with did with the
+``Implement'' factor in \cite{kjernsmo_doe_intro}. That is, if the
+effect of this factor and its interactions are negative\todo{finish a
+  short story about testing}.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-25 dissertation/cache-proxy.tex
hypothesis test explained

@@
 then we can examine the experiment similarly to what with did with the
-``Implement'' factor in \cite{kjernsmo_doe_intro}. That is, if the
-effect of this factor and its interactions are negative\todo{finish a
-  short story about testing}.
+``Implement'' factor in \cite{kjernsmo_doe_intro}. That is, the first
+step to understand the contributions to the endpoint system load by
+the different factors, is to create a normal plot and then use the
+Lenth criterion for significance. For the next step, to test the
+alternative hypothesis that the endpoint system load has decreased,
+form an average of the effects for all combinations of the factor in
+point~1 and the insignificant factors determined in the first step,
+based on the average over the rest of the factors. Based on this, a
+common one-sided two-sample $t$-test can be performed.
 

### 2016-05-25 dissertation/cache-proxy.tex
comment on orthogonality

@@
 i.e. that the proxy was not in use and the remote endpoint would
-tackle the whole load. To achieve this, a fractional factorial
-experiment would likely result where certain combinations would be
-banned. I did not think this complete through to a concrete
-experiment.
+tackle the whole load. I did not think this complete through to a
+concrete experiment, but it might be a challenging problem to maintain
+orthogonality, i.e. ensure that all level combinations can occur in
+the same number of runs.
 

### 2016-05-25 dissertation/cache-proxy.tex
minor

@@
 period on the order of a magnitude of the original endpoint (which
-took ca. 1~second).
+took on the order of 1~second).
 
@@
 GB of RAM, residing a LAN with a gigabit Ethernet link between
-them. Running the query on this query took more than 1000 seconds
+them. Running the query on this hardware took more than 1000 seconds
 without any performance analysis enabled. This is clearly more than
@@
 \pmodule{AtteanX::QueryPlanner::Cache::LDF}. Additionally, I noted
-that downloading the triple pattern fragment \sparql{?s foaf:name
-  ?name .}  resulted in more than 40~000 HTTP requests due to
+that downloading the triple pattern fragment
+\sparql{?s~foaf:name~?name~.}  
+resulted in more than 40~000 HTTP requests due to
 paging. Another relevant factor might be to deploy a Triple Pattern

### 2016-05-25 dissertation/related.tex
dividino critisism

@@
 
+Dividino et al. \cite{Dividino2015} proposed strategies to keep caches
+of Linked Open Data up to date, in face of changing data. The authors
+sets out to learn the change frequency of resources, and based on that
+schedule updates of the resources. This does then not depend on the
+resource metadata (neither HTTP headers nor metadata in the RDF
+itself, which I studied in \cite{kjernsmo_survey_2015}), which is
+motivated by their finding that only 8\% of \httph{Last-Modified}
+headers they surveyed reflected the true change frequency. While I
+have no reason to doubt this finding, and the monitoring approach is
+interesting, I think the dismissal of resource metadata is
+insufficiently motivated, and therefore the evaluation is
+inadequate. First, the \httph{Last-Modified} header is far from the
+only cache-relevant header, as should be evident by the fact that it
+is not defined in the cache standard, RFC7234, but in the conditional
+request standard, RFC7232. Moreover, RFC7234 contains a Section~4.2.2.
+titled ``Calculating Heuristic Freshness''. It is my opinion that the
+paper should be framed within the loose constraints of this section,
+where it could make a substantial contribution. As it stands, it would
+serve to violate Web standards, which in my opinion, is something the
+Semantic Web community should be more cautious about. If further
+considered within the frames of RFC7234's Section~4.2.2., it also
+proposes a simple heuristic for calculating freshness lifetime. I
+found that the distribution of this heuristic freshness lifetime
+agrees well with The Dynamic Linked Data Observatory (DyLDO)
+\cite{dyldo2}. However, a weakness of my study, which could make this
+observation irrelevant, is if the change frequency of individual
+resources were wrong, in particular, if the use of the simple
+heuristic lead to using stale content.  RFC7234's
+Section~4.2.2. explicitly prohibits the use of a heuristic if an
+explicit expiration time is present (defined in the caching standard
+RFC7234), yet the paper does not attempt to include those in their
+assessment. Furthermore, I found in \cite{kjernsmo_survey_2015} that
+cache prohibitions were common. While they in many cases may be
+misguided, they nevertheless express a strong statement from the
+server administrators, that should be respected. RDF data,
+e.g. \rdfterm{dct:valid} and \rdfterm{dct:modified} could also be used
+to compute a expiration time. Finally, an opaque validator,
+\httph{ETag} is also a part of RFC7232. After considering cache
+prohibitions, explicit expiration times, opaque validators, simpler
+heuristics, etc., the space where learning the change frequency could
+be applied is likely much smaller. However, if understood as a
+sophisticated heuristic with the constraints of RFC7234's
+Section~4.2.2., the contribution is valuable.
+
 While the approaches are diverse, these studies pull in the same

### 2016-05-26 dissertation/related.tex
typo
improve language

@@
 serve to violate Web standards, which in my opinion, is something the
-Semantic Web community should be more cautious about. If further
-considered within the frames of RFC7234's Section~4.2.2., it also
-proposes a simple heuristic for calculating freshness lifetime. I
-found that the distribution of this heuristic freshness lifetime
-agrees well with The Dynamic Linked Data Observatory (DyLDO)
-\cite{dyldo2}. However, a weakness of my study, which could make this
-observation irrelevant, is if the change frequency of individual
+Semantic Web community should be more cautious about. RFC7234's
+Section~4.2.2. also proposes a simple heuristic for calculating
+freshness lifetime. I found that the distribution of this heuristic
+freshness lifetime agrees well with The Dynamic Linked Data
+Observatory (DyLDO) \cite{dyldo2}. However, a weakness of my study
+which could make this observation irrelevant, is that I considered
+only summary statistics, . If the change frequency of individual
 resources were wrong, in particular, if the use of the simple
-heuristic lead to using stale content.  RFC7234's
-Section~4.2.2. explicitly prohibits the use of a heuristic if an
-explicit expiration time is present (defined in the caching standard
-RFC7234), yet the paper does not attempt to include those in their
-assessment. Furthermore, I found in \cite{kjernsmo_survey_2015} that
-cache prohibitions were common. While they in many cases may be
-misguided, they nevertheless express a strong statement from the
-server administrators, that should be respected. RDF data,
-e.g. \rdfterm{dct:valid} and \rdfterm{dct:modified} could also be used
-to compute a expiration time. Finally, an opaque validator,
-\httph{ETag} is also a part of RFC7232. After considering cache
-prohibitions, explicit expiration times, opaque validators, simpler
-heuristics, etc., the space where learning the change frequency could
-be applied is likely much smaller. However, if understood as a
-sophisticated heuristic with the constraints of RFC7234's
-Section~4.2.2., the contribution is valuable.
+heuristic lead to using stale content, it would be inappropriate to
+use the simple heuristic.  RFC7234's explicitly prohibits the use of a
+heuristic if an explicit expiration time is present, yet the paper
+does not attempt to include those in their assessment. Furthermore, I
+found in \cite{kjernsmo_survey_2015} that cache prohibitions were
+common. While they in many cases may be misguided, they nevertheless
+express a strong statement from the server administrators, that should
+be respected. RDF data, e.g. \rdfterm{dct:valid} and
+\rdfterm{dct:modified} could also be used to compute a expiration
+time. Finally, an opaque validator, \httph{ETag} is also a part of
+RFC7232. After considering cache prohibitions, explicit expiration
+times, opaque validators, simpler heuristics, etc., the space where
+learning the change frequency could be applied is likely much
+smaller. However, if understood as a sophisticated heuristic within
+the constraints of RFC7234's Section~4.2.2., the contribution is
+valuable.
 

### 2016-05-26 dissertation/cache-proxy.tex
typo
small

@@
 subsequent queries to cause Cartesian joins. Much of the above cost
-model is design to prevent this from having a detrimental impact.
+model is designed to prevent this from having a detrimental impact.
 
@@
 In more detail, query logs can be obtained by from
-e.g. USEWOD~\cite{eps385344} or the Linked SPARQL SPARQL Queries
+e.g. USEWOD~\cite{eps385344} or the Linked SPARQL Queries
 Dataset \cite{Saleem2015}. An extraction of one of those datasets
@@
 
+\section{Discussion}
+
+\subsection{Join rotation and Iterative Dynamic
+  Programming}
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-26 dissertation/cache-proxy.tex
discuss join rotation

@@
 
+While the system used a simple brute-force query planner in the above
+evaluation, using an Iterative Dynamic Programming planner
+\cite{Kossmann:2000:IDP:352958.352982} would simply amount to
+composing \pmodule{Attean::API::IDPJoinPlanner} in the planner class.
+
+While this performed better in most cases, the test suite contains a
+query where it does not:
+\begin{verbatim}
+?s <http://example.org/m/p> "1" ;
+   <http://example.org/m/p> ?o ;
+   <http://example.org/m/q> _:xyz ;
+   <http://example.org/m/q> <http://example.org/m/a> .
+?a <http://example.org/m/b> <http://example.org/m/c> .
+\end{verbatim}
+Note that the above Basic Graph Pattern contains a triple pattern that
+does not share a variable (\sparql{?a}) with any other triple pattern,
+thus creating a problematic Cartesian join. Additionally, in this
+test, the results of the two first triple patterns have been
+cached. \textit{A priori}, I have assumed that the best plan is a plan
+where the cache is used for these two triple patterns, the last triple
+pattern is answered by a Linked Data Fragments server and the last two
+triple patterns are coalesced into a new SPARQL query, and evaluated
+remotely. This is also coded as a pass criterion for the test. With
+the simple query planner, this test passes, but not with the IDP
+planner, as it results in a plan where the two triple patterns that
+should run is a single SPARQL query is split into two. This resulted
+in an interesting observation:
+
+As noted in Section~\ref{sec:implqueryplan}, we rely on rotating joins
+to gather triple pattern plans into a larger structure that can be
+evaluated together, like a remote SPARQL BGP. On careful inspection, I
+found that the join rotation would have had to descend into
+great-grandchildren of the top join plan. However, with the IDP
+planner, plans are aggressively pruned, and therefore, this
+constellation is not considered. This led us (with Gregory Todd
+Williams) to conjecture that this weakness is inherit to IDP planning
+when using join rotation. We did not attempt to examine this any
+further, however.
+
+To remedy this problem, we considered creating a custom query planner
+for Basic Graph Patterns, but I did not implement it. The outline
+would be as follows:
+\begin{enumerate}
+\item Find triple patterns that are connected with a common variable,
+  and group them to ``components''.
+\item If a triple pattern have a cached result, create a plan that
+  accesses the cache. 
+\item If the triple pattern in the previous step caused a Cartesian
+  join to be created within a component, create an equivalent
+  additional plan for a SPARQL remote evaluation that does not access
+  the local cache.
+\item Any remaining components consisting of a single triple pattern
+  results in a LDF plans.
+\item Any remaining components are turned into SPARQL remote BGP
+  plans.
+\end{enumerate}
+
+This procedure would allow the cost model to balance the cost between
+doing a Cartesian join and doing a remote SPARQL evaluation of data
+already in the cache.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}

### 2016-05-26 dissertation/cache-proxy.tex
small

@@
 Williams) to conjecture that this weakness is inherit to IDP planning
+(and possibly a general problem with any early-pruning planner) 
 when using join rotation. We did not attempt to examine this any

### 2016-05-26 dissertation/cache-proxy.tex
small

@@
 doing a Cartesian join and doing a remote SPARQL evaluation of data
-already in the cache.
+already in the cache, as plans for both these options would be provided.
 

### 2016-05-26 dissertation/cache-proxy.tex
grammar
note problem of SPARQL LDF link

@@
 the results of the triple pattern and enter it into the cache, or use
-a single triple pattern SPARQL query if not.
+a single triple pattern SPARQL query if not. The current
+implementation assumes that both and LDF server and a SPARQL endpoint
+can be used to answer any triple pattern. Relaxing this assumption
+depends on that the LDF specifications gain stronger semantics to link
+SPARQL endpoints and LDF servers together. With that, the assumption
+could be abandoned trivially.
 
@@
 \item Any remaining components consisting of a single triple pattern
-  results in a LDF plans.
+  result in LDF plans.
 \item Any remaining components are turned into SPARQL remote BGP

### 2016-05-26 dissertation/cache-proxy.tex
explain TPF vs LDF

@@
 response \circled{5}.
+
+Finally note that in this text, the term Triple Pattern Fragment is
+understood as a specialisation of Linked Data Fragments, in which a
+Triple Pattern Fragment server must be able to answer any triple
+pattern, whereas for a Linked Data Fragments server, this is not
+required. However, there is no difference in the current
+implementation, so in practice, the terms could be used
+interchangeably.
 

### 2016-05-26 dissertation/cache-proxy.tex
typo

@@
 contributions from me, the vast majority of the code has been written
-my Gregory Todd Williams. The main work in this project has gone into
-the module \pmodule{AtteanX::Query::Cache}, where I have authored the vast
-majority of the code. The relative difference in authorship can be
-further examined by checking the linked Github repositories, or by
+by Gregory Todd Williams. The main work in this project has gone into
+the module \pmodule{AtteanX::Query::Cache}, where I have authored the
+vast majority of the code. The relative difference in authorship can
+be further examined by checking the linked Github repositories, or by
 using git2prov\footnote{see \url{http://git2prov.org/}} to create RDF

### 2016-05-26 dissertation/cache-proxy.tex
discuss the fail

@@
 
-\subsection{Planned evaluation}
+\subsection{Planned evaluation}\label{sec:planned}
 
@@
 
+\subsection{Remedies for performance issues}\label{sec:discussfail}
+
+One may raise the issue that it was a bad idea to begin with to use a
+scripting language like Perl to write a query engine with performance
+requirements. Indeed, that may be the case. However, generally
+accepted practice is to first find the actual performance issues and
+then optimise using a bridge to first and foremost the C family of
+languages as needed. Importantly, with the width of the scope, I only
+had some months to implement the query engine. This is not a software
+engineering project, but given past industrial experience, it seems
+unlikely that I could have achieved it within the allowed time frame
+using older paradigms, such as the factory method pattern.
+
+The profiling points out a clearly where large gains could be
+made. First and foremost, parsing must be avoided, and that could be
+avoided by ensuring that IRIs where canonicalised before they were
+written to the cache, so that comparing them later would be a simple
+string comparison. Secondly, the Redis data structure store could be
+better exploited to store strings that could be used to construct  
+literals and IRI objects without parsing.
+
+However, since computing the join itself is too slow, it is not
+necessarily a solution worth pursuing. The main motivation behind
+caching hashes and arrays in Redis was to save space; since at least
+one term would be bound in any cached pattern, it shouldn't be
+necessary to insert more data into the cache. Since the performance of
+doing the join is poor, this may not be justified. Instead, it may be
+a better idea to use a conventional quad store for the cache. In that
+case, the planner would require a simple modification so that triple
+patterns are gathered into locally evaluated BGPs and remotely
+evaluated BGPs as well as LDFs. To take full advantage of this
+approach, SPARQL 1.1 \sparql{VALUES} would also need to be supported,
+so that the local quad store could perform all join operations. To
+this end, I wrote a rudimentary Virtuoso store driver for Attean, but
+at that point, the project was overdue and needed to be concluded.
+
+It seems likely that the project, with more time, could be brought to
+the point where a the evaluation in Section~\ref{sec:planned} could be
+applied, and new insight could be gained.
+
+Thus, the main contribution from this project is the remarkable ease
+with which it was implemented.
+
 \bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}
@@
 \end{document}
+
+%%  LocalWords:  canonicalised

### 2016-05-26 dissertation/cache-proxy.tex
fix ref
something about Models

@@
 the best for future requests, is sent. This analysis is discussed in
-\todo{ref}. The response \circled{2} is not sent back to the client,
-but will enter the cache. Lets say that the analyser was successful
-and that the cache can be used to answer request \circled{3}. In that
-case, the proxy will send a rewritten query for the rest of the data
-it needs to answer the query to the origin server, indicated by the
-dashed arrows. This may be a SPARQL query, a TPF request or a
-combination thereof, depending on what the planner finds least
-expensive. The proxy will then evaluate the entire query and send the
-final result back as response \circled{3}. In addition, the analyser
-may have identified another Triple Pattern Fragment that it takes as
-likely useful for future requests, and sends a request and caches the
-response \circled{4}. At some point, it may be able to answer the
-entire query using only cached results, as illustrated by request and
-response \circled{5}.
+Section~\ref{sec:analpre}. The response \circled{2} is not sent back
+to the client, but will enter the cache. Lets say that the analyser
+was successful and that the cache can be used to answer request
+\circled{3}. In that case, the proxy will send a rewritten query for
+the rest of the data it needs to answer the query to the origin
+server, indicated by the dashed arrows. This may be a SPARQL query, a
+TPF request or a combination thereof, depending on what the planner
+finds least expensive. The proxy will then evaluate the entire query
+and send the final result back as response \circled{3}. In addition,
+the analyser may have identified another Triple Pattern Fragment that
+it takes as likely useful for future requests, and sends a request and
+caches the response \circled{4}. At some point, it may be able to
+answer the entire query using only cached results, as illustrated by
+request and response \circled{5}.
 
@@
 class. Composing other roles will require further methods to be
-implemented.\todo{Explain Model}
+implemented.
+
+Stores can be rather diverse, so in addition, Attean provides another
+abstraction layer, called a Model. Models will consistently have
+quads, in the case where underlying store is a triple store, a graph
+name will be set, or it may wrap several triple stores with a graph
+name for each. In addition, Models provide higher level methods, like
+for example a \pcode{subject} method to list all subjects, and
+corresponding for other terms. Thus, upper layers of an application
+should use the Model. We have also chosen to implement significant
+parts of the query planning in the Model.
 
 When a query is parsed, algebra objects are generated. Based on the
-algebra tree, models (which may delegate the task to the store) may
-generate plans for any algebra object of their choosing by
-implementing the \pmodule{Attean::API::CostPlanner} role. This amounts
-to implementing \pcode{plans\_for\_algebra} and
+algebra tree, models may generate plans for any algebra object of
+their choosing by implementing the \pmodule{Attean::API::CostPlanner}
+role. This amounts to implementing \pcode{plans\_for\_algebra} and
 \pcode{cost\_for\_plan} methods. For a given algebra, Attean's query

### 2016-05-27 dissertation/introdefhistory.tex
remove final para
Most of Martins grammar comments

@@
 The Semantic Web can be described as a machine-readable Web of Data,
-defined by a technology stack defined by the World Wide Web
+defined by a technology stack managed by the World Wide Web
 Consortium. While individual technologies can be given precise,
@@
 players adopt a fairly consistent set of vocabularies, or by major
-players are able to force convergence towards e.g. \texttt{Schema.org}
+players that are able to force convergence towards e.g. \texttt{Schema.org}
 is an example of the social, economic and cultural mechanisms that
@@
 Scientific American 2001\cite{berners2001semantic}, but it does not attempt such an
-operational definition. Neither \cite{semwebroadmap} nor \cite{berners2000weaving} contains
+operational definition. Neither \cite{semwebroadmap} nor \cite{berners2000weaving} contain
 such definitions. However, it is clear that an essential
@@
 When the RDF working drafts were first published, I was first
-inundated by the amount of new text to read, but at the same time I
+overwhelmed by the amount of new text to read, but at the same time I
 had a growing realisation that getting information out was not solving
-the problem The Norwegian Skeptics Society needed to solve: We needed
+the problem \textit{The Norwegian Skeptics Society} needed to solve: We needed
 to push information into closed minds. My initial plan to address that
@@
 my study was something else entirely: Cosmology, and so the interest
-laid dormant for a few years as I finished my Cand. Scient.-degree. 
+lay dormant for a few years as I finished my Cand. Scient.-degree. 
 After that, I got some limited funding to write what today would be
@@
 
-Unfortunately, it wasn't allotted time to enhance and document the
+Unfortunately, management did not permit more time to enhance and document the
 SPARQL Endpoint further, and it didn't see much practical use. 
@@
 development.
-
-While the motivation and the problem statements of this work
-undoubtedly have been influenced by my intellectual investment in the
-Semantic Web, the papers should show that the investigation has not.

### 2016-05-27 dissertation/motivation.tex
Most of Martins grammar comments

@@
 transport me to campus within a certain time frame, close to
-ski tracks, abundance of uranium in the ground, number of children for
+ski tracks, low concentration of uranium in the ground, number of children for
 my children to play with, and an affordable price level. And once the
@@
 
-Of fundamental importance is that is that even though each of these
+Of fundamental importance is that even though each of these
 problems are rarely encountered, there are so many such problems that
@@
 Technically, it might be an easier problem to solve if the data could
-be centralized, and that solution has been argued by
+be centralized, and that solution has been argued for by
 \cite{DBLP:conf/semweb/BetzGHS12}. Nevertheless, even if all their
@@
 arguments against centralisation are social: If important components
-of the Semantic Web is under the control of some central authority, it
+of the Semantic Web are under the control of some central authority, it
 is likely that one cannot publish anything, link to anything, or

### 2016-05-27 dissertation/papersincontext.tex
Most of Martins grammar comments

@@
 %TODO: not the Dagstuhl?
-
+% TODO: relate to problems
 This paper is relevant to the topic of the dissertation for two
@@
 dissertation. Unfortunately, this also implies that I do not have the
-rigorous evaluation that I would preferred to have in this
+rigorous evaluation that I would have preferred to have in this
 dissertation, it is unattainable.
@@
 Williams' SPARQL implementation written in Perl is one of the SPARQL
-1.1 reference implementations, and is known to be fully compliant with
+1.1 reference implementations. It is also fully compliant with
 the specification. Through our work, and the availability of
@@
 
-I committed to do the survey after Jürgen Umbrich challenged me to see
+I committed to do this survey after Jürgen Umbrich challenged me to see
 if the standards were actually followed. Rather than just survey the
@@
 specifically benchmarking, to test assertions of performance lead to
-exploration of statistical methods in referenced in
+exploration of statistical methods referenced in
 Section~\ref{sec:condoe}, and to organise the Empirical 2014 Workshop

### 2016-05-27 dissertation/problems.tex
Most of Martins grammar comments

@@
 
-In \cite{berners2000weaving}, Tim Berners-Lee take schema
-heterogeneity for granted and that it would be resolved by using
+In \cite{berners2000weaving}, Tim Berners-Lee et al take schema
+heterogeneity for granted and postulate that it would be resolved by using
 inference, saying:
@@
 \begin{problem}\label{prob:sayanything}
-With a centralised ontology approach, how are anybody enabled to say
+With a centralised ontology approach, how can we ensure that anybody is enabled to say
 anything?
@@
 methods. Conventionally, one or more benchmarks have been used for
-this purpose, and there exists a number such
+this purpose, and there exists a number of such
 benchmarks. Unfortunately, due to this complexity, they have to test a
@@
 today in terms of data profiles, workloads, etc. Since data profiles
-and workloads are important to device a test to evaluate assertions
+and workloads are important to devise a test to evaluate assertions
 about performance, the fundamental problem arises that we currently
@@
 \begin{problem}\label{prob:lawsfuture}
-Establish laws of the Semantic Web information universe can reasonably be
+Establish laws of the Semantic Web information universe that can reasonably be
 expected to have validity far beyond the present.
@@
 
-My next concern was the implications of the
+My next concern were the implications of the
 problems~\ref{prob:zipf}~and~\ref{prob:uniform} in a federated SPARQL
@@
 arbitrary SPARQL queries are in general unsustainable, and it is
-likely that the problems in Section~\ref{sec:perfproblems} is partly
+likely that the problems in Section~\ref{sec:perfproblems} are partly
 to blame for it. Our work, and most of the related work in
@@
 are orthogonal to the other problems discussed here. Thus, it was
-natural to define them out of scope. 
+natural to consider them out of scope. 
 % TODO: However, I shall assume that no inference is needed in the

### 2016-05-27 dissertation/related.tex
Most of Martins grammar comments

@@
 Verborgh et al. \cite{ldf1} share many of the observations and goals
-that motivates this study. They introduce a hypermedia system that can
+that motivate this study. They introduce a hypermedia system that can
 be used to answer individual triple patterns (known as Triple Pattern
@@
 explores querying Linked Data, i.e. query data that is not contained
-in an a-priory defined collection of RDF data, but where a resources
+in an a-priory defined collection of RDF data, but where resources
 are traversed to compute a result. Hartig provides solid formal
@@
 
-Acosta et al. \cite{acosta2014shepherd} has ongoing work named ``SHEPHERD'' to
+Acosta et al. \cite{acosta2014shepherd} have ongoing work named ``SHEPHERD'' to
 create a SPARQL processor that takes into account observations done by
@@
 Internet and integration with HTTP-based caching.  Montoya et
-al. \cite{DBLP:journals/corr/MontoyaSMV15} proposes a query federation
+al. \cite{DBLP:journals/corr/MontoyaSMV15} propose a query federation
 system that takes advantage of client-side data replication. Both
@@
 Umbrich et al. \cite{umbrich2012hybrid} considered a situation where
-parts of the graph changes at different paces. For the slow-changing
+parts of the graph change at different paces. For the slow-changing
 parts, they prefetch an entire dataset to a local store to relieve

### 2016-05-30 dissertation/motivation.tex
quote Crosscloud

@@
 
+The Crosscloud\footnote{See \url{http://crosscloud.org/}, quote
+  retrieved on 2016-05-30.} project is motivated from the same
+observation:
+\begin{quote}
+Today, in a world of cloud-hosted software, every application is a
+kind of trap. Even if it lets you export your data in a form other
+systems can read, many of the best apps have social features. You
+can't switch because your friends or colleagues are still using the
+old site — the new site will be a ghost town. This stifles competion,
+blocks innovation, and leaves users less happy with the systems they
+are using. While some developers might want to lock-in users, we trust
+that many value user happiness and would open their systems if it was
+technologically practical. Our goal is to make it practical.
+\end{quote}
+
 The naive enthusiast may now forward the claim that Linked Data is the

### 2016-05-30 dissertation/introdefhistory.tex
s/openness/universality/

@@
 in the 1990-ties. It is characterised first and foremost by its
-openness, anyone can set up a computer, connect it to the Internet and
-start serving data or documents from it.
+universality, anyone can set up a computer, connect it to the Internet and
+start serving data or documents from it, and further adapt it to their purpose.
 

### 2016-05-30 dissertation/introdefhistory.tex
cite chalmers on objectivity

@@
 
+The problems are also discussed throughout
+\cite{chalmers1999whatis}. It is interesting to consider the example
+on page 25, where the author describes a method forwarded by Galileo
+Galilei to measure the diameter of a star. The method was objective in
+the sense if followed, it would yield the same value today as it did
+when Galileo employed it. However, it relied on faulty assumptions and
+is therefore invalid. The same kind of problem may occur in the
+evaluation of e.g. SPARQL engines. If a benchmark relies on a faulty
+workload, the application of the benchmark may seem objective to the
+investigator, but its external validity should be
+challenged. Moreover, if the benchmark relied on a use case supplied
+by e.g. an industrial partner, the validity could be further
+challenged.
+
 The impact of the Web in particular and information technologies in
@@
 own personal motivations, visions and possible biases to let the
-reader decide whether the researcher's subjective beliefs has
-influenced the validity of their contribution.
+reader decide whether the researcher's subjective beliefs or other
+external pressures has influenced the validity of their
+contribution. It is also important to be humble and admit that even
+though we as researchers treasure our objectivity, like Galileo, we
+may not be in the ideal position to judge our own objectivity. We
+should therefore detail our background to better equip the reader to
+judge.
 
@@
   \url{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
-triples, but it quickly grew to \~15~million triples.
+triples, but it quickly grew to approx. 15~million triples.
 

### 2016-05-30 dissertation/motivation.tex
punctiation

@@
 interview in Wired Magazine in 2014\footnote{see
-  \url{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}
+  \url{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}.
 This is the key motivation to still design a distributed Semantic Web,

### 2016-05-30 dissertation/related.tex
punctiation
introduction and topic division

@@
 reuse.
+
+The related work on the topic of evaluations based on contemporary
+statistics, and on the topic of philosophy is scarce, and have been
+dealt with in the papers. For the other topics dealt with in this
+thesis, a more extensive treatment is needed to place this work in
+context:
+
+\subsection{Hypermedia}
 
@@
 and further improvements have been made in more recent papers.
+
+\subsection{Query answering with cache}
+
+\todo{\cite{adali1996query}}
 
@@
 which could make this observation irrelevant, is that I considered
-only summary statistics, . If the change frequency of individual
+only summary statistics. If the change frequency of individual
 resources were wrong, in particular, if the use of the simple
@@
 
+\subsection{Tools for developer efficiency}
+
+\hskip
+
 While the approaches are diverse, these studies pull in the same
@@
 economically feasible.
-
-For more detailed related work, see the corresponding section in the
-individual papers.

### 2016-05-30 dissertation/related.tex
Update ref to Gabriella
add todos

@@
 down. In \cite{verborgh2014querying}, they demonstrated SPARQL evaluation over TPF,
-and further improvements have been made in more recent papers.
+and further improvements have been made in more recent papers.\todo{more here}
 
@@
 feasibility, soundness and completeness. The most relevant paper to
-this work is \cite{hartig2011caching}.
+this work is \cite{hartig2011caching}.\todo{more here}
 
@@
 present work, except for our emphasis on deployment on proxies in the
-Internet and integration with HTTP-based caching.  Montoya et
-al. \cite{DBLP:journals/corr/MontoyaSMV15} propose a query federation
+Internet and integration with HTTP-based caching.  \todo{reread the
+  paper more here}
+
+Montoya et al. \cite{montoya2015federated} propose a query federation
 system that takes advantage of client-side data replication. Both
 these studies claim to support the entire SPARQL query language, but
-in both cases, it is not clear that they actually do that.
+in both cases, it is not clear that they actually do that.\todo{more here}
 
@@
 remote endpoints from some processing, and thereby speed up query
-responses.
+responses. \todo{more here}
 
@@
 that controlled the changes to the dataset, and so could relieve the
-endpoint of the burden of evaluating all queries.
+endpoint of the burden of evaluating all queries.\todo{more here}
 
@@
 \subsection{Tools for developer efficiency}
+\todo{more here}
 

### 2016-05-31 dissertation/related.tex
minor

@@
 explores querying Linked Data, i.e. query data that is not contained
-in an a-priory defined collection of RDF data, but where resources
+in an \textit{a priori} defined collection of RDF data, but where resources
 are traversed to compute a result. Hartig provides solid formal
@@
 
-Acosta et al. \cite{acosta2014shepherd} have ongoing work named ``SHEPHERD'' to
+Acosta et al. has published a poster on ``SHEPHERD''
+\cite{acosta2014shepherd}, presumed to be have ongoing, to
 create a SPARQL processor that takes into account observations done by

### 2016-05-31 dissertation/cache-proxy.tex
restrict to practicals and HTTP

@@
 knowledge of the data than what is exposed through the SPARQL Endpoint
-or hypermedia metadata. In practice, there is very little
-information. If the cache is to be shared, then it would typically
-reside on the forward proxy, or in a CDN.
+or hypermedia metadata. While this would be known as a mediator in the
+database literature and has also been a field of study for 20 years, I
+focus on a practical motivation stemming mainly from what is currently
+available, and in practice, there is very little information. Also,
+this study is also practically restricted to HTTP. If the cache is to
+be shared, then it would typically reside on the forward proxy, or in
+a CDN.
 

### 2016-05-31 dissertation/related.tex
add todo
HERMES frist draft

@@
 
-\todo{\cite{adali1996query}}
+Two papers that span nearly two decades of research are of particular
+interest: The HERMES system \cite{adali1996query} considered query
+processing and cost-model based optimisation in a mediator system
+where the mediator does not have access to source statistics
+information. Recently, Papailiou et al. \cite{papailiou2015graph}
+made an extensive study of SPARQL caching, addressing some of the same
+problems as the present work.
+
+Even though these do not share our perspective, which is deployment on
+proxies in the Internet and integration with HTTP-based caching, these
+contributions provide important context.
+
+HERMES \cite{adali1996query} provided motivation for
+\cite{sec:conpush}, but is also important prior art for the work in
+Chapter~\ref{sec:tpfcacheplanning}. An important concept in the HERMES
+system is ``invariants''. They are, in our terminology, patterns that
+do not need to be looked up remotely, but can be integrated in the
+query answer by the mediator. The expectation in HERMES is that such
+invariants will be encoded by a domain expert. The Attean framework
+provides facilities to do so by using traits. Attean can also pass the
+task of computing a scalar cost metric by a similar mechanism, but
+HERMES has a more advanced cost model in that it has a cost vector
+with estimates for time required to find the first answer, time to
+find all answers and the cardinality of the answer set. Caching
+summary statistics is another important feature of HERMES, and it also
+contributes lossy and lossless summaries, for the purpose of saving
+space and reduce the time it takes to compute a cost estimate. HERMES
+further has a query rewriter and optimiser, that rewrites the query
+taking into account the cache and the invariants. While HERMES is more
+general than what I detail in Chapter~\ref{sec:tpfcacheplanning}, we
+note that while invariants could be coded in Attean, a SPARQL query
+cache should be oblivious to domain knowledge and therefore shouldn't
+assume any invariants, except in the possible but unlikely case where
+the remote server declares that an infinite freshness lifetime for the
+result of a certain graph pattern. While statistics summarisation is
+relevant and should be studied in future work, we note that with a
+Linked Data Fragments server, a cardinality estimate is required to be
+available for every triple pattern, and is also cacheable with the
+same constraints as the data themselves.
+
+\todo{import from paper}
 
@@
 \cite{acosta2014shepherd} whether the SHEPHERD system can cache results locally.
-
-Recently, Papailiou et al. \cite{papailiou2015graph} made an extensive
-study of SPARQL caching, addressing some of the same problems as the
-present work, except for our emphasis on deployment on proxies in the
-Internet and integration with HTTP-based caching.  \todo{reread the
-  paper more here}
 
@@
 economically feasible.
+
+%%  LocalWords:  summarisation

### 2016-05-31 dissertation/related.tex
canonical labelling algo

@@
 same constraints as the data themselves.
+
+Interestingly, \cite{papailiou2015graph} opens by noting that many
+optimisations used by SQL databases are ineffective in RDF databases,
+due to that RDF schema, even if they exist, are not as helpful as the
+obligatory schema in SQL databases. This provides an opportunity for
+novel research. The solution proposed in this paper is not a cache on
+a proxy or a mediator like my primary focus is. The execution engine
+in this case, has access to both the primary indexes of the RDF store
+and the cache. A key contribution of the paper is a canonical
+labelling algorithm that can create a string that will be unique for a
+Basic Graph Pattern including optional blocks. This is done by
+transforming a SPARQL query to a directed vertex-coloured graph, where
+each triple pattern is represented by a vertex, and triple patterns
+that share a common variable is linked with an edge. For each vertex,
+a label that consists of three IDs are created. Bound terms are
+translated to IDs using a dictionary, while variables are translated
+to zero. Then, the graph is directed and edges labelled according to
+the type of join they represent. The vertex and edge labels are then
+translated to non-negative integers (colours) using a sort key, to
+form an directed vertex-and-edge-coloured graph that is the authors
+claim is isomorphic with the original SPARQL. Finally, the graph is
+transformed to a simpler directed vertex-coloured graph. With this
+graph, the author's method can produce a canonical label for the query
+graph.
+
+The canonical labelling algorithm presented in this paper provides an
+opportunity for further work, when considered in an HTTP context: The
+\httph{ETag} header, defined in RFC7232 \cite{rfc7232}, can optionally
+contain an opaque identifier for a specific version of a resource
+representation, e.g. a SPARQL result. Such a canonical label could be
+used in conjunction with an algorithm that tracks changes to the RDF
+graph to produce an \httph{ETag}. For slow-changing Linked Data, which
+is a common case, it is a possibility that a canonical label in
+combination with a time-stamp of the entire RDF Graph could yield
+considerable benefits. However, the presented algorithm does not cover
+SPARQL in its entirety, for example, it does not handle named graphs,
+solution modifiers and filters, nor is it clear how it handles
+property paths (which in many cases should be trivial, since a
+property path often has an equivalent Basic Graph Pattern). Union
+queries or negation is also not mentioned, so more work is needed for
+this method to have general applicability.
 

### 2016-05-31 dissertation/related.tex
start of query planner description

@@
 graph.
+
+The query planner first examines all connected subgraphs of a full
+SPARQL query, and generates canonical labels for each one. Since it
+insists on caching connected subgraphs, it avoids the problem I
+encountered in Chapter~\ref{sec:tpfcacheplanning} that the cache may
+break up connected subgraphs and thus create Cartesian joins.
 

### 2016-06-01 dissertation/related.tex
profitable query patterns more on the use on a mediator

@@
 encountered in Chapter~\ref{sec:tpfcacheplanning} that the cache may
-break up connected subgraphs and thus create Cartesian joins.
+break up connected subgraphs and thus create Cartesian joins. To
+further enhance the query planner, the cache can optionally include
+indexing. The authors also introduce the concept of
+\textit{profitable} query patterns. It is not quite clear from the
+text, but it seems this implies that the cache can also prefetch
+results into the cache, if it finds that the result may benefit
+subsequent queries.
 
@@
 queries or negation is also not mentioned, so more work is needed for
-this method to have general applicability.
+this method to have general applicability. Nevertheless, if the entire
+Group Graph Pattern of a certain query is supported, and the query
+projects all variables, e.g. it wraps the pattern with \sparql{SELECT
+  *} or \sparql{CONSTRUCT}, the \httph{ETag} would be trivial to
+construct using the presented algorithm in a non-trivial number of
+beneficial cases. With this, cached result verification could be done
+by a mediator using HTTP requests with the \httph{If-None-Match}
+header and so make it possible to use parts of the engine of this
+paper usable in the Internet infrastructure.
 

### 2016-06-01 dissertation/cache-proxy.tex
add label
Merge the documents
add something about expiry

@@
-\documentclass[a4paper, 12pt]{article}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
-\usepackage{multirow}
-%\usepackage{times}
-\usepackage{listings}
-\usepackage{supertabular}
-\usepackage{cite}
-\usepackage{amsthm}
-\usepackage{tikz}
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
-\newcommand{\todo}[1]{\ensuremath{^{\textrm{\tiny{TODO}\normalsize}}}\footnote{\textbf{TODO:}~#1}}
-
-\include{notasjon}
-\include{helpers}
-\usepackage[pdftex,linktocpage]{hyperref}
-\hyperbaseurl{https://github.com/}
-
-\begin{document}
-\title{Prefetching SPARQL query cacher}
-\author{Kjetil Kjernsmo}
-%\institute{Department of Informatics,
-%Postboks 1080 Blindern,
-%0316 Oslo, Norway}
-
-%\email{kjekje@ifi.uio.no}
-
-\maketitle
+\chapter{Prefetching SPARQL query cacher}\label{sec:tpfcacheplanning}
 
@@
 To further understand the roles of the different modules, note the
-following: \pmodule{LWP::UserAgent::CHICaching} is a traits-based \cite{traits} implementation of the
-majority of RFC7234. \pmodule{AtteanX::Endpoint} is an implementation of the
-server side of SPARQL 1.1 Protocol.
+following: \pmodule{LWP::UserAgent::CHICaching} is a traits-based
+\cite{traits} implementation of the majority of RFC7234.
+The underlying caching framework used by this module, \pmodule{CHI},
+will take care of expiry and purging of entries, since this module
+will supply an explicit freshness lifetime. 
+\pmodule{AtteanX::Endpoint} is an implementation of the server side of
+SPARQL 1.1 Protocol.
 
@@
 
-\bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne}
-\bibliographystyle{plain}
-
-\end{document}
-
 %%  LocalWords:  canonicalised

### 2016-06-01 dissertation/related.tex
minor
Merge the documents
federation studies

@@
 HERMES \cite{adali1996query} provided motivation for
-\cite{sec:conpush}, but is also important prior art for the work in
+\ref{sec:conpush}, but is also important prior art for the work in
 Chapter~\ref{sec:tpfcacheplanning}. An important concept in the HERMES
@@
 feasibility, soundness and completeness. The most relevant paper to
-this work is \cite{hartig2011caching}.\todo{more here}
+this work is \cite{hartig2011caching}.
 
@@
 \cite{acosta2014shepherd} whether the SHEPHERD system can cache results locally.
-
-Montoya et al. \cite{montoya2015federated} propose a query federation
-system that takes advantage of client-side data replication. Both
-these studies claim to support the entire SPARQL query language, but
-in both cases, it is not clear that they actually do that.\todo{more here}
 
@@
 
+Naturally, the problem of caching in a proxy has much in common with
+the federation problem, as the cache, as seen from the query execution
+engine's perspective, could be view as yet another possible data
+source, and the number of HTTP requests to the remote endpoint can be
+wisely assumed to be kept small. Schwarte et
+al. \cite{springerlink:10.1007/978-3-642-25073-6-38} addressed these
+problems with FedX, in which they sought to group connected subgraphs
+with their concept of exclusive groups.  Montoya et
+al. \cite{montoya2015federated} propose a query federation system that
+minimise the data needed to be transferred while preserving
+completeness in the common case where the data relevant to a query is
+available on different endpoints. Görlitz and Staab \cite{splendid}
+exploited cardinality estimates sometimes available in VoID
+\cite{voidnote} descriptions to improve join order execution plans. It
+could be advantageous for the system in
+Chapter~\ref{sec:tpfcacheplanning} to exploit as well, and
+straightforward to implement, but since it has not seen wide uptake
+and the statistics is mandated by Triple Pattern Fragments, and
+therefore available in some cases to the planner, we chose not to.
+
 \subsection{Tools for developer efficiency}
 \todo{more here}
-
-\hskip
 
@@
 
-%%  LocalWords:  summarisation
+%%  LocalWords:  summarisation Schwarte

### 2016-06-01 dissertation/related.tex
martins proxy not good enough

@@
 
+The following studies do not consider how the results of one query can
+benefit the execution of a subsequent query. Nevertheless, they are
+important to consider as prior art:
 Umbrich et al. \cite{umbrich2012hybrid} considered a situation where
@@
 remote endpoints from some processing, and thereby speed up query
-responses. \todo{more here}
+responses. Of central importance to this approach is the notion of
+coherence of query patterns, defined through the ratio between results
+that were found in the live remote endpoint but not in the local cache
+and the results from the live remote endpoint. Based on this, the need
+for correct, up to date answers can be balanced against the need for
+good response times.  This approach differs radically from our
+approach in Chapter~\ref{sec:tpfcacheplanning} in that it replicates
+an entire dataset rather than just fragments.
+
+Naturally, the problem of caching in a proxy has much in common with
+the federation problem, as the cache, as seen from the query execution
+engine's perspective, could be view as yet another possible data
+source, and the number of HTTP requests to the remote endpoint can be
+wisely assumed to be kept small. Schwarte et
+al. \cite{springerlink:10.1007/978-3-642-25073-6-38} addressed these
+problems with FedX, in which they sought to group connected subgraphs
+with their concept of exclusive groups.  Montoya et
+al. \cite{montoya2015federated} propose a query federation system that
+minimise the data needed to be transferred while preserving
+completeness in the common case where the data relevant to a query is
+available on different endpoints. Görlitz and Staab \cite{splendid}
+exploited cardinality estimates sometimes available in VoID
+\cite{voidnote} descriptions to improve join order execution plans. It
+could be advantageous for the system in
+Chapter~\ref{sec:tpfcacheplanning} to exploit as well, and
+straightforward to implement, but since it has not seen wide uptake
+and the statistics is mandated by Triple Pattern Fragments, and
+therefore available in some cases to the planner, we chose not to.
 
 Martin et al. \cite{sparqlproxy} implemented a reverse caching proxy
-that controlled the changes to the dataset, and so could relieve the
-endpoint of the burden of evaluating all queries.\todo{more here}
+that controlled the changes to the dataset rather than rely on HTTP
+headers to track changes, and so could relieve the endpoint of the
+burden of evaluating all queries. Such a proxy would work poorly in
+the scenario considered in Chapter~\ref{sec:tpfcacheplanning}, as
+there is no way to mandate use of the proxy in this scenario.
 
@@
 
-Naturally, the problem of caching in a proxy has much in common with
-the federation problem, as the cache, as seen from the query execution
-engine's perspective, could be view as yet another possible data
-source, and the number of HTTP requests to the remote endpoint can be
-wisely assumed to be kept small. Schwarte et
-al. \cite{springerlink:10.1007/978-3-642-25073-6-38} addressed these
-problems with FedX, in which they sought to group connected subgraphs
-with their concept of exclusive groups.  Montoya et
-al. \cite{montoya2015federated} propose a query federation system that
-minimise the data needed to be transferred while preserving
-completeness in the common case where the data relevant to a query is
-available on different endpoints. Görlitz and Staab \cite{splendid}
-exploited cardinality estimates sometimes available in VoID
-\cite{voidnote} descriptions to improve join order execution plans. It
-could be advantageous for the system in
-Chapter~\ref{sec:tpfcacheplanning} to exploit as well, and
-straightforward to implement, but since it has not seen wide uptake
-and the statistics is mandated by Triple Pattern Fragments, and
-therefore available in some cases to the planner, we chose not to.
-
 \subsection{Tools for developer efficiency}

### 2016-06-01 dissertation/related.tex
reargue TPF

@@
 
-Verborgh et al. \cite{ldf1} share many of the observations and goals
-that motivate this study. They introduce a hypermedia system that can
-be used to answer individual triple patterns (known as Triple Pattern
-Fragments (TPF)). Their stated goal is to transfer as much as possible
-of the burden of evaluating a query from the server to the
+Hypermedia, in the context of the Web, has been strongly influenced by
+Fielding's definition of the REST architectural style, see
+\cite{Fielding_2000_Architectural-Styles}~Chapter~5, in particular, a
+constraint that he formulated, known as ``Hypermedia as the Engine of
+Application State'' (abbreviated HATEOAS). The implications for an
+application using RDF is that the application should be able to tell
+from the RDF data \emph{only} how any interactions should occur. The
+semantics expressed with RDF may enable this.
+
+Verborgh et al. \cite{ldf1} wanted a system that would be able to
+answer individual triple patterns to enable clients to answer queries,
+and to do so, they employed hypermedia, and formulated this as Triple
+Pattern Fragments. The key to this hypermedia is that an answer does
+not only include the data, but also metadata, importantly a
+cardinality estimate for the triple pattern, and also control
+information, that tells a client explicitly in every response how
+further interaction should be done, like obtaining further data.
+
+Overall, this study share many of the observations and goals that
+motivate this study. Their stated goal is to transfer as much as
+possible of the burden of evaluating a query from the server to the
 client. This can happen because a simple pattern match for a single
@@
 by a simple Web server. The results may also be paged, so that each
-individual response may be kept small. However, TPF mandates that
-every response must contain a cardinality estimate for every triple
-pattern, an operation that may be quite expensive for the server, and
+individual response may be kept small.
+
+However, it is not necessarily the case that evaluating the entire
+query on the client side is the most efficient, depending in
+particular on the perspective of the different
+actors. Chapter~\ref{sec:tpfcacheplanning} adopts Triple Pattern
+Fragments, but makes different assumptions, e.g. tries to avoid a
+large number of HTTP requests.
+
+Also, TPF mandates that every response must contain a cardinality
+estimate for every triple pattern, an operation that may be quite
+expensive for the server, and
 %TODO: ref characteristic sets?
@@
 planners may not provide it to keep the cost of planning itself
-down. In \cite{verborgh2014querying}, they demonstrated SPARQL evaluation over TPF,
-and further improvements have been made in more recent papers.\todo{more here}
+down.
+
+In \cite{verborgh2014querying}, they demonstrated SPARQL evaluation
+over TPF, and TPF has become an active area of research.
 
 \subsection{Query answering with cache}
+
+First, we note some claims that SPARQL query caching is not possible:
+In \cite{hogan2014paths}, the authors examine cacheable as one of the
+desiderata for sustainable data access. They claim, without further
+justification, that SPARQL isn't cacheable. With this work and much of
+the below related work, I suggest they are mistaken. Not only can
+whole SPARQL queries be cached, but it is possible to break down the
+query to parts that can be cached.
 
@@
 
-\todo{import from paper}
-
 Olaf Hartig has a large number of papers that are adjacent to the
@@
 
+Williams and Weaver \cite{kaseicache} consider the implications of the
+HTTP protocol on query caching in a quad store on the indexing methods
+that were employed. They propose to introduce a modification timestamp
+in index structures, and find that the query engine would be able to
+identify pieces that have been changed during execution, and could
+therefore expose a \httph{Last-Modified} header and validate an
+earlier response in accordance with RFC7232 \cite{rfc7232} cheaper
+than to evaluate the query. This also represent an alternative to HTTP
+\httph{ETag} as in the discussion of \cite{papailiou2015graph}.
+
 Dividino et al. \cite{Dividino2015} proposed strategies to keep caches

### 2016-06-01 dissertation/related.tex
link my own to ldf

@@
 
-Verborgh et al. \cite{ldf1} wanted a system that would be able to
-answer individual triple patterns to enable clients to answer queries,
-and to do so, they employed hypermedia, and formulated this as Triple
-Pattern Fragments. The key to this hypermedia is that an answer does
-not only include the data, but also metadata, importantly a
-cardinality estimate for the triple pattern, and also control
-information, that tells a client explicitly in every response how
-further interaction should be done, like obtaining further data.
+An example of this use is the work considered in
+Section~\ref{sec:conlapis}, another example is the new direction
+proposed by Verborgh et al. \cite{ldf1} who wanted a system that would
+be able to answer individual triple patterns to enable clients to
+answer queries, and to do so, they employed hypermedia, and formulated
+this as Triple Pattern Fragments. The key to this hypermedia is that
+an answer does not only include the data, but also metadata,
+importantly a cardinality estimate for the triple pattern, and also
+control information, that tells a client explicitly in every response
+how further interaction should be done, like obtaining further data.
 

### 2016-06-02 dissertation/introdefhistory.tex
Add the working ontologist book
point out beliefs

@@
 Scientific American 2001\cite{berners2001semantic}, but it does not attempt such an
-operational definition. Neither \cite{semwebroadmap} nor \cite{berners2000weaving} contain
+operational definition. Neither \cite{semwebroadmap},
+\cite{berners2000weaving} nor \cite{Allemang:2008:SWW:1386668} contain
 such definitions. However, it is clear that an essential
@@
 (e.g. ``We note that...'').
+
+Lastly, while the discussion in Section~\ref{sec:conphil} should make
+clear that I think that there are epistemological problems underneath
+the entire field of study, i.e. there is reason to question the
+validity of any result, there are invariably certain statements that
+have a particularly weak foundation, especially in the motivation. I
+shall attempt to be careful to use the term ``belief'' to qualify such
+statements.
 

### 2016-06-02 dissertation/introdefhistory.tex
add a structure overview

@@
 investigation rather than being the end of any conversations.
+
+This chapter serves to tie the accepted papers into a whole, provide
+and overarching rationale and problem statements, overview of the
+contributions and a discussion of the problematic sides of the study
+seen from a philosophy of science viewpoint. It is structured as
+follows: Section~\ref{sec:nodef} discusses the possibility of defining
+the Semantic Web, Section~\ref{sec:objectivity} discusses the problem
+of maintaining objectivity, Section~\ref{sec:history} attempts to
+satisfy the previous discussion by detailing the author's personal
+involvement in the Semantic Web. Section~\ref{sec:motivation} has an
+overview of the motivation behind this work.
+Section~\ref{sec:related} puts the present work into the context
+within the existing literature. Section~\ref{sec:problems} summarises
+important research problems and details the motivation, and
+Section~\ref{sec:papersincontext} addresses how the contributions
+address these problems and discusses weaknesses of the papers with the
+benefit of hindsight.
 
@@
 
-\section{Objectivity of the study}
+\section{Objectivity of the study}\label{sec:objectivity}
 
@@
 have a particularly weak foundation, especially in the motivation. I
-shall attempt to be careful to use the term ``belief'' to qualify such
+shall be careful to use the term ``belief'' to qualify such
 statements.

### 2016-06-02 dissertation/motivation.tex
hypermedia motivation headline for overarching

@@
 \section{Motivation}\label{sec:motivation}
+
+\subsection{Cruise altitude view}
 
@@
 
+\subsection{Hypermedia}
+
+The hypermedia idea is attractive as it is the extension of the ``View
+Source'' idea that brought me to the Web. In its original form, I
+believe that this feature is essential for broad developer
+adoption\footnote{See also
+  \url{http://archive.oreilly.com/pub/a/network/2000/04/13/CFPkeynote.html}
+  for a view aligned with this.}. With hypermedia, a developer would
+not need to read extensive documentation, but can start using the data
+to create applications that interact with the data with a very low
+barrier to entry. When designing the hypermedia discussed in
+Section~\ref{sec:conlapis}, this was my primary design motivation.
+
+It is also possible to think of hypermedia in terms of machine
+interaction. A classical SQL database is often the antithesis of
+hypermedia, as the semantics of columns is usually not clear, to
+interpret the schema, domain expertise is needed, the documentation
+may reside in documents not even accessible to machines, for various
+reasons.
+
+RDF is well suited to include information that clarifies the semantics
+with the data so it could be argued that most RDF documents would be
+hypermedia, but what is often overlooked is that the documents need to
+be explicit in linking schema from all documents. Moreover, control
+information to enable anticipated and serendipitous applications must
+be made available in the RDF. E.g. It is not sufficient for a pizza
+restaurant to publish data on pizzas, the RDF must also express how to
+order to enable a software agent to organise a party.
+
 %%  LocalWords:  hegemonies

### 2016-06-02 dissertation/motivation.tex
dev motivation

@@
 \section{Motivation}\label{sec:motivation}
+
+The following section provides an overview of the motivation behind
+this work in general and each direction in particular. Detailed
+problem statements are given in Section~\ref{sec:problems}.
 
@@
 
+\subsection{Developer friendliness}
+
+As noted, my hypermedia motivation came from the perspective as a
+developer, but hypermedia has a substantial research literature behind
+it. While software engineering certainly is a large field, the
+direction of work concerning developer friendliness comes mainly from
+personal experience, from developing solutions in the industry, but
+also from working with other developers who are at best indifferent to
+Semantic Web technologies, and in many cases antagonistic towards it. 
+As noted in Section~\ref{sec:history}, Aaron Swartz echoed the
+sentiment already a decade ago that the Semantic Web had failed
+because it did not provide tangible benefits, and had gotten a
+reputation for being unfriendly to developers. It is crucial to
+address this problem.
+
 %%  LocalWords:  hegemonies

### 2016-06-02 dissertation/motivation.tex
philosophy of science

@@
 
+\subsection{Philosophy of Science}
+
+While a failure to be developer friendly may be one reason why the
+Semantic Web has not succeeded at a grand scale, my training in
+natural science lead me to ask if there may be other reasons that we
+do not see the remarkable rate of progress of the natural sciences in
+this field. These questions are as hard to ask and to live with as to
+answer.
+
 %%  LocalWords:  hegemonies

### 2016-06-02 dissertation/motivation.tex
caching section

@@
 
-However, we note that the Internet infrastructure is accommodating for
-solving parts of this problem with the presence of caches at several
-different levels. Some clients also have significant processing
-capacity, and these are among the factors that could make SPARQL query
-answering on the open Web practically feasible. The focus of this work
-is to ease the burden of servers by allowing other parties to take
-parts of the query evaluation.
-
 \subsection{Hypermedia}
@@
 
-\subsection{Philosophy of Science}
+\subsection{Philosophy of Science and statistical methods}
 
@@
 
+I became early quite confident that the practice of benchmarking was
+an easy target, with its lack of summary statistics and inability to
+support proper hypothesis testing, standardising to deal with
+complexity rather than methodological advances, and lack of ability to
+formulate the experiment as severe enough.
+
+\subsection{Caching in the Internet}
+
+Clearly, the above motivation points out federated SPARQL as a key
+solution to the problem of query answering over decentralised sources,
+and that was originally where I intended my main contributions. During
+the work, it became clear that the stability of individual endpoints
+would need to addressed first. In the detailed problem descriptions,
+this will be further discussed, the choice of caching is the main
+focus is somewhat arbitrary.
+
+It is crucial, however, that we note that the Internet infrastructure
+is accommodating for solving parts of this problem with the presence
+of caches at several different levels. We would need to create an
+infrastructure to take advantage of these caches, and encourage data
+publishers to include metadata to allow caches to be managed in a
+standard-defined way. The scale at which this can already be done was
+at first important to understand.
+
+We further note that some clients also have significant processing
+capacity and could process SPARQL queries themselves. Others, like
+microcontrollers on the Internet of Things, cannot. This calls for a
+diverse infrastructure that can accommodate all kinds of
+clients. Nevertheless, these are among the factors that
+could make SPARQL query answering on the open Web practically
+feasible.
+
+The convergence of all the directions of this work was intended to be
+a contribution to ease the burden of servers by allowing other parties
+to take parts of the query evaluation.
+
 %%  LocalWords:  hegemonies

### 2016-06-02 dissertation/related.tex
really do the developer stuff?
ldp in related work

@@
 over TPF, and TPF has become an active area of research.
+
+Under the auspices of the  World Wide Web Consortium a standard for
+read-write applications using RDF has been developed: The Linked Data
+Platform \cite{ldp1}. The abstract in the specification describes it
+as:
+\begin{quote}
+Linked Data Platform (LDP) defines a set of rules for HTTP operations
+on web resources, some based on RDF, to provide an architecture for
+read-write Linked Data on the web.
+\end{quote}
+While developers who are familiar with HTTP-based applications and
+Linked Data may feel familiar with the platform, it still requires
+that they read and understand the specification to be able to use it,
+and therefore, it is not hypermedia, and I think it misses out on many
+of the benefits of hypermedia. A discussion at the Developers Workshop
+at the Extended Semantic Web Conference 2015\footnote{see
+  \url{https://www.youtube.com/watch?v=F4WN4XEpViA} for a recording of
+  the discussion session.} found rough consensus around a comment
+formulated by me:
+\begin{quote}
+LDP has to be superseded by hypermedia in some way.
+\end{quote}
 
@@
 
-\subsection{Tools for developer efficiency}
-\todo{more here}
+%\subsection{Tools for developer efficiency}
+%\todo{more here}
 

### 2016-06-03 dissertation/motivation.tex
Cache problems and solutions

@@
 
-\subsection{Caching in the Internet}
+\subsection{Caching in the Internet}\label{sec:motivcache}
 
@@
 focus is somewhat arbitrary.
+
+An important motivation was not problems, however, but opportunity:
+The proliferation of Content Delivery Networks is an opportunity for
+improvement, the emergence of Triple Pattern Fragments, and traits
+based planning in Section~\ref{sec:conpush} all represent important
+opportunities to seize.
 

### 2016-06-03 dissertation/problems.tex
Cache problems and solutions

@@
 
+\begin{problem}\label{prob:cachecartesian}
+Caching of results of individual triple patterns may cause subsequent
+queries to break into blocks that must be evaluated with Cartesian joins.
+\end{problem}
+
 \begin{problem}\label{prob:queryeq}
@@
 could be used in HTTP caching.
+\end{problem}
+
+\begin{problem}\label{prob:nostats}
+A caching proxy far removed from the database it caches may have
+little or no statistics for query optimisation.
 \end{problem}
@@
 ontology matching or ontology alignment, and is a very active area of
-research.
-
-% TODO, take some open problems in ontology alignment
+research.\todo{take some open problems in ontology alignment}
 
@@
 used to describe simple query interfaces, define how read-write
-operations are to be made, etc. %TODO more here
+operations are to be made, etc. \todo{more here}
 
@@
 problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}. This
-is explained in Section~\ref{sec:conpush}
+is explained in Section~\ref{sec:conpush}.
 
@@
 
-TODO: Paper yet to be done on caching.
+Finally, the intention was to bring convergence to the topics of this
+thesis around a system to cache results that could be reused in
+subsequent queries. As noted in Section~\ref{sec:motivcache}, this was
+motivated by opportunity as well as problems, but it is intended to
+address
+problems~\ref{prob:syntacticcache}~and~\ref{prob:cachecartesian}. In
+addition, problems~\ref{prob:zipf}~and~\ref{prob:uniform} are
+compounded by Problem~\ref{prob:nostats}, but alleviated somewhat by
+the presence of cardinality estimates given by Triple Pattern
+Fragments, which we exploit.
 
@@
 
-% TODO: Discuss characteristic sets?
+\todo{Discuss characteristic sets?}
 
@@
 algorithms that have a reasonable chance of returning the same digest
-if two queries are equivalent. This problem was addressed by
-\cite{papailiou2015graph} and their approach might be portable to be
-used in an HTTP context.
+if two queries are equivalent. As noted in Section~\ref{sec:relcache},
+the approach \cite{papailiou2015graph} took to canonical labelling may
+be extended to address this problem.
 

### 2016-06-03 dissertation/papersincontext.tex
add label
make todos visible

@@
-\section{The papers in context}
+\section{The papers in context}\label{sec:papersincontext}
 
@@
 
-%TODO: not the Dagstuhl?
-% TODO: relate to problems
+\todo{not the Dagstuhl?}
+\todo{relate to problems}
 This paper is relevant to the topic of the dissertation for two

### 2016-06-03 dissertation/introdefhistory.tex
add todo #7

@@
 address these problems and discusses weaknesses of the papers with the
-benefit of hindsight.
+benefit of hindsight.\todo{Big problem statement here}
 

### 2016-06-05 dissertation/introdefhistory.tex
industry viewpoint impact

@@
 by e.g. an industrial partner, the validity could be further
-challenged.
+challenged based on e.g. the restriction on viewpoint that this
+represents.
 
 The impact of the Web in particular and information technologies in
-general, on society is great, and research projects are often funded
+general on society is great, and research projects are often funded
 because of their relevance to society. With this perspective, this

### 2016-06-05 dissertation/introdefhistory.tex
proofread

@@
 
-In 1996, I started the Web pages of the Norwegian Skeptics Society
+In 1996, I started the Web pages of the \textit{Norwegian Skeptics Society}
 \texttt{skepsis.no}, and therefore got the control over an entire Web
@@
 However, all the above had happened on my spare time, as the topic of
-my study was something else entirely: Cosmology, and so the interest
-lay dormant for a few years as I finished my Cand. Scient.-degree. 
-After that, I got some limited funding to write what today would be
-called a Semantic Content Management System. The project was far too
-ambitious and largely a failure, but it got me some valuable
-experience with Semantic Web technology.
+my study was something else entirely: Cosmology. Therefore, the
+interest lay dormant for a few years as I finished my
+Cand. Scient.-degree.  After that, I got some limited funding to write
+what today would be called a Semantic Content Management System. The
+project was far too ambitious and largely a failure, but it got me
+some valuable experience with Semantic Web technology.
 
@@
 industry to pursue what was developing as my main interest, query
-answering over data on the open Web, for generic application
+answering over data on the open Web for generic application
 development.

### 2016-06-05 dissertation/motivation.tex
proofread

@@
 house was being built, I had to choose the toilet seat depending on
-the frame in the wall and even a toilet seat that would fit.
+the frame in the wall.
 
@@
 be made available in the RDF. E.g. It is not sufficient for a pizza
-restaurant to publish data on pizzas, the RDF must also express how to
-order to enable a software agent to organise a party.
+restaurant to publish data on pizzas and link to the schema, the RDF
+must also express how to order to enable a software agent to organise
+a party.
 
@@
 would need to addressed first. In the detailed problem descriptions,
-this will be further discussed, the choice of caching is the main
+this will be further discussed as the choice of caching as a point of
 focus is somewhat arbitrary.

### 2016-06-05 dissertation/problems.tex
make todos visible
TPF problem on microcontroller

@@
 distributions.
+\end{problem}
+
+Heterogeneity should also be expected in different types of servers
+and clients, in particular
+
+\begin{problem}\label{prob:microcontroller}
+A client running on a microcontroller does not have significant
+processing capability.
 \end{problem}
@@
 How can the adoption of existing vocabularies be encouraged? 
-% TODO: Find the ref by Frank on the economics
+\todo{ Find the ref by Frank on the economics}
 \end{problem}
@@
 address
-problems~\ref{prob:syntacticcache}~and~\ref{prob:cachecartesian}. In
-addition, problems~\ref{prob:zipf}~and~\ref{prob:uniform} are
-compounded by Problem~\ref{prob:nostats}, but alleviated somewhat by
-the presence of cardinality estimates given by Triple Pattern
-Fragments, which we exploit.
+problems~\ref{prob:syntacticcache}~and~\ref{prob:cachecartesian}.
+Triple Pattern Fragments shifts most of the processing from servers to
+clients, but Problem~\ref{prob:microcontrollers} makes that in some
+cases infeasible.  In addition,
+problems~\ref{prob:zipf}~and~\ref{prob:uniform} are compounded by
+Problem~\ref{prob:nostats}, but alleviated somewhat by the presence of
+cardinality estimates given by Triple Pattern Fragments, which we
+exploit.
 
@@
 natural to consider them out of scope. 
-% TODO: However, I shall assume that no inference is needed in the
-% following examples.
+\todo{However, I shall assume that no inference is needed in the
+following examples.}
 
@@
 since it is hypermedia, it can be described in the messages of the
-protocol.
+protocol. However, the solution in \cite{verborgh2014querying} does
+not take into account Problem~\ref{prob:microcontroller}, so query
+planning that shifts the burden to a cost-model driven balance between
+server, proxy and client is an interesting direction.
 

### 2016-06-05 dissertation/motivation.tex
Minor

@@
 
-It is crucial, however, that we note that the Internet infrastructure
-is accommodating for solving parts of this problem with the presence
-of caches at several different levels. We would need to create an
-infrastructure to take advantage of these caches, and encourage data
-publishers to include metadata to allow caches to be managed in a
-standard-defined way. The scale at which this can already be done was
-at first important to understand.
+We therefore proceed with the understanding that the Internet
+infrastructure is accommodating for solving parts of this problem with
+the presence of caches at several different levels. We would need to
+create an infrastructure to take advantage of these caches, and
+encourage data publishers to include metadata to allow caches to be
+managed in a standard-defined way. The scale at which this can already
+be done was at first important to understand.
 

### 2016-06-05 dissertation/problems.tex
note HTTP/2 on cost of requests
typo

@@
 Triple Pattern Fragments shifts most of the processing from servers to
-clients, but Problem~\ref{prob:microcontrollers} makes that in some
+clients, but Problem~\ref{prob:microcontroller} makes that in some
 cases infeasible.  In addition,
@@
 planning that shifts the burden to a cost-model driven balance between
-server, proxy and client is an interesting direction.
+server, proxy and client is an interesting direction. In this context,
+it should be noted that HTTP/2 \cite{rfc7540} is intended to make
+multiple requests less expensive, and it is interesting to study the
+practical impact of this as it could impact the cost of making many
+HTTP requests substantially.
 

### 2016-06-05 dissertation/related.tex
Cache problems and solutions
improve Attean on HERMES a note

@@
 
-\subsection{Query answering with cache}
+\subsection{Query answering with cache}\label{sec:relcache}
 
-First, we note some claims that SPARQL query caching is not possible:
+First, we note a claim that SPARQL query caching is not possible:
 In \cite{hogan2014paths}, the authors examine cacheable as one of the
@@
 query answer by the mediator. The expectation in HERMES is that such
-invariants will be encoded by a domain expert. The Attean framework
-provides facilities to do so by using traits. Attean can also pass the
-task of computing a scalar cost metric by a similar mechanism, but
+invariants will be encoded by a domain expert. The Attean framework,
+which is used extensively in this work and is detailed in Section~\ref{sec:attean}
+provides facilities to do this in an easy and extensible way, and also
+accommodates for the task of computing a scalar cost metric, but
 HERMES has a more advanced cost model in that it has a cost vector

### 2016-06-05 dissertation/related.tex
proofread

@@
 Naturally, the problem of caching in a proxy has much in common with
-the federation problem, as the cache, as seen from the query execution
-engine's perspective, could be view as yet another possible data
+the federation problem as the cache, as seen from the query execution
+engine's perspective, could be viewed as yet another possible data
 source, and the number of HTTP requests to the remote endpoint can be
@@
 is not defined in the cache standard, RFC7234, but in the conditional
-request standard, RFC7232. Moreover, RFC7234 contains a Section~4.2.2.
+request standard, RFC7232. Moreover, RFC7234 contains a Section~4.2.2
 titled ``Calculating Heuristic Freshness''. It is my opinion that the
@@
 Semantic Web community should be more cautious about. RFC7234's
-Section~4.2.2. also proposes a simple heuristic for calculating
+Section~4.2.2 also proposes a simple heuristic for calculating
 freshness lifetime. I found that the distribution of this heuristic
@@
 smaller. However, if understood as a sophisticated heuristic within
-the constraints of RFC7234's Section~4.2.2., the contribution is
+the constraints of RFC7234's Section~4.2.2, the contribution is
 valuable.

### 2016-06-05 dissertation/papersincontext.tex
shortcoming of survey

@@
 
+The main shortcoming of this paper was that I didn't attempt to
+validate the recorded freshness lifetime against the true change
+frequency recorded by \cite{dyldo2} and discussed by
+\cite{Dividino2015}. It would be interesting to see how common it is
+that the given freshness lifetime, both given as standard-compliant
+caching and from existing heuristics, is wrong. There are two types of
+errors: That the freshness lifetime is too short, i.e., we could have
+cached it longer, and that it is too long, i.e., we would serve stale
+data.
+
 \subsection{Philosophy of Science}\label{sec:conphil}

### 2016-06-05 dissertation/cache-proxy.tex
minor
clarify

@@
 dissertation: It should relieve the remote endpoint of some of the
-burden to evaluated the query and it should add to the robustness of the
+burden to evaluate the query and it should add to the robustness of the
 open Web SPARQL infrastructure. It uses hypermedia to answer individual
@@
 focus on a practical motivation stemming mainly from what is currently
-available, and in practice, there is very little information. Also,
-this study is also practically restricted to HTTP. If the cache is to
-be shared, then it would typically reside on the forward proxy, or in
-a CDN.
+available. In current practice, very little information is available
+to the mediator. Also, this study is also practically restricted to
+HTTP. If the cache is to be shared, then it would typically reside on
+the forward proxy, or in a CDN.
 
@@
 
-Thus, the main contribution from this project is the remarkable ease
+Thus, the main contribution from this project is the ease
 with which it was implemented.

### 2016-06-05 dissertation/cache-proxy.tex
abbrievate TPF and LDF

@@
 open Web SPARQL infrastructure. It uses hypermedia to answer individual
-triple patterns, based on Triple Pattern Fragments  \cite{ldf1}. The
+triple patterns, based on Triple Pattern Fragments (TPF) \cite{ldf1}. The
 system can only answer triple patterns, but other than that, supports
@@
 
-Finally note that in this text, the term Triple Pattern Fragment is
-understood as a specialisation of Linked Data Fragments, in which a
-Triple Pattern Fragment server must be able to answer any triple
-pattern, whereas for a Linked Data Fragments server, this is not
-required. However, there is no difference in the current
+Finally note that in this text, the term Triple Pattern Fragment (TPF)
+is understood as a specialisation of Linked Data Fragments (LDF), in
+which a Triple Pattern Fragment server must be able to answer any
+triple pattern, whereas for a Linked Data Fragments server, this is
+not required. However, there is no difference in the current
 implementation, so in practice, the terms could be used
@@
 cardinality estimates, both provided by the underlying client code as
-they are a part of the core functionality of Triple Pattern Fragments,
+they are a part of the core functionality of TPF,
 as well as \pcode{cost\_for\_plan}.
@@
 Another persistent script is subscribed to this latter channel. If the
-Model supports Linked Data Fragments, a retriever will then download
+Model supports LDF, a retriever will then download
 the results of the triple pattern and enter it into the cache, or use
@@
 endpoint in combination with a local cache. The second extends this to
-also be able to generate and use Triple Pattern Fragments plans. The
+also be able to generate and use TPF plans. The
 final is a query planner for the analyser in Section~\ref{sec:analpre}
@@
 However, since the combination of remotely executed SPARQL with a
-local cache and Triple Pattern Fragments is rather complex, the cost
+local cache and TPF is rather complex, the cost
 heuristics are fairly elaborate. Moreover, since the objective of the
@@
 selectivity of the individual triple patterns. The retriever managed
-to pull all the data from the Linked Data Fragments version of DBpedia
+to pull all the data from the LDF version of DBpedia
 within some hours.
@@
   and off.
-\item Use of a caching Linked Data Fragments client, levels on and
+\item Use of a caching LDF client, levels on and
   off.
@@
 would correspond to the null hypothesis, certain levels are somewhat
-problematic. For example, the Linked Data Fragments client in point~2
+problematic. For example, the LDF client in point~2
 might be an implementation of \cite{verborgh2014querying}, but that
@@
 where the cache is used for these two triple patterns, the last triple
-pattern is answered by a Linked Data Fragments server and the last two
+pattern is answered by a LDF server and the last two
 triple patterns are coalesced into a new SPARQL query, and evaluated

### 2016-06-05 dissertation/cache-proxy.tex
improve on graph names

@@
 dissertation: It should relieve the remote endpoint of some of the
-burden to evaluate the query and it should add to the robustness of the
-open Web SPARQL infrastructure. It uses hypermedia to answer individual
-triple patterns, based on Triple Pattern Fragments (TPF) \cite{ldf1}. The
-system can only answer triple patterns, but other than that, supports
-SPARQL 1.1 in its entirety. This was made possible to develop quickly
-by the efforts put into the query planning in the Attean framework
-described in Section~\ref{sec:conpush}. Caching based on RFC7234
-\cite{rfc7234} was shown to be of possible use in the survey I
-conducted. Finally, it was planned to be evaluated using
-Design of Experiments.
+burden to evaluate the query and it should add to the robustness of
+the open Web SPARQL infrastructure. It uses hypermedia to answer
+individual triple patterns, based on Triple Pattern Fragments (TPF)
+\cite{ldf1}. The system does not support named graphs (i.e. only
+triple stores), but other than that, supports SPARQL 1.1 in its
+entirety. This was made possible to develop quickly by the efforts put
+into the query planning in the Attean framework described in
+Section~\ref{sec:conpush}. Caching based on RFC7234 \cite{rfc7234} was
+shown to be of possible use in the survey I conducted. Finally, it was
+planned to be evaluated using Design of Experiments.
 

### 2016-06-05 dissertation/cache-proxy.tex
add todo
improve transition
proofreading

@@
 \begin{table}
-\caption{The modules that enable the query cache}\label{tab:modules}
+\caption{The modules that enable the query cache\todo{make it fit. Rotate?}}\label{tab:modules}
 \begin{tabular}{ | l | p{3cm} | l |}
@@
 \pmodule{AtteanX::Endpoint} is an implementation of the server side of
-SPARQL 1.1 Protocol.
+SPARQL 1.1 Protocol. The system is further detailed in the following sections:
 
@@
 traits and possibly the class' own methods, and then instantiated. A
-trait can also require a set of methods, and as such, is similar to
+trait can also require a set of methods or attributes, and as such, is similar to
 interfaces, but since its focus is code reuse, it can also provide a
@@
 possible plans and then estimate the cost of all plans, and finally
-return the best 5. \pmodule{Attean::API::IDPJoinPlanner}, which
+return the best 5. \pmodule{Attean::API::IDPJoinPlanner}
 implements an Iterative Dynamic Programming planner

### 2016-06-05 dissertation/cache-proxy.tex
add todo on cost model based on doe
move the todo

@@
 \begin{table}
-\caption{The modules that enable the query cache\todo{make it fit. Rotate?}}\label{tab:modules}
+\caption{The modules that enable the query cache}\label{tab:modules}
 \begin{tabular}{ | l | p{3cm} | l |}
@@
 \end{table}
+\todo{make it fit. Rotate?}
 
@@
 based on the average over the rest of the factors. Based on this, a
-common one-sided two-sample $t$-test can be performed.
+common one-sided two-sample $t$-test can be performed.\todo{Discuss
+  cost model based on DoE?}
 

### 2016-06-05 dissertation/index.tex
Fix citations; new label on PRM selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@393 43ef1190-90fd-4dd7-a769-d0121f02b80e
Add refs
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@398 43ef1190-90fd-4dd7-a769-d0121f02b80e
fixups
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@400 43ef1190-90fd-4dd7-a769-d0121f02b80e
add machinelearning bib
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@403 43ef1190-90fd-4dd7-a769-d0121f02b80e
Say intro
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@406 43ef1190-90fd-4dd7-a769-d0121f02b80e
Add new dis

@@
-\documentclass[a4paper, 12pt, draft]{scrartcl}
+\documentclass[a4paper, 12pt, draft]{report}
 \usepackage{graphicx}
-\usepackage{natbib}
 \usepackage[utf8]{inputenc}
 \usepackage{multirow}
-\usepackage{times}
+%\usepackage{times}
 \usepackage{listings}
@@
 \usepackage{cite}
+\usepackage{amsthm}
+\usepackage{tikz}
+\usepackage[obeyDraft]{todonotes}
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
+
 \include{notasjon}
 \include{helpers}
-%\usepackage[pdftex,linktocpage,ps2pdf,dvips]{hyperref}
+\usepackage[pdftex,linktocpage]{hyperref}
+\hyperbaseurl{https://github.com/}
+
+\newtheorem{problem}{Problem}[subsection]
 
 \begin{document}
-\title{Dissertation}
+\title{SPARQL on the open, decentralised Web}
 \author{Kjetil Kjernsmo}
@@
 
-\input{introsota}
+\begin{abstract}
+  This dissertation discusses a broad range of problems concerning the
+  use of the SPARQL query language on the open, public Web. It is
+  motivated from seeing decentralisation of data and infrastructure as
+  an important social goal, and SPARQL as an enabling technology to
+  solve problems using the Semantic Web. The dissertation makes
+  contributions in hypermedia, where RDF is used to create a format
+  that can tell humans and machines alike how to manipulate resources
+  on the Web; philosophy of science, where important foundational
+  problems around how to create valid knowledge and objectivity is
+  discussed; statistical methods that better satisfies the
+  requirements from philosophy of science than current practice; how
+  to improve developer efficiency with novel programming paradigms;
+  and finally how caching infrastructure in the Internet may be used
+  to make query answering across the Web more robust.
+\end{abstract}
 
-\bibliography{selectivity}
-\bibliographystyle{plainnat}
+\tableofcontents
+
+\chapter{Introduction}
+
+\input{introdefhistory}
+\input{motivation}
+\input{related}
+\input{problems}
+\input{papersincontext}
+\input{cache-proxy}
+
+\bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne,optimization}
+\bibliographystyle{plain}
 

### 2016-06-05 dissertation/introdefhistory.tex
Add a stub conclusions #9

@@
 address these problems and discusses weaknesses of the papers with the
-benefit of hindsight.\todo{Big problem statement here}
+benefit of hindsight. Key insights are finally summarised in
+Section~\ref{sec:conclusions}.\todo{Big problem statement here}
 

### 2016-06-05 dissertation/conclusions.tex
some conclusions #9

\section{Conclusions}\label{sec:conclusions}

\begin{itemize}
\item Hypermedia is one key to developing applications on the Semantic
  Web.
\item Traits eases the task of developing experimental systems and
  usage of optimisations in underlying systems.
\item While the adoption of caching and conditional request as defined
  in HTTP is not widespread, the adoption is sufficient to start using
  it in practical systems.
\item Evaluation methodologies suffer from a poor foundation in
  philosophy of science, and research into the epistemology of the
  field is important.
\item The author admits that the study is lacking in terms of
  objectivity, but notes that the problems are of a general nature.
\item Statistical Design of Experiments provides a path to a critical
  practice of evaluations, needed to improve the foundations of the
  field.
\end{itemize}

### 2016-06-21 dissertation/related.tex
Add discussion of Semantic caching #10

@@
 query to parts that can be cached.
+
+As such, this mirrors an extensive development known as ``Semantic
+Caching'' in the database community. Ahmad et al. \cite{4777801}
+surveyed the state of the art in that area. They divided caching
+approaches into \emph{page}, \emph{tuple} and \emph{semantic} caches,
+and noted that in the two former cases, if only partial results were
+available in the cache, the entire result would have to be fetched
+from the remote database. Indeed, a page or tuple cache may be used to
+cache the results of a whole SPARQL query in HTTP context, but SPARQL
+queries are also trivial to partition by e.g. triple patterns or Basic
+Graph Patterns, which provides further opportunities for caching where
+only partial results are available.\todo{Have I understood this correctly?}
 

### 2016-06-21 dissertation/related.tex
link to partitioning

@@
 Graph Patterns, which provides further opportunities for caching where
-only partial results are available.\todo{Have I understood this correctly?}
+only partial results are available.\todo{Have I understood this
+  correctly? See \url{http://ieeexplore.ieee.org/xpls/icp.jsp?arnumber=1161590#fig1}}
 

### 2016-06-21 dissertation/cache-proxy.tex
Capitalisation; thanks Helen

@@
-\chapter{Prefetching SPARQL query cacher}\label{sec:tpfcacheplanning}
+\chapter{Prefetching SPARQL Query Cacher}\label{sec:tpfcacheplanning}
 
@@
 
-\subsection{The Attean framework}\label{sec:attean}
+\subsection{The Attean Framework}\label{sec:attean}
 
@@
 
-\subsection{SPARQL protocol client}\label{sec:sparqlclient}
+\subsection{SPARQL Protocol Client}\label{sec:sparqlclient}
 
@@
 
-\subsection{Linked Data Fragment client}\label{sec:ldfclient}
+\subsection{Linked Data Fragment Client}\label{sec:ldfclient}
 
@@
 
-\subsection{The caching proxy}\label{sec:cacher}
+\subsection{The Caching Proxy}\label{sec:cacher}
 
@@
 
-\subsubsection{Analysis and prefetching}\label{sec:analpre}
+\subsubsection{Analysis and Prefetching}\label{sec:analpre}
 
@@
 
-\subsubsection{Query planner}\label{sec:implqueryplan}
+\subsubsection{Query Planner}\label{sec:implqueryplan}
 
@@
 
-\subsection{Actual evaluation}
+\subsection{Actual Evaluation}
 
@@
 
-\subsection{Planned evaluation}\label{sec:planned}
+\subsection{Planned Evaluation}\label{sec:planned}
 
@@
 
-\subsection{Join rotation and Iterative Dynamic
+\subsection{Join Rotation and Iterative Dynamic
   Programming}
@@
 
-\subsection{Remedies for performance issues}\label{sec:discussfail}
+\subsection{Remedies for Performance Issues}\label{sec:discussfail}
 

### 2016-06-21 dissertation/papersincontext.tex
Capitalisation; thanks Helen

@@
-\section{The papers in context}\label{sec:papersincontext}
+\section{The Papers in Context}\label{sec:papersincontext}
 
@@
 
-\subsection{Read-write hypermedia}\label{sec:conlapis}
+\subsection{Read-write Hypermedia}\label{sec:conlapis}
 
@@
 
-\subsection{Pushing complexity down the stack}\label{sec:conpush}
+\subsection{Pushing Complexity Down the Stack}\label{sec:conpush}
 
@@
 
-\subsection{Survey of HTTP caching}\label{sec:consanity}
+\subsection{Survey of HTTP Caching}\label{sec:consanity}
 

### 2016-06-21 dissertation/problems.tex
short section heading
Capitalisation; thanks Helen

@@
 
-\subsection{Performance and stability problems}\label{sec:perfproblems}
+\subsection{Performance and Stability Problems}\label{sec:perfproblems}
 
@@
 
-\subsection{Problems of shared understanding}\label{sec:semproblems}
+\subsection{Problems of Shared Understanding}\label{sec:semproblems}
 
@@
 
-\subsection{Problems with evaluations}\label{sec:evalproblems}
+\subsection{Problems With Evaluations}\label{sec:evalproblems}
 
@@
 
-\subsection{Development problems}\label{sec:devproblems}
+\subsection{Development Problems}\label{sec:devproblems}
 
@@
 
-\subsection{Problems addressed by this work}\label{sec:probaddress}
+\subsection{Problems Addressed by This Work}\label{sec:probaddress}
 
@@
 
-\subsection{Future work; Discussion of other problems}
+\subsection[Future Work]{Future work; Discussion of Other Problems}
 

### 2016-06-21 dissertation/related.tex
typo #10
Capitalisation; thanks Helen

@@
-\section{Related work}\label{sec:related}
+\section{Related Work}\label{sec:related}
 
@@
 
-\subsection{Query answering with cache}\label{sec:relcache}
+\subsection{Query Answering With Cache}\label{sec:relcache}
 
@@
 only partial results are available.\todo{Have I understood this
-  correctly? See \url{http://ieeexplore.ieee.org/xpls/icp.jsp?arnumber=1161590#fig1}}
+  correctly? See \url{http://ieeexplore.ieee.org/xpls/icp.jsp?arnumber=1161590\#fig1}}
 
@@
 
-HERMES \cite{adali1996query} provided motivation for
-\ref{sec:conpush}, but is also important prior art for the work in
-Chapter~\ref{sec:tpfcacheplanning}. An important concept in the HERMES
-system is ``invariants''. They are, in our terminology, patterns that
-do not need to be looked up remotely, but can be integrated in the
-query answer by the mediator. The expectation in HERMES is that such
-invariants will be encoded by a domain expert. The Attean framework,
-which is used extensively in this work and is detailed in Section~\ref{sec:attean}
-provides facilities to do this in an easy and extensible way, and also
-accommodates for the task of computing a scalar cost metric, but
-HERMES has a more advanced cost model in that it has a cost vector
-with estimates for time required to find the first answer, time to
-find all answers and the cardinality of the answer set. Caching
-summary statistics is another important feature of HERMES, and it also
-contributes lossy and lossless summaries, for the purpose of saving
-space and reduce the time it takes to compute a cost estimate. HERMES
-further has a query rewriter and optimiser, that rewrites the query
-taking into account the cache and the invariants. While HERMES is more
-general than what I detail in Chapter~\ref{sec:tpfcacheplanning}, we
-note that while invariants could be coded in Attean, a SPARQL query
-cache should be oblivious to domain knowledge and therefore shouldn't
-assume any invariants, except in the possible but unlikely case where
-the remote server declares that an infinite freshness lifetime for the
-result of a certain graph pattern. While statistics summarisation is
-relevant and should be studied in future work, we note that with a
-Linked Data Fragments server, a cardinality estimate is required to be
-available for every triple pattern, and is also cacheable with the
-same constraints as the data themselves.
+HERMES \cite{adali1996query} provided motivation for the paper in
+Section~\ref{sec:conpush}, but is also important prior art for the
+work in Chapter~\ref{sec:tpfcacheplanning}. An important concept in
+the HERMES system is ``invariants''. They are, in our terminology,
+patterns that do not need to be looked up remotely, but can be
+integrated in the query answer by the mediator. The expectation in
+HERMES is that such invariants will be encoded by a domain expert. The
+Attean framework, which is used extensively in this work and is
+detailed in Section~\ref{sec:attean} provides facilities to do this in
+an easy and extensible way, and also accommodates for the task of
+computing a scalar cost metric, but HERMES has a more advanced cost
+model in that it has a cost vector with estimates for time required to
+find the first answer, time to find all answers and the cardinality of
+the answer set. Caching summary statistics is another important
+feature of HERMES, and it also contributes lossy and lossless
+summaries, for the purpose of saving space and reduce the time it
+takes to compute a cost estimate. HERMES further has a query rewriter
+and optimiser, that rewrites the query taking into account the cache
+and the invariants. While HERMES is more general than what I detail in
+Chapter~\ref{sec:tpfcacheplanning}, we note that while invariants
+could be coded in Attean, a SPARQL query cache should be oblivious to
+domain knowledge and therefore shouldn't assume any invariants, except
+in the possible but unlikely case where the remote server declares
+that an infinite freshness lifetime for the result of a certain graph
+pattern. While statistics summarisation is relevant and should be
+studied in future work, we note that with a Linked Data Fragments
+server, a cardinality estimate is required to be available for every
+triple pattern, and is also cacheable with the same constraints as the
+data themselves.
 
@@
 Acosta et al. has published a poster on ``SHEPHERD''
-\cite{acosta2014shepherd}, presumed to be have ongoing, to
+\cite{acosta2014shepherd}, presumed to be ongoing work, to
 create a SPARQL processor that takes into account observations done by

### 2016-06-21 dissertation/index.tex
Add a stub conclusions #9
Use the uiophd class
use UK ENglish
Capitalisation; thanks Helen
Add the papers

@@
-\documentclass[a4paper, 12pt, draft]{report}
+\documentclass[a4paper, 12pt]{uiophd}
 \usepackage{graphicx}
@@
 \usepackage{multirow}
-%\usepackage{times}
 \usepackage{listings}
@@
 \usepackage{tikz}
+\usepackage{pdfpages}
 \usepackage[obeyDraft]{todonotes}
@@
 \begin{document}
-\title{SPARQL on the open, decentralised Web}
+\title{SPARQL on the Open, Decentralised Web}
 \author{Kjetil Kjernsmo}
@@
 
+\frontmatter
 \maketitle
@@
 \input{papersincontext}
+\input{conclusions}
 \input{cache-proxy}
@@
 
+\chapter{Papers}
+
+\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2012/hypermedia-rdf.pdf}
+\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2013/iswc.pdf}
+\includepdf[pages=1-last,openright]{paper14.pdf}
+\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2015/cache-survey.pdf}
+\includepdf[pages=1-last,openright]{report.pdf}
+\includepdf[pages=1-last,openright]{noise.pdf}
 \end{document}

### 2016-06-21 dissertation/motivation.tex
Capitalisation; thanks Helen
distrib -> desentral

@@
 
-\subsection{Cruise altitude view}
+\subsection{Cruise Altitude View}
 
@@
   \url{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}.
-This is the key motivation to still design a distributed Semantic Web,
+This is the key motivation to still design a decentralised Semantic Web,
 regardless of the technical difficulties that may arise.
@@
 
-\subsection{Developer friendliness}
+\subsection{Developer Friendliness}
 
@@
 
-\subsection{Philosophy of Science and statistical methods}
+\subsection{Philosophy of Science and Statistical Methods}
 

### 2016-06-21 dissertation/introdefhistory.tex
Capitalisation; thanks Helen
detail challenges to validity

@@
 
-\section{Semantic Web -- A definition?}\label{sec:nodef}
+\section{Semantic Web -- A Definition?}\label{sec:nodef}
 
@@
 
-\section{Objectivity of the study}\label{sec:objectivity}
+\section{Objectivity of the Study}\label{sec:objectivity}
 
@@
 workload, the application of the benchmark may seem objective to the
-investigator, but its external validity should be
+investigator, but its external validity should obviously be
 challenged. Moreover, if the benchmark relied on a use case supplied
@@
 challenged based on e.g. the restriction on viewpoint that this
-represents.
+represents. However, such challenges are scarcely seen in the literature.
 
@@
 
-\section{A personal history}\label{sec:history}
+\section{A Personal History}\label{sec:history}
 

### 2016-06-22 dissertation/introdefhistory.tex

@@
 challenged based on e.g. the restriction on viewpoint that this
-represents. However, such challenges are scarcely seen in the literature.
+represents.
 
@@
 though we as researchers treasure our objectivity, like Galileo, we
-may not be in the ideal position to judge our own objectivity. We
-should therefore detail our background to better equip the reader to
-judge.
+may not be in the ideal position to judge our own objectivity, as
+should be clear from the few challenges that research methods are
+getting in the literature. We should therefore detail our background
+to better equip the reader to judge.
 

### 2016-06-22 dissertation/papersincontext.tex
link to the noise essay, note the interview as included

@@
 in the role of the interviewer, we reformulated it as
-\cite{kjernsmo_noise_2015_interview}. It was performed with  Jacco van
-Ossenbruggen as the interviewer in front of the audience.
+\cite{kjernsmo_noise_2015_interview}, which was published in the
+proceedings and is included in the dissertation. It was performed with
+Jacco van Ossenbruggen as the interviewer in front of the
+audience. The essay, at
+\url{http://folk.uio.no/kjekje/2015/noise-essay.html} is however a
+more thorough account.
 
-As a philosophical essay, the contributions are hard to detail. It is
+As a philosophical text, the contributions are hard to detail. It is
 written in a provocative style, to encourage discussions of

### 2016-06-22 dissertation/index.tex
Problem counter
typo in abstract

@@
 
-\newtheorem{problem}{Problem}[subsection]
+\newtheorem{problem}{Problem}
 
@@
   problems around how to create valid knowledge and objectivity is
-  discussed; statistical methods that better satisfies the
+  discussed; statistical methods that better satisfy the
   requirements from philosophy of science than current practice; how

### 2016-06-22 dissertation/introdefhistory.tex
Some low-hanging fruit

@@
 
-This chapter serves to tie the accepted papers into a whole, provide
-and overarching rationale and problem statements, overview of the
+This chapter serves to tie the published papers into a whole, provide
+an overarching rationale and problem statements, overview of the
 contributions and a discussion of the problematic sides of the study
@@
 important research problems and details the motivation, and
-Section~\ref{sec:papersincontext} addresses how the contributions
+Section~\ref{sec:papersincontext} details how the contributions
 address these problems and discusses weaknesses of the papers with the
@@
 that achieves success, though unlikely at this point, it may be
-something in the extension of microformats\footnote{see
+something in the extension of microformats\footnote{See
   \url{http://microformats.org/}}, microdata\cite{microdata}, or
@@
 
-Therefore, rather than attempt a definition, understanding is rather
+Therefore, rather than attempt a definition, understanding is better
 promoted by following the example of \cite{berners2001semantic} and
-declare a vision.
+declaring a vision.
 

### 2016-06-22 dissertation/cache-proxy.tex
Capitilize all see in footnotes

@@
 be further examined by checking the linked Github repositories, or by
-using git2prov\footnote{see \url{http://git2prov.org/}} to create RDF
+using git2prov\footnote{See \url{http://git2prov.org/}} to create RDF
 data of the commit history. Table~\ref{tab:modules} shows the modules

### 2016-06-22 dissertation/introdefhistory.tex
Capitilize all see in footnotes

@@
 virtues that Tim Berners-Lee has attributes the Web's success
-to\footnote{see
+to\footnote{See
   \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.
@@
 execute arbitrary queries. On 2005-11-30, I published the first public
-SPARQL endpoint with actual production data\footnote{see
+SPARQL endpoint with actual production data\footnote{See
   \url{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
@@
 who was asked by the group on his opinion, and he emailed the
-following statement\footnote{see
+following statement\footnote{See
   \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
@@
 for SPARQL 1.1, however: Full text index and a feature known as
-``Limit Per Resource''\footnote{see
+``Limit Per Resource''\footnote{See
   \url{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The

### 2016-06-22 dissertation/introduction.tex
A start of a new intro

--- 2016-06-22 dissertation/introdefhistory.tex
@@
-\section{Overview}\label{sec:introoverview}
+\chapter{Introduction}
 
@@
 methods for evaluating performance, standards compliance and even
-philosophy of science. With a scope of this breadth, the investigation
-is shallow, it is hoped that some key insights will provoke further
-investigation rather than being the end of any conversations.
+philosophy of science.
 
-This chapter serves to tie the published papers into a whole, provide
-an overarching rationale and problem statements, overview of the
-contributions and a discussion of the problematic sides of the study
-seen from a philosophy of science viewpoint. It is structured as
-follows: Section~\ref{sec:nodef} discusses the possibility of defining
-the Semantic Web, Section~\ref{sec:objectivity} discusses the problem
-of maintaining objectivity, Section~\ref{sec:history} attempts to
-satisfy the previous discussion by detailing the author's personal
-involvement in the Semantic Web. Section~\ref{sec:motivation} has an
-overview of the motivation behind this work.
-Section~\ref{sec:related} puts the present work into the context
-within the existing literature. Section~\ref{sec:problems} summarises
-important research problems and details the motivation, and
-Section~\ref{sec:papersincontext} details how the contributions
-address these problems and discusses weaknesses of the papers with the
-benefit of hindsight. Key insights are finally summarised in
-Section~\ref{sec:conclusions}.\todo{Big problem statement here}
+\section{Contributions}
 
-\section{Semantic Web -- A Definition?}\label{sec:nodef}
+\subsection{Hypermedia}
 
-The Semantic Web can be described as a machine-readable Web of Data,
-defined by a technology stack managed by the World Wide Web
-Consortium. While individual technologies can be given precise,
-operational definitions, the above definition is naive and can hardly
-be used operationally to derive new knowledge. The Semantic Web is a
-complex, human artefact that must be understood not only in technical
-terms, but also in social, economic and cultural terms. First, the Web
-is not only readable, but also writeable. Whether the essential parts
-of a data integration problem will be solved by e.g. pervasive
-ontology alignment, a linked open vocabularies approach where many
-players adopt a fairly consistent set of vocabularies, or by major
-players that are able to force convergence towards e.g. \texttt{Schema.org}
-is an example of the social, economic and cultural mechanisms that
-shape the future of the Semantic Web, see
-Section~\ref{sec:semproblems} for a more elaborate
-discussion. Likewise, it may not be the W3C Semantic Web stack at all
-that achieves success, though unlikely at this point, it may be
-something in the extension of microformats\footnote{See
-  \url{http://microformats.org/}}, microdata\cite{microdata}, or
-something else entirely. For a thoughtful criticism of Linked Data
-with a proposed alternative, see \cite{darobin1}, but again, it is
-likely that social, economic and cultural factors will be as decisive
-as technical. One may say that e.g. microformats cannot help achieve
-the visions that may be stated for the Semantic Web, but that again
-emphasises visions, not operational definitions.
+\begin{enumerate}
+\item A thorough discussion of the implications of Hypermedia
+  Types\cite{hypermediatypes} for RDF.
+\item A sketch of a vocabulary for read-write RDF hypermedia.
+\end{enumerate}
 
-One such vision is the article by Berners-Lee, Hendler and Lassila in
-Scientific American 2001\cite{berners2001semantic}, but it does not attempt such an
-operational definition. Neither \cite{semwebroadmap},
-\cite{berners2000weaving} nor \cite{Allemang:2008:SWW:1386668} contain
-such definitions. However, it is clear that an essential
-characteristic of the Web, and by extension, the Semantic Web, is its
-\emph{universality} and therefore, any actor should be as free as
-possible to adapt it to their visions and needs. A pursuit of a
-definition of the term ``Semantic Web'' would require a philosophical
-treatment which would be beyond the scope of this thesis, and even
-with that, may not be a fruitful exercise from a technical
-perspective.
+\subsection{Design of Experiments}
 
-Therefore, rather than attempt a definition, understanding is better
-promoted by following the example of \cite{berners2001semantic} and
-declaring a vision.
+\begin{enumerate}
+\item Introduction of a path to critical practice of evaluations, that makes
+  use of contemporary statistical techniques, to establish a practice
+  that can be used to refute assertions on performance.
+\item A didactical experiment to help researchers understand the statistics.
+\item The novel application of a well established method in
+  statistics, rarely used in Computer Science, to SPARQL endpoint
+  evaluation.
 
-\section{Objectivity of the Study}\label{sec:objectivity}
+\end{enumerate}
 
-A vision has necessarily a strong personal element. This challenges
-the ideals of objectivity in scientific investigations. A longer
-discussion of the desirability and attainability of objectivity in
-science is beyond the scope of this thesis, but objectivity can be
-challenged from many different angles
-\cite{sep-scientific-objectivity}.
+\subsection{Development problems}
 
-The problems are also discussed throughout
-\cite{chalmers1999whatis}. It is interesting to consider the example
-on page 25, where the author describes a method forwarded by Galileo
-Galilei to measure the diameter of a star. The method was objective in
-the sense if followed, it would yield the same value today as it did
-when Galileo employed it. However, it relied on faulty assumptions and
-is therefore invalid. The same kind of problem may occur in the
-evaluation of e.g. SPARQL engines. If a benchmark relies on a faulty
-workload, the application of the benchmark may seem objective to the
-investigator, but its external validity should obviously be
-challenged. Moreover, if the benchmark relied on a use case supplied
-by e.g. an industrial partner, the validity could be further
-challenged based on e.g. the restriction on viewpoint that this
-represents.
+\begin{enumerate}
+\item A framework to enable the use of low-level optimisations in
+  databases.
+\item Simplification when implementing experimental features in
+  SPARQL.
+\item Make it possible to compose custom query planners that doesn't
+  require inheritance.
+\end{enumerate}
 
-The impact of the Web in particular and information technologies in
-general on society is great, and research projects are often funded
-because of their relevance to society. With this perspective, this
-research is hardly value-free. As individual researchers, we may
-develop a strong relationship to the systems we create, as we have a
-great intellectual investment in them. This may influence our ability
-to reject a flawed hypothesis even if we should. Even though we note
-in the discussion of the contributions in Section~\ref{sec:conphil}
-the weaknesses of Thomas Kuhn's philosophy of science, his insistence
-of understanding science as a social endeavour stands firm.
+\subsection{Survey of HTTP Caching}
 
-These challenges are inescapable, and therefore, it is important that
-the researcher acknowledges such problems and explicitly details their
-own personal motivations, visions and possible biases to let the
-reader decide whether the researcher's subjective beliefs or other
-external pressures has influenced the validity of their
-contribution. It is also important to be humble and admit that even
-though we as researchers treasure our objectivity, like Galileo, we
-may not be in the ideal position to judge our own objectivity, as
-should be clear from the few challenges that research methods are
-getting in the literature. We should therefore detail our background
-to better equip the reader to judge.
+\begin{enumerate}
+\item An understanding of actual usage of caching headers (metadata)
+  on the open Semantic Web.
+\end{enumerate}
 
-I shall attempt to do so in the following, but also note that this
-influences the writing style. There are several conventions of
-scientific writing that stands in opposition to the goal of letting
-the reader understand the possible influence of subjective
-beliefs. The frequent use of passive voice is one such convention,
-that in my opinion tries to hide subjectivity rather than alleviate
-its harmful effects. If the methodology used in the investigation is
-flawed, it should not be obscured by writing style, the methodology
-will need to be enhanced. Double blind trials in medicine are an
-example of why it is important to acknowledge the investigator's
-role. That is not to say that the use of passive voice is always
-inappropriate, when the role of the investigator cannot be
-misunderstood, it can be used just like any other language construct.
+\subsection{Philosophy of Science}
 
-Of less importance is the use of ``we'', when it is clear that the
-work has been done by a single investigator. I hold as a general
-principle that the role of the investigator should never be obscured,
-and therefore, such use is inappropriate. The use of pronouns should
-not differ substantially in scientific writing from other writing, and
-I will use them accordingly. In particular, I will use ``we'' when
-more than one person has been involved, or when the reader is included
-(e.g. ``We note that...'').
-
-Lastly, while the discussion in Section~\ref{sec:conphil} should make
-clear that I think that there are epistemological problems underneath
-the entire field of study, i.e. there is reason to question the
-validity of any result, there are invariably certain statements that
-have a particularly weak foundation, especially in the motivation. I
-shall be careful to use the term ``belief'' to qualify such
-statements.
-
-\section{A Personal History}\label{sec:history}
-
-As noted in the previous section, objectivity may be challenged by a
-researcher's intellectual investment in a certain field of study. I
-admit that I have a very large intellectual investment in the success
-of the Semantic Web. This section serves two purposes: It should give
-the reader a good understanding of this intellectual investment, and
-enable them to assess the impact of any subjective beliefs. Secondly,
-it serves to clarify \emph{my} vision, which we noted in
-Section~\ref{sec:nodef} is more important than to attempt a
-definition. The vision strongly influences the motivation detailed in
-Section~\ref{sec:motivation}.
-
-I created my first Web page in December 1994, and strongly appreciated
-the value of the fact that I could easily contribute to the Web, and
-therefore, I quickly posted anything that I had of value, for example
-a collection of my best mountaineering pictures, constrained not by
-the platform, but by my own time. Initially, I learnt purely by
-example, i.e. using the ``View Source'' menu item when I encountered
-web pages that I liked. I also had the power to mint identifiers.
-
-With identifiers, and the hypertext language HTML came the fact that I
-could link to anyone, and anyone could link to me was highly
-empowering: Discovering that people I didn't know linked to my
-material was a strong, social reward. After some time, I had tens of
-thousands of visitors every month.
-
-In 1996, I started the Web pages of the \textit{Norwegian Skeptics Society}
-\texttt{skepsis.no}, and therefore got the control over an entire Web
-server. That was another revelation, as I could actually run code, and
-from that arose the need to not just learn by ``View Source'', but
-start to read the specifications, notably HTML, CSS and
-HTTP. Appreciating the design of these specifications took a while,
-but eventually, the value of orthogonal specifications and separation
-of concerns became apparent. Also, at the same time, I realised that I
-could be standing on the shoulders of giants by using Free Software.
-
-When the RDF working drafts were first published, I was first
-overwhelmed by the amount of new text to read, but at the same time I
-had a growing realisation that getting information out was not solving
-the problem \textit{The Norwegian Skeptics Society} needed to solve: We needed
-to push information into closed minds. My initial plan to address that
-was to enable a higher degree of targeting of information, and a Web-wide
-conversation. To do that, I wanted to create a large thesaurus of
-topics of interest, and create annotations of the type ``this article
-is a rebuttal of that article'', and an index that browsers could
-query, so that when a user viewed an article, they would also get a
-critical context. At the same time, Netscape was open-sourced as
-Mozilla, and so, I was quite convinced I could contribute the code
-needed for this to be accepted.
-
-I shared these ideas on some mailing lists in August 1998, and one of
-the persons who responded was Dan Brickley, who would chair the RDF
-working group. He quickly convinced me that RDF was not something to
-be afraid of, quite the contrary, it was exactly what I needed.
-
-However, all the above had happened on my spare time, as the topic of
-my study was something else entirely: Cosmology. Therefore, the
-interest lay dormant for a few years as I finished my
-Cand. Scient.-degree.  After that, I got some limited funding to write
-what today would be called a Semantic Content Management System. The
-project was far too ambitious and largely a failure, but it got me
-some valuable experience with Semantic Web technology.
-
-I was convinced that the things that made the Web great to work with
-had to be present in the Semantic Web as well: Everybody is empowered
-to publish, find and quickly make use of Free Software, learn by
-``View Source'' (an idea which becomes even more powerful with
-semantics, as if you can understand what you are viewing simply by
-looking at the message, you can begin processing it), anyone can mint
-identifiers, anybody can reuse those identifiers in their own
-contexts, and eventually graduate to reading technical materials as
-needed. Many of these things are virtues of the Semantic Web as it is
-an extension of the Web, others have not been sufficiently cultivated
-or even appreciated. Moreover, my experience corresponds to the
-virtues that Tim Berners-Lee has attributes the Web's success
-to\footnote{See
-  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.
-
-It was not until I joined Opera Software in 2005 to work on the now
-defunct Opera Community social networking site, that I had a bit more
-time to work on Semantic Web technology. When I joined, they had
-rudimentary support for an RDF/XML-based ``Friend of a Friend'' (FOAF)
-export, which I improved before the initial release. At the time, it
-was not anticipated that everything would have a URI, so Dan Brickley
-was a proponent of the view that most things, including people, would
-be identified mainly by their properties, e.g. their email address. It
-was a certain tension between that view and the ``give everything a
-URI''-movement heralded by Tim Berners-Lee at the time, but I decided
-to place myself in between, by giving everything a URI, but at the
-same time ensure that enough properties would identify people. I also
-included outward links, so that people could integrate the database
-export we provided with their hand-maintained FOAF. I also added
-support for photo gallery metadata and linking tags as given in a
-personal SKOS-ontology to Wordnet to enable assigning clearer meaning
-to tags. This effort was acknowledged in the Linked Data Design
-Issue\cite{linkeddataissue} that formed the basis for the more recent Linking
-Open Data project.
-
-However, at this point, I started to realise that it was difficult to
-develop applications around traversing Linked Data, and that
-prompted me to consider the possibility to offer the ability to
-execute arbitrary queries. On 2005-11-30, I published the first public
-SPARQL endpoint with actual production data\footnote{See
-  \url{http://www.onlamp.com/pub/wlg/8609}}, consisting of 2695114
-triples, but it quickly grew to approx. 15~million triples.
-
-Unfortunately, management did not permit more time to enhance and document the
-SPARQL Endpoint further, and it didn't see much practical use. 
-I was, however, allowed to join several World Wide Web Consortium
-groups. First, the Web Content Label Incubator Group, which was tasked
-with discussing a replacement for the then archaic Platform for
-Internet Content Selection, and resolved to use RDF for this
-purpose, and it transitioned into the POWDER Working Group, which
-provided the specifications.
-
-More importantly, I joined the Semantic Web Education and Outreach
-(SWEO) Interest Group, where I admitted my frustration with the lack
-of practical progress with Semantic Web technologies and the relative
-lack of uptake. This sentiment was echoed by the late Aaron Swartz,
-who was asked by the group on his opinion, and he emailed the
-following statement\footnote{See
-  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
-\begin{quote}
-I'm not sure what SWEO is, but my feeling is and pretty much always
-has been that the Semantic Web people need to start putting together
-Genuinely Useful stuff that can be done Right Now.
-\end{quote}
-
-This was one of the motivations I had for starting the SWEO Community
-Projects, where a questionnaire was posted on the Web, and people from
-the Semantic Web community were challenged to come up with a project
-that would provide practical benefits in the short term. We received
-10 proposals, out of which 3 was selected for backing by the SWEO
-IG. None of them were successful, but a fourth proposal, submitted by
-Chris Bizer and Richard Cyganiak built momentum quickly and was
-therefore also selected for backing, despite some criticism by several
-IG members, myself included. The proposal was titled ``Linking Open
-Data'' and sought to take already abundant open data and model it
-using guidelines of the Linked Data Design
-Issue\cite{linkeddataissue}. LODstats\cite{auer2012lodstats} provides
-extensive statistics on the results of this project, and has of this
-writing seen nearly 10~000 data sets.
-
-At this point, I left Opera to work as a consultant on Semantic Web
-technologies. One of the projects that were successfully completed was
-called Sublima\cite{sublima}. It was driven entirely by RDF and SPARQL,
-and featured faceted navigation, navigation in a thesaurus, and full
-text search. Experiences from this project influenced requirements of
-SPARQL 1.1 \cite{sparql11query}, and I was an editor of the SPARQL 1.1 Features and
-Rationale specification\cite{sparql11new}. Parts of the property paths
-feature and aggregate queries were used extensively. Write operations
-were also implemented with SPARQL using what was then specified only
-in a member submission to the W3C\cite{seaborne2008sparql}. This was the basis for
-the update language in SPARQL 1.1 but differed substantially in
-surface syntax, partly because our work showed how certain usage patterns
-could be simplified.
-
-Two features that arose from Sublima requirements were not accepted
-for SPARQL 1.1, however: Full text index and a feature known as
-``Limit Per Resource''\footnote{See
-  \url{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
-latter feature arises from the fact that usually, the user is
-indifferent to the number of solutions to a query, and therefore
-limiting by the number of solutions is unhelpful. This is especially
-true since other features of SPARQL, such as optional clauses, are
-helpful in dealing with heterogeneous data. In the Sublima case, it
-was interesting to limit by the number of articles that were returned,
-but the number of solutions depended on the number of authors of an
-article and the concepts used to classify an article. To achieve the
-desired effect with SPARQL 1.1, one would have to write a very complex
-subquery, which is undesirable for such an important feature.
-
-Another project, based on much of the same code
-base replaced the SKOS ontology of the Sublima project with an OWL
-ontology, and also used reasoning to aid navigation in multimedia
-libraries.
-
-However, it became clear that I could not have sufficient time in the
-industry to pursue what was developing as my main interest, query
-answering over data on the open Web for generic application
-development.
+\begin{enumerate}
+\item A provocation to discuss epistemological questions.
+\end{enumerate}

### 2016-06-22 dissertation/papersincontext.tex
def not the dagstuhl position paper
A start of a new intro

@@
 
-\todo{not the Dagstuhl?}
 \todo{relate to problems}
@@
 understand when a developer does ``View Source''.
-
-Main contributions:
-\begin{enumerate}
-\item A thorough discussion of the implications of Hypermedia
-  Types\cite{hypermediatypes} for RDF.
-\item A sketch of a vocabulary for read-write RDF hypermedia.
-\end{enumerate}
 
@@
 
-Main contributions:
-\begin{enumerate}
-\item Introduction of a path to critical practice of evaluations, that makes
-  use of contemporary statistical techniques, to establish a practice
-  that can be used to refute assertions on performance.
-\item A didactical experiment to help researchers understand the statistics.
-\item The novel application of a well established method in
-  statistics, rarely used in Computer Science, to SPARQL endpoint
-  evaluation.
-
-\end{enumerate}
-
 \subsection{Pushing Complexity Down the Stack}\label{sec:conpush}
@@
 \cite{williamspushing}.
-
-Main contributions:
-\begin{enumerate}
-\item A framework to enable the use of low-level optimisations in
-  databases.
-\item Simplification when implementing experimental features in
-  SPARQL.
-\item Make it possible to compose custom query planners that doesn't
-  require inheritance.
-\end{enumerate}
 

### 2016-06-23 dissertation/introduction.tex
Some new intro
basic RDF

@@
 form\cite{semwebroadmap}.
+
+The Semantic Web is built on a number of specifications that has been
+developed under the auspices of the World Wide Web Consortium. The
+core technology is known as the Resource Description Framework
+(RDF). The framework is defined in a suite of specifications, defining
+concepts and abstract syntax, semantics, and several concrete
+syntaxes.
+
+Conceptually, an RDF statement is a triple, where the terms are a
+subject, a predicate and an object. The building blocks for these
+triples are literals, blank nodes and Internationalized Resource
+Identifiers (IRI). IRIs builds on the familiar URL: Uniform Resource
+Locator, informally known as a Web address,
+e.g. \url{https://www.w3.org/RDF/} is an address for a document about
+RDF. On the Semantic Web, the URL is first generalised to a Uniform
+Resource Identifier, URI, that can not only name a document on the
+Web, but any resource. A resource can be a physical thing, like a
+person, or an abstract thing, like the type of weather at some place
+at some time, numbers, strings, etc. Then, URIs are extended to IRIs
+to enable the use of any character in them. Literals are strings with
+a datatype (denoted by a IRI), and optionally a language tag. Finally,
+blank nodes are unnamed nodes. RDF defines which of these building
+blocks may be used for what term: The subject may be a IRI or a blank
+node, the predicate only a IRI, whereas the object may be a literal,
+blank node or IRI.
 

### 2016-06-23 dissertation/introduction.tex
Turtle example

@@
 blank node or IRI.
+
+The use of IRIs makes it possible to view RDF statements as a directed
+graph, where the subject IRI is a vertex, the predicate is an edge to
+an object vertex\footnote{In general it is more precisely described as a
+  partially labelled directed pseudograph with unique node labels}.
+
+For historical reasons, the first syntax that was standardised for RDF
+was an syntax based on the tree-structured Extensible Markup Language
+(XML). The simplest standardised syntax is known as N-Triples, where
+each statement is written on a single line, where each IRI is written
+out in full. The basic building blocks of N-Triples has been reused to
+create a compact and natural text form syntax, known as Turtle. Turtle
+has gained traction, and is well established as a common syntax. A
+simple graph expressed in Turtle could be:
+
+\begin{verbatim}
+@prefix foaf: <http://xmlns.com/foaf/0.1/>.
+@prefix dct:  <http://purl.org/dc/terms/>.
+
+<http://www.kjetil.kjernsmo.net/foaf#me> a foaf:Person ;
+    foaf:name "Kjetil Kjernsmo" ;
+    foaf:publications <http://folk.uio.no/kjekje/2013/iswc.pdf> .
+<http://folk.uio.no/kjekje/2013/iswc.pdf> a foaf:Document ;
+    dct:title "Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation" ; 
+    dct:creator <http://www.kjetil.kjernsmo.net/foaf#me>, 
+                <http://data.semanticweb.org/person/john-s-tyssedal> .
+\end{verbatim}
+
+The example opens with prefix declarations, to allow abbreviating
+IRIs, e.g. the predicate \rdfterm{foaf:name} is an abbreviation of the
+IRI \rdfterm{http://xmlns.com/foaf/0.1/name}. The IRI
+\rdfterm{http://www.kjetil.kjernsmo.net/foaf#me} is a name for this
+author. The \rdfterm{a} predicate is standards-defined abbreviation
+for \rdfterm{http://www.w3.org/1999/02/22-rdf-syntax-ns#type}, which
+is used for declaring the subject an instance of a class, in this case
+\rdfterm{foaf:Person}. Then, the \rdfterm{foaf:publications} predicate
+is used to link to an author-published version of one of the
+publications of this dissertation, and further information is then
+provided about this publication, including a link to the co-author. As
+we see, semicolons are used to separate statements that share a
+subject, commas separate statements with the same subject and
+predicate, while periods terminate a statement.
+
+Also, note that all IRIs in this example are \emph{dereferenceable},
+meaning that if an HTTP client is used, a representation of the
+resource can be retrieved across the Internet. This, along with that
+it is expressed as RDF, makes the above so-called Linked Data.
 

### 2016-06-23 dissertation/introduction.tex
a bit on SPARQL

@@
 form\cite{semwebroadmap}.
+
+\subsection{Basic technology}
 
@@
 
+Note that all the IRIs uses the HTTP schema. The Hypertext Transfer
+Protocol (HTTP) is the main application protocol used on the Internet
+for the Web, and is also the protocol used most for the Semantic
+Web. It is a request-response protocol in a client-server computing
+model, but also accommodates for intermediate proxies.
+
 Also, note that all IRIs in this example are \emph{dereferenceable},
@@
 it is expressed as RDF, makes the above so-called Linked Data.
+
+\subsection{Interacting With Data}
+
+To read and write this graph, several techniques have been
+developed. For reading, one is to simply parse the string into a
+suited data structure and use any tool available in the chosen
+programming language to examine the data structure, or simply write
+data to e.g. a Turtle string.
+
+A more sophisticated approach is to match parts of the graphs using
+\emph{triple patterns}. A triple pattern may contain variables for
+certain terms, for example:
+
+\begin{verbatim}
+?subject a ?class .
+\end{verbatim}
+This triple pattern will match the two triples 
+\begin{verbatim}
+<http://www.kjetil.kjernsmo.net/foaf#me> a foaf:Person .
+<http://folk.uio.no/kjekje/2013/iswc.pdf> a foaf:Document .
+\end{verbatim}
+from the above example. Frameworks in popular programming languages
+commonly have a method that implements this.
+
+The triple pattern is also a fundamental building block of the SPARQL
+query language, which is the language that is the main topic of this
+dissertation. SPARQL is an extensive, standardised query language with
+both read and write operations. The syntax is near Turtle, but as the
+above triple patterns, introduces variables. A set of triple patterns
+may be used to create a conjunctive query, and along with a filter
+that can be used to further constrain the query with boolean
+expressions. Together, they are known as a Basic Graph Pattern
+(BGP). More advanced parts of the language is built around BGPs.
+
+SPARQL also introduced a mechanism for naming a graph or parts of a
+graph with a IRI, an approach that was subsequently adopted by
+RDF. Today, RDF is not just a triple, but may be a quad.
 

### 2016-06-23 dissertation/introduction.tex
hypermedia

@@
 
+\subsection{Hypermedia}
+
+However, in terms of utility when programming, advanced query
+languages such as SPARQL may not be needed for many applications, and
+may represent an overly high barrier to entry.
+
+Hypermedia is the idea that all information needed to drive
+interaction in an application must be readily available in the
+request-response dialogue. For human interactions, this has been
+understood as a requirement for a long time, one has to tell the user
+how to interact with the application, if not, the user is unlikely to
+be able to solve the task. Contemporary efforts now extend this to
+machine-machine interactions.
+
+In this context, hypermedia types is a way to classify what kind of
+operations a certain media type can help the application perform.
+
 The present work sits in the confluence of several contemporary

### 2016-06-23 dissertation/introduction.tex
more on HTTP
query planning

@@
 Web. It is a request-response protocol in a client-server computing
-model, but also accommodates for intermediate proxies.
+model, but also accommodates for intermediate proxies. HTTP messages
+consist of a header, typically with metadata, and a body, which is the
+content itself. The standard defines several headers that can be used
+to control caching on either the client, the server or an intermediate
+proxy.
 
@@
 \end{verbatim}
-from the above example. Frameworks in popular programming languages
-commonly have a method that implements this.
+from the above example (prefixes omitted for brevity). Frameworks in
+popular programming languages commonly have a method that implements
+this.
 
@@
 RDF. Today, RDF is not just a triple, but may be a quad.
+
+To execute a query, a SPARQL engine would typically parse the query to
+create a tree of algebra objects (in the case where an object oriented
+programming language is used) in the process. Then, a query planner
+component would traverse the algebra tree, and for each algebra object
+or branch of algebra objects, the query planner finds various ways to
+access the data in the underlying store or execute other
+operations. In many programming frameworks, the only way to access
+data is through matching a triple pattern. If that is the case and
+the incoming query consists of just a BGP, the query planner's task
+would simply be to match all the triple patterns and join their
+results, and finally apply any constraining filters.
+
+However, even though the result does not depend on the order of
+execution, the time it takes to find the result often does. For
+example, consider the following BGP:
+\begin{verbatim}
+?subject foaf:name ?name ;
+         foaf:workplaceHomepage <http://example.com/> ;
+         foaf:made ?contribution .
+\end{verbatim}
+In most cases, if the triple pattern with the \rdfterm{foaf:name}
+predicate is evaluated first, it is likely to match many resources,
+since nearly everything may have a name, and therefore, it does not
+reduce the number of statements that needs to be searched when
+matching the next triple pattern by much. Instead, the
+\rdfterm{foaf:workplaceHomepage} may be matched first, as we may guess
+the fact that the object is given will be more restrictive, but that
+too fails if everyone in the database works for the same organisation
+and has used this predicate. It is the query planner's task to choose
+between equivalent plans to ensure that the most efficient evaluation
+is chosen. As is clear from this example, it is often insufficient to
+rely on such heuristics, and therefore, it is common to rely on a
+statistical digest in the query planner. Still, the task of query
+planning remains a very complex one, and often, insufficient
+statistics is available, so that planning is done using heuristics or
+based on assumptions that do not hold.
+
+Indeed, detailed knowledge of the data, such as a comprehensive
+statistical digest may be available close to the database, but is not
+exposed in ways that higher levels in the technology stack can make
+use of. In our case, this is especially true for a caching proxy.
 

### 2016-06-23 dissertation/introduction.tex
minor
some fixes reformulate a contribution

@@
 \chapter{Introduction}
+
+\section{Semantic Web Technology}
 
@@
 IRI \rdfterm{http://xmlns.com/foaf/0.1/name}. The IRI
-\rdfterm{http://www.kjetil.kjernsmo.net/foaf#me} is a name for this
+\rdfterm{http://www.kjetil.kjernsmo.net/foaf\#me} is a name for this
 author. The \rdfterm{a} predicate is standards-defined abbreviation
-for \rdfterm{http://www.w3.org/1999/02/22-rdf-syntax-ns#type}, which
+for \rdfterm{http://www.w3.org/1999/02/22-rdf-syntax-ns\#type}, which
 is used for declaring the subject an instance of a class, in this case
@@
 statistics is available, so that planning is done using heuristics or
-based on assumptions that do not hold.
+based on assumptions that do not hold. These are only some of the
+concerns that affect performance of the overall query engine.
 
@@
 
+\section{Contributions}\label{sec:contribsum}
+
 The present work sits in the confluence of several contemporary
@@
 philosophy of science.
-
-\section{Contributions}
 
@@
   SPARQL.
-\item Make it possible to compose custom query planners that doesn't
-  require inheritance.
+\item Novel programming techniques for custom query planners.
 \end{enumerate}

### 2016-06-23 dissertation/introdefhistory.tex
new start on overview

@@
+\chapter{A View of the Semantic Web}
+
 \section{Overview}\label{sec:introoverview}
 
-The World Wide Web, or just the Web for short, is a well-known global
-information space invented by Tim Berners-Lee in 1989, further emerged
-in the 1990-ties. It is characterised first and foremost by its
-universality, anyone can set up a computer, connect it to the Internet and
-start serving data or documents from it, and further adapt it to their purpose.
-
-The Semantic Web is an extension of the Web that has been under
-development since 1997\footnote{The first working group draft of the
-  RDF specification is dated 1997-08-01, see
-  \url{http://www.w3.org/TR/WD-rdf-syntax-971002/}.}, to extend the
-Web with languages for expressing information in a machine processable
-form\cite{semwebroadmap}.
-
-The present work sits in the confluence of several contemporary
-efforts in the Semantic Web community. The focus is on query answering
-with the SPARQL query language, with emphasis on exploiting the World
-Wide Web, but it touches upon query federation, hypermedia, empiricial
-methods for evaluating performance, standards compliance and even
+As noted in Section~\ref{sec:contribsum}, the scope of the study is
+very broad, ranging from the practical programming details to
 philosophy of science. With a scope of this breadth, the investigation
@@
 
-This chapter serves to tie the published papers into a whole, provide
-an overarching rationale and problem statements, overview of the
-contributions and a discussion of the problematic sides of the study
-seen from a philosophy of science viewpoint. It is structured as
-follows: Section~\ref{sec:nodef} discusses the possibility of defining
-the Semantic Web, Section~\ref{sec:objectivity} discusses the problem
-of maintaining objectivity, Section~\ref{sec:history} attempts to
-satisfy the previous discussion by detailing the author's personal
-involvement in the Semantic Web. Section~\ref{sec:motivation} has an
-overview of the motivation behind this work.
-Section~\ref{sec:related} puts the present work into the context
-within the existing literature. Section~\ref{sec:problems} summarises
-important research problems and details the motivation, and
+The introduction provided a very basic view of the technology, this
+chapter will further discuss some important aspects of the Semantic
+Web, as it provides guidance for the investigation detailed in the
+papers.
+
+As such, the following serves to tie the published papers into a
+whole, provide an overarching rationale and problem statements,
+overview of the contributions and a discussion of the problematic
+sides of the study seen from a philosophy of science viewpoint. It is
+structured as follows: Section~\ref{sec:nodef} discusses the
+possibility of defining the Semantic Web,
+Section~\ref{sec:objectivity} discusses the problem of maintaining
+objectivity, Section~\ref{sec:history} attempts to satisfy the
+previous discussion by detailing the author's personal involvement in
+the Semantic Web. Section~\ref{sec:motivation} has an overview of the
+motivation behind this work.  Section~\ref{sec:related} puts the
+present work into the context within the existing
+literature. Section~\ref{sec:problems} summarises important research
+problems and details the motivation, and
 Section~\ref{sec:papersincontext} details how the contributions

### 2016-06-24 dissertation/conclusions.tex
Move future work to conclusions

@@
+\section[Future Work]{Future work; Discussion of Other Problems}
+
+As noted in Section~\ref{sec:motivation} it is clear that evaluating
+arbitrary SPARQL queries are in general unsustainable, and it is
+likely that the problems in Section~\ref{sec:perfproblems} are partly
+to blame for it. Our work, and most of the related work in
+Section~\ref{sec:related} approach this work from the direction of
+moving stress from the server, but it is also clear that another way
+to address Problem~\ref{prob:sparqlcomplex} is for the SPARQL engine
+to identify queries it would be too heavy to answer already when
+parsing or planning, and also expose that in the service description
+so that clients can avoid asking them. The HTTP/2 specification
+\cite{rfc7540} has some generic facilities that may be helpful:
+
+\begin{quote}
+\texttt{ENHANCE\_YOUR\_CALM (0xb)}:  The endpoint detected that its peer is
+      exhibiting a behavior that might be generating excessive load.
+\end{quote}
+
+Problem~\ref{prob:queryeq} is best addressed by creating hashing
+algorithms that have a reasonable chance of returning the same digest
+if two queries are equivalent. As noted in Section~\ref{sec:relcache},
+the approach \cite{papailiou2015graph} took to canonical labelling may
+be extended to address this problem.
+
+I have not approached any of the problems of
+Section~\ref{sec:semproblems}, mainly because I think it is an open
+question which of these approaches will be adopted in practice, and
+whatever the solution may be, their designers should make sure they
+are orthogonal to the other problems discussed here. Thus, it was
+natural to consider them out of scope. 
+\todo{However, I shall assume that no inference is needed in the
+following examples.}
+
+Clearly, Problem~\ref{prob:lawsfuture} is a very interesting one, with
+important implications for evaluation, but also because it shifts the
+focus of our attention. In this work, the methodology used to study
+Problem~\ref{prob:sanity} could be used to gain such understanding,
+but at present, the field is too immature to attempt the formulation
+of any laws. As noted in Section~\ref{sec:history}, this study is
+partly motivated by my subjective experience of why the Web was so
+successful. Clearly, this should be studied more objectively, but I am
+not aware of any such study. Such study could also be applied to see
+if the experiences from the emergence of the Web can be applied to
+accelerating the Semantic Web, like I subjectively propose.
+
+Problem~\ref{prob:tpf} was along with
+problems~\ref{prob:sparqlcomplex},~\ref{prob:endpointunpred}~and~\ref{prob:syntacticcache}
+part of the motivation of \cite{ldf1} and \cite{verborgh2014querying},
+and this builds in part on the insight I contributed when addressing
+Problem~\ref{prob:lapis}. Their solution was to define an ontology to
+describe an interface that can answer a single triple pattern, but
+since it is hypermedia, it can be described in the messages of the
+protocol. However, the solution in \cite{verborgh2014querying} does
+not take into account Problem~\ref{prob:microcontroller}, so query
+planning that shifts the burden to a cost-model driven balance between
+server, proxy and client is an interesting direction. In this context,
+it should be noted that HTTP/2 \cite{rfc7540} is intended to make
+multiple requests less expensive, and it is interesting to study the
+practical impact of this as it could impact the cost of making many
+HTTP requests substantially.
+
+I believe that the core of Problem~\ref{prob:dynaprog} lies in the
+prevalence of statically typed languages. From an engineering
+perspective, static typing seems like a good idea, since errors should
+surface at compile time. However, it seems like it comes at a high
+cost: Programmers seem incapable of thinking beyond the constraints it
+sets. Nevertheless, the theoretical framework that seems to solve the
+problem was proposed in 1991 in the book ``The art of the metaobject
+protocol'' \cite{kiczales1991art}. Perl has a mature implementation of
+the metaobject protocol, and the topic has been discussed on several
+hackathons, and some code has been written, but the solution seems
+elusive.
+
 \section{Conclusions}\label{sec:conclusions}

### 2016-06-24 dissertation/problems.tex
Capitilize all see in footnotes
Move future work to conclusions

@@
 Frank van Harmelen has encouraged a direction that seeks to study laws
-of the Semantic Web information universe\footnote{see
+of the Semantic Web information universe\footnote{See
   e.g. \url{http://videolectures.net/iswc2011_van_harmelen_universal/}
@@
 exploit.
-
-\subsection[Future Work]{Future work; Discussion of Other Problems}
-
-As noted in Section~\ref{sec:motivation} it is clear that evaluating
-arbitrary SPARQL queries are in general unsustainable, and it is
-likely that the problems in Section~\ref{sec:perfproblems} are partly
-to blame for it. Our work, and most of the related work in
-Section~\ref{sec:related} approach this work from the direction of
-moving stress from the server, but it is also clear that another way
-to address Problem~\ref{prob:sparqlcomplex} is for the SPARQL engine
-to identify queries it would be too heavy to answer already when
-parsing or planning, and also expose that in the service description
-so that clients can avoid asking them. The HTTP/2 specification
-\cite{rfc7540} has some generic facilities that may be helpful:
-
-\begin{quote}
-\texttt{ENHANCE\_YOUR\_CALM (0xb)}:  The endpoint detected that its peer is
-      exhibiting a behavior that might be generating excessive load.
-\end{quote}
-
-\todo{Discuss characteristic sets?}
-
-Problem~\ref{prob:queryeq} is best addressed by creating hashing
-algorithms that have a reasonable chance of returning the same digest
-if two queries are equivalent. As noted in Section~\ref{sec:relcache},
-the approach \cite{papailiou2015graph} took to canonical labelling may
-be extended to address this problem.
-
-I have not approached any of the problems of
-Section~\ref{sec:semproblems}, mainly because I think it is an open
-question which of these approaches will be adopted in practice, and
-whatever the solution may be, their designers should make sure they
-are orthogonal to the other problems discussed here. Thus, it was
-natural to consider them out of scope. 
-\todo{However, I shall assume that no inference is needed in the
-following examples.}
-
-Clearly, Problem~\ref{prob:lawsfuture} is a very interesting one, with
-important implications for evaluation, but also because it shifts the
-focus of our attention. In this work, the methodology used to study
-Problem~\ref{prob:sanity} could be used to gain such understanding,
-but at present, the field is too immature to attempt the formulation
-of any laws. As noted in Section~\ref{sec:history}, this study is
-partly motivated by my subjective experience of why the Web was so
-successful. Clearly, this should be studied more objectively, but I am
-not aware of any such study. Such study could also be applied to see
-if the experiences from the emergence of the Web can be applied to
-accelerating the Semantic Web, like I subjectively propose.
-
-Problem~\ref{prob:tpf} was along with
-problems~\ref{prob:sparqlcomplex},~\ref{prob:endpointunpred}~and~\ref{prob:syntacticcache}
-part of the motivation of \cite{ldf1} and \cite{verborgh2014querying},
-and this builds in part on the insight I contributed when addressing
-Problem~\ref{prob:lapis}. Their solution was to define an ontology to
-describe an interface that can answer a single triple pattern, but
-since it is hypermedia, it can be described in the messages of the
-protocol. However, the solution in \cite{verborgh2014querying} does
-not take into account Problem~\ref{prob:microcontroller}, so query
-planning that shifts the burden to a cost-model driven balance between
-server, proxy and client is an interesting direction. In this context,
-it should be noted that HTTP/2 \cite{rfc7540} is intended to make
-multiple requests less expensive, and it is interesting to study the
-practical impact of this as it could impact the cost of making many
-HTTP requests substantially.
-
-I believe that the core of Problem~\ref{prob:dynaprog} lies in the
-prevalence of statically typed languages. From an engineering
-perspective, static typing seems like a good idea, since errors should
-surface at compile time. However, it seems like it comes at a high
-cost: Programmers seem incapable of thinking beyond the constraints it
-sets. Nevertheless, the theoretical framework that seems to solve the
-problem was proposed in 1991 in the book ``The art of the metaobject
-protocol'' \cite{kiczales1991art}. Perl has a mature implementation of
-the metaobject protocol, and the topic has been discussed on several
-hackathons, and some code has been written, but the solution seems
-elusive.

### 2016-06-24 dissertation/introduction.tex
note todo

@@
 Web with languages for expressing information in a machine processable
-form\cite{semwebroadmap}.
+form\cite{semwebroadmap}.\todo{citations in intro}
 

### 2016-06-24 dissertation/details.tex
HTTP details

\section{Preliminaries}

\subsection{Caching in the Hypertext Transfer Protocol Standard}\label{sec:httpcache}

Internet standards come in the shape of Requests for Comments (RFC),
that it is hoped will gain traction to become \textit{de facto}
standards. Two such RFCs regulate caching in the Internet when HTTP is
used: RFC7234 ``Caching'' \cite{rfc7234} and RFC7232 ``Conditional
Requests'' \cite{rfc7232}. In \cite{kjernsmo_survey_2015}, I provided
some details on these standards in Section~1.1, that is worth
reiterating to understand the problems and contributions of this
work.

RFC7234 defines how clients and proxies may reuse a prior response
without contacting the origin server at all. The server can do this by
supplying a freshness lifetime. It also defines when clients and/or
proxies may \emph{not} reuse a prior response, and what can be done
with stale responses. It also contains a Section~4.2.2 titled
``Calculating Heuristic Freshness'', which sets some loose constraints
for when a prior response may be used even when the server has not
supplied a freshness lifetime. These constraints include that
heuristics may not be used if there is an explicit freshness lifetime,
or the server prohibits caching. The section also proposes a simple
heuristic for calculating freshness lifetime.

RFC7232 defines a protocol for asking the origin server if the cached
response is still fresh. Two response headers are relevant to achieve
this: \httph{ETag} and \httph{Last-Modified}.  \httph{ETag} may
contain an opaque identifier for a specific version of a resource
representation, for example a query result. A client or proxy may then
ask the origin server if the resource is still fresh, by sending a
request with a \httph{If-None-Match} header. If the server is able to
validate the resource representation, it will respond with a 304
status code, otherwise it will return an updated representation. It
may be of use if servers are able to generate and validate such opaque
identifiers cheaply.

The \httph{Last-Modified} header should contain the time of last
modification of the resource representation. A similar protocol is
used to validate it.

### 2016-06-24 dissertation/related.tex
Capitilize all see in footnotes
HTTP details

@@
 of the benefits of hypermedia. A discussion at the Developers Workshop
-at the Extended Semantic Web Conference 2015\footnote{see
+at the Extended Semantic Web Conference 2015\footnote{See
   \url{https://www.youtube.com/watch?v=F4WN4XEpViA} for a recording of
@@
 opportunity for further work, when considered in an HTTP context: The
-\httph{ETag} header, defined in RFC7232 \cite{rfc7232}, can optionally
-contain an opaque identifier for a specific version of a resource
-representation, e.g. a SPARQL result. Such a canonical label could be
+\httph{ETag} header, discussed in Section~\ref{sec:httpcache}, can be
+used for a SPARQL result. Such a canonical label could be
 used in conjunction with an algorithm that tracks changes to the RDF

### 2016-06-25 dissertation/related.tex
improve description of Dividino shorten and ease the critisism

@@
 of Linked Open Data up to date, in face of changing data. The authors
-sets out to learn the change frequency of resources, and based on that
-schedule updates of the resources. This does then not depend on the
-resource metadata (neither HTTP headers nor metadata in the RDF
-itself, which I studied in \cite{kjernsmo_survey_2015}), which is
-motivated by their finding that only 8\% of \httph{Last-Modified}
-headers they surveyed reflected the true change frequency. While I
-have no reason to doubt this finding, and the monitoring approach is
-interesting, I think the dismissal of resource metadata is
-insufficiently motivated, and therefore the evaluation is
-inadequate. First, the \httph{Last-Modified} header is far from the
-only cache-relevant header, as should be evident by the fact that it
-is not defined in the cache standard, RFC7234, but in the conditional
-request standard, RFC7232. Moreover, RFC7234 contains a Section~4.2.2
-titled ``Calculating Heuristic Freshness''. It is my opinion that the
-paper should be framed within the loose constraints of this section,
-where it could make a substantial contribution. As it stands, it would
-serve to violate Web standards, which in my opinion, is something the
-Semantic Web community should be more cautious about. RFC7234's
-Section~4.2.2 also proposes a simple heuristic for calculating
-freshness lifetime. I found that the distribution of this heuristic
-freshness lifetime agrees well with The Dynamic Linked Data
-Observatory (DyLDO) \cite{dyldo2}. However, a weakness of my study
-which could make this observation irrelevant, is that I considered
-only summary statistics. If the change frequency of individual
-resources were wrong, in particular, if the use of the simple
-heuristic lead to using stale content, it would be inappropriate to
-use the simple heuristic.  RFC7234's explicitly prohibits the use of a
-heuristic if an explicit expiration time is present, yet the paper
-does not attempt to include those in their assessment. Furthermore, I
-found in \cite{kjernsmo_survey_2015} that cache prohibitions were
-common. While they in many cases may be misguided, they nevertheless
-express a strong statement from the server administrators, that should
-be respected. RDF data, e.g. \rdfterm{dct:valid} and
-\rdfterm{dct:modified} could also be used to compute a expiration
-time. Finally, an opaque validator, \httph{ETag} is also a part of
-RFC7232. After considering cache prohibitions, explicit expiration
-times, opaque validators, simpler heuristics, etc., the space where
-learning the change frequency could be applied is likely much
-smaller. However, if understood as a sophisticated heuristic within
-the constraints of RFC7234's Section~4.2.2, the contribution is
-valuable.
+set out to learn the change frequency of resources representations,
+and based on that schedule updates of them as efficiently as possible
+under bandwidth constraints. Instead of relying on resource metadata,
+the authors propose an algorithm that iteratively ensures the quality
+of the cache by examining the resource representations that have been
+downloaded and compute certain metrics, such a dynamicity and
+changerate.
+
+While the approach is interesting and useful, I nevertheless have some
+points of critisism: They motivate their approach from an earlier
+finding that only 8\% of \httph{Last-Modified} headers they surveyed
+reflected the true change frequency. While I have no reason to doubt
+this finding, the authors boldly go on to claim that `` The only
+alternative [to \httph{Last-Modified}] is to actually retrieve the
+data from the sources and check it for changes.'', but this
+contradicts the results I found in \cite{kjernsmo_survey_2015}. The
+authors have not considered all alternatives, including not the
+alternative presented by the HTTP caching standard RFC7234 discussed
+in Section~\ref{sec:httpcache}, which invalidates that
+assertion. Moreover, the author's approach would violate RFC7234 in
+the case where their algorithm caches results that should not be
+cached or has an explicit freshness lifetime, as discussed in
+Section~\ref{sec:httpcache}. As a general observation, it is my belief
+that the Semantic Web community should be more cautious about
+violating core Web standards.
 

### 2016-07-22 dissertation/index.tex
note draft
A start of a new intro
add mainmatter marker
More restructuring
Helen's corrections to intro

@@
   on the Web; philosophy of science, where important foundational
-  problems around how to create valid knowledge and objectivity is
-  discussed; statistical methods that better satisfy the
+  problems around how to create valid knowledge and objectivity are
+  discussed; statistical methods that better satisfies the
   requirements from philosophy of science than current practice; how
@@
   to make query answering across the Web more robust.
+
+  \textbf{This is an unfinished draft of the thesis.}
 \end{abstract}
@@
 \tableofcontents
+\mainmatter
 
-\chapter{Introduction}
+\input{introduction}
 
@@
 \input{motivation}
+
+\chapter{Technology Details}
+\input{details}
 \input{related}
+
+\chapter{Problems and Solutions}
+
 \input{problems}

### 2016-07-22 dissertation/introdefhistory.tex
Helen's corrections to intro

@@
 likely that social, economic and cultural factors will be as decisive
-as technical. One may say that e.g. microformats cannot help achieve
+as technical. One may say that, for example microformats cannot help achieve
 the visions that may be stated for the Semantic Web, but that again
@@
 challenged. Moreover, if the benchmark relied on a use case supplied
-by e.g. an industrial partner, the validity could be further
+by for example an industrial partner, the validity could be further
 challenged based on e.g. the restriction on viewpoint that this
@@
 looking at the message, you can begin processing it), anyone can mint
-identifiers, anybody can reuse those identifiers in their own
+identifiers and anybody can reuse those identifiers in their own
 contexts, and eventually graduate to reading technical materials as
@@
 an extension of the Web, others have not been sufficiently cultivated
-or even appreciated. Moreover, my experience corresponds to the
-virtues that Tim Berners-Lee has attributes the Web's success
-to\footnote{See
+or even appreciated. Moreover, my experience corresponds with the
+virtues that Tim Berners-Lee has attributed the Web's success\footnote{See
   \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.
@@
 I was, however, allowed to join several World Wide Web Consortium
-groups. First, the Web Content Label Incubator Group, which was tasked
-with discussing a replacement for the then archaic Platform for
+groups. Firstly, the Web Content Label Incubator Group, which was
+given the task of discussing a replacement for the then archaic Platform for
 Internet Content Selection, and resolved to use RDF for this
-purpose, and it transitioned into the POWDER Working Group, which
+purpose. It transitioned into the POWDER Working Group, which
 provided the specifications.
 
-More importantly, I joined the Semantic Web Education and Outreach
+Secondly and more importantly, I joined the Semantic Web Education and Outreach
 (SWEO) Interest Group, where I admitted my frustration with the lack
@@
 lack of uptake. This sentiment was echoed by the late Aaron Swartz,
-who was asked by the group on his opinion, and he emailed the
-following statement\footnote{See
-  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
+who when asked by the group on his opinion, emailed the
+following statement:
 \begin{quote}
@@
 has been that the Semantic Web people need to start putting together
-Genuinely Useful stuff that can be done Right Now.
+Genuinely Useful stuff that can be done Right Now.\footnote{See
+  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}
 \end{quote}
 
-This was one of the motivations I had for starting the SWEO Community
+This was one of my motivations for starting the SWEO Community
 Projects, where a questionnaire was posted on the Web, and people from
@@
 that would provide practical benefits in the short term. We received
-10 proposals, out of which 3 was selected for backing by the SWEO
+10 proposals, out of which 3 were selected for backing by the SWEO
 IG. None of them were successful, but a fourth proposal, submitted by
@@
 Issue\cite{linkeddataissue}. LODstats\cite{auer2012lodstats} provides
-extensive statistics on the results of this project, and has of this
-writing seen nearly 10~000 data sets.
+extensive statistics on the results of this project, and has by the
+time of this writing seen nearly 10~000 data sets.
 
 At this point, I left Opera to work as a consultant on Semantic Web
-technologies. One of the projects that were successfully completed was
+technologies. One of the projects that were successfully completed were
 called Sublima\cite{sublima}. It was driven entirely by RDF and SPARQL,
@@
 Two features that arose from Sublima requirements were not accepted
-for SPARQL 1.1, however: Full text index and a feature known as
+for SPARQL 1.1: Full text index and a feature known as
 ``Limit Per Resource''\footnote{See
@@
 
-However, it became clear that I could not have sufficient time in the
+However, it became clear that I would not have sufficient time in the
 industry to pursue what was developing as my main interest, query

### 2016-07-22 dissertation/introduction.tex
intro prelim
Helen's corrections to intro

@@
 
-\section{Semantic Web Technology}
+\section{Semantic Web Technology}\label{sec:semwebtech}
 
@@
 information space invented by Tim Berners-Lee in 1989, further emerged
-in the 1990-ties. It is characterised first and foremost by its
+in the 1990s. It is characterised first and foremost by its
 universality, anyone can set up a computer, connect it to the Internet and

### 2016-07-22 dissertation/motivation.tex
Capitilize all see in footnotes
Helens corrections

@@
 travel based on what I want to experience rather than make me choose a
-destination based on lacking information.
+destination based on limited information.
 
@@
 hegemonies. Tim Berners-Lee discussed this in \cite{berners2000weaving} and an
-interview in Wired Magazine in 2014\footnote{see
+interview in Wired Magazine in 2014\footnote{See
   \url{http://www.wired.co.uk/news/archive/2014-02/06/tim-berners-lee-reclaim-the-web}}.

### 2016-07-22 dissertation/motivation.tex
ref for connollys bane
note the relation to researc

@@
  The bane of my existence is doing things I know the computer could do
- for me.
+ for me.\footnote{See \url{http://www.nature.com/nature/webmatters/xml/xml.html}}
 \end{quote}
@@
 interaction. A classical SQL database is often the antithesis of
-hypermedia, as the semantics of columns is usually not clear, to
+hypermedia, as the semantics of columns are usually not clear. To
 interpret the schema, domain expertise is needed, the documentation
@@
 information to enable anticipated and serendipitous applications must
-be made available in the RDF. E.g. It is not sufficient for a pizza
+be made available in the RDF. For example, It is not sufficient for a pizza
 restaurant to publish data on pizzas and link to the schema, the RDF
@@
 As noted, my hypermedia motivation came from the perspective as a
-developer, but hypermedia has a substantial research literature behind
-it. While software engineering certainly is a large field, the
+developer, not from the research literature. Nevertheless, hypermedia
+has a substantial research literature behind it. 
+While software engineering certainly is a large field, the 
 direction of work concerning developer friendliness comes mainly from

### 2016-07-22 dissertation/motivation.tex
more Helens corrections

@@
 While a failure to be developer friendly may be one reason why the
-Semantic Web has not succeeded at a grand scale, my training in
+Semantic Web has not succeeded on a grand scale, my training in
 natural science lead me to ask if there may be other reasons that we
@@
 support proper hypothesis testing, standardising to deal with
-complexity rather than methodological advances, and lack of ability to
+complexity rather than methodological advances, and lacking in ability to
 formulate the experiment as severe enough.
@@
 the work, it became clear that the stability of individual endpoints
-would need to addressed first. In the detailed problem descriptions,
-this will be further discussed as the choice of caching as a point of
+would need to addressed first. This will be further discussed in the
+detailed problem descriptions, as the choice of caching as a point of
 focus is somewhat arbitrary.
@@
 An important motivation was not problems, however, but opportunity:
-The proliferation of Content Delivery Networks is an opportunity for
+The proliferation of Content Delivery Networks was an opportunity for
 improvement, the emergence of Triple Pattern Fragments, and traits
-based planning in Section~\ref{sec:conpush} all represent important
+based planning in Section~\ref{sec:conpush} all represented important
 opportunities to seize.

### 2016-07-22 dissertation/related.tex
Helen's corrections

@@
 There is a substantial amount of relevant literature and ongoing
-research that is well aligned with the present work, either because
-the authors share parts of the motivation, or more serendipitous
-reuse.
+research that are well aligned with the present work, either because
+the authors share parts of the motivation, or this work finds serendipitous
+reuse of prior work.
 
 The related work on the topic of evaluations based on contemporary
-statistics, and on the topic of philosophy is scarce, and have been
+statistics, and on the topic of philosophy is scarce, and has been
 dealt with in the papers. For the other topics dealt with in this
@@
 be able to answer individual triple patterns to enable clients to
-answer queries, and to do so, they employed hypermedia, and formulated
+answer queries, and in doing so, they employed hypermedia, and formulated
 this as Triple Pattern Fragments. The key to this hypermedia is that
@@
 control information, that tells a client explicitly in every response
-how further interaction should be done, like obtaining further data.
+how further interaction should be done, like in obtaining further data.
 
@@
 actors. Chapter~\ref{sec:tpfcacheplanning} adopts Triple Pattern
-Fragments, but makes different assumptions, e.g. tries to avoid a
+Fragments, but makes different assumptions, for example by trying to avoid a
 large number of HTTP requests.
@@
 estimate for every triple pattern, an operation that may be quite
-expensive for the server, and
+expensive for the server.
 %TODO: ref characteristic sets?
-indeed, this information may be so expensive to provide that SPARQL
-planners may not provide it to keep the cost of planning itself
+Indeed, this information may be so expensive to provide that SPARQL
+planners may not do so, in order to keep the cost of planning itself
 down.
@@
 HERMES \cite{adali1996query} provided motivation for the paper in
-Section~\ref{sec:conpush}, but is also important prior art for the
+Section~\ref{sec:conpush}, but is also important for the
 work in Chapter~\ref{sec:tpfcacheplanning}. An important concept in
@@
 novel research. The solution proposed in this paper is not a cache on
-a proxy or a mediator like my primary focus is. The execution engine
+a proxy or a mediator like that which is my primary focus. The execution engine
 in this case, has access to both the primary indexes of the RDF store
@@
 translated to non-negative integers (colours) using a sort key, to
-form an directed vertex-and-edge-coloured graph that is the authors
+form a directed vertex-and-edge-coloured graph that the authors
 claim is isomorphic with the original SPARQL. Finally, the graph is
@@
 the SPARQLES survey \cite{buil2013sparql}, to decompose a SPARQL query into
-subqueries that is indicated to be easier for the server to evaluate,
+subqueries that are indicated to be easier for the server to evaluate,
 and so is well aligned with our goal to ease the load of the
@@
 straightforward to implement, but since it has not seen wide uptake
-and the statistics is mandated by Triple Pattern Fragments, and
+and the statistics are mandated by Triple Pattern Fragments, and
 therefore available in some cases to the planner, we chose not to.

### 2016-07-22 dissertation/problems.tex
Helen's corrections

@@
 
-He further goes on to explain how linking such terms will help make
+He goes on to explain how linking such terms will help make
 that inference, and how partial understanding is common in everyday
@@
 
-To understand whether a result from a study on e.g. performance is
+To understand whether a result from a study on for example, performance is
 valid, an empirical study must be conducted, as a SPARQL system is
@@
 can support realistic experiments even though a future Semantic Web
-may be several orders of magnitude larger than today's.
+may be several orders of magnitude larger than that of today.
 
@@
 epistemological problems. For example, falsificationism, as argued by
-Karl Popper and adopted in \cite{avinatguide} will be unsuited, for
+Karl Popper and adopted in \cite{avinatguide} will be unsuitable, for
 reasons that more recent philosophers have pointed out, see
 \cite{chalmers1999whatis} for a comprehensive discussion: If it is
-insisted that hypotheses must be falsifiable, how can a researcher
+insisted upon that hypotheses must be falsifiable, how can a researcher
 know if their hypothesis should be rejected, or if it is something
@@
 and serialisation, where an underlying implementation might have
-relevant information to enhance e.g. the performance of an operation,
+relevant information to enhance, for example, the performance of an operation,
 but where the API does not allow it to expose it so that upper layers
@@
 
-Finally, the intention was to bring convergence to the topics of this
+Finally, the intention was to bring convergence in the topics of this
 thesis around a system to cache results that could be reused in

### 2016-07-22 dissertation/conclusions.tex
Helen's corrections

@@
 to blame for it. Our work, and most of the related work in
-Section~\ref{sec:related} approach this work from the direction of
+Section~\ref{sec:related} approach this problem from the direction of
 moving stress from the server, but it is also clear that another way
@@
 to identify queries it would be too heavy to answer already when
-parsing or planning, and also expose that in the service description
+parsing or planning, and also to expose that in the service description
 so that clients can avoid asking them. The HTTP/2 specification

### 2016-07-22 dissertation/introduction.tex
Helen's corrections

@@
 \begin{enumerate}
-\item Introduction of a path to critical practice of evaluations, that makes
+\item Introduction of a path for critical practice of evaluations, that makes
   use of contemporary statistical techniques, to establish a practice
@@
 \item A didactical experiment to help researchers understand the statistics.
-\item The novel application of a well established method in
+\item The novel application of a well-established method in
   statistics, rarely used in Computer Science, to SPARQL endpoint

### 2016-07-22 dissertation/papersincontext.tex
Helen's corrections

@@
 challenge the client first, but this should not happen on a resource
-that the client is permitted to read. Thus, the proposed hypermedia
+that the client is permitted to read. Therefore, the proposed hypermedia
 needs a property so that the client understands what to do if it wants
@@
 To understand the context of this study, I noted none of the
-benchmarks that was in use for the evaluation of SPARQL engines could
+benchmarks that were in use for the evaluation of SPARQL engines could
 be used for basic statistical hypothesis testing, and clearly,
@@
 in its own right, clearly beyond the scope of this
-dissertation. Unfortunately, this also implies that I do not have the
-rigorous evaluation that I would have preferred to have in this
-dissertation, it is unattainable.
+dissertation. Unfortunately, this also implies that this dissertation
+does not feature an evaluation that lives up to the rigorous standards
+I am advocating as it is unattainable within the time frame of this
+work.
 
@@
 International Semantic Web Conference, by the lead author Gregory Todd
-Williams, we were the two authors. The reference is
-\cite{williamspushing}.
+Williams. The reference is \cite{williamspushing}.
 
-This work was the outcome of several hackathons that I had organised
+This work was the outcome of several hackathons that I organised
 in the Perl community. It is not very well known that Gregory Todd
@@
 
-Dissatisfaction with the methodology used in the literature,
-specifically benchmarking, to test assertions of performance lead to
-exploration of statistical methods referenced in
-Section~\ref{sec:condoe}, and to organise the Empirical 2014 Workshop
-at the Extended Semantic Web Conference (which I was unfortunately
-prevented from attending myself). Still, frustrated by the slow
-progress of the last 16 years, and difficulty with comparing methods
-to the more mature natural sciences, I still felt the need to express
-a philosophical view of science, which I did by submitting an article
-to the ``Negative or inconclusive results in Semantic Web'' at the
-Extended Semantic Web Conference 2015. The essay, see
+This direction was motivated by a dissatisfaction with the methodology
+used in the literature, specifically benchmarking, to test assertions
+of performance. That lead to exploration of statistical methods referenced
+in Section~\ref{sec:condoe}, and to the organising of the Empirical 2014
+Workshop at the Extended Semantic Web Conference (which I was
+unfortunately prevented from attending myself). Still, frustrated by
+the slow progress of the last 16 years, and the difficulty in comparing
+methods to the more mature natural sciences, I still felt the need to
+express a philosophical view of science, which I did by submitting an
+article to the ``Negative or inconclusive results in Semantic Web''
+workshop at the Extended Semantic Web Conference 2015. The essay, see
 \cite{kjernsmo_noise_2015} was tangential to the Call for Papers, but

### 2016-07-22 dissertation/cache-proxy.tex
Helen's corrections

@@
 The system is in the convergence of the directions described in this
-dissertation: It should relieve the remote endpoint of some of the
+thesis: It should relieve the remote endpoint of some of the
 burden to evaluate the query and it should add to the robustness of
@@
 \cite{ldf1}. The system does not support named graphs (i.e. only
-triple stores), but other than that, supports SPARQL 1.1 in its
-entirety. This was made possible to develop quickly by the efforts put
+triple stores), but other than that, it supports SPARQL 1.1 in its
+entirety. It was possible to develop this quickly by the efforts put
 into the query planning in the Attean framework described in
@@
   down) and responses (when pointing up) in a deployment scenario for
-  when a the query cache considered in this paper is deployed on a
-  caching proxy in the Internet infrastructure, see the text for
-  details.}\label{fig:messaging}
+  when the query cache considered in this paper is deployed on a
+  caching proxy in the Internet infrastructure (see the text for
+  details).}\label{fig:messaging}
 \end{center}
@@
 transmitted directly from the SPARQL Endpoint at the database back to
-the client. The entire response may be cached at the proxy in case it
+the client. The entire response may be cached on the proxy in case it
 can be reused in its entirety. However, the main point is that the
@@
 \circled{3}. In that case, the proxy will send a rewritten query for
-the rest of the data it needs to answer the query to the origin
+the rest of the data it needs to be able to answer the query to the origin
 server, indicated by the dashed arrows. This may be a SPARQL query, a
@@
 the analyser may have identified another Triple Pattern Fragment that
-it takes as likely useful for future requests, and sends a request and
+it takes as likely to be useful for future requests, and sends a request and
 caches the response \circled{4}. At some point, it may be able to
@@
 even more. As such, it builds on the code of hundreds of authors, too
-long to enumerate. I shall constrain myself to list the modules that
-have code that has been motivated from this study. Modules that have
+many to enumerate. I shall constrain myself to list the modules that
+have code that has been motivated by this study. Modules that have
 other primary authors than myself have been started prior to this
@@
 
-The system also depends on modules I have written that is not part of
+The system also depends on modules I have written that are not part of
 the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
@@
 Traits are groups of methods that serve as a primitive unit of code
-reuse. Traits are not constructed to objects themselves and do not
-use inheritance for composition, instead a class is composed by a set of
-traits and possibly the class' own methods, and then instantiated. A
-trait can also require a set of methods or attributes, and as such, is similar to
-interfaces, but since its focus is code reuse, it can also provide a
-default implementation.
+reuse. Traits are not constructed into objects themselves and do not
+use inheritance for composition, instead a class is composed by a set
+of traits and possibly the class' own methods, and then
+instantiated. A trait can also require a set of methods or attributes,
+and as such, is similar to an interface, but since its focus is code
+reuse, it can also provide a default implementation.
 
@@
 combination with traits-based composition, but only when the base
-class provide only fundamental functionality that is very likely to be
+class provide just fundamental functionality that is very likely to be
 common to all implementations. For an example of this, see
@@
 proportional to the number of triple patterns in the plan, but
-penalises plans that does not have triple patterns that are connected
+penalises plans that do not have triple patterns that are connected
 through a variable (i.e. will cause a Cartesian join) with a factor
@@
 very successful. The cache will store the results as an array if two
-RDF terms were bound in the triple pattern, or a hash if only one term
-was bound. The strings put into the cache are serialized N-Triples
+RDF terms are bound in the triple pattern, or a hash if only one term
+is bound. The strings put into the cache are serialized N-Triples
 strings.
@@
 everything else is provided by the roles. Conventional single
-inheritance is constrained to tree structure, and would result in a
+inheritance is constrained to a tree structure, and would result in a
 very complex system. Another option is to use the factory method
@@
 heuristics are fairly elaborate. Moreover, since the objective of the
-study is to take load off of the remote endpoint, the heuristics is
+study is to take load off of the remote endpoint, the heuristics are
 intended to make the proxy work harder than the remote endpoint.
@@
 
-Next, we detail the cost of a join where one of more children in the
+Next, we detail the cost of a join where one or more children in the
 tree is a remote SPARQL plan. It is usually not wanted to execute
@@
 with emphasis on testing cases that I assumed \textit{a priori} to be
-difficult. The interested reader is referred to the published code,
+difficult. The interested reader may like to refer to the published code,
 given in Section~\ref{sec:impl}. Some of the tests consist of a number
-of queries, and some of the tests concerns the problem I found in a
+of queries, and some of the tests concern the problem I found in a
 preliminary study, that the cached triples would often break
@@
 intended and that analysis and prefetching is able to retrieve and
-cache results, an elaborate evaluation would be needed to further
-explore whether the system actually helps remote endpoints and is able
+cache results, an elaborate evaluation would be needed to explore
+further whether the system actually helps remote endpoints and is able
 to serve clients with adequate performance. Unfortunately, the
@@
 prevented further investigation. Thus, this section consists of an
-account of the evaluation that were performed and the methodology I
+account of the evaluation that was performed and the methodology I
 planned to employ.
@@
 
-In more detail, query logs can be obtained by from
-e.g. USEWOD~\cite{eps385344} or the Linked SPARQL Queries
+In more detail, query logs can be obtained from
+for example USEWOD~\cite{eps385344} or the Linked SPARQL Queries
 Dataset \cite{Saleem2015}. An extraction of one of those datasets
@@
 
-On the server, some metric of system load would recorded. In
+On the server, some metric of system load would be recorded. In
 statistical experimentation parlance, this is known as the ``response
@@
 
-As for the overall design of the experiment, the reader is referred to
+As for the overall design of the experiment, the reader may refer to
 the paper discussed in Section~\ref{sec:condoe} for an introduction to
@@
 planner, as it results in a plan where the two triple patterns that
-should run is a single SPARQL query is split into two. This resulted
+should run is a single SPARQL query are split into two. This resulted
 in an interesting observation:
@@
 (and possibly a general problem with any early-pruning planner) 
-when using join rotation. We did not attempt to examine this any
-further, however.
+when using join rotation. However, we did not attempt to examine this any
+further.
 
 To remedy this problem, we considered creating a custom query planner
-for Basic Graph Patterns, but I did not implement it. The outline
-would be as follows:
+for Basic Graph Patterns, but this was not implemented. The outline
+would have been as follows:
 \begin{enumerate}
@@
   and group them to ``components''.
-\item If a triple pattern have a cached result, create a plan that
+\item If a triple pattern has a cached result, create a plan that
   accesses the cache. 
-\item If the triple pattern in the previous step caused a Cartesian
+\item If the triple pattern in the previous step causes a Cartesian
   join to be created within a component, create an equivalent
@@
 
-One may raise the issue that it was a bad idea to begin with to use a
+The issue could be raised that it was a bad idea to begin with to use a
 scripting language like Perl to write a query engine with performance
@@
 approach, SPARQL 1.1 \sparql{VALUES} would also need to be supported,
-so that the local quad store could perform all join operations. To
-this end, I wrote a rudimentary Virtuoso store driver for Attean, but
+so that the local quad store could perform all join operations. With
+this in mind, I wrote a rudimentary Virtuoso store driver for Attean, but
 at that point, the project was overdue and needed to be concluded.
@@
 It seems likely that the project, with more time, could be brought to
-the point where a the evaluation in Section~\ref{sec:planned} could be
+the point where the evaluation in Section~\ref{sec:planned} could be
 applied, and new insight could be gained.

### 2016-07-22 dissertation/introdefhistory.tex
don't break sentence

@@
 As noted in the previous section, objectivity may be challenged by a
-researcher's intellectual investment in a certain field of study. I
+researcher's intellectual investment in a certain field of study and I
 admit that I have a very large intellectual investment in the success

### 2016-07-22 dissertation/introdefhistory.tex
divide sentence

@@
 semantics, as if you can understand what you are viewing simply by
-looking at the message, you can begin processing it), anyone can mint
+looking at the message, you can begin processing it). Also, anyone can mint
 identifiers and anybody can reuse those identifiers in their own

### 2016-07-23 dissertation/introduction.tex
Notes on IRI scheme

@@
 
-Note that all the IRIs uses the HTTP schema. The Hypertext Transfer
+Note that all the IRIs uses the HTTP scheme. The Hypertext Transfer
 Protocol (HTTP) is the main application protocol used on the Internet
@@
 it is expressed as RDF, makes the above so-called Linked Data.
+
+HTTP IRIs are not the only possible choice, to the contrary, the
+scheme and what may be associated with it, such as a protocol, is an
+orthogonal issue to the data model and syntax, and any IRI scheme can
+be used.
 

### 2016-07-23 dissertation/cache-proxy.tex
add todo

@@
 doing a Cartesian join and doing a remote SPARQL evaluation of data
-already in the cache, as plans for both these options would be provided.
+already in the cache, as plans for both these options would be
+provided.\todo{Discuss algebra equivalences here?}
 

### 2016-07-23 dissertation/papersincontext.tex
Small

@@
 Workshop at the Extended Semantic Web Conference (which I was
-unfortunately prevented from attending myself). Still, frustrated by
+unfortunately prevented from attending myself). Frustrated by
 the slow progress of the last 16 years, and the difficulty in comparing

### 2016-07-23 dissertation/conclusions.tex
create etags

@@
 if two queries are equivalent. As noted in Section~\ref{sec:relcache},
-the approach \cite{papailiou2015graph} took to canonical labelling may
-be extended to address this problem.
+\cite{kaseicache} can be used for this, and the approach
+\cite{papailiou2015graph} took to canonical labelling may be extended
+to address this problem
 

### 2016-07-23 dissertation/conclusions.tex
More of Martin's feedback

@@
 
-As noted in Section~\ref{sec:motivation} it is clear that evaluating
+As noted in Section~\ref{sec:motivation}, it is clear that evaluating
 arbitrary SPARQL queries are in general unsustainable, and it is

### 2016-07-23 dissertation/introdefhistory.tex
More of Martin's feedback

@@
 
-With identifiers, and the hypertext language HTML came the fact that I
-could link to anyone, and anyone could link to me was highly
-empowering: Discovering that people I didn't know linked to my
-material was a strong, social reward. After some time, I had tens of
-thousands of visitors every month.
+With identifiers, and the hypertext language HTML came the ability to
+link to anyone, and anyone could link to me was highly empowering:
+Discovering that people I didn't know linked to my material was a
+strong, social reward. After some time, I had tens of thousands of
+visitors every month.
 

### 2016-07-23 dissertation/motivation.tex
add todo
More of Martin's feedback

@@
 with the data so it could be argued that most RDF documents would be
-hypermedia, but what is often overlooked is that the documents need to
-be explicit in linking schema from all documents. Moreover, control
-information to enable anticipated and serendipitous applications must
-be made available in the RDF. For example, It is not sufficient for a pizza
-restaurant to publish data on pizzas and link to the schema, the RDF
-must also express how to order to enable a software agent to organise
-a party.
+hypermedia, but what is often overlooked is that for RDF documents to
+be self-descriptive, they need to explicitly link their relevant
+schema. Moreover, control information to enable anticipated and
+serendipitous applications must be made available in the RDF. For
+example, It is not sufficient for a pizza restaurant to publish data
+on pizzas and link to the schema, the RDF must also express how to
+order to enable a software agent to organise a party.
 
@@
 this field. These questions are as hard to ask and to live with as to
-answer.
+answer. \todo{Too fluffy?}
 
@@
 
-Clearly, the above motivation points out federated SPARQL as a key
+Based on the this discussion, federated SPARQL could be forwarded as a 
 solution to the problem of query answering over decentralised sources,
@@
 a contribution to ease the burden of servers by allowing other parties
-to take parts of the query evaluation.
+to perform parts of the query evaluation.
 

### 2016-07-23 dissertation/papersincontext.tex
More of Martin's feedback

@@
 used in the literature, specifically benchmarking, to test assertions
-of performance. That lead to exploration of statistical methods referenced
+of performance. That led to exploration of statistical methods referenced
 in Section~\ref{sec:condoe}, and to the organising of the Empirical 2014

### 2016-07-23 dissertation/problems.tex
More of Martin's feedback

@@
 
-The same kind of problems occur in many operations, such as parsing
+The same kind of problem occurs in many operations, such as parsing
 and serialisation, where an underlying implementation might have
@@
 
-Another type of problems occur when working with dynamic RDF that may
+Another type of problem occurs when working with dynamic RDF that may
 mix terminological and assertional information with object-oriented
@@
 Triple Pattern Fragments shifts most of the processing from servers to
-clients, but Problem~\ref{prob:microcontroller} makes that in some
-cases infeasible.  In addition,
+clients, but Problem~\ref{prob:microcontroller} makes that infeasible
+in some cases.  In addition,
 problems~\ref{prob:zipf}~and~\ref{prob:uniform} are compounded by

### 2016-07-23 dissertation/related.tex
do not use prior art
More of Martin's feedback

@@
 answer queries, and in doing so, they employed hypermedia, and formulated
-this as Triple Pattern Fragments. The key to this hypermedia is that
+this as Triple Pattern Fragments (TPF). The key to this hypermedia is that
 an answer does not only include the data, but also metadata,
@@
 
-Overall, this study share many of the observations and goals that
+Overall, the work on query answering with Triple Pattern Fragments
+share many of the observations and goals that
 motivate this study. Their stated goal is to transfer as much as
@@
 triple patterns and store them in a file system, which can be managed
-by a simple Web server. The results may also be paged, so that each
+by a simple Web server. The results may also be paginated, so that each
 individual response may be kept small.
@@
 First, we note a claim that SPARQL query caching is not possible:
-In \cite{hogan2014paths}, the authors examine cacheable as one of the
+In \cite{hogan2014paths}, the authors examine cacheability as one of the
 desiderata for sustainable data access. They claim, without further
@@
 HERMES is that such invariants will be encoded by a domain expert. The
-Attean framework, which is used extensively in this work and is
+Attean framework, which is used extensively in my work and is
 detailed in Section~\ref{sec:attean} provides facilities to do this in
@@
 in the possible but unlikely case where the remote server declares
-that an infinite freshness lifetime for the result of a certain graph
+an infinite freshness lifetime for the result of a certain graph
 pattern. While statistics summarisation is relevant and should be
@@
 obligatory schema in SQL databases. This provides an opportunity for
-novel research. The solution proposed in this paper is not a cache on
+novel research. The solution proposed in \cite{papailiou2015graph}  is not a cache on
 a proxy or a mediator like that which is my primary focus. The execution engine
-in this case, has access to both the primary indexes of the RDF store
+in this case, has access to both the primary indices of the RDF store
 and the cache. A key contribution of the paper is a canonical
@@
 translated to non-negative integers (colours) using a sort key, to
-form a directed vertex-and-edge-coloured graph that the authors
-claim is isomorphic with the original SPARQL. Finally, the graph is
+form a directed vertex-and-edge-coloured graph that according to the
+authors is isomorphic with the original SPARQL. Finally, the graph is
 transformed to a simpler directed vertex-coloured graph. With this
@@
 benefit the execution of a subsequent query. Nevertheless, they are
-important to consider as prior art:
+important to consider as prior work:
 Umbrich et al. \cite{umbrich2012hybrid} considered a situation where
@@
 therefore expose a \httph{Last-Modified} header and validate an
-earlier response in accordance with RFC7232 \cite{rfc7232} cheaper
-than to evaluate the query. This also represent an alternative to HTTP
+earlier response in accordance with RFC7232 \cite{rfc7232} in a cheaper way
+than by evaluating the query. This also represent an alternative to HTTP
 \httph{ETag} as in the discussion of \cite{papailiou2015graph}.

### 2016-07-23 dissertation/conclusions.tex
Martins comments

@@
   Web.
-\item Traits eases the task of developing experimental systems and
+\item Traits ease the task of developing experimental systems and
   usage of optimisations in underlying systems.
-\item While the adoption of caching and conditional request as defined
+\item While the adoption of caching and conditional requests as defined
   in HTTP is not widespread, the adoption is sufficient to start using
-  it in practical systems.
+  them in practical systems.
 \item Evaluation methodologies suffer from a poor foundation in

### 2016-07-24 dissertation/introduction.tex
More of Martin's feedback
Use an example environment

@@
 
+\begin{example}{Turtle syntax}\label{ex:turtle}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 
@@
 
+\begin{example}{A triple pattern}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 This triple pattern will match the two triples 
+\begin{example}{Triples matched by triple pattern}
 \begin{verbatim}
@@
 However, even though the result does not depend on the order of
-execution, the time it takes to find the result often does. For
+execution of a BGP, the time it takes to find the result often does. For
 example, consider the following BGP:
+\begin{example}{A Basic Graph Pattern}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 In most cases, if the triple pattern with the \rdfterm{foaf:name}

### 2016-07-24 dissertation/details.tex
intro prelim
details on cartesians

@@
 \section{Preliminaries}
+
+This section extends the basic introduction in
+Section~\ref{sec:semwebtech} where required to understand the
+following discussion.
 
@@
 used to validate it.
+
+\subsection{Topics in query evaluation}\label{sec:prelimquery}
+
+As shown in \cite{perez2009semantics} that SPARQL has some highly
+complex parts. In particular, nested \sparql{OPTIONAL}s will be
+difficult to evaluate. It is possible for a user to either inadvertently
+or with malicious intent (in e.g. an denial of service attack) submit
+a query that will cause excessive load to the server.
+
+Another problem worthy of special attention is the case where parts of
+the query has no connecting variable. Let us consider the
+simple example of when two triple patterns do not share a variable:
+\begin{verbatim}
+?person foaf:name ?name .
+?contribution dct:title ?title .
+\end{verbatim}
+When this is evaluated, it will cause what is known as a ``Cartesian
+join'', in which the result will contain \emph{all} the solutions of
+the first triple pattern joined with \emph{all} the solutions of the
+second triple pattern. This will usually be a large result, and it is
+also resource consuming to compute.
+
+In some cases, the users may have written such a query, in which case
+the query planner only has the options to execute the query or reject
+it entirely. In other cases, the query planner may arrive at such
+queries as part of query rewriting, in which case it must ensure that
+such plans are not preferred.

### 2016-07-24 dissertation/problems.tex
ref new query eval prelim

@@
 
-It was found in \cite{perez2009semantics} that
+As discussed in \ref{sec:prelimquery},
 
 \begin{problem}\label{prob:sparqlcomplex}
-SPARQL has some highly complex parts.
+Some features of SPARQL may cause excessive endpoint load.
 \end{problem}
 
-In particular, nested \sparql{OPTIONAL}s will be difficult to
-evaluate. A Cartesian join is also likely to cause trouble, but even
-seemingly straightforward queries may be heavy. We also note from
-experience that:
+We also note from experience that:
 

### 2016-07-24 dissertation/problems.tex
on distributions

@@
 Moreover, an ambition of the Semantic Web is to deal well with
-heterogeneous data, and it has been known for long \cite{1385469} that:
+heterogeneous data and it has been known for long \cite{1385469} that
+power laws are frequently encountered when fitting a distribution to
+data. As noted in the discussion of Example~\ref{ex:bgp} in
+Section~\ref{sec:introinteract}, a query planner would often rely on
+statistics, but:
 
-\begin{problem}\label{prob:zipf}
-The distribution of data is often dominated by a power-law.
+\begin{problem}\label{prob:unknowndist}
+The data distribution is often unknown to the query engine.
 \end{problem}
-However, 
-\begin{problem}\label{prob:uniform}
-The data distribution is often unknown to the query engine, and so,
-following the principle of indifference, assumes uniform
-distributions.
+Only simple summary statistics may be available. The discussion of
+Example~\ref{ex:bgp} illustrates that even if the distribution of data
+given for example a predicate is known, 
+\begin{problem}\label{prob:unknownconddist}
+\emph{conditional} data distributions are usually unknown to the query engine.
 \end{problem}
+For example, the population of geographical places may be uniformly
+distributed in a given data set, but the population given that the
+places are of type city may be a power law. This could have very
+significant impact on the overall performance, since in the latter
+case, the query planner may consider a BGP like:
+\begin{example}{Skew distribution}
+\begin{verbatim}
+?place ex:population ?population ;
+       a ex:City .
+\end{verbatim}
+\end{example}
+as a unit.\todo{Move to preliminaries?}
 

### 2016-07-24 dissertation/problems.tex
rewrite microcontroller problem

@@
 Heterogeneity should also be expected in different types of servers
-and clients, in particular
+and clients. In particular, microcontrollers and to a lesser extent,
+mobile phones, are resource constrained devices. Thus:
 
 \begin{problem}\label{prob:microcontroller}
-A client running on a microcontroller does not have significant
-processing capability.
+Clients run on devices with a wide range of processing capability.
 \end{problem}

### 2016-07-25 dissertation/problems.tex
more explanations in problems #6

@@
 
+Even if this statistics could be found, it is not clear that it should:
+\begin{problem}\label{prob:largestats}
+Comprehensive statistics may be costly to compute and requires much
+storage space.
+\end{problem}
+Therefore, it is important to find a balance.
+
 Heterogeneity should also be expected in different types of servers
@@
 
+The performance issue discussed in Section~\ref{sec:prelimquery} has
+relevance in the context of caching:
+
 \begin{problem}\label{prob:cachecartesian}
@@
 
+To understand this, consider Example~\ref{ex:cartesian}, where the BGP
+was a part of a query with 3 triple patterns. In addition to the two
+triple patterns in Example~\ref{ex:cartesian}, a triple pattern that
+connected the two, for example
+\rdfterm{?person~foaf:publications~?paper~.}. If the result of this
+triple pattern is cached, what is left to evaluate are the two triple
+patterns in Example~\ref{ex:cartesian}, which is likely a very
+expensive operation.
+
+It would be useful when caching if one could compare the incoming
+query to a query with an already cached result and determine that they
+were equivalent, but unfortunately:
 \begin{problem}\label{prob:queryeq}
@@
 
+Metadata is important for query planning, and as explained in
+Section~\ref{sec:httpcache}, the HTTP protocol provides facilities for
+sharing such information. However:
 \begin{problem}\label{prob:sanity}
@@
 
+Even though statistics may be available internally in a query engine,
+a caching proxy is unlikely to have direct access to that data, and therefore:
 \begin{problem}\label{prob:nostats}

### 2016-07-25 dissertation/details.tex
add ref #6

@@
 simple example of when two triple patterns do not share a variable:
+\begin{example}{BGP with Cartesian join}\label{ex:cartesian}
 \begin{verbatim}
 ?person foaf:name ?name .
-?contribution dct:title ?title .
+?paper  dct:title ?title .
 \end{verbatim}
+\end{example}
 When this is evaluated, it will cause what is known as a ``Cartesian

### 2016-07-25 dissertation/problems.tex
eval problems #6

@@
 ontology matching or ontology alignment, and is a very active area of
-research.\todo{take some open problems in ontology alignment}
+research. This problem, as well as related problems, such as entity
+resolution, are considered orthogonal to the problems in this thesis,
+so the below problems are issues that arise from the ideal of an open
+decentralised Web.
 
@@
 How can the adoption of existing vocabularies be encouraged? 
-\todo{ Find the ref by Frank on the economics}
 \end{problem}
@@
 and workloads are important to devise a test to evaluate assertions
-about performance, the fundamental problem arises that we currently
-test against something that does not exist, and this differs markedly
-from the methodology of natural science.
+about performance, the fundamental problem arises that: 
+\begin{problem}\label{prob:testingnonexist}
+We currently test against something that does not exist and therefore
+usage patterns cannot be predicted,
+\end{problem}
+and this differs markedly from the methodology of natural science.
 
@@
 of the Semantic Web information universe\footnote{See
-  e.g. \url{http://videolectures.net/iswc2011_van_harmelen_universal/}
-  and
-  \url{http://sssw.org/2015/invited-speakers-tutors/frank-vanharmelen/}}.
+  e.g. \url{http://videolectures.net/iswc2011_van_harmelen_universal/}}.
 If this direction is pursued with success, then it may be possible to
@@
 \begin{problem}\label{prob:lawsfuture}
-Establish laws of the Semantic Web information universe that can reasonably be
-expected to have validity far beyond the present.
+No methodology has been established to find laws of the Semantic Web
+information universe that can reasonably be expected to have validity
+far beyond the present.
 \end{problem}

### 2016-07-25 dissertation/problems.tex
improve wording #6

@@
 
-To understand whether a result from a study on for example, performance is
-valid, an empirical study must be conducted, as a SPARQL system is
-generally too complex to be evaluated solely by formal
-methods. Conventionally, one or more benchmarks have been used for
-this purpose, and there exists a number of such
-benchmarks. Unfortunately, due to this complexity, they have to test a
-very large number of parameters, but are unable to do that, and they
-may also oversimplify the test by attempting to neutralise certain
-optimisation techniques. Further, they have no structured approach to
-investigate flaws in the benchmark itself, and have no meaningful
-summary statistics that can be used to understand the overall
+To understand whether a result from a study on for example,
+performance is valid, an empirical study must be conducted, as a
+SPARQL system is generally too complex to be evaluated solely by
+formal methods. Conventionally, one or more benchmarks have been used
+for this purpose, and there exists a number of such
+benchmarks. Unfortunately, due to the complexity of the systems under
+evaluation, benchmarks have to test a very large number of parameters,
+which is methodologically beyond their reach. They may also
+oversimplify the test by attempting to neutralise certain optimisation
+techniques. Further, they have no structured approach to investigate
+flaws in the benchmark itself, and have no meaningful summary
+statistics that can be used to understand the overall
 performance. While one may standardise benchmarks, doing so makes it

### 2016-07-25 dissertation/problems.tex
move the sparql hypermedia problem typo #6

@@
 used to describe simple query interfaces, define how read-write
-operations are to be made, etc. \todo{more here}
+operations are to be made, etc.
+
+\begin{problem}\label{prob:tpf}
+Since the SPARQL language is described in an external specification,
+it cannot be hypermedia.
+\end{problem}
+
+It is likely that the adoption of the open Semantic Web would be
+accelerated if large numbers of developers felt the technology was
+more accessible to them. There are a number of broad as well as
+detailed problems that must be solved to achieve this:
 
@@
 server that offers a read-write interface.
-\end{problem}
-
-\begin{problem}\label{prob:tpf}
-Since the SPARQL language is described in an external specification,
-it cannot be hypermedia.
 \end{problem}
@@
 \begin{problem}\label{prob:dynaprog}
-Current programming paradigm deals poorly with dynamic RDF data
+Current programming paradigms deal poorly with dynamic RDF data
 containing a mix of terminological and assertional information.

### 2016-07-25 dissertation/problems.tex
improve explanation of get_pattern etc

@@
 Until recently, a query engine would need to break down a SPARQL query
-to individual triple pattern matches, which would then use for example
+to individual triple pattern matches, see
+Section~\ref{sec:introinteract} for how frameworks tend to interact
+with data. Typically, the query engine would then use for example
 the \jcode{listStatements()} method of Jena, \jcode{filter()} method
@@
 \end{problem}
+
+For example, an underlying SQL store could not compile one query for a
+BGP, and could not use a \jcode{WHERE} clause to evaluate a SPARQL
+\rdfterm{FILTER}. To remedy this, one could have a method for
+evaluating BGPs with (such as the  \pcode{get\_pattern} method of
+\pmodule{RDF::Trine}) and without filters, but:
 

### 2016-07-25 dissertation/papersincontext.tex
Inline problems #12

@@
 
-This section will put the included papers into context. It also enjoys
-the benefit of hindsight, and will therefore address some of the
-shortcomings that have surfaced since their publication.
+This section will put the included papers into context, and enumerates
+the problem (from Section~\ref{sec:problems}) each paper is intended
+to address. It also enjoys the benefit of hindsight, and will
+therefore address some of the shortcomings that have surfaced since
+their publication.
 
@@
 
-\todo{relate to problems}
 This paper is relevant to the topic of the dissertation for two
@@
 
-The main idea in this paper is to look upon RDF as a natural language
+The main idea in this paper is to address Problem~\ref{prob:lapis} 
+by looking upon RDF as a natural language
 sentence with subject, predicate and object, and that sentence should
@@
 
+Problem~\ref{prob:badstats} motivates this paper. 
 To understand the context of this study, I noted none of the
@@
 
+Originally, this paper was intended as the start of a methodology to
+evaluate a federated system that addressed the
+problems~\ref{prob:unknowndist},~\ref{prob:unknownconddist}~and~\ref{prob:largestats},
+but as \cite{buil2013sparql} confirmed my suspicions that there were
+severe stability problems with SPARQL endpoints, my attention shifted
+towards the more immediate Problem~\ref{prob:syntacticcache}.
+
 \subsection{Pushing Complexity Down the Stack}\label{sec:conpush}
@@
 
-This work was the outcome of several hackathons that I organised
-in the Perl community. It is not very well known that Gregory Todd
-Williams' SPARQL implementation written in Perl is one of the SPARQL
-1.1 reference implementations. It is also fully compliant with
-the specification. Through our work, and the availability of
-traits-based programming in Perl, see \cite{traits}, we had extensive
-discussions on how to leave optimisations to lower levels, e.g. the
-underlying database engine, when they had the potential to provide a
-better plan or cost estimate. While it is certainly possible to do
-this without this new programming paradigm, traits made the
-methodology much more comprehensible, and so helped the subsequent
-work substantially.
+This work was the outcome of several hackathons that I organised in
+the Perl community, where
+problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}
+were discussed at length after having become a major practical
+issue. It is not very well known that Gregory Todd Williams' SPARQL
+implementation written in Perl is one of the SPARQL 1.1 reference
+implementations. It is also fully compliant with the
+specification. Through our work, and the availability of traits-based
+programming in Perl, see \cite{traits}, we had extensive discussions
+on how to leave optimisations to lower levels, e.g. the underlying
+database engine, when they had the potential to provide a better plan
+or cost estimate. While it is certainly possible to do this without
+this new programming paradigm, traits made the methodology much more
+comprehensible, and so helped the subsequent work substantially.
 
@@
 I committed to do this survey after Jürgen Umbrich challenged me to see
-if the standards were actually followed. Rather than just survey the
+if the standards were actually followed, and thus address
+Problem~\ref{prob:sanity} before further work to solve practical
+caching problems. Rather than just survey the
 usage for SPARQL Endpoints, I decided to do it for the breadth of the
@@
 methods to the more mature natural sciences, I still felt the need to
-express a philosophical view of science, which I did by submitting an
+address Problem~\ref{prob:badphil} and express a philosophical view of 
+science, which I did by submitting an
 article to the ``Negative or inconclusive results in Semantic Web''
@@
 
-As a philosophical text, the contributions are hard to detail. It is
+As a philosophical text, the contributions are hard to detail. It
+should establish Problem~\ref{prob:testingnonexist} as a fundamental
+problem. It is further
 written in a provocative style, to encourage discussions of
@@
 aware of this keynote by the reviewer.
+
+\section{Unpublished Work on Query Caching}
+
+Finally, the intention was to bring convergence in the topics of this
+thesis around a system to cache results that could be reused in
+subsequent queries. As noted in Section~\ref{sec:motivcache}, this was
+motivated by opportunity as well as problems, but it is intended to
+address
+problems~\ref{prob:syntacticcache}~and~\ref{prob:cachecartesian}. This
+work was not finished within the time frame of the dissertation, but
+is discussed at length in Chapter~\ref{sec:tpfcacheplanning}. This
+work relates to the problems in Section~\ref{sec:problems} as follows:
+Triple Pattern Fragments shifts most of the processing from servers to
+clients, but Problem~\ref{prob:microcontroller} makes that infeasible
+in some cases.  In addition,
+problems~\ref{prob:unknowndist},~\ref{prob:unknownconddist}~and~\ref{prob:largestats},
+are compounded by Problem~\ref{prob:nostats}, but alleviated somewhat
+by the presence of cardinality estimates given by Triple Pattern
+Fragments, which we exploit.

### 2016-07-25 dissertation/problems.tex
Inline problems #12

@@
 \end{problem}
-
-\subsection{Problems Addressed by This Work}\label{sec:probaddress}
-
-Given the technical experience outlined in Section~\ref{sec:history},
-I first addressed Problem~\ref{prob:lapis}, by suggesting a simple
-vocabulary for the server to tell the client what primitive operations
-it is allowed to perform on a given resource. See
-Section~\ref{sec:conlapis} for more on this paper.
-
-My next concern were the implications of the
-problems~\ref{prob:zipf}~and~\ref{prob:uniform} in a federated SPARQL
-engine. While studying the evaluations done in the relevant literature
-in the field, I decided to start an effort to address
-Problem~\ref{prob:badstats} first by adopting methodology from the
-statistical field of ``Design of Experiments'' (DoE). See
-Section~\ref{sec:condoe} for details on this work.
-
-Then, as \cite{buil2013sparql} confirmed my suspicions that there were
-severe stability problems with SPARQL endpoints,
-Problem~\ref{prob:syntacticcache} became my main focus, and I decided
-to write a query planner that could reuse previous results or prefetch
-data the system deemed relevant, but then we realised that the current
-programming paradigm was quite unhelpful when customising a query
-planner for such a purpose. Rather than writing a prototypical query
-engine for illustrative purposes, we wanted to integrate it into a
-fully compliant SPARQL query engine, and address
-problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}. This
-is explained in Section~\ref{sec:conpush}.
-
-Before addressing Problem~\ref{prob:syntacticcache}, I found that the
-understanding of the actual usage of caching was inadequate, so I
-decided to solve Problem~\ref{prob:sanity} first. This work is
-discussed in Section~\ref{sec:consanity}.
-
-Problem~\ref{prob:badphil} has been a growing concern throughout this
-work. A proper evaluation should be an important part of the
-dissertation, and yet, the problems discussed in
-Section~\ref{sec:evalproblems} are so severe that the time frame of
-this work does not allow a satisfactory solution to these
-problems. The paper discussed in Section~\ref{sec:conphil} is
-motivated in part from these problems.
-
-Finally, the intention was to bring convergence in the topics of this
-thesis around a system to cache results that could be reused in
-subsequent queries. As noted in Section~\ref{sec:motivcache}, this was
-motivated by opportunity as well as problems, but it is intended to
-address
-problems~\ref{prob:syntacticcache}~and~\ref{prob:cachecartesian}.
-Triple Pattern Fragments shifts most of the processing from servers to
-clients, but Problem~\ref{prob:microcontroller} makes that infeasible
-in some cases.  In addition,
-problems~\ref{prob:zipf}~and~\ref{prob:uniform} are compounded by
-Problem~\ref{prob:nostats}, but alleviated somewhat by the presence of
-cardinality estimates given by Triple Pattern Fragments, which we
-exploit.

### 2016-07-25 dissertation/problems.tex
remove labels on unapproached problems

@@
 
-\begin{problem}\label{prob:infgivecode}
+\begin{problem}%\label{prob:infgivecode}
 How can suitable inference engines be made available to all users?
@@
 
-\begin{problem}\label{prob:sayanything}
+\begin{problem}%\label{prob:sayanything}
 With a centralised ontology approach, how can we ensure that anybody is enabled to say
@@
 
-\begin{problem}\label{prob:useexist}
+\begin{problem}%\label{prob:useexist}
 How can the adoption of existing vocabularies be encouraged? 

### 2016-07-25 dissertation/conclusions.tex
remove todo
Add characteristic sets cite
cap

@@
 
+Problems~\ref{prob:unknowndist},~\ref{prob:unknownconddist}~and~\ref{prob:largestats}
+has a promising approach in \cite{5767868}.
+
 I have not approached any of the problems of
@@
 natural to consider them out of scope.
-\todo{However, I shall assume that no inference is needed in the
-following examples.}
 
@@
 sets. Nevertheless, the theoretical framework that seems to solve the
-problem was proposed in 1991 in the book ``The art of the metaobject
-protocol'' \cite{kiczales1991art}. Perl has a mature implementation of
+problem was proposed in 1991 in the book ``The Art of the Metaobject
+Protocol'' \cite{kiczales1991art}. Perl has a mature implementation of
 the metaobject protocol, and the topic has been discussed on several

### 2016-07-25 dissertation/conclusions.tex
Mention new experimentalism

@@
 
+Finally, the problems~\ref{prob:badphil}~and~\ref{prob:badstats}
+should prompt an elaborate approach to understand and apply the
+practical implications of contemporary philosophy of
+science. \cite{Mayo2005-MAYEAP} proposes a framework that defines what
+constitutes a \emph{severe} test, and one possible direction is to
+apply this to query endpoint evaluation.
+
 \section{Conclusions}\label{sec:conclusions}

### 2016-07-25 dissertation/conclusions.tex
mention new experimentalism add new

@@
 but at present, the field is too immature to attempt the formulation
-of any laws. As noted in Section~\ref{sec:history}, this study is
+of any new laws. As noted in Section~\ref{sec:history}, this study is
 partly motivated by my subjective experience of why the Web was so

### 2016-07-25 dissertation/conclusions.tex
remove a too bombastic statement

--- 2016-07-25 dissertation/conclusions.tex
@@
 focus of our attention. In this work, the methodology used to study
-Problem~\ref{prob:sanity} could be used to gain such understanding,
-but at present, the field is too immature to attempt the formulation
-of any laws. As noted in Section~\ref{sec:history}, this study is
-partly motivated by my subjective experience of why the Web was so
-successful. Clearly, this should be studied more objectively, but I am
-not aware of any such study. Such study could also be applied to see
-if the experiences from the emergence of the Web can be applied to
-accelerating the Semantic Web, like I subjectively propose.
+Problem~\ref{prob:sanity} could be used to gain such understanding. As
+noted in Section~\ref{sec:history}, this study is partly motivated by
+my subjective experience of why the Web was so successful. Clearly,
+this should be studied more objectively, but I am not aware of any
+such study. Such study could also be applied to see if the experiences
+from the emergence of the Web can be applied to accelerating the
+Semantic Web, like I subjectively propose.
 

### 2016-07-25 dissertation/papersincontext.tex
A bit of proofreading

@@
 challenge the client first, but this should not happen on a resource
-that the client is permitted to read. Therefore, the proposed hypermedia
-needs a property so that the client understands what to do if it wants
-to do write operations.
+that the client is permitted to read without authentication. 
+Therefore, the proposed hypermedia needs a property so that the client
+understands what to do if it wants to do write operations.
 
@@
 is discussed at length in Chapter~\ref{sec:tpfcacheplanning}. This
-work relates to the problems in Section~\ref{sec:problems} as follows:
+work further relates to the problems in Section~\ref{sec:problems} as follows:
 Triple Pattern Fragments shifts most of the processing from servers to

### 2016-07-25 dissertation/problems.tex
A bit of proofreading

@@
 
-As discussed in \ref{sec:prelimquery},
+As discussed in Section~\ref{sec:prelimquery},
 
@@
 
-To understand this, consider Example~\ref{ex:cartesian}, where the BGP
-was a part of a query with 3 triple patterns. In addition to the two
-triple patterns in Example~\ref{ex:cartesian}, a triple pattern that
-connected the two, for example
+To understand this, consider Example~\ref{ex:cartesian}, and extend
+the BGP to a query with 3 triple patterns. In addition to the two
+triple patterns in Example~\ref{ex:cartesian}, add a triple pattern that
+connect the two, for example
 \rdfterm{?person~foaf:publications~?paper~.}. If the result of this
@@
 so the below problems are issues that arise from the ideal of an open
-decentralised Web.
+decentralised Web. These problems are therefore not discussed further.
 

### 2016-07-25 dissertation/introdefhistory.tex
Link to more in structure

@@
 the Semantic Web. Section~\ref{sec:motivation} has an overview of the
-motivation behind this work.  Section~\ref{sec:related} puts the
+motivation behind this work.  Section~\ref{sec:prelims} provides some
+further technical details and Section~\ref{sec:related} puts the
 present work into the context within the existing
 literature. Section~\ref{sec:problems} summarises important research
-problems and details the motivation, and
-Section~\ref{sec:papersincontext} details how the contributions
-address these problems and discusses weaknesses of the papers with the
-benefit of hindsight. Key insights are finally summarised in
+problems and details the motivation.
+Sections~\ref{sec:papersincontext}~and~\ref{sec:unpublished} details
+how the contributions address these problems and discusses weaknesses
+of the papers with the benefit of hindsight. Future work is discussed
+in Section~\ref{sec:future}. Key insights are finally summarised in
 Section~\ref{sec:conclusions}.\todo{Big problem statement here}

### 2016-07-27 dissertation/cache-proxy.tex
Clean out my own stuff from HERMES and move some

@@
 the forward proxy, or in a CDN.
+
+It is further interesting to note the contrast to the HERMES
+\cite{adali1996query} system's concept of invariants discussed in
+Section~\ref{sec:relcache}: a SPARQL query cache should be oblivious
+to domain knowledge and therefore shouldn't assume any invariants,
+except in the possible but unlikely case where the remote server
+declares an infinite freshness lifetime for the result of a certain
+graph pattern.
 

### 2016-07-27 dissertation/related.tex
Clean out my own stuff from HERMES and move some

@@
 HERMES \cite{adali1996query} provided motivation for the paper in
-Section~\ref{sec:conpush}, but is also important for the
-work in Chapter~\ref{sec:tpfcacheplanning}. An important concept in
-the HERMES system is ``invariants''. They are, in our terminology,
-patterns that do not need to be looked up remotely, but can be
-integrated in the query answer by the mediator. The expectation in
-HERMES is that such invariants will be encoded by a domain expert. The
-Attean framework, which is used extensively in my work and is
-detailed in Section~\ref{sec:attean} provides facilities to do this in
-an easy and extensible way, and also accommodates for the task of
-computing a scalar cost metric, but HERMES has a more advanced cost
-model in that it has a cost vector with estimates for time required to
+Section~\ref{sec:conpush}, but is also important for the work in
+Chapter~\ref{sec:tpfcacheplanning}. An important concept in the HERMES
+system is ``invariants''. They are, in our terminology, patterns that
+do not need to be looked up remotely, but can be integrated in the
+query answer by the mediator. The expectation in HERMES is that such
+invariants will be encoded by a domain expert. The HERMES planner
+bases decision on a cost vector with estimates for time required to
 find the first answer, time to find all answers and the cardinality of
@@
 and optimiser, that rewrites the query taking into account the cache
-and the invariants. While HERMES is more general than what I detail in
-Chapter~\ref{sec:tpfcacheplanning}, we note that while invariants
-could be coded in Attean, a SPARQL query cache should be oblivious to
-domain knowledge and therefore shouldn't assume any invariants, except
-in the possible but unlikely case where the remote server declares
-an infinite freshness lifetime for the result of a certain graph
-pattern. While statistics summarisation is relevant and should be
+and the invariants. While statistics summarisation is relevant and should be
 studied in future work, we note that with a Linked Data Fragments

### 2016-07-27 dissertation/conclusions.tex
Link to more in structure
Move etag labelling to future work

@@
-\section[Future Work]{Future work; Discussion of Other Problems}
+\section[Future Work]{Future work; Discussion of Other Problems}\label{sec:future}
 
@@
 
-Problem~\ref{prob:queryeq} is best addressed by creating hashing
-algorithms that have a reasonable chance of returning the same digest
-if two queries are equivalent. As noted in Section~\ref{sec:relcache},
-\cite{kaseicache} can be used for this, and the approach
-\cite{papailiou2015graph} took to canonical labelling may be extended
-to address this problem
+Problem~\ref{prob:queryeq} is has an interesting possible solution in
+the application of hashing algorithms that have a reasonable chance of
+returning the same digest if two queries are equivalent. As noted in
+Section~\ref{sec:relcache}, \cite{kaseicache} can be used for this. It
+is also interesting to discuss the approach \cite{papailiou2015graph}
+took to canonical labelling in HTTP context:
+
+The \httph{ETag} header, discussed in Section~\ref{sec:httpcache}, can
+be used for a SPARQL result. Such a canonical label could be used in
+conjunction with an algorithm that tracks changes to the RDF graph to
+produce an \httph{ETag}. For slow-changing Linked Data, which is a
+common case, it is a possibility that a canonical label in combination
+with a time-stamp of the entire RDF Graph could yield considerable
+benefits. However, the presented algorithm does not cover SPARQL in
+its entirety, for example, it does not handle named graphs, solution
+modifiers and filters, nor is it clear how it handles property paths
+(which in many cases should be trivial, since a property path often
+has an equivalent Basic Graph Pattern). Union queries or negation is
+also not mentioned, so more work is needed for this method to have
+general applicability. Nevertheless, if the entire Group Graph Pattern
+of a certain query is supported, and the query projects all variables,
+e.g. it wraps the pattern with \sparql{SELECT *} or
+\sparql{CONSTRUCT}, the \httph{ETag} would be trivial to construct
+using the presented algorithm in a non-trivial number of beneficial
+cases. With this, cached result verification could be done by a
+mediator using HTTP requests with the \httph{If-None-Match} header and
+so make it possible to use parts of the engine of this paper usable in
+the Internet infrastructure.
 

### 2016-07-27 dissertation/related.tex
Mention connected subgraphs in the problems instead
Move etag labelling to future work

@@
 The query planner first examines all connected subgraphs of a full
-SPARQL query, and generates canonical labels for each one. Since it
-insists on caching connected subgraphs, it avoids the problem I
-encountered in Chapter~\ref{sec:tpfcacheplanning} that the cache may
-break up connected subgraphs and thus create Cartesian joins. To
+SPARQL query, and generates canonical labels for each one. To
 further enhance the query planner, the cache can optionally include
@@
 subsequent queries.
-
-The canonical labelling algorithm presented in this paper provides an
-opportunity for further work, when considered in an HTTP context: The
-\httph{ETag} header, discussed in Section~\ref{sec:httpcache}, can be
-used for a SPARQL result. Such a canonical label could be
-used in conjunction with an algorithm that tracks changes to the RDF
-graph to produce an \httph{ETag}. For slow-changing Linked Data, which
-is a common case, it is a possibility that a canonical label in
-combination with a time-stamp of the entire RDF Graph could yield
-considerable benefits. However, the presented algorithm does not cover
-SPARQL in its entirety, for example, it does not handle named graphs,
-solution modifiers and filters, nor is it clear how it handles
-property paths (which in many cases should be trivial, since a
-property path often has an equivalent Basic Graph Pattern). Union
-queries or negation is also not mentioned, so more work is needed for
-this method to have general applicability. Nevertheless, if the entire
-Group Graph Pattern of a certain query is supported, and the query
-projects all variables, e.g. it wraps the pattern with \sparql{SELECT
-  *} or \sparql{CONSTRUCT}, the \httph{ETag} would be trivial to
-construct using the presented algorithm in a non-trivial number of
-beneficial cases. With this, cached result verification could be done
-by a mediator using HTTP requests with the \httph{If-None-Match}
-header and so make it possible to use parts of the engine of this
-paper usable in the Internet infrastructure.
 

### 2016-07-27 dissertation/index.tex
readd draft keyword
Use an example environment
add todo

@@
-\documentclass[a4paper, 12pt]{uiophd}
+\documentclass[a4paper, 12pt, draft]{uiophd}
 \usepackage{graphicx}
@@
 \newtheorem{problem}{Problem}
+\newtheorem{example}{Example}
 
@@
 \begin{abstract}
-  This dissertation discusses a broad range of problems concerning the
+  This dissertation\todo{thesis or dissertation, or no change?} discusses a broad range of problems concerning the
   use of the SPARQL query language on the open, public Web. It is

### 2016-07-27 dissertation/cache-proxy.tex
qualify HERMES

--- 2016-07-23 dissertation/cache-proxy.tex
@@
 the forward proxy, or in a CDN.
+
+It is further interesting to note the contrast to the HERMES
+\cite{adali1996query} system's concept of invariants discussed in
+Section~\ref{sec:relcache}: a SPARQL query cache be oblivious
+to domain knowledge in a CDN deployment and therefore shouldn't assume any invariants,
+except in the possible but unlikely case where the remote server
+declares an infinite freshness lifetime for the result of a certain
+graph pattern.
 

### 2016-07-27 dissertation/cache-proxy.tex
note forward proxy and HERMES
smaller corrections, thanks to @kasei.

@@
 This chapter describes an effort to create a proxy that can cache the
-results of not only SPARQL queries, but also the results of individual
+results of not only complete SPARQL queries, but also the results of individual
 triple patterns. It will asynchronously analyse executed queries, and
@@
 
-The present system may be deployed at any of these levels. However,
+The present system may be deployed at any of these levels (and similar
+figures could be drawn). However,
 since the database and the reverse proxy in front of it may have
@@
 declares an infinite freshness lifetime for the result of a certain
-graph pattern.
+graph pattern. A forward proxy, on the other hand, may have a deeper
+understanding on the needs of the client, and could possibly implement
+invariants. The Attean framework accommodates for this.
 
@@
 are being passed, where the cache is assumed to reside on a
-intermediate caching proxy in the Internet, i.e. a CDN. The figure may
-be viewed as having time on the $x$-axis, with the arrows pointing
+intermediate caching proxy in the Internet, i.e. a CDN. The figure
+may be viewed as having time on the $x$-axis, with the arrows pointing
 down being requests and the arrows point up being responses. In this
@@
 was successful and that the cache can be used to answer request
-\circled{3}. In that case, the proxy will send a rewritten query for
-the rest of the data it needs to be able to answer the query to the origin
-server, indicated by the dashed arrows. This may be a SPARQL query, a
-TPF request or a combination thereof, depending on what the planner
-finds least expensive. The proxy will then evaluate the entire query
-and send the final result back as response \circled{3}. In addition,
-the analyser may have identified another Triple Pattern Fragment that
-it takes as likely to be useful for future requests, and sends a request and
-caches the response \circled{4}. At some point, it may be able to
-answer the entire query using only cached results, as illustrated by
-request and response \circled{5}.
+\circled{3}. In that case, the proxy will send rewritten queries for
+the rest of the data it needs to be able to answer the query to the
+origin server, indicated by the dashed arrows. This may be zero or more SPARQL
+query, TPF requests or a combination thereof, depending on what the
+planner finds least expensive. The proxy will then evaluate the entire
+query and send the final result back as response \circled{3}. In
+addition, the analyser may have identified another Triple Pattern
+Fragment that it takes as likely to be useful for future requests, and
+sends a request and caches the response \circled{4}. At some point, it
+may be able to answer the entire query using only cached results, as
+illustrated by request and response \circled{5}.
 
@@
 the more well-known Jena framework, and the opportunities presented by
-the introduction of traits \cite{traits}, or roles as they are known,
+the introduction of traits \cite{traits}, or roles as they are known
 to the Perl world.
@@
 
-Perl is an interpreted language and is not known to run fast. Instead,
+Perl is an interpreted language and is not focused on speed. Instead,
 the focus is on developer efficiency. However, our motivation for
 Attean, expressed in \cite{williamspushing}, was to enable the use
-of faster implementations by using roles to simplify code reuse.
+of faster implementations by using roles. It also serves to simplify 
+code reuse.
 
@@
 bulkupdateable, can enable the use of ETags or modification times for
-conditional requests (as defined in RFC7232 \cite{rfc7234}, and makes
+conditional requests (as defined in RFC7232 \cite{rfc7234}), and makes
 possible further extensions. Each of these primitives is represented
@@
 \pcode{count\_quads\_estimate}, \pcode{get\_graphs} and
-\pcode{size}. These methods may be overridden when composing a
-class. Composing other roles will require further methods to be
-implemented.
+\pcode{size}. These methods may be overridden with more efficient
+implementations when composing a class. Composing other roles will
+require further methods to be implemented.
 
 Stores can be rather diverse, so in addition, Attean provides another
-abstraction layer, called a Model. Models will consistently have
+abstraction layer, called a Model. Models consistently operate over
 quads, in the case where underlying store is a triple store, a graph
 name will be set, or it may wrap several triple stores with a graph
-name for each. In addition, Models provide higher level methods, like
+name for each. In addition, Models provide higher level methods,
 for example a \pcode{subject} method to list all subjects, and
 corresponding for other terms. Thus, upper layers of an application
-should use the Model. We have also chosen to implement significant
+should use the Model abstraction. We have also chosen to implement significant
 parts of the query planning in the Model.
@@
 \cite{williamspushing}. Additionally, Attean has a simpler mechanism
-for single triple or quad patterns: By default, triple or quad
-patterns are evaluated by calling the store's \pcode{get\_triples} or
-\pcode{get\_quads}, but this may be augmented by adding a wrapper
-around an \pcode{access\_plans} method used by the query planner to
-produce further plans.
+for single quad patterns: By default, quad patterns are evaluated by
+calling the Model's \pcode{get\_quads}, but this may be augmented by
+adding a wrapper around an \pcode{access\_plans} method used by the
+query planner to produce further plans.
 
@@
 
-The cache itself is assumed to only cache triples where at least one
-term is bound. I have used the Redis data structure store for the
+The cache itself is assumed to only cache the results of triple
+patterns where at least one term is bound. 
+I have used the Redis data structure store for the
 cache as well. This choice is somewhat arbitrary, and has not proved

### 2016-07-27 dissertation/cache-proxy.tex
triple store -> triplestore, thanks @kasei

@@
 \cite{ldf1}. The system does not support named graphs (i.e. only
-triple stores), but other than that, it supports SPARQL 1.1 in its
+triplestores), but other than that, it supports SPARQL 1.1 in its
 entirety. It was possible to develop this quickly by the efforts put
@@
 abstraction layer, called a Model. Models consistently operate over
-quads, in the case where underlying store is a triple store, a graph
-name will be set, or it may wrap several triple stores with a graph
+quads, in the case where underlying store is a triplestore, a graph
+name will be set, or it may wrap several triplestores with a graph
 name for each. In addition, Models provide higher level methods,
@@
 Graph Pattern optimisation from \cite{verborgh2014querying}. Included with the client
-code is an implementation of a triple store of the legacy
+code is an implementation of a triplestore of the legacy
 \pmodule{RDF::Trine} framework that predates Attean. He also started

### 2016-07-27 dissertation/cache-proxy.tex
Add mboxes around some words and code to not hyphenate them

@@
 Protocol client that implements Attean APIs. The store implementation
-itself composes the TripleStore API and implements methods to retrieve
+itself composes the \mbox{TripleStore} API and implements methods to retrieve
 the results of a single triple pattern and exact cardinality for a
@@
 \pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
-model implementation that composes the Model and CostPlanner
+model implementation that composes the Model and \mbox{CostPlanner}
 APIs. This contains a \pcode{cost\_for\_plan} implementation that will
@@
 separate module, but that module was subsequently adopted by me, and I
-have written most of the functionality. It composes the TripleStore
-and CostPlanner APIs and implements triple pattern queries and
+have written most of the functionality. It composes the \mbox{TripleStore}
+and \mbox{CostPlanner} APIs and implements triple pattern queries and
 cardinality estimates, both provided by the underlying client code as

### 2016-07-28 dissertation/cache-proxy.tex
Add some substructure Add appendix and move modules and stuff there
Note that @kasei would have been a co-author of a paper

@@
 may prefetch the results of certain triple patterns into the cache.
+
+\subsection{Scenario}
 
@@
 
+\subsection{Key Problems}
+
 \section{Implementation}\label{sec:impl}
@@
 published to the Comprehensive Perl Archive Network as Free Software,
-and most of which have become a part the Free Software ecosystem. To
-install the system on the top of a Debian system with only essential
-packages requires 146 Debian packages. Running the test suite requires
-even more. As such, it builds on the code of hundreds of authors, too
-many to enumerate. I shall constrain myself to list the modules that
-have code that has been motivated by this study. Modules that have
-other primary authors than myself have been started prior to this
-present project, but may have substantial contributions from this
-project. The Attean framework was started as part of the effort in
-Section~\ref{sec:conpush}, but even though it has been developed
-further with the requirements that arose from this project and has
-contributions from me, the vast majority of the code has been written
-by Gregory Todd Williams. The main work in this project has gone into
-the module \pmodule{AtteanX::Query::Cache}, where I have authored the
-vast majority of the code. The relative difference in authorship can
-be further examined by checking the linked Github repositories, or by
-using git2prov\footnote{See \url{http://git2prov.org/}} to create RDF
-data of the commit history. Table~\ref{tab:modules} shows the modules
-that have been influenced or written as part of this project.
-
-\begin{table}
-\caption{The modules that enable the query cache}\label{tab:modules}
-\begin{tabular}{ | l | p{3cm} | l |}
-  \hline
-  Module & Authors & Github URL \\ \hline
-
-  \pmodule{AtteanX::Query::Cache} & Kjetil Kjernsmo, Gregory Todd Williams &
-  \url{kjetilk/p5-atteanx-query-cache} \\ % 0.001_04
-
-  \pmodule{AtteanX::Store::SPARQL} & Kjetil Kjernsmo &
-  \url{kjetilk/p5-atteanx-store-sparql} \\ % 0.008
-
-  \pmodule{AtteanX::Store::LDF} & Kjetil Kjernsmo, Patrick Hochstenbach &
-  \url{phochste/AtteanX-Store-LDF} \\ % 0.02
-
-  \pmodule{LWP::UserAgent::CHICaching} & Kjetil Kjernsmo &
-  \url{kjetilk/p5-lwp-useragent-chicaching} \\ % 0.04
-
-  \pmodule{Attean} & Gregory Todd Williams, Kjetil Kjernsmo &
-  \url{kasei/attean} \\ % 0.015
-
-  \pmodule{AtteanX::Endpoint} & Gregory Todd Williams &
-  \url{kasei/atteanx-endpoint} \\ % 0.001
-
-  \pmodule{RDF::LDF} &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
-  Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\ % 0.17
-
-  \hline
-\end{tabular}
-\end{table}
-\todo{make it fit. Rotate?}
-
-The system also depends on modules I have written that are not part of
-the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
-\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.
-
-To further understand the roles of the different modules, note the
-following: \pmodule{LWP::UserAgent::CHICaching} is a traits-based
-\cite{traits} implementation of the majority of RFC7234.
-The underlying caching framework used by this module, \pmodule{CHI},
-will take care of expiry and purging of entries, since this module
-will supply an explicit freshness lifetime. 
-\pmodule{AtteanX::Endpoint} is an implementation of the server side of
-SPARQL 1.1 Protocol. The system is further detailed in the following sections:
+and most of which have become a part the Free Software ecosystem. For
+a detailed account of the modules that are a part of the system and
+the authors that have contributed code in this ecosystem, see
+Appendix~\ref{app:modules}.
 
@@
 
+\begin{subappendices}
+%\renewcommand{\setthesection}{\Alph{section}}
+%\renewcommand{\setthesubsection}{\Alph{subsection}}
+
+%\section{Appendix}
+\section{Appendix: Modules and Their Contributors}\label{app:modules}
+
+To install the system on the top of a Debian system with only
+essential packages requires 146 Debian packages. Running the test
+suite requires even more. As such, it builds on the code of hundreds
+of authors, too many to enumerate. I shall constrain myself to list
+the modules that have code that has been motivated by this
+study. Modules that have other primary authors than myself have been
+started prior to this present project, but may have substantial
+contributions from this project. The Attean framework was started as
+part of the effort in Section~\ref{sec:conpush}, but even though it
+has been developed further with the requirements that arose from this
+project and has contributions from me, the vast majority of the code
+has been written by Gregory Todd Williams, who would have been a
+co-author of this work had it been published in the peer-reviewed
+literature. The main work in this project has gone into the module
+\pmodule{AtteanX::Query::Cache}, where I have authored the vast
+majority of the code. The relative difference in authorship can be
+further examined by checking the linked Github repositories, or by
+using git2prov\footnote{See \url{http://git2prov.org/}} to create RDF
+data of the commit history. Table~\ref{tab:modules} shows the modules
+that have been influenced or written as part of this project.
+
+\begin{table}
+\caption{The modules that enable the query cache}\label{tab:modules}
+\begin{tabular}{ | l | p{3cm} | l |}
+  \hline
+  Module & Authors & Github URL \\ \hline
+
+  \pmodule{AtteanX::Query::Cache} & Kjetil Kjernsmo, Gregory Todd Williams &
+  \url{kjetilk/p5-atteanx-query-cache} \\ % 0.001_04
+
+  \pmodule{AtteanX::Store::SPARQL} & Kjetil Kjernsmo &
+  \url{kjetilk/p5-atteanx-store-sparql} \\ % 0.008
+
+  \pmodule{AtteanX::Store::LDF} & Kjetil Kjernsmo, Patrick Hochstenbach &
+  \url{phochste/AtteanX-Store-LDF} \\ % 0.02
+
+  \pmodule{LWP::UserAgent::CHICaching} & Kjetil Kjernsmo &
+  \url{kjetilk/p5-lwp-useragent-chicaching} \\ % 0.04
+
+  \pmodule{Attean} & Gregory Todd Williams, Kjetil Kjernsmo &
+  \url{kasei/attean} \\ % 0.015
+
+  \pmodule{AtteanX::Endpoint} & Gregory Todd Williams &
+  \url{kasei/atteanx-endpoint} \\ % 0.001
+
+  \pmodule{RDF::LDF} &  Patrick Hochstenbach, Gregory Todd Williams, Jakob Voß,
+  Kjetil Kjernsmo & \url{phochste/RDF-LDF} \\ % 0.17
+
+  \hline
+\end{tabular}
+\end{table}
+\todo{make it fit. Rotate?}
+
+The system also depends on modules I have written that are not part of
+the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
+\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.
+
+To further understand the roles of the different modules, note the
+following: \pmodule{LWP::UserAgent::CHICaching} is a traits-based
+\cite{traits} implementation of the majority of RFC7234.
+The underlying caching framework used by this module, \pmodule{CHI},
+will take care of expiry and purging of entries, since this module
+will supply an explicit freshness lifetime. 
+\pmodule{AtteanX::Endpoint} is an implementation of the server side of
+SPARQL 1.1 Protocol.
+
+\end{subappendices}
+
 %%  LocalWords:  canonicalised

### 2016-07-28 dissertation/cache-proxy.tex
move the description of the clients
reword

@@
 the authors that have contributed code in this ecosystem, see
-Appendix~\ref{app:modules}.
+Appendix~\ref{app:modules}. Some parts are essential for the
+discussion and to understand the contributions, and therefore detailed
+in the following:
 
@@
 Section~\ref{sec:implqueryplan}.
-
-\subsection{SPARQL Protocol Client}\label{sec:sparqlclient}
-
-The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
-Protocol client that implements Attean APIs. The store implementation
-itself composes the \mbox{TripleStore} API and implements methods to retrieve
-the results of a single triple pattern and exact cardinality for a
-given triple pattern, by using an aggregate query. In addition, it can
-generate plans for Basic Graph Pattern algebra objects that have more
-than one triple pattern. If that is the case, it will return an
-instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
-also defined by the module. If it is not the case, the module also has
-an implementation of the \pcode{access\_plan} method in a
-\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
-composed by a query planner to provide a single triple pattern
-\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
-model implementation that composes the Model and \mbox{CostPlanner}
-APIs. This contains a \pcode{cost\_for\_plan} implementation that will
-provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
-proportional to the number of triple patterns in the plan, but
-penalises plans that do not have triple patterns that are connected
-through a variable (i.e. will cause a Cartesian join) with a factor
-10. See Section~\ref{sec:costheuristics} for details on the cost
-heuristics.
-
-\subsection{Linked Data Fragment Client}\label{sec:ldfclient}
-
-The Linked Data Fragment client work was started prior to this project
-by Patrick Hochstenbach, and consists of the client code for Triple
-Pattern Fragments in \pmodule{RDF::LDF}. It also includes the Basic
-Graph Pattern optimisation from \cite{verborgh2014querying}. Included with the client
-code is an implementation of a triplestore of the legacy
-\pmodule{RDF::Trine} framework that predates Attean. He also started
-an Attean store implementation \pmodule{AtteanX::Store::LDF} as a
-separate module, but that module was subsequently adopted by me, and I
-have written most of the functionality. It composes the \mbox{TripleStore}
-and \mbox{CostPlanner} APIs and implements triple pattern queries and
-cardinality estimates, both provided by the underlying client code as
-they are a part of the core functionality of TPF,
-as well as \pcode{cost\_for\_plan}.
-
-It also supplies a plan class \pmodule{AtteanX::Plan::LDF::Triple}
-used in the query planning.
 
@@
 
-To install the system on the top of a Debian system with only
-essential packages requires 146 Debian packages. Running the test
+Installing the system on the top of a Debian system with only
+essential packages require 146 Debian packages. Running the test
 suite requires even more. As such, it builds on the code of hundreds
@@
 \pmodule{AtteanX::Endpoint} is an implementation of the server side of
-SPARQL 1.1 Protocol.
+SPARQL 1.1 Protocol. Another two modules require further details:
+
+\subsection{SPARQL Protocol Client}\label{sec:sparqlclient}
+
+The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
+Protocol client that implements Attean APIs. The store implementation
+itself composes the \mbox{TripleStore} API and implements methods to retrieve
+the results of a single triple pattern and exact cardinality for a
+given triple pattern, by using an aggregate query. In addition, it can
+generate plans for Basic Graph Pattern algebra objects that have more
+than one triple pattern. If that is the case, it will return an
+instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
+also defined by the module. If it is not the case, the module also has
+an implementation of the \pcode{access\_plan} method in a
+\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
+composed by a query planner to provide a single triple pattern
+\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
+model implementation that composes the Model and \mbox{CostPlanner}
+APIs. This contains a \pcode{cost\_for\_plan} implementation that will
+provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
+proportional to the number of triple patterns in the plan, but
+penalises plans that do not have triple patterns that are connected
+through a variable (i.e. will cause a Cartesian join) with a factor
+10. See Section~\ref{sec:costheuristics} for details on the cost
+heuristics.
+
+\subsection{Linked Data Fragment Client}\label{sec:ldfclient}
+
+The Linked Data Fragment client work was started prior to this project
+by Patrick Hochstenbach, and consists of the client code for Triple
+Pattern Fragments in \pmodule{RDF::LDF}. It also includes the Basic
+Graph Pattern optimisation from \cite{verborgh2014querying}. Included with the client
+code is an implementation of a triplestore of the legacy
+\pmodule{RDF::Trine} framework that predates Attean. He also started
+an Attean store implementation \pmodule{AtteanX::Store::LDF} as a
+separate module, but that module was subsequently adopted by me, and I
+have written most of the functionality. It composes the \mbox{TripleStore}
+and \mbox{CostPlanner} APIs and implements triple pattern queries and
+cardinality estimates, both provided by the underlying client code as
+they are a part of the core functionality of TPF,
+as well as \pcode{cost\_for\_plan}.
+
+It also supplies a plan class \pmodule{AtteanX::Plan::LDF::Triple}
+used in the query planning.
 

### 2016-07-28 dissertation/cache-proxy.tex
use same prefix

@@
 the authors that have contributed code in this ecosystem, see
-Appendix~\ref{app:modules}. Some parts are essential for the
+Appendix~\ref{sec:modules}. Some parts are essential for the
 discussion and to understand the contributions, and therefore detailed
@@
 planner evaluates a query into the above cache, using a trivial
-extension to the plan class in Section~\ref{sec:ldfclient}.
+extension to the plan class packaged with the LDF client code
+described in Appendix~\ref{sec:ldfclient}.
 
@@
 \pmodule{AtteanX::Query::AccessPlan::Cache}, out of which the three
-first are a part of the Attean framework, the next was described in
-Section~\ref{sec:sparqlclient} and the last is a part of the
+first are a part of the Attean framework, the next is described in
+Appendix~\ref{sec:sparqlclient} and the last is a part of the
 query cache system itself.
@@
 %\section{Appendix}
-\section{Appendix: Modules and Their Contributors}\label{app:modules}
+\section{Appendix: Modules and Their Contributors}\label{sec:modules}
 

### 2016-07-28 dissertation/cache-proxy.tex
start of an planning elaboration

@@
 
-\subsection{Key Problems}
+\subsection{Understanding the Query Planning}
+
+We saw in Section~\ref{sec:introinteract} that a query would be parsed
+to a tree of algebra objects and then a query planner would traverse
+the algebra tree to create a plan tree by finding all the ways the
+data can be accessed, or operations executed over them. For example,
+if the input query has a BGP with two triple patterns and there are
+two types of join operations, it may result in four trees, where each
+of the join operation plans are paired with both access plans for the
+triple patterns.
+
+It is next the query planner's task to determine which of the possible
+plans it should execute. A common choice to perform this task is to
+estimate the cost executing various choices.
+
+The system we propose is based on this paradigm, any choices we make
+are therefore to be encoded into a cost model. The cost model itself
+is not considered a part of the contributions, it is a simple cost
+model based on heuristics that arises from the following
+discussion. The heuristics themselves are detailed in
+Section~\ref{sec:costheuristics}.
+
+Recall Problem~\ref{prob:cachecartesian}, where a BGP with three
+triple patterns could break into two components if the triple pattern
+connecting the two others was cached. It is one of the requirements
+for the heuristics of the cost model that it is be able to avoid such
+situations.
+
+As mentioned, the proposed query planner takes may access a remote
+SPARQL endpoint and a Linked Data Fragments Server as well as the
+local cache. I have also said that a key motivation behind the study
+is to relieve the remote endpoint of load. It was argued in
+\cite{verborgh2014querying} that LDF shifts the burden of query
+evaluation to the client, and so relieves the server of load. It is
+not clear from the literature that this strikes the right balance, it
+is my belief that the balance must depend on the concrete server and
+client capabilities, keeping in mind
+Problem~\ref{prob:microcontroller}. Future work to determine the right
+balance in a given situation would be welcome in this area, but I
+believe it would be advantageous to build on a cost model.
+
+Following this discussion, consider this example:
+\begin{example}{A four-triple pattern BGP}
+\begin{verbatim}
+?s <http://example.org/m/p> "1" ;
+   <http://example.org/m/q> <http://example.org/m/a> ;
+   <http://example.org/m/p> ?o .
+?o <http://example.org/m/b> "2" .
+\end{verbatim}
+\end{example}
+Say that the first triple pattern has a result in the cache. This will
+not cause Problem~\ref{prob:cachecartesian} to occur (it would have
+occurred if the third triple pattern had a cached result, however), so
+we can safely use the cache. The results of the three remaining triple
+patterns must be fetched from a remote server. There are essentially
+two reasonable choices: Either fetch the result from an LDF server as
+argued by \cite{verborgh2014querying} and join them locally or send a
+single, rewritten SPARQL query to the remote endpoint for the
+evaluation. When the results are returned from both the cache and the
+remote endpoint, they must then finally be joined to produce the final
+result returned to the client. A string representation of the plan
+tree from the current implementation reflects this and reads:
+\begin{example}{An example Attean plan}
+\begin{verbatim}
+- Hash Join { s } (distinct cost: 74)
+-   SPARQLBGP (cost: 60)
+-     Quad { ?s, <http://example.org/m/p>, ?o, <http://test.invalid/graph> }
+-     Quad { ?o, <http://example.org/m/b>, "2", <http://test.invalid/graph> }
+-     Quad { ?s, <http://example.org/m/q>, <http://example.org/m/a>, <http://test.invalid/graph> }
+-   Iterator (?s with 2 elements) (cost: 2)
+\end{verbatim}
+\end{example}
 

### 2016-07-28 dissertation/cache-proxy.tex
add simple example

@@
 estimate the cost executing various choices.
+
+The resulting plan trees may be represented as strings as follows:
+\begin{example}{Four Attean plan trees}
+\begin{verbatim}
+- Hash Join { s } (distinct cost: 27)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+
+- Hash Join { s } (distinct cost: 28)
+-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+
+- NestedLoop Join (distinct cost: 180)
+-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+
+- NestedLoop Join (distinct cost: 180)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+\end{verbatim}
+\end{example}
+In the above example, the first plan tree will be returned from the
+query planner for execution, since it has the lowest estimated cost
+(27).
 

### 2016-07-28 dissertation/cache-proxy.tex
small change
more explanation

@@
 result returned to the client. A string representation of the plan
-tree from the current implementation reflects this and reads:
-\begin{example}{An example Attean plan}
+tree from the current implementation reflects this and reads (note
+that accessing the cache is represented by an iterator plan):
+\begin{example}{An example Attean plan with a remotely executed BGP}
 \begin{verbatim}

### 2016-07-28 dissertation/cache-proxy.tex
describe join rotation

@@
 query planner for execution, since it has the lowest estimated cost
-(27).
+(27). Since the SPARQL engine operates over quads, we have assigned a
+dummy graph name when constructing the Model, namely
+\rdfterm{<http://example.org/>}.
 
@@
 Following this discussion, consider this example:
-\begin{example}{A four-triple pattern BGP}
+\begin{example}{A four-triple pattern BGP}\label{ex:bgp4}
 \begin{verbatim}
@@
 
+Merging the quad plans into a single SPARQL BGP plan is not
+straightforward in the presence of a quad pattern that is to be looked
+up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
+quad patterns\todo{correctly expressed?}  in the order in
+Example~\ref{ex:bgp4}, the initial join tree produced by parsing may
+be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
+a plan on the form $A \bowtie BCD$ (all commutations will be
+generated). To facilitate this, we employ \emph{join rotation}. A
+simple example of join rotation would be this: We first generate
+subplans for both $((A \bowtie B )\bowtie C) \bowtie D) $ and $ (A
+\bowtie B )\bowtie (C \bowtie D) $. Then, $C$ and $D$ can be merged to
+a SPARQL BGP to $ (A \bowtie B )\bowtie CD $. Another rotation can
+then be performed: $ A \bowtie (B \bowtie CD) $ and then coalesced to
+the final result. The planner will generate alternative plans for both
+rotated and unrotated plans, and cost is estimated for both, so we
+have to be careful to design the cost model to allow the rotation to
+result in SPARQL BGP plans.
+
 \section{Implementation}\label{sec:impl}

### 2016-07-28 dissertation/cache-proxy.tex
space

@@
 up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
-quad patterns\todo{correctly expressed?}  in the order in
+quad patterns \todo{correctly expressed?} in the order in
 Example~\ref{ex:bgp4}, the initial join tree produced by parsing may

### 2016-07-28 dissertation/cache-proxy.tex
Note IDP

@@
 
+As can be imagined from the fact that a BGP with just to triple
+patterns result in 4 plan trees, a naive query planner may generate a
+large number of plans for complex queries. In addition, the proposed
+system will generate even more plans, since it considers three sources
+of the same data, the cache, the remote endpoint and an LDF
+server. Various methods have been developed in the literature to limit
+the growth of plans. In addition to a naive planner, we also have an
+Iterative Dynamic Programming planner
+\cite{Kossmann:2000:IDP:352958.352982}, which prunes plans
+aggressively based on their estimated cost in each iteration of query
+plan generation.
+
 \section{Implementation}\label{sec:impl}
@@
 possible plans and then estimate the cost of all plans, and finally
-return the best 5. \pmodule{Attean::API::IDPJoinPlanner}
-implements an Iterative Dynamic Programming planner
-\cite{Kossmann:2000:IDP:352958.352982}, which prunes plans
-aggressively based on their estimated cost in each iteration of query
-plan generation. When a final plan tree has been found, the query may
-be executed.
+return the best 5. \pmodule{Attean::API::IDPJoinPlanner} implements an
+Iterative Dynamic Programming planner
+\cite{Kossmann:2000:IDP:352958.352982}. When a final plan tree has
+been found, the query may be executed.
 

### 2016-07-28 dissertation/cache-proxy.tex
large cleanup of join rotation

@@
 
-\subsection{Understanding the Query Planning}
+\subsection{Understanding the Query Planning}\label{sec:understandquery}
 
@@
 be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
-a plan on the form $A \bowtie BCD$ (all commutations will be
-generated). To facilitate this, we employ \emph{join rotation}. A
-simple example of join rotation would be this: We first generate
-subplans for both $((A \bowtie B )\bowtie C) \bowtie D) $ and $ (A
-\bowtie B )\bowtie (C \bowtie D) $. Then, $C$ and $D$ can be merged to
-a SPARQL BGP to $ (A \bowtie B )\bowtie CD $. Another rotation can
-then be performed: $ A \bowtie (B \bowtie CD) $ and then coalesced to
-the final result. The planner will generate alternative plans for both
-rotated and unrotated plans, and cost is estimated for both, so we
-have to be careful to design the cost model to allow the rotation to
-result in SPARQL BGP plans.
+a plan on the form $A \bowtie BCD$, i.e. $BCD$ is reduced to a more
+fundamental plan type, in this example a SPARQL BGP. This will
+minimise the number of HTTP requests, in much the same way as the
+exclusive groups of
+\cite{springerlink:10.1007/978-3-642-25073-6-38}. To facilitate this,
+we employ \emph{join rotation}. A simple example of join rotation
+would be this: We first generate subplans for both $((A \bowtie B
+)\bowtie C) \bowtie D) $ and $ (A \bowtie B )\bowtie (C \bowtie D)
+$. Then, $C$ and $D$ can be merged to a SPARQL BGP to $ (A \bowtie B
+)\bowtie CD $. Another rotation can then be performed: $ A \bowtie (B
+\bowtie CD) $ and then coalesced to the final result. The planner will
+generate alternative plans for both rotated and unrotated plans, and
+all generated plans are subject to cost estimation, so whether a large
+BGP plan will be evaluated remotely, or as a join of local cached
+results and a smaller remote BGP, or a remote Triple Pattern Fragments
+server, is up to the cost estimation. This will be discussed in detail
+in Section~\ref{sec:costheuristics}.
 
@@
 
-It is interesting to detail the function of
-\pmodule{AtteanX::API::JoinRotatingPlanner}. Its main function is to
-produce alternatives for join query plans. For example, if we have
-three quad patterns $A$, $B$ and $C$, plans for both $ ( A \bowtie B)
-\bowtie C $ and $ A \bowtie ( B \bowtie C ) $ will be generated. In
-some cases, it may be possible to reduce a join to a more fundamental
-plan type, and this may be done next. In our case, this is done in
-\pmodule{AtteanX::QueryPlanner::Cache}, where adjacent quad patterns
-are coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
+The implementation of the join rotation detailed in
+Section~\ref{sec:understandquery} is in the role
+\pmodule{AtteanX::API::JoinRotatingPlanner}. In the planner
+\pmodule{AtteanX::QueryPlanner::Cache} adjacent quad patterns are
+coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
 coalesced into larger BGP plans, so that they can be evaluated by the
-remote SPARQL endpoint as a single unit. This will minimise the number
-of HTTP requests, in much the same way as the exclusive groups of
-\cite{springerlink:10.1007/978-3-642-25073-6-38}. Note, however, that
-the generated plans are subject to cost estimation, so whether a large
-BGP plan will be evaluated remotely, or as a join of local cached
-results and a smaller remote BGP, or in the case of
-\pmodule{AtteanX::QueryPlanner::Cache::LDF}, a remote Triple Pattern
-Fragments server, is up to the cost estimation. This will be discussed
-in detail in Section~\ref{sec:costheuristics}.
+remote SPARQL endpoint as a single unit. The extension
+\pmodule{AtteanX::QueryPlanner::Cache::LDF} adds the option of
+querying remote LDF servers.
 

### 2016-07-28 dissertation/cache-proxy.tex
downplay cost heuristics
comment we can't eval
remove comma

@@
 model based on heuristics that arises from the following
-discussion. The heuristics themselves are detailed in
+discussion. The heuristics themselves are best considered an
+implementation detail and is described in
 Section~\ref{sec:costheuristics}.
@@
 variables, to mildly penalise Cartesian joins. If the plan is a hash
-join the difference is that, we add rather than multiply the cost of
+join the difference is that we add rather than multiply the cost of
 the of the subplans, and multiply by a factor of 100 rather than 10 so
@@
 plans, by multiplying the cost of the join plan by the number of the
-Triple Pattern Fragment plans in the tree.
+Triple Pattern Fragment plans in the tree. Whether this is a
+reasonable choice would have been a subject of the planned evaluation.
 

### 2016-07-28 dissertation/cache-proxy.tex
mark example

@@
 query where it does not:
+\begin{example}{Query with results of 2 triple patterns cached and
+    unconnected triple pattern.}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 Note that the above Basic Graph Pattern contains a triple pattern that

### 2016-07-28 dissertation/cache-proxy.tex
different ref

@@
 
-As noted in Section~\ref{sec:implqueryplan}, we rely on rotating joins
+As noted in Section~\ref{sec:understandquery}, we rely on rotating joins
 to gather triple pattern plans into a larger structure that can be

### 2016-07-28 dissertation/cache-proxy.tex
ref examples and problems

@@
 subplans, or 1000 if there is not. This is to prevent the remote
-endpoint from having to compute Cartesian joins as much as possible. A
+endpoint from having to compute Cartesian joins as much as possible,
+thus addressing Problem~\ref{prob:cachecartesian}. A
 remote SPARQL query with only one triple pattern is assumed to always
@@
 plans, by multiplying the cost of the join plan by the number of the
-Triple Pattern Fragment plans in the tree. Whether this is a
-reasonable choice would have been a subject of the planned evaluation.
+Triple Pattern Fragment plans in the tree. Example~\ref{ex:bgp4} is an
+example of a situation where this heuristic ensures a larger
+BGP. Whether this is a reasonable choice would have been a subject of
+the planned evaluation.
 

### 2016-07-28 dissertation/cache-proxy.tex
Martin's proofreading

@@
 certainty. This chapter will detail the system, show its design and
-features and discuss its failures.
+features and discuss its strengths and weaknesses.
 
@@
 may use conventional database caching techniques. The database cache
-and reverse proxy is typically controlled by the data provider.
+and reverse proxy are typically controlled by the data provider.
 
@@
 Section~\ref{sec:analpre}. The response \circled{2} is not sent back
-to the client, but will enter the cache. Lets say that the analyser
+to the client, but will enter the cache. Assuming that the analyser
 was successful and that the cache can be used to answer request
@@
 community gained with \pmodule{RDF::Trine}, the Perl counterpart to
-the more well-known Jena framework, and the opportunities presented by
+the better known Jena framework, and the opportunities presented by
 the introduction of traits \cite{traits}, or roles as they are known
@@
 bulkupdateable, can enable the use of ETags or modification times for
-conditional requests (as defined in RFC7232 \cite{rfc7234}), and makes
-possible further extensions. Each of these primitives is represented
+conditional requests (as defined in RFC7232 \cite{rfc7234}), and
+enables further extensions. Each of these primitives is represented
 by a role that implements default functionality or simply requires
@@
 a single triple pattern SPARQL query if not. The current
-implementation assumes that both and LDF server and a SPARQL endpoint
+implementation assumes that both an LDF server and a SPARQL endpoint
 can be used to answer any triple pattern. Relaxing this assumption
-depends on that the LDF specifications gain stronger semantics to link
+depends on the LDF specifications gaining stronger semantics to link
 SPARQL endpoints and LDF servers together. With that, the assumption
@@
 
-I decided to set the entire system up as a proxy server, and run a
+I decided to set up the entire system as a proxy server, and run a
 realistic query against DBpedia repeatedly. The intention was to let
@@
 The cache was kept in a Redis data structure database on a machine
-with a Intel Core2 Duo E8500 CPU with 16 GB of RAM. The proxy server
-with the query engine ran on a Intel Xeon CPU E3-1230 v5 CPU with 32
+with an Intel Core2 Duo E8500 CPU with 16 GB of RAM. The proxy server
+with the query engine ran on an Intel Xeon CPU E3-1230 v5 CPU with 32
 GB of RAM, residing a LAN with a gigabit Ethernet link between
@@
 correspond to the usage of
-\pmodule{AtteanX::QueryPlanner::Cache::LDF}. Additionally, I noted
-that downloading the triple pattern fragment
-\sparql{?s~foaf:name~?name~.}  
-resulted in more than 40~000 HTTP requests due to
-paging. Another relevant factor might be to deploy a Triple Pattern
-Fragments server that does not page, and have levels with paging on
-and off. Integrating this factor in the experimental design would
-further complicate it.
+\pmodule{AtteanX::QueryPlanner::Cache::LDF}. 
+%Additionally, I noted
+%that downloading the triple pattern fragment
+%\sparql{?s~foaf:name~?name~.}  
+%resulted in more than 40~000 HTTP requests due to
+%paging. Another relevant factor might be to deploy a Triple Pattern
+%Fragments server that does not page, and have levels with paging on
+%and off. Integrating this factor in the experimental design would
+%further complicate it.
 
@@
 the factor in the first point with level~1 as the cache being enabled,
-then we can examine the experiment similarly to what with did with the
+then we can examine the experiment similarly to what we did with the
 ``Implement'' factor in \cite{kjernsmo_doe_intro}. That is, the first
@@
 then optimise using a bridge to first and foremost the C family of
-languages as needed. Importantly, with the width of the scope, I only
+languages as needed. Importantly, with the breadth of the scope, I only
 had some months to implement the query engine. This is not a software
@@
 
-The profiling points out a clearly where large gains could be
+The profiling points out clearly where large gains could be
 made. First and foremost, parsing must be avoided, and that could be

### 2016-07-29 dissertation/cache-proxy.tex
Add arch section
more on architecture

@@
 
+\subsection{Architecture}\label{sec:arch}
+
+The architecture is depicted in Figure~\ref{fig:arch}. A SPARQL
+protocol implementation will accept a client's query (and could cache
+it's result in its entirety, but this is neither implemented nor
+depicted). Two things happen to a valid query:
+
+\begin{itemize}
+\item The query (or indeed the algebra) is sent asynchronously to a
+  set of analysers that is tasked with predicting what parts of the
+  incoming query may be useful to answer future queries. The result of
+  the analysis is sent to a prefetcher that will decide how to fetch
+  the needed data, currently it only operates on triple patterns, and
+  therefore, and since it is also assumed that a triple pattern
+  fragments server can answer any triple pattern, it instructs a TPF
+  client to fetch and enter the result into the cache.  
+\item The SPARQL protocol implementation will also send the query to
+  a parser that will create an algebra tree from it. The algebra tree
+  is then passed to a query engine for immediate execution. The query
+  engine has an elaborate query planner that can create plans for
+  accessing a cache, a TPF server and a SPARQL Endpoint. When the
+  planner finishes planning, the evaluator may use a TPF client or a
+  SPARQL protocol client to fetch data that is not in the cache. The
+  results of single triple patterns may be inserted in the cache by
+  the TPF client.
+\end{itemize}
+
+\begin{figure}[bh]
+\begin{center}
+\includegraphics{architecture.pdf}
+\caption{Architectural overview. The proposed system is represented
+  inside the box titled ``Proxy''. It interacts with a client, a
+  Triple Patterns Fragments Server and a SPARQL Endpoint. The most
+  important component is the query engine, which consists of a query
+  planner and a query evaluator.}\label{fig:arch}
+\end{center}
+\end{figure}
+
 \section{Implementation}\label{sec:impl}

### 2016-07-29 dissertation/cache-proxy.tex
move on arch
fix placement
simplify a bit

@@
 interchangeably.
+
+\subsection{Architecture}\label{sec:arch}
+
+The architecture is depicted in Figure~\ref{fig:arch}. A SPARQL
+protocol implementation will accept a client's query (and could cache
+it's result in its entirety, but this is neither implemented nor
+depicted). Two things happen to a valid query:
+
+\begin{itemize}
+\item The query (or indeed the algebra) is sent asynchronously to a
+  set of analysers that is tasked with predicting what parts of the
+  incoming query may be useful to answer future queries. The result of
+  the analysis is sent to a prefetcher that will decide how to fetch
+  the needed data, currently it only operates on triple patterns, and
+  therefore, and since it is also assumed that a triple pattern
+  fragments server can answer any triple pattern, it instructs a TPF
+  client to fetch and enter the result into the cache.  
+\item The SPARQL protocol implementation will also send the query to
+  a parser that will create an algebra tree from it. The algebra tree
+  is then passed to a query engine for immediate execution. The query
+  engine has an elaborate query planner that can create plans for
+  accessing a cache, a TPF server and a SPARQL Endpoint. When the
+  planner finishes planning, the evaluator may use a TPF client or a
+  SPARQL protocol client to fetch data that is not in the cache. The
+  results of single triple patterns may be inserted in the cache by
+  the TPF client.
+\end{itemize}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{architecture.pdf}
+\caption{Architectural overview. The proposed system is represented
+  inside the box titled ``Proxy''. It interacts with a client, a
+  Triple Patterns Fragments Server and a SPARQL Endpoint. The most
+  important component is the query engine, which consists of a query
+  planner and a query evaluator.}\label{fig:arch}
+\end{center}
+\end{figure}
 
@@
 
-\subsection{Architecture}\label{sec:arch}
-
-The architecture is depicted in Figure~\ref{fig:arch}. A SPARQL
-protocol implementation will accept a client's query (and could cache
-it's result in its entirety, but this is neither implemented nor
-depicted). Two things happen to a valid query:
-
-\begin{itemize}
-\item The query (or indeed the algebra) is sent asynchronously to a
-  set of analysers that is tasked with predicting what parts of the
-  incoming query may be useful to answer future queries. The result of
-  the analysis is sent to a prefetcher that will decide how to fetch
-  the needed data, currently it only operates on triple patterns, and
-  therefore, and since it is also assumed that a triple pattern
-  fragments server can answer any triple pattern, it instructs a TPF
-  client to fetch and enter the result into the cache.  
-\item The SPARQL protocol implementation will also send the query to
-  a parser that will create an algebra tree from it. The algebra tree
-  is then passed to a query engine for immediate execution. The query
-  engine has an elaborate query planner that can create plans for
-  accessing a cache, a TPF server and a SPARQL Endpoint. When the
-  planner finishes planning, the evaluator may use a TPF client or a
-  SPARQL protocol client to fetch data that is not in the cache. The
-  results of single triple patterns may be inserted in the cache by
-  the TPF client.
-\end{itemize}
-
-\begin{figure}[bh]
-\begin{center}
-\includegraphics{architecture.pdf}
-\caption{Architectural overview. The proposed system is represented
-  inside the box titled ``Proxy''. It interacts with a client, a
-  Triple Patterns Fragments Server and a SPARQL Endpoint. The most
-  important component is the query engine, which consists of a query
-  planner and a query evaluator.}\label{fig:arch}
-\end{center}
-\end{figure}
-
 \section{Implementation}\label{sec:impl}
+
+This section explores the implementation of the general architecture
+outlined in Section~\ref{sec:arch}.
 
@@
 
-When a query is parsed, algebra objects are generated. Based on the
-algebra tree, models may generate plans for any algebra object of
+Based on the algebra tree from a parsed query (see
+Section~\ref{sec:understandquery} for details), models may generate
+plans for any algebra object of
 their choosing by implementing the \pmodule{Attean::API::CostPlanner}
@@
 
-The caching proxy consists of several components:
+The caching proxy consists of several components, detailing the
+overview in Figure~\ref{fig:arch}:
 

### 2016-07-29 dissertation/cache-proxy.tex
new intro
describe Redis

@@
 
-The caching proxy consists of several components, detailing the
-overview in Figure~\ref{fig:arch}:
+It is interesting to add further implementation details to the
+overview in Figure~\ref{fig:arch} with a list of the implemented
+components of the caching proxy:
 
@@
 query to a persistent analyser script that subscribes to a channel on
-a Redis\footnote{See \url{http://redis.io/}} data structure store.
+a Redis\footnote{See \url{http://redis.io/}} data structure store. The
+Redis data structure store is a mostly-memory ``NoSQL'' database for
+storing primitive data structures identified by a key.
 

### 2016-07-29 dissertation/cache-proxy.tex
more redis
detail the analyser in a section of its own

@@
 
+\subsection{Analysis for prefetching}\ref{sec:analpre}
+
+The objective of the analyser in Figure~\ref{fig:arch} is to predict
+what remote data is most useful to answer future queries. I haven't
+attempted to formalise the notion of what a useful cache entry is, nor
+have I attempted to exploit the concept of profitable query patterns
+of \cite{papailiou2015graph}, mostly because this work came at a time
+where my work was already near the deadline. I opted for a simple
+approach to consider only individual triple patterns.
+
+With this constraint, the analysers have been implemented with two
+different assumptions of usefulness: In the first case, I try to
+capture frequently occurring bound predicates. This is due to the
+observation that predicates are often bound, and so predicates that
+are common will occur in the future, and therefore, prefetching and
+caching results of a triple pattern where the predicate is bound to a
+frequently occurring is likely to be beneficial. This is done by
+counting the occurrences of predicates, and when the number exceeds a
+certain threshold, the prefetcher is instructed to fetch the data.
+
+The second analyser works along a different axis: It tries to find the
+triple pattern that, if cached, would provide the greatest cost
+benefit. To do so, it will rerun the query planner on the incoming
+query while simulating that the results of a certain triple pattern
+was in the cache. The simulation will run for \emph{all} triple
+patterns that do not already have a cached result. If the cost
+reduction of the best plan is larger than a certain limit, the
+prefetcher is invoked.
+
 \section{Implementation}\label{sec:impl}
@@
 
-\subsubsection{Analysis and Prefetching}\label{sec:analpre}
+\subsubsection{Analysis and Prefetching}\label{sec:analimpl}
 
@@
 Redis data structure store is a mostly-memory ``NoSQL'' database for
-storing primitive data structures identified by a key.
+storing primitive data structures identified by a key. It also
+features a publish-subscribe system.
 
 The analyser script can be configured to use any number of analysers.
-Two such analysers have been implemented, for now constrained to
-single triple patterns. One will count the predicates in consecutive
-queries it sees, and publish triple patterns to another Redis channel
-with the predicate if the count exceeds a certain configurable
-limit. This is an attempt to capture and cache popular parts of
-queries.
-
-The second analyser will rerun the query planner on the incoming query
-while simulating that the results of a certain triple pattern was in
-the cache. If the cost reduction of the best plan is larger than a
-configurable limit, it will publish that triple pattern to the same
-Redis channel as above.
+Two such analysers have been implemented as according to the ideas in
+Section~\ref{sec:analpre}. Any further analysers that operate on
+single triple patterns are trivial to implement, whereas analysers for
+more advanced query patterns would require new plan classes as the
+current cache system relies on using the \pcode{access\_plans} method,
+which is applicable only to single triple/quad patterns. Any triple
+patterns found by the analysers is published by them to another Redis
+channel for the prefetcher to retrieve.
 
@@
 also be able to generate and use TPF plans. The
-final is a query planner for the analyser in Section~\ref{sec:analpre}
+final is a query planner for the analyser in Section~\ref{sec:analimpl}
 that reruns the query planning with all triple patterns in a simulated
@@
 also defined by the module. If it is not the case, the module also has
-an implementation of the \pcode{access\_plan} method in a
+an implementation of the \pcode{access\_plans} method in a
 \pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be

### 2016-07-29 dissertation/cache-proxy.tex
Some todos; mention that analysers are important
Make the big table in Appendix landscape
on cache maintenance

@@
   Triple Patterns Fragments Server and a SPARQL Endpoint. The most
-  important component is the query engine, which consists of a query
-  planner and a query evaluator.}\label{fig:arch}
+  important components are the query engine, which consists of a query
+  planner and a query evaluator, and the analysers that find
+  subqueries to cache.}\label{fig:arch}
 \end{center}
@@
 
-\subsection{Analysis for prefetching}\ref{sec:analpre}
+\subsection{Analysis for prefetching}\label{sec:analpre}
 
@@
 described in Appendix~\ref{sec:ldfclient}.
+
+Cache maintenance, for example purging elements that are not useful or
+has expired from the cache, is an orthogonal problem to the proposed
+system. It is currently done straightforwardly in the User Agent
+described in Appendix~\ref{sec:modules} which supports HTTP caching as
+specified in RFC7234 \cite{rfc7234}, but further sophistication could
+be added without modifications to the rest of the system.
 
@@
 
+\begin{landscape}
 \begin{table}
@@
 \end{table}
-\todo{make it fit. Rotate?}
+\end{landscape}
 

### 2016-07-29 dissertation/cache-proxy.tex
names for the analysers

@@
 With this constraint, the analysers have been implemented with two
-different assumptions of usefulness: In the first case, I try to
-capture frequently occurring bound predicates. This is due to the
+different assumptions of usefulness: The first analyser, called
+predicate-count, captures frequently occurring bound predicates. This is due to the
 observation that predicates are often bound, and so predicates that
@@
 
-The second analyser works along a different axis: It tries to find the
+The second analyser, named cost-reduction, works along a different
+axis: It tries to find the 
 triple pattern that, if cached, would provide the greatest cost

### 2016-07-29 dissertation/cache-proxy.tex
small

@@
 The second analyser, named cost-reduction, works along a different
-axis: It tries to find the 
+axis: It will find the 
 triple pattern that, if cached, would provide the greatest cost

### 2016-07-29 dissertation/cache-proxy.tex
Excuse the cost model

@@
 models, I have considered this to be future work, and only a small
-heuristic cost model is present in the current system.
-
-However, since the combination of remotely executed SPARQL with a
-local cache and TPF is rather complex, the cost
-heuristics are fairly elaborate. Moreover, since the objective of the
-study is to take load off of the remote endpoint, the heuristics are
-intended to make the proxy work harder than the remote endpoint.
+heuristic cost model is present in the current system. This is guided
+by intuition and verified mainly by a number of tests in the code for
+expected corner cases. Importantly the different parts of the cost
+model are tuned to balance each other, not actually estimate the cost
+of any actual operation.
+
+Admittedly, with this approach, using the term ``cost model'' is
+somewhat of a misnomer. An important motivation for this work is to
+relieve the remote endpoint of stress, and so a principled approach
+could attempt to estimate the cost for the remote endpoint, but
+rather, the intuition explained in Section~\ref{sec:understandquery}
+motivates the choices made in the following.
+
+Moreover, since the combination of remotely executed SPARQL with a
+local cache and TPF is rather complex, the cost heuristics are fairly
+elaborate. Therefore, an attempt to create a proper cost model would
+make that a primary focus of the research, rather than the empirical
+analysis of a first-generation caching proxy integrating SPARQL and
+LDF we set out to achieve. That being said, I have chosen to keep the
+``cost model'' terminology since the usage of the metric resembles how
+it is used in the literature.
 

### 2016-07-29 dissertation/cache-proxy.tex
explain the two first costs

@@
 
-The cost of evaluating a Triple Pattern Fragment plan $C_{tpf}$ is estimated to
-be 
+The cost of evaluating a Triple Pattern Fragment plan $C_{tpf}$ is
+designated to be
 \begin{equation}
@@
 reason we use a floor function is that the Attean cost model only
-supports integers.
-
-The cost of accessing the cache is assumed to be small compared to the
-above, and is estimated as 
+supports integers. This choice is mainly motivated from the assumption
+that evaluating TPFs is a relatively cheap operation for the remote
+endpoint, and so the cost should be contained within an interval, so
+that other costs can easily be balanced against it.
+
+It is then assumed that the cost of accessing the cache is usually
+lower than the above, i.e. it should usually be below the lowest cost
+of a TPF (yet have no upper bound), and is therefore designated as
 \begin{equation}

### 2016-07-29 dissertation/cache-proxy.tex
elaborate the costs

@@
 much depends on the type plans. Conventionally, for a nested loop join
-plan, we multiply the cost of the subplans, increment the resulting
-cost by one if the right hand plan is more expensive than the left
-hand plan, and multiply by a factor of 10 if there are no shared
-variables, to mildly penalise Cartesian joins. If the plan is a hash
-join the difference is that we add rather than multiply the cost of
-the of the subplans, and multiply by a factor of 100 rather than 10 so
+plan, the we first assign 
+\begin{equation}\label{eq:cnl}
+T_{A \bowtie_{nl} B} = 10 ((1 + C_A) C_B ) ,
+\end{equation}
+where $A$ and $B$ signifies the left and right subplans, and $C_A$ and
+$C_B$ their respective cost estimates (which is originates from the
+other heuristics in this section).  Following the default
+implementation in Attean, we multiply the cost in Eq.~(\ref{eq:cnl})
+by a factor of 10 if there are no shared variables, to mildly penalise
+Cartesian joins.
+
+If the plan is a hash join the, we continue to follow the example of
+the Attean implementation and set
+\begin{equation}\label{eq:ch}
+T_{A \bowtie_{h} B} = 10 ((1 + C_A) + C_B ) .
+\end{equation}
+Further, if there are no shared variables, the cost of
+Eq.~(\ref{eq:ch}) is multiplied by a factor of 100 rather than 10 so
 that if it is necessary to evaluate a Cartesian join, it is more
-likely done using a nested loop join. Then, the resulting cost is
-multiplied by the number of remote SPARQL Basic Graph Patterns in the
-tree and the factor 1.2, and floored to the nearest integer.
+likely done using a nested loop join.
+
+Then, to make local joins somewhat more expensive than similar BGPs\todo{check},
+for both these join types, we designate the cost of joins to be
+\begin{equation}\label{eq:cjoins}
+C_{A \bowtie_* B} = \floor{ 1.2 n_{rs} T_{A \bowtie_{*} B} } ,
+\end{equation}
+where the asterisk denotes either $nl$ for nested loop or $h$ for hash
+join, respectively, and $n_{rs}$ is the number of remote SPARQL Basic
+Graph Patterns in the tree.
 

### 2016-07-29 dissertation/cache-proxy.tex
somewhat sloppy for more bgps

@@
 
-Then, to make local joins somewhat more expensive than similar BGPs\todo{check},
+Then, to ensure that the join rotation and coalescing can be
+efficient, we penalise plans that have more of them, so
 for both these join types, we designate the cost of joins to be

### 2016-07-29 dissertation/cache-proxy.tex
some proofreading

@@
 of \cite{papailiou2015graph}, mostly because this work came at a time
-where my work was already near the deadline. I opted for a simple
+when my work was already near the deadline. I opted for a simple
 approach to consider only individual triple patterns.
@@
 
-It is then assumed that the cost of accessing the cache is usually
-lower than the above, i.e. it should usually be below the lowest cost
+I have assumed that the cost of accessing the cache is usually
+lower than the above, that is, it should usually be below the lowest cost
 of a TPF (yet have no upper bound), and is therefore designated as
@@
 
-If the plan is a hash join the, we continue to follow the example of
+If the plan is a hash join, we continue to follow the example of
 the Attean implementation and set

### 2016-07-29 dissertation/conclusions.tex
Some todos; mention that analysers are important
More future work

@@
 \end{quote}
+
+In Section~\ref{sec:understandquery}, it is noted that
+Problem~\ref{prob:microcontroller} makes it important for a query
+planner that integrates LDF query answering to find a balance between
+client and server capability.
+
+The lack of a formalisation of the notion of a useful cache entry is
+noted in Section~\ref{sec:analpre}, and while there has been some work
+on this topic, I am not aware of any attempt to do so in the context
+of the present topic.
 
@@
 Problems~\ref{prob:unknowndist},~\ref{prob:unknownconddist}~and~\ref{prob:largestats}
-has a promising approach in \cite{5767868}.
+has a promising approach in \cite{5767868}, but a way to expose this
+information to a client or a proxy is lacking. We saw in
+\cite{splendid} that a VoID description can be used with some success,
+but this needs further elaboration.
 

### 2016-07-30 dissertation/abstract.tex
Move abstract to a file

--- 2016-07-27 dissertation/index.tex
@@
-\documentclass[a4paper, 12pt, draft]{uiophd}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
-\usepackage{multirow}
-\usepackage{listings}
-\usepackage{supertabular}
-\usepackage{cite}
-\usepackage{amsthm}
-\usepackage{tikz}
-\usepackage{pdfpages}
-\usepackage[obeyDraft]{todonotes}
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
-
-\include{notasjon}
-\include{helpers}
-\usepackage[pdftex,linktocpage]{hyperref}
-\hyperbaseurl{https://github.com/}
-
-\newtheorem{problem}{Problem}
-\newtheorem{example}{Example}
-
-\begin{document}
-\title{SPARQL on the Open, Decentralised Web}
-\author{Kjetil Kjernsmo}
-%\institute{Department of Informatics,
-%Postboks 1080 Blindern,
-%0316 Oslo, Norway}
-
-%\email{kjekje@ifi.uio.no}
-
-\frontmatter
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\tableofcontents
-\mainmatter
-
-\input{introduction}
-
-\input{introdefhistory}
-\input{motivation}
-
-\chapter{Technology Details}
-\input{details}
-\input{related}
-
-\chapter{Problems and Solutions}
-
-\input{problems}
-\input{papersincontext}
-\input{conclusions}
-\input{cache-proxy}
-
-\bibliography{management,dataprofiles,federation,dynamicity,hypermedia,specs,webarch,practical,semweb,caching,critisism,data,philosophy,benchmarks,rfc,programming,egne,optimization}
-\bibliographystyle{plain}
-
-\chapter{Papers}
-
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2012/hypermedia-rdf.pdf}
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2013/iswc.pdf}
-\includepdf[pages=1-last,openright]{paper14.pdf}
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2015/cache-survey.pdf}
-\includepdf[pages=1-last,openright]{report.pdf}
-\includepdf[pages=1-last,openright]{noise.pdf}
-\end{document}

### 2016-07-30 dissertation/index.tex
Use the appendix package
more on architecture
Make the big table in Appendix landscape
Move abstract to a file

--- 2016-07-22 dissertation/index.tex
@@
 \usepackage{pdfpages}
-\usepackage[obeyDraft]{todonotes}
+\usepackage[toc]{appendix}
+\usepackage{todonotes}
+\usepackage{pdflscape}
 \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
@@
 
-\begin{abstract}
-  This dissertation discusses a broad range of problems concerning the
-  use of the SPARQL query language on the open, public Web. It is
-  motivated from seeing decentralisation of data and infrastructure as
-  an important social goal, and SPARQL as an enabling technology to
-  solve problems using the Semantic Web. The dissertation makes
-  contributions in hypermedia, where RDF is used to create a format
-  that can tell humans and machines alike how to manipulate resources
-  on the Web; philosophy of science, where important foundational
-  problems around how to create valid knowledge and objectivity are
-  discussed; statistical methods that better satisfies the
-  requirements from philosophy of science than current practice; how
-  to improve developer efficiency with novel programming paradigms;
-  and finally how caching infrastructure in the Internet may be used
-  to make query answering across the Web more robust.
-
-  \textbf{This is an unfinished draft of the thesis.}
-\end{abstract}
+\input{abstract}
 

### 2016-07-30 dissertation/acknowledgments.tex
First stab at acknowledgements

\section*{Acknowledgements}

First and foremost it is my pleasure to thank my caring wife Hege,
whose invaluable support and affection has been critical to carry
through this project, and then my children Synne, Eivind and Marius,
two of whom have been born while working on this, for putting up with
a father that sometimes has his mind in the computer. My deceased
parents Ragnhild and Dag has had a major impact with their stimulus
throughout childhood and their enduring support. Arguably, allowing me
to use the 1985 PC with no graphics adapter to write programs was the
start of this project.

My advisers Martin Giese and Arild Waaler deserves a big thanks for
kickstarting the project, help adjusting its course over the years and
thorough and enlightened criticism of manuscripts prior to submission.

I would like to thank my co-authors: In particular Gregory Todd
Williams, who have not only co-authored a paper, but written much of
the underlying software I use, formed the standards that this work
builds on, and been a great discussion partner and friend. John
S. Tyssedal deserves a big thanks for agreeing to write a paper where
we started out with very little common ground. Discussions with Ruben
Verborgh always creates the right amount sparks to be enlightening,
and it has been a pleasure to become friends.

It is also a pleasure to thank the entire Semantic Web community for
their openness and welcoming attitude. Whether you come as a hacker,
or want to pursue standardisation, you are likely to meet people like
Dan Brickley, Charles Nevile, Andy Seaborne, Steve Harris, Phil Archer
or Ivan Herman, all of whom have been important for my understanding
of the Semantic Web. The academic community, with their important ISWC
and ESWC conferences are equally welcoming, and has culture of sharing
insight that many other academic communities would envy. Many thanks
are due to Oscar Corcho for first integrating me into the community,
Maria-Esther Vidal and Jacco van Ossenbruggen for encouraging me to
further pursue the speculations around the role of empirical methods,
and general support. Jürgen Umbrich and Axel Polleres deserves a big
thanks for welcoming me in Vienna and many discussions.

I have made many friends in the community, for which I am thankful. Of
these, the following in no particular order have in one way or the
other shaped this work through personal contact beyond published
papers: Olaf Görlitz, Norman Gray, Ansgar Scherp, Frank Van Harmelen,
Olaf Hartig, Javier David Fernández García, Aidan Hogan, Maribel
Acosta and Philip Stutz.

The Perl community have also been important in developing the ideas
and the code of this thesis. In addition to Gregory Todd Williams, I
would like to thank Chris Prather, Kip Hampton, Robin Berjon, Michael
Nachbaur and Jonas Smedegaard.

Finally, my industry colleagues David Norheim and Robert Engels have
also been important in shaping my technical understanding and my
ideas.

### 2016-07-30 dissertation/introduction.tex
Add label
Fix mssing
software contributions

@@
 
-\subsection{Interacting With Data}
+\subsection{Interacting With Data}\label{sec:introinteract}
 
@@
 \end{verbatim}
+\end{example}
 from the above example (prefixes omitted for brevity). Frameworks in
@@
 example, consider the following BGP:
-\begin{example}{A Basic Graph Pattern}
+\begin{example}{A Basic Graph Pattern}\label{ex:bgp}
 \begin{verbatim}
@@
 \end{enumerate}
+
+\subsection{Software}
+
+\begin{enumerate}
+\item 5 Free Software packages that have been accepted into the Debian
+  GNU/Linux operating system.
+\item Further 5 packages that already were accepted in Debian have been
+  enhanced in the course of this work.
+\item Accepted contributions to the code base of another 5 external projects.
+\end{enumerate}

### 2016-07-30 dissertation/acknowledgments.tex
add @tobyink

@@
 It is also a pleasure to thank the entire Semantic Web community for
-their openness and welcoming attitude. Whether you come as a hacker,
+their openness and welcoming attitude. Whether you come as a hacker
 or want to pursue standardisation, you are likely to meet people like
@@
 and the code of this thesis. In addition to Gregory Todd Williams, I
-would like to thank Chris Prather, Kip Hampton, Robin Berjon, Michael
-Nachbaur and Jonas Smedegaard.
+would like to thank Toby Inkster, Chris Prather, Kip Hampton, Robin
+Berjon, Michael Nachbaur and Jonas Smedegaard.
 

### 2016-07-30 dissertation/acknowledgments.tex
another ack

@@
 would like to thank Toby Inkster, Chris Prather, Kip Hampton, Robin
-Berjon, Michael Nachbaur and Jonas Smedegaard.
+Berjon, Shawn M.  Moore, Michael Nachbaur and Jonas Smedegaard.
 

### 2016-07-30 dissertation/acknowledgments.tex
and Helen

@@
 
-Finally, my industry colleagues David Norheim and Robert Engels have
+My industry colleagues David Norheim and Robert Engels have
 also been important in shaping my technical understanding and my
 ideas.
+
+Finally, many thanks to Helen Murray for linguistic assistance.

### 2016-07-30 dissertation/papersincontext.tex
Link to more in structure
inline citations

@@
 workshop at the Ninth Extended Semantic Web Conference. I was the sole
-author. The reference is \cite{kjernsmo_lapis_2012}.
+author. The citation is:
+
+\begin{quote}
+Kjetil Kjernsmo.
+\newblock The necessity of hypermedia rdf and an approach to achieve it.
+\newblock In {\em Proceedings of the First Linked APIs workshop at the Ninth
+  Extended Semantic Web Conference}, May 2012.
+\end{quote}
 
@@
 Endpoint Evaluation'', which was co-authored with John
-S. Tyssedal. The reference is \cite{kjernsmo_doe_intro}.
+S. Tyssedal. The citation is:
+
+\begin{quote}
+Kjetil Kjernsmo and John~S. Tyssedal.
+\newblock Introducing statistical design of experiments to sparql endpoint
+  evaluation.
+\newblock In Harith Alani, Lalana Kagal, Achille Fokoue, Paul Groth, Chris
+  Biemann, Josiane~Xavier Parreira, Lora Aroyo, Natasha Noy, Chris Welty, and
+  Krzysztof Janowicz, editors, {\em The Semantic Web – ISWC 2013}, volume
+  8219 of {\em Lecture Notes in Computer Science}, pages 360--375. Springer
+  Berlin Heidelberg, 2013.
+\end{quote}
 
@@
 International Semantic Web Conference, by the lead author Gregory Todd
-Williams. The reference is \cite{williamspushing}.
+Williams. The citation is:
+
+\begin{quote}
+Gregory~Todd Williams and Kjetil Kjernsmo.
+\newblock Pushing complexity down the stack.
+\newblock In {\em Proceedings of the ISWC Developers Workshop 2014}, volume
+  1268 of {\em {CEUR} Workshop Proceedings}. CEUR-WS.org, May 2014.
+\end{quote}
 
@@
 Semantic Web Conference 2015 entitled ``A Survey of HTTP Caching
-Implementations on the Open Semantic Web'' with reference
-\cite{kjernsmo_survey_2015} and a technical report
-\cite{kjernsmo_add_survey_2015} that elaborates on data reduction
-methods and statistical methods.
+Implementations on the Open Semantic Web'' with citation:
+
+\begin{quote}
+Kjetil Kjernsmo.
+\newblock A survey of http caching implementations on the open semantic web.
+\newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
+  Philippe Cudré-Mauroux, and Antoine Zimmermann, editors, {\em The Semantic
+  Web. Latest Advances and New Domains}, volume 9088 of {\em Lecture Notes in
+  Computer Science}, pages 286--301. Springer International Publishing, 2015.
+\end{quote}
+
+and a technical report that elaborates on data reduction
+methods and statistical methods with the citation:
+
+\begin{quote}
+Kjetil Kjernsmo.
+\newblock Addendum to a survey of {HTTP} caching on the {Semantic Web}.
+\newblock Technical Report 444, Department of Informatics, University of Oslo,
+  March 2015.
+\end{quote}
 
@@
 the chairs decided to accept it as an interview. With Ruben Verborgh
-in the role of the interviewer, we reformulated it as
-\cite{kjernsmo_noise_2015_interview}, which was published in the
+in the role of the interviewer, we reformulated it and it was published in the
 proceedings and is included in the dissertation. It was performed with
@@
 \url{http://folk.uio.no/kjekje/2015/noise-essay.html} is however a
-more thorough account.
+more thorough account. The citation of the interview is:
+\begin{quote}
+Kjetil Kjernsmo and Ruben Verborgh.
+\newblock How can scientific methods provide guidance for semantic web research
+  and development.
+\newblock In {\em Workshop on Negative or Inconclusive Results in Semantic Web
+  Proceedings}, volume 1435 of {\em {CEUR} Workshop Proceedings}. CEUR-WS.org,
+  2015.
+\end{quote}
 
@@
 
-\section{Unpublished Work on Query Caching}
+\section{Unpublished Work on Query Caching}\label{sec:unpublished}
 

### 2016-08-03 dissertation/abstract.tex
Martins final comments

@@
 \begin{abstract}
-  This dissertation\todo{thesis or dissertation, or no change?} discusses a broad range of problems concerning the
+  This dissertation discusses a broad range of problems concerning the
   use of the SPARQL query language on the open, public Web. It is
@@
 
-  \textbf{This is an unfinished draft of the thesis.}
 \end{abstract}

### 2016-08-03 dissertation/introduction.tex
minor
Martins final comments

@@
 The World Wide Web, or just the Web for short, is a well-known global
-information space invented by Tim Berners-Lee in 1989, further emerged
+information space invented by Tim Berners-Lee in 1989, that further emerged
 in the 1990s. It is characterised first and foremost by its
@@
 Web with languages for expressing information in a machine processable
-form\cite{semwebroadmap}.\todo{citations in intro}
+form\cite{semwebroadmap}.
 
@@
 e.g. \url{https://www.w3.org/RDF/} is an address for a document about
-RDF. On the Semantic Web, the URL is first generalised to a Uniform
+RDF. It is also the document that lists the specifications that the
+following outline relies. 
+On the Semantic Web, the URL is first generalised to a Uniform
 Resource Identifier, URI, that can not only name a document on the
@@
     foaf:name "Kjetil Kjernsmo" ;
-    foaf:publications <http://folk.uio.no/kjekje/2013/iswc.pdf> .
+    foaf:made <http://folk.uio.no/kjekje/2013/iswc.pdf> .
 <http://folk.uio.no/kjekje/2013/iswc.pdf> a foaf:Document ;
@@
 is used for declaring the subject an instance of a class, in this case
-\rdfterm{foaf:Person}. Then, the \rdfterm{foaf:publications} predicate
+\rdfterm{foaf:Person}. Then, the \rdfterm{foaf:made} predicate
 is used to link to an author-published version of one of the
@@
   enhanced in the course of this work.
-\item Accepted contributions to the code base of another 5 external projects.
+\item Contributions to the code base were accepted by another 5
+  external projects.
 \end{enumerate}

### 2016-08-03 dissertation/introduction.tex
Martins final comments

@@
 expressions. Together, they are known as a Basic Graph Pattern
-(BGP). More advanced parts of the language is built around BGPs.
+(BGP). More advanced parts of the language are built around BGPs.
 
@@
 graph with a IRI, an approach that was subsequently adopted by
-RDF. Today, RDF is not just a triple, but may be a quad.
+RDF. Therefore, RDF is today often implemented not just a expressed as
+a triple, but may be a expressed with a quad.
 
@@
 
-The present work sits in the confluence of several contemporary
+The present work is located at the confluence of several contemporary
 efforts in the Semantic Web community. The focus is on query answering

### 2016-08-03 dissertation/introduction.tex
improve pseudograh desc

@@
 graph, where the subject IRI is a vertex, the predicate is an edge to
-an object vertex\footnote{In general it is more precisely described as a
-  partially labelled directed pseudograph with unique node labels}.
+an object vertex\footnote{A working draft of the RDF Semantics
+  specification dated 23 January 2003, see
+  \url{https://www.w3.org/TR/2003/WD-rdf-mt-20030123/}, described the
+  graph structure as a partially labeled directed pseudograph with
+  unique node labels.}.
 

### 2016-08-03 dissertation/motivation.tex
remove todo
Martins final comments

@@
 this field. These questions are as hard to ask and to live with as to
-answer. \todo{Too fluffy?}
+answer.
 
-I became early quite confident that the practice of benchmarking was
+I became quite confident at an early stage that the practice of benchmarking was
 an easy target, with its lack of summary statistics and inability to

### 2016-08-03 dissertation/acknowledgments.tex
and patrick
Martins final comments

@@
 
-My advisers Martin Giese and Arild Waaler deserves a big thanks for
+My advisers Martin Giese and Arild Waaler deserve a big thanks for
 kickstarting the project, help adjusting its course over the years and
@@
 further pursue the speculations around the role of empirical methods,
-and general support. Jürgen Umbrich and Axel Polleres deserves a big
+and general support. Jürgen Umbrich and Axel Polleres deserve a big
 thanks for welcoming me in Vienna and many discussions.
@@
 and the code of this thesis. In addition to Gregory Todd Williams, I
-would like to thank Toby Inkster, Chris Prather, Kip Hampton, Robin
-Berjon, Shawn M.  Moore, Michael Nachbaur and Jonas Smedegaard.
+would like to thank Toby Inkster, Patrick Hochstenbach, Chris Prather,
+Kip Hampton, Robin Berjon, Shawn M.  Moore, Michael Nachbaur and Jonas
+Smedegaard.
 

### 2016-08-03 dissertation/related.tex
mention caching in relation to TPF
remove todo

@@
 expensive for the server.
-%TODO: ref characteristic sets?
 Indeed, this information may be so expensive to provide that SPARQL
@@
 In \cite{verborgh2014querying}, they demonstrated SPARQL evaluation
-over TPF, and TPF has become an active area of research.
+over TPF, and TPF has become an active area of research. The impact of
+caching is also measured in this work, as the result of a single
+triple pattern is far easier to cache using HTTP mechanisms than the
+result of an entire SPARQL query.
 
@@
 Graph Patterns, which provides further opportunities for caching where
-only partial results are available.\todo{Have I understood this
-  correctly? See \url{http://ieeexplore.ieee.org/xpls/icp.jsp?arnumber=1161590\#fig1}}
+only partial results are available.
 

### 2016-08-03 dissertation/papersincontext.tex
Caps the titles

@@
 Kjetil Kjernsmo.
-\newblock The necessity of hypermedia rdf and an approach to achieve it.
+\newblock The Necessity of Hypermedia RDF and an Approach to Achieve it.
 \newblock In {\em Proceedings of the First Linked APIs workshop at the Ninth
@@
 Kjetil Kjernsmo and John~S. Tyssedal.
-\newblock Introducing statistical design of experiments to sparql endpoint
-  evaluation.
+\newblock Introducing Statistical Design of Experiments to SPARQL Endpoint
+  Evaluation.
 \newblock In Harith Alani, Lalana Kagal, Achille Fokoue, Paul Groth, Chris
@@
 Gregory~Todd Williams and Kjetil Kjernsmo.
-\newblock Pushing complexity down the stack.
+\newblock Pushing Complexity Down the Stack.
 \newblock In {\em Proceedings of the ISWC Developers Workshop 2014}, volume
@@
 Kjetil Kjernsmo.
-\newblock A survey of http caching implementations on the open semantic web.
-\newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
+\newblock A Survey of HTTP Caching Implementations on the Open Semantic Web.
+\Newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
   Philippe Cudré-Mauroux, and Antoine Zimmermann, editors, {\em The Semantic

### 2016-08-03 dissertation/cache-proxy.tex
Martins final comments

@@
 \begin{verbatim}
-?s <http://example.org/m/p> "1" ;
-   <http://example.org/m/q> <http://example.org/m/a> ;
-   <http://example.org/m/p> ?o .
-?o <http://example.org/m/b> "2" .
+?s <http://example.org/m/p> "1" .                      # A
+?s <http://example.org/m/q> <http://example.org/m/a> . # B
+?s <http://example.org/m/p> ?o .                       # C
+?o <http://example.org/m/b> "2" .                      # D
 \end{verbatim}
@@
 up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
-quad patterns \todo{correctly expressed?} in the order in
+triple patterns in the order expressed in
 Example~\ref{ex:bgp4}, the initial join tree produced by parsing may
-be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
+be $ (((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
 a plan on the form $A \bowtie BCD$, i.e. $BCD$ is reduced to a more
@@
 we employ \emph{join rotation}. A simple example of join rotation
-would be this: We first generate subplans for both $((A \bowtie B
+would be this: We first generate subplans for both $(((A \bowtie B
 )\bowtie C) \bowtie D) $ and $ (A \bowtie B )\bowtie (C \bowtie D)

### 2016-08-03 dissertation/introdefhistory.tex
add todo
remove todo
Big Problem statement #7

@@
 Web, as it provides guidance for the investigation detailed in the
-papers.
+papers. The overarching problem this work tries to address is this:
+
+\begin{quote}
+To achieve a Semantic Web where anyone can say anything about
+anything, and where everyone is enabled to analyse data that have been
+contributed, there is a difficult balance between centralising
+infrastructure and decentralising it, and between processing on
+clients and servers.
+\end{quote}
 
@@
 in Section~\ref{sec:future}. Key insights are finally summarised in
-Section~\ref{sec:conclusions}.\todo{Big problem statement here}
+Section~\ref{sec:conclusions}.
 

### 2016-08-03 dissertation/related.tex
Improve desc of Hartig, #21

@@
 feasibility, soundness and completeness. The most relevant paper to
-this work is \cite{hartig2011caching}.
+this work is \cite{hartig2011caching}. In that work, he considers the
+impact of caching on performance and completeness. Completeness is
+important in the context of query answering when the collection is not
+defined, and the cache may therefore provide results that were not
+accessible for some reason, at query time. Hartig finds many cases
+where it is profitable to cache, but also cases where it is not.
 

### 2016-08-03 dissertation/cache-proxy.tex
remove todo
Shorten example.org/m

@@
 \begin{verbatim}
-?s <http://example.org/m/p> "1" .                      # A
-?s <http://example.org/m/q> <http://example.org/m/a> . # B
-?s <http://example.org/m/p> ?o .                       # C
-?o <http://example.org/m/b> "2" .                      # D
+?s <http://ex.org/p> "1" .                      # A
+?s <http://ex.org/q> <http://ex.org/a> . # B
+?s <http://ex.org/p> ?o .                       # C
+?o <http://ex.org/b> "2" .                      # D
 \end{verbatim}
@@
 -   SPARQLBGP (cost: 60)
--     Quad { ?s, <http://example.org/m/p>, ?o, <http://test.invalid/graph> }
--     Quad { ?o, <http://example.org/m/b>, "2", <http://test.invalid/graph> }
--     Quad { ?s, <http://example.org/m/q>, <http://example.org/m/a>, <http://test.invalid/graph> }
+-     Quad { ?s, <http://ex.org/p>, ?o, <http://test.invalid/graph> }
+-     Quad { ?o, <http://ex.org/b>, "2", <http://test.invalid/graph> }
+-     Quad { ?s, <http://ex.org/q>, <http://ex.org/a>, <http://test.invalid/graph> }
 -   Iterator (?s with 2 elements) (cost: 2)
@@
 based on the average over the rest of the factors. Based on this, a
-common one-sided two-sample $t$-test can be performed.\todo{Discuss
-  cost model based on DoE?}
+common one-sided two-sample $t$-test can be performed.
 
@@
 \begin{verbatim}
-?s <http://example.org/m/p> "1" ;
-   <http://example.org/m/p> ?o ;
-   <http://example.org/m/q> _:xyz ;
-   <http://example.org/m/q> <http://example.org/m/a> .
-?a <http://example.org/m/b> <http://example.org/m/c> .
+?s <http://ex.org/p> "1" ;
+   <http://ex.org/p> ?o ;
+   <http://ex.org/q> _:xyz ;
+   <http://ex.org/q> <http://ex.org/a> .
+?a <http://ex.org/b> <http://ex.org/c> .
 \end{verbatim}

### 2016-08-03 dissertation/cache-proxy.tex
Shorten test.invalid to ex.org

@@
 -   SPARQLBGP (cost: 60)
--     Quad { ?s, <http://ex.org/p>, ?o, <http://test.invalid/graph> }
--     Quad { ?o, <http://ex.org/b>, "2", <http://test.invalid/graph> }
--     Quad { ?s, <http://ex.org/q>, <http://ex.org/a>, <http://test.invalid/graph> }
+-     Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> }
+-     Quad { ?o, <http://ex.org/b>, "2", <http://ex.org/graph> }
+-     Quad { ?s, <http://ex.org/q>, <http://ex.org/a>, <http://ex.org/graph> }
 -   Iterator (?s with 2 elements) (cost: 2)

### 2016-08-03 dissertation/cache-proxy.tex
Example.org to ex.org/graph

@@
 - Hash Join { s } (distinct cost: 27)
--   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
--   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
 
 - Hash Join { s } (distinct cost: 28)
--   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
--   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
--   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
--   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
+-   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
 \end{verbatim}
@@
 dummy graph name when constructing the Model, namely
-\rdfterm{<http://example.org/>}.
+\rdfterm{<http://ex.org/graph>}.
 

### 2016-08-03 dissertation/cache-proxy.tex
example.org to ex.org

@@
 - Hash Join { s } (distinct cost: 27)
--   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
--   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
 
 - Hash Join { s } (distinct cost: 28)
--   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
--   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
--   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://example.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
--   Quad { ?s, <http://example.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
 \end{verbatim}

### 2016-08-03 dissertation/introduction.tex
Go through text flow #18

@@
 IRIs, e.g. the predicate \rdfterm{foaf:name} is an abbreviation of the
-IRI \rdfterm{http://xmlns.com/foaf/0.1/name}. The IRI
-\rdfterm{http://www.kjetil.kjernsmo.net/foaf\#me} is a name for this
+IRI \rdfterm{\nolinkurl{http://xmlns.com/foaf/0.1/name}}. The IRI
+\rdfterm{\nolinkurl{http://www.kjetil.kjernsmo.net/foaf\#me}} is a name for this
 author. The \rdfterm{a} predicate is standards-defined abbreviation
-for \rdfterm{http://www.w3.org/1999/02/22-rdf-syntax-ns\#type}, which
+for \rdfterm{\nolinkurl{http://www.w3.org/1999/02/22-rdf-syntax-ns\#type}}, which
 is used for declaring the subject an instance of a class, in this case
@@
 \begin{enumerate}
-\item 5 Free Software packages that have been accepted into the Debian
-  GNU/Linux operating system.
+\item 5 Free Software packages that have been accepted into the recent
+  versions of the Debian GNU/Linux operating system.
 \item Further 5 packages that already were accepted in Debian have been

### 2016-08-03 dissertation/related.tex
Forgot the lampo reference
Go through text flow #18

@@
 Interestingly, \cite{papailiou2015graph} opens by noting that many
-optimisations used by SQL databases are ineffective in RDF databases,
+optimisations used by SQL\linebreak[4] databases are ineffective in RDF databases,
 due to that RDF schema, even if they exist, are not as helpful as the
@@
 where it is profitable to cache, but also cases where it is not.
+
+Lampo et al. \cite{lampo2011cache} found similar results. They found
+that queries with star-shaped groups, that is, triple patterns that
+share a common variable in the subject position, take best advantage
+of a cache, and suggest rewriting queries to take better advantage of
+a cache.
 

### 2016-08-03 dissertation/cache-proxy.tex
Go through text flow #18
Go through text flow in cache #18

@@
 Attean internals. The system has in fact three query planners:
-\pmodule{AtteanX::QueryPlanner::Cache},
-\pmodule{AtteanX::QueryPlanner::Cache::LDF} and
-\pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}. Each is an
+\begin{itemize}
+\item \pmodule{AtteanX::QueryPlanner::Cache},
+\item \pmodule{AtteanX::QueryPlanner::Cache::LDF} and
+\item \pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}.
+\end{itemize}
+Each is an
 extension of the previous. The first is written to use a remote
@@
 blocks), etc. In the case of our query planner that combines a remote
-endpoint with a local cache, named
+endpoint with a local cache, the module named above
 \pmodule{AtteanX::QueryPlanner::Cache}, the basic query planner of the
@@
 (i.e. inherited from), but also composes the following roles:
+\begin{itemize}
+\item 
 \pmodule{Attean::API::NaiveJoinPlanner},
-\pmodule{Attean::API::SimpleCostPlanner},
-\pmodule{AtteanX::API::JoinRotatingPlanner},
-\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
-\pmodule{AtteanX::Query::AccessPlan::Cache}, out of which the three
+\item \pmodule{Attean::API::SimpleCostPlanner},
+\item \pmodule{AtteanX::API::JoinRotatingPlanner},
+\item \pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} and
+\item \pmodule{AtteanX::Query::AccessPlan::Cache},
+\end{itemize}
+out of which the three
 first are a part of the Attean framework, the next is described in
@@
 \pmodule{AtteanX::API::JoinRotatingPlanner}. In the planner
-\pmodule{AtteanX::QueryPlanner::Cache} adjacent quad patterns are
+\pmodule{AtteanX::Query\linebreak[4]Planner::Cache} adjacent quad patterns are
 coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
@@
 key challenge is to balance the costs given by different models in the
-system, and this is done in
+system, and this is done in the module
 \pmodule{AtteanX::Model::SPARQLCache::LDF}. Let us, however, start
@@
 should be small, and that plans where the joins have been rotated and
-coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
+coalesced by the implementation in the role \pmodule{AtteanX::API::JoinRotatingPlanner} should be
 less expensive. Therefore, such plans should be penalised, but by how
@@
 The system also depends on modules I have written that are not part of
-the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
-\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.
+the project, but that I have enhanced as time has allowed: 
+\begin{itemize}
+\item \pmodule{RDF::LinkedData},
+\item \pmodule{RDF::Generator::Void},
+\item \pmodule{Test::RDF},
+\item \pmodule{URI::NamespaceMap} and
+\item \pmodule{RDF::NS::Curated}.
+\end{itemize}
 
@@
 Protocol client that implements Attean APIs. The store implementation
-itself composes the \mbox{TripleStore} API and implements methods to retrieve
+itself composes the supplied \mbox{TripleStore} API and implements methods to retrieve
 the results of a single triple pattern and exact cardinality for a

### 2016-08-03 dissertation/introduction.tex
Make examples fit #18

@@
 \begin{example}{Turtle syntax}\label{ex:turtle}
+\small
 \begin{verbatim}
@@
 <http://folk.uio.no/kjekje/2013/iswc.pdf> a foaf:Document ;
-    dct:title "Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation" ; 
+  dct:title "Introducing Statistical Design of [...]" ; 
   dct:creator <http://www.kjetil.kjernsmo.net/foaf#me>, 
-                <http://data.semanticweb.org/person/john-s-tyssedal> .
+              <http://[...]/person/john-s-tyssedal> .
 \end{verbatim}
+\normalsize
 \end{example}
@@
 subject, commas separate statements with the same subject and
-predicate, while periods terminate a statement.
+predicate, while periods terminate a statement. \texttt{[...]} is just
+to shorten the lines in the example and is not a part of the syntax.
 
@@
 \begin{example}{Triples matched by triple pattern}
+\small
 \begin{verbatim}
@@
 \end{verbatim}
+\normalsize
 \end{example}

### 2016-08-03 dissertation/cache-proxy.tex
Make examples fit #18

@@
 - Hash Join { s } (distinct cost: 27)
--   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
--   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <g> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <g> } (cost: 12)
 
 - Hash Join { s } (distinct cost: 28)
--   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
--   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <g> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <g> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
--   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <g> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <g> } (cost: 15)
 
 - NestedLoop Join (distinct cost: 180)
--   Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> } (cost: 15)
--   Quad { ?s, <http://ex.org/p>, "1", <http://ex.org/graph> } (cost: 12)
+-   Quad { ?s, <http://ex.org/p>, ?o, <g> } (cost: 15)
+-   Quad { ?s, <http://ex.org/p>, "1", <g> } (cost: 12)
 \end{verbatim}
@@
 dummy graph name when constructing the Model, namely
-\rdfterm{<http://ex.org/graph>}.
+\rdfterm{<g>}.
 

### 2016-08-03 dissertation/cache-proxy.tex
Make examples fit #18
Not even call it a cost model

@@
 \begin{example}{Four Attean plan trees}
+\small
 \begin{verbatim}
@@
 (27). Since the SPARQL engine operates over quads, we have assigned a
-dummy graph name when constructing the Model, namely
-\rdfterm{<g>}.
+dummy graph name when constructing the Model, in this case further
+shortened to \rdfterm{<g>}.
 
@@
 are therefore to be encoded into a cost model. The cost model itself
-is not considered a part of the contributions, it is a simple cost
-model based on heuristics that arises from the following
+is not considered a part of the contributions, as it is just
+based on heuristics that arises from the following
 discussion. The heuristics themselves are best considered an

### 2016-08-03 dissertation/cache-proxy.tex
Make examples fit #18

@@
 \begin{example}{An example Attean plan with a remotely executed BGP}
+\small
 \begin{verbatim}
@@
 -   SPARQLBGP (cost: 60)
--     Quad { ?s, <http://ex.org/p>, ?o, <http://ex.org/graph> }
--     Quad { ?o, <http://ex.org/b>, "2", <http://ex.org/graph> }
--     Quad { ?s, <http://ex.org/q>, <http://ex.org/a>, <http://ex.org/graph> }
+-     Quad { ?s, <http://ex.org/p>, ?o, <g> }
+-     Quad { ?o, <http://ex.org/b>, "2", <g> }
+-     Quad { ?s, <http://ex.org/q>, <http://ex.org/a>, <g> }
 -   Iterator (?s with 2 elements) (cost: 2)
 \end{verbatim}
+\normalsize
 \end{example}

### 2016-08-03 dissertation/cache-proxy.tex
Make table fit #18

@@
 6355072	 	& 79.7s & 1856s & \pcode{Attean::CodeIterator::next} \\
-2 & 77.4s & 4540s & \pcode{Attean::Plan::HashJoin::\_\_ANON\_\_[Attean/Plan.pm:362]} \\
+2 & 77.4s & 4540s & \pcode{Attean::Plan::HashJoin::\_\_ANON\_\_[...]} \\
 4217970		& 70.0s & 847s & \pcode{Attean::Literal::new} \\
@@
 2419260	        & 50.3s & 208s & \pcode{Attean::API::Result::join} \\
-4217968	        & 48.7s	& 1463s & \pcode{AtteanX::Query::AccessPlan::Cache::\_\_ANON\_\_[AtteanX/Query/AccessPlan/Cache.pm:72]} \\
+4217968	        & 48.7s	& 1463s & \pcode{AtteanX::Query::AccessPlan::Cache[...]} \\
 24186545	& 48.1s	& 56.6s & \pcode{Role::Tiny::does\_role} \\
 28128408	& 44.4s	& 53.8s	& \pcode{Attean::Result::value} \\
-16871877	& 44.1s	& 470s & \pcode{Attean::Literal::\_\_ANON\_\_[(eval 234)[Class/Method/Modifiers.pm:93]:1] (merge of 4 subs)}\\
-4217969		& 40.0s & 92.0s & \pcode{Attean::API::Literal::\_\_ANON\_\_[Attean/API/Term.pm:160]} \\
+16871877	& 44.1s	& 470s & \pcode{Attean::Literal::\_\_ANON\_\_[(eval 234)[...]}\\
+4217969		& 40.0s & 92.0s & \pcode{Attean::API::Literal::\_\_ANON\_\_[...]} \\
 13309111	& 39.5s & 39.5s & \pcode{IRI::CORE:regcomp (opcode)}

### 2016-08-03 dissertation/cache-proxy.tex
remove todo

@@
 already in the cache, as plans for both these options would be
-provided.\todo{Discuss algebra equivalences here?}
+provided.
 

### 2016-08-03 dissertation/introduction.tex
Some fixes #17

@@
 RDF. It is also the document that lists the specifications that the
-following outline relies. 
+following outline relies on. 
 On the Semantic Web, the URL is first generalised to a Uniform
@@
   \url{https://www.w3.org/TR/2003/WD-rdf-mt-20030123/}, described the
-  graph structure as a partially labeled directed pseudograph with
-  unique node labels.}.
+  graph structure as a ``partially labeled directed pseudograph with
+  unique node labels''.}.
 
@@
 graph with a IRI, an approach that was subsequently adopted by
-RDF. Therefore, RDF is today often implemented not just a expressed as
+RDF. Therefore, RDF is today often implemented not just expressed as
 a triple, but may be a expressed with a quad.

### 2016-08-03 dissertation/introduction.tex
mention writing RDF graphs #17

@@
 
+More techniques have been developed for writing data. The simplest is
+to write RDF in one of its serialisations to a file. Basic operations
+like writing single triples or quads may be done with a framework
+directly onto a triplestore. Since write operations are specified in
+HTTP, write operations may use HTTP verbs if they are authorised to do
+so. Finally, SPARQL has been extended with an update language in
+version 1.1.
+
 \subsection{Hypermedia}

### 2016-08-03 dissertation/cache-proxy.tex
Some fixes closes #17

@@
 \cite{adali1996query} system's concept of invariants discussed in
-Section~\ref{sec:relcache}: a SPARQL query cache be oblivious
+Section~\ref{sec:relcache}: a SPARQL query cache should be oblivious
 to domain knowledge in a CDN deployment and therefore shouldn't assume any invariants,
@@
 graph pattern. A forward proxy, on the other hand, may have a deeper
-understanding on the needs of the client, and could possibly implement
+understanding of the needs of the client, and could possibly implement
 invariants. The Attean framework accommodates for this.
@@
 Section~\ref{sec:analpre}. The response \circled{2} is not sent back
-to the client, but will enter the cache. Assuming that the analyser
+to the client, but will enter the cache. We then assume that the analyser
 was successful and that the cache can be used to answer request
@@
 origin server, indicated by the dashed arrows. This may be zero or more SPARQL
-query, TPF requests or a combination thereof, depending on what the
+queries, TPF requests or a combination thereof, depending on what the
 planner finds least expensive. The proxy will then evaluate the entire
@@
 
-As can be imagined from the fact that a BGP with just to triple
+As can be imagined from the fact that a BGP with just two triple
 patterns result in 4 plan trees, a naive query planner may generate a
@@
 by a role that implements default functionality or simply requires
-such functionality to be required. For instance, a quadstore is
+such functionality to be present. For instance, a quadstore is
 required by \pmodule{Attean::API::QuadStore} to implement the
@@
 them \cite{rfc3987}, and the \pmodule{IRI} module is a rigorous
-implementation of the standard that does syntax based resolution if
-the if an object is constructed with a base. IRI parsing happens not
+implementation of the standard that does syntax based resolution
+if an object is constructed with a base. IRI parsing happens not
 only when an \pmodule{Attean::IRI} object is constructed, but may also

### 2016-08-03 dissertation/introdefhistory.tex
Frame the big problem statement, closes #7
Go through text flow #18
Some fixes closes #17

@@
 investigation rather than being the end of any conversations. 
+The overarching problem this work tries to address is this:
 
-The introduction provided a very basic view of the technology, this
-chapter will further discuss some important aspects of the Semantic
-Web, as it provides guidance for the investigation detailed in the
-papers. The overarching problem this work tries to address is this:
-
+\begin{framed}
 \begin{quote}
@@
 \end{quote}
+\end{framed}
+
+The introduction provided a very basic view of the technology, this
+chapter will further discuss some important aspects of the Semantic
+Web, as it provides guidance for the investigation detailed in the
+papers.
 
@@
 problems and details the motivation.
-Sections~\ref{sec:papersincontext}~and~\ref{sec:unpublished} details
+Sections~\ref{sec:papersincontext}~and~\ref{sec:unpublished} detail
 how the contributions address these problems and discusses weaknesses
@@
 
-In 1996, I started the Web pages of the \textit{Norwegian Skeptics Society}
+In 1996, I started the Web pages of the \textit{Norwegian Skeptics
+  Society} using the domain
 \texttt{skepsis.no}, and therefore got the control over an entire Web

### 2016-11-14 dissertation/problems.tex
Mention connected subgraphs in the problems instead
Use sparql macro
remove todo
Go through text flow #18
Note that not all problems addressed, #31, and link to table

@@
 for the whole Semantic Web community, their only intended scope is the
-presented vision.
+presented vision. Moreover, many of them are not addressed in this
+thesis, but discussed for comprehensiveness. An overview of the
+problems and their relation to the published works are included in
+Table~\ref{tab:problempaper}.
 
@@
 \end{example}
-as a unit.\todo{Move to preliminaries?}
+as a unit.
 
@@
 connect the two, for example
-\rdfterm{?person~foaf:publications~?paper~.}. If the result of this
+\sparql{?person~foaf:made~?paper~.}. If the result of this
 triple pattern is cached, what is left to evaluate are the two triple
 patterns in Example~\ref{ex:cartesian}, which is likely a very
-expensive operation.
+expensive operation. Another approach to this problem is that of
+\cite{papailiou2015graph}, as they insist on caching only connected 
+subgraphs.
 
@@
 BGP, and could not use a \jcode{WHERE} clause to evaluate a SPARQL
-\rdfterm{FILTER}. To remedy this, one could have a method for
+\sparql{FILTER}. To remedy this, one could have a method for
 evaluating BGPs with (such as the  \pcode{get\_pattern} method of

### 2016-11-14 dissertation/papersincontext.tex
Mention frank's laws
The problem-paper mapping table #31

@@
 their publication.
+
+The discussion of the problems include references to the problems
+discussed in Section~\ref{sec:problems}, and the relationship can also
+be found in the overview Table~\ref{tab:problempaper}.
 
@@
 \newblock A Survey of HTTP Caching Implementations on the Open Semantic Web.
-\Newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
+\newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
   Philippe Cudré-Mauroux, and Antoine Zimmermann, editors, {\em The Semantic
@@
 data.
+
+Also, the investigation would be deeper if it tried to address
+Problem~\ref{prob:lawsfuture}. A single analysis would not be
+sufficient to do that, laws would require multiple studies over time.
 
@@
 Fragments, which we exploit.
+
+\begin{table}{h}
+  \caption{The relationship between the identified problems and the
+    papers that make contributions towards solving them.}\label{tab:problempaper}
+
+  \begin{tabular}{|l|l|}
+    \hline
+    Paper & Problems \\
+    \hline
+    Hypermedia RDF (Section~\ref{sec:conlapis}) &
+    Developer-understandable interface (Problem~\ref{prob:lapis}) \\
+    Design of Experiments (Section~\ref{sec:condoe}) & Statistical
+    methodologies (Problem~\ref{prob:badstats}) \\
+    Low-level complexity (Section~\ref{sec:conpush}) &
+    Practical evaluation
+    problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}
+    \\
+    Caching survey (Section~\ref{sec:consanity}) &
+    Unknown cache support (Problem~\ref{prob:sanity}) \\
+    Philosophy of Science (Section~\ref{sec:conphil}) & Inadequate
+    philosophical foundation (Problem~\ref{prob:badphil}) \\
+    \hline
+  \end{tabular}
+\end{table}

### 2016-11-15 dissertation/introduction.tex
Create problem overview #27 Map papers to problems #31 Reference papers from contrib #26

@@
 
-\section{Contributions}\label{sec:contribsum}
+\section{Research Problems Overview}\label{sec:problemsum}
 
@@
 methods for evaluating performance, standards compliance and even
-philosophy of science.
+philosophy of science. A comprehensive treatment of research problems
+are given in Section~\ref{sec:problems}, but to understand the
+following contributions, the overarching problems are as follows:
 
-\subsection{Hypermedia}
+It is difficult to many developers to interact with RDF, and
+contributions to alleviate that are summarised in
+Section~\ref{sec:schm}.
+
+We have found that in few cases, at least when evaluating SPARQL
+Engines, the evaluation has had a basis in sound statistics or in
+philosophy of science and therefore the validity of the conclusions
+drawn from them should be challenged. We contributed a new direction
+to the statistical problem in Section~\ref{sec:scdoe} and a
+philosophical discussion in Section~\ref{sec:scphil}.
+
+To create a framework that allows the flexible and at the same time
+efficient manner is difficult. Conventionally, the query engine would
+break down a query to individual triple pattern matches, which did not
+allow for much optimisation, or to easily modify the query
+planning. Our contributions towards these problems are summarised in
+Section~\ref{sec:scdev}.
+
+As noted, the HTTP standard and the Web architecture allows for
+caching responses, and in Section~\ref{sec:scsurvey} we detail
+contributions to understand the contemporary usage of these
+mechanisms. This is particularly interesting in the context of SPARQL
+query execution on the Web, as currently, the infrastructure cannot
+support evaluating arbitrary queries across the Web in a reliable
+manner. A situation where the client, a caching proxy and a server may
+share the work when evaluating a query results in a number of research
+problems that are detailed in Section~\ref{sec:perfproblems}, and we
+attempted to contribute an approach to some of these problems in
+Chapter~\ref{sec:tpfcacheplanning} without reaching any solid
+conclusions.
+
+\section{Contributions}\label{sec:contribsum}
+
+In this section, we summarise the contributions of this work, with
+reference to the detailed problem descriptions in
+Section~\ref{sec:problems}, as well as
+Section~\ref{sec:papersincontext}, where the papers are described and
+put in context.
+
+\subsection{Hypermedia}\label{sec:schm}
+
+The following contributions were made towards Problem~\ref{prob:lapis}
+in the paper described in Section~\ref{sec:conlapis}:
 
@@
 
-\subsection{Design of Experiments}
+\subsection{Design of Experiments}\label{sec:scdoe}
+
+The following contributions were made towards Problem~\ref{prob:badstats}
+in the paper described in Section~\ref{sec:condoe}:
 
@@
 
-\subsection{Development problems}
+\subsection{Development problems}\label{sec:scdev}
+
+The following contributions were made towards
+problems~\ref{prob:breakdown},~\ref{prob:complexapi}~and~\ref{prob:dontjustpass}
+in the paper described in Section~\ref{sec:conpush}:
 
@@
 
-\subsection{Survey of HTTP Caching}
+\subsection{Survey of HTTP Caching}\label{sec:scsurvey}
+
+The following contribution was made towards Problem~\ref{prob:sanity}
+in the papers described in Section~\ref{sec:consanity}:
 
@@
 
-\subsection{Philosophy of Science}
+\subsection{Philosophy of Science}\label{sec:scphil}
+
+The following contribution was made towards Problem~\ref{prob:badphil}
+in the paper described in Section~\ref{sec:conphil}:
 
@@
 
+In addition to papers, the following contributions were made in terms
+of Free Software:
+
 \begin{enumerate}
-\item 5 Free Software packages that have been accepted into the recent
+\item 5 packages that have been accepted into the recent
   versions of the Debian GNU/Linux operating system.

### 2016-11-15 dissertation/introduction.tex
Move the big problem statement

@@
 methods for evaluating performance, standards compliance and even
-philosophy of science. A comprehensive treatment of research problems
-are given in Section~\ref{sec:problems}, but to understand the
-following contributions, the overarching problems are as follows:
+philosophy of science.
+
+The overarching problem this work tries to address is this:
+
+\begin{framed}
+\begin{quote}
+To achieve a Semantic Web where anyone can say anything about
+anything, and where everyone is enabled to analyse data that have been
+contributed, there is a difficult balance between centralising
+infrastructure and decentralising it, and between processing on
+clients and servers.
+\end{quote}
+\end{framed}
+
+A comprehensive treatment of research problems is given in
+Section~\ref{sec:problems}, but to understand the following
+contributions, a superficial overview is provided:
 

### 2016-11-16 dissertation/introduction.tex
small change #27

@@
 philosophy of science and therefore the validity of the conclusions
-drawn from them should be challenged. We contributed a new direction
-to the statistical problem in Section~\ref{sec:scdoe} and a
-philosophical discussion in Section~\ref{sec:scphil}.
+drawn from them should be challenged. We summarise contributions to a
+new direction to the statistical problem in Section~\ref{sec:scdoe}
+and a philosophical discussion in Section~\ref{sec:scphil}.
 

### 2016-11-16 dissertation/index.tex
First stab at acknowledgements
Martins final comments
Insert blank page
Frame the big problem statement, closes #7
Pass hyphen to url #18
Add some PDF info
Remove todonotes
List of papers in Paper Chapter, fixes #29

@@
-\documentclass[a4paper, 12pt]{uiophd}
+\documentclass[a4paper, 12pt, twoside]{uiophd}
 \usepackage{graphicx}
@@
 \usepackage{tikz}
+\usepackage{framed}
 \usepackage{pdfpages}
@@
 \include{helpers}
-\usepackage[pdftex,linktocpage]{hyperref}
+\PassOptionsToPackage{hyphens}{url}\usepackage[pdftex,linktocpage]{hyperref}
 \hyperbaseurl{https://github.com/}
-
+\hypersetup{  
+  pdfinfo={  
+    Title={SPARQL on the Open, Decentralised Web},  
+    Author={Kjetil Kjernsmo}
+  }  
+} 
 \newtheorem{problem}{Problem}
@@
 \maketitle
-
+\newpage
+This page was intentionally left blank
+\newpage
 \input{abstract}
+\input{acknowledgments}
 
@@
 
-\chapter{Technology Details}
+\chapter{Preliminaries and Related Work}
 \input{details}
@@
 
+The following papers are used for this thesis:
+\begin{itemize}
+\item Kjetil Kjernsmo.
+\newblock The Necessity of Hypermedia RDF and an Approach to Achieve it.
+\newblock In {\em Proceedings of the First Linked APIs workshop at the Ninth
+  Extended Semantic Web Conference}, May 2012.
+\item Kjetil Kjernsmo and John~S. Tyssedal.
+\newblock Introducing Statistical Design of Experiments to SPARQL Endpoint
+  Evaluation.
+\newblock In Harith Alani, Lalana Kagal, Achille Fokoue, Paul Groth, Chris
+  Biemann, Josiane~Xavier Parreira, Lora Aroyo, Natasha Noy, Chris Welty, and
+  Krzysztof Janowicz, editors, {\em The Semantic Web – ISWC 2013}, volume
+  8219 of {\em Lecture Notes in Computer Science}, pages 360--375. Springer
+  Berlin Heidelberg, 2013.
+\item Gregory~Todd Williams and Kjetil Kjernsmo.
+\newblock Pushing Complexity Down the Stack.
+\newblock In {\em Proceedings of the ISWC Developers Workshop 2014}, volume
+  1268 of {\em {CEUR} Workshop Proceedings}. CEUR-WS.org, May 2014.
+\item Kjetil Kjernsmo.
+\newblock A Survey of HTTP Caching Implementations on the Open Semantic Web.
+\newblock In Fabien Gandon, Marta Sabou, Harald Sack, Claudia d’Amato,
+  Philippe Cudré-Mauroux, and Antoine Zimmermann, editors, {\em The Semantic
+  Web. Latest Advances and New Domains}, volume 9088 of {\em Lecture Notes in
+  Computer Science}, pages 286--301. Springer International
+  Publishing, 2015.
+\item Kjetil Kjernsmo.
+\newblock Addendum to a survey of {HTTP} caching on the {Semantic Web}.
+\newblock Technical Report 444, Department of Informatics, University of Oslo,
+  March 2015.
+\item Kjetil Kjernsmo and Ruben Verborgh.
+\newblock How can scientific methods provide guidance for semantic web research
+  and development.
+\newblock In {\em Workshop on Negative or Inconclusive Results in Semantic Web
+  Proceedings}, volume 1435 of {\em {CEUR} Workshop Proceedings}. CEUR-WS.org,
+  2015.
+\end{itemize}
+
 \includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2012/hypermedia-rdf.pdf}

### 2016-11-16 dissertation/conclusions.tex
add jurgen-paper on data summary
Expand on the conclusions, fixes #24

@@
 \cite{splendid} that a VoID description can be used with some success,
-but this needs further elaboration.
+but this needs further elaboration. Histograms have also been proposed
+for this purpose in \cite{Umbrich2011}.
 
@@
 
-\begin{itemize}
-\item Hypermedia is one key to developing applications on the Semantic
-  Web.
-\item Traits ease the task of developing experimental systems and
-  usage of optimisations in underlying systems.
-\item While the adoption of caching and conditional requests as defined
-  in HTTP is not widespread, the adoption is sufficient to start using
-  them in practical systems.
-\item Evaluation methodologies suffer from a poor foundation in
-  philosophy of science, and research into the epistemology of the
-  field is important.
-\item The author admits that the study is lacking in terms of
+The conclusions of the papers can be summarised as follows:
+
+Hypermedia can contribute a ``View Source'' capability to application
+development and data exploration, and thinking about an RDF statement
+with control information as a natural language sentence can make RDF
+understandable to many developers. Therefore, it is one key to
+developing applications on the Semantic Web.
+
+When engineering systems to encourage contributions from many
+different parties, it is important to carefully consider how such
+systems can be extended and adapted, and how contributions can be made
+a part of the systems and so enhance the base that researchers and
+other users can benefit from. Traits ease the task of extending
+database systems with experimental features and eases the usage of
+optimisations in underlying systems.
+
+Caching in the Web is a prominent topic, but we found that while the
+adoption of caching and conditional requests as defined in HTTP is not
+widespread, the adoption is sufficient to start using them in
+practical systems. We also found that many resources change at a very
+slow pace, but that it is not reflected in the freshness lifetime set
+by the server, but may more commonly be computed by common heuristics.
+
+Evaluation methodologies suffer from a poor foundation in philosophy
+of science, and research into the epistemology of the field is
+important. This is due to several interrelated factors: The field of
+research differs from natural science in that it studies a human
+artefact that does not yet exist. An example of this problem is that
+there is no formulated coherent theory of the Semantic Web, so the
+external validity of any hypothesis relative to such a theory can
+always be contested. If one would wish to follow the criterion that
+science is distinguished by falsifiability, it is currently difficult
+to tell if the evaluation or the hypothesis is to be rejected.
+
+Moreover, along the above directions, and based on the fact that the
+author has a large intellectual investment in the success of the
+Semantic Web, the author admits that the study is lacking in terms of
 objectivity, but notes that the problems are of a general nature.
-\item Statistical Design of Experiments provides a path to a critical
-  practice of evaluations, needed to improve the foundations of the
-  field.
-\end{itemize}
+
+Empirical evaluations should be developed from the principles of
+philosophy of science and statistics, and the practice of benchmarking
+does not adequately follow these principles. Statistical Design of
+Experiments provides a path to a critical practice of evaluations,
+needed to improve the foundations of the field. The experiment we
+designed provided for both an instructive example of a new methodology
+and how to use the methodology to evaluate the experiment itself.

### 2016-11-16 dissertation/conclusions.tex
flow into prefetcher, fixes #25

@@
 and how to use the methodology to evaluate the experiment itself.
+
+Finally, the work extended to a SPARQL query engine that could run on
+a proxy. It could cache intermediate results of triple pattern
+matches, either in real time or by prefetching results that might be
+valuable. In query planning and evaluation, it could use the cache, a
+remote SPARQL endpoint and a Triple Pattern Fragments
+server. Unfortunately, this work had to be ended before it could reach
+a conclusion, but a detailed account of the work is included in
+Chapther~\ref{sec:tpfcacheplanning}.

### 2016-11-17 dissertation/index.tex
scaling the papers, fixes #23

@@
 
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2012/hypermedia-rdf.pdf}
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2013/iswc.pdf}
-\includepdf[pages=1-last,openright]{paper14.pdf}
-\includepdf[pages=1-last,openright]{/home/kjetil/uio/www_docs/2015/cache-survey.pdf}
+\includepdf[pages=1-last,openright,scale=1.2]{/home/kjetil/uio/www_docs/2012/hypermedia-rdf.pdf}
+\includepdf[pages=1-last,openright,scale=1.2]{/home/kjetil/uio/www_docs/2013/iswc.pdf}
+\includepdf[pages=1-last,openright,scale=1.2]{paper14.pdf}
+\includepdf[pages=1-last,openright,scale=1.2]{/home/kjetil/uio/www_docs/2015/cache-survey.pdf}
 \includepdf[pages=1-last,openright]{report.pdf}
-\includepdf[pages=1-last,openright]{noise.pdf}
+\includepdf[pages=1-last,openright,scale=1.2]{noise.pdf}
 \end{document}

### 2016-11-21 dissertation/conclusions.tex
Proofreading corrections

@@
 
-Problem~\ref{prob:queryeq} is has an interesting possible solution in
+Problem~\ref{prob:queryeq} has an interesting possible solution in
 the application of hashing algorithms that have a reasonable chance of

### 2016-12-09 dissertation/acknowledgments.tex

@@
 Olaf Hartig, Javier David Fernández García, Aidan Hogan, Maribel
-Acosta and Philip Stutz.
+Acosta, Josef Noll and Philip Stutz.
 

### 2016-12-09 dissertation/introdefhistory.tex

@@
 investigation rather than being the end of any conversations.
-The overarching problem this work tries to address is this:
-
-\begin{framed}
-\begin{quote}
-To achieve a Semantic Web where anyone can say anything about
-anything, and where everyone is enabled to analyse data that have been
-contributed, there is a difficult balance between centralising
-infrastructure and decentralising it, and between processing on
-clients and servers.
-\end{quote}
-\end{framed}
 

### 2016-12-09 dissertation/introduction.tex

@@
 dissertation. SPARQL is an extensive, standardised query language with
-both read and write operations. The syntax is near Turtle, but as the
-above triple patterns, introduces variables. A set of triple patterns
+both read and write operations. The syntax is near Turtle but as the
+above triple pattern shows, introduces variables. A set of triple patterns
 may be used to create a conjunctive query, and along with a filter
@@
 
-It is difficult to many developers to interact with RDF, and
+Many developers find it difficult to interact with RDF, and
 contributions to alleviate that are summarised in
